### 2017-03-23 realtime-filter/realtime-filter.tex
Outline of a paper ...

\documentclass[draft,a4paper,onecolumn]{IEEEtran}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
\let\hlipl\hlkwb

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}

\title{Modelling a Public Transport System in Real Time using a Particle Filter}
\author{Tom~Elliott}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

\maketitle

\begin{abstract}
  Model all vehicles in the public transport system in real time.
  Particle filter model for each vehicle (bus),
  using observations of position (GPS coordinates) to infer distance into trip
  and velocity.
\end{abstract}

\begin{IEEEkeywords}
  Particle filter, transit, real-time
\end{IEEEkeywords}

\section{Introduction}
\label{sec:intro}

Basically talk about how the Kalman filter is good, fast, but not as reliable as we want.
Some faults, namely assumptions of normality, symmetry, and that we have to transform observations
(rather than transform parameters to match observations).
Particle filter has been used, and allow us to get around these problems.
Some examples of particle filter usage \ldots

\section{Particle Filtering}
\label{sec:pf}

Theory etc of how the particle filter works.

\section{GTFS Data and Road Segmentation}
\label{sec:gtfs}

Describe the data (GTFS static and real-time). (Maybe this go in section~\ref{sec:intro}?).

And the idea behind \emph{segmenting} bus routes between intersections.

\section{Transit Vehicle Transition Function}
\label{sec:transition}

The logic behind the transition function specific for a transit vehicle
(in particular, a bus).

Define all parameters, priors, etc.

\section{Likelihood Function}
\label{sec:likelihood}

The somewhat complex likelihood function.
Basic - distance from observation, \emph{or} proximity to stop/intersection.

More complex - include stop time updates.

\section{Implementation and Results}
\label{sec:results}

Implementation in \texttt{C++} and running in real-time.
Some pretty pictures of the results of the particle filter? 
How well it performs, etc.

\section{Discusion and Future Work}
\label{sec:discussion}

Talk about the good and the bad, and how we hope to use the particle filter in future to make predictions.

\end{document}

### 2017-04-04 realtime-filter/realtime-filter.Rnw
add some text...

--- 2017-03-23 realtime-filter/realtime-filter.tex
@@
-\documentclass[draft,a4paper,onecolumn]{IEEEtran}\usepackage[]{graphicx}\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-\let\hlipl\hlkwb
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
+\documentclass{IEEEtran}
+%\documentclass[draft,a4paper,onecolumn]{IEEEtran}
 
@@
 \author{Tom~Elliott}
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+
 \begin{document}
@@
 
-\section{Particle Filtering}
-\label{sec:pf}
-
 Theory etc of how the particle filter works.
@@
 
-Describe the data (GTFS static and real-time). (Maybe this go in section~\ref{sec:intro}?).
+Our work makes use of publically available GTFS data,
+which consists of the static component (timetables, shapes, etc),
+and the real-time component (vehicle positions and stop time updates).
+In its raw form, we are able to obtain the data,
+model an individual bus, and determine its (unmeasurable) state
+(see section~\ref{sec:transition}).
 
-And the idea behind \emph{segmenting} bus routes between intersections.
+However, in order to model the network as a whole, 
+we need to make some modifications to the data to enable us to 
+separate vehicles from routes;
+current methodologies, such as those by \ldots,
+only use trips belonging to the same route.
+In most networks, however,
+there is a degree of overlap between routes,
+and it is our intention to not only detect these overlaps,
+but make full use of them.
+
+We therefore need to be able to break routes into a series
+of connecting segments, 
+which can then be common between multiple different routes.
+The simplest way to find all points at which any two routes 
+may converge or diverge is at intersections.
+Presently, we are using traffic light positions 
+(easily obtained from OSM)
+as an approximation for all intersections;
+thus, there are some overlapping segments.
 

### 2017-04-04 realtime-filter/realtime-filter.tex
add some text...

@@
 \author{Tom~Elliott}
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+
+\usepackage{Sweave}
 \begin{document}

### 2017-04-11 realtime-filter/realtime-filter.Rnw
a few wordings

@@
-\documentclass{IEEEtran}
-%\documentclass[draft,a4paper,onecolumn]{IEEEtran}
+%\documentclass{IEEEtran}
+\documentclass[draft,a4paper,onecolumn]{IEEEtran}
 
@@
 
-\title{Modelling a Public Transport System in Real Time using a Particle Filter}
+\title{Modelling the Real Time State of a Public Transport Road Network}
 \author{Tom~Elliott}
@@
 \begin{IEEEkeywords}
-  Particle filter, transit, real-time
+  Public transport, real-time, particle filter, kalman filter
 \end{IEEEkeywords}
@@
 
-Basically talk about how the Kalman filter is good, fast, but not as reliable as we want.
-Some faults, namely assumptions of normality, symmetry, and that we have to transform observations
-(rather than transform parameters to match observations).
-Particle filter has been used, and allow us to get around these problems.
-Some examples of particle filter usage \ldots
+Modelling busses in real-time has been done for a while now.
+Modelling the underlying state of the road network in order to improve 
+travel time predictions has not been explored \ldots
+Some looking into complex models for ``future'' travel times \cite{Julio2016}.
 
-Theory etc of how the particle filter works.
+The basic idea:
+GPS position $\rightarrow$ 
+vehicle state (distance, segment, speed) $\rightarrow$
+road segment state (speed) $\rightarrow$
+travel time predictions
 
-\section{GTFS Data and Road Segmentation}
+\section{Data Preparation: Road Segmentation}
 \label{sec:gtfs}
+
+Real-time data can be used as-is.
+
+However, the shape data needs modification to become composed of \emph{segments},
+which are shared between multiple (unrelated) routes.
 
@@
 
+\bibliographystyle{IEEEtran}
+\bibliography{IEEEabrv,../reflist.bib}
+
 \end{document}

### 2017-04-11 realtime-filter/realtime-filter.tex
a few wordings

--- 2017-03-23 realtime-filter/realtime-filter.tex
@@
+%\documentclass{IEEEtran}
 \documentclass[draft,a4paper,onecolumn]{IEEEtran}\usepackage[]{graphicx}\usepackage[]{color}
@@
 
-Basically talk about how the Kalman filter is good, fast, but not as reliable as we want.
-Some faults, namely assumptions of normality, symmetry, and that we have to transform observations
-(rather than transform parameters to match observations).
-Particle filter has been used, and allow us to get around these problems.
-Some examples of particle filter usage \ldots
+Modelling busses in real-time has been done for a while now.
+Modelling the underlying state of the road network in order to improve 
+travel time predictions has not been explored \ldots
+Some looking into complex models for ``future'' travel times \cite{Julio2016}.
 
-\section{Particle Filtering}
-\label{sec:pf}
+The basic idea:
+GPS position $\rightarrow$ 
+vehicle state (distance, segment, speed) $\rightarrow$
+road segment state (speed) $\rightarrow$
+travel time predictions
 
-Theory etc of how the particle filter works.
-
-\section{GTFS Data and Road Segmentation}
+\section{Data Preparation: Road Segmentation}
 \label{sec:gtfs}
 
-Describe the data (GTFS static and real-time). (Maybe this go in section~\ref{sec:intro}?).
+Real-time data can be used as-is.
 
-And the idea behind \emph{segmenting} bus routes between intersections.
+However, the shape data needs modification to become composed of \emph{segments},
+which are shared between multiple (unrelated) routes.
+
+Our work makes use of publically available GTFS data,
+which consists of the static component (timetables, shapes, etc),
+and the real-time component (vehicle positions and stop time updates).
+In its raw form, we are able to obtain the data,
+model an individual bus, and determine its (unmeasurable) state
+(see section~\ref{sec:transition}).
+
+However, in order to model the network as a whole, 
+we need to make some modifications to the data to enable us to 
+separate vehicles from routes;
+current methodologies, such as those by \ldots,
+only use trips belonging to the same route.
+In most networks, however,
+there is a degree of overlap between routes,
+and it is our intention to not only detect these overlaps,
+but make full use of them.
+
+We therefore need to be able to break routes into a series
+of connecting segments, 
+which can then be common between multiple different routes.
+The simplest way to find all points at which any two routes 
+may converge or diverge is at intersections.
+Presently, we are using traffic light positions 
+(easily obtained from OSM)
+as an approximation for all intersections;
+thus, there are some overlapping segments.
 

### 2017-04-12 glossary.tex
write some things

\usepackage[acronym]{glossaries}
\newacronym[shortplural={ATIS}]{atis}{ATIS}{advanced traveller information system}
\newacronym{avl}{AVL}{automatic vehicle location}
\newacronym{apc}{APC}{automatic passenger counter}
\newacronym{rti}{RTI}{real-time information}
\newacronym{gps}{GPS}{Global Positioning System}
\newacronym{api}{API}{application programming interface}
\newacronym{gtfs}{GTFS}{general transit feed specification}
\newacronym{knn}{KNN}{$k$-nearest neighbour}
\newacronym{ann}{ANN}{artificial neural networks}
\newacronym{svm}{SVM}{support vector machines}
\newacronym{kf}{KF}{Kalman filter}
\newacronym{pf}{PF}{particle filter}
\newacronym{mcmc}{MCMC}{Markov chain Monte Carlo}
\newacronym[longplural={expected times of arrival}]{eta}{ETA}{expected time of arrival}
\newacronym{at}{AT}{Auckland Transport}
\newacronym{pt}{PT}{public transport}
\newacronym{ui}{UI}{user interface}

### 2017-04-12 realtime-filter/realtime-filter.Rnw
write some things

@@
-%\documentclass{IEEEtran}
-\documentclass[draft,a4paper,onecolumn]{IEEEtran}
+\documentclass{IEEEtran}
+%\documentclass[draftcls,a4paper,onecolumn]{IEEEtran}
 
@@
 \usepackage{bm}
+\usepackage{bbm}
+\usepackage{subfig}
+
+\input{../symbols.tex}
+\input{../glossary.tex}
 
@@
   and velocity.
+  Then use particle filter estimates of vehicle speed to estimate bus speed along
+  road segments, independently of route.
+
 \end{abstract}
@@
 
-\section{Data Preparation: Road Segmentation}
-\label{sec:gtfs}
+\section{The Data: GTFS database and route segmentation}
+\label{sec:data}
 
-Real-time data can be used as-is.
+Transit data has become increasingly accessible over the year.
+Attibuted mostly to Google's \gls{gtfs}.
+Two components to \gls{gtfs}: static and real-time.
 
-However, the shape data needs modification to become composed of \emph{segments},
-which are shared between multiple (unrelated) routes.
+Static contains information about routes, trips (one instance of a route at a scheduled time of day),
+stop positions, stop times, route shapes.
 
-Our work makes use of publically available GTFS data,
-which consists of the static component (timetables, shapes, etc),
-and the real-time component (vehicle positions and stop time updates).
-In its raw form, we are able to obtain the data,
-model an individual bus, and determine its (unmeasurable) state
-(see section~\ref{sec:transition}).
+Real-time provides vehicle positions (as GPS coordinates),
+and trip updates (including stop time updates, i.e., arrival and departure times/delays).
+Updated typically every 30~seconds or so.
 
-However, in order to model the network as a whole, 
-we need to make some modifications to the data to enable us to 
-separate vehicles from routes;
-current methodologies, such as those by \ldots,
-only use trips belonging to the same route.
-In most networks, however,
-there is a degree of overlap between routes,
-and it is our intention to not only detect these overlaps,
-but make full use of them.
+Modelling a single trip or route is fairly simple to do with the raw \gls{gtfs} data.
+However, difficult to combine information from different routes traveling along the same road segments.
 
-We therefore need to be able to break routes into a series
-of connecting segments, 
-which can then be common between multiple different routes.
-The simplest way to find all points at which any two routes 
-may converge or diverge is at intersections.
-Presently, we are using traffic light positions 
-(easily obtained from OSM)
-as an approximation for all intersections;
-thus, there are some overlapping segments.
+We define a \emph{road segment} as a stretch of road used in a transit route between two intersections,
+\emph{or} between a start/end stop of a route and an intersection.
 
-\section{Transit Vehicle Transition Function}
-\label{sec:transition}
+Do this by identifying all intersections
+(approximated by all traffic lights for now, since that's the data we have)
+and segment routes at these points.
+Modify \gls{gtfs} data so \emph{shapes} table becomes
+\texttt{shapes}, \texttt{segments}, and \texttt{segment\_shapes}.
 
-The logic behind the transition function specific for a transit vehicle
-(in particular, a bus).
+Thus as we add new routes, we don't need to create new segments
+(except at start/end\ldots see section~\ref{sec:discussion}).
 
-Define all parameters, priors, etc.
+<<route_segmentation,echo=FALSE,out.width='0.5\\linewidth',fig.cap="Simplified unidirectional road network diagram, with (a) overlapping routes, and (b) the segments obtained by splitting the routes at intersections.
+Route 1 $ = \\{r_1, r_2, r_3\\}$, route 2 $=\\{r_4,r_2,r_5,r_6\\}$, and route 3 $=\\{r_7,r_5,r_8,r_9\\}$.",fig.scap="Simplied diagram of overlapping routes",fig.subcap=c("Routes","Segments")>>=
+par(mar = rep(2.1, 4))
+e <- 0.01
+r1 <- rbind(c(0.5, 1), c(1 - e, 1), c(1 - e, 2), c(1 - e, 2.5))
+r2 <- rbind(c(1 + e, 0.5), c(1 + e, 1), c(1 + e, 2 - e), c(2, 2 - e), c(2.5, 2 - e))
+r3 <- rbind(c(0.5, 2 + e), c(1, 2 + e), c(2 + e, 2 + e), c(2 + e, 1), c(2 + e, 0.5))
+grid <- rbind(c(1, 1), c(1, 2), c(2, 1), c(2, 2))
 
-\section{Likelihood Function}
-\label{sec:likelihood}
+plot(grid, xlim = c(0.5, 2.5), ylim = c(0.5, 3), type = "n",
+     xaxs = "i", yaxs = "i", bty = "n", xaxt = "n", yaxt = "n",
+     xlab = "", ylab = "", pch = 19, asp = 1)
+pal <- RColorBrewer::brewer.pal(3, "Dark2")
+arrows(1-e, 2, 1-e, 2.5, lwd = 5, col = pal[1], length = 0.15)
+arrows(2, 2-e, 2.5, 2-e, lwd = 5, col = pal[2], length = 0.15)
+arrows(2+e, 1, 2+e, 0.5, lwd = 5, col = pal[3], length = 0.15)
+lines(r1, col = pal[1], lwd = 5)
+lines(r2, col = pal[2], lwd = 5)
+lines(r3, col = pal[3], lwd = 5)
+legend("topright", legend = paste("Route", 1:3),
+       col = pal, bty = "n", lwd = 5, cex = 2)
 
-The somewhat complex likelihood function.
-Basic - distance from observation, \emph{or} proximity to stop/intersection.
+plot(grid, xlim = c(0.5, 2.5), ylim = c(0.5, 3), type = "n",
+     xaxs = "i", yaxs = "i", bty = "n", xaxt = "n", yaxt = "n",
+     xlab = "", ylab = "", pch = 19, asp = 1)
+ea <- 0.06
+arrows(c(0.5, 1, 1,  
+         1, 1+ea, 2+ea,  
+         0.5, 2, 2),
+       c(1, 1+ea, 2+ea,  
+         0.5, 2, 2,  
+         2, 2-ea, 1-ea),
+       c(1-ea, 1, 1,  
+         1, 2-ea, 2.5,  
+         1-ea, 2, 2),
+       c(1, 2-ea, 2.5,  
+         1-ea, 2, 2,  
+         2, 1+ea, 0.5), 
+       code = 2, length = 0.15,
+       lwd = c(2, 4, 2,  2, 4, 2,  2, 2))
+#points(grid, pch = 19, cex = 4, col = "white")
+points(grid, pch = 19, cex = 2, col = "black")
 
-More complex - include stop time updates.
+tx <- rbind(c(0.7, 1.1), c(0.9, 1.5), c(0.9, 2.3),
+            c(1.1, 0.7), c(1.5, 2.1), c(2.3, 2.1),
+            c(0.7, 1.9), c(2.1, 1.5), c(2.1, 0.7))
+for (i in 1:nrow(tx)) {
+    text(tx[i,1], tx[i,2], 
+         substitute(r[x], list(x = i)),
+         cex = 3, family="serif", font=3)
+}
 
-\section{Implementation and Results}
+@
+
+\section{Real-time vehicle model: particle filter}
+\label{sec:pf}
+
+Need to estimate vehicle state at time $t_k$, $\bX_k = [d_k, v_k, \ldots]$, 
+from observations, $\bY_k = [\phi_k, \lambda_k]$.
+Use a transition function to predict,
+likelihood function to weight and update (via weighted resample).
+
+\subsection{Predict: transition function}
+The idea behind using particle filter (as opposed to other models)
+is that it is well suited modelling busses \cite{Hans2015},
+as it has good coverage of the plausible locations and thus overcomes
+many problems associated with modelling transit vehicles,
+such as multimodality (due to loops, traffic lights, and stops).
+
+Also easy to describe the model;
+only have to deal with an individual object (particle) which follows laws of physics.
+Rather than dealing with a state \emph{distribution}.
+
+Downside is computational demand.
+
+Algorithm~1 describes the particle filter transition algorithm that
+describes the motion of a bus between observations.
+
+\subsection{Update: weighted resampling}
+
+Several pieces of data: position, arrival time, and departure time.
+
+For position, need to allow bus to be ``at stop'' (but not exactly),
+or ``queued up at an intersection''.
+
+For arrival/departure times, when the data is there, it's fairly easy;
+however, if data is \emph{missing}, we don't know whether bus hasn't arrived yet,
+or if bus didn't report arrival.
+Also if bus doesn't report both arrival time and departure time,
+don't know if it simply didn't stop (i.e., arrival = departure),
+or if we missed one of them.
+
+So, depending on the data, we apply different likelihood functions.
+
+\subsubsection[L1]{$L_1$: not at a stop or intersection}
+
+In situations where the bus is neither at a bus stop nor at an intersection,
+the likelihood is simply determined from the distance between the particle and 
+vehicle's reported position.
+The particle filter framework makes this easy,
+as each particle has a single value of distance into trip, $d_k^{(i)}$,
+with which we can compute the latitude and longitude values based on 
+the routes shape information, $z_k^{(i)} = h(d_k{(i)})$.
+
+\subsubsection[L1b]{$L_1'$: at stop}
+
+If the particle is considered to be at a stop, then we apply the following likelihood.
+Define a parameter $\varepsilon_S$ as a boundary around the stop;
+if observation is within that boundary likelihood is 1, else 0.
+
+\subsubsection[L1c]{$L_1''$: at intersection}
+
+If the particle is considered to be at an intersection, we apply the following likelihood.
+Similar to $L_1'$, however need to check observation is heading into the intersection
+(and not out of it) by using two boundaries around the intersection and a point $\varepsilon_I$ away.
+
+Then 
+\begin{equation}
+  L_1(\bX_k^{(i)} | \bY_k) =
+  \begin{cases}
+    \frac{1}{2\pi\sigma_y} e^{\frac{\|g(h(d_k^{(i)})|\bY_k)\|^2}{2\sigma_y^2}} & \text{ if particle neither at stop nor intersection } \\
+    \frac{1}{c} \mathbbm{1}_{d(h(S_{s_k{[i]}}^d), Y_k) < \varepsilon_S}    & \text{ if particle at stop } \\
+    \frac{1}{c} \mathbbm{1}_{d(h(R_{r_k{[i]}}^d), Y_k) < \varepsilon_I \cap d(h(R_{r_k{[i]}}^d-\varepsilon_I), Y_k) < \varepsilon_I}    & \text{ if particle at intersection }
+  \end{cases}
+\end{equation}
+
+\subsubsection[L2]{$L_2$: not at a stop}
+
+If the particle is not at a stop, we determine a likelihood for the arrival and departure
+times at the last stop, if they are available.
+
+First check that the most recent stop time update agrees with the particle;
+if the stop time update is for stop $j$, and the particles state has $s_k^{(i)} < j$,
+then the particle is not plausible;
+otherwise use normal distribution for arrival and departure times,
+or whichever are available.
+
+\subsubsection[L2b]{$L_2'$: at a stop}
+
+If particle is at a stop, again determine whether or not it agrees with data.
+If it does, then check data doesn't imply particle has departed,
+then just use likelihood on arrival time.
+
+\section{Real-time transport road network model: Kalman filter}
+\label{sec:kf}
+
+Now that each vehicle is modelled by a sample of particles,
+we can use the speed estimates of busses along segments to estimate
+the speed along them, regardless of the route.
+In future we can add correlation structure between adjacent segments.
+
+Use Kalman filter because we have one parameter, speed, 
+which we expect to be unimodal and approximately Gaussian
+(some busses travel faster or slower).
+
+Use weighted speed estimates as data,
+and weighted variance as measurement error.
+
+\section{Arrival time prediction: application to a small subset of routes}
 \label{sec:results}
 
-Implementation in \texttt{C++} and running in real-time.
-Some pretty pictures of the results of the particle filter? 
-How well it performs, etc.
+Compare several methods:
+current delay, current speed, and Kalman filter speed estimates.
+Also include dwell distibutions at stops and intersections.
 
@@
 
-Talk about the good and the bad, and how we hope to use the particle filter in future to make predictions.
+\begin{itemize}
+\item removing ``partial'' segments at start/end of routes
+\end{itemize}
 

### 2017-04-12 realtime-filter/realtime-filter.tex
write some things

@@
-%\documentclass{IEEEtran}
-\documentclass[draft,a4paper,onecolumn]{IEEEtran}\usepackage[]{graphicx}\usepackage[]{color}
+\documentclass{IEEEtran}\usepackage[]{graphicx}\usepackage[]{color}
 %% maxwidth is the original width if it is less than linewidth
@@
 \usepackage{alltt}
+%\documentclass[draftcls,a4paper,onecolumn]{IEEEtran}
 
@@
 
-\section{Data Preparation: Road Segmentation}
-\label{sec:gtfs}
+\section{The Data: GTFS database and route segmentation}
+\label{sec:data}
 
-Real-time data can be used as-is.
+Transit data has become increasingly accessible over the year.
+Attibuted mostly to Google's \gls{gtfs}.
+Two components to \gls{gtfs}: static and real-time.
 
-However, the shape data needs modification to become composed of \emph{segments},
-which are shared between multiple (unrelated) routes.
+Static contains information about routes, trips (one instance of a route at a scheduled time of day),
+stop positions, stop times, route shapes.
 
-Our work makes use of publically available GTFS data,
-which consists of the static component (timetables, shapes, etc),
-and the real-time component (vehicle positions and stop time updates).
-In its raw form, we are able to obtain the data,
-model an individual bus, and determine its (unmeasurable) state
-(see section~\ref{sec:transition}).
+Real-time provides vehicle positions (as GPS coordinates),
+and trip updates (including stop time updates, i.e., arrival and departure times/delays).
+Updated typically every 30~seconds or so.
 
-However, in order to model the network as a whole, 
-we need to make some modifications to the data to enable us to 
-separate vehicles from routes;
-current methodologies, such as those by \ldots,
-only use trips belonging to the same route.
-In most networks, however,
-there is a degree of overlap between routes,
-and it is our intention to not only detect these overlaps,
-but make full use of them.
+Modelling a single trip or route is fairly simple to do with the raw \gls{gtfs} data.
+However, difficult to combine information from different routes traveling along the same road segments.
 
-We therefore need to be able to break routes into a series
-of connecting segments, 
-which can then be common between multiple different routes.
-The simplest way to find all points at which any two routes 
-may converge or diverge is at intersections.
-Presently, we are using traffic light positions 
-(easily obtained from OSM)
-as an approximation for all intersections;
-thus, there are some overlapping segments.
+We define a \emph{road segment} as a stretch of road used in a transit route between two intersections,
+\emph{or} between a start/end stop of a route and an intersection.
 
-\section{Transit Vehicle Transition Function}
-\label{sec:transition}
+Do this by identifying all intersections
+(approximated by all traffic lights for now, since that's the data we have)
+and segment routes at these points.
+Modify \gls{gtfs} data so \emph{shapes} table becomes
+\texttt{shapes}, \texttt{segments}, and \texttt{segment\_shapes}.
 
-The logic behind the transition function specific for a transit vehicle
-(in particular, a bus).
+Thus as we add new routes, we don't need to create new segments
+(except at start/end\ldots see section~\ref{sec:discussion}).
 
-Define all parameters, priors, etc.
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\subfloat[Routes\label{fig:route_segmentation1}]{\includegraphics[width=0.5\linewidth]{figure/route_segmentation-1} }
+\subfloat[Segments\label{fig:route_segmentation2}]{\includegraphics[width=0.5\linewidth]{figure/route_segmentation-2} }\caption[Simplied diagram of overlapping routes]{Simplified unidirectional road network diagram, with (a) overlapping routes, and (b) the segments obtained by splitting the routes at intersections.
+Route 1 $ = \{r_1, r_2, r_3\}$, route 2 $=\{r_4,r_2,r_5,r_6\}$, and route 3 $=\{r_7,r_5,r_8,r_9\}$.}\label{fig:route_segmentation}
+\end{figure}
 
-\section{Likelihood Function}
-\label{sec:likelihood}
+\end{knitrout}
 
-The somewhat complex likelihood function.
-Basic - distance from observation, \emph{or} proximity to stop/intersection.
+\section{Real-time vehicle model: particle filter}
+\label{sec:pf}
 
-More complex - include stop time updates.
+Need to estimate vehicle state at time $t_k$, $\bX_k = [d_k, v_k, \ldots]$, 
+from observations, $\bY_k = [\phi_k, \lambda_k]$.
+Use a transition function to predict,
+likelihood function to weight and update (via weighted resample).
 
-\section{Implementation and Results}
+\subsection{Predict: transition function}
+The idea behind using particle filter (as opposed to other models)
+is that it is well suited modelling busses \cite{Hans2015},
+as it has good coverage of the plausible locations and thus overcomes
+many problems associated with modelling transit vehicles,
+such as multimodality (due to loops, traffic lights, and stops).
+
+Also easy to describe the model;
+only have to deal with an individual object (particle) which follows laws of physics.
+Rather than dealing with a state \emph{distribution}.
+
+Downside is computational demand.
+
+Algorithm~1 describes the particle filter transition algorithm that
+describes the motion of a bus between observations.
+
+\subsection{Update: weighted resampling}
+
+Several pieces of data: position, arrival time, and departure time.
+
+For position, need to allow bus to be ``at stop'' (but not exactly),
+or ``queued up at an intersection''.
+
+For arrival/departure times, when the data is there, it's fairly easy;
+however, if data is \emph{missing}, we don't know whether bus hasn't arrived yet,
+or if bus didn't report arrival.
+Also if bus doesn't report both arrival time and departure time,
+don't know if it simply didn't stop (i.e., arrival = departure),
+or if we missed one of them.
+
+So, depending on the data, we apply different likelihood functions.
+
+\subsubsection[L1]{$L_1$: not at a stop or intersection}
+
+In situations where the bus is neither at a bus stop nor at an intersection,
+the likelihood is simply determined from the distance between the particle and 
+vehicle's reported position.
+The particle filter framework makes this easy,
+as each particle has a single value of distance into trip, $d_k^{(i)}$,
+with which we can compute the latitude and longitude values based on 
+the routes shape information, $z_k^{(i)} = h(d_k{(i)})$.
+
+\subsubsection[L1b]{$L_1'$: at stop}
+
+If the particle is considered to be at a stop, then we apply the following likelihood.
+Define a parameter $\varepsilon_S$ as a boundary around the stop;
+if observation is within that boundary likelihood is 1, else 0.
+
+\subsubsection[L1c]{$L_1''$: at intersection}
+
+If the particle is considered to be at an intersection, we apply the following likelihood.
+Similar to $L_1'$, however need to check observation is heading into the intersection
+(and not out of it) by using two boundaries around the intersection and a point $\varepsilon_I$ away.
+
+Then 
+\begin{equation}
+  L_1(\bX_k^{(i)} | \bY_k) =
+  \begin{cases}
+    \frac{1}{2\pi\sigma_y} e^{\frac{\|g(h(d_k^{(i)})|\bY_k)\|^2}{2\sigma_y^2}} & \text{ if particle neither at stop nor intersection } \\
+    \frac{1}{c} \mathbbm{1}_{d(h(S_{s_k{[i]}}^d), Y_k) < \varepsilon_S}    & \text{ if particle at stop } \\
+    \frac{1}{c} \mathbbm{1}_{d(h(R_{r_k{[i]}}^d), Y_k) < \varepsilon_I \cap d(h(R_{r_k{[i]}}^d-\varepsilon_I), Y_k) < \varepsilon_I}    & \text{ if particle at intersection }
+  \end{cases}
+\end{equation}
+
+\subsubsection[L2]{$L_2$: not at a stop}
+
+If the particle is not at a stop, we determine a likelihood for the arrival and departure
+times at the last stop, if they are available.
+
+First check that the most recent stop time update agrees with the particle;
+if the stop time update is for stop $j$, and the particles state has $s_k^{(i)} < j$,
+then the particle is not plausible;
+otherwise use normal distribution for arrival and departure times,
+or whichever are available.
+
+\subsubsection[L2b]{$L_2'$: at a stop}
+
+If particle is at a stop, again determine whether or not it agrees with data.
+If it does, then check data doesn't imply particle has departed,
+then just use likelihood on arrival time.
+
+\section{Real-time transport road network model: Kalman filter}
+\label{sec:kf}
+
+Now that each vehicle is modelled by a sample of particles,
+we can use the speed estimates of busses along segments to estimate
+the speed along them, regardless of the route.
+In future we can add correlation structure between adjacent segments.
+
+Use Kalman filter because we have one parameter, speed, 
+which we expect to be unimodal and approximately Gaussian
+(some busses travel faster or slower).
+
+Use weighted speed estimates as data,
+and weighted variance as measurement error.
+
+\section{Arrival time prediction: application to a small subset of routes}
 \label{sec:results}
 
-Implementation in \texttt{C++} and running in real-time.
-Some pretty pictures of the results of the particle filter? 
-How well it performs, etc.
+Compare several methods:
+current delay, current speed, and Kalman filter speed estimates.
+Also include dwell distibutions at stops and intersections.
 

### 2017-04-19 realtime-filter/realtime-filter.Rnw
kf eqns

@@
-\documentclass{IEEEtran}
-%\documentclass[draftcls,a4paper,onecolumn]{IEEEtran}
+%\documentclass{IEEEtran}
+\documentclass[draftcls,a4paper,onecolumn]{IEEEtran}
 
@@
 
+We run the Kalman filter every time we run the particle filter,
+since it has a low cost, allowing the most recent data to be incorporated.
+We have $N$ weighted particle estimates of vehicle state for each bus,
+with which we can compute the mean distance into trip and velocity for bus $j$,
+\begin{equation}
+  \label{eq:weighted_mean}
+  \bar d_j = \sum_{i=1}^N w_j^{(i)} d_j^{(i)} \quad\text{and}\quad
+  \bar v_j = \sum_{i=1}^N w_j^{(i)} v_j^{(i)},
+\end{equation}
+as well as the weighted sample variance \cite{cn},
+\begin{equation}
+  \label{eq:weighted_variance}
+  \bar S_j^2 = \frac{\sum_{i=1}^N w_j^{(i)} (v_j^{(i)} - \bar v_j)^2}{1 - \sum_{i=1}^N (w_j^{(i)})^2},
+\end{equation}
+since $\sum_{i=1}^N w_j^{(i)} = 1\quad \forall j$.
+
+Using $\bar d_j$, we can find the segment id corresponding to each busses most probable location,
+and then combine all vehicle speed estimates for each segment.
+
+In situations where multiple vehicles are in the same segment $\ell$, 
+we can simply combine the estimates, using $M_\ell = $ number of vehicles in segment $\ell$ 
+\cite{cn}.
+\begin{equation}
+  \label{eq:kf_data_estimates}
+  \bar z_\ell = \frac{1}{M_\ell} \sum_{j = 1}^{M_\ell} \bar v_j \quad\text{and}\quad
+  \hat \bR_{\ell,\ell}^2 = \frac{1}{M_\ell} \sum_{j=1}^{M_\ell} \left(\bar v_j^2 + \bar S_j^2\right) - \hat v_\ell^2
+\end{equation}
+
+Next is just the simple matter of applying the Kalman filter algorithm to the transport road network state,
+using the data $\bz_k$ as the observation and $\bR_k$,
+as the measurement error matrix.
+
 \section{Arrival time prediction: application to a small subset of routes}

### 2017-04-19 realtime-filter/realtime-filter.tex
kf eqns

@@
-\documentclass{IEEEtran}\usepackage[]{graphicx}\usepackage[]{color}
+%\documentclass{IEEEtran}
+\documentclass[draftcls,a4paper,onecolumn]{IEEEtran}\usepackage[]{graphicx}\usepackage[]{color}
 %% maxwidth is the original width if it is less than linewidth
@@
 \usepackage{alltt}
-%\documentclass[draftcls,a4paper,onecolumn]{IEEEtran}
 

### 2017-04-21 realtime-filter/realtime-filter.Rnw
Type up eqns

@@
-%\documentclass{IEEEtran}
-\documentclass[draftcls,a4paper,onecolumn]{IEEEtran}
+\documentclass[9pt]{IEEEtran}
+%\documentclass[draftcls,a4paper,onecolumn]{IEEEtran}
 
@@
 
-So, depending on the data, we apply different likelihood functions.
+So, depending on the particles' states, we apply different likelihood functions.
 
-\subsubsection[L1]{$L_1$: not at a stop or intersection}
-
+\begin{enumerate}
+%\subsubsection[L1]{$L_1$: not at a stop or intersection}
+\item
 In situations where the bus is neither at a bus stop nor at an intersection,
@@
 
-\subsubsection[L1b]{$L_1'$: at stop}
+\begin{equation}
+  \label{eq:l1}
+  \log L_1 (\bX_k^{(i)} | \bY_k) = - \log(2\pi\sigma_y) - \frac{\|g(h(d_k^{(i)})|\bY_k)\|^2}{2\sigma_y^2}
+\end{equation}
 
+%\subsubsection[L1b]{$L_1'$: at stop}
+\item
 If the particle is considered to be at a stop, then we apply the following likelihood.
@@
 if observation is within that boundary likelihood is 1, else 0.
+Let $B_s$ be a circle with radius $\varepsilon_S$ and centered at stop $j$, then
 
-\subsubsection[L1c]{$L_1''$: at intersection}
+\begin{equation}
+  \label{eq:l1b}
+  \log L_1(\bX_k^{(i)} | \bY_k)=
+  \begin{cases}
+    - \log(\pi \varepsilon_S^2) & \bY_k \in B_s \\
+    - \infty & \text{otherwise}
+  \end{cases}
+\end{equation}
 
+%\subsubsection[L1c]{$L_1''$: at intersection}
+\item
 If the particle is considered to be at an intersection, we apply the following likelihood.
@@
 (and not out of it) by using two boundaries around the intersection and a point $\varepsilon_I$ away.
+Let $B_I^1$ and $B_I^2$ be two circles with radii $\varepsilon_I$,
+the first centered on the intersection and the second centered $\varepsilon_I$ meters before it, then
 
-Then 
 \begin{equation}
-  L_1(\bX_k^{(i)} | \bY_k) =
+  \label{eq:l1c}
+  \log L_1 (\bX_k^{(i)} | \bY_k) =
   \begin{cases}
-    \frac{1}{2\pi\sigma_y} e^{\frac{\|g(h(d_k^{(i)})|\bY_k)\|^2}{2\sigma_y^2}} & \text{ if particle neither at stop nor intersection } \\
-    \frac{1}{c} \mathbbm{1}_{d(h(S_{s_k{[i]}}^d), Y_k) < \varepsilon_S}    & \text{ if particle at stop } \\
-    \frac{1}{c} \mathbbm{1}_{d(h(R_{r_k{[i]}}^d), Y_k) < \varepsilon_I \cap d(h(R_{r_k{[i]}}^d-\varepsilon_I), Y_k) < \varepsilon_I}    & \text{ if particle at intersection }
+    - \log(A) & \bY_k \in B_I^1 \cap B_I^2 \\
+    - \infty & \text{otherwise}
+  \end{cases}
+\end{equation}
+where $A = 2\varepsilon_I^2 \left( \cos^{-1}(\frac{1}{2}) - \frac{\sqrt{3}}{4} \right)$,
+the area of $B_I^1\cap B_I^2$.
+
+\end{enumerate}
+
+%\subsubsection[L2]{$L_2$: not at a stop}
+
+The second part of the likelihood involves the stop time update,
+and how the likelihood is formed depends on what data is available,
+where $\bT_k = \left[ j, T^a_j , T^d_j \right]^T$ 
+are the stop index, arrival and departure times,
+respectively, either of which can be NA.
+
+\begin{enumerate}
+\item 
+  If both $T^a_j$ and $T^d_j$ are NA, 
+  then there is no additional likelihood;
+  $ \log L_2  (\bX_k^{(i)} | \bT_k) = 0 $.
+
+\item
+  If departure time, $T^d_j$, is available only, for stop $j$, then we
+  know the bus has at least departed stop $j$; therefore,
+  \begin{equation}
+    \label{eq:l2}
+      \log L_2 (\bX_k^{(i)} | \bT_k) =
+      \begin{cases}
+        - \log(2\pi\sigma_d^2) - \frac{1}{2\sigma_d^2}(T^d_j - D_j^{(i)})^2 \\
+        \hfill\quad \text{if } s_k^{(i)} \geq j \text{ and $D_j^{(i)}$ is set} \\
+        - \infty \\
+        \hfill\quad \text{otherwise}
       \end{cases}
@@
 
-\subsubsection[L2]{$L_2$: not at a stop}
+\item
+  If arrival time, $T^a_j$, is available only, then similarly
+  \begin{equation}
+    \label{eq:l2b}
+    \log L_2  (\bX_k^{(i)} | \bT_k) =
+    \begin{cases}
+      - \log(2\pi\sigma_d^2) - \frac{(T^a_j - A_j^{(i)})^2}{2\sigma_d^2} & s_k^{(i)} \geq j \\
+      - \infty & \text{otherwise}
+    \end{cases}.
+  \end{equation}
+  Note that this does not assume the particle hasn't yet left the stop;
+  the departure time could simply be missing data.
 
-If the particle is not at a stop, we determine a likelihood for the arrival and departure
-times at the last stop, if they are available.
+\item
+  If both arrival time and departure time are available, then
+  \begin{equation}
+    \label{eq:l2c}
+    \log L_2  (\bX_k^{(i)} | \bT_k) =
+    \begin{cases}
+      \begin{array}
+        - \log(2\pi\sigma_d^2) - \frac{(T^a_j - A_j^{(i)})^2}{2\sigma_d^2} \\
+        + \log(\|\bT_k\|) - \|\bT_k\|
+        (D_k^{(i)} - A_k^{(i)})
+      \end{array} \\
+      \hfill\quad \text{if } s_k^{(i)} \geq j \text{ and $D_j^{(i)}$ is set}  \\
+      - \infty\\
+      \hfill\quad \text{otherwise}
+    \end{cases}
+  \end{equation}
+  which makes use of the observed dwell time at the stop.
 
-First check that the most recent stop time update agrees with the particle;
-if the stop time update is for stop $j$, and the particles state has $s_k^{(i)} < j$,
-then the particle is not plausible;
-otherwise use normal distribution for arrival and departure times,
-or whichever are available.
-
-\subsubsection[L2b]{$L_2'$: at a stop}
-
-If particle is at a stop, again determine whether or not it agrees with data.
-If it does, then check data doesn't imply particle has departed,
-then just use likelihood on arrival time.
+\end{enumerate}
 

### 2017-04-21 realtime-filter/realtime-filter.tex
Type up eqns

@@
-%\documentclass{IEEEtran}
-\documentclass[draftcls,a4paper,onecolumn]{IEEEtran}\usepackage[]{graphicx}\usepackage[]{color}
+\documentclass[9pt]{IEEEtran}\usepackage[]{graphicx}\usepackage[]{color}
 %% maxwidth is the original width if it is less than linewidth

### 2017-04-24 realtime-filter/realtime-filter.Rnw
writing algorithm and stuff

@@
-\documentclass[9pt]{IEEEtran}
-%\documentclass[draftcls,a4paper,onecolumn]{IEEEtran}
+%\documentclass[9pt]{IEEEtran}
+\documentclass[draftcls,a4paper,onecolumn]{IEEEtran}
 
@@
 \usepackage{subfig}
+\usepackage{algpseudocode}
+\algblockdefx[Phase]{Phase}{EndPhase}[2]{\textbf{Phase #1:} #2}{\textbf{End Phase}\\}
 
@@
 
-Need to estimate vehicle state at time $t_k$, $\bX_k = [d_k, v_k, \ldots]$, 
-from observations, $\bY_k = [\phi_k, \lambda_k]$.
-Use a transition function to predict,
-likelihood function to weight and update (via weighted resample).
+The primary goal of the particle filter is to estimate the unobservatble
+vehicle state,
+consisting of distance into trip ($d$), velocity ($v$),
+current stop sequence ($s$), 
+arrival and departure times at the last stop ($A_s$ and $D_s$),
+current segment index ($r$),
+the time spent queuing at the beginning of the current segment ($Q_r$),
+and the time at which the vehicle began traveling along the current segment ($B_r$);
+the vehicle state at time $k$ is denoted
+\begin{equation}
+  \label{eq:vehiclestate}
+  \bX_k = [d_k, v_k, s_k, A_{s_k}, D_{s_k}, r_k, Q_{r_k}, B_{r_k}]^T.
+\end{equation}
+
+The vehicle state is related to an observable state $\bY_k = [\phi_k, \lambda_k]$,
+(which contains the latitude and longitude, respectively)
+through a \emph{measurment function}, $\bY_k = h(\bX_k) = h(d_k)$,
+which depends only on the distance into trip.
+
+There is a second source of data, the \emph{stop time updates}:
+\mbox{$\bT_k = \left[ j, T^a_j , T^d_j \right]^T$}
+are the stop index, arrival and departure times,
+respectively; 
+either of the arrival or departure times can be missing.
+
+The particle filter is used to model a transit vehicle,
+so we need to know about the behaviour of these.
+The vehicle itself has a purpose of driving along an \emph{a priori} known route,
+stopping at stops with known location along the way to pick up
+and drop off passengers, if there are any.
+Between stops, vehicles travel as close to the posted speed limit as possible,
+typically only stopping at intersections (if necessary).
+Due to many conditional situations, 
+the particle filter becomes a vastly superior model when compared to traditional
+Kalman filter based methods, as it can easily encompass each of these plausible states.
+
+On approach to a stop $j$, the bus will stop with probability $\pi_j$ 
+(i.e., it is a bernoulli trial for each particle).
+If it stops, it will wait a minimum of $\gamma$ seconds while the doors are opened and closed,
+and an additional $\bar t$ seconds as passengers board and disembark.
+$\gamma$ is considered a fixed parameter, across all busses, 
+while $\bar t$ is given an exponential distribution that, for now, has a fixed mean, $\tau$
+(section~\ref{sec:discussion} discusses modifications to this).
+
+On approach to an intersection (the end of a segment),
+the behaviour is slightly different;
+the vehicle doesn't reach the intersection itself, instead it joins the queue
+(if there is one), and queues until it is able to pass through.
+Similarly to stops, currently this is just modelled using fixed parameter values;
+$\rho$ is the probability of stopping at the intersection
+(for example, green vs.\ red traffic lights),
+and $\nu$ is the average queue time.
+
+To implement the particle filter model, 
+we simply initialise $N$ particles (using superscript $^{(i)}$ to identify an individual particle)
+by sampling distance and velocity, and then determining $s_k$ and $r_k$ from the 
+vectors of stop and segment distances, $\bS^d$ and $\bR^d$, respectively.
+The remaining state parameters are NULL.
+Each particle is then passed through the \emph{transition function} independently (the prediction step),
+and then the vehicle state estimate is updated using the data by computing likelihood weights
+and resampling the particles with replacement.
+The steps are described below.
 
@@
 describes the motion of a bus between observations.
+The $k$ subscripts and $(i)$ superscripts have been dropped from the
+vehicle state parameters (e.g., $d \equiv d_k^{(i)}$).
+
+The additional parameters are the noise terms, $\sigma_v^2$ and $\sigma_r^2$,
+which represent system noise and between-segment speed variance respectively.
+The former is assumed to be small, and represents the average change in speed
+of a bus per second \emph{within a segment},
+while the latter is the variance in speed \emph{between adjacent segments},
+and allows for the bus to quickly change speed when passing through an intersection.
+
+\begin{figure}[bt]
+  \begin{algorithmic}
+    \Function{transition}{$\bX$, $\delta$, $t$}
+
+    \Phase{1}{initial wait time}
+    \State $w\gets 0$
+    \If {$A_s \neq$ NULL $\cap\ D_s =$ NULL}
+      \Comment{at stop}
+      \State $w\gets w + \max\{0, \gamma - (t - A_s)\}$
+      \Comment{min dwell time}
+      \State $\bar t \sim \mathcal{E}(\tau^{-1})$
+      \State $w\gets w + \bar t, \delta\gets \delta - w$
+      \If {$\delta > 0$}
+        \State $D_s\gets t - \delta$
+        \Comment{depart stop}
+      \EndIf
+    \ElsIf {$Q_r \neq$ NULL $\cap\ B_r =$ NULL}
+      \Comment{at intersection}
+      \State $\bar t \sim \mathcal{E}(\nu^{-1})$
+      \State $Q_r\gets Q_r + \bar t$, $w\gets w+\bar t$, $\delta\gets \delta - w$
+      \If {$\delta > 0$}
+        \State $B_s\gets t - \delta$
+        \Comment{pass through intersection}
+      \EndIf
+    \EndIf
+    \EndPhase
+
+    \Phase{2}{System Noise}
+    \State $v_k^{(i)}\sim \mathcal{N}(v, \delta\sigma_v^2)$
+    \Comment{add system noise}
+    \EndPhase
+
+    \Phase{3}{Transition Forward}
+    \State see figure~\ref{alg:transitionPhase3}
+    \EndPhase
+
+    \State \textbf{return} $\bX$
+    \EndFunction
+  \end{algorithmic}
+  \caption{Particle filter transition function. Arguments are vehicle state~($\bX$), time since previous observation~($\delta$), and time of current observation ($t$).}
+  \label{alg:transition}
+\end{figure}
+
+\begin{figure}[bt]
+  \begin{algorithmic}
+    \Phase{3}{Transition Forward}
+    \While {$\delta > 0$}
+      \If {$S^d_{s+1} \leq R^d_{r+1}$}
+        \Comment{approaching stop}
+        \State ETA  $\gets \frac{1}{v} (S^d_{s+1} - d)$
+        \If {ETA $\leq \delta$}
+        \Comment{reaches stop}
+          \State $\delta\gets \delta - $ ETA, $s \gets s + 1$
+          \State $A \gets t - \delta$, $D\gets $ null, $d\gets S^d_s$
+          \If {$s = M$}
+            \State\textbf{exit}
+            \Comment{reached last stop}
+          \EndIf
+          \State $p\gets \mathcal{B}(\pi)$, $\bar t \sim \mathcal{E}(\tau^{-1})$
+          \State dwell $\gets p(\gamma + \bar t)$, $\delta\gets \delta\ - $ dwell
+          \If {$\delta > 0$}
+            \State $D\gets t - \delta$
+            \Comment{departs stop}
+          \EndIf
+        \Else
+          \Comment{doesn't reach stop}
+          \State $d\gets d + \delta v$, $\delta \gets 0$
+        \EndIf
+      \Else
+        \Comment{approaching intersection}
+        \State ETA $\gets \frac{1}{v}(R^d_{r+1} - d)$
+        \If {ETA $\leq \delta$}
+          \Comment{reaches intersection}
+          \State $\delta\gets \delta - $ ETA, $r\gets r + 1$
+          \State $Q \gets 0$, $B\gets $ null, $d\gets R^d_r$
+          \State $p \sim \mathcal{B}(\rho)$, $\bar t \sim \mathcal{E}(\nu^{-1})$
+          \State queue $\gets p\bar t$, $\delta\gets \delta\ - $ queue
+          \State $Q\gets Q\ +$ queue
+          \If {$\delta > 0$}
+            \Comment{makes it through intersection}
+            \State $B\gets t - \delta$
+            \State $v \sim \mathcal{N}(v, \sigma_r^2)$
+          \EndIf
+        \Else
+          \Comment{doesn't reach intersection}
+          \State $d\gets d + \delta v$, $\delta \gets 0$
+        \EndIf
+      \EndIf
+    \EndWhile
+    \EndPhase
+  \end{algorithmic}
+  \caption{Transition function, phase 3: the main phase in which particles are moved forwards along the route.}
+  \label{alg:transitionPhase3}
+\end{figure}
 
@@
 with which we can compute the latitude and longitude values based on 
-the routes shape information, $z_k^{(i)} = h(d_k{(i)})$.
+the routes shape information, $z_k^{(i)} = h(d_k^{(i)})$.
+The function $g (\bZ | \bY)$ performs an \emph{equirectangular projection} \cite{Snyder1997}
+of the coordinates $\bX$, centered on $\bY$, 
+onto a flat surface so that the Euclidean distance between points is equivalent
+to the Great Circle distance between their associated coordinates.
+This allows us to use a bivariate normal likelihood on the particles' positions,
+with a GPS error of $\sigma_y^2$.
 
@@
 are the stop index, arrival and departure times,
-respectively, either of which can be NA.
+respectively, either of which can be NULL.
 
@@
 \item 
-  If both $T^a_j$ and $T^d_j$ are NA, 
-  then there is no additional likelihood;
-  $ \log L_2  (\bX_k^{(i)} | \bT_k) = 0 $.
+  If both $T^a_j$ and $T^d_j$ are NULL, 
+  then there is no change to the likelihood;
+  $ L_2  (\bX_k^{(i)} | \bT_k) = 1 $.
 
@@
     \label{eq:l2}
-      \log L_2 (\bX_k^{(i)} | \bT_k) =
+    L_2 (\bX_k^{(i)} | \bT_k) =
       \begin{cases}
-        - \log(2\pi\sigma_d^2) - \frac{1}{2\sigma_d^2}(T^d_j - D_j^{(i)})^2 \\
-        \hfill\quad \text{if } s_k^{(i)} \geq j \text{ and $D_j^{(i)}$ is set} \\
-        - \infty \\
-        \hfill\quad \text{otherwise}
+        f_N(D_j^{(i)} | T_j^d, \sigma_d^2) & s_k^{(i)} \geq j \cap D_j^{(i)} \neq \text{NULL} \\
+        0 & \text{otherwise},
       \end{cases}
   \end{equation}
+  where $f_N(x | \mu, \sigma^2)$ is the probability density of a normal distribution
+  with mean $\mu$ and variance $\sigma^2$.
 
@@
     \label{eq:l2b}
-    \log L_2  (\bX_k^{(i)} | \bT_k) =
+    L_2  (\bX_k^{(i)} | \bT_k) =
     \begin{cases}
-      - \log(2\pi\sigma_d^2) - \frac{(T^a_j - A_j^{(i)})^2}{2\sigma_d^2} & s_k^{(i)} \geq j \\
-      - \infty & \text{otherwise}
-    \end{cases}.
+      f_N(A_j^{(i)} | T_j^a, \sigma_d^2) & s_k^{(i)} \geq j \\
+      0 & \text{otherwise}.
+    \end{cases}
   \end{equation}
@@
     \label{eq:l2c}
-    \log L_2  (\bX_k^{(i)} | \bT_k) =
+    L_2  (\bX_k^{(i)} | \bT_k) =
     \begin{cases}
-      \begin{array}
-        - \log(2\pi\sigma_d^2) - \frac{(T^a_j - A_j^{(i)})^2}{2\sigma_d^2} \\
-        + \log(\|\bT_k\|) - \|\bT_k\|
-        (D_k^{(i)} - A_k^{(i)})
-      \end{array} \\
-      \hfill\quad \text{if } s_k^{(i)} \geq j \text{ and $D_j^{(i)}$ is set}  \\
-      - \infty\\
-      \hfill\quad \text{otherwise}
+      \begin{array}{l}
+        f_N(A_j^{(i)} | T_j^a, \sigma_d^2) f_E(D_k^{(i)} - A_k^{(i)} | T_k^\delta) \quad\\
+        \hfill s_k^{(i)} \geq j \cap D_j^{(i)} \neq \text{NULL} \\[0.5em]
+        0 \qquad\qquad \text{otherwise},
+      \end{array}
     \end{cases}
   \end{equation}
-  which makes use of the observed dwell time at the stop.
+  where $T_k^\delta = T_k^d - T_k^a$ is the observed dwell time at the stop,
+  and $f_E(x | \lambda)$ is the probability density of an exponential distribution 
+  with rate $\lambda$.
 
 \end{enumerate}
+
+We now determine the likelihood of each particle by combining $L_1$ and $L_2$;
+\begin{equation}
+  \label{eq:particle_likelihood}
+  L (\bX_k^{(i)} | \bY_k, \bT_k) = L_1(\bX_k^{(i)}|\bY_k) L_2(\bX_k^{(i)} | \bT_k).
+\end{equation}
+This is then used to compute the weight of each particle,
+\begin{equation}
+  \label{eq:particle_weights}
+  w_i = \frac{L (\bX_k^{(i)} | \bY_k, \bT_k)}{\sum_{j=1}^N L (\bX_k^{(j)} | \bY_k, \bT_k)}.
+\end{equation}
+
+The resampling stage takes place at the beginning of each iteration for computational 
+reasons (see section~\ref{sec:results}).
+so we therefore use the weighted estimates of parameters in the next section.
+However, the resampling is simply done using a standard weighted resampling 
+algorithm with replacement.
 
@@
 
+- mention why weighted estimates stored rather than resampled 
+(i.e., to keep trajectories of disbanded particles for analysis).
+
 \section{Discusion and Future Work}
@@
 \item removing ``partial'' segments at start/end of routes
+\item enforcing max length of segments
 \end{itemize}

### 2017-04-24 realtime-filter/realtime-filter.tex
writing algorithm and stuff

@@
-\documentclass[9pt]{IEEEtran}\usepackage[]{graphicx}\usepackage[]{color}
+%\documentclass[9pt]{IEEEtran}
+\documentclass[draftcls,a4paper,onecolumn]{IEEEtran}\usepackage[]{graphicx}\usepackage[]{color}
 %% maxwidth is the original width if it is less than linewidth

### 2017-04-26 realtime-filter/realtime-filter.Rnw
fix up things

@@
   \label{eq:l1}
-  \log L_1 (\bX_k^{(i)} | \bY_k) = - \log(2\pi\sigma_y) - \frac{\|g(h(d_k^{(i)})|\bY_k)\|^2}{2\sigma_y^2}
+  \log L_1 (\bY_k | \bX_k^{(i)}) = - \log(2\pi\sigma_y) - \frac{\|g(h(d_k^{(i)})|\bY_k)\|^2}{2\sigma_y^2}
 \end{equation}
@@
   \label{eq:l1b}
-  \log L_1(\bX_k^{(i)} | \bY_k)=
+  \log L_1(\bY_k | \bX_k^{(i)})=
   \begin{cases}
@@
   \label{eq:l1c}
-  \log L_1 (\bX_k^{(i)} | \bY_k) =
+  \log L_1 (\bY_k | \bX_k^{(i)}) =
   \begin{cases}
@@
   then there is no change to the likelihood;
-  $ L_2  (\bX_k^{(i)} | \bT_k) = 1 $.
+  $ L_2  (\bT_k | \bX_k^{(i)}) = 1 $; else
+
+\item
+  If $s_k^{(i)} \neq j$ (particle's stop index isn't the same as the stop time update's stop sequence),
+  then we check particle is plausible given the data:
+  \begin{equation}
+    \label{eq:l2a}
+    L_2 (\bT_k | \bX_k^{(i)}) =
+    \mathbbm{1}_{s_k^{(i)} > j\ \cap\ A_{s_k}^{(i)} > T_j^a\ \cap\ A_{s_k}^{(i)} > T_j^d},
+  \end{equation}
+  (defining $x > y$ as TRUE if either $x$ or $y$ are NULL to simplify notation); else
 
@@
   If departure time, $T^d_j$, is available only, for stop $j$, then we
-  know the bus has at least departed stop $j$; therefore,
+  know the bus has departed stop $j$; therefore,
   \begin{equation}
     \label{eq:l2}
-    L_2 (\bX_k^{(i)} | \bT_k) =
-      \begin{cases}
-        f_N(D_j^{(i)} | T_j^d, \sigma_d^2) & s_k^{(i)} \geq j \cap D_j^{(i)} \neq \text{NULL} \\
-        0 & \text{otherwise},
-      \end{cases}
+    L_2 (\bT_k | \bX_k^{(i)}) =
+        f_N(T_j^d | D_j^{(i)}, \sigma_d^2) \mathbbm{1}_{D_j^{(i)}}
   \end{equation}
   where $f_N(x | \mu, \sigma^2)$ is the probability density of a normal distribution
-  with mean $\mu$ and variance $\sigma^2$.
+  with mean $\mu$ and variance $\sigma^2$,
+  and $\mathbbm{1}_Z = 0$ if $Z$ is NULL, and $1$ otherwise; else
 
@@
     \label{eq:l2b}
-    L_2  (\bX_k^{(i)} | \bT_k) =
-    \begin{cases}
-      f_N(A_j^{(i)} | T_j^a, \sigma_d^2) & s_k^{(i)} \geq j \\
-      0 & \text{otherwise}.
-    \end{cases}
+    L_2  (\bT_k | \bX_k^{(i)}) =
+      f_N(T_j^a | A_j^{(i)}, \sigma_d^2)
   \end{equation}
   Note that this does not assume the particle hasn't yet left the stop;
-  the departure time could simply be missing data.
+  the departure time could simply be missing data; else
 
@@
     \label{eq:l2c}
-    L_2  (\bX_k^{(i)} | \bT_k) =
-    \begin{cases}
-      \begin{array}{l}
-        f_N(A_j^{(i)} | T_j^a, \sigma_d^2) f_E(D_k^{(i)} - A_k^{(i)} | T_k^\delta) \quad\\
-        \hfill s_k^{(i)} \geq j \cap D_j^{(i)} \neq \text{NULL} \\[0.5em]
-        0 \qquad\qquad \text{otherwise},
-      \end{array}
-    \end{cases}
+    L_2  (\bT_k | \bX_k^{(i)}) =
+        f_N(T_j^a| A_j^{(i)}, \sigma_d^2) f_E(T_k^\delta | D_k^{(i)} - A_k^{(i)}) \mathbbm{1}_{D_j^{(i)}}
   \end{equation}
@@
   \label{eq:particle_likelihood}
-  L (\bX_k^{(i)} | \bY_k, \bT_k) = L_1(\bX_k^{(i)}|\bY_k) L_2(\bX_k^{(i)} | \bT_k).
+  L ([\bY_k, \bT_k] | \bX_k^{(i)}) = L_1(\bY_k|\bX_k^{(i)}) L_2(\bT_k | \bX_k^{(i)}).
 \end{equation}
@@
   \label{eq:particle_weights}
-  w_i = \frac{L (\bX_k^{(i)} | \bY_k, \bT_k)}{\sum_{j=1}^N L (\bX_k^{(j)} | \bY_k, \bT_k)}.
+  w_i = \frac{L ([\bY_k, \bT_k] | \bX_k^{(i)})}{\sum_{j=1}^N   L ([\bY_k, \bT_k] | \bX_k^{(j)})}.
 \end{equation}
@@
   \bar z_\ell = \frac{1}{M_\ell} \sum_{j = 1}^{M_\ell} \bar v_j \quad\text{and}\quad
-  \hat \bR_{\ell,\ell}^2 = \frac{1}{M_\ell} \sum_{j=1}^{M_\ell} \left(\bar v_j^2 + \bar S_j^2\right) - \hat v_\ell^2
+  \hat \bR_{\ell,\ell}^2 = \frac{1}{M_\ell} \sum_{j=1}^{M_\ell} \left(\bar v_j^2 + \bar S_j^2\right) - \bar z_\ell^2
 \end{equation}

### 2017-05-02 realtime-filter/realtime-filter.Rnw
fix forumulae
things
kf results + pictures

@@
 
-\title{Modelling the Real Time State of a Public Transport Road Network}
+\title{Modelling the real-time state of a public transport road network}
 \author{Tom~Elliott}
@@
 \begin{abstract}
-  Model all vehicles in the public transport system in real time.
+  Model all vehicles in the public transport system in real-time.
   Particle filter model for each vehicle (bus),
@@
 Define a parameter $\varepsilon_S$ as a boundary around the stop;
-if observation is within that boundary likelihood is 1, else 0.
+if observation is within that boundary likelihood is 1, else 0 (figure~\ref{fig:likelihood_circles}a).
 Let $B_s$ be a circle with radius $\varepsilon_S$ and centered at stop $j$, then
@@
 Similar to $L_1'$, however need to check observation is heading into the intersection
-(and not out of it) by using two boundaries around the intersection and a point $\varepsilon_I$ away.
+(and not out of it) by using two boundaries around the intersection and a point $\varepsilon_I$ away
+(figure~\ref{fig:likelihood_circles}b).
 Let $B_I^1$ and $B_I^2$ be two circles with radii $\varepsilon_I$,
@@
 
-%\subsubsection[L2]{$L_2$: not at a stop}
+<<likelihood_circles,echo=FALSE,out.width='0.5\\linewidth',fig.cap="Likelihood functions for particle's at stops and intersections.
+The grey region represents the area where particle's are accepted.",fig.subcap=c("Stop", "Intersection")>>=
+par(mar = rep(2.1, 4))
+plot(NA, bty = "n", type = "n", xaxt = "n", yaxt = "n", xlim = 0:1, ylim = 0:1, asp = 1)
+lines(c(0, 1), c(0.5, 0.5))
+th <- seq(0, 360, length = 100) * pi / 180
+polygon(cos(th) / 4 + 0.5, sin(th) / 4 + 0.5, col = "#cccccc")
+lines(c(0, 1), c(0.5, 0.5), lty = "9999")
+points(0.5, 0.5, pch = 19, cex = 4)
+
+plot(NA, bty = "n", type = "n", xaxt = "n", yaxt = "n", xlim = 0:1, ylim = 0:1, asp = 1)
+lines(c(0, 1), c(0.5, 0.5))
+th <- seq(0, 360, length = 100) * pi / 180
+th2 <- seq(-180, 180, length = 100) * pi / 180
+lines(c1x <- cos(th) / 4 + 0.7, c1y <- sin(th) / 4 + 0.5, lty = "99")
+lines(c2x <- cos(th2) / 4 + 0.7 - 0.25, c2y <- sin(th2) / 4 + 0.5, lty = "99")
+wc1 <- c1x <= 0.7 - 0.25/2
+wc2 <- c2x >= 0.7 - 0.25/2
+polygon(c(c1x[wc1], c2x[wc2]), c(c1y[wc1], c2y[wc2]), col = "#cccccc")
+lines(c(0, 1), c(0.5, 0.5), lty = "9999")
+points(0.7, 0.5, pch = 19, cex = 4)
+points(0.7 - 0.25, 0.5, pch = 15, cex = 3)
+@
 
@@
     L_2 (\bT_k | \bX_k^{(i)}) =
-    \mathbbm{1}_{s_k^{(i)} > j\ \cap\ A_{s_k}^{(i)} > T_j^a\ \cap\ A_{s_k}^{(i)} > T_j^d},
+    \mathbbm{1}_{s_k^{(i)} > j\ \cap\ A_{s_k}^{(i)} > T_j^a\ \cap\ A_{s_k}^{(i)} > T_j^d\
+      \cap\ T^a_j > t_k\ \cap\ T^d_j > t_k}
   \end{equation}
@@
   \label{eq:weighted_variance}
-  \bar S_j^2 = \frac{\sum_{i=1}^N w_j^{(i)} (v_j^{(i)} - \bar v_j)^2}{1 - \sum_{i=1}^N (w_j^{(i)})^2},
+  \bar S_j^2 = \sum_{i=1}^N w_j^{(i)} (v_j^{(i)} - \bar v_j)^2,
 \end{equation}
@@
 
+The transport network state is composed of $L$ segments,
+defined by the mean speed in each segment, 
+\mbox{$\bnu = [\nu_1, \nu_2, \cdots, \nu_L]^T$},
+and the variance of speed within each segment, $\bxi$, 
+a diagonal matrix where the $\xi_{\ell\ell}$ is the variance of the segment $\ell$.
+The Kalman filter predict step is
+\begin{align}
+  \bnu_{k|k-1} &= \bnu_{k-1|k-1} \\
+  \bxi_{k|k-1} &= \bxi_{k-1|k-1} + \bQ_k
+\end{align}
+where $\bQ_k \sim \mathcal{N}(0, \Delta_k\sigma_{\nu}^2I_L)$ and $\sigma_{\nu}^2$ is the system noise parameter
+and $I_n$ is an $n\times n$ identity matrix.
+
+The update equations simplify to
+\begin{align}
+  \bK_k &= \frac{\bxi_{k|k-1}}{\bxi_{k|k-1} + \hat R^2} \\
+  \nu_{k|k} &= \bnu_{k|k-1} + \bK_k(\bar\bz_k - \bnu_{k|k-1}) \\
+  \bxi_{k|k} &= (I_L - \bK_k) \bxi_{k|k-1}
+\end{align}
+which make use of the segment means and variances computed in (\ref{eq:kf_data_estimates}).
+
 \section{Arrival time prediction: application to a small subset of routes}
@@
 (i.e., to keep trajectories of disbanded particles for analysis).
+
+- model implemented in C++
 

### 2017-05-02 realtime-filter/realtime-filter.tex
fix forumulae
things
kf results + pictures

@@
 
-%\subsubsection[L2]{$L_2$: not at a stop}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\subfloat[Stop\label{fig:likelihood_circles1}]{\includegraphics[width=0.5\linewidth]{figure/likelihood_circles-1} }
+\subfloat[Intersection\label{fig:likelihood_circles2}]{\includegraphics[width=0.5\linewidth]{figure/likelihood_circles-2} }\caption[Likelihood functions for particle's at stops and intersections]{Likelihood functions for particle's at stops and intersections.
+The grey region represents the area where particle's are accepted.}\label{fig:likelihood_circles}
+\end{figure}
+
+\end{knitrout}
 

### 2017-05-03 realtime-filter/realtime-filter.Rnw
text and references

@@
 \algblockdefx[Phase]{Phase}{EndPhase}[2]{\textbf{Phase #1:} #2}{\textbf{End Phase}\\}
+
+\usepackage{cite}
 
@@
 
-Modelling busses in real-time has been done for a while now.
+Modelling busses in real-time has been done for a while now 
+\cite{Cats_2015,Chen_2014,Hans_2015,Jeong_2005,TCRP_2003,Wessel_2016,Xinghao_2013}.
 Modelling the underlying state of the road network in order to improve 
 travel time predictions has not been explored \ldots
-Some looking into complex models for ``future'' travel times \cite{Julio2016}.
+Some looking into complex models for ``future'' travel times 
+\cite{Julio_2016,Xinghao_2013,Yichen_Zheng_2016,Yu_2011}.
 
@@
 Transit data has become increasingly accessible over the year.
-Attibuted mostly to Google's \gls{gtfs}.
+Attibuted mostly to Google's \gls{gtfs} \cite{GoogleDevelopers_2006}.
 Two components to \gls{gtfs}: static and real-time.
@@
 However, difficult to combine information from different routes traveling along the same road segments.
+Some attempts \cite{Yu_2011}, but assumes knowledge of converging routes.
 
@@
 \emph{or} between a start/end stop of a route and an intersection.
+Add max segment length to attempt to reduce within-segment variability.
 
@@
 The idea behind using particle filter (as opposed to other models)
-is that it is well suited modelling busses \cite{Hans2015},
+is that it is well suited modelling busses \cite{Hans_2015},
 as it has good coverage of the plausible locations and thus overcomes
@@
 the routes shape information, $z_k^{(i)} = h(d_k^{(i)})$.
-The function $g (\bZ | \bY)$ performs an \emph{equirectangular projection} \cite{Snyder1997}
+The function $g (\bZ | \bY)$ performs an \emph{equirectangular projection} \cite{Snyder_1998}
 of the coordinates $\bX$, centered on $\bY$, 

### 2017-05-03 realtime-filter/realtime-filter.tex
text and references

@@
 
+\usepackage{cite}
+
 \input{../symbols.tex}

### 2017-05-05 realtime-filter/realtime-filter.Rnw
tidy up formulae and stuff

@@
 \begin{IEEEkeywords}
-  Public transport, real-time, particle filter, kalman filter
+  Public transport, real-time, particle filter, kalman filter, road state
 \end{IEEEkeywords}
@@
 Attibuted mostly to Google's \gls{gtfs} \cite{GoogleDevelopers_2006}.
-Two components to \gls{gtfs}: static and real-time.
+Two components to \gls{gtfs}: real-time and static.
+
+Real-time provides vehicle positions (as GPS coordinates),
+and trip updates (including stop time updates, i.e., arrival and departure times/delays).
+Updated typically every 30~seconds or so,
+and is used by the real-time model (section~\ref{sec:pf}).
 
@@
 stop positions, stop times, route shapes.
-
-Real-time provides vehicle positions (as GPS coordinates),
-and trip updates (including stop time updates, i.e., arrival and departure times/delays).
-Updated typically every 30~seconds or so.
 
@@
 We define a \emph{road segment} as a stretch of road used in a transit route between two intersections,
-\emph{or} between a start/end stop of a route and an intersection.
-Add max segment length to attempt to reduce within-segment variability.
+and partial segments between a start/end stop of a route and an intersection.
+Add max segment length to attempt to reduce within-segment variability on roads with few intersections 
+(e.g., highways).
 
-Do this by identifying all intersections
-(approximated by all traffic lights for now, since that's the data we have)
-and segment routes at these points.
+Do this by identifying all intersections,
+which can be downloaded from OpenStreetMap \cite{OpenStreetMap_2017}
+and segment routes at these points,
+as shown in figure~\ref{fig:route_segmentation}.
+For now, only using traffic lights as simplification.
 Modify \gls{gtfs} data so \emph{shapes} table becomes
 \texttt{shapes}, \texttt{segments}, and \texttt{segment\_shapes}.
+We assume that if two routes both go through intersections $A$ and $B$,
+they follow the same road to do so
+(once we scale up to all intersections, this will be true almost surely).
 
@@
 
-The vehicle state is related to an observable state $\bY_k = [\phi_k, \lambda_k]$,
-(which contains the latitude and longitude, respectively)
-through a \emph{measurment function}, $\bY_k = h(\bX_k) = h(d_k)$,
+To estimate $\bX_k$, we use the observed data,
+which comes from the realtime GTFS feed;
+\begin{equation}
+  \label{eq:data}
+  \bY_k = [t_k, \phi_k, \lambda_k, j, T_j^a, T_j^d]^T,
+\end{equation}
+which is the observation time, latitude, longitude, stop index,
+arrival time, and departure time, respectively.
+The relationship between the GPS position ($\tilde \bY_k = [\phi_k, \lambda_k]^T$)
+is denoted by a \emph{measurment function}, $\tilde\bY_k = h(\bX_k) = h(d_k)$,
 which depends only on the distance into trip.
-
-There is a second source of data, the \emph{stop time updates}:
-\mbox{$\bT_k = \left[ j, T^a_j , T^d_j \right]^T$}
-are the stop index, arrival and departure times,
-respectively; 
-either of the arrival or departure times can be missing.
+The \emph{stop time updates}, ($j, T^a_j , T^d_j$),
+are recorded when the bus arrives or departs a stop;
+either or both of the arrival or departure times can be missing.
 
@@
 the particle filter becomes a vastly superior model when compared to traditional
-Kalman filter based methods, as it can easily encompass each of these plausible states.
+Kalman filter based methods, 
+as it can easily deal with multimodality.
 
@@
 (for example, green vs.\ red traffic lights),
-and $\nu$ is the average queue time.
+and $\kappa$ is the average queue time.
 
-To implement the particle filter model, 
-we simply initialise $N$ particles (using superscript $^{(i)}$ to identify an individual particle)
+To implement the particle filter model (for a single vehicle $c$), 
+we simply initialise $N_c$ particles (using superscript~${(i)}$ to identify an individual particle)
 by sampling distance and velocity, and then determining $s_k$ and $r_k$ from the 
-vectors of stop and segment distances, $\bS^d$ and $\bR^d$, respectively.
+vectors of stop and segment distances, $\bS^d_r$ and $\bR^d_r$, respectively
+(conditional on the route, $r$).
 The remaining state parameters are NULL.
@@
       \Comment{at intersection}
-      \State $\bar t \sim \mathcal{E}(\nu^{-1})$
+      \State $\bar t \sim \mathcal{E}(\kappa^{-1})$
       \State $Q_r\gets Q_r + \bar t$, $w\gets w+\bar t$, $\delta\gets \delta - w$
@@
           \State $Q \gets 0$, $B\gets $ null, $d\gets R^d_r$
-          \State $p \sim \mathcal{B}(\rho)$, $\bar t \sim \mathcal{E}(\nu^{-1})$
+          \State $p \sim \mathcal{B}(\rho)$, $\bar t \sim \mathcal{E}(\kappa^{-1})$
           \State queue $\gets p\bar t$, $\delta\gets \delta\ - $ queue
@@
 the routes shape information, $z_k^{(i)} = h(d_k^{(i)})$.
-The function $g (\bZ | \bY)$ performs an \emph{equirectangular projection} \cite{Snyder_1998}
-of the coordinates $\bX$, centered on $\bY$, 
+The function $g (\bZ | \tilde\bY)$ performs an \emph{equirectangular projection} \cite{Snyder_1998}
+of the coordinates $\bX$, centered on $\tilde\bY$, 
 onto a flat surface so that the Euclidean distance between points is equivalent
@@
   \label{eq:l1}
-  \log L_1 (\bY_k | \bX_k^{(i)}) = - \log(2\pi\sigma_y) - \frac{\|g(h(d_k^{(i)})|\bY_k)\|^2}{2\sigma_y^2}
+  \log L_1 (\bY_k | \bX_k^{(i)}) = - \log(2\pi\sigma_y) - \frac{\|g(h(d_k^{(i)})|\tilde\bY_k)\|^2}{2\varepsilon_y^2}
 \end{equation}
@@
   \begin{cases}
-    - \log(\pi \varepsilon_S^2) & \bY_k \in B_s \\
+    - \log(\pi \varepsilon_S^2) & \tilde\bY_k \in B_s \\
     - \infty & \text{otherwise}
@@
   \begin{cases}
-    - \log(A) & \bY_k \in B_I^1 \cap B_I^2 \\
+    - \log(A) & \tilde\bY_k \in B_I^1 \cap B_I^2 \\
     - \infty & \text{otherwise}
@@
     L_2 (\bT_k | \bX_k^{(i)}) =
-    \mathbbm{1}_{s_k^{(i)} > j\ \cap\ A_{s_k}^{(i)} > T_j^a\ \cap\ A_{s_k}^{(i)} > T_j^d\
-      \cap\ T^a_j > t_k\ \cap\ T^d_j > t_k}
+    \mathbbm{1}_{s_k^{(i)} > j\ \cap\ A_{s_k}^{(i)} > \max\{T_j^a,T_j^d\} > t_k}
   \end{equation}
@@
   \label{eq:particle_likelihood}
-  L ([\bY_k, \bT_k] | \bX_k^{(i)}) = L_1(\bY_k|\bX_k^{(i)}) L_2(\bT_k | \bX_k^{(i)}).
+  L (\bY_k | \bX_k^{(i)}) = L_1(\tilde\bY_k|\bX_k^{(i)}) L_2(\bT_k | \bX_k^{(i)}).
 \end{equation}
@@
   \label{eq:particle_weights}
-  w_i = \frac{L ([\bY_k, \bT_k] | \bX_k^{(i)})}{\sum_{j=1}^N   L ([\bY_k, \bT_k] | \bX_k^{(j)})}.
+  w_i = \frac{L (\bY_k | \bX_k^{(i)})}{\sum_{j=1}^N   L (\bY_k | \bX_k^{(j)})}.
 \end{equation}
@@
 
+Want to find the state,
+\begin{equation}
+  \label{eq:kf_state}
+  \Theta = [\theta_1, \cdots, \theta_L]^T\quad\text{and}\quad
+  \Xi = \diag(\xi_1, \cdots, \xi_L) = \bxi\eye_L,
+\end{equation}
+which are the mean and variance of each segement, respectively.
+
 We run the Kalman filter every time we run the particle filter,
 since it has a low cost, allowing the most recent data to be incorporated.
-We have $N$ weighted particle estimates of vehicle state for each bus,
-with which we can compute the mean distance into trip and velocity for bus $j$,
+For each vehicle $c$, we have $N_c$ particles which we use to
+compute the mean distance into trip and velocity,
 \begin{equation}
   \label{eq:weighted_mean}
-  \bar d_j = \sum_{i=1}^N w_j^{(i)} d_j^{(i)} \quad\text{and}\quad
-  \bar v_j = \sum_{i=1}^N w_j^{(i)} v_j^{(i)},
+  \bar d_c = \sum_{i=1}^{N_c} w_c^{(i)} d_c^{(i)} \quad\text{and}\quad
+  \bar v_c = \sum_{i=1}^{N_c} w_c^{(i)} v_c^{(i)},
 \end{equation}
-as well as the weighted sample variance \cite{cn},
+as well as the weighted sample variance,
 \begin{equation}
   \label{eq:weighted_variance}
-  \bar S_j^2 = \sum_{i=1}^N w_j^{(i)} (v_j^{(i)} - \bar v_j)^2,
+  \bar S_c^2 = \sum_{i=1}^{N_c} w_c^{(i)} (v_c^{(i)} - \bar v_c)^2,
 \end{equation}
-since $\sum_{i=1}^N w_j^{(i)} = 1\quad \forall j$.
+since $\sum_{i=1}^{N_c} w_c^{(i)} = 1\quad \forall c = 1,\ldots,C$.
 
-Using $\bar d_j$, we can find the segment id corresponding to each busses most probable location,
+Using $\bar d_c$, we can find the segment id corresponding to each busses most probable location,
 and then combine all vehicle speed estimates for each segment.
@@
 In situations where multiple vehicles are in the same segment $\ell$, 
-we can simply combine the estimates, using $M_\ell = $ number of vehicles in segment $\ell$ 
+we can simply combine the estimates, using $C_\ell = $ number of vehicles in segment $\ell$ 
 \cite{cn}.
@@
   \label{eq:kf_data_estimates}
-  \bar z_\ell = \frac{1}{M_\ell} \sum_{j = 1}^{M_\ell} \bar v_j \quad\text{and}\quad
-  \hat \bR_{\ell,\ell}^2 = \frac{1}{M_\ell} \sum_{j=1}^{M_\ell} \left(\bar v_j^2 + \bar S_j^2\right) - \bar z_\ell^2
+  \hat V_\ell = \frac{1}{C_\ell} \sum_{c=1}^{C_\ell} \bar v_j \quad\text{and}\quad
+  \hat P_\ell = \frac{1}{C_\ell} \sum_{j=1}^{C_\ell} \left(\bar v_c^2 + \bar S_c^2\right) - \hat V_\ell^2
 \end{equation}
 
-Next is just the simple matter of applying the Kalman filter algorithm to the transport road network state,
-using the data $\bz_k$ as the observation and $\bR_k$,
+Next is just the simple matter of applying the Kalman filter algorithm \cite{Anderson_2012} 
+to the transport road network state,
+using the data \mbox{$\hat\bV_k = [\hat V_1, \ldots, \hat V_L]$} as the observation vector
+and \mbox{$\hat\bP_k = \diag(\hat P_1, \ldots, \hat P_L)$},
 as the measurement error matrix.
-
-The transport network state is composed of $L$ segments,
-defined by the mean speed in each segment, 
-\mbox{$\bnu = [\nu_1, \nu_2, \cdots, \nu_L]^T$},
-and the variance of speed within each segment, $\bxi$, 
-a diagonal matrix where the $\xi_{\ell\ell}$ is the variance of the segment $\ell$.
-The Kalman filter predict step is
+The Kalman filter prediction equations are
 \begin{align}
-  \bnu_{k|k-1} &= \bnu_{k-1|k-1} \\
-  \bxi_{k|k-1} &= \bxi_{k-1|k-1} + \bQ_k
+  \Theta_{k|k-1} &= \Theta_{k-1|k-1} \\
+  \Xi_{k|k-1} &= \Xi_{k-1|k-1} + \bW_k
 \end{align}
-where $\bQ_k \sim \mathcal{N}(0, \Delta_k\sigma_{\nu}^2I_L)$ and $\sigma_{\nu}^2$ is the system noise parameter
-and $I_n$ is an $n\times n$ identity matrix.
+where $\bW_k \sim \mathcal{N}(0, \Delta_k\zeta^2\eye_L)$, $\zeta^2$ is the system noise parameter,
+and $\Delta_k = t_k - t_{k-1}$.
 
@@
 \begin{align}
-  \bK_k &= \frac{\bxi_{k|k-1}}{\bxi_{k|k-1} + \hat R^2} \\
-  \nu_{k|k} &= \bnu_{k|k-1} + \bK_k(\bar\bz_k - \bnu_{k|k-1}) \\
-  \bxi_{k|k} &= (I_L - \bK_k) \bxi_{k|k-1}
+  \bK_k &= \frac{\Xi_{k|k-1}}{\Xi_{k|k-1} + \hat\bP_k} \\
+  \Theta_{k|k} &= \Theta_{k|k-1} + \bK_k(\bV_k - \Theta_{k|k-1}) \\
+  \Xi_{k|k} &= (\eye_L - \bK_k) \Xi_{k|k-1}
 \end{align}
-which make use of the segment means and variances computed in (\ref{eq:kf_data_estimates}).
+which make use of the segment means and variances computed in~(\ref{eq:kf_data_estimates}).
 
@@
 
+\appendix[List of Symbols]
+
+\begin{tabular}{ll}
+  $N_c$ & number of particles for vehicle $c$ \\
+  $C$ & number of vehicles \\
+  $C_\ell$ & number of vehicles in segment $\ell$ \\
+  $M$ ($M_r$) & number of stops (for given route $r$) \\
+  $L$ ($L_r$) & number of segments (for given route $r$) \\
+  $\bS^d_r, \bR^d_r$ & stop, intersection distances along route $r$ \\
+  $t_k$ & time at observation $k$ \\
+  $\bY_k$ & observation $k$ \\
+  $\phi,\lambda$ & latitude, latitude \\
+  $T^a_j, T^d_j$ & arrival, departure time at stop $j \in \{1, \ldots, M_s\}$ \\
+  $\bX_k$ & vehicle state associated with obs $k$ \\
+  $d$ & distance into trip (meters) \\
+  $v$ & velocity (speed, meters per second) \\
+  $s$ & stop index $\in \{1, \ldots, M_s\}$ \\
+  $A_s, D_s$ & arrival, departure time at stop $s$ \\
+  $r$ & segment index $\in \{1, \ldots, M_r\}$ \\
+  $Q_r, B_r$ & queue time (seconds), begin time for segment $r$ \\
+  $z$ & computed GPS position of a particle \\
+  $\delta$ & time since last observation (seconds) \\
+  $\pi$ & probability of stopping at a stop \\
+  $\gamma$ & minimum dwell time (seconds) \\
+  $\tau$ & mean dwell time (seconds) \\
+  $\rho$ & probability of stopping at an intersection \\
+  $\kappa$ & mean queue time at intersection \\
+  $\sigma_v^2$ & system noise (particle filter) \\
+  $\sigma_r^2$ & between segment velocity variability \\
+  $\varepsilon_y^2, \varepsilon_d^2$ & GPS, arrival/departure time error \\
+  $\varepsilon_S, \varepsilon_I$ & stop, intersection radius (meters) \\
+  $w$ & particle weight \\
+  $\Theta, \Xi$ & transport road state mean, variance \\
+  $\hat V_\ell$ & mean speed of recent vehicles in segment $\ell$ \\
+  $\hat P_\ell$ & estimated population variance of vehicle speeds in segment $\ell$ \\
+  $\zeta^2$ & system noise (Kalman filter)
+\end{tabular}
+
+\vspace{1em}
+Times are UNIX timestamps (seconds since 1 Janurary, 1970) unless stated otherwise.
+
 \end{document}

### 2017-06-27 realtime-filter/realtime-filter.Rnw
Write a little something about arrival time pred...

@@
 \usepackage{algpseudocode}
-\algblockdefx[Phase]{Phase}{EndPhase}[2]{\textbf{Phase #1:} #2}{\textbf{End Phase}\\}
+\algblockdefx[Phase]{Phase}{EndPhase}[2]{\textbf{Phase #1:} #2}%
+{\textbf{End Phase}\\}
 
@@
 Real-time provides vehicle positions (as GPS coordinates),
-and trip updates (including stop time updates, i.e., arrival and departure times/delays).
+and trip updates (including stop time updates, 
+i.e., arrival and departure times/delays).
 Updated typically every 30~seconds or so,
@@
 
-Static contains information about routes, trips (one instance of a route at a scheduled time of day),
+Static contains information about routes, 
+trips (one instance of a route at a scheduled time of day),
 stop positions, stop times, route shapes.
@@
 
-\section{Arrival time prediction: application to a small subset of routes}
-\label{sec:results}
+\section{Arrival time prediction}
+\label{sec:arrival_time}
 
@@
 Also include dwell distibutions at stops and intersections.
+
+The final step is to predict the arrival time of every bus 
+at all remaining stops along their respective routes.
+Using the real-time transport network road state, 
+we can predict how long it will take a bus to travel 
+to each stop.
+
+For segments \emph{without} stops,
+we simply sample a travel time from the segment's state estimate
+for each particle.
+For segments with stops, however,
+we only want a \emph{partial} travel time.
+Most simply, this is based on proportionally how far into the segment
+the stop is.
+More complex situations could be investigated in the future,
+one example might be to use \emph{sub-segments}.
+
+\section{Application to Auckland Transport}
+\label{sec:results}
 

### 2017-07-31 realtime-filter/realtime-filter.Rnw
Working on symbols

@@
 Modelling busses in real-time has been done for a while now 
-\cite{Cats_2015,Chen_2014,Hans_2015,Jeong_2005,TCRP_2003,Wessel_2016,Xinghao_2013}.
+\cite{Cats_2015,Chen_2014,Hans_2015,Jeong_2005,TCRP_2003,Wessel_2016,
+  Xinghao_2013}.
 Modelling the underlying state of the road network in order to improve 
@@
 
-Modelling a single trip or route is fairly simple to do with the raw \gls{gtfs} data.
-However, difficult to combine information from different routes traveling along the same road segments.
+Modelling a single trip or route is fairly simple to do with 
+the raw \gls{gtfs} data.
+However, difficult to combine information from different routes 
+traveling along the same road segments.
 Some attempts \cite{Yu_2011}, but assumes knowledge of converging routes.
 
-We define a \emph{road segment} as a stretch of road used in a transit route between two intersections,
-and partial segments between a start/end stop of a route and an intersection.
-Add max segment length to attempt to reduce within-segment variability on roads with few intersections 
-(e.g., highways).
+We define a \emph{road segment} as a stretch of road used 
+in a transit route between two intersections,
+and partial segments between a start/end stop of a route 
+and an intersection.
+Add max segment length to attempt to reduce within-segment variability 
+on roads with few intersections (e.g., highways).
 
@@
 as shown in figure~\ref{fig:route_segmentation}.
-For now, only using traffic lights as simplification.
+For now, only using traffic lights and roundabouts as simplification.
 Modify \gls{gtfs} data so \emph{shapes} table becomes
@@
 they follow the same road to do so
-(once we scale up to all intersections, this will be true almost surely).
+(routes retain their own shape, so in cases where this \emph{is not} true,
+the shape will still be valid).
 
@@
 
-<<route_segmentation,echo=FALSE,out.width='0.5\\linewidth',fig.cap="Simplified unidirectional road network diagram, with (a) overlapping routes, and (b) the segments obtained by splitting the routes at intersections.
-Route 1 $ = \\{r_1, r_2, r_3\\}$, route 2 $=\\{r_4,r_2,r_5,r_6\\}$, and route 3 $=\\{r_7,r_5,r_8,r_9\\}$.",fig.scap="Simplied diagram of overlapping routes",fig.subcap=c("Routes","Segments")>>=
+<<route_segmentation,echo=FALSE,out.width='0.5\\linewidth',fig.cap="Simplified unidirectional road network diagram, with (a) overlapping routes, and (b) the segments obtained by splitting the routes at intersections (solid dots).
+Route 1 $ = \\{R_1, R_2, R_3\\}$, route 2 $=\\{R_4,R_2,R_5,R_6\\}$, and route 3 $=\\{R_7,R_5,R_8,R_9\\}$.",fig.scap="Simplied diagram of overlapping routes",fig.subcap=c("Routes","Segments")>>=
 par(mar = rep(2.1, 4))
@@
 r1 <- rbind(c(0.5, 1), c(1 - e, 1), c(1 - e, 2), c(1 - e, 2.5))
-r2 <- rbind(c(1 + e, 0.5), c(1 + e, 1), c(1 + e, 2 - e), c(2, 2 - e), c(2.5, 2 - e))
-r3 <- rbind(c(0.5, 2 + e), c(1, 2 + e), c(2 + e, 2 + e), c(2 + e, 1), c(2 + e, 0.5))
+r2 <- rbind(c(1 + e, 0.5), c(1 + e, 1), c(1 + e, 2 - e), c(2, 2 - e), 
+            c(2.5, 2 - e))
+r3 <- rbind(c(0.5, 2 + e), c(1, 2 + e), c(2 + e, 2 + e), c(2 + e, 1), 
+            c(2 + e, 0.5))
 grid <- rbind(c(1, 1), c(1, 2), c(2, 1), c(2, 2))
@@
        lwd = c(2, 4, 2,  2, 4, 2,  2, 2))
-#points(grid, pch = 19, cex = 4, col = "white")
 points(grid, pch = 19, cex = 2, col = "black")
 
-tx <- rbind(c(0.7, 1.1), c(0.9, 1.5), c(0.9, 2.3),
-            c(1.1, 0.7), c(1.5, 2.1), c(2.3, 2.1),
-            c(0.7, 1.9), c(2.1, 1.5), c(2.1, 0.7))
+tx <- rbind(c(0.7, 1.0), c(1, 1.5), c(1, 2.3),
+            c(1.0, 0.7), c(1.5, 2.0), c(2.3, 2.0),
+            c(0.7, 1.98), c(2.0, 1.5), c(2.0, 0.7))
 for (i in 1:nrow(tx)) {
     text(tx[i,1], tx[i,2], 
-         substitute(r[x], list(x = i)),
-         cex = 3, family="serif", font=3)
+         substitute(R[x], list(x = i)),
+         cex = 2.5, family="serif", font = 3,
+         pos=c(3, 2, 2, 4, 3, 3, 1, 4, 4)[i])
 }
@@
 vehicle state,
-consisting of distance into trip ($d$), velocity ($v$),
-current stop sequence ($s$), 
-arrival and departure times at the last stop ($A_s$ and $D_s$),
-current segment index ($r$),
-the time spent queuing at the beginning of the current segment ($Q_r$),
-and the time at which the vehicle began traveling along the current segment ($B_r$);
-the vehicle state at time $k$ is denoted
+consisting of distance into trip, $x$, 
+velocity, $\dot x$,
+current stop sequence, $j \in \{1,\ldots,J\}$, 
+arrival and dwell times at each stop, 
+$\mat A = [A_2, \cdots, A_J]^T$ and 
+$\tilde {\mat  D} = [\tilde D_1, \cdots, \tilde D_{J-1}]^T$ respectively
+(note that the first stop doesn't have an arrival time,
+and the last doesn't have a departure),
+current segment index, $\ell$,
+the time spent queuing at the intersection at the beginning
+of each segment,
+$\mat q = [q_2, \cdots, q_L]^T$ (note no queueing time for the first segment),
+and the travel time (less dwell time) along each segment,
+$\mat b = [b_1, \cdots, b_L]^T$.
+The vehicle state at time $k$ is denoted
 \begin{equation}
   \label{eq:vehiclestate}
-  \bX_k = [d_k, v_k, s_k, A_{s_k}, D_{s_k}, r_k, Q_{r_k}, B_{r_k}]^T.
+  \bX_k = 
+  [x_k, \dot x_k, j, \mat A^T, \tilde{\mat D}^T, \ell, \mat q^T, \mat b^T]^T
 \end{equation}
@@
 To estimate $\bX_k$, we use the observed data,
-which comes from the realtime GTFS feed;
+which comes from the realtime GTFS feed in two parts:
+the first is vehicle position, 
+$\tilde y_k = [\mathrm{lng}_k, \mathrm{lat}_l]^T = 
+\frac{180}{\pi}[\phi_k, \lambda_k]^T$,
+where $\phi$ and $\lambda$ are the latitude and longitude in radians, 
+respectively;
+the second part is the \emph{stop time updates},
+$\mat T_k = [j_k, T_j^a, T_j^d]^T$,
+which are the current stop sequence (at time $k$), $j_k$,
+the arrival time, $T_j^a$, and departure time, $T_j^d$.
+Thus the data obtained at time $t_k$ can be represented by
 \begin{equation}
   \label{eq:data}
-  \bY_k = [t_k, \phi_k, \lambda_k, j, T_j^a, T_j^d]^T,
+  \bY_k = [t_k, \tilde y_k, \mat T_k]^T.
 \end{equation}
-which is the observation time, latitude, longitude, stop index,
-arrival time, and departure time, respectively.
-The relationship between the GPS position ($\tilde \bY_k = [\phi_k, \lambda_k]^T$)
-is denoted by a \emph{measurment function}, $\tilde\bY_k = h(\bX_k) = h(d_k)$,
+The relationship between the GPS position, $\tilde y$,
+and vehicle's distance into trip, $x$,
+is denoted by a \emph{measurment function}, 
+$\tilde y_k = h(\bX_k) = h(x_k)$,
 which depends only on the distance into trip.
-The \emph{stop time updates}, ($j, T^a_j , T^d_j$),
+The \emph{stop time updates}, $\mat T_k$,
 are recorded when the bus arrives or departs a stop;
@@
 
-On approach to a stop $j$, the bus will stop with probability $\pi_j$ 
+On approach to the $j$th stop along a route, 
+$s_j = s(j | \text{route})$,
+the bus will stop with probability $\pi_{s_j}$ 
 (i.e., it is a bernoulli trial for each particle).
-If it stops, it will wait a minimum of $\gamma$ seconds while the doors are opened and closed,
-and an additional $\bar t$ seconds as passengers board and disembark.
+If it stops, it will wait a minimum of $\gamma$ seconds 
+while the doors are opened and closed,
+and an additional $u$ seconds as passengers board and disembark.
 $\gamma$ is considered a fixed parameter, across all busses, 
-while $\bar t$ is given an exponential distribution that, for now, has a fixed mean, $\tau$
-(section~\ref{sec:discussion} discusses modifications to this).
+while $u$ is given an exponential distribution with mean $\tau_{s_j}$.
+For now, $\pi_s = \pi$ and $\tau_s = \tau$ $\forall s$
+(i.e., every stop has the same stopping probability and mean dwell time; 
+section~\ref{sec:discussion} discusses modifications to this).
 
@@
 Similarly to stops, currently this is just modelled using fixed parameter values;
-$\rho$ is the probability of stopping at the intersection
+$\rho_{r_\ell}$ is the probability of stopping at the intersection
 (for example, green vs.\ red traffic lights),
-and $\kappa$ is the average queue time.
+and $\theta_{r_\ell}$ is the average queue time.
+Similiarly to stops, $\rho_r = \rho$ and $\theta_r = \theta$ $\forall r$.
 
-To implement the particle filter model (for a single vehicle $c$), 
-we simply initialise $N_c$ particles (using superscript~${(i)}$ to identify an individual particle)
-by sampling distance and velocity, and then determining $s_k$ and $r_k$ from the 
-vectors of stop and segment distances, $\bS^d_r$ and $\bR^d_r$, respectively
-(conditional on the route, $r$).
-The remaining state parameters are NULL.
-Each particle is then passed through the \emph{transition function} independently (the prediction step),
-and then the vehicle state estimate is updated using the data by computing likelihood weights
+To implement the particle filter model (for a single vehicle $v$), 
+we simply initialise $I$ particles 
+(using superscript~${(i)}$ to identify an individual particle)
+by sampling distance and velocity from a prior distribution
+(see section~\ref{sec:results}), 
+and the computing the indices,
+$j = \arg\max_{j:s_j^d \leq x} \{s_j^d\}$ 
+and $\ell = \arg\max_{\ell:r_\ell^d \leq x} \{r_\ell^d\}$ from the 
+vectors of stop and segment distances, $\mat s^d$ and $\mat r^d$, respectively.
+The remaining state parameters are left uninitialized.
+Each particle is then passed through the \emph{transition function}, $f$,
+independently (the prediction step),
+and then the vehicle state estimate is updated 
+using the data by computing likelihood weights
 and resampling the particles with replacement.
 The steps are described below.
+
+THE ABOVE WILL CHANGE:
+\begin{itemize}
+\item project each particle to the end of the route,
+  including queuing and dwell times
+\item update vehicle positions
+\item update trip updates
+\item each time, reweight (and resample?) based on proximity
+\end{itemize}
 
@@
 Also easy to describe the model;
-only have to deal with an individual object (particle) which follows laws of physics.
+only have to deal with an individual object (particle) which follows 
+laws of physics.
 Rather than dealing with a state \emph{distribution}.
@@
 The $k$ subscripts and $(i)$ superscripts have been dropped from the
-vehicle state parameters (e.g., $d \equiv d_k^{(i)}$).
+vehicle state parameters (e.g., $x \equiv x_k^{(i)}$).
 
-The additional parameters are the noise terms, $\sigma_v^2$ and $\sigma_r^2$,
-which represent system noise and between-segment speed variance respectively.
-The former is assumed to be small, and represents the average change in speed
-of a bus per second \emph{within a segment},
-while the latter is the variance in speed \emph{between adjacent segments},
-and allows for the bus to quickly change speed when passing through an intersection.
+The noise parameter, $\sigma_v^2$
+represents system noise, assumed to be small, 
+and represents the average change in speed of a bus per second.
 
@@
 
-So, depending on the particles' states, we apply different likelihood functions.
-
-\begin{enumerate}
-%\subsubsection[L1]{$L_1$: not at a stop or intersection}
-\item
-In situations where the bus is neither at a bus stop nor at an intersection,
-the likelihood is simply determined from the distance between the particle and 
+\subsubsection{GTFS vehicle position}
+The likelihood is simply determined from the distance between the particle and 
 vehicle's reported position.
 The particle filter framework makes this easy,
-as each particle has a single value of distance into trip, $d_k^{(i)}$,
+as each particle has a single value of distance into trip, $x_k^{(i)}$,
 with which we can compute the latitude and longitude values based on 
@@
   \label{eq:l1}
-  \log L_1 (\bY_k | \bX_k^{(i)}) = - \log(2\pi\sigma_y) - \frac{\|g(h(d_k^{(i)})|\tilde\bY_k)\|^2}{2\varepsilon_y^2}
+  \log L_1 (\bY_k | \bX_k^{(i)})
+  = \log L_1 (\tilde y_k | x_k^{(i)})
+  = - \log(2\pi\sigma_y) - \frac{\|g(h(x_k^{(i)})|\tilde y_k)\|^2}{2\sigma_y^2}
 \end{equation}
 
-%\subsubsection[L1b]{$L_1'$: at stop}
-\item
-If the particle is considered to be at a stop, then we apply the following likelihood.
-Define a parameter $\varepsilon_S$ as a boundary around the stop;
-if observation is within that boundary likelihood is 1, else 0 (figure~\ref{fig:likelihood_circles}a).
-Let $B_s$ be a circle with radius $\varepsilon_S$ and centered at stop $j$, then
-
-\begin{equation}
-  \label{eq:l1b}
-  \log L_1(\bY_k | \bX_k^{(i)})=
-  \begin{cases}
-    - \log(\pi \varepsilon_S^2) & \tilde\bY_k \in B_s \\
-    - \infty & \text{otherwise}
-  \end{cases}
-\end{equation}
-
-%\subsubsection[L1c]{$L_1''$: at intersection}
-\item
-If the particle is considered to be at an intersection, we apply the following likelihood.
-Similar to $L_1'$, however need to check observation is heading into the intersection
-(and not out of it) by using two boundaries around the intersection and a point $\varepsilon_I$ away
-(figure~\ref{fig:likelihood_circles}b).
-Let $B_I^1$ and $B_I^2$ be two circles with radii $\varepsilon_I$,
-the first centered on the intersection and the second centered $\varepsilon_I$ meters before it, then
-
-\begin{equation}
-  \label{eq:l1c}
-  \log L_1 (\bY_k | \bX_k^{(i)}) =
-  \begin{cases}
-    - \log(A) & \tilde\bY_k \in B_I^1 \cap B_I^2 \\
-    - \infty & \text{otherwise}
-  \end{cases}
-\end{equation}
-where $A = 2\varepsilon_I^2 \left( \cos^{-1}(\frac{1}{2}) - \frac{\sqrt{3}}{4} \right)$,
-the area of $B_I^1\cap B_I^2$.
-
-\end{enumerate}
-
-<<likelihood_circles,echo=FALSE,out.width='0.5\\linewidth',fig.cap="Likelihood functions for particle's at stops and intersections.
-The grey region represents the area where particle's are accepted.",fig.subcap=c("Stop", "Intersection")>>=
-par(mar = rep(2.1, 4))
-plot(NA, bty = "n", type = "n", xaxt = "n", yaxt = "n", xlim = 0:1, ylim = 0:1, asp = 1)
-lines(c(0, 1), c(0.5, 0.5))
-th <- seq(0, 360, length = 100) * pi / 180
-polygon(cos(th) / 4 + 0.5, sin(th) / 4 + 0.5, col = "#cccccc")
-lines(c(0, 1), c(0.5, 0.5), lty = "9999")
-points(0.5, 0.5, pch = 19, cex = 4)
-
-plot(NA, bty = "n", type = "n", xaxt = "n", yaxt = "n", xlim = 0:1, ylim = 0:1, asp = 1)
-lines(c(0, 1), c(0.5, 0.5))
-th <- seq(0, 360, length = 100) * pi / 180
-th2 <- seq(-180, 180, length = 100) * pi / 180
-lines(c1x <- cos(th) / 4 + 0.7, c1y <- sin(th) / 4 + 0.5, lty = "99")
-lines(c2x <- cos(th2) / 4 + 0.7 - 0.25, c2y <- sin(th2) / 4 + 0.5, lty = "99")
-wc1 <- c1x <= 0.7 - 0.25/2
-wc2 <- c2x >= 0.7 - 0.25/2
-polygon(c(c1x[wc1], c2x[wc2]), c(c1y[wc1], c2y[wc2]), col = "#cccccc")
-lines(c(0, 1), c(0.5, 0.5), lty = "9999")
-points(0.7, 0.5, pch = 19, cex = 4)
-points(0.7 - 0.25, 0.5, pch = 15, cex = 3)
-@
-
+\subsubsection{GTFS trip updates}
 The second part of the likelihood involves the stop time update,
@@
 
-\begin{enumerate}
-\item 
-  If both $T^a_j$ and $T^d_j$ are NULL, 
-  then there is no change to the likelihood;
-  $ L_2  (\bT_k | \bX_k^{(i)}) = 1 $; else
-
-\item
-  If $s_k^{(i)} \neq j$ (particle's stop index isn't the same as the stop time update's stop sequence),
-  then we check particle is plausible given the data:
+We only apply the likelihood when the data is recieved, 
+so the two necessary likelihood functions are,
+for an arrival time,
 \begin{equation}
   \label{eq:l2a}
-    L_2 (\bT_k | \bX_k^{(i)}) =
-    \mathbbm{1}_{s_k^{(i)} > j\ \cap\ A_{s_k}^{(i)} > \max\{T_j^a,T_j^d\} > t_k}
+  \log L_2 (\bY_k | \bX_k^{(i)})
+  = \log L_2 (T_j^a | A_j^{(i)}) 
+  = - \frac{1}{2}\log (2\pi\varepsilon^2) - 
+  \frac{(T_j^a - A_j^{(i)})^2}{2\varepsilon^2},
 \end{equation}
-  (defining $x > y$ as TRUE if either $x$ or $y$ are NULL to simplify notation); else
-
-\item
-  If departure time, $T^d_j$, is available only, for stop $j$, then we
-  know the bus has departed stop $j$; therefore,
+and for a departure time
+(note that the departure time for a particle
+is simply $D_j^{(i)} = A_j^{(i)} + \tilde D_j^{(i)}$),
 \begin{equation}
-    \label{eq:l2}
-    L_2 (\bT_k | \bX_k^{(i)}) =
-        f_N(T_j^d | D_j^{(i)}, \sigma_d^2) \mathbbm{1}_{D_j^{(i)}}
+  \label{eq:l2d}
+  \log L_2 (\bY_k | \bX_k^{(i)})
+  = \log L_2 (T_j^d | D_j^{(i)}) 
+  = - \frac{1}{2}\log (2\pi\varepsilon^2) - 
+  \frac{(T_j^d - D_j^{(i)})^2}{2\varepsilon^2}
 \end{equation}
-  where $f_N(x | \mu, \sigma^2)$ is the probability density of a normal distribution
-  with mean $\mu$ and variance $\sigma^2$,
-  and $\mathbbm{1}_Z = 0$ if $Z$ is NULL, and $1$ otherwise; else
-
-\item
-  If arrival time, $T^a_j$, is available only, then similarly
-  \begin{equation}
-    \label{eq:l2b}
-    L_2  (\bT_k | \bX_k^{(i)}) =
-      f_N(T_j^a | A_j^{(i)}, \sigma_d^2)
-  \end{equation}
-  Note that this does not assume the particle hasn't yet left the stop;
-  the departure time could simply be missing data; else
-
-\item
-  If both arrival time and departure time are available, then
-  \begin{equation}
-    \label{eq:l2c}
-    L_2  (\bT_k | \bX_k^{(i)}) =
-        f_N(T_j^a| A_j^{(i)}, \sigma_d^2) f_E(T_k^\delta | D_k^{(i)} - A_k^{(i)}) \mathbbm{1}_{D_j^{(i)}}
-  \end{equation}
-  where $T_k^\delta = T_k^d - T_k^a$ is the observed dwell time at the stop,
-  and $f_E(x | \lambda)$ is the probability density of an exponential distribution 
-  with rate $\lambda$.
-
-\end{enumerate}
 
@@
   \label{eq:particle_weights}
-  w_i = \frac{L (\bY_k | \bX_k^{(i)})}{\sum_{j=1}^N   L (\bY_k | \bX_k^{(j)})}.
+  w_i = \frac{L (\bY_k | \bX_k^{(i)})}{\sum_{j=1}^I   L (\bY_k | \bX_k^{(j)})}.
 \end{equation}
-
-The resampling stage takes place at the beginning of each iteration for computational 
-reasons (see section~\ref{sec:results}).
-so we therefore use the weighted estimates of parameters in the next section.
-However, the resampling is simply done using a standard weighted resampling 
-algorithm with replacement.
+These weights are then used to perform a weighted resample with replacement.
 
@@
 
-Want to find the state,
+Want to find the state of all $M$ segments in the network,
 \begin{equation}
   \label{eq:kf_state}
-  \Theta = [\theta_1, \cdots, \theta_L]^T\quad\text{and}\quad
-  \Xi = \diag(\xi_1, \cdots, \xi_L) = \bxi\eye_L,
+  \bbeta = [\beta_1, \cdots, \beta_M]^T\quad\text{and}\quad
 \end{equation}
-which are the mean and variance of each segement, respectively.
+which is the travel time along each segment,
+along with the variance-covariance matrix $\Sigma$.
+For now, we are assuming segments are independent,
+so $\Sigma$ is a diagonal matrix.
 
@@
 since it has a low cost, allowing the most recent data to be incorporated.
-For each vehicle $c$, we have $N_c$ particles which we use to
-compute the mean distance into trip and velocity,
+For each vehicle $v$, we have $I_v$ particles which we use to
+compute the mean distance into trip,
 \begin{equation}
   \label{eq:weighted_mean}
-  \bar d_c = \sum_{i=1}^{N_c} w_c^{(i)} d_c^{(i)} \quad\text{and}\quad
-  \bar v_c = \sum_{i=1}^{N_c} w_c^{(i)} v_c^{(i)},
+  \bar x_v = \sum_{i=1}^{I_v} w_v^{(i)} x_v^{(i)}
 \end{equation}
-as well as the weighted sample variance,
+which we can use to estimate which segment the vehicle is in
+(using the same method as in the initialisation step).
+
+When the vehicle passes through a segment, $r_\ell$,
+we set the travel time of the vehicle using the 
+weighted mean and sample variance of the particles,
 \begin{equation}
-  \label{eq:weighted_variance}
-  \bar S_c^2 = \sum_{i=1}^{N_c} w_c^{(i)} (v_c^{(i)} - \bar v_c)^2,
+  \label{eq:weighted_travel}
+  \bar b_{v,\ell} = \sum_{i=1}^{I_v} w_v^{(i)} b_{v\ell}^{(i)} \quad\text{and}\quad
+  \bar S_{v,\ell}^2 = \sum_{i=1}^{I_v} w_v^{(i)}(b_{v\ell}^{(i)} - \bar b_{v\ell})
 \end{equation}
-since $\sum_{i=1}^{N_c} w_c^{(i)} = 1\quad \forall c = 1,\ldots,C$.
 
-Using $\bar d_c$, we can find the segment id corresponding to each busses most probable location,
-and then combine all vehicle speed estimates for each segment.
-
-In situations where multiple vehicles are in the same segment $\ell$, 
-we can simply combine the estimates, using $C_\ell = $ number of vehicles in segment $\ell$ 
-\cite{cn}.
+Having processed all vehicles, we move on to updating each \emph{segment},
+$R_m$, $m = 1, \ldots, M$.
+Define
+\begin{equation*}
+  V_m = \{v:\text{ vehicle $v$ passed through segment } R_m\}
+\end{equation*}
+and let $K_m$ be the length of $V_m$,
+and $\ell(R_m)$ be the segment index of the given stop
+(i.e., $\ell(r_\ell) = \ell$).
 \begin{equation}
   \label{eq:kf_data_estimates}
-  \hat V_\ell = \frac{1}{C_\ell} \sum_{c=1}^{C_\ell} \bar v_j \quad\text{and}\quad
-  \hat P_\ell = \frac{1}{C_\ell} \sum_{j=1}^{C_\ell} \left(\bar v_c^2 + \bar S_c^2\right) - \hat V_\ell^2
+  \hat B_m = \frac{1}{K_m} \sum_{v \in V_m} \bar b_{v,\ell(R_m)} 
+  \quad\text{and}\quad
+  \hat Q_m = \frac{1}{K_m} \sum_{v \in V_m} 
+  \left(\bar b_{v\ell(R_m)}^2 + \bar S_{v\ell(R_m)}^2\right) - \hat B_m^2
 \end{equation}
 
-Next is just the simple matter of applying the Kalman filter algorithm \cite{Anderson_2012} 
+Next is just the simple matter of applying the Kalman filter algorithm 
+\cite{Anderson_2012} 
 to the transport road network state,
-using the data \mbox{$\hat\bV_k = [\hat V_1, \ldots, \hat V_L]$} as the observation vector
-and \mbox{$\hat\bP_k = \diag(\hat P_1, \ldots, \hat P_L)$},
+using the data \mbox{$\hat {\mat B}_m = [\hat B_1, \ldots, \hat B_M]$} 
+as the observation vector
+and \mbox{$\hat {\mat Q}_k = \diag(\hat Q_1, \ldots, \hat Q_M)$},
 as the measurement error matrix.
@@
 \begin{align}
-  \Theta_{k|k-1} &= \Theta_{k-1|k-1} \\
-  \Xi_{k|k-1} &= \Xi_{k-1|k-1} + \bW_k
+  \bbeta_{c|c-1} &= \bbeta_{c-1|c-1} \\
+  \Sigma_{c|c-1} &= \Sigma_{c-1|c-1} + \bW_c
 \end{align}
-where $\bW_k \sim \mathcal{N}(0, \Delta_k\zeta^2\eye_L)$, $\zeta^2$ is the system noise parameter,
-and $\Delta_k = t_k - t_{k-1}$.
+where $\bW_c \sim \mathcal{N}(0, \delta_c\xi^2\eye_M)$, 
+$\xi^2$ is the system noise parameter,
+and $\delta_c = t_c - t_{c-1}$ is the time since the segment
+was last updated.
 
@@
 \begin{align}
-  \bK_k &= \frac{\Xi_{k|k-1}}{\Xi_{k|k-1} + \hat\bP_k} \\
-  \Theta_{k|k} &= \Theta_{k|k-1} + \bK_k(\bV_k - \Theta_{k|k-1}) \\
-  \Xi_{k|k} &= (\eye_L - \bK_k) \Xi_{k|k-1}
+  \bK_c &= \frac{\Sigma_{c|c-1}}{\Sigma_{c|c-1} + \hat {\mat Q}_c} \\
+  \bbeta_{c|c} &= \bbeta_{c|c-1} + \bK_c(\hat {\mat B}_c - \bbeta_{c|c-1}) \\
+  \Sigma_{c|c} &= (\eye_M - \bK_c) \Sigma_{c|c-1}
 \end{align}
@@
 \begin{tabular}{ll}
+  &\textbf{Data} \\
+  $t_k$, $t_c$ & time at observation $k$, $c$ \\
+  $\delta$ & time since last observation \\
+  $\bY_k$ & observation $k$ \\
+  $\tilde y_k$ & GPS position of vehicle (degrees) \\
+  $\phi,\lambda$ & latitude, longitude (radians) \\
+  $T^a_j, T^d_j$ & arrival, departure time at stop $j \in \{1, \ldots, J\}$ \\
+  $\mat S$ & stop information (position) \\
+  $\mat s$ & stop information conditional on route (distance) \\
+  $\mat R$ & segment information (path) \\
+  $\mat r$ & segment information conditional on route 
+             (distance traveled, length) \\
+  &\textbf{Parameters} \\
+  $\pi$ & probability of stopping at a stop \\
+  $\gamma$ & minimum dwell time (seconds) \\
+  $\tau$ & mean dwell time (seconds) \\
+  $\rho$ & probability of stopping at an intersection \\
+  $\theta$ & mean queue time (seconds) at intersection \\
+  $\beta$ & mean travel time (seconds) for segment \\
+  $\bbeta$, $\Sigma$ & transit network state (mean, covariance) \\
+  $\eta$ & stop size (m) \\
+  $\kappa$ & intersection size (m) \\
+  $\sigma_x$ & system noise (variability in speed) \\
+  $\sigma_y$ & GPS error \\
+  $\varepsilon$ & arrival/departure time error \\
+  $\xi$ & system noise (Kalman filter) \\
+  &\textbf{Variables} \\
+  $\bX_k$ & vehicle state associated with obs $k$ \\
+  $x$, $\dot x$ & vehicle distance into trip, velocity \\
+  $j$, $\ell$ & stop, segment index \\
+  $A_j$, $\tilde D_j$, $D_j$ & arrival, dwell, departure time \\
+  $q_j$, $b_j$ & queue, travel time \\
+  $z$ & computed GPS position of particle \\
+  $w$ & particle weight \\
+  $V$ & number of vehicles \\
+  $I$ & number of particles (per vehicle) \\
+  $J$ & number of stops (conditional on route) \\
+  $L$ & number of segments (conditional on route) \\
+  $N$ & total number of stops in network \\
+  $M$ & total number of segments in network \\
+  &\textbf{Functions} \\
+
+\end{tabular}
+
+\begin{tabular}{ll}
   $N_c$ & number of particles for vehicle $c$ \\
@@
   $L$ ($L_r$) & number of segments (for given route $r$) \\
-  $\bS^d_r, \bR^d_r$ & stop, intersection distances along route $r$ \\
-  $t_k$ & time at observation $k$ \\
-  $\bY_k$ & observation $k$ \\
-  $\phi,\lambda$ & latitude, latitude \\
-  $T^a_j, T^d_j$ & arrival, departure time at stop $j \in \{1, \ldots, M_s\}$ \\
-  $\bX_k$ & vehicle state associated with obs $k$ \\
-  $d$ & distance into trip (meters) \\
-  $v$ & velocity (speed, meters per second) \\
-  $s$ & stop index $\in \{1, \ldots, M_s\}$ \\
-  $A_s, D_s$ & arrival, departure time at stop $s$ \\
-  $r$ & segment index $\in \{1, \ldots, M_r\}$ \\
-  $Q_r, B_r$ & queue time (seconds), begin time for segment $r$ \\
-  $z$ & computed GPS position of a particle \\
-  $\delta$ & time since last observation (seconds) \\
-  $\pi$ & probability of stopping at a stop \\
-  $\gamma$ & minimum dwell time (seconds) \\
-  $\tau$ & mean dwell time (seconds) \\
-  $\rho$ & probability of stopping at an intersection \\
-  $\kappa$ & mean queue time at intersection \\
-  $\sigma_v^2$ & system noise (particle filter) \\
-  $\sigma_r^2$ & between segment velocity variability \\
-  $\varepsilon_y^2, \varepsilon_d^2$ & GPS, arrival/departure time error \\
-  $\varepsilon_S, \varepsilon_I$ & stop, intersection radius (meters) \\
-  $w$ & particle weight \\
-  $\Theta, \Xi$ & transport road state mean, variance \\
+
   $\hat V_\ell$ & mean speed of recent vehicles in segment $\ell$ \\

### 2017-07-31 realtime-filter/realtime-filter.tex
Working on symbols

@@
 \subfloat[Routes\label{fig:route_segmentation1}]{\includegraphics[width=0.5\linewidth]{figure/route_segmentation-1} }
-\subfloat[Segments\label{fig:route_segmentation2}]{\includegraphics[width=0.5\linewidth]{figure/route_segmentation-2} }\caption[Simplied diagram of overlapping routes]{Simplified unidirectional road network diagram, with (a) overlapping routes, and (b) the segments obtained by splitting the routes at intersections.
-Route 1 $ = \{r_1, r_2, r_3\}$, route 2 $=\{r_4,r_2,r_5,r_6\}$, and route 3 $=\{r_7,r_5,r_8,r_9\}$.}\label{fig:route_segmentation}
+\subfloat[Segments\label{fig:route_segmentation2}]{\includegraphics[width=0.5\linewidth]{figure/route_segmentation-2} }\caption[Simplied diagram of overlapping routes]{Simplified unidirectional road network diagram, with (a) overlapping routes, and (b) the segments obtained by splitting the routes at intersections (solid dots).
+Route 1 $ = \{R_1, R_2, R_3\}$, route 2 $=\{R_4,R_2,R_5,R_6\}$, and route 3 $=\{R_7,R_5,R_8,R_9\}$.}\label{fig:route_segmentation}
 \end{figure}
@@
   \label{eq:l1}
-  \log L_1 (\bY_k | \bX_k^{(i)}) = - \log(2\pi\sigma_y) - \frac{\|g(h(d_k^{(i)})|\tilde\bY_k)\|^2}{2\varepsilon_y^2}
+  \log L_1 (\bY_k | \bX_k^{(i)})
+  = \log L_1 (\tilde y_k | x_k^{(i)})
+  = - \log(2\pi\sigma_y) - \frac{\|g(h(x_k^{(i)})|\tilde y_k)\|^2}{2\sigma_y^2}
 \end{equation}
 
-%\subsubsection[L1b]{$L_1'$: at stop}
-\item
-If the particle is considered to be at a stop, then we apply the following likelihood.
-Define a parameter $\varepsilon_S$ as a boundary around the stop;
-if observation is within that boundary likelihood is 1, else 0 (figure~\ref{fig:likelihood_circles}a).
-Let $B_s$ be a circle with radius $\varepsilon_S$ and centered at stop $j$, then
-
-\begin{equation}
-  \label{eq:l1b}
-  \log L_1(\bY_k | \bX_k^{(i)})=
-  \begin{cases}
-    - \log(\pi \varepsilon_S^2) & \tilde\bY_k \in B_s \\
-    - \infty & \text{otherwise}
-  \end{cases}
-\end{equation}
-
-%\subsubsection[L1c]{$L_1''$: at intersection}
-\item
-If the particle is considered to be at an intersection, we apply the following likelihood.
-Similar to $L_1'$, however need to check observation is heading into the intersection
-(and not out of it) by using two boundaries around the intersection and a point $\varepsilon_I$ away
-(figure~\ref{fig:likelihood_circles}b).
-Let $B_I^1$ and $B_I^2$ be two circles with radii $\varepsilon_I$,
-the first centered on the intersection and the second centered $\varepsilon_I$ meters before it, then
-
-\begin{equation}
-  \label{eq:l1c}
-  \log L_1 (\bY_k | \bX_k^{(i)}) =
-  \begin{cases}
-    - \log(A) & \tilde\bY_k \in B_I^1 \cap B_I^2 \\
-    - \infty & \text{otherwise}
-  \end{cases}
-\end{equation}
-where $A = 2\varepsilon_I^2 \left( \cos^{-1}(\frac{1}{2}) - \frac{\sqrt{3}}{4} \right)$,
-the area of $B_I^1\cap B_I^2$.
-
-\end{enumerate}
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
-\subfloat[Stop\label{fig:likelihood_circles1}]{\includegraphics[width=0.5\linewidth]{figure/likelihood_circles-1} }
-\subfloat[Intersection\label{fig:likelihood_circles2}]{\includegraphics[width=0.5\linewidth]{figure/likelihood_circles-2} }\caption[Likelihood functions for particle's at stops and intersections]{Likelihood functions for particle's at stops and intersections.
-The grey region represents the area where particle's are accepted.}\label{fig:likelihood_circles}
-\end{figure}
-
-\end{knitrout}
-
+\subsubsection{GTFS trip updates}
 The second part of the likelihood involves the stop time update,

### 2017-07-31 realtime-filter/realtime-filter.Rnw
finished symbols;

@@
 {\textbf{End Phase}\\}
+%\usepackage{longtable}
+\usepackage{xtab,booktabs}
 
@@
 
-\title{Modelling the real-time state of a public transport road network}
+\title{Real-time modelling of a public transport network to improve arrival-time predictions}
 \author{Tom~Elliott}
@@
 the routes shape information, $z_k^{(i)} = h(d_k^{(i)})$.
-The function $g (\bZ | \tilde\bY)$ performs an \emph{equirectangular projection} \cite{Snyder_1998}
-of the coordinates $\bX$, centered on $\tilde\bY$, 
+The function $g (z | \tilde y)$ performs an \emph{equirectangular projection} 
+\cite{Snyder_1998}
+of the coordinates $z$, centered on $\tilde y$, 
 onto a flat surface so that the Euclidean distance between points is equivalent
@@
   \log L_1 (\bY_k | \bX_k^{(i)})
-  = \log L_1 (\tilde y_k | x_k^{(i)})
+%  = \log L_1 (\tilde y_k | x_k^{(i)})
   = - \log(2\pi\sigma_y) - \frac{\|g(h(x_k^{(i)})|\tilde y_k)\|^2}{2\sigma_y^2}
@@
   \log L_2 (\bY_k | \bX_k^{(i)})
-  = \log L_2 (T_j^a | A_j^{(i)}) 
+%  = \log L_2 (T_j^a | A_j^{(i)}) 
   = - \frac{1}{2}\log (2\pi\varepsilon^2) - 
@@
   \log L_2 (\bY_k | \bX_k^{(i)})
-  = \log L_2 (T_j^d | D_j^{(i)}) 
+%  = \log L_2 (T_j^d | D_j^{(i)}) 
   = - \frac{1}{2}\log (2\pi\varepsilon^2) - 
@@
   \label{eq:kf_state}
-  \bbeta = [\beta_1, \cdots, \beta_M]^T\quad\text{and}\quad
+  \bbeta = [\beta_1, \cdots, \beta_M]^T
 \end{equation}
@@
 weighted mean and sample variance of the particles,
-\begin{equation}
+\begin{align}
   \label{eq:weighted_travel}
-  \bar b_{v,\ell} = \sum_{i=1}^{I_v} w_v^{(i)} b_{v\ell}^{(i)} \quad\text{and}\quad
-  \bar S_{v,\ell}^2 = \sum_{i=1}^{I_v} w_v^{(i)}(b_{v\ell}^{(i)} - \bar b_{v\ell})
-\end{equation}
+  \bar b_{v,\ell} &= \sum_{i=1}^{I_v} w_v^{(i)} b_{v\ell}^{(i)} \\
+  \bar S_{v,\ell}^2 &= \sum_{i=1}^{I_v} w_v^{(i)}(b_{v\ell}^{(i)} - \bar b_{v\ell})
+\end{align}
 
@@
 (i.e., $\ell(r_\ell) = \ell$).
-\begin{equation}
+\begin{align}
   \label{eq:kf_data_estimates}
-  \hat B_m = \frac{1}{K_m} \sum_{v \in V_m} \bar b_{v,\ell(R_m)} 
-  \quad\text{and}\quad
-  \hat Q_m = \frac{1}{K_m} \sum_{v \in V_m} 
+  \hat B_m &= \frac{1}{K_m} \sum_{v \in V_m} \bar b_{v,\ell(R_m)}  \\
+  \hat Q_m &= \frac{1}{K_m} \sum_{v \in V_m} 
   \left(\bar b_{v\ell(R_m)}^2 + \bar S_{v\ell(R_m)}^2\right) - \hat B_m^2
-\end{equation}
+\end{align}
 
@@
 as the measurement error matrix.
+The transition matrix for the network state, $\mat F_c$,
+is currently an identity matrix,
+but in future we'll allow for the update to account for
+changes in adjacent segments.
 The Kalman filter prediction equations are
 \begin{align}
-  \bbeta_{c|c-1} &= \bbeta_{c-1|c-1} \\
+  \bbeta_{c|c-1} &= \mat F_c\bbeta_{c-1|c-1} \\
   \Sigma_{c|c-1} &= \Sigma_{c-1|c-1} + \bW_c
@@
 
-\begin{tabular}{ll}
-  &\textbf{Data} \\
+%\begin{xtabular}{lp{6cm}}
+\begin{xtabular}{ll}
+  \toprule
+  &\textbf{Data} \\ \midrule
   $t_k$, $t_c$ & time at observation $k$, $c$ \\
-  $\delta$ & time since last observation \\
+  $\delta_x$ & time since last observation, $t_x - t_{x-1}$\\
   $\bY_k$ & observation $k$ \\
@@
   $T^a_j, T^d_j$ & arrival, departure time at stop $j \in \{1, \ldots, J\}$ \\
-  $\mat S$ & stop information (position) \\
-  $\mat s$ & stop information conditional on route (distance) \\
-  $\mat R$ & segment information (path) \\
-  $\mat r$ & segment information conditional on route 
-             (distance traveled, length) \\
-  &\textbf{Parameters} \\
+  $S_n$ & unique identifier of stop $n$ \\
+  $s_j$, $s_j^x$, $s_j^p$ & unique identifier of $j$th stop (of relevent route), shape distance traveled, position \\
+  $R_m$ & unique identifier of segment $m$  \\
+  $r_\ell$, $r_\ell^x$, $r_\ell^p$ & unique identifier of $\ell$th segment (of relevent route), shape distance traveled, path \\
+  \bottomrule
+  \\ \toprule
+  &\textbf{Parameters} \\ \midrule
   $\pi$ & probability of stopping at a stop \\
@@
   $\rho$ & probability of stopping at an intersection \\
-  $\theta$ & mean queue time (seconds) at intersection \\
-  $\beta$ & mean travel time (seconds) for segment \\
+  $\theta_m$ & mean queue time (seconds) at intersection of segment $m$ \\
+  $\beta_m$ & mean travel time (seconds) for segment $m$ \\
   $\bbeta$, $\Sigma$ & transit network state (mean, covariance) \\
@@
   $\xi$ & system noise (Kalman filter) \\
-  &\textbf{Variables} \\
+  \bottomrule
+  \\ \toprule
+  &\textbf{Variables} \\ \midrule
   $\bX_k$ & vehicle state associated with obs $k$ \\
@@
   $j$, $\ell$ & stop, segment index \\
-  $A_j$, $\tilde D_j$, $D_j$ & arrival, dwell, departure time \\
+  $A_j$, $\tilde D_j$, $D_j$ & arrival, dwell, departure time at stop $j$ \\
   $q_j$, $b_j$ & queue, travel time \\
-  $z$ & computed GPS position of particle \\
+  $z^{(i)}$ & computed GPS position of particle $(i)$ \\
+  $p^{(i)}$ & indicator to whether particle $(i)$ stopped \\
   $w$ & particle weight \\
-  $V$ & number of vehicles \\
-  $I$ & number of particles (per vehicle) \\
-  $J$ & number of stops (conditional on route) \\
-  $L$ & number of segments (conditional on route) \\
-  $N$ & total number of stops in network \\
-  $M$ & total number of segments in network \\
-  &\textbf{Functions} \\
-
-\end{tabular}
-
-\begin{tabular}{ll}
-  $N_c$ & number of particles for vehicle $c$ \\
-  $C$ & number of vehicles \\
-  $C_\ell$ & number of vehicles in segment $\ell$ \\
-  $M$ ($M_r$) & number of stops (for given route $r$) \\
-  $L$ ($L_r$) & number of segments (for given route $r$) \\
-
-  $\hat V_\ell$ & mean speed of recent vehicles in segment $\ell$ \\
-  $\hat P_\ell$ & estimated population variance of vehicle speeds in segment $\ell$ \\
-  $\zeta^2$ & system noise (Kalman filter)
-\end{tabular}
+  $B_m$, $Q_m$ & mean, covariance of travel time of all recent vehicles along segment $m$ \\
+  $\mat F_c$ & network state transition matrix \\
+  $V$, $v$ & number of vehicles, indexed by $v$ \\
+  $I$, $i$ & number of particles (per vehicle), indexed by $i$ \\
+  $J$, $j$ & number of stops (conditional on route), indexed by $j$ \\
+  $L$, $\ell$ & number of segments (conditional on route), indexed by $\ell$ \\
+  $N$, $n$ & total number of stops in network, indexed by $n$ \\
+  $M$, $m$ & total number of segments in network, indexed by $m$ \\
+  $k$ & index for vehicle observations \\
+  $c$ & index for transit network state updates \\
+  \bottomrule
+  \\ \toprule
+  &\textbf{Functions} \\ \midrule
+  $f$ & vehicle transition function \\
+  $h$ & vehicle measurement function \\
+  $g$ & equirectangular projection \\
+  \bottomrule
+\end{xtabular}
 

### 2017-08-03 realtime-filter/realtime-filter.Rnw
edits

--- 2017-06-27 realtime-filter/realtime-filter.Rnw
@@
 Do this by identifying all intersections,
-which can be downloaded from OpenStreetMap \cite{OpenStreetMap_2017}
+which can be downloaded from \gls{osm} \cite{OpenStreetMap_2017}
 and segment routes at these points,
 as shown in figure~\ref{fig:route_segmentation}.
-For now, only using traffic lights as simplification.
-Modify \gls{gtfs} data so \emph{shapes} table becomes
-\texttt{shapes}, \texttt{segments}, and \texttt{segment\_shapes}.
+For now, only using traffic lights and roundabouts 
+(as these are easy to access from \gls{osm}).
+Modify \gls{gtfs} data
+by adding \texttt{intersections}, \texttt{segments}, and \texttt{shape\_segments} tables.
 We assume that if two routes both go through intersections $A$ and $B$,
@@
 (once we scale up to all intersections, this will be true almost surely).
-
-Thus as we add new routes, we don't need to create new segments
-(except at start/end\ldots see section~\ref{sec:discussion}).
+We also allow segments to begin or end at a bus stop, 
+which allows routes with the same origin or destination to share segments.
 

### 2017-08-03 realtime-filter/realtime-filter.Rnw

--- 2017-07-31 realtime-filter/realtime-filter.Rnw
@@
 Do this by identifying all intersections,
-which can be downloaded from OpenStreetMap \cite{OpenStreetMap_2017}
+which can be downloaded from \gls{osm} \cite{OpenStreetMap_2017}
 and segment routes at these points,
 as shown in figure~\ref{fig:route_segmentation}.
-For now, only using traffic lights and roundabouts as simplification.
-Modify \gls{gtfs} data so \emph{shapes} table becomes
-\texttt{shapes}, \texttt{segments}, and \texttt{segment\_shapes}.
+For now, only using traffic lights and roundabouts 
+(as these are easy to access from \gls{osm}).
+Modify \gls{gtfs} data
+by adding \texttt{intersections}, \texttt{segments}, and \texttt{shape\_segments} tables.
 We assume that if two routes both go through intersections $A$ and $B$,
 they follow the same road to do so
-(routes retain their own shape, so in cases where this \emph{is not} true,
-the shape will still be valid).
-
-Thus as we add new routes, we don't need to create new segments
-(except at start/end\ldots see section~\ref{sec:discussion}).
+(once we scale up to all intersections, this will be true almost surely).
+We also allow segments to begin or end at a bus stop, 
+which allows routes with the same origin or destination to share segments.
 
@@
 
+In our particle filter implementation,
+for a single vehicle $v$ we initialize $I$ particles,
+and using the superscript $(i) : i = 1, \ldots, I$ to denote 
+a value for a single particle.
+Each particle is initialized with a full trajectory (see below),
+from the begining of the route at time $t_o^{(i)}$ (trip start time is unknown),
+until they reach the last stop $s_J$.
+
 \subsection{Predict: transition function}
@@
 Downside is computational demand.
+
+In our chosen approach, a particle represents a full journey,
+which allows us to estimate travel times for previous segments, 
+as well as produce arrival time predictions in the same step.
+The trajectory, which is demonstrated in figure~\ref{fig:trajectory},
+is computed at each second of the journey, started at $t_0^{(i)}$.
+Each particle is indendently run through a \emph{transition} algorithm,
+outlined in figure~\ref{alg:transition}.
+
+<<trajectory,echo=FALSE,out.width='\\linewidth',fig.cap="desc",fig.scap="short desc",fig.width=10,fig.height=5>>=
+plot.new()
+plot.window(c(0, 600), c(0, 1000), xaxs = "i", yaxs = "i")
+axis(2, c(0, 400, 600, 1100), c(0, expression(s[1]^d), expression(r[1]^d), ""), las = 1, family = "serif")
+axis(1, c(0, 700), c(expression(t[0]), ""))
+mtext("Time (s)", 1, 2)
+abline(h = 400, lty = 2, col = "#999999")
+abline(h = 600, lty = "7727", lwd = 2, col = "#999999")
+
+@
+
+\vspace{1em}\hrule\vspace{1em}
 
@@
 
-%\begin{xtabular}{lp{6cm}}
 \begin{xtabular}{ll}

### 2017-08-03 realtime-filter/realtime-filter.tex

--- 2017-07-31 realtime-filter/realtime-filter.tex
@@
 Downside is computational demand.
+
+In our chosen approach, a particle represents a full journey,
+which allows us to estimate travel times for previous segments, 
+as well as produce arrival time predictions in the same step.
+The trajectory, which is demonstrated in figure~\ref{fig:trajectory},
+is computed at each second of the journey, started at $t_0^{(i)}$.
+Each particle is indendently run through a \emph{transition} algorithm,
+outlined in figure~\ref{alg:transition}.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+\includegraphics[width=\linewidth]{figure/trajectory-1} \caption[short desc]{desc}\label{fig:trajectory}
+\end{figure}
+
+\end{knitrout}
+
+\vspace{1em}\hrule\vspace{1em}
 

### 2018-02-02 realtime-filter/realtime-filter.Rnw
Update KF with EKF info

--- 2017-07-31 realtime-filter/realtime-filter.Rnw
@@
 
-Modelling busses in real-time has been done for a while now 
+Modelling buses in real-time has been done for a while now 
 \cite{Cats_2015,Chen_2014,Hans_2015,Jeong_2005,TCRP_2003,Wessel_2016,
@@
 and an additional $u$ seconds as passengers board and disembark.
-$\gamma$ is considered a fixed parameter, across all busses, 
+$\gamma$ is considered a fixed parameter, across all buses, 
 while $u$ is given an exponential distribution with mean $\tau_{s_j}$.
@@
 The idea behind using particle filter (as opposed to other models)
-is that it is well suited modelling busses \cite{Hans_2015},
+is that it is well suited modelling buses \cite{Hans_2015},
 as it has good coverage of the plausible locations and thus overcomes
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Real-time transport road network model: Kalman filter}
@@
 
+The main goal of our work is to estimate arrival times of buses,
+which requires knowledge of how long it takes a bus to
+travel along roads within the network.
+Thus, we need to estimate the \emph{road network state} at time $c$,
+\begin{equation}
+  \label{eq:kf_state}
+  \bbeta_c = [\beta^1_c, \cdots, \beta^M_c]^T
+\end{equation}
+where $\beta^j_c$ is the travel time along road segment $j$.
+The basic idea is to update this state every time a bus travels
+along a segment of road.
+This way, subsequent buses can use the most recent travel time
+information to estimate how long it will take them to reach 
+upcoming stops.
+
+Due to the simplicity of the model (a single parameter, travel time),
+we are using an \gls{ekf} algorithm; 
+we are modeling each segment independently, although in the future
+we plan to take into account adjacent segments.
+However, since we cannot expect regular observations,
+we need the state to "fall back" on our prior estimate of travel time%
+---based on historical data---over time.
+
+The first stage is to derive the transition function $f_c$ that
+trends towards the prior mean, $\mu_j(t)$, and variance, $\omega_j(t)^2$,
+of travel times along segment $j$ at time $t$ (based on historical data).
+For now, we will be using $\mu_j = \mu_j(t)$ and $\omega_j = \omega_j(t)\ \forall t$.
+We also need to allow $f_c$ to cope with irregular update intervals 
+($\Delta_c = $ time since segment was last updated).
+
+We first define a one-second transition function $f$ such that the state mean
+converges towards the prior mean,
+\begin{equation}
+  \label{eq:kf_trans1}
+  \beta_{c} = f(\beta_{c-1}, \mu, \lambda) = \beta_{c-1} + \lambda(\mu - \beta_{c-1}),
+\end{equation}
+where $\lambda \in (0, 1)$ is a tuning parameter that determines how quickly
+the process converges to the prior.
+We next define $f_n$ as a recusive composition of $n$ transitions,
+\begin{equation}
+  \label{eq:kf_transf}
+f_n(\beta, \mu, \lambda) =
+(f^1\circ \cdots \circ f^n) (\beta, \mu, \lambda),\quad
+\text{where } f^j \equiv f\ \forall j.
+\end{equation}
+
+For the \gls{ekf}, we need the derivative of $f_n$, which we get via the chain rule.
+To simplify notation, we're dropping the $\mu$ and $\beta$ parameters since they constant.
+\begin{align}
+  \label{eq:kf_deriv}
+  F_n = f_n'(\beta)\nonumber
+    &= f'[(f^1\circ \cdots \circ f^{n-1})(\beta)]
+      (f^1\circ \cdots \circ f^{n-1})'(\beta) \nonumber \\
+    &= f'[f_{n-1}(\beta)] F_{n-1} \nonumber \\
+  \intertext{which, by recursion, we get}
+    &= f'[f_{n-1}(\beta)] f'[f_{n-2}(\beta)] \cdots f'[f_1(\beta)] f'(\beta) \nonumber \\
+    &= f'(\beta) \prod_{i=1}^{n-1} f'[f_i(\beta)] \nonumber \\
+  \intertext{Next we can calculate 
+  $f'(\beta, \mu, \lambda) = \frac{\partial}{\partial \beta} \beta + \lambda(\mu - \beta) = 1 - \lambda$,
+  so the derivative becomes}
+    F_n &= (1-\lambda) \prod_{i=1}^n (1 - \lambda) = (1-\lambda)^n
+\end{align}
+
+That ensure that as $n$ goes to infinity, $f_n(\beta, \mu, \lambda) \to \mu$.
+To do the same with the variance, we need to find a value for $Q$,
+the variance of system noise, which is involved in the prediction of the state variance, $\Sigma$:
+\begin{align}
+  \label{eq:kf_varpred}
+  \Sigma_{c|c-1} &= F_n^2\Sigma_{c-1|c-1} + Q \nonumber \\
+  \intertext{At equilibrium, we want $\Sigma_{c|c-1} = \Sigma_{c-1|c-1} = \omega$, so}
+  \omega &= F_n^2 \omega + Q = (1-\lambda)^{2n} \omega + Q \nonumber \\
+  \intertext{which, solving for $Q$, gives us}
+  Q &= \omega(1 - (1 - \lambda)^{2n})
+\end{align}
+
+The last step is to choose a value of $\lambda$.
+However, any fixed value will cause the system to converge quickly when the difference is large;
+instead, we want the convergence to be a function of the ratio of information held
+in the prior and posterior states. 
+For example, if we have a lot of buses traveling at the same speed along a segment, 
+we will have a lot of information and want the state estimate to remain there
+longer that if only one bus went through. 
+So we will use $\lambda = \phi \frac{\Sigma_{c-1|c-1}}{\Sigma_{c-1|c-1} + \omega}$.
+
+\vspace{2em}
+\hrule
 Now that each vehicle is modelled by a sample of particles,
-we can use the speed estimates of busses along segments to estimate
+we can use the speed estimates of buses along segments to estimate
 the speed along them, regardless of the route.
@@
 which we expect to be unimodal and approximately Gaussian
-(some busses travel faster or slower).
+(some buses travel faster or slower).
 
@@
 Want to find the state of all $M$ segments in the network,
-\begin{equation}
-  \label{eq:kf_state}
-  \bbeta = [\beta_1, \cdots, \beta_M]^T
-\end{equation}
+% \begin{equation}
+%   \label{eq:kf_state}
+%   \bbeta = [\beta_1, \cdots, \beta_M]^T
+% \end{equation}
 which is the travel time along each segment,

### 2018-02-05 realtime-filter/realtime-filter.Rnw
matrix notation and new symbols

@@
 \end{equation}
-where $\beta^j_c$ is the travel time along road segment $j$.
+where $\beta^m_c$ is the travel time along road segment $m$.
 The basic idea is to update this state every time a bus travels
@@
 
-Due to the simplicity of the model (a single parameter, travel time),
-we are using an \gls{ekf} algorithm; 
-we are modeling each segment independently, although in the future
-we plan to take into account adjacent segments.
+We are using an \gls{ekf} algorithm to model the network state,
+treating each segment independently for now, although in the future
+we plan to take into account adjacent segments
+(see section~\ref{sec:discussion}).
 However, since we cannot expect regular observations,
-we need the state to "fall back" on our prior estimate of travel time%
+we need the state to ``fall back'' on a prior travel time distribution%
 ---based on historical data---over time.
 
-The first stage is to derive the transition function $f_c$ that
-trends towards the prior mean, $\mu_j(t)$, and variance, $\omega_j(t)^2$,
-of travel times along segment $j$ at time $t$ (based on historical data).
-For now, we will be using $\mu_j = \mu_j(t)$ and $\omega_j = \omega_j(t)\ \forall t$.
-We also need to allow $f_c$ to cope with irregular update intervals 
-($\Delta_c = $ time since segment was last updated).
+The model formulation for the network state is therefore
+\begin{align}
+  \label{eq:kf_model}
+  \bbeta_c &= \mat{F}_c(\bbeta_{c-1}, \bnu_c, \bpsi_c) + \bu_c \\
+  \bB_c &= \bbeta_c + \be_c,
+\end{align}
+where $\mat{F}_c$ is the transition function at time $c$ that
+trends towards the prior mean, $\bnu_c$, at some rate determined
+by the turning parameter $\bpsi_c$.
+Note that all of these parameters are vectors with $M$-components.
 
-We first define a one-second transition function $f$ such that the state mean
+The first stage is to derive the transition function $\mat{F}_c$ that
+trends towards the prior mean, $\bnu_c$, based on historical data.
+We also need to allow $\mat{F}_c$ to cope with irregular update intervals 
+($\Delta_c = t_c - t_{c-1} = $ time since segment was last updated).
+
+We first define a one-second transition function $\mat{F}$ such that the state mean
 converges towards the prior mean,
@@
   \label{eq:kf_trans1}
-  \beta_{c} = f(\beta_{c-1}, \mu, \lambda) = \beta_{c-1} + \lambda(\mu - \beta_{c-1}),
+  \bbeta_{c} = \mat{F}_c(\bbeta_{c-1}, \bnu, \bpsi) = \bbeta_{c-1} + \bpsi_c\circ(\bnu - \bbeta_{c-1}),
 \end{equation}
-where $\lambda \in (0, 1)$ is a tuning parameter that determines how quickly
+where $\psi^m_c \in (0, 1)\ \forall m$ is a tuning parameter that determines how quickly
 the process converges to the prior.
-We next define $f_n$ as a recusive composition of $n$ transitions,
+We next define $\mat{F}_n$ as a recusive composition of $n$ transitions,
 \begin{equation}
   \label{eq:kf_transf}
-f_n(\beta, \mu, \lambda) =
-(f^1\circ \cdots \circ f^n) (\beta, \mu, \lambda),\quad
-\text{where } f^j \equiv f\ \forall j.
+\mat{F}_n(\bbeta, \bnu, \bpsi) =
+(\mat{F}^1\circ \cdots \circ \mat{F}^n) (\bbeta, \bnu, \bpsi),\quad
+\text{where } \mat{F}^j \equiv \mat{F}\ \forall j.
 \end{equation}
 
-For the \gls{ekf}, we need the derivative of $f_n$, which we get via the chain rule.
-To simplify notation, we're dropping the $\mu$ and $\beta$ parameters since they constant.
+For the \gls{ekf}, we need the derivative of $\mat{F}_n$, which we get via the chain rule.
+To simplify notation, we're dropping the $\bnu$ and $\bpsi$ parameters since they're constant,
+and also only considering each segment independently, so
 \begin{align}
   \label{eq:kf_deriv}
-  F_n = f_n'(\beta)\nonumber
-    &= f'[(f^1\circ \cdots \circ f^{n-1})(\beta)]
-      (f^1\circ \cdots \circ f^{n-1})'(\beta) \nonumber \\
-    &= f'[f_{n-1}(\beta)] F_{n-1} \nonumber \\
+  \F_n' (\bbeta) &= 
+    \begin{bmatrix}
+      F_n' (\beta^1) \\ \vdots \\ F_n' (\beta^M)
+    \end{bmatrix}\nonumber \\
+  \text{and } F_n'(\beta)\nonumber
+    &= F'[(F^1\circ \cdots \circ F^{n-1})(\beta)]
+      (F^1\circ \cdots \circ F^{n-1})'(\beta) \nonumber \\
+    &= F'[\F_{n-1}(\beta)] F'_{n-1}(\beta) \nonumber \\
   \intertext{which, by recursion, we get}
-    &= f'[f_{n-1}(\beta)] f'[f_{n-2}(\beta)] \cdots f'[f_1(\beta)] f'(\beta) \nonumber \\
-    &= f'(\beta) \prod_{i=1}^{n-1} f'[f_i(\beta)] \nonumber \\
+    &= F'[F_{n-1}(\beta)] F'[F_{n-2}(\beta)] \cdots F'[F_1(\beta)] F'(\beta) \nonumber \\
+    &= F'(\beta) \prod_{i=1}^{n-1} F'[F_i(\beta)] \nonumber \\
   \intertext{Next we can calculate 
-  $f'(\beta, \mu, \lambda) = \frac{\partial}{\partial \beta} \beta + \lambda(\mu - \beta) = 1 - \lambda$,
+  $F'(\beta, \nu, \psi) = \frac{\partial}{\partial \beta} \left[\beta + \psi(\nu - \beta)\right] = 1 - \psi$,
   so the derivative becomes}
-    F_n &= (1-\lambda) \prod_{i=1}^n (1 - \lambda) = (1-\lambda)^n
+    F'_n(\beta) &= (1-\psi) \prod_{i=1}^n (1 - \psi) = (1-\psi)^n
 \end{align}
 
-That ensure that as $n$ goes to infinity, $f_n(\beta, \mu, \lambda) \to \mu$.
-To do the same with the variance, we need to find a value for $Q$,
-the variance of system noise, which is involved in the prediction of the state variance, $\Sigma$:
+That ensure that as $n$ goes to infinity, $\F_n(\bbeta, \bnu, \bpsi) \to \bnu$.
+To do the same with the variance, we need to find a value for $\bQ$,
+the variance of system noise, which is involved in the prediction of the state variance, $\bP$:
 \begin{align}
   \label{eq:kf_varpred}
-  \Sigma_{c|c-1} &= F_n^2\Sigma_{c-1|c-1} + Q \nonumber \\
-  \intertext{At equilibrium, we want $\Sigma_{c|c-1} = \Sigma_{c-1|c-1} = \omega$, so}
-  \omega &= F_n^2 \omega + Q = (1-\lambda)^{2n} \omega + Q \nonumber \\
+  \bP_{c|c-1} &= \F_n \bP_{c-1|c-1}\F_n^T + \bQ \nonumber \\
+  \intertext{At equilibrium, we want $\bP_{c|c-1} = \bP_{c-1|c-1} = \bxi$,
+  which we can also compute independently for each segment (since the off-diagonals will be 0):}
+  \xi^m &= F'_n(\beta^m) \xi^m F'_n(\beta^m)^T + Q_{m,m} = (1-\psi^m)^{2n} \xi^m + Q_{m,m} \nonumber \\
   \intertext{which, solving for $Q$, gives us}
-  Q &= \omega(1 - (1 - \lambda)^{2n})
+  Q_{m,m} &= \xi^m(1 - (1 - \psi^m)^{2n})
 \end{align}
 
-The last step is to choose a value of $\lambda$.
+The last step is to choose a value of $\bpsi$.
 However, any fixed value will cause the system to converge quickly when the difference is large;
@@
 longer that if only one bus went through. 
-So we will use $\lambda = \phi \frac{\Sigma_{c-1|c-1}}{\Sigma_{c-1|c-1} + \omega}$.
+So we will use $\psi^m = \phi \frac{P^m_{c-1|c-1}}{P^m_{c-1|c-1} + \xi^m}$.
 

### 2018-02-07 realtime-filter/realtime-filter.Rnw
Some stuff about EKF update

@@
 the variance of system noise, which is involved in the prediction of the state variance, $\bP$:
+\begin{equation}
+  \label{eq:kf_varpred}
+  \bP_{c|c-1} = \F_n \bP_{c-1|c-1}\F_n^T + \bQ 
+\end{equation}
+At equilibrium, we want $\bP_{c|c-1} = \bP_{c-1|c-1} = \bxi$,
+which we can also compute independently for each segment (since the off-diagonals will be 0):
 \begin{align}
-  \label{eq:kf_varpred}
-  \bP_{c|c-1} &= \F_n \bP_{c-1|c-1}\F_n^T + \bQ \nonumber \\
-  \intertext{At equilibrium, we want $\bP_{c|c-1} = \bP_{c-1|c-1} = \bxi$,
-  which we can also compute independently for each segment (since the off-diagonals will be 0):}
-  \xi^m &= F'_n(\beta^m) \xi^m F'_n(\beta^m)^T + Q_{m,m} = (1-\psi^m)^{2n} \xi^m + Q_{m,m} \nonumber \\
+  \xi^m &= F'_n(\beta^m) \xi^m F'_n(\beta^m)^T + Q_{m,m}\\
+    &= (1-\psi^m)^{2n} \xi^m + Q_{m,m} \nonumber \\
   \intertext{which, solving for $Q$, gives us}
@@
 longer that if only one bus went through. 
-So we will use $\psi^m = \phi \frac{P^m_{c-1|c-1}}{P^m_{c-1|c-1} + \xi^m}$.
+So we will use $\psi^m = \phi \frac{P^m_{c-1|c-1}}{P^m_{c-1|c-1} + \xi^m}$,
+where $\phi$ is a tuning parameter.
 
-\vspace{2em}
-\hrule
-Now that each vehicle is modelled by a sample of particles,
-we can use the speed estimates of buses along segments to estimate
-the speed along them, regardless of the route.
-In future we can add correlation structure between adjacent segments.
+Now that the prediction step has been defined,
+we can consider how to update the state given a new observation;
+but first, we need to define what an "observation" is.
 
-Use Kalman filter because we have one parameter, speed, 
-which we expect to be unimodal and approximately Gaussian
-(some buses travel faster or slower).
+Due to the nature of the application,
+we will not be updating the state \emph{immediately} after a vehicle
+traverses a road segment;
+instead, we will complete the particle filter algorithm for all buses,
+and then deal with the data obtained.
 
-Use weighted speed estimates as data,
-and weighted variance as measurement error.
-
-Want to find the state of all $M$ segments in the network,
-% \begin{equation}
-%   \label{eq:kf_state}
-%   \bbeta = [\beta_1, \cdots, \beta_M]^T
-% \end{equation}
-which is the travel time along each segment,
-along with the variance-covariance matrix $\Sigma$.
-For now, we are assuming segments are independent,
-so $\Sigma$ is a diagonal matrix.
-
-We run the Kalman filter every time we run the particle filter,
-since it has a low cost, allowing the most recent data to be incorporated.
-For each vehicle $v$, we have $I_v$ particles which we use to
-compute the mean distance into trip,
-\begin{equation}
-  \label{eq:weighted_mean}
-  \bar x_v = \sum_{i=1}^{I_v} w_v^{(i)} x_v^{(i)}
-\end{equation}
-which we can use to estimate which segment the vehicle is in
-(using the same method as in the initialisation step).
-
-When the vehicle passes through a segment, $r_\ell$,
-we set the travel time of the vehicle using the 
-weighted mean and sample variance of the particles,
+For each vehicle $v$, we have $I_v$ particles,
+with weights $w_v^{(i)}, i=1\ldots I_v$, which we use to
+estimate the vehicle's travel time through segment $r_\ell$
+by using the weighted mean and sample variance:
 \begin{align}
@@
   \bar b_{v,\ell} &= \sum_{i=1}^{I_v} w_v^{(i)} b_{v\ell}^{(i)} \\
-  \bar S_{v,\ell}^2 &= \sum_{i=1}^{I_v} w_v^{(i)}(b_{v\ell}^{(i)} - \bar b_{v\ell})
+  \bar S_{v,\ell}^2 &= \sum_{i=1}^{I_v} w_v^{(i)}(b_{v\ell}^{(i)} - \bar b_{v\ell})^2
 \end{align}
@@
 \end{equation*}
-and let $K_m$ be the length of $V_m$,
+and let $|V_m|$ be the number of elements in $V_m$,
 and $\ell(R_m)$ be the segment index of the given stop
@@
   \label{eq:kf_data_estimates}
-  \hat B_m &= \frac{1}{K_m} \sum_{v \in V_m} \bar b_{v,\ell(R_m)}  \\
-  \hat Q_m &= \frac{1}{K_m} \sum_{v \in V_m} 
+  \hat B_m &= \frac{1}{|V_m|} \sum_{v \in V_m} \bar b_{v,\ell(R_m)}  \\
+  \hat E_m &= \frac{1}{|V_m|} \sum_{v \in V_m} 
   \left(\bar b_{v\ell(R_m)}^2 + \bar S_{v\ell(R_m)}^2\right) - \hat B_m^2
@@
 as the observation vector
-and \mbox{$\hat {\mat Q}_k = \diag(\hat Q_1, \ldots, \hat Q_M)$},
+and \mbox{$\hat {\mat E}_k = \diag(\hat E_1, \ldots, \hat E_M)$},
 as the measurement error matrix.
-The transition matrix for the network state, $\mat F_c$,
-is currently an identity matrix,
-but in future we'll allow for the update to account for
-changes in adjacent segments.
-The Kalman filter prediction equations are
+
+The update equations for the \gls{ekf} state are now
 \begin{align}
-  \bbeta_{c|c-1} &= \mat F_c\bbeta_{c-1|c-1} \\
-  \Sigma_{c|c-1} &= \Sigma_{c-1|c-1} + \bW_c
+  \bK_c &= \bP_{c|c-1}(\bP_{c|c-1} + \hat {\mat E}_c)^{-1} \\
+  \bbeta_{c|c} &= \bbeta_{c|c-1} + \bK_c(\hat {\mat B}_c - \bbeta_{c|c-1}) \\
+  \bP_{c|c} &= (\eye_M - \bK_c) \bP_{c|c-1}.
 \end{align}
-where $\bW_c \sim \mathcal{N}(0, \delta_c\xi^2\eye_M)$, 
-$\xi^2$ is the system noise parameter,
-and $\delta_c = t_c - t_{c-1}$ is the time since the segment
-was last updated.
-
-The update equations simplify to
-\begin{align}
-  \bK_c &= \frac{\Sigma_{c|c-1}}{\Sigma_{c|c-1} + \hat {\mat Q}_c} \\
-  \bbeta_{c|c} &= \bbeta_{c|c-1} + \bK_c(\hat {\mat B}_c - \bbeta_{c|c-1}) \\
-  \Sigma_{c|c} &= (\eye_M - \bK_c) \Sigma_{c|c-1}
-\end{align}
-which make use of the segment means and variances computed in~(\ref{eq:kf_data_estimates}).
 

### 2018-05-27 anzjs2018/anzjs2018.tex
Start template paper

\documentclass[times, doublespace]{anzsauth}
\usepackage{moreverb}
\usepackage{url}
\usepackage{grffile}
\usepackage{lineno}
\usepackage{lipsum}
\usepackage[UKenglish]{isodate}

\def\volumeyear{2018}
\linenumbers

\begin{document}
\cleanlookdateon
\runningheads{Short title of paper}{TOM~ELLIOTT AND THOMAS~LUMLEY}
\title{Long name of the paper}
\author{Tom Elliott\corrauth and Thomas Lumley}
\affiliation{University of Auckland}
\address{
    Department of Statistics, University of Auckland, Private Bag 92019, Auckland 1142, New Zealand\\
    Email: \texttt{tom.elliott@auckland.ac.nz}
}

\begin{abstract}
Lorem ipsum blah blah blah.
\end{abstract}

\keywords{particle filter; Kalman filter; transit modeling;
          transit networks; statistical computing}

\maketitle
\section{Introduction}
\label{sec:intro}

Overview of the history, problem, proposed solution I guess...

\section{GTFS network construction}
\label{sec:gtfs}

Before we can usefully model buses in real-time, 
we need to construct a \emph{transit network}, 
consisting of \emph{intersections} (nodes),
connected by \emph{road segments} (edges).
In this way, each \emph{route} 
(a journey taken by a transit vehicle, from an origin to a destination by a fixed path)
can be represented as a sequence of road segments,
each of which we will model as vehicles travel along them (see section~\ref{sec:kf}).

\section{Real-time vehicle model}
\label{sec:pf}

The real-time vehicle model that runs ... in real time?

\section{Network model}
\label{sec:kf}

Two parts to this - first, the prior prediction step; then, the data update step.

\subsection{Prediction step}
\label{sec:kf-pred}

\subsection{Update step}
\label{sec:kf-update}

\end{document}

### 2018-05-30 anzjs2018/anzjs2018.tex
few words

@@
 \usepackage{lineno}
+% \usepackage{lmodern}
 \usepackage{lipsum}
 \usepackage[UKenglish]{isodate}
+
+\input{../glossary.tex}
 
@@
 \keywords{particle filter; Kalman filter; transit modeling;
-          transit networks; statistical computing}
+          transit networks; statistical computing; applications}
 
@@
 
-Overview of the history, problem, proposed solution I guess...
+Bus arrival-time prediction is far from a novel concept: 
+ever since the first vehicle tracking technology was implemented
+in transit vehicles in the late 1980's, 
+real-time models have been implemented to provide \gls{rti} for commuters.
+The simplest form of this is, the \gls{eta}
+(usually displayed as \emph{minutes until arrival}),
+was implemented in {{region}} by \cite{Wall99analgorithm} 
+using their proposed \gls{kf} algorithm.
+In their work, they used historical data to obtain \glspl{eta}.
+
+Over the years, advances in computing power have enabled new
+modeling approaches,
+such as support vector machines \citep
 
@@
 
-Before we can usefully model buses in real-time, 
+Before we can usefully model buses in real-tim{}e, 
 we need to construct a \emph{transit network},
-consisting of \emph{intersections} (nodes),
+consisting of \emph{intersections} (nodes)
 connected by \emph{road segments} (edges).
@@
 can be represented as a sequence of road segments,
-each of which we will model as vehicles travel along them (see section~\ref{sec:kf}).
+each of which we model as vehicles travel along them (see section~\ref{sec:kf}).
 
@@
 
+\bibliographystyle{anzsj}
+\bibliography{../reflist.bib}
+
 \end{document}

### 2018-06-11 anzjs2018/anzjs2018.tex
Wordssss :book:

@@
 modeling approaches,
-such as support vector machines \citep
+such as support vector machines \citep{Yu_2006},
+artificial neural networks \citep{Yu_2011},
+and more recently particle filters \citep{Hans_2015}.
 

### 2018-08-07 anzjs2018/anzjs2018.tex
Abstract and outline

--- 2018-05-27 anzjs2018/anzjs2018.tex
@@
 \address{
-    Department of Statistics, University of Auckland, Private Bag 92019, Auckland 1142, New Zealand\\
+    Department of Statistics, University of Auckland,\\
+    Private Bag 92019, Auckland 1142, New Zealand\\
     Email: \texttt{tom.elliott@auckland.ac.nz}
@@
 \begin{abstract}
-Lorem ipsum blah blah blah.
+Predicting transit vehicle arrival time is a somewhat complex procress.
+Road networks are dynamic, and can change from free-flowing to highly congested
+very quickly.
+Any reliable prediction framework needs to be able to respond to these changes,
+but also to future trends, such as typical changes to traffic flow
+at peak times, based on hisorical data.
+Of course, the major constraint on any prediction framework is going to be
+computational efficiency.
+Due to the realtime nature of the problem, predictions need to be 
+available as soon as possible after observing the vehicles' positions,
+which are updated approximately every 30~seconds.
+In this paper, we describe our C++ implementation of a particle filter to estimate
+vehicle state in combination with a network state model,
+allowing us to model transit vehicle flow through the network.
+While our application here is being testing in Auckland, New Zealand,
+we hope that our methods are general enough to be applied to any
+transit network that uses GTFS.
+Preliminary results show that, with suitable hardware,
+the framework is fast enough while retaining the complexity necessary
+to incorporate as much information as necessary to predict arrival times.
 \end{abstract}
@@
 \keywords{particle filter; Kalman filter; transit modeling;
-          transit networks; statistical computing}
+          transit networks; statistical computing; gtfs}
 
@@
 
-Overview of the history, problem, proposed solution I guess...
+The need for realtime information.
+Where it goes wrong: 
+- bad schedule calibration;
+- no accounting for realtime traffic 
+Literature review of historical and present methods?
 
-\section{GTFS network construction}
+Particle filter has proven useful to this kind of dynamic problem,
+where the state distribution is non-gaussian, non-symmetric, 
+multimodal...
+
+Any other recent attempts at network-estimation?
+There was a KF using point-to-point speed estimations?
+
+Computational aspect - historically, models had to be simple 
+(e.g., KF) so they could run in realtime.
+These days, running a somewhat powerful VM on a remote server
+is almost standard practice, so we can take full advantage 
+and use a more computationally demanding (particle filter)
+but more accurate estimation technique - paired with a more complex model.
+
+We first give an overview of the transit network,
+and how we construct it from raw GTFS shape data.
+This paper assumes a very basic model,
+with a focus on the implementation in a realtime setting.
+
+\section{Transit network construction}
 \label{sec:gtfs}
@@
 
-\section{Real-time vehicle model}
+\section{The Models}
+\label{sec:models}
+
+\subsection{Real-time vehicle model}
 \label{sec:pf}
@@
 
-\section{Network model}
+\subsection{Network model}
 \label{sec:kf}
@@
 
-\subsection{Prediction step}
-\label{sec:kf-pred}
+\section{Implementing in realtime}
+\label{sec:rt}
 
-\subsection{Update step}
-\label{sec:kf-update}
+How we implement it, choice of software (Rcpp = R + C++).
+R: dealing with data structures is easier, maintainability, interfacing
+C++: speed
+
+Overall structure:
+- load
+- fetch positions
+- initialize or mutate+update
+- update network
+- make ETA predictions (vehicle state + network state)
+
+Some of the key things:
+- minimise copy, parallelisation using OMP
+- moving as much computation ``outside'' of the main loop as possible
+  (e.g., ``pre''-predict vehicle/network states so only update required)
+  so ETAs are generated ASAP after retrieving data
+- keeping the GTFS database up-to-date by fetching new data each morning
+- distribution - a cloud database vs a single protobuf file with everything 
+  (maintenance/reliability/speed/size)
+
+\section{Preliminary Results}
+\label{sec:results}
+
+Currently only have timings (i.e., yes this is plausible).
+- Function of number of particles/number of cores/number of buses.
+- Network coverage? i.e., how many road segments actually get enough data
+to generate useful numbers
+
+- any problems?
+
+\section{Discussion and Future Work}
+\label{sec:discussion}
+
+It all seems to work (hopefully I don't need to change this D:)
+
+Next steps include
+- generating useful priors for ``fall back'' (i.e., no data, future) prediction
+- develop and implement a more complex network update model (correlations, etc)
+- prediction estimates (point vs interval)
+
+\cite{Hans_2015}
+
+\bibliographystyle{anzsj}
+\bibliography{refs.bib}
 

### 2018-08-07 anzjs2018/anzjs2018.tex

@@
 \keywords{particle filter; Kalman filter; transit modeling;
-          transit networks; statistical computing; gtfs}
+          transit networks; statistical computing; applications; gtfs}
 
@@
 Literature review of historical and present methods?
+
+% Bus arrival-time prediction is far from a novel concept: 
+% ever since the first vehicle tracking technology was implemented
+% in transit vehicles in the late 1980's, 
+% real-time models have been implemented to provide \gls{rti} for commuters.
+% The simplest form of this is, the \gls{eta}
+% (usually displayed as \emph{minutes until arrival}),
+% was implemented in {{region}} by \cite{Wall99analgorithm} 
+% using their proposed \gls{kf} algorithm.
+% In their work, they used historical data to obtain \glspl{eta}.
+
+% Over the years, advances in computing power have enabled new
+% modeling approaches,
+% such as support vector machines \citep{Yu_2006},
+% artificial neural networks \citep{Yu_2011},
+% and more recently particle filters \citep{Hans_2015}.
 
@@
 \bibliographystyle{anzsj}
-\bibliography{refs.bib}
+\bibliography{../reflist.bib}
 

### 2018-08-17 anzjs2018/anzjs2018.tex
change sections

@@
 \title{Long name of the paper}
-\author{Tom Elliott\corrauth and Thomas Lumley}
+\author{Tom Elliott\corrauth~and~Thomas~Lumley}
 \affiliation{University of Auckland}
@@
 
-\section{Transit network construction}
+\section{Working with GTFS data}
 \label{sec:gtfs}
 
-Before we can usefully model buses in real-tim{}e, 
+What is GTFS? Its purpose. Its accessibility (global, free).
+Thus a good thing to use.
+
+\subsection{Transit network construction}
+\label{sec:network_build}
+
+Before we can usefully model buses in real-time, 
 we need to construct a \emph{transit network},
@@
 each of which we model as vehicles travel along them (see section~\ref{sec:kf}).
+
+\subsection{Realtime vehicle locations}
+\label{sec:realtime_data}
+
+Recieve vehicle updates ~30 seconds (but the actual update frequency varies continually, even for a single bus).
+Key features of the data - position, trip, timestamp, etc.
+Also note there is trip updates feed, but this is less useful (the arrival/departure times themselves haven't proved reliable).
 

### 2018-08-17 anzjs2018/anzjs2018.tex
Update address

--- 2018-08-07 anzjs2018/anzjs2018.tex
@@
 \address{
-    Department of Statistics, University of Auckland,\\
-    Private Bag 92019, Auckland 1142, New Zealand\\
+    Department of Statistics, University of Auckland, 
+    Auckland 1142, New Zealand\\
     Email: \texttt{tom.elliott@auckland.ac.nz}
@@
 \begin{abstract}
-Predicting transit vehicle arrival time is a somewhat complex procress.
-Road networks are dynamic, and can change from free-flowing to highly congested
-very quickly.
-Any reliable prediction framework needs to be able to respond to these changes,
-but also to future trends, such as typical changes to traffic flow
-at peak times, based on hisorical data.
-Of course, the major constraint on any prediction framework is going to be
-computational efficiency.
-Due to the realtime nature of the problem, predictions need to be 
-available as soon as possible after observing the vehicles' positions,
-which are updated approximately every 30~seconds.
-In this paper, we describe our C++ implementation of a particle filter to estimate
-vehicle state in combination with a network state model,
-allowing us to model transit vehicle flow through the network.
-While our application here is being testing in Auckland, New Zealand,
-we hope that our methods are general enough to be applied to any
-transit network that uses GTFS.
-Preliminary results show that, with suitable hardware,
-the framework is fast enough while retaining the complexity necessary
-to incorporate as much information as necessary to predict arrival times.
+Lorem ipsum blah blah blah.
 \end{abstract}
@@
 \keywords{particle filter; Kalman filter; transit modeling;
-          transit networks; statistical computing; applications; gtfs}
+          transit networks; statistical computing; applications}
 
@@
 
-The need for realtime information.
-Where it goes wrong: 
-- bad schedule calibration;
-- no accounting for realtime traffic 
-Literature review of historical and present methods?
+Bus arrival-time prediction is far from a novel concept: 
+ever since the first vehicle tracking technology was implemented
+in transit vehicles in the late 1980's, 
+real-time models have been implemented to provide \gls{rti} for commuters.
+The simplest form of this is, the \gls{eta}
+(usually displayed as \emph{minutes until arrival}),
+was implemented in {{region}} by \cite{Wall99analgorithm} 
+using their proposed \gls{kf} algorithm.
+In their work, they used historical data to obtain \glspl{eta}.
 
-% Bus arrival-time prediction is far from a novel concept: 
-% ever since the first vehicle tracking technology was implemented
-% in transit vehicles in the late 1980's, 
-% real-time models have been implemented to provide \gls{rti} for commuters.
-% The simplest form of this is, the \gls{eta}
-% (usually displayed as \emph{minutes until arrival}),
-% was implemented in {{region}} by \cite{Wall99analgorithm} 
-% using their proposed \gls{kf} algorithm.
-% In their work, they used historical data to obtain \glspl{eta}.
+Over the years, advances in computing power have enabled new
+modeling approaches,
+such as support vector machines \citep{Yu_2006},
+artificial neural networks \citep{Yu_2011},
+and more recently particle filters \citep{Hans_2015}.
 
-% Over the years, advances in computing power have enabled new
-% modeling approaches,
-% such as support vector machines \citep{Yu_2006},
-% artificial neural networks \citep{Yu_2011},
-% and more recently particle filters \citep{Hans_2015}.
-
-Particle filter has proven useful to this kind of dynamic problem,
-where the state distribution is non-gaussian, non-symmetric, 
-multimodal...
-
-Any other recent attempts at network-estimation?
-There was a KF using point-to-point speed estimations?
-
-Computational aspect - historically, models had to be simple 
-(e.g., KF) so they could run in realtime.
-These days, running a somewhat powerful VM on a remote server
-is almost standard practice, so we can take full advantage 
-and use a more computationally demanding (particle filter)
-but more accurate estimation technique - paired with a more complex model.
-
-We first give an overview of the transit network,
-and how we construct it from raw GTFS shape data.
-This paper assumes a very basic model,
-with a focus on the implementation in a realtime setting.
-
-\section{Transit network construction}
+\section{GTFS network construction}
 \label{sec:gtfs}
@@
 
-\section{The Models}
-\label{sec:models}
-
-\subsection{Real-time vehicle model}
+\section{Real-time vehicle model}
 \label{sec:pf}
@@
 
-\subsection{Network model}
+\section{Network model}
 \label{sec:kf}
@@
 
-\section{Implementing in realtime}
-\label{sec:rt}
+\subsection{Prediction step}
+\label{sec:kf-pred}
 
-How we implement it, choice of software (Rcpp = R + C++).
-R: dealing with data structures is easier, maintainability, interfacing
-C++: speed
-
-Overall structure:
-- load
-- fetch positions
-- initialize or mutate+update
-- update network
-- make ETA predictions (vehicle state + network state)
-
-Some of the key things:
-- minimise copy, parallelisation using OMP
-- moving as much computation ``outside'' of the main loop as possible
-  (e.g., ``pre''-predict vehicle/network states so only update required)
-  so ETAs are generated ASAP after retrieving data
-- keeping the GTFS database up-to-date by fetching new data each morning
-- distribution - a cloud database vs a single protobuf file with everything 
-  (maintenance/reliability/speed/size)
-
-\section{Preliminary Results}
-\label{sec:results}
-
-Currently only have timings (i.e., yes this is plausible).
-- Function of number of particles/number of cores/number of buses.
-- Network coverage? i.e., how many road segments actually get enough data
-to generate useful numbers
-
-- any problems?
-
-\section{Discussion and Future Work}
-\label{sec:discussion}
-
-It all seems to work (hopefully I don't need to change this D:)
-
-Next steps include
-- generating useful priors for ``fall back'' (i.e., no data, future) prediction
-- develop and implement a more complex network update model (correlations, etc)
-- prediction estimates (point vs interval)
-
-\cite{Hans_2015}
+\subsection{Update step}
+\label{sec:kf-update}
 

### 2018-08-17 anzjs2018/anzjs2018.tex
Wrote a couple paragraphs

--- 2018-08-17 anzjs2018/anzjs2018.tex
@@
 - no accounting for realtime traffic 
+- mention how only the delays are used (in Akl), which rely on (often inaccurate) schedules
 Literature review of historical and present methods?
@@
 
-What is GTFS? Its purpose. Its accessibility (global, free).
-Thus a good thing to use.
+GTFS (general transit feed specification)
+is an API (application programming interface) specification for transit data,
+developed and maintained by Google \citep{GoogleDevelopers_2006}.
+It is used by over 900~transit providers around the world,
+including here in Auckland, New Zealand
+(source \url{http://transitfeeds.com}),
+allowing an application developed locally to
+be deployed to another GTFS-based public transport system with minimal modification.
+
+The two main components of GTFS are \emph{static} and \emph{realtime}.
+Static GTFS includes information about routes, trips, shapes, and stops:
+a \emph{route} is a sequence of two or more stops displayed as a single service;
+a \emph{trip} is an instance of a route occuring at a specific time of day;
+a \emph{shape} is the sequence of points defining a vehicle's path along a route;
+and a \emph{stop} is a physical location where passengers can embark and disembark
+the vehicle.
+GTFS realtime feeds provide information such as vehicle positions and trip delays,
+and are usually accessed via an API.
+% The latter is often the only covariate used for predicting arrival time,
+% as previously discussed:
+% while it is a computationally inexpensive method (simply addition),
+% it relies on accurate schedules and cannot respond quickly to realtime events.
 
@@
 
-Before we can usefully model buses in real-time, 
-we need to construct a \emph{transit network},
-consisting of \emph{intersections} (nodes)
-connected by \emph{road segments} (edges).
-In this way, each \emph{route} 
-(a journey taken by a transit vehicle, from an origin to a destination by a fixed path)
-can be represented as a sequence of road segments,
-each of which we model as vehicles travel along them (see section~\ref{sec:kf}).
+Arguably one of the most important predictors of arrival time is
+the travel time along intermediate roads,
+however in most applications this vital information is unavailable,
+at least directly. 
+Several predictive approaches used \emph{headway},
+the time between consecutive trips along the same route
+\citep{Hans_2015}.
+While this is reasonable for high frequency routes,
+low frequncy routes will be unable to react quickly to changes in congestion,
+and it is for these routes which reliable ETAs are arguably more important,
+since the cost of missing a bus is greater.
+
+One solution would be to use information obtained from
+vehicles servicing other routes but traveling along the same roads
+to estimate arrival times;
+however, there is no direct way of knowing which routes share common roads.
+We are propose using an algorithm to convert the raw GTFS shape files
+into a \emph{road network},
+as demonstated in Figure~\ref{fig:network_creation}.
+This involves detecting locations where one or more routes meet 
+(intersections or \emph{nodes}),
+and the connecting paths (road segments or \emph{edges}).
+
+% Before we can usefully model buses in real-time, 
+% we need to construct a \emph{transit network},
+% consisting of \emph{intersections} (nodes)
+% connected by \emph{road segments} (edges).
+% In this way, each \emph{route} 
+% (a journey taken by a transit vehicle, from an origin to a destination by a fixed path)
+% can be represented as a sequence of road segments,
+% each of which we model as vehicles travel along them (see section~\ref{sec:kf}).
 

### 2018-08-21 anzjs2018/anzjs2018.tex
paragraph

@@
 
-The real-time vehicle model that runs ... in real time?
+To generate arrival time predictions, 
+several aspects of the vehicle's \emph{state} need to first be estimated,
+such as its speed.
+The \emph{unmeasurable state} of a vehicle at time $t_k$ is denoted by
+\begin{equation}
+    X_k = \begin{bmatrix} x_k & \dot x_k & \ddot x_k \end{bmatrix}^\top,
+\end{equation}
+where $x$ is the distance traveled along a route,
+$\dot x$ is the speed (the first derivative of distance traveled as a function of time),
+and $\ddot x$ is acceleration (the second derivative).
+
+The next state is modeled as a function of the current state,
+using a transition function $f$
 

### 2018-08-21 anzjs2018/anzjs2018.tex
more words

--- 2018-08-17 anzjs2018/anzjs2018.tex
@@
 
-Recieve vehicle updates ~30 seconds (but the actual update frequency varies continually, even for a single bus).
-Key features of the data - position, trip, timestamp, etc.
-Also note there is trip updates feed, but this is less useful (the arrival/departure times themselves haven't proved reliable).
+GTFS realtime allows developers to query the current positions of vehicles
+in the transit network.
+The data consists of the time $t_k$ that the observation was made,
+the GPS position of the vehicle, $Y_k$, 
+and some other information about the trip being serviced.
+Vehicle positions are usually updated within 30~seconds,
+but this can vary from 10~seconds to several minutes,
+so there is often a lot of uncertainty about the trajectory
+between two observations, particularly when there is a bus stop
+or intersection between them.
+
+Another complication with the Auckland Transport realtime feed is that
+the buses are programmed to report their location when arriving at
+bus stops and some major intersections.
+Often these positions appear to be preemptive (i.e., that bus is almost there, 
+but not quiet),
+and subsequent observations place the bus \emph{behind} the stop or intersection.
+This must therefore be accounted for in the model proposed in the next section.
 
@@
 
-The real-time vehicle model that runs ... in real time?
+In order to estimate vehicle speed from a series of noisy GPS observations,
+we need a model capable of overcoming several key problems:
+varying update intervals, multimodality, assymetry, 
+and some other artefacts such as mentioned in Section~\ref{sec:realtime_data}.
+One estimation technique that suits itself well to this type of problem
+is the particle filter, 
+which has been used in several other bus modeling applications
+\citep{Hans_2015}.
+The primary advantage of this approach is that we need not make any 
+unnecessary assumptions, particularly on the shape of the distributions 
+as is the case in other commonly used methods such as the Kalman filter.
 

### 2018-08-21 anzjs2018/anzjs2018.tex

@@
 
+To generate arrival time predictions, 
+several aspects of the vehicle's \emph{state} need to first be estimated,
+such as its speed.
+The \emph{unmeasurable state} of a vehicle at time $t_k$ is denoted by
+\begin{equation}
+    X_k = \begin{bmatrix} x_k & \dot x_k & \ddot x_k \end{bmatrix}^\top,
+\end{equation}
+where $x$ is the distance traveled along a route,
+$\dot x$ is the speed (the first derivative of distance traveled as a function of time),
+and $\ddot x$ is acceleration (the second derivative).
+
+The next state is modeled as a function of the current state,
+using a transition function $f$
+
 \subsection{Network model}

### 2018-08-27 anzjs2018/anzjs2018.tex
Restructure project

--- 2018-08-21 anzjs2018/anzjs2018.tex
@@
 \maketitle
-\section{Introduction}
-\label{sec:intro}
-
-The need for realtime information.
-Where it goes wrong: 
-- bad schedule calibration;
-- no accounting for realtime traffic 
-- mention how only the delays are used (in Akl), which rely on (often inaccurate) schedules
-Literature review of historical and present methods?
-
-% Bus arrival-time prediction is far from a novel concept: 
-% ever since the first vehicle tracking technology was implemented
-% in transit vehicles in the late 1980's, 
-% real-time models have been implemented to provide \gls{rti} for commuters.
-% The simplest form of this is, the \gls{eta}
-% (usually displayed as \emph{minutes until arrival}),
-% was implemented in {{region}} by \cite{Wall99analgorithm} 
-% using their proposed \gls{kf} algorithm.
-% In their work, they used historical data to obtain \glspl{eta}.
-
-% Over the years, advances in computing power have enabled new
-% modeling approaches,
-% such as support vector machines \citep{Yu_2006},
-% artificial neural networks \citep{Yu_2011},
-% and more recently particle filters \citep{Hans_2015}.
-
-Particle filter has proven useful to this kind of dynamic problem,
-where the state distribution is non-gaussian, non-symmetric, 
-multimodal...
-
-Any other recent attempts at network-estimation?
-There was a KF using point-to-point speed estimations?
-
-Computational aspect - historically, models had to be simple 
-(e.g., KF) so they could run in realtime.
-These days, running a somewhat powerful VM on a remote server
-is almost standard practice, so we can take full advantage 
-and use a more computationally demanding (particle filter)
-but more accurate estimation technique - paired with a more complex model.
-
-We first give an overview of the transit network,
-and how we construct it from raw GTFS shape data.
-This paper assumes a very basic model,
-with a focus on the implementation in a realtime setting.
-
-\section{Working with GTFS data}
-\label{sec:gtfs}
-
-GTFS (general transit feed specification)
-is an API (application programming interface) specification for transit data,
-developed and maintained by Google \citep{GoogleDevelopers_2006}.
-It is used by over 900~transit providers around the world,
-including here in Auckland, New Zealand
-(source \url{http://transitfeeds.com}),
-allowing an application developed locally to
-be deployed to another GTFS-based public transport system with minimal modification.
-
-The two main components of GTFS are \emph{static} and \emph{realtime}.
-Static GTFS includes information about routes, trips, shapes, and stops:
-a \emph{route} is a sequence of two or more stops displayed as a single service;
-a \emph{trip} is an instance of a route occuring at a specific time of day;
-a \emph{shape} is the sequence of points defining a vehicle's path along a route;
-and a \emph{stop} is a physical location where passengers can embark and disembark
-the vehicle.
-GTFS realtime feeds provide information such as vehicle positions and trip delays,
-and are usually accessed via an API.
-% The latter is often the only covariate used for predicting arrival time,
-% as previously discussed:
-% while it is a computationally inexpensive method (simply addition),
-% it relies on accurate schedules and cannot respond quickly to realtime events.
-
-\subsection{Transit network construction}
-\label{sec:network_build}
-
-Arguably one of the most important predictors of arrival time is
-the travel time along intermediate roads,
-however in most applications this vital information is unavailable,
-at least directly. 
-Several predictive approaches used \emph{headway},
-the time between consecutive trips along the same route
-\citep{Hans_2015}.
-While this is reasonable for high frequency routes,
-low frequncy routes will be unable to react quickly to changes in congestion,
-and it is for these routes which reliable ETAs are arguably more important,
-since the cost of missing a bus is greater.
-
-One solution would be to use information obtained from
-vehicles servicing other routes but traveling along the same roads
-to estimate arrival times;
-however, there is no direct way of knowing which routes share common roads.
-We are propose using an algorithm to convert the raw GTFS shape files
-into a \emph{road network},
-as demonstated in Figure~\ref{fig:network_creation}.
-This involves detecting locations where one or more routes meet 
-(intersections or \emph{nodes}),
-and the connecting paths (road segments or \emph{edges}).
-
-% Before we can usefully model buses in real-time, 
-% we need to construct a \emph{transit network},
-% consisting of \emph{intersections} (nodes)
-% connected by \emph{road segments} (edges).
-% In this way, each \emph{route} 
-% (a journey taken by a transit vehicle, from an origin to a destination by a fixed path)
-% can be represented as a sequence of road segments,
-% each of which we model as vehicles travel along them (see section~\ref{sec:kf}).
-
-\subsection{Realtime vehicle locations}
-\label{sec:realtime_data}
-
-Recieve vehicle updates ~30 seconds (but the actual update frequency varies continually, even for a single bus).
-Key features of the data - position, trip, timestamp, etc.
-Also note there is trip updates feed, but this is less useful (the arrival/departure times themselves haven't proved reliable).
-
-\section{The Models}
-\label{sec:models}
-
-\subsection{Real-time vehicle model}
-\label{sec:pf}
-
-To generate arrival time predictions, 
-several aspects of the vehicle's \emph{state} need to first be estimated,
-such as its speed.
-The \emph{unmeasurable state} of a vehicle at time $t_k$ is denoted by
-\begin{equation}
-    X_k = \begin{bmatrix} x_k & \dot x_k & \ddot x_k \end{bmatrix}^\top,
-\end{equation}
-where $x$ is the distance traveled along a route,
-$\dot x$ is the speed (the first derivative of distance traveled as a function of time),
-and $\ddot x$ is acceleration (the second derivative).
-
-The next state is modeled as a function of the current state,
-using a transition function $f$
-
-\subsection{Network model}
-\label{sec:kf}
-
-Two parts to this - first, the prior prediction step; then, the data update step.
-
-\section{Implementing in realtime}
-\label{sec:rt}
-
-How we implement it, choice of software (Rcpp = R + C++).
-R: dealing with data structures is easier, maintainability, interfacing
-C++: speed
-
-Overall structure:
-- load
-- fetch positions
-- initialize or mutate+update
-- update network
-- make ETA predictions (vehicle state + network state)
-
-Some of the key things:
-- minimise copy, parallelisation using OMP
-- moving as much computation ``outside'' of the main loop as possible
-  (e.g., ``pre''-predict vehicle/network states so only update required)
-  so ETAs are generated ASAP after retrieving data
-- keeping the GTFS database up-to-date by fetching new data each morning
-- distribution - a cloud database vs a single protobuf file with everything 
-  (maintenance/reliability/speed/size)
-
-\section{Preliminary Results}
-\label{sec:results}
-
-Currently only have timings (i.e., yes this is plausible).
-- Function of number of particles/number of cores/number of buses.
-- Network coverage? i.e., how many road segments actually get enough data
-to generate useful numbers
-
-- any problems?
-
-\section{Discussion and Future Work}
-\label{sec:discussion}
-
-It all seems to work (hopefully I don't need to change this D:)
-
-Next steps include
-- generating useful priors for ``fall back'' (i.e., no data, future) prediction
-- develop and implement a more complex network update model (correlations, etc)
-- prediction estimates (point vs interval)
-
-\cite{Hans_2015}
-
+\input{sections/01_introduction}
+\input{sections/02_gtfs}
+\input{sections/03_models}
+\input{sections/04_implement}
+\input{sections/05_results}
+\input{sections/06_discuss}
 \bibliographystyle{anzsj}

### 2018-08-27 anzjs2018/sections/01_introduction.tex
Restructure project

\section{Introduction}
\label{sec:intro}

The need for realtime information.
Where it goes wrong: 
- bad schedule calibration;
- no accounting for realtime traffic 
- mention how only the delays are used (in Akl), which rely on (often inaccurate) schedules
Literature review of historical and present methods?

% Bus arrival-time prediction is far from a novel concept: 
% ever since the first vehicle tracking technology was implemented
% in transit vehicles in the late 1980's, 
% real-time models have been implemented to provide \gls{rti} for commuters.
% The simplest form of this is, the \gls{eta}
% (usually displayed as \emph{minutes until arrival}),
% was implemented in {{region}} by \cite{Wall99analgorithm} 
% using their proposed \gls{kf} algorithm.
% In their work, they used historical data to obtain \glspl{eta}.

% Over the years, advances in computing power have enabled new
% modeling approaches,
% such as support vector machines \citep{Yu_2006},
% artificial neural networks \citep{Yu_2011},
% and more recently particle filters \citep{Hans_2015}.

Particle filter has proven useful to this kind of dynamic problem,
where the state distribution is non-gaussian, non-symmetric, 
multimodal...

Any other recent attempts at network-estimation?
There was a KF using point-to-point speed estimations?

Computational aspect - historically, models had to be simple 
(e.g., KF) so they could run in realtime.
These days, running a somewhat powerful VM on a remote server
is almost standard practice, so we can take full advantage 
and use a more computationally demanding (particle filter)
but more accurate estimation technique - paired with a more complex model.

We first give an overview of the transit network,
and how we construct it from raw GTFS shape data.
This paper assumes a very basic model,
with a focus on the implementation in a realtime setting.

### 2018-08-27 anzjs2018/sections/02_gtfs.tex
Restructure project

\section{Working with GTFS data}
\label{sec:gtfs}

GTFS (general transit feed specification)
is an API (application programming interface) specification for transit data,
developed and maintained by Google \citep{GoogleDevelopers_2006}.
It is used by over 900~transit providers around the world,
including here in Auckland, New Zealand
(source \url{http://transitfeeds.com}),
allowing an application developed locally to
be deployed to another GTFS-based public transport system with minimal modification.

The two main components of GTFS are \emph{static} and \emph{realtime}.
Static GTFS includes information about routes, trips, shapes, and stops:
a \emph{route} is a sequence of two or more stops displayed as a single service;
a \emph{trip} is an instance of a route occuring at a specific time of day;
a \emph{shape} is the sequence of points defining a vehicle's path along a route;
and a \emph{stop} is a physical location where passengers can embark and disembark
the vehicle.
GTFS realtime feeds provide information such as vehicle positions and trip delays,
and are usually accessed via an API.
% The latter is often the only covariate used for predicting arrival time,
% as previously discussed:
% while it is a computationally inexpensive method (simply addition),
% it relies on accurate schedules and cannot respond quickly to realtime events.

\subsection{Transit network construction}
\label{sec:network_build}

Arguably one of the most important predictors of arrival time is
the travel time along intermediate roads,
however in most applications this vital information is unavailable,
at least directly. 
Several predictive approaches used \emph{headway},
the time between consecutive trips along the same route
\citep{Hans_2015}.
While this is reasonable for high frequency routes,
low frequncy routes will be unable to react quickly to changes in congestion,
and it is for these routes which reliable ETAs are arguably more important,
since the cost of missing a bus is greater.

One solution would be to use information obtained from
vehicles servicing other routes but traveling along the same roads
to estimate arrival times;
however, there is no direct way of knowing which routes share common roads.
We are propose using an algorithm to convert the raw GTFS shape files
into a \emph{road network},
as demonstated in Figure~\ref{fig:network_creation}.
This involves detecting locations where one or more routes meet 
(intersections or \emph{nodes}),
and the connecting paths (road segments or \emph{edges}).

% Before we can usefully model buses in real-time, 
% we need to construct a \emph{transit network},
% consisting of \emph{intersections} (nodes)
% connected by \emph{road segments} (edges).
% In this way, each \emph{route} 
% (a journey taken by a transit vehicle, from an origin to a destination by a fixed path)
% can be represented as a sequence of road segments,
% each of which we model as vehicles travel along them (see section~\ref{sec:kf}).

\subsection{Realtime vehicle locations}
\label{sec:realtime_data}

GTFS realtime allows developers to query the current positions of vehicles
in the transit network.
The data consists of the time $t_k$ that the observation was made,
the GPS position of the vehicle, $Y_k$, 
and some other information about the trip being serviced.
Vehicle positions are usually updated within 30~seconds,
but this can vary from 10~seconds to several minutes,
so there is often a lot of uncertainty about the trajectory
between two observations, particularly when there is a bus stop
or intersection between them.

Another complication with the Auckland Transport realtime feed is that
the buses are programmed to report their location when arriving at
bus stops and some major intersections.
Often these positions appear to be preemptive (i.e., that bus is almost there, 
but not quiet),
and subsequent observations place the bus \emph{behind} the stop or intersection.
This must therefore be accounted for in the model proposed in the next section.

### 2018-08-27 anzjs2018/sections/03_models.tex
Restructure project

\section{The Models}
\label{sec:models}

\subsection{Real-time vehicle model}
\label{sec:pf}

In order to estimate vehicle speed from a series of noisy GPS observations,
we need a model capable of overcoming several key problems:
varying update intervals, multimodality, assymetry, 
and some other artefacts such as mentioned in Section~\ref{sec:realtime_data}.
One estimation technique that suits itself well to this type of problem
is the particle filter, 
which has been used in several other bus modeling applications
\citep{Hans_2015}.
The primary advantage of this approach is that we need not make any 
unnecessary assumptions, particularly on the shape of the distributions 
as is the case in other commonly used methods such as the Kalman filter.

To generate arrival time predictions, 
several aspects of the vehicle's \emph{state} need to first be estimated,
such as its speed.
The \emph{unmeasurable state} of a vehicle at time $t_k$ is denoted by
\begin{equation}
    X_k = \begin{bmatrix} x_k & \dot x_k & \ddot x_k \end{bmatrix}^\top,
\end{equation}
where $x$ is the distance traveled along a route,
$\dot x$ is the speed (the first derivative of distance traveled as a function of time),
and $\ddot x$ is acceleration (the second derivative).

\subsection{Network model}
\label{sec:kf}

Two parts to this - first, the prior prediction step; then, the data update step.

### 2018-08-27 anzjs2018/sections/04_implementation.tex
Restructure project

\section{Implementing in realtime}
\label{sec:rt}

How we implement it, choice of software (Rcpp = R + C++).
R: dealing with data structures is easier, maintainability, interfacing
C++: speed

Overall structure:
- load
- fetch positions
- initialize or mutate+update
- update network
- make ETA predictions (vehicle state + network state)

Some of the key things:
- minimise copy, parallelisation using OMP
- moving as much computation ``outside'' of the main loop as possible
  (e.g., ``pre''-predict vehicle/network states so only update required)
  so ETAs are generated ASAP after retrieving data
- keeping the GTFS database up-to-date by fetching new data each morning
- distribution - a cloud database vs a single protobuf file with everything 
  (maintenance/reliability/speed/size)

### 2018-08-27 anzjs2018/sections/06_discuss.tex
Restructure project

\section{Discussion and Future Work}
\label{sec:discussion}

It all seems to work (hopefully I don't need to change this D:)

Next steps include
- generating useful priors for ``fall back'' (i.e., no data, future) prediction
- develop and implement a more complex network update model (correlations, etc)
- prediction estimates (point vs interval)

### 2018-08-28 anzjs2018/sections/03_models.tex
Write some words about moels

@@
 
-In order to estimate vehicle speed from a series of noisy GPS observations,
-we need a model capable of overcoming several key problems:
-varying update intervals, multimodality, assymetry, 
-and some other artefacts such as mentioned in Section~\ref{sec:realtime_data}.
-One estimation technique that suits itself well to this type of problem
-is the particle filter, 
-which has been used in several other bus modeling applications
-\citep{Hans_2015}.
-The primary advantage of this approach is that we need not make any 
-unnecessary assumptions, particularly on the shape of the distributions 
-as is the case in other commonly used methods such as the Kalman filter.
+The vehicle model involves infering the \emph{unobservable state} $\bX_k$ 
+of a vehicle at time $t_k$ from the \emph{observed state} $\by_k$,
+which in this case is the GPS location of the vehicle,
+where $\lambda_k$ and $\phi_k$ are the longitude and latitude, respectively.
+\begin{equation}
+\label{eq:vehicle_observation}
+\by_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}
+\end{equation}
+To construct the unobservable state, 
+each vehicle is assumed to follow the trajectory defined by a trajectory function of time, $x(t)$,
+such that the distance traveled along a route by a vehicle at time $t_k$ is given by
+$x_k = x(t_k)$.
+Additionally, speed and acceleration are given by
+the first and second derivatives of the trajectory function,
+$\dot x_k = x'(t_k)$ and $\ddot x_k = x''(t_k)$, respectively.
+Exactly how the state is constructed will depend on the model being used.
 
-To generate arrival time predictions, 
-several aspects of the vehicle's \emph{state} need to first be estimated,
-such as its speed.
-The \emph{unmeasurable state} of a vehicle at time $t_k$ is denoted by
+Since observations of the vehicle are irregular, 
+a \emph{transition function} $f$ is used to
+predict the vehicle's state at time $t_k$ given the current knowledge of the state at time $t_{k-1}$,
+as well as accounting for \emph{system noise}, $\sigma_x^2$.
+We also define a \emph{measurement function} $h$ which describes the relationship between
+the observable and unobservable states,
+and accounting for \emph{measurement error}, $\bv_k$,
 \begin{equation}
-    X_k = \begin{bmatrix} x_k & \dot x_k & \ddot x_k \end{bmatrix}^\top,
+\label{eq:vehicle_measurement}
+\by_k = h(\bX_k) + \bv_k = h(x_k) + \bv_k,
 \end{equation}
-where $x$ is the distance traveled along a route,
-$\dot x$ is the speed (the first derivative of distance traveled as a function of time),
-and $\ddot x$ is acceleration (the second derivative).
+so all models need to estimate at least $x_k$, the distance traveled along the route.
+
+The model is split into two steps, prediction and update.
+The prediction involves using the transition function to predict
+the next state, $\hat\bX_{k|k-1}$,
+while the update step using a likelihood function and the measurement function
+to update the state to account for the observed vehicle location, $\by_k$,
+giving a final estimate of $\hat\bX_{k|k}$.
+
+\subsubsection{Predicting state using the transtion function}
+\label{sec:pf_prediction}
+
+The transition function is a model of bus behaviour that allows
+predictions of future states to be made given the most recently 
+updated state.
+Developing a good model for bus behaviour allows us to estimate
+parameters of interest, such as vehicle speed,
+dwell time at bus stops, and travel time along individual roads,
+the last of which is of course our primary objective.
+In this section, we propose [HOW MANY?] increasingly complex models
+of bus behaviour.
+
+%% MODEL 1
+The first model, $\bM_0$, assumes a constant speed between observations
+using the state
+$\bX_k = \left[\begin{smallmatrix}x_k && \dot x_k\end{smallmatrix}\right]^\top$
+and Gaussian system noise with mean zero and variance adjusted by
+$\Delta_k = t_k - t_{k-1}$.
+\begin{equation*}
+\hat\bX_{k|k-1} = f_0(\bX_{k-1|k-1}, \Delta_k, \sigma_x^2) =
+\begin{bmatrix}
+x_{k|k-1} & \dot x_{k|k-1}
+\end{bmatrix}
+\end{equation*}
+where
+\begin{align*}
+x_{k|k-1} &= x_{k-1|k-1} + \Delta_k \dot x_{k|k-1} \\
+x_{k|k-1} &= \dot x_{k-1|k-1} + w_k \\
+w_k &\sim \mathcal{N}_T(0, \Delta_k \sigma_x^2)
+\end{align*}
+Here, $\mathcal{N}_T$ denotes a truncated normal distribution such that
+the system noise is truncated to ensure the speed remains between 0 and 30~m/s.
+
+While $\bM_0$ may perform well on average,
+and indeed if all we were interested in was estimating $x_k$ it may be adequate;
+however, it is missing two important features of bus behaviour:
+\begin{itemize}
+\item variable speeds along roads, \emph{which is what we are truly interested in}
+\item unknown dwell times at bus stops
+\end{itemize}
+
+The first of these is incorporated into $\bM_1$,
+which allows vehicle speeds to vary over time.
+This model uses the same state as $\bM_0$,
+but a different transition function $f_1$.
+In this case, the state equations depend on a sequence of system noise,
+$\bw_k = \{\bw_k^1, \cdots, \bw_k^{\Delta_k}\}$
+\begin{align*}
+x_{k|k-1} &= x_{k-1|k-1} + \Delta_k \dot x_{k-1|k-1} + \sum_{i=1}^{\Delta_k} w_k^i \\
+\dot x_{k|k-1} &= \dot x_{k-1|k-1} + \sum_{i=1}^{\Delta_k} w_k^i \\
+w_k^i &\sim \mathcal{N}_T(0, \sigma_x^2)
+\end{align*}
+Again, system noise is assumed to have a Gaussian distribution, 
+truncated to ensure that the vehicle's speed is always between 0 and 30~m/s.
+This is achieved by iteratively estimating each second the the vehicle's state,
+each time ensuring the speed remains in the specified range.
 

### 2018-08-30 anzjs2018/sections/03_models.tex
words

@@
 $\bw_k = \{\bw_k^1, \cdots, \bw_k^{\Delta_k}\}$
-\begin{align*}
-x_{k|k-1} &= x_{k-1|k-1} + \Delta_k \dot x_{k-1|k-1} + \sum_{i=1}^{\Delta_k} w_k^i \\
+\begin{align}
+\label{eq:transition_f1}
+x_{k|k-1} &= x_{k-1|k-1} + \Delta_k \dot x_{k-1|k-1} + \sum_{i=1}^{\Delta_k} w_k^i \nonumber \\
 \dot x_{k|k-1} &= \dot x_{k-1|k-1} + \sum_{i=1}^{\Delta_k} w_k^i \\
-w_k^i &\sim \mathcal{N}_T(0, \sigma_x^2)
-\end{align*}
+w_k^i &\sim \mathcal{N}_T(0, \sigma_x^2) \nonumber
+\end{align}
 Again, system noise is assumed to have a Gaussian distribution, 
@@
 
+The next model, $\bM_2$, allows for dwell times at bus stops.
+Since buses can either stop and wait, or not stop,
+multimodality is introduced to the system,
+which requires a modeling framework, such as the particle filter used here,
+to handle it.
+For each bus stop $j = 1, \ldots, M$ along a route,
+its distance into the trip, $S_j^d$, 
+is used to determine when a vehicle passes a stop.
+When this occurs, the dwell time $\delta_j \geq 0$ at stop $j$ is modelled
+using the stopping probability, $\pi_j$,
+a common dwell time parameter $\gamma$ which models the
+deceleration, opening and closing of doors, and acceleration at bus stops,
+and a stop dwell time $\tau_j$ for passengers alighting and disembarking.
+Therefore the stop dwell time is
+\begin{equation}
+\label{eq:dwell_time}
+\delta_j = p_j(\gamma + \tilde\delta_j),\quad
+p_j \sim \mathrm{Bern}(\pi_j),\quad
+\tilde\delta_j \sim \mathcal{E}(\tau_j)
+\end{equation}
+
+To define the transition function for this model, $f_2$,
+we need to account for multiple dwell times.
+Let $s_n$ be the index of the next stop,
+and sample dwell times for each upcoming stop $j = s_n, \ldots, M$,
+noting that these can be zero.
+Assuming the vehicle passes stops $s_n$ up to $J-1$,
+does it reach stop $J$ in the time remaining?
+\begin{equation}
+\tilde X_k^J = \min\left\{
+S_J^d, f_1(\hat X_{k|k-1}, \Delta_k - \sum_{j=s_n}^{J-1}\delta_j)\right\}
+\end{equation}
+Taking the maximum of the sequence gives the predicted distance
+traveled by the vehicle, so
+\begin{equation}
+\hat X_{k|k-1} = f_2 (\hat X_{k-1|k-1}, \Delta_k) =
+\max_{J = s_n, \ldots, M}\left\{
+    \tilde X_k^J
+\right\}
+\end{equation}
+
+Care needs to be made when generating the sequence,
+so in practice the state will be calculated recursively to ensure
+the system noise is consistent.
+\begin{equation*}
+\hat{\tilde X}_k^J =
+\begin{cases}
+    f_1(\hat{\tilde X}_k^{J+1}, \delta_J) & J < M \\
+    \hat X_{k-1|k-1} & J = M
+\end{cases}
+\end{equation*}
+
 \subsection{Network model}

### 2018-09-08 anzjs2018/sections/03_models.tex
model writing

@@
 
+The final model $\bM_3$ demonstrated here will allow acceleration to vary.
+\begin{align*}
+\label{eq:model_m3}
+x_{k|k-1} &= x(t_k) \\
+\intertext{where states are recursively calculated back to $t_{k-1}$}
+x(t) &= x(t - 1) + \dot x(t) \\
+\dot x(t) &= \dot x(t-1) + \ddot x(t) \\
+\ddot x(t) &= \ddot x(t-1) + w_t,\quad
+w_t \sim \mathcal{N}(0,\sigma_x^2)
+\end{align*}
+By truncating the acceleration, 
+the speed of the vehicle can be maintained between 0 and 30~m/s.
+Additionally, when approaching a bus stop that will be stopped at,
+the acceleration will be trunctated to below zero,
+however no physical deceleration will be modeled as this is too complex
+and stopping behaviour of buses can be sudden 
+(for example when passengers signal late that they wish to board or disembark).
+After servicing a stop, the bus will need to accelerate up to speed again.
+
+\subsubsection{Updating state using the observation likelihood}
+\label{sec:pf_update}
+
+The update step involves comparing the predicted state $\bX_{k|k-1}$
+with the observed state $\bY_k$,
+using the measurement function $h$ to convert between the two state spaces,
+and a likelihood function.
+The measurement function is a deterministic function that uses only the 
+distance traveled, $x_k$, along a known path $\bR$ to obtain a GPS coordinate.
+The inverse measurement function, which is used in other approaches,
+is \emph{not deterministic}, and instead uses an optimisation to
+find the ``estimated'' distance traveled,
+which invariably leads to issues in the implementation.
+By using a particle filter approach, we avoid using $h^{-1}$
+and instead determine the liklihood using $\bY_k$ directly.
+
+For the likelihood, the GPS error is assumed symmetrical about the true
+position of the vehicle, with a radius of $\sigma_y^2$ meters.
+Define a projection function $g(\cdot | \bY)$ that projects points
+onto a cartesian surface (``Flat Earth''),
+so that the distance between points is the real Earth distance.
+Then the distance between the points is
+\begin{align*}
+\delta_k &= \sqrt{\left(g(h(X_k))_1 - g(\bY_k)_1\right)^2 +
+    \left(g(h(X_k))_2 - g(\bY_k)_2\right)^2} \\
+\intertext{which can be standardised by the GPS error}
+\tilde \delta_k &=
+\sqrt{\left(\frac{g(h(X_k))_1 - g(\bY_k)_1}{\sigma_y}\right)^2 +
+    \left(\frac{g(h(X_k))_2 - g(\bY_k)_2}{\sigma_y}\right)^2} =\frac{\delta_k}{\sigma_y} \\
+\intertext{Since we assume symmetric variance, this is actually}
+\tilde\delta_k &= \sqrt{Z_1^2 + Z_2^2}, \quad
+Z_1,Z_2 \sim \mathcal{N}(0,1)
+\end{align*}
+The sum of two normal random variables has an Exponential distribution with rate $\frac{1}{2}$,
+therefore 
+\begin{equation}
+\label{eq:lhood_exp}
+\left(\frac{\delta_k}{\sigma_y}\right)^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
+\end{equation}
+
+Combining this together, we use a distance function $d(\bY_1, \bY_2)$ that calculates
+the geographic distance between two GPS coordinates,
+the likelihood is
+\begin{equation}
+\label{eq:lhood}
+f(\bY_k | \hat\bX_{k|k-1}, \sigma_y^2) =
+\frac{1}{2} \exp\left\{-\frac{d(\bY_k, h(\hat\bX_{k|k-1}))^2}{2\sigma^2}\right\}
+\end{equation}
+
 \subsection{Network model}

### 2018-10-01 anzjs2018/sections/03_models.tex
network model

@@
 
-Two parts to this - first, the prior prediction step; then, the data update step.
+The second part of this appraoch involves a network model of transit network travel times along links in the network.
+Reliable ETAs will require a historical data component,
+used as a prior in the absense of data as well as for
+predicting future states, 
+and a realtime update component where the travel times of vehicles
+are used to update the current estimate of network state.
+
+The model will be another recursive Bayesian model as used for the vehicle state,
+as this provides a simple way to model realtime data.
+Since travel times can safely be assumed to be unimodal,
+we use an information filter (IF),
+which is a transformation of the Kalman filter
+using information instead of covariance,
+and allows the simple combination of multiple observations,
+for example when multiple buses traverse a road in the same update.
+The prediction step will involve a transition function that incorporates historical data
+and the update step will weight between observed and predicted travel times.
+
+The network state $\Theta_c = \{\theta_c^j\}_{j = 1}^J$ is the travel time 
+of transit vehicles along road segment $j$ at time $t_c$.
+In this implementation, we are assuming independence between each segment
+to provide a simple basis for the model; of course in reality this
+is not true and future work will investigate ways of incorporating this.
+
+Using an IF requires a transition matrix, $\mathbf{F}_c$;
+however, in this case the travel time model is constant so no
+transition matrix is required (i.e., $\mathbf{F}_c = 1$).
+Let $\Delta_c = t_c - t_{c-1}$ and $P_c^j$ the state uncertainty
+with system noise $Q_c = \Delta_c \sigma_b^2$, then the update equations are
+simply the normal KF update equations,
+\begin{align}
+\label{eq:kf_transition}
+\hat\theta^j_{c|c-1} &= \hat\theta^j_{c|c-1} \\
+P^j_{c|c-1} &= P^j_{c-1|c-1} + Q_c
+\end{align}
+
+For the update step, however, we need to transform into an information
+space parameterised by the information matrix 
+$\mathbf{Z}^j_{c|c-1} = P_{c|c-1}^{-1}$ 
+and the information vector $\mathbf{\hat z}^j_{c|c-1} = \hat\theta_{c|c-1} P_{c|c-1}^{-1}$.
+The measurement data obtained from the vehicle model are
+the travel time of vehicle $m$ along segment $j$,
+$\bar b^m_j$, and the uncertainty $s^m_j$.
+These can be transformed to a measurement information covaraince matrix
+and vector using the measurement matrix $\mathbf{H} = 1$ since,
+in this case, the observed state is the underlying state we are estimating (travel time).
+\begin{equation}
+I^m_{jc} = \frac{1}{(s^m_j)^{2}}\quad\text{and}\quad
+i^m_{jc} = \frac{\bar b^m_j}{(s^m_j)^2}
+\end{equation}
+The information update is now the sum of the information for all vehicles
+that traversed the segment since the last update, so
+\begin{align*}
+\mathbf{Z}^j_{c|c} &= \mathbf{Z}^j_{c|c-1} + \sum_{m=1}^M I^m_{jc} \\
+\mathbf{\hat i}^j_{c|c} &= \mathbf{\hat z}^j_{c|c-1} + \sum_{m=1}^M i^m_{jc}
+\end{align*}
+which are easily transformed back to get the desired state parameters.
+\begin{equation}
+\hat\theta^j_{c|c} = \frac{\mathbf{\hat z}^j_{c|c}}{\mathbf{Z}^j_{c|c}} 
+\quad\text{and}\quad
+P^j_{c|c} = \frac{1}{\mathbf{Z}^j_{c|c}}
+\end{equation}

### 2018-10-03 anzjs2018/sections/03_models.tex
Writing model information

--- 2018-08-30 anzjs2018/sections/03_models.tex
@@
 \label{sec:models}
+
+The real-time nature of this application (means that for) both
+the vehicle and network models a recursive Bayesian filtering approach is logical,
+and indeed real-time tracking and robotics applications often use such (models).
+In each of the following models, 
+we assume an underlying Markov process with state $\mathcal{X}_k$,
+of which observations $\mathcal{Y}_k$ are made,
+which leads us to the following general model,
+\begin{equation}
+\label{eq:rbe_model}
+\begin{split}
+\mathcal{X}_k &= f(\mathcal{X}_{k-1}, \omega_k) \\
+\mathcal{Y}_k &= h(\mathcal{X}_k, \nu_k)
+\end{split}
+\end{equation}
+where $\omega_k\sim\mathrm{N}(0,\mathcal{Q})$ is the system noise,
+and $\nu_k\sim\mathrm{N}(0,\mathcal{R})$ is measurement error.
+The transition function $f$ determines the relationship between consecutive states,
+while the measurement function $h$ is a deterministic function describing the
+relationship between the underlying and observed states.
+
+The following sections describe two Bayesian filter models based on (\ref{eq:rbe_model}) used in this application.
+The first model is implemented using a particle filter to estimate vehicle states,
+with the primary objective of estimating vehicle travel times along roads,
+while the second using these estimated times to update road states,
+and is implemented using a Kalman filter.
 
@@
 
-The vehicle model involves infering the \emph{unobservable state} $\bX_k$ 
-of a vehicle at time $t_k$ from the \emph{observed state} $\by_k$,
-which in this case is the GPS location of the vehicle,
-where $\lambda_k$ and $\phi_k$ are the longitude and latitude, respectively.
+The underlying vehicle state at time $t_k$ consists of
+the vehicle's distance traveled $x$ in meters along the route and
+its speed $\dot x$ in meters per second.
+Additionally, we include travel times along road segments along the route, 
+$\bz = (z_1,\ldots,z_L)^\top$, in seconds,
+which are independent of time but included in the state as they are estimated
+sequentially as the vehicle traverses the route.
+These are combined into the state vector,
 \begin{equation}
-\label{eq:vehicle_observation}
-\by_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}
+\label{eq:vehicle_state}
+\bX_k = 
+\begin{bmatrix}
+    x_k \\ \dot x_k \\ \bz
+\end{bmatrix}.
 \end{equation}
-To construct the unobservable state, 
-each vehicle is assumed to follow the trajectory defined by a trajectory function of time, $x(t)$,
-such that the distance traveled along a route by a vehicle at time $t_k$ is given by
-$x_k = x(t_k)$.
-Additionally, speed and acceleration are given by
-the first and second derivatives of the trajectory function,
-$\dot x_k = x'(t_k)$ and $\ddot x_k = x''(t_k)$, respectively.
-Exactly how the state is constructed will depend on the model being used.
-
-Since observations of the vehicle are irregular, 
-a \emph{transition function} $f$ is used to
-predict the vehicle's state at time $t_k$ given the current knowledge of the state at time $t_{k-1}$,
-as well as accounting for \emph{system noise}, $\sigma_x^2$.
-We also define a \emph{measurement function} $h$ which describes the relationship between
-the observable and unobservable states,
-and accounting for \emph{measurement error}, $\bv_k$,
+Observations of the vehicle are made at time $t_k$ using a GPS,
+giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle,
+giving us the observation vector
 \begin{equation}
-\label{eq:vehicle_measurement}
-\by_k = h(\bX_k) + \bv_k = h(x_k) + \bv_k,
-\end{equation}
-so all models need to estimate at least $x_k$, the distance traveled along the route.
-
-The model is split into two steps, prediction and update.
-The prediction involves using the transition function to predict
-the next state, $\hat\bX_{k|k-1}$,
-while the update step using a likelihood function and the measurement function
-to update the state to account for the observed vehicle location, $\by_k$,
-giving a final estimate of $\hat\bX_{k|k}$.
-
-\subsubsection{Predicting state using the transtion function}
-\label{sec:pf_prediction}
-
-The transition function is a model of bus behaviour that allows
-predictions of future states to be made given the most recently 
-updated state.
-Developing a good model for bus behaviour allows us to estimate
-parameters of interest, such as vehicle speed,
-dwell time at bus stops, and travel time along individual roads,
-the last of which is of course our primary objective.
-In this section, we propose [HOW MANY?] increasingly complex models
-of bus behaviour.
-
-%% MODEL 1
-The first model, $\bM_0$, assumes a constant speed between observations
-using the state
-$\bX_k = \left[\begin{smallmatrix}x_k && \dot x_k\end{smallmatrix}\right]^\top$
-and Gaussian system noise with mean zero and variance adjusted by
-$\Delta_k = t_k - t_{k-1}$.
-\begin{equation*}
-\hat\bX_{k|k-1} = f_0(\bX_{k-1|k-1}, \Delta_k, \sigma_x^2) =
-\begin{bmatrix}
-x_{k|k-1} & \dot x_{k|k-1}
-\end{bmatrix}
-\end{equation*}
-where
-\begin{align*}
-x_{k|k-1} &= x_{k-1|k-1} + \Delta_k \dot x_{k|k-1} \\
-x_{k|k-1} &= \dot x_{k-1|k-1} + w_k \\
-w_k &\sim \mathcal{N}_T(0, \Delta_k \sigma_x^2)
-\end{align*}
-Here, $\mathcal{N}_T$ denotes a truncated normal distribution such that
-the system noise is truncated to ensure the speed remains between 0 and 30~m/s.
-
-While $\bM_0$ may perform well on average,
-and indeed if all we were interested in was estimating $x_k$ it may be adequate;
-however, it is missing two important features of bus behaviour:
-\begin{itemize}
-\item variable speeds along roads, \emph{which is what we are truly interested in}
-\item unknown dwell times at bus stops
-\end{itemize}
-
-The first of these is incorporated into $\bM_1$,
-which allows vehicle speeds to vary over time.
-This model uses the same state as $\bM_0$,
-but a different transition function $f_1$.
-In this case, the state equations depend on a sequence of system noise,
-$\bw_k = \{\bw_k^1, \cdots, \bw_k^{\Delta_k}\}$
-\begin{align}
-\label{eq:transition_f1}
-x_{k|k-1} &= x_{k-1|k-1} + \Delta_k \dot x_{k-1|k-1} + \sum_{i=1}^{\Delta_k} w_k^i \nonumber \\
-\dot x_{k|k-1} &= \dot x_{k-1|k-1} + \sum_{i=1}^{\Delta_k} w_k^i \\
-w_k^i &\sim \mathcal{N}_T(0, \sigma_x^2) \nonumber
-\end{align}
-Again, system noise is assumed to have a Gaussian distribution, 
-truncated to ensure that the vehicle's speed is always between 0 and 30~m/s.
-This is achieved by iteratively estimating each second the the vehicle's state,
-each time ensuring the speed remains in the specified range.
-
-The next model, $\bM_2$, allows for dwell times at bus stops.
-Since buses can either stop and wait, or not stop,
-multimodality is introduced to the system,
-which requires a modeling framework, such as the particle filter used here,
-to handle it.
-For each bus stop $j = 1, \ldots, M$ along a route,
-its distance into the trip, $S_j^d$, 
-is used to determine when a vehicle passes a stop.
-When this occurs, the dwell time $\delta_j \geq 0$ at stop $j$ is modelled
-using the stopping probability, $\pi_j$,
-a common dwell time parameter $\gamma$ which models the
-deceleration, opening and closing of doors, and acceleration at bus stops,
-and a stop dwell time $\tau_j$ for passengers alighting and disembarking.
-Therefore the stop dwell time is
-\begin{equation}
-\label{eq:dwell_time}
-\delta_j = p_j(\gamma + \tilde\delta_j),\quad
-p_j \sim \mathrm{Bern}(\pi_j),\quad
-\tilde\delta_j \sim \mathcal{E}(\tau_j)
+\label{eq:vehicle_obs}
+\bY_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}.
 \end{equation}
 
-To define the transition function for this model, $f_2$,
-we need to account for multiple dwell times.
-Let $s_n$ be the index of the next stop,
-and sample dwell times for each upcoming stop $j = s_n, \ldots, M$,
-noting that these can be zero.
-Assuming the vehicle passes stops $s_n$ up to $J-1$,
-does it reach stop $J$ in the time remaining?
+We decided to employ a particle filter to estimate $\bX_k$,
+since it is a highly flexible approach and has been used in recent 
+transit vehicle modeling applications \citep{Hans_2015}.
+Our main justification for using it is that it handles multimodality very well,
+which is a common feature of the proposal distribution particularly around bus stops.
+Another advantage is the intuitive likelihood function, 
+which is described later.
+Conversely, the particle filter is a computationally demanding method,
+as thousands of particles are required per bus.
+Section~\ref{sec:rt} describes how we were able to implement the particle filter in real-time.
+
+In the particle filter, the posterior distribution of the state at time $t_k$,
+is estimated by a set of discrete points, or particles,
 \begin{equation}
-\tilde X_k^J = \min\left\{
-S_J^d, f_1(\hat X_{k|k-1}, \Delta_k - \sum_{j=s_n}^{J-1}\delta_j)\right\}
+p(\bX_k | \bY_k) \approx \tilde\bX_{k|k} := (\bX_k^{(i)})_{i=1}^N
 \end{equation}
-Taking the maximum of the sequence gives the predicted distance
-traveled by the vehicle, so
+each of which is independently updated or \emph{mutated} using the transition function $f$,
 \begin{equation}
-\hat X_{k|k-1} = f_2 (\hat X_{k-1|k-1}, \Delta_k) =
-\max_{J = s_n, \ldots, M}\left\{
-    \tilde X_k^J
-\right\}
+p(\bX_k | \bX_{k-1}) \approx \tilde\bX_{k|k-1} := 
+\left(f(\bX_{k-1}^{(i)}, \psi)\right)_{i=1}^N
 \end{equation}
+using a parameter vector $\psi$ containing all of the necessary parameters
+for the model (including system noise).
+After mutation, \emph{selection} of a new set of particles is performed by
+importance resampling using likelihood weights.
 
-Care needs to be made when generating the sequence,
-so in practice the state will be calculated recursively to ensure
-the system noise is consistent.
-\begin{equation*}
-\hat{\tilde X}_k^J =
-\begin{cases}
-    f_1(\hat{\tilde X}_k^{J+1}, \delta_J) & J < M \\
-    \hat X_{k-1|k-1} & J = M
-\end{cases}
-\end{equation*}
+\subsubsection{Vehicle transition function}
+\label{sec:pf_prediction}
+As we are using a particle filter to implement the vehicle model,
+the transition function $f$ can be a complex description of bus behaviour,
+most notably
+\begin{itemize}
+\item non-constant speed along roads (acceleration as system noise),
+\item stopping and waiting at bus stops while passengers board and disembark, and
+\item stopping and waiting at intersections.
+\end{itemize}
+The details of this are given in the algorithm defined in the Appendix.
+
+For each particle, the transition function generates a plausible trajectory,
+using system noise parameter $\sigma^2$ which describes 
+how the vehicles acceleration changes as a random process,
+and propogated using Newton's laws of motion \cite{}
+\begin{equation}
+
+\end{equation}
+This includes stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
+dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
+board and disembark (conditional on the bus stopping),
+and the minimum dwell time at stops, $\gamma$.
+For these parameters, we used constant values for all stops,
+and based the values on those used by \cite{Hans_2015};
+future work will look at modeling these separately in real-time also.
+
+Also in this application, due to the complexity of the problem,
+we use bus stops to define road segments.
+Therefore no intersection model is required,
+only the segment travel times as predicted by the network model at time $t_k$,
+$\btheta(t_k) = (\theta_1(t_k), \ldots, \theta_L(t_k))^\top$
+(see section~\ref{sec:kf}).
+Future work will adapt this to use intersection that are indpendent 
+of bus stops, and therefore completely independent of routes,
+which will allow increasing the complexity of the model to incorporate
+intersection stopping probabilities and waiting times.
+
+% The vehicle model involves infering the \emph{unobservable state} $\bX_k$ 
+% of a vehicle at time $t_k$ from the \emph{observed state} $\by_k$,
+% which in this case is the GPS location of the vehicle,
+% where $\lambda_k$ and $\phi_k$ are the longitude and latitude, respectively.
+% \begin{equation}
+% \label{eq:vehicle_observation}
+% \by_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}
+% \end{equation}
+% To construct the unobservable state, 
+% each vehicle is assumed to follow the trajectory defined by a trajectory function of time, $x(t)$,
+% such that the distance traveled along a route by a vehicle at time $t_k$ is given by
+% $x_k = x(t_k)$.
+% Additionally, speed and acceleration are given by
+% the first and second derivatives of the trajectory function,
+% $\dot x_k = x'(t_k)$ and $\ddot x_k = x''(t_k)$, respectively.
+% Exactly how the state is constructed will depend on the model being used.
+
+% Since observations of the vehicle are irregular, 
+% a \emph{transition function} $f$ is used to
+% predict the vehicle's state at time $t_k$ given the current knowledge of the state at time $t_{k-1}$,
+% as well as accounting for \emph{system noise}, $\sigma_x^2$.
+% We also define a \emph{measurement function} $h$ which describes the relationship between
+% the observable and unobservable states,
+% and accounting for \emph{measurement error}, $\bv_k$,
+% \begin{equation}
+% \label{eq:vehicle_measurement}
+% \by_k = h(\bX_k) + \bv_k = h(x_k) + \bv_k,
+% \end{equation}
+% so all models need to estimate at least $x_k$, the distance traveled along the route.
+
+% The model is split into two steps, prediction and update.
+% The prediction involves using the transition function to predict
+% the next state, $\hat\bX_{k|k-1}$,
+% while the update step using a likelihood function and the measurement function
+% to update the state to account for the observed vehicle location, $\by_k$,
+% giving a final estimate of $\hat\bX_{k|k}$.
+
+% \subsubsection{Predicting state using the transtion function}
+% \label{sec:pf_prediction}
+
+% The transition function is a model of bus behaviour that allows
+% predictions of future states to be made given the most recently 
+% updated state.
+% Developing a good model for bus behaviour allows us to estimate
+% parameters of interest, such as vehicle speed,
+% dwell time at bus stops, and travel time along individual roads,
+% the last of which is of course our primary objective.
+% In this section, we propose [HOW MANY?] increasingly complex models
+% of bus behaviour.
+
+% %% MODEL 1
+% The first model, $\bM_0$, assumes a constant speed between observations
+% using the state
+% $\bX_k = \left[\begin{smallmatrix}x_k && \dot x_k\end{smallmatrix}\right]^\top$
+% and Gaussian system noise with mean zero and variance adjusted by
+% $\Delta_k = t_k - t_{k-1}$.
+% \begin{equation*}
+% \hat\bX_{k|k-1} = f_0(\bX_{k-1|k-1}, \Delta_k, \sigma_x^2) =
+% \begin{bmatrix}
+% x_{k|k-1} & \dot x_{k|k-1}
+% \end{bmatrix}
+% \end{equation*}
+% where
+% \begin{align*}
+% x_{k|k-1} &= x_{k-1|k-1} + \Delta_k \dot x_{k|k-1} \\
+% x_{k|k-1} &= \dot x_{k-1|k-1} + w_k \\
+% w_k &\sim \mathcal{N}_T(0, \Delta_k \sigma_x^2)
+% \end{align*}
+% Here, $\mathcal{N}_T$ denotes a truncated normal distribution such that
+% the system noise is truncated to ensure the speed remains between 0 and 30~m/s.
+
+% While $\bM_0$ may perform well on average,
+% and indeed if all we were interested in was estimating $x_k$ it may be adequate;
+% however, it is missing two important features of bus behaviour:
+% \begin{itemize}
+% \item variable speeds along roads, \emph{which is what we are truly interested in}
+% \item unknown dwell times at bus stops
+% \end{itemize}
+
+% The first of these is incorporated into $\bM_1$,
+% which allows vehicle speeds to vary over time.
+% This model uses the same state as $\bM_0$,
+% but a different transition function $f_1$.
+% In this case, the state equations depend on a sequence of system noise,
+% $\bw_k = \{\bw_k^1, \cdots, \bw_k^{\Delta_k}\}$
+% \begin{align}
+% \label{eq:transition_f1}
+% x_{k|k-1} &= x_{k-1|k-1} + \Delta_k \dot x_{k-1|k-1} + \sum_{i=1}^{\Delta_k} w_k^i \nonumber \\
+% \dot x_{k|k-1} &= \dot x_{k-1|k-1} + \sum_{i=1}^{\Delta_k} w_k^i \\
+% w_k^i &\sim \mathcal{N}_T(0, \sigma_x^2) \nonumber
+% \end{align}
+% Again, system noise is assumed to have a Gaussian distribution, 
+% truncated to ensure that the vehicle's speed is always between 0 and 30~m/s.
+% This is achieved by iteratively estimating each second the the vehicle's state,
+% each time ensuring the speed remains in the specified range.
+
+% The next model, $\bM_2$, allows for dwell times at bus stops.
+% Since buses can either stop and wait, or not stop,
+% multimodality is introduced to the system,
+% which requires a modeling framework, such as the particle filter used here,
+% to handle it.
+% For each bus stop $j = 1, \ldots, M$ along a route,
+% its distance into the trip, $S_j^d$, 
+% is used to determine when a vehicle passes a stop.
+% When this occurs, the dwell time $\delta_j \geq 0$ at stop $j$ is modelled
+% using the stopping probability, $\pi_j$,
+% a common dwell time parameter $\gamma$ which models the
+% deceleration, opening and closing of doors, and acceleration at bus stops,
+% and a stop dwell time $\tau_j$ for passengers alighting and disembarking.
+% Therefore the stop dwell time is
+% \begin{equation}
+% \label{eq:dwell_time}
+% \delta_j = p_j(\gamma + \tilde\delta_j),\quad
+% p_j \sim \mathrm{Bern}(\pi_j),\quad
+% \tilde\delta_j \sim \mathcal{E}(\tau_j)
+% \end{equation}
+
+% To define the transition function for this model, $f_2$,
+% we need to account for multiple dwell times.
+% Let $s_n$ be the index of the next stop,
+% and sample dwell times for each upcoming stop $j = s_n, \ldots, M$,
+% noting that these can be zero.
+% Assuming the vehicle passes stops $s_n$ up to $J-1$,
+% does it reach stop $J$ in the time remaining?
+% \begin{equation}
+% \tilde X_k^J = \min\left\{
+% S_J^d, f_1(\hat X_{k|k-1}, \Delta_k - \sum_{j=s_n}^{J-1}\delta_j)\right\}
+% \end{equation}
+% Taking the maximum of the sequence gives the predicted distance
+% traveled by the vehicle, so
+% \begin{equation}
+% \hat X_{k|k-1} = f_2 (\hat X_{k-1|k-1}, \Delta_k) =
+% \max_{J = s_n, \ldots, M}\left\{
+%     \tilde X_k^J
+% \right\}
+% \end{equation}
+
+% Care needs to be made when generating the sequence,
+% so in practice the state will be calculated recursively to ensure
+% the system noise is consistent.
+% \begin{equation*}
+% \hat{\tilde X}_k^J =
+% \begin{cases}
+%     f_1(\hat{\tilde X}_k^{J+1}, \delta_J) & J < M \\
+%     \hat X_{k-1|k-1} & J = M
+% \end{cases}
+% \end{equation*}
+
+% The final model $\bM_3$ demonstrated here will allow acceleration to vary.
+% \begin{align*}
+% \label{eq:model_m3}
+% x_{k|k-1} &= x(t_k) \\
+% \intertext{where states are recursively calculated back to $t_{k-1}$}
+% x(t) &= x(t - 1) + \dot x(t) \\
+% \dot x(t) &= \dot x(t-1) + \ddot x(t) \\
+% \ddot x(t) &= \ddot x(t-1) + w_t,\quad
+% w_t \sim \mathcal{N}(0,\sigma_x^2)
+% \end{align*}
+% By truncating the acceleration, 
+% the speed of the vehicle can be maintained between 0 and 30~m/s.
+% Additionally, when approaching a bus stop that will be stopped at,
+% the acceleration will be trunctated to below zero,
+% however no physical deceleration will be modeled as this is too complex
+% and stopping behaviour of buses can be sudden 
+% (for example when passengers signal late that they wish to board or disembark).
+% After servicing a stop, the bus will need to accelerate up to speed again.
+
+% \subsubsection{Updating state using the observation likelihood}
+% \label{sec:pf_update}
+
+% The update step involves comparing the predicted state $\bX_{k|k-1}$
+% with the observed state $\bY_k$,
+% using the measurement function $h$ to convert between the two state spaces,
+% and a likelihood function.
+% The measurement function is a deterministic function that uses only the 
+% distance traveled, $x_k$, along a known path $\bR$ to obtain a GPS coordinate.
+% The inverse measurement function, which is used in other approaches,
+% is \emph{not deterministic}, and instead uses an optimisation to
+% find the ``estimated'' distance traveled,
+% which invariably leads to issues in the implementation.
+% By using a particle filter approach, we avoid using $h^{-1}$
+% and instead determine the liklihood using $\bY_k$ directly.
+
+% For the likelihood, the GPS error is assumed symmetrical about the true
+% position of the vehicle, with a radius of $\sigma_y^2$ meters.
+% Define a projection function $g(\cdot | \bY)$ that projects points
+% onto a cartesian surface (``Flat Earth''),
+% so that the distance between points is the real Earth distance.
+% Then the distance between the points is
+% \begin{align*}
+% \delta_k &= \sqrt{\left(g(h(X_k))_1 - g(\bY_k)_1\right)^2 +
+%     \left(g(h(X_k))_2 - g(\bY_k)_2\right)^2} \\
+% \intertext{which can be standardised by the GPS error}
+% \tilde \delta_k &=
+% \sqrt{\left(\frac{g(h(X_k))_1 - g(\bY_k)_1}{\sigma_y}\right)^2 +
+%     \left(\frac{g(h(X_k))_2 - g(\bY_k)_2}{\sigma_y}\right)^2} =\frac{\delta_k}{\sigma_y} \\
+% \intertext{Since we assume symmetric variance, this is actually}
+% \tilde\delta_k &= \sqrt{Z_1^2 + Z_2^2}, \quad
+% Z_1,Z_2 \sim \mathcal{N}(0,1)
+% \end{align*}
+% The sum of two normal random variables has an Exponential distribution with rate $\frac{1}{2}$,
+% therefore 
+% \begin{equation}
+% \label{eq:lhood_exp}
+% \left(\frac{\delta_k}{\sigma_y}\right)^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
+% \end{equation}
+
+% Combining this together, we use a distance function $d(\bY_1, \bY_2)$ that calculates
+% the geographic distance between two GPS coordinates,
+% the likelihood is
+% \begin{equation}
+% \label{eq:lhood}
+% f(\bY_k | \hat\bX_{k|k-1}, \sigma_y^2) =
+% \frac{1}{2} \exp\left\{-\frac{d(\bY_k, h(\hat\bX_{k|k-1}))^2}{2\sigma^2}\right\}
+% \end{equation}
 
@@
 
-Two parts to this - first, the prior prediction step; then, the data update step.
+% The second part of this appraoch involves a network model of transit network travel times along links in the network.
+% Reliable ETAs will require a historical data component,
+% used as a prior in the absense of data as well as for
+% predicting future states, 
+% and a realtime update component where the travel times of vehicles
+% are used to update the current estimate of network state.
+
+% The model will be another recursive Bayesian model as used for the vehicle state,
+% as this provides a simple way to model realtime data.
+% Since travel times can safely be assumed to be unimodal,
+% we use an information filter (IF),
+% which is a transformation of the Kalman filter
+% using information instead of covariance,
+% and allows the simple combination of multiple observations,
+% for example when multiple buses traverse a road in the same update.
+% The prediction step will involve a transition function that incorporates historical data
+% and the update step will weight between observed and predicted travel times.
+
+% The network state $\Theta_c = \{\theta_c^j\}_{j = 1}^J$ is the travel time 
+% of transit vehicles along road segment $j$ at time $t_c$.
+% In this implementation, we are assuming independence between each segment
+% to provide a simple basis for the model; of course in reality this
+% is not true and future work will investigate ways of incorporating this.
+
+% Using an IF requires a transition matrix, $\mathbf{F}_c$;
+% however, in this case the travel time model is constant so no
+% transition matrix is required (i.e., $\mathbf{F}_c = 1$).
+% Let $\Delta_c = t_c - t_{c-1}$ and $P_c^j$ the state uncertainty
+% with system noise $Q_c = \Delta_c \sigma_b^2$, then the update equations are
+% simply the normal KF update equations,
+% \begin{align}
+% \label{eq:kf_transition}
+% \hat\theta^j_{c|c-1} &= \hat\theta^j_{c|c-1} \\
+% P^j_{c|c-1} &= P^j_{c-1|c-1} + Q_c
+% \end{align}
+
+% For the update step, however, we need to transform into an information
+% space parameterised by the information matrix 
+% $\mathbf{Z}^j_{c|c-1} = P_{c|c-1}^{-1}$ 
+% and the information vector $\mathbf{\hat z}^j_{c|c-1} = \hat\theta_{c|c-1} P_{c|c-1}^{-1}$.
+% The measurement data obtained from the vehicle model are
+% the travel time of vehicle $m$ along segment $j$,
+% $\bar b^m_j$, and the uncertainty $s^m_j$.
+% These can be transformed to a measurement information covaraince matrix
+% and vector using the measurement matrix $\mathbf{H} = 1$ since,
+% in this case, the observed state is the underlying state we are estimating (travel time).
+% \begin{equation}
+% I^m_{jc} = \frac{1}{(s^m_j)^{2}}\quad\text{and}\quad
+% i^m_{jc} = \frac{\bar b^m_j}{(s^m_j)^2}
+% \end{equation}
+% The information update is now the sum of the information for all vehicles
+% that traversed the segment since the last update, so
+% \begin{align*}
+% \mathbf{Z}^j_{c|c} &= \mathbf{Z}^j_{c|c-1} + \sum_{m=1}^M I^m_{jc} \\
+% \mathbf{\hat i}^j_{c|c} &= \mathbf{\hat z}^j_{c|c-1} + \sum_{m=1}^M i^m_{jc}
+% \end{align*}
+% which are easily transformed back to get the desired state parameters.
+% \begin{equation}
+% \hat\theta^j_{c|c} = \frac{\mathbf{\hat z}^j_{c|c}}{\mathbf{Z}^j_{c|c}} 
+% \quad\text{and}\quad
+% P^j_{c|c} = \frac{1}{\mathbf{Z}^j_{c|c}}
+% \end{equation}

### 2018-10-04 anzjs2018/sections/03_models.tex
models

@@
 \end{itemize}
-The details of this are given in the algorithm defined in the Appendix.
 
@@
 using system noise parameter $\sigma^2$ which describes 
-how the vehicles acceleration changes as a random process,
-and propogated using Newton's laws of motion \cite{}
-\begin{equation}
-
-\end{equation}
-This includes stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
+how the vehicles acceleration changes as a random process.
+% In its simplest form, the transition function becomes,
+% from Newton's laws of motions,
+% \begin{equation}
+% \begin{split}
+%     x_k &= x_{k-1} + \delta_k \dot x_{k-1} \\
+%     \dot x_k &= \dot x_{k-1} + w_k,
+%     \quad w_k \sim \mathrm{N}(0, \delta_k \sigma^2)
+% \end{split}
+% \end{equation}
+However, the vehicle does not simply travel constantly along the route,
+as it needs to service bus stops along the way.
+Therefore, the transition function includes 
+stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
 dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
@@
 future work will look at modeling these separately in real-time also.
-
 Also in this application, due to the complexity of the problem,
@@
 intersection stopping probabilities and waiting times.
+The details of the transition function are given in the algorithm defined in the Appendix,
+which implements those features discussed above.
+
+\subsubsection{Updating state using the observation likelihood}
+\label{sec:pf_update}
+
+After mutating the particle set, the posterior distribution is obtained by
+\emph{selection}, which is implemented by importance resampling using
+likelihood weights.
+To do so, we need to define the measurement function $h$,
+and an additional function $g$ which transforms GPS coordinates onto a flat
+surface (the geographical equirectangular projection,
+which allows us to place a bivate normal likelihood on the data.
+The model for the observation generation is,
+assuming GPS error $\epsilon^2$,
+\begin{equation}
+\label{eq:pf_obs_model}
+g(\bY_k) = g(h(\bX_k)) + \br_k,
+\quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
+\end{equation}
 

### 2018-10-05 anzjs2018/sections/03_models.tex
Likeilihood update

@@
 \end{equation}
+However, due to the assumption that error in longitude and latitude are uncorrelated, 
+the geographical distance between the observation and the true vehicle can be expressed
+in terms of two independent standard normal random variables $z_1$ and $z_2$
+\begin{equation}
+\label{eq:obs_dist}
+dist(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
+    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}
+\end{equation}
+As the sum of two independent, standard normal random variables 
+is $\chi^2$ distributed with 2~degrees of freedom,
+which is itself exponential,
+rearranging (\ref{eq:obs_dist}) yields
+\begin{equation}
+\label{eq:obs_exp}
+\left(\frac{dist(\bY_k, h(\bX_k))}{\epsilon}\right)^2 =
+z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
+\end{equation}
 
-% The vehicle model involves infering the \emph{unobservable state} $\bX_k$ 
-% of a vehicle at time $t_k$ from the \emph{observed state} $\by_k$,
-% which in this case is the GPS location of the vehicle,
-% where $\lambda_k$ and $\phi_k$ are the longitude and latitude, respectively.
-% \begin{equation}
-% \label{eq:vehicle_observation}
-% \by_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}
-% \end{equation}
-% To construct the unobservable state, 
-% each vehicle is assumed to follow the trajectory defined by a trajectory function of time, $x(t)$,
-% such that the distance traveled along a route by a vehicle at time $t_k$ is given by
-% $x_k = x(t_k)$.
-% Additionally, speed and acceleration are given by
-% the first and second derivatives of the trajectory function,
-% $\dot x_k = x'(t_k)$ and $\ddot x_k = x''(t_k)$, respectively.
-% Exactly how the state is constructed will depend on the model being used.
+The likelihood of the data given a particle's state estimate 
+is therefore easy to calculate using (\ref{eq:obs_exp})
+\begin{equation}
+p(\bY_k | \bX_k^{(i)}, \epsilon) =
+\frac{1}{2}\exp\left\{
+-\frac{1}{2} \left(\frac{dist(\bY_k, h(\bX_k^{(i)}))}{\epsilon}\right)^2
+\right\}
+\end{equation}
+It is worth noting that this representation of the likelihood is only
+possible within the particle filter;
+in other approaches, such as the Kalman filter,
+the likelihood cannot be calculated using the distance as the estimate
+is a distribution, and so instead a reverse non-deterministic transformation
+is required, which introduces additional error and uncertainty into the model.
 
-% Since observations of the vehicle are irregular, 
-% a \emph{transition function} $f$ is used to
-% predict the vehicle's state at time $t_k$ given the current knowledge of the state at time $t_{k-1}$,
-% as well as accounting for \emph{system noise}, $\sigma_x^2$.
-% We also define a \emph{measurement function} $h$ which describes the relationship between
-% the observable and unobservable states,
-% and accounting for \emph{measurement error}, $\bv_k$,
-% \begin{equation}
-% \label{eq:vehicle_measurement}
-% \by_k = h(\bX_k) + \bv_k = h(x_k) + \bv_k,
-% \end{equation}
-% so all models need to estimate at least $x_k$, the distance traveled along the route.
-
-% The model is split into two steps, prediction and update.
-% The prediction involves using the transition function to predict
-% the next state, $\hat\bX_{k|k-1}$,
-% while the update step using a likelihood function and the measurement function
-% to update the state to account for the observed vehicle location, $\by_k$,
-% giving a final estimate of $\hat\bX_{k|k}$.
-
-% \subsubsection{Predicting state using the transtion function}
-% \label{sec:pf_prediction}
-
-% The transition function is a model of bus behaviour that allows
-% predictions of future states to be made given the most recently 
-% updated state.
-% Developing a good model for bus behaviour allows us to estimate
-% parameters of interest, such as vehicle speed,
-% dwell time at bus stops, and travel time along individual roads,
-% the last of which is of course our primary objective.
-% In this section, we propose [HOW MANY?] increasingly complex models
-% of bus behaviour.
-
-% %% MODEL 1
-% The first model, $\bM_0$, assumes a constant speed between observations
-% using the state
-% $\bX_k = \left[\begin{smallmatrix}x_k && \dot x_k\end{smallmatrix}\right]^\top$
-% and Gaussian system noise with mean zero and variance adjusted by
-% $\Delta_k = t_k - t_{k-1}$.
-% \begin{equation*}
-% \hat\bX_{k|k-1} = f_0(\bX_{k-1|k-1}, \Delta_k, \sigma_x^2) =
-% \begin{bmatrix}
-% x_{k|k-1} & \dot x_{k|k-1}
-% \end{bmatrix}
-% \end{equation*}
-% where
-% \begin{align*}
-% x_{k|k-1} &= x_{k-1|k-1} + \Delta_k \dot x_{k|k-1} \\
-% x_{k|k-1} &= \dot x_{k-1|k-1} + w_k \\
-% w_k &\sim \mathcal{N}_T(0, \Delta_k \sigma_x^2)
-% \end{align*}
-% Here, $\mathcal{N}_T$ denotes a truncated normal distribution such that
-% the system noise is truncated to ensure the speed remains between 0 and 30~m/s.
-
-% While $\bM_0$ may perform well on average,
-% and indeed if all we were interested in was estimating $x_k$ it may be adequate;
-% however, it is missing two important features of bus behaviour:
-% \begin{itemize}
-% \item variable speeds along roads, \emph{which is what we are truly interested in}
-% \item unknown dwell times at bus stops
-% \end{itemize}
-
-% The first of these is incorporated into $\bM_1$,
-% which allows vehicle speeds to vary over time.
-% This model uses the same state as $\bM_0$,
-% but a different transition function $f_1$.
-% In this case, the state equations depend on a sequence of system noise,
-% $\bw_k = \{\bw_k^1, \cdots, \bw_k^{\Delta_k}\}$
-% \begin{align}
-% \label{eq:transition_f1}
-% x_{k|k-1} &= x_{k-1|k-1} + \Delta_k \dot x_{k-1|k-1} + \sum_{i=1}^{\Delta_k} w_k^i \nonumber \\
-% \dot x_{k|k-1} &= \dot x_{k-1|k-1} + \sum_{i=1}^{\Delta_k} w_k^i \\
-% w_k^i &\sim \mathcal{N}_T(0, \sigma_x^2) \nonumber
-% \end{align}
-% Again, system noise is assumed to have a Gaussian distribution, 
-% truncated to ensure that the vehicle's speed is always between 0 and 30~m/s.
-% This is achieved by iteratively estimating each second the the vehicle's state,
-% each time ensuring the speed remains in the specified range.
-
-% The next model, $\bM_2$, allows for dwell times at bus stops.
-% Since buses can either stop and wait, or not stop,
-% multimodality is introduced to the system,
-% which requires a modeling framework, such as the particle filter used here,
-% to handle it.
-% For each bus stop $j = 1, \ldots, M$ along a route,
-% its distance into the trip, $S_j^d$, 
-% is used to determine when a vehicle passes a stop.
-% When this occurs, the dwell time $\delta_j \geq 0$ at stop $j$ is modelled
-% using the stopping probability, $\pi_j$,
-% a common dwell time parameter $\gamma$ which models the
-% deceleration, opening and closing of doors, and acceleration at bus stops,
-% and a stop dwell time $\tau_j$ for passengers alighting and disembarking.
-% Therefore the stop dwell time is
-% \begin{equation}
-% \label{eq:dwell_time}
-% \delta_j = p_j(\gamma + \tilde\delta_j),\quad
-% p_j \sim \mathrm{Bern}(\pi_j),\quad
-% \tilde\delta_j \sim \mathcal{E}(\tau_j)
-% \end{equation}
-
-% To define the transition function for this model, $f_2$,
-% we need to account for multiple dwell times.
-% Let $s_n$ be the index of the next stop,
-% and sample dwell times for each upcoming stop $j = s_n, \ldots, M$,
-% noting that these can be zero.
-% Assuming the vehicle passes stops $s_n$ up to $J-1$,
-% does it reach stop $J$ in the time remaining?
-% \begin{equation}
-% \tilde X_k^J = \min\left\{
-% S_J^d, f_1(\hat X_{k|k-1}, \Delta_k - \sum_{j=s_n}^{J-1}\delta_j)\right\}
-% \end{equation}
-% Taking the maximum of the sequence gives the predicted distance
-% traveled by the vehicle, so
-% \begin{equation}
-% \hat X_{k|k-1} = f_2 (\hat X_{k-1|k-1}, \Delta_k) =
-% \max_{J = s_n, \ldots, M}\left\{
-%     \tilde X_k^J
-% \right\}
-% \end{equation}
-
-% Care needs to be made when generating the sequence,
-% so in practice the state will be calculated recursively to ensure
-% the system noise is consistent.
-% \begin{equation*}
-% \hat{\tilde X}_k^J =
-% \begin{cases}
-%     f_1(\hat{\tilde X}_k^{J+1}, \delta_J) & J < M \\
-%     \hat X_{k-1|k-1} & J = M
-% \end{cases}
-% \end{equation*}
-
-% The final model $\bM_3$ demonstrated here will allow acceleration to vary.
-% \begin{align*}
-% \label{eq:model_m3}
-% x_{k|k-1} &= x(t_k) \\
-% \intertext{where states are recursively calculated back to $t_{k-1}$}
-% x(t) &= x(t - 1) + \dot x(t) \\
-% \dot x(t) &= \dot x(t-1) + \ddot x(t) \\
-% \ddot x(t) &= \ddot x(t-1) + w_t,\quad
-% w_t \sim \mathcal{N}(0,\sigma_x^2)
-% \end{align*}
-% By truncating the acceleration, 
-% the speed of the vehicle can be maintained between 0 and 30~m/s.
-% Additionally, when approaching a bus stop that will be stopped at,
-% the acceleration will be trunctated to below zero,
-% however no physical deceleration will be modeled as this is too complex
-% and stopping behaviour of buses can be sudden 
-% (for example when passengers signal late that they wish to board or disembark).
-% After servicing a stop, the bus will need to accelerate up to speed again.
-
-% \subsubsection{Updating state using the observation likelihood}
-% \label{sec:pf_update}
-
-% The update step involves comparing the predicted state $\bX_{k|k-1}$
-% with the observed state $\bY_k$,
-% using the measurement function $h$ to convert between the two state spaces,
-% and a likelihood function.
-% The measurement function is a deterministic function that uses only the 
-% distance traveled, $x_k$, along a known path $\bR$ to obtain a GPS coordinate.
-% The inverse measurement function, which is used in other approaches,
-% is \emph{not deterministic}, and instead uses an optimisation to
-% find the ``estimated'' distance traveled,
-% which invariably leads to issues in the implementation.
-% By using a particle filter approach, we avoid using $h^{-1}$
-% and instead determine the liklihood using $\bY_k$ directly.
-
-% For the likelihood, the GPS error is assumed symmetrical about the true
-% position of the vehicle, with a radius of $\sigma_y^2$ meters.
-% Define a projection function $g(\cdot | \bY)$ that projects points
-% onto a cartesian surface (``Flat Earth''),
-% so that the distance between points is the real Earth distance.
-% Then the distance between the points is
-% \begin{align*}
-% \delta_k &= \sqrt{\left(g(h(X_k))_1 - g(\bY_k)_1\right)^2 +
-%     \left(g(h(X_k))_2 - g(\bY_k)_2\right)^2} \\
-% \intertext{which can be standardised by the GPS error}
-% \tilde \delta_k &=
-% \sqrt{\left(\frac{g(h(X_k))_1 - g(\bY_k)_1}{\sigma_y}\right)^2 +
-%     \left(\frac{g(h(X_k))_2 - g(\bY_k)_2}{\sigma_y}\right)^2} =\frac{\delta_k}{\sigma_y} \\
-% \intertext{Since we assume symmetric variance, this is actually}
-% \tilde\delta_k &= \sqrt{Z_1^2 + Z_2^2}, \quad
-% Z_1,Z_2 \sim \mathcal{N}(0,1)
-% \end{align*}
-% The sum of two normal random variables has an Exponential distribution with rate $\frac{1}{2}$,
-% therefore 
-% \begin{equation}
-% \label{eq:lhood_exp}
-% \left(\frac{\delta_k}{\sigma_y}\right)^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
-% \end{equation}
-
-% Combining this together, we use a distance function $d(\bY_1, \bY_2)$ that calculates
-% the geographic distance between two GPS coordinates,
-% the likelihood is
-% \begin{equation}
-% \label{eq:lhood}
-% f(\bY_k | \hat\bX_{k|k-1}, \sigma_y^2) =
-% \frac{1}{2} \exp\left\{-\frac{d(\bY_k, h(\hat\bX_{k|k-1}))^2}{2\sigma^2}\right\}
-% \end{equation}
+The posterior distribution is obtained by \emph{selection},
+or resampling with replacement using likelihood weights
+\begin{equation*}
+W_k^{(i)} = \frac{p(\bY_k | \bX_k^{(i)}, \epsilon)}{
+    \sum_{j=1}^N p(\bY_k | \bX_k^{(j)}, \epsilon)
+} 
+W_{k-1}^{(i)} 
+\end{equation*}
+However, to improve efficiency, the state $\tilde \bX_k$ is only resampled
+when the effective sample size 
+$N_{\text{eff}}$ is less than some threshold,
+\begin{equation*}
+N_{\text{eff}} = \frac{1}{\sum_{i=1}^N (W_k^{(i)})} < N_{\text{thres}}
+\end{equation*}
+When resampling is performed, $W_k^{(i)} := N^{-1}$ for all $i$.
 

### 2018-10-05 anzjs2018/sections/03_models.tex
network model

@@
 
-% The second part of this appraoch involves a network model of transit network travel times along links in the network.
-% Reliable ETAs will require a historical data component,
-% used as a prior in the absense of data as well as for
-% predicting future states, 
-% and a realtime update component where the travel times of vehicles
-% are used to update the current estimate of network state.
+The second part of this appraoch involves a network model of transit network travel times along links in the network.
+Reliable ETAs will require a historical data component,
+used as a prior in the absense of data as well as for
+predicting future states, 
+and a realtime update component where the travel times of vehicles
+are used to update the current estimate of network state.
 
-% The model will be another recursive Bayesian model as used for the vehicle state,
-% as this provides a simple way to model realtime data.
-% Since travel times can safely be assumed to be unimodal,
-% we use an information filter (IF),
-% which is a transformation of the Kalman filter
-% using information instead of covariance,
-% and allows the simple combination of multiple observations,
-% for example when multiple buses traverse a road in the same update.
-% The prediction step will involve a transition function that incorporates historical data
-% and the update step will weight between observed and predicted travel times.
+The model will be another recursive Bayesian model as used for the vehicle state,
+as this provides a simple way to model realtime data.
+Since travel times can safely be assumed to be unimodal,
+we use an information filter (IF),
+which is a transformation of the Kalman filter
+using information instead of covariance,
+and allows the simple combination of multiple observations,
+for example when multiple buses traverse a road in the same update.
+The prediction step will involve a transition function that incorporates historical data
+and the update step will weight between observed and predicted travel times.
 
-% The network state $\Theta_c = \{\theta_c^j\}_{j = 1}^J$ is the travel time 
-% of transit vehicles along road segment $j$ at time $t_c$.
-% In this implementation, we are assuming independence between each segment
-% to provide a simple basis for the model; of course in reality this
-% is not true and future work will investigate ways of incorporating this.
+The network state $\boldsymbol\theta_c = \{\theta_c^\ell\}_{\ell = 1}^L$ is the travel time 
+of transit vehicles along road segment $\ell$ at time $t_c$.
+In this implementation, we are assuming independence between each segment
+to provide a simple basis for the model; of course in reality this
+is not true and future work will investigate ways of incorporating this.
 
-% Using an IF requires a transition matrix, $\mathbf{F}_c$;
-% however, in this case the travel time model is constant so no
-% transition matrix is required (i.e., $\mathbf{F}_c = 1$).
-% Let $\Delta_c = t_c - t_{c-1}$ and $P_c^j$ the state uncertainty
-% with system noise $Q_c = \Delta_c \sigma_b^2$, then the update equations are
-% simply the normal KF update equations,
-% \begin{align}
-% \label{eq:kf_transition}
-% \hat\theta^j_{c|c-1} &= \hat\theta^j_{c|c-1} \\
-% P^j_{c|c-1} &= P^j_{c-1|c-1} + Q_c
-% \end{align}
+Since the road state model is very simple, 
+no transition matrix or measurement matrix is required,
+and the model from (\ref{eq:rbe_model}) reduces to
+\begin{equation}
+\begin{split}
+\hat \theta_c^\ell &= \theta_{c-1}^\ell + \Delta_c^\ell v_c^\ell \\
+Z_c^\ell &= \theta_c^\ell + e_c^\ell
+\end{split}
+\end{equation}
+where $\Delta_c = t_c - t_{c-1}$,
+$v_c^\ell \sim \mathrm{N}(0, \nu^2)$ is the system noise,
+and $e_c^\ell \sim \mathrm{N}(0, R_c^\ell)$ is the measurement error
+as estimated from the vehicle model using particle uncertainty.
 
-% For the update step, however, we need to transform into an information
-% space parameterised by the information matrix 
-% $\mathbf{Z}^j_{c|c-1} = P_{c|c-1}^{-1}$ 
-% and the information vector $\mathbf{\hat z}^j_{c|c-1} = \hat\theta_{c|c-1} P_{c|c-1}^{-1}$.
-% The measurement data obtained from the vehicle model are
-% the travel time of vehicle $m$ along segment $j$,
-% $\bar b^m_j$, and the uncertainty $s^m_j$.
-% These can be transformed to a measurement information covaraince matrix
-% and vector using the measurement matrix $\mathbf{H} = 1$ since,
-% in this case, the observed state is the underlying state we are estimating (travel time).
-% \begin{equation}
-% I^m_{jc} = \frac{1}{(s^m_j)^{2}}\quad\text{and}\quad
-% i^m_{jc} = \frac{\bar b^m_j}{(s^m_j)^2}
-% \end{equation}
-% The information update is now the sum of the information for all vehicles
-% that traversed the segment since the last update, so
-% \begin{align*}
-% \mathbf{Z}^j_{c|c} &= \mathbf{Z}^j_{c|c-1} + \sum_{m=1}^M I^m_{jc} \\
-% \mathbf{\hat i}^j_{c|c} &= \mathbf{\hat z}^j_{c|c-1} + \sum_{m=1}^M i^m_{jc}
-% \end{align*}
-% which are easily transformed back to get the desired state parameters.
-% \begin{equation}
-% \hat\theta^j_{c|c} = \frac{\mathbf{\hat z}^j_{c|c}}{\mathbf{Z}^j_{c|c}} 
-% \quad\text{and}\quad
-% P^j_{c|c} = \frac{1}{\mathbf{Z}^j_{c|c}}
-% \end{equation}
+The state is estimated by its mean $\hat \theta_c^\ell = \mathrm{E}(\theta_c^\ell)$
+and uncertainty $P_c^\ell = \mathrm{var}(\theta_c^\ell)$,
+which are predicted from the previous state estimate using the prediction
+equations
+\begin{align*}
+\label{eq:kf_transition}
+\hat \theta^\ell_{c|c-1} &= \hat \theta^\ell_{c|c-1} \\
+P^\ell_{c|c-1} &= P^\ell_{c-1|c-1} + \Delta_c^\ell \nu_c^\ell
+\end{align*}
+
+For the update step, however, we need to transform into an information
+space parameterised by the information matrix $U^\ell_c$
+and the information vector $u^\ell_c$,
+\begin{equation}
+\label{eq:information_transform}
+U^\ell_{c|c-1} = \frac{1}{P_{c|c-1}}
+\quad\text{and}\quad
+\hat u^\ell_{c|c-1} = \frac{\hat \theta^\ell_{c|c-1}}{P^\ell_{c|c-1}}
+\end{equation}
+The measurement data obtained from the vehicle model are
+the travel time of vehicle $m$ along segment $\ell$,
+$\bar z_c^{m\ell}$, and the uncertainty $s^{m\ell}_c$.
+These can be transformed to a measurement information covaraince matrix
+and vector using the measurement matrix,
+\begin{equation*}
+B^{m\ell}_c = \frac{1}{(s^{m\ell}_c)^{2}}\quad\text{and}\quad
+b^{m\ell}_c = \frac{\bar z^{m\ell}_c}{(s^{m\ell}_c)^2}
+\end{equation*}
+The information update is now the sum of the information for all vehicles
+that traversed the segment since the last update, so
+\begin{align*}
+U^\ell_{c|c} &= U^\ell_{c|c-1} + \sum_{m=1}^M B^{m\ell}_{c} \\
+\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^M b^{m\ell}_{c}
+\end{align*}
+which are back transformed to get the updated state parameters
+using the inverse of (\ref{eq:information_transform})
+\begin{equation*}
+\hat \theta^\ell_{c|c} = \frac{\mathbf{\hat u}^\ell_{c|c}}{\mathbf{U}^\ell_{c|c}} 
+\quad\text{and}\quad
+P^\ell_{c|c} = \frac{1}{\mathbf{U}^\ell_{c|c}}.
+\end{equation*}

### 2018-10-08 anzjs2018/sections/04_implementation.tex
Implementation info

@@
-\section{Implementing in realtime}
+\section{Real-time implementation and preliminary results}
 \label{sec:rt}
 
-How we implement it, choice of software (Rcpp = R + C++).
-R: dealing with data structures is easier, maintainability, interfacing
-C++: speed
+There are two components to the application:
+the static GTFS~schedule and network data,
+and the real-time modeling and prediction.
+We chose \verb+Rcpp+ to develop our program,
+giving us the advantages of an R~package for data manipulation 
+(notably \verb+RSQLite+ and \verb+dplyr+)
+and distribution, 
+as well as the speed and memory management capabilities of \verb|C++|. 
+As a result, all of the following has been implemented in the R package
+\verb+transitr+, available on Github (\url{https://github.com/tmelliott/transitr}).
+Here our focus is on the real-time aplication, so we are only discussing the real-time 
+features of model's implementation.
 
-Overall structure:
-- load
-- fetch positions
-- initialize or mutate+update
-- update network
-- make ETA predictions (vehicle state + network state)
+The general structure of the main function is:
+\begin{enumerate}
+\item Load GTFS data from database
+\item Indefinitely repeat when new data recieved \ldots
+\begin{enumerate}
+    \item Update or create new vehicle objects from new data
+    \item Run particle filter on each vehicle to estimate update state
+    \item Collect travel time information for each vehicle for any completed roads
+    \item Update road network with any completed road segments
+    \item Generate ETAs for vehicles
+    \item Write ETAs to extended Google Protobuf binary file for distribution
+\end{enumerate}
+\end{enumerate}
 
-Some of the key things:
-- minimise copy, parallelisation using OMP
-- moving as much computation ``outside'' of the main loop as possible
-  (e.g., ``pre''-predict vehicle/network states so only update required)
-  so ETAs are generated ASAP after retrieving data
-- keeping the GTFS database up-to-date by fetching new data each morning
-- distribution - a cloud database vs a single protobuf file with everything 
-  (maintenance/reliability/speed/size)
+The main concern is speed: 
+we desire ETAs to be available as soon as possible after obtaining the data.
+To do this, we use raw pointers to read-only GTFS objects,
+and particles are only copied when resampling is required.
+This is also why we chose to use protobuf as the output format,
+since this is fast to write and distribute.
+Steps 2b--e are performed in parallel, 
+which in our test environment of a virtual machine with 8~cores, 
+significantly speeds up computation.
+
+As far as results, we have two categories:
+implementation and faesibility,
+and model performance;
+we do not have any complex arrival time estimation method,
+so no results for that are shown here.
+
+% How we implement it, choice of software (Rcpp = R + C++).
+% R: dealing with data structures is easier, maintainability, interfacing
+% C++: speed
+
+% Overall structure:
+% - load
+% - fetch positions
+% - initialize or mutate+update
+% - update network
+% - make ETA predictions (vehicle state + network state)
+
+% Some of the key things:
+% - minimise copy, parallelisation using OMP
+% - moving as much computation ``outside'' of the main loop as possible
+%   (e.g., ``pre''-predict vehicle/network states so only update required)
+%   so ETAs are generated ASAP after retrieving data
+% - keeping the GTFS database up-to-date by fetching new data each morning
+% - distribution - a cloud database vs a single protobuf file with everything 
+%   (maintenance/reliability/speed/size)

### 2018-10-10 anzjs2018/sections/01_introduction.tex
intro very drafty

@@
 
-The need for realtime information.
-Where it goes wrong: 
-- bad schedule calibration;
-- no accounting for realtime traffic 
-- mention how only the delays are used (in Akl), which rely on (often inaccurate) schedules
-Literature review of historical and present methods?
+From the comments posted on social media during outages,
+it would seem that \rt information (RTI) has surpassed being a novelty
+to become an integral component of public transport.
+In Auckland, the RTI provided by Auckland Transport is unreliable at best,
+and often succumbs to several major problems that public transport users
+are all too familiar with.
+The first is inaccurate estimated times of arrival (ETAs),
+which tend to increase over time as the bus gets progressively later
+due to traffic congestion, or simply because the schedule is poorly calibrated.
+The second is buses that are shown as on time but in fact are not running,
+and therefore have not registered on the \rt system in which case the default,
+scheduled times are used as ETAs.
 
-% Bus arrival-time prediction is far from a novel concept: 
-% ever since the first vehicle tracking technology was implemented
-% in transit vehicles in the late 1980's, 
-% real-time models have been implemented to provide \gls{rti} for commuters.
-% The simplest form of this is, the \gls{eta}
-% (usually displayed as \emph{minutes until arrival}),
-% was implemented in {{region}} by \cite{Wall99analgorithm} 
-% using their proposed \gls{kf} algorithm.
-% In their work, they used historical data to obtain \glspl{eta}.
+Arrival time prediction is only as accurate as the underlying model,
+and while a lot of work has gone into developing public transport vehicle models
+\citep{Cathey_2003,Jeong_2005,Yu_2011,Hans_2015},
+in many public transport systems no formal model is used.
+Instead, ETAs are solely based on the scheduled arrival time
+adjusted by the vehicle's delay at the most recent stop, if available.
+This assumes that the schedule is valid,
+and that there is no unusual congestion along the route,
+neither of which are valid assumptions,
+particularly in our test area of Auckland, New Zealand,
+where infrastructure for buses (such as priority lanes) is very limited.
 
-% Over the years, advances in computing power have enabled new
-% modeling approaches,
-% such as support vector machines \citep{Yu_2006},
-% artificial neural networks \citep{Yu_2011},
-% and more recently particle filters \citep{Hans_2015}.
+In order to improve upon this, we first needed to develop a vehicle model
+that could infer the position and speed of vehicles accurately.
+Several options were explored, 
+including the Kalman filter models used by \cite{Dailey_2001} and \cite{Cathey_2003},
+machine learning models as proposed by \cite{Yu_2006} and \cite{Chang_2010},
+however finally decided on using a particle filter approach as demonstrated
+by \cite{Hans_2015}.
+The main reason for this choice is that particle filter models can be 
+flexible and aren't limited by many assumptions,
+instead only by computational demands.
 
-Particle filter has proven useful to this kind of dynamic problem,
-where the state distribution is non-gaussian, non-symmetric, 
-multimodal...
+The second consideration is how to make \emph{reliable} arrival time estimates.
+Ideally, congestion of intermediate roads should be incorporated into the predictive model,
+and one way of doing this is to use the travel time of recent buses along
+the same roads to improve ETA reliability.
+In the past, several novel attempts at this were made,
+from previous-trip (of the same route) as presented by \cite{Yu_2010},
+and more complex versions of this where routes that congverged could
+inform on each other \citep{Yu_2011}.
+In general, however, no comprehensive network modeling approach has been attempted
+as far as being trained and used for public transport predictions.
 
-Any other recent attempts at network-estimation?
-There was a KF using point-to-point speed estimations?
+In this applications paper, we describe a real-time vehicle and network state
+model which uses the vehicles traveling throughout the transit road network
+to update road travel time information.
+The vehicle model is implemented using a particle filter,
+so we also demonstrate the faesibility of this as a valid real-time option.
+From the particle filter estimates of vehicle trajectories we 
+can estimate the vehicles travel time along roads within the network,
+and use these to update the network state.
+This can then be used in future work to develop a more reliable 
+arrival time estimation method.
 
-Computational aspect - historically, models had to be simple 
-(e.g., KF) so they could run in realtime.
-These days, running a somewhat powerful VM on a remote server
-is almost standard practice, so we can take full advantage 
-and use a more computationally demanding (particle filter)
-but more accurate estimation technique - paired with a more complex model.
+% The need for realtime information.
+% Where it goes wrong: 
+% - bad schedule calibration;
+% - no accounting for realtime traffic 
+% - mention how only the delays are used (in Akl), which rely on (often inaccurate) schedules
+% Literature review of historical and present methods?
 
-We first give an overview of the transit network,
-and how we construct it from raw GTFS shape data.
-This paper assumes a very basic model,
-with a focus on the implementation in a realtime setting.
+% % Bus arrival-time prediction is far from a novel concept: 
+% % ever since the first vehicle tracking technology was implemented
+% % in transit vehicles in the late 1980's, 
+% % real-time models have been implemented to provide \gls{rti} for commuters.
+% % The simplest form of this is, the \gls{eta}
+% % (usually displayed as \emph{minutes until arrival}),
+% % was implemented in {{region}} by \cite{Wall99analgorithm} 
+% % using their proposed \gls{kf} algorithm.
+% % In their work, they used historical data to obtain \glspl{eta}.
+
+% % Over the years, advances in computing power have enabled new
+% % modeling approaches,
+% % such as support vector machines \citep{Yu_2006},
+% % artificial neural networks \citep{Yu_2011},
+% % and more recently particle filters \citep{Hans_2015}.
+
+% Particle filter has proven useful to this kind of dynamic problem,
+% where the state distribution is non-gaussian, non-symmetric, 
+% multimodal...
+
+% Any other recent attempts at network-estimation?
+% There was a KF using point-to-point speed estimations?
+
+% Computational aspect - historically, models had to be simple 
+% (e.g., KF) so they could run in realtime.
+% These days, running a somewhat powerful VM on a remote server
+% is almost standard practice, so we can take full advantage 
+% and use a more computationally demanding (particle filter)
+% but more accurate estimation technique - paired with a more complex model.
+
+% We first give an overview of the transit network,
+% and how we construct it from raw GTFS shape data.
+% This paper assumes a very basic model,
+% with a focus on the implementation in a realtime setting.

### 2018-10-11 anzjs2018/sections/01_introduction.tex
type yp edits

@@
 
-From the comments posted on social media during outages,
-it would seem that \rt information (RTI) has surpassed being a novelty
-to become an integral component of public transport.
+% From the comments posted on social media during outages,
+% it would seem that \rt information (RTI) has surpassed being a novelty
+% to become an integral component of public transport.
 In Auckland, the RTI provided by Auckland Transport is unreliable at best,
@@
 The first is inaccurate estimated times of arrival (ETAs),
-which tend to increase over time as the bus gets progressively later
+which tend to either increase or decrease over time as the bus gets progressively later or earlier
 due to traffic congestion, or simply because the schedule is poorly calibrated.
 The second is buses that are shown as on time but in fact are not running,
-and therefore have not registered on the \rt system in which case the default,
+and therefore have not registered on the \rt system in which case the 
 scheduled times are used as ETAs.

### 2018-10-11 anzjs2018/sections/02_gtfs.tex
type yp edits

@@
 (source \url{http://transitfeeds.com}),
-allowing an application developed locally to
+allowing an application developed in Auckland to
 be deployed to another GTFS-based public transport system with minimal modification.
@@
 and a \emph{stop} is a physical location where passengers can embark and disembark
-the vehicle.
+the vehicle \citep{GoogleDevelopers_2006}.
 GTFS realtime feeds provide information such as vehicle positions and trip delays,
@@
 however, there is no direct way of knowing which routes share common roads.
-We are propose using an algorithm to convert the raw GTFS shape files
+Here we propose using an algorithm to convert the raw GTFS shape files
 into a \emph{road network},
@@
 and the connecting paths (road segments or \emph{edges}).
-
-% Before we can usefully model buses in real-time, 
-% we need to construct a \emph{transit network},
-% consisting of \emph{intersections} (nodes)
-% connected by \emph{road segments} (edges).
-% In this way, each \emph{route} 
-% (a journey taken by a transit vehicle, from an origin to a destination by a fixed path)
-% can be represented as a sequence of road segments,
-% each of which we model as vehicles travel along them (see section~\ref{sec:kf}).
+However, in this paper we use bus stops as nodes and the routes between them as edges.
+This captures most overlaps, 
+with the exception of some express services and multiple-stop locations,
+but these are typically in high traffic areas and should be OK.
+Figure~\ref{fig:nw_construct} demonstrates how several overlapping routes 
+are merged to form a road network.
+In section~\ref{sec:kf} we present a model for estimating veihcle travel time
+along a road segment.
 
@@
 
-GTFS realtime allows developers to query the current positions of vehicles
+GTFS \rt allows developers to query the current positions of vehicles
 in the transit network.
 The data consists of the time $t_k$ that the observation was made,
-the GPS position of the vehicle, $Y_k$, 
+the GPS position of the vehicle, $\bY_k$, 
 and some other information about the trip being serviced.
-Vehicle positions are usually updated within 30~seconds,
-but this can vary from 10~seconds to several minutes,
+Vehicle positions are updated with a frequency of anywhere from 10~seconds to several minutes,
 so there is often a lot of uncertainty about the trajectory
@@
 Often these positions appear to be preemptive (i.e., that bus is almost there, 
-but not quiet),
+but not quite),
 and subsequent observations place the bus \emph{behind} the stop or intersection.
-This must therefore be accounted for in the model proposed in the next section.
+To handle this, we compute the approximate distance traveled, $\tilde x_k$,
+of the vehicle by finding the nearest point on the path to the observation;
+if this has decreased, the previous observation is rejected and the vehicle reverted
+to its previous state.
+To enable this, vehicles retain the current and previous states (see section~\ref{sec:rt}).

### 2018-10-11 anzjs2018/sections/03_models.tex
type yp edits

@@
 
-The real-time nature of this application (means that for) both
-the vehicle and network models a recursive Bayesian filtering approach is logical,
-and indeed real-time tracking and robotics applications often use such (models).
-In each of the following models, 
+The real-time nature of this application suggests that
+a recursive Bayesian filtering approach would be appropriate, as is often used in
+real-time tracking and robotics applications.
+We are using two separate models, one each for the vehicle and network states.
+In each,
 we assume an underlying Markov process with state $\mathcal{X}_k$,
 of which observations $\mathcal{Y}_k$ are made,
-which leads us to the following general model,
+giving us to the following general model,
 \begin{equation}
@@
 
-The following sections describe two Bayesian filter models based on (\ref{eq:rbe_model}) used in this application.
-The first model is implemented using a particle filter to estimate vehicle states,
+The following sections describe the two Bayesian filter models based on (\ref{eq:rbe_model}) used in this application.
+The first is implemented using a particle filter to estimate vehicle states,
 with the primary objective of estimating vehicle travel times along roads,
 while the second using these estimated times to update road states,
-and is implemented using a Kalman filter.
+and is implemented using a variant of the Kalman filter referred to as the \emph{information filter}
+\citep{}.
 
@@
 \end{equation}
-Observations of the vehicle are made at time $t_k$ using a GPS,
-giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle,
-giving us the observation vector
+Observations of the vehicle are made using a GPS at time $t_k$,
+giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle
+and stored in the observation vector
 \begin{equation}
@@
 
-We decided to employ a particle filter to estimate $\bX_k$,
-since it is a highly flexible approach and has been used in recent 
+We chose to employ a particle filter to estimate $\bX_k$,
+to do its high flexibility and use in recent 
 transit vehicle modeling applications \citep{Hans_2015}.
 Our main justification for using it is that it handles multimodality very well,
+as demonstrated in figure~\ref{fig:pf_state},
 which is a common feature of the proposal distribution particularly around bus stops.
-Another advantage is the intuitive likelihood function, 
-which is described later.
+Another advantage is the intuitive likelihood function (see section~\ref{sec:pf_update}).
 Conversely, the particle filter is a computationally demanding method,
 as thousands of particles are required per bus.
-Section~\ref{sec:rt} describes how we were able to implement the particle filter in real-time.
+Section~\ref{sec:rt} describes how we were able to implement the particle filter in real-time and the timings.
 
-In the particle filter, the posterior distribution of the state at time $t_k$,
-is estimated by a set of discrete points, or particles,
+In the particle filter, the posterior distribution of the state at time $t_{k-1}$,
+is represented by a set of discrete points, or particles, (fig~\ref{fig:pf_state_prev})
 \begin{equation}
-p(\bX_k | \bY_k) \approx \tilde\bX_{k|k} := (\bX_k^{(i)})_{i=1}^N
+p(\bX_{k-1} | \bY_{k-1}) \approx \tilde\bX_{k-1|k-1} := (\bX_{k-1}^{(i)})_{i=1}^N
 \end{equation}
-each of which is independently updated or \emph{mutated} using the transition function $f$,
+each of which is independently updated or \emph{mutated} using the transition function $f$ (fig~\ref{fig:pf_state_predict})
 \begin{equation}
@@
 After mutation, \emph{selection} of a new set of particles is performed by
-importance resampling using likelihood weights.
+importance resampling using likelihood weights,
+determined by distance between particle's map position and $\bY_k$ (fig~\ref{fig:pf_state_map})
+to obtain a new sample $\tilde\bX_{k|k}$, (fig~\ref{fig:pf_state_update}).
 
@@
 \label{sec:pf_prediction}
-As we are using a particle filter to implement the vehicle model,
-the transition function $f$ can be a complex description of bus behaviour,
-most notably
+
+Another advantage of using a particle filter to implement the vehicle model is that
+the transition function $f$ can implement several bus behaviours,
+most importantly
 \begin{itemize}
 \item non-constant speed along roads (acceleration as system noise),
-\item stopping and waiting at bus stops while passengers board and disembark, and
-\item stopping and waiting at intersections.
+\item optional stopping and waiting at bus stops while passengers board and disembark, and
+\item optional stopping and waiting at intersections.
 \end{itemize}
@@
 how the vehicles acceleration changes as a random process.
-% In its simplest form, the transition function becomes,
-% from Newton's laws of motions,
-% \begin{equation}
-% \begin{split}
-%     x_k &= x_{k-1} + \delta_k \dot x_{k-1} \\
-%     \dot x_k &= \dot x_{k-1} + w_k,
-%     \quad w_k \sim \mathrm{N}(0, \delta_k \sigma^2)
-% \end{split}
-% \end{equation}
-However, the vehicle does not simply travel constantly along the route,
+However, the vehicle does not travel constantly along the route,
 as it needs to service bus stops along the way.
@@
 and based the values on those used by \cite{Hans_2015};
-future work will look at modeling these separately in real-time also.
-Also in this application, due to the complexity of the problem,
-we use bus stops to define road segments.
-Therefore no intersection model is required,
+future work will look at modeling these separately in real-time similarly to \ref{sec:kf}.
+As we are using bus stops to determined segments,
+no intersection model is required,
 only the segment travel times as predicted by the network model at time $t_k$,
@@
 
-After mutating the particle set, the posterior distribution is obtained by
+After mutating the particle set, the posterior distribution $\tilde\bX_{k|k}$ is obtained by
 \emph{selection}, which is implemented by importance resampling using
 likelihood weights.
-To do so, we need to define the measurement function $h$,
+To do so, we define the measurement function $h$,
 and an additional function $g$ which transforms GPS coordinates onto a flat
-surface (the geographical equirectangular projection,
-which allows us to place a bivate normal likelihood on the data.
+surface (the geographical equirectangular projection),
+which allows us to place a bivate normal likelihood on the data
+as shown in figure~\ref{fig:pf_lhood},
 The model for the observation generation is,
@@
 The posterior distribution is obtained by \emph{selection},
-or resampling with replacement using likelihood weights
+or resampling with replacement using the likelihood weights
 \begin{equation*}
@@
 When resampling is performed, $W_k^{(i)} := N^{-1}$ for all $i$.
+When sampling is not performed,
+state estimates use $\{\bX_k^{(i)}, W_k^{(i)}\}$ to obtain weighted estimates.
 
@@
 
-The second part of this appraoch involves a network model of transit network travel times along links in the network.
-Reliable ETAs will require a historical data component,
-used as a prior in the absense of data as well as for
-predicting future states, 
-and a realtime update component where the travel times of vehicles
-are used to update the current estimate of network state.
+[[  blurb  ]]
+% The second part of this appraoch involves a network model of transit network travel times along links in the network.
+% Reliable ETAs will require a historical data component,
+% used as a prior in the absense of data as well as for
+% predicting future states, 
+% and a realtime update component where the travel times of vehicles
+% are used to update the current estimate of network state.
 
-The model will be another recursive Bayesian model as used for the vehicle state,
-as this provides a simple way to model realtime data.
-Since travel times can safely be assumed to be unimodal,
-we use an information filter (IF),
-which is a transformation of the Kalman filter
-using information instead of covariance,
-and allows the simple combination of multiple observations,
-for example when multiple buses traverse a road in the same update.
-The prediction step will involve a transition function that incorporates historical data
-and the update step will weight between observed and predicted travel times.
+To obtain reliable ETAs,
+it is necessary to estimate travel times along roads,
+as this allows the model to react to \rt congestion as detected by other buses.
+Future work will aim to construct a more complex model incorporating historical trends 
+and correlations between roads, 
+however for the current work we are demonstrating our approach
+using a simple model.
+
+The model assums a Gaussian state distribution for travel time,
+truncated necessarily at zero;
+this will not be an issue as the state prediction and data will always be non-negative.
+However, the model must be able to handle multiple observations in a single update,
+which often occurs along major bus routes when several buses are traveling together.
+Therefore we have chosen the information filter as our estimation method.
+
+The information filter is a transormation of the Kalman filter in which the
+\emph{information matrix} and \emph{information vector} are used in place of 
+the covariance matrix and state vector, respectively.
+The advantage is that multiple observations can be added together when updating the state.
+Otherwise the information filter follows the same procedure as other recursive Bayesian models.
 
@@
 of transit vehicles along road segment $\ell$ at time $t_c$.
-In this implementation, we are assuming independence between each segment
+Observations of the state, $Z_c^\ell$, are obtained from the particle filter above.
+We are assuming independence between each segment
 to provide a simple basis for the model; of course in reality this
 is not true and future work will investigate ways of incorporating this.
-
-Since the road state model is very simple, 
+Thanks to the model simplicity,
 no transition matrix or measurement matrix is required,
@@
 
-For the update step, however, we need to transform into an information
+For the update step, the parameters are transformed into an information
 space parameterised by the information matrix $U^\ell_c$
@@
 $\bar z_c^{m\ell}$, and the uncertainty $s^{m\ell}_c$.
-These can be transformed to a measurement information covaraince matrix
-and vector using the measurement matrix,
+These are transformed to a measurement information covariance matrix $B$
+and vector $b$ using
 \begin{equation*}
@@
 \begin{equation*}
-\hat \theta^\ell_{c|c} = \frac{\mathbf{\hat u}^\ell_{c|c}}{\mathbf{U}^\ell_{c|c}} 
+\hat \theta^\ell_{c|c} = \frac{\hat u^\ell_{c|c}}{U^\ell_{c|c}} 
 \quad\text{and}\quad
-P^\ell_{c|c} = \frac{1}{\mathbf{U}^\ell_{c|c}}.
+P^\ell_{c|c} = \frac{1}{U^\ell_{c|c}}.
 \end{equation*}

### 2018-10-16 anzjs2018/sections/02_gtfs.tex
Figures for network and pf

@@
 but these are typically in high traffic areas and should be OK.
-Figure~\ref{fig:nw_construct} demonstrates how several overlapping routes 
+Figure~\ref{fig:network_creation_2} demonstrates how several overlapping routes 
 are merged to form a road network.
@@
 along a road segment.
+
+\begin{figure}[tb]
+    \centering
+    \begin{subfigure}{0.7\textwidth}
+        \centering
+        \includegraphics[width=0.95\textwidth]{figures/02_network_segments_1.pdf}
+        \caption{Raw GTFS route shapes}
+        \label{fig:network_creation_1}
+    \end{subfigure} \\
+    \begin{subfigure}{0.7\textwidth}
+        \centering
+        \includegraphics[width=0.95\textwidth]{figures/02_network_segments_2.pdf}
+        \caption{GTFS-based road network}
+        \label{fig:network_creation_2}
+    \end{subfigure}
+    \caption{Construction of a route network involves combining routes at nodes %
+        (here we are using stops) which are connected by edges (roads). %
+        In (a), the three unique routes, represented by different line types, clearly %
+        overlap in several places. In (b) these have been merged, and the width of each line %
+        represents how many routes use that link.}
+    \label{fig:network_creation}
+\end{figure}
 

### 2018-10-16 anzjs2018/sections/03_models.tex
Figures for network and pf

@@
 importance resampling using likelihood weights,
-determined by distance between particle's map position and $\bY_k$ (fig~\ref{fig:pf_state_map})
+determined by distance between particle's map position and $\bY_k$ 
 to obtain a new sample $\tilde\bX_{k|k}$, (fig~\ref{fig:pf_state_update}).
+
+\begin{figure}[tb]
+    \centering
+    \begin{subfigure}[t]{0.48\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/03_particle_filter_1.pdf}
+        \caption{The vehicle state represented by a sample of discrete points is related to the observable state (GPS position) through the transition function $h$.}
+        \label{fig:pf_state_prev}
+    \end{subfigure}\;\;
+    \begin{subfigure}[t]{0.48\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/03_particle_filter_2.pdf}
+        \caption{The transition function $f$ models the behaviour of each particle,
+            including changes in acceleration and stopping behaviour at bus stops.}
+        \label{fig:pf_state_predict}
+    \end{subfigure}\\
+    \begin{subfigure}[t]{0.48\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/03_particle_filter_6.pdf}
+        \caption{Each particle's distance from the true observation is calculated using
+            the geographic distance between two coordinates, which can then be used
+            to calculate it's weight.}
+        \label{fig:pf_state_update}
+    \end{subfigure}\;\;
+    \begin{subfigure}[t]{0.48\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/03_particle_filter_4.pdf}
+        \caption{Resampling occurs using likelihood weights that use the distance
+            between the particle's location and the vehicles reported location (red cross),
+            resulting in the updated state.}
+        \label{fig:pf_state_predict2}
+    \end{subfigure}
+    \caption{The vehicles state is estimated using a set of discrete points, or particles,
+        each of which is independently mutated to make predictions of the future state. The history of the state estimates can be used to estimate parameters of interest, such as
+        travel times along roads.}
+    \label{fig:pf_state}
+\end{figure}
 

### 2018-10-16 anzjs2018/figures/02_network_segments.R

@@ -1,0 +1,90 @@
+library(tidyverse)
+
+stops <- tibble(x = c(0, 0, 1, 1, 1.2, 2, 2, 3, 4, 5, 5, 6, 6),
+                y = c(2, 1, 2, 1, 0.3, 1.5, 0.8, 1.5, 1.5, 2, 1, 2, 1))
+
+roads <- tibble(id = c(1, 1, 2, 2, 3, 3, 3, 4, 4, 4, 5, 5, 6, 6, 7, 7, 7, 8, 8, 
+                       9, 9, 9, 10, 10, 10, 11, 11, 12, 12, 13, 13,
+                       14, 14, 15, 15, 16, 16, 16, 17, 17),
+                x = c(0, 1, 0, 1, 1, 1.2, 2, 1, 1.2, 2, 2, 3, 1.2, 2, 2, 2.5, 3, 
+                      3, 4, 4, 4.3, 5, 4, 4.3, 5, 5, 6, 5, 6, 2, 3, 3, 4, 3, 4, 4, 4.295, 5,
+                      5, 6),
+                y = c(2, 2, 1, 1, 2, 1.51, 1.51, 1, 1.49, 1.49, 1.49, 1.49, 0.3, 0.8, 0.8, 1.47, 1.47,
+                      1.49, 1.49, 1.51, 1.51, 2, 1.49, 1.49, 1, 2, 2, 1.01, 1.01, 1.51, 1.51,
+                      1.51, 1.51, 1.47, 1.47, 1.47, 1.47, 0.975, 0.99, 0.99)) %>%
+    mutate(lty = ifelse(id %in% c(6, 7, 15, 16, 17), 2, 
+                        ifelse(id %in% c(1, 3, 13, 14, 9, 11), 3, 1)))
+
+p1 <- ggplot() +
+    geom_point(aes(x, y), data = stops, size = 5, pch = 19) +
+    geom_path(aes(x, y, group = id, lty = as.factor(lty)), data = roads) + 
+    theme_void() + theme(legend.position = 'none') +
+    coord_fixed(1)
+
+roads2 <- roads %>% filter(!id %in% c(5, 14, 15, 16, 17)) %>%
+    group_by(id) %>% do({
+        if ((.)$id[1] == 12) {
+            (.) %>% mutate(y = 1)
+        } else if ((.)$id[1] %in% c(13, 8)) {
+            (.) %>% mutate(y = 1.5)
+        } else if ((.)$id[1] %in% c(10)) {
[…]

### 2018-10-16 anzjs2018/figures/03_particle_filter.R

@@ -1,0 +1,146 @@
+library(tidyverse)
+library(RSQLite)
+library(dbplyr)
+library(grid)
+library(gridExtra)
+
+gtfs <- dbConnect(SQLite(), file.path("..", "..", "transitr", "fulldata.db"))
+route <- gtfs %>% tbl("routes") %>% 
+    filter(route_short_name == "110" && route_long_name %like% '%To City%') %>%
+    arrange(desc(version)) %>% head(1) %>% collect()
+trip <- gtfs %>% tbl("trips") %>% 
+    filter(route_id == route$route_id) %>% head(1) %>% collect()
+shape <- gtfs %>% tbl("shapes") %>%
+    filter(shape_id == trip$shape_id) %>% arrange(shape_pt_sequence) %>% collect() %>%
+    mutate(shape_dist_traveled = (.) %>% 
+        select(shape_pt_lon, shape_pt_lat) %>%
+        as.matrix() %>% geosphere::distGeo() %>% cumsum %>% c(0, .))
+dbDisconnect(gtfs)
+
+shapef <- shape %>% filter(shape_dist_traveled > 8000 & shape_dist_traveled < 15000)
+
+p0 <- ggplot() +
+    geom_path(aes(shape_pt_lon, shape_pt_lat), data = shapef) +
+    coord_fixed(1.6) +
+    theme_minimal() +
+    theme(panel.grid = element_blank(), axis.text = element_blank()) +
+    scale_x_continuous(position = 'top') +
+    ylab("Latitude") + xlab("Longitude")
+d0 <- ggplot() +
+    geom_path(aes(x = xx, y = yy), data = tibble(xx = c(8000, 15000), yy = c(0, 0))) +
+    theme_minimal() + 
+    theme(panel.grid = element_blank(), axis.text = element_blank()) +
+    xlab("Distance Traveled") + ylab("") +
+    xlim(8000, 15000)
+
+h <- function(x, shape = shapef) {
+    x <- do.call(rbind, lapply(x, function(X) {
+        if (X <= min(shapef$shape_dist_traveled))
+            return(shapef[1, c('shape_pt_lon', 'shape_pt_lat')])
+        if (X >= max(shapef$shape_dist_traveled))
+            return(shapef[nrow(shapef), c('shape_pt_lon', 'shape_pt_lat')])
[…]

### 2018-10-17 anzjs2018/sections/02_gtfs.tex
edits

@@
 GTFS (general transit feed specification)
-is an API (application programming interface) specification for transit data,
-developed and maintained by Google \citep{GoogleDevelopers_2006}.
-It is used by over 900~transit providers around the world,
+is an API (application programming interface) specification for transit data
+detailing it should be organised,
+making access easier for application developers.
+Developed and maintained by Google \citep{GoogleDevelopers_2006},
+who make use of it in their Google Maps Transit Directions,
+it is used by over 900~transit providers around the world,
 including here in Auckland, New Zealand
-(source \url{http://transitfeeds.com}),
-allowing an application developed in Auckland to
-be deployed to another GTFS-based public transport system with minimal modification.
+(source \url{http://transitfeeds.com}).
+An advantage of this for us is that,
+provided our application depends solely on GTFS data,
+after developing it locally in Auckland it can be deployed to any other GTFS-based
+public transport system with minimal modification.
 
-The two main components of GTFS are \emph{static} and \emph{realtime}.
-Static GTFS includes information about routes, trips, shapes, and stops:
+There are two components to GTFS.
+The first, \emph{GTFS static}, includes information about stops, routes, trips, and shapes:
+a \emph{stop} is a physical location where passengers can embark and disembark the vehicle;
 a \emph{route} is a sequence of two or more stops displayed as a single service;
-a \emph{trip} is an instance of a route occuring at a specific time of day;
-a \emph{shape} is the sequence of points defining a vehicle's path along a route;
-and a \emph{stop} is a physical location where passengers can embark and disembark
-the vehicle \citep{GoogleDevelopers_2006}.
-GTFS realtime feeds provide information such as vehicle positions and trip delays,
-and are usually accessed via an API.
-% The latter is often the only covariate used for predicting arrival time,
-% as previously discussed:
-% while it is a computationally inexpensive method (simply addition),
-% it relies on accurate schedules and cannot respond quickly to realtime events.
+a \emph{trip} is an instance of a route occuring at a specific time of day; and
+a \emph{shape} is the sequence of points defining a vehicle's path along a route
+\citep{GoogleDevelopers_2006}.
+The second component is \emph{GTFS realtime},
+which is only available in a subset of the providers due to the requirement of 
+onboard GPS and a central server.
+It provides a standardised format for sharing vehicle positions and trip delays,
+which are often accessed via an API which developers can use in \rt applications.
+
+Before we can develop our application,
+we need to modify the raw GTFS information to make it possible to detect common roads.
 
@@
 at least directly. 
-Several predictive approaches used \emph{headway},
-the time between consecutive trips along the same route
-\citep{Hans_2015}.
-While this is reasonable for high frequency routes,
+Several predictive approaches using \emph{headway},
+the time between consecutive trips along the same route,
+have been proposed \citep{Hans_2015};
+however, while this is reasonable for high frequency routes,
 low frequncy routes will be unable to react quickly to changes in congestion,
@@
 vehicles servicing other routes but traveling along the same roads
-to estimate arrival times;
-however, there is no direct way of knowing which routes share common roads.
-Here we propose using an algorithm to convert the raw GTFS shape files
-into a \emph{road network},
-as demonstated in Figure~\ref{fig:network_creation}.
-This involves detecting locations where one or more routes meet 
-(intersections or \emph{nodes}),
-and the connecting paths (road segments or \emph{edges}).
-However, in this paper we use bus stops as nodes and the routes between them as edges.
-This captures most overlaps, 
-with the exception of some express services and multiple-stop locations,
-but these are typically in high traffic areas and should be OK.
-Figure~\ref{fig:network_creation_2} demonstrates how several overlapping routes 
+to estimate arrival times.
+The simplest method of identifying overlapping routes is to compare
+the stop sequence: 
+routes with a common sub-sequence of stops must be traveling the same route between them.
+While there are several exceptions, for example express routes 
+and multi-stop locations where there are high numbers of routes,
+it is a viable enough simplication for the purposes of this paper.
+Figure~\ref{fig:network_creation} demonstrates how several overlapping routes 
 are merged to form a road network.
-In section~\ref{sec:kf} we present a model for estimating veihcle travel time
+In section~\ref{sec:kf} we present a model for estimating vehicle travel time
 along a road segment.
@@
 and some other information about the trip being serviced.
-Vehicle positions are updated with a frequency of anywhere from 10~seconds to several minutes,
+Vehicle positions are updated with a frequency of anywhere between 10~seconds and several minutes,
 so there is often a lot of uncertainty about the trajectory
-between two observations, particularly when there is a bus stop
+between two observations, particularly when there is one or more bus stop
 or intersection between them.
@@
 bus stops and some major intersections.
-Often these positions appear to be preemptive (i.e., that bus is almost there, 
-but not quite),
-and subsequent observations place the bus \emph{behind} the stop or intersection.
+Often these positions appear to be preemptive 
+(i.e., the bus is almost there, but not quite),
+and subsequent observations place the bus \emph{behind} the stop or intersection
+(e.g., in a queue of traffic at traffic lights).
 To handle this, we compute the approximate distance traveled, $\tilde x_k$,
@@
 to its previous state.
-To enable this, vehicles retain the current and previous states (see section~\ref{sec:rt}).

### 2018-10-17 anzjs2018/sections/03_models.tex
edits

@@
 
-The real-time nature of this application suggests that
-a recursive Bayesian filtering approach would be appropriate, as is often used in
-real-time tracking and robotics applications.
-We are using two separate models, one each for the vehicle and network states.
-In each,
-we assume an underlying Markov process with state $\mathcal{X}_k$,
-of which observations $\mathcal{Y}_k$ are made,
-giving us to the following general model,
+Due to the real-time nature of this application, we are using
+a recursive Bayesian filtering approach as is often used in
+real-time tracking applications.
+We are using two distinct models for the vehicle and network states,
+and in each assume an underlying Markov process with state $\mathcal{X}_k$
+with system noise $\omega_k\sim\mathrm{N}(0,\mathcal{Q})$ 
+of which observations $\mathcal{Y}_k$ are made
+with error $\nu_k\sim\mathrm{N}(0,\mathcal{R})$,
+giving the following general model below.
 \begin{equation}
@@
 \end{equation}
-where $\omega_k\sim\mathrm{N}(0,\mathcal{Q})$ is the system noise,
-and $\nu_k\sim\mathrm{N}(0,\mathcal{R})$ is measurement error.
 The transition function $f$ determines the relationship between consecutive states,
@@
 
-The following sections describe the two Bayesian filter models based on (\ref{eq:rbe_model}) used in this application.
+The following sections describe the two models used in this application.
 The first is implemented using a particle filter to estimate vehicle states,
-with the primary objective of estimating vehicle travel times along roads,
-while the second using these estimated times to update road states,
-and is implemented using a variant of the Kalman filter referred to as the \emph{information filter}
-\citep{}.
+with the primary objective of infering vehicle travel times along roads
+from a sequence of GPS positions,
+while the second uses these travel times to update road states 
+and is implemented using an \emph{information filter},
+which is a variant of the Kalman filter \citep{}.
 
@@
 the vehicle's distance traveled $x$ in meters along the route and
-its speed $\dot x$ in meters per second.
-Additionally, we include travel times along road segments along the route, 
+it's speed $\dot x$ in meters per second.
+Additionally, travel times along road segments along the route, 
 $\bz = (z_1,\ldots,z_L)^\top$, in seconds,
-which are independent of time but included in the state as they are estimated
+are independent of time but included in the state so they can be estimated
 sequentially as the vehicle traverses the route.
@@
 \end{equation}
-Observations of the vehicle are made using a GPS at time $t_k$,
-giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle
-and stored in the observation vector
+Observations $\bY_k$ of the vehicle are made using a GPS at time $t_k$
+with GPS error $\epsilon^2$,
+giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle,
 \begin{equation}
@@
 We chose to employ a particle filter to estimate $\bX_k$,
-to do its high flexibility and use in recent 
+due to do its high flexibility and use in recent 
 transit vehicle modeling applications \citep{Hans_2015}.
-Our main justification for using it is that it handles multimodality very well,
-as demonstrated in figure~\ref{fig:pf_state},
-which is a common feature of the proposal distribution particularly around bus stops.
-Another advantage is the intuitive likelihood function (see section~\ref{sec:pf_update}).
-Conversely, the particle filter is a computationally demanding method,
-as thousands of particles are required per bus.
-Section~\ref{sec:rt} describes how we were able to implement the particle filter in real-time and the timings.
-
-In the particle filter, the posterior distribution of the state at time $t_{k-1}$,
-is represented by a set of discrete points, or particles, (fig~\ref{fig:pf_state_prev})
-\begin{equation}
-p(\bX_{k-1} | \bY_{k-1}) \approx \tilde\bX_{k-1|k-1} := (\bX_{k-1}^{(i)})_{i=1}^N
-\end{equation}
-each of which is independently updated or \emph{mutated} using the transition function $f$ (fig~\ref{fig:pf_state_predict})
-\begin{equation}
-p(\bX_k | \bX_{k-1}) \approx \tilde\bX_{k|k-1} := 
-\left(f(\bX_{k-1}^{(i)}, \psi)\right)_{i=1}^N
-\end{equation}
-using a parameter vector $\psi$ containing all of the necessary parameters
-for the model (including system noise).
-After mutation, \emph{selection} of a new set of particles is performed by
-importance resampling using likelihood weights,
-determined by distance between particle's map position and $\bY_k$ 
-to obtain a new sample $\tilde\bX_{k|k}$, (fig~\ref{fig:pf_state_update}).
+The primary advante of a particle filter is its handling of multimodality,
+as in figure~\ref{fig:pf_state_predict},
+which is a common feature of the proposal distribution, particularly around bus stops.
+A second advantage is the intuitive likelihood function (see section~\ref{sec:pf_update}).
+Conversely, particle filter methods are computationally demanding,
+requiring an increasing number of particles as the model complexity and
+number of parameters in increase \citep{}.
+Section~\ref{sec:rt} describes our implementation of the particle filter in real-time 
+and it's timings.
 
@@
 
+In a particle filter, the posterior distribution of the state at time $t_{k-1}$,
+is represented by a set of discrete points, or particles, (fig~\ref{fig:pf_state_prev}),
+each with an associated weight $W_{k-1}^{(i)}$
+\begin{equation}
+p(\bX_{k-1} | \bY_{k-1}) \approx \tilde\bX_{k-1|k-1} 
+:= \{\bX_{k-1}^{(i)}, W_{k-1}^{(i)}\}_{i=1}^N
+\end{equation}
+each of which is independently updated or \emph{mutated} using the transition function $f$ (fig~\ref{fig:pf_state_predict})
+\begin{equation}
+p(\bX_k | \bX_{k-1}) \approx \tilde\bX_{k|k-1} := 
+\left\{f(\bX_{k-1}^{(i)}, \psi), W_{k-1}^{(i)}\right\}_{i=1}^N
+\end{equation}
+using parameter vector $\psi$ containing all of the necessary parameters
+for the model (including system noise).
+After mutation, 
+the particles are reweighted using the likelihood $p(\bY_k | \bX_k^{(i)})$ 
+(sec.~\ref{sec:pf_update}) and standardised,
+\begin{equation*}
+W_k^{(i)} = \frac{W_{k-1}^{(i)} p(\bY_k | \bX_k^{(i)})}{
+    \sum_{j=1}^N W_{k-1}^{(j)} p(\bY_k | \bX_k^{(j)})
+}
+\end{equation*}
+generating a posterior estimate of the state
+\begin{equation}
+p(\bX_k | \bY_k) \approx \tilde\bX_{k|k} := 
+\{\bX_{k}^{(i)}, W_{k}^{(i)}\}_{i=1}^N.
+\end{equation}
+
+In order to prevvent degeneration of the particle sample,
+\emph{importance resampling} is employed whenever the 
+effective sample size $N_{\text{eff}} = 1/(\sum_i (W_k^{(i)})^2)$
+falls below a specified threshold $N_{\text{thres}}$.
+The resampling is performed with replacement using importance weights $W_k^{(i)}$,
+as demonstrated in Figure~\ref{fig:pf_state_update}.
+Afterwards, the weights are reinitialized to $W_k^{(i)} = N^{-1}$.
+
 \subsubsection{Vehicle transition function}
@@
 
-Another advantage of using a particle filter to implement the vehicle model is that
-the transition function $f$ can implement several bus behaviours,
-most importantly
+Using a particle filter allows us to develop a vehicle model in which
+the transition function $f$ implements several key bus behaviours:
 \begin{itemize}
-\item non-constant speed along roads (acceleration as system noise),
+\item non-constant speed along roads,
 \item optional stopping and waiting at bus stops while passengers board and disembark, and
-\item optional stopping and waiting at intersections.
+\item optional stopping and waiting at intersections/traffic lights (future work).
 \end{itemize}
@@
 using system noise parameter $\sigma^2$ which describes 
-how the vehicles acceleration changes as a random process.
-However, the vehicle does not travel constantly along the route,
-as it needs to service bus stops along the way.
+how the vehicle's acceleration changes as a random process.
+The vehicle does not travel constantly along the route,
+however, as it needs to service bus stops along the way.
 Therefore, the transition function includes 
@@
 board and disembark (conditional on the bus stopping),
-and the minimum dwell time at stops, $\gamma$.
+and the minimum dwell time at stops, $\gamma$,
+for the doors to open and close.
 For these parameters, we used constant values for all stops,
 and based the values on those used by \cite{Hans_2015};
-future work will look at modeling these separately in real-time similarly to \ref{sec:kf}.
-As we are using bus stops to determined segments,
-no intersection model is required,
-only the segment travel times as predicted by the network model at time $t_k$,
+future work will investigate modeling these in real-time similarly to Section~\ref{sec:kf}.
+No intersection model is implemented here,
+as the segment road segments are between stops,
+and the travel times predicted by the network model at time $t_k$,
 $\btheta(t_k) = (\theta_1(t_k), \ldots, \theta_L(t_k))^\top$
 (see section~\ref{sec:kf}).
-Future work will adapt this to use intersection that are indpendent 
-of bus stops, and therefore completely independent of routes,
-which will allow increasing the complexity of the model to incorporate
-intersection stopping probabilities and waiting times.
-The details of the transition function are given in the algorithm defined in the Appendix,
+The transition function algorithm is shown in Figure~\ref{fig:algorithm},
 which implements those features discussed above.
@@
 
-After mutating the particle set, the posterior distribution $\tilde\bX_{k|k}$ is obtained by
-\emph{selection}, which is implemented by importance resampling using
-likelihood weights.
-To do so, we define the measurement function $h$,
+After mutating the particle set, the posterior distribution $\tilde\bX_{k|k}$ 
+is updated by reweighting each particle using the likelihood.
+The measurement function $h$,
 and an additional function $g$ which transforms GPS coordinates onto a flat
 surface (the geographical equirectangular projection),
-which allows us to place a bivate normal likelihood on the data
-as shown in figure~\ref{fig:pf_lhood},
+allow the placement of a bivate normal likelihood on the data
+as shown in figure~\ref{fig:pf_lhood}.
 The model for the observation generation is,
@@
 \end{equation}
-However, due to the assumption that error in longitude and latitude are uncorrelated, 
-the geographical distance between the observation and the true vehicle can be expressed
-in terms of two independent standard normal random variables $z_1$ and $z_2$
+Now, the geographical distance between the observation and the true vehicle can be expressed
+in terms of two independent random variables $z_1, z_2 \sim \mathrm{N}(0,1)$
 \begin{equation}
@@
 dist(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
-    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}
+    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}.
 \end{equation}
-As the sum of two independent, standard normal random variables 
+The sum of two independent, standard normal random variables 
 is $\chi^2$ distributed with 2~degrees of freedom,
 which is itself exponential,
-rearranging (\ref{eq:obs_dist}) yields
+so rearranging (\ref{eq:obs_dist}) yields
 \begin{equation}
@@
 It is worth noting that this representation of the likelihood is only
-possible within the particle filter;
+possible due to the discretization of the state provided by the particle filter;
 in other approaches, such as the Kalman filter,
 the likelihood cannot be calculated using the distance as the estimate
-is a distribution, and so instead a reverse non-deterministic transformation
-is required, which introduces additional error and uncertainty into the model.
-
-The posterior distribution is obtained by \emph{selection},
-or resampling with replacement using the likelihood weights
-\begin{equation*}
-W_k^{(i)} = \frac{p(\bY_k | \bX_k^{(i)}, \epsilon)}{
-    \sum_{j=1}^N p(\bY_k | \bX_k^{(j)}, \epsilon)
-} 
-W_{k-1}^{(i)} 
-\end{equation*}
-However, to improve efficiency, the state $\tilde \bX_k$ is only resampled
-when the effective sample size 
-$N_{\text{eff}}$ is less than some threshold,
-\begin{equation*}
-N_{\text{eff}} = \frac{1}{\sum_{i=1}^N (W_k^{(i)})} < N_{\text{thres}}
-\end{equation*}
-When resampling is performed, $W_k^{(i)} := N^{-1}$ for all $i$.
-When sampling is not performed,
-state estimates use $\{\bX_k^{(i)}, W_k^{(i)}\}$ to obtain weighted estimates.
+is a distribution, and so instead a reverse non-deterministic transformation,
+the inverse measurement function,
+is required to estimate the ``observed'' distance traveled $\hat Z_k~=~h^{-1}(\bY_k)$, 
+which introduces additional error and uncertainty into the model.
 
@@
 
-[[  blurb  ]]
-% The second part of this appraoch involves a network model of transit network travel times along links in the network.
-% Reliable ETAs will require a historical data component,
-% used as a prior in the absense of data as well as for
-% predicting future states, 
-% and a realtime update component where the travel times of vehicles
-% are used to update the current estimate of network state.
+The purpose of the road network model is to provide a way of using
+spatial information about vehicle speeds to inform arrival time predictions.
+The simplest notion would be to map each vehicle speed estimate to a physical road;
+however, this has two problems
+\begin{enumerate}
+\item roads usually have two directions, so vehicle directionality would need to be computed,
+\item road data is not provided in GTFS, and can be difficult to obtain cleanly without
+    manual intervention.
+\end{enumerate}
+The most desirable solution would be to use the GTFS shape data to construct a network
+by merging overlapping routes and using merge points as intersections
+to define road segments.
+Initial efforts at applying this had poor results due to overly long or short segments,
+so for this current work we are approximating the road network by using the stops 
+along a route as intersection, and the legs between them as road segments,
+as shown in Figure~\ref{fig:network_creation}.
 
 To obtain reliable ETAs,
-it is necessary to estimate travel times along roads,
-as this allows the model to react to \rt congestion as detected by other buses.
+road travel times need to be estimated in \rt,
+as this allows the model to react to congestion changes as detected by preceeding buses.
 Future work will aim to construct a more complex model incorporating historical trends 
@@
 however for the current work we are demonstrating our approach
-using a simple model.
+using a simple model,
+in which each road segment is treated independently.
 
@@
 of transit vehicles along road segment $\ell$ at time $t_c$.
-Observations of the state, $Z_c^\ell$, are obtained from the particle filter above.
-We are assuming independence between each segment
-to provide a simple basis for the model; of course in reality this
-is not true and future work will investigate ways of incorporating this.
-Thanks to the model simplicity,
+Observations of the state, $Z_c^\ell$, are obtained from the particle filter 
+in Section~\ref{sec:pf}.
+Due to the simplicity of the model,
 no transition matrix or measurement matrix is required,
-and the model from (\ref{eq:rbe_model}) reduces to
+and the model from (\ref{eq:rbe_model}) reduces simply to
 \begin{equation}
 \begin{split}
-\hat \theta_c^\ell &= \theta_{c-1}^\ell + \Delta_c^\ell v_c^\ell \\
+\theta_c^\ell &= \theta_{c-1}^\ell + \Delta_c^\ell v_c^\ell \\
 Z_c^\ell &= \theta_c^\ell + e_c^\ell
@@
 and $e_c^\ell \sim \mathrm{N}(0, R_c^\ell)$ is the measurement error
-as estimated from the vehicle model using particle uncertainty.
+as estimated from the vehicle model using uncertainty of particle estimates.
 
@@
 $\bar z_c^{m\ell}$, and the uncertainty $s^{m\ell}_c$.
-These are transformed to a measurement information covariance matrix $B$
-and vector $b$ using
+These are transformed to a measurement information covariance matrix $B_c^{m\ell}$
+and measurement information vector $b_c^{m\ell}$,
 \begin{equation*}
@@
 \end{equation*}
-The information update is now the sum of the information for all vehicles
+The information update is now the sum of the information over all $M$ vehicles
 that traversed the segment since the last update, so
@@
 \end{align*}
-which are back transformed to get the updated state parameters
+which are back transformed
 using the inverse of (\ref{eq:information_transform})
+to obtain the desired parameter estimates
 \begin{equation*}

### 2018-10-23 anzjs2018/sections/04_implementation.tex
We want to track the figures

@@
 
+The simulations we discuss here use a subset of real data from 8~October, 2018.
+This allows us to run the model with different parameters over the same data.
+The parameters varied were the number of particles,
+GPS error $\epsilon^2$, and system noise $\sigma^2$.
+The primary goal is to reduce the rate of degeneration and ``losing'' the bus 
+(i.e., no particles are near the observed location),
+and reducing the variance of parameter estimates; here we use vehicle speed as the example.
+
+\subsection{Program Timings}
+\label{sec:timings}
+
+During development of our application,
+our primary concert was the computational faesibility of a particle filter.
+For each iteration, 
+the timings of the individual components in 2a--f above are recorded.
+Since the number of vehicles traveling at any given time changes throughout the day,
+we used the average timings from 13:45--14:00 to compare.
+
+Figure~\ref{fig:timings} shows the timing results for varying number of particles.
+
+\begin{figure}[tb]
+    \centering
+    \includegraphics[width=0.7\textwidth]{figures/04_model_results_timing.pdf}
+    \caption{The timings for various parts of the application, and overall. %
+        The trend is approximately linear with the number of particles.}
+    \label{fig:figure1}
+\end{figure}
+
+\subsection{Model performance}
+\label{sec:model_perf}
+
+To check the model results, 
+we need to realise that we are not trying to accurately predict
+future locations of the vehicle,
+but instead are trying to use a sequence of GPS positions to infer travel
+times along intermediate roads.
+Therefore, several tools used to assess the validity of a particle filter
+are unnecessary.
+Instead, we will be comparing the rate of degeneration 
+(i.e., the particle sample loses the actual bus and needs to be reinitialized)
+and the mean distance between the state estimate and the observation.
+[We also explore the uncertainty of travel time estimates,
+as well as the variability between consecutive buses]\footnote{assuming we can get the results in time}.
+
 % How we implement it, choice of software (Rcpp = R + C++).

### 2018-10-23 anzjs2018/figures/02_network_segments.R

@@ -49,42 +49,42 @@ ggsave("figures/02_network_segments_2.pdf", p2, device = "pdf", width = 6, heigh
 
 
 
-library(RSQLite)
-library(dbplyr)
+# library(RSQLite)
+# library(dbplyr)
 
-DBDIR <- file.path("..", "..", "transitr")
+# DBDIR <- file.path("..", "..", "transitr")
 
-gtfs <- dbConnect(SQLite(), file.path(DBDIR, "fulldata.db"))
+# gtfs <- dbConnect(SQLite(), file.path(DBDIR, "fulldata.db"))
 
-## get some routes
-vmax <- gtfs %>% tbl("routes") %>% 
-    summarize(version = max(version, na.rm = TRUE)) %>% collect %>% pluck('version')
-routes_to_use <- c("22N", "22A", "24B", "27W", "30", "70")
-routes <- gtfs %>% tbl("routes") %>% 
-    filter(route_short_name %in% routes_to_use && version == vmax &&
-           (route_long_name %like% '%To City%' || route_long_name %like% '%To Britomart%')) %>%
-    select(route_id, route_short_name) %>% collect %>%
-    group_by(route_short_name) %>%
-    summarize(route_id = first(route_id))
+# ## get some routes
+# vmax <- gtfs %>% tbl("routes") %>% 
+#     summarize(version = max(version, na.rm = TRUE)) %>% collect %>% pluck('version')
+# routes_to_use <- c("22N", "22A", "24B", "27W", "30", "70")
+# routes <- gtfs %>% tbl("routes") %>% 
+#     filter(route_short_name %in% routes_to_use && version == vmax &&
+#            (route_long_name %like% '%To City%' || route_long_name %like% '%To Britomart%')) %>%
+#     select(route_id, route_short_name) %>% collect %>%
+#     group_by(route_short_name) %>%
+#     summarize(route_id = first(route_id))
 
-trips <- gtfs %>% tbl("trips") %>% filter(route_id %in% routes$route_id) %>% 
-    collect %>% group_by(trip_id) %>% summarize(shape_id = first(shape_id))
+# trips <- gtfs %>% tbl("trips") %>% filter(route_id %in% routes$route_id) %>% 
+#     collect %>% group_by(trip_id) %>% summarize(shape_id = first(shape_id))
 
-shapes <- gtfs %>% tbl("shapes") %>%
[…]

### 2018-10-23 anzjs2018/figures/04_model_results.R

@@ -1,0 +1,64 @@
+library(tidyverse)
+
+times <- do.call(bind_rows, lapply(list.files("../../transitr/simulations", pattern = "sim_", full = TRUE), function(dir) {
+    if (!file.exists(file.path(dir, "timings.csv"))) return(NULL)
+    sim <- basename(dir)
+    siminfo <- strsplit(sim, "_")[[1]][-1]
+    if (grepl("e", siminfo[3])) siminfo[3] <- format(as.numeric(siminfo[3]), scientific = FALSE)
+    siminfo <- as.numeric(gsub("-", ".", siminfo))
+    read_csv(file.path(dir, "timings.csv")) %>%
+        mutate(sim = sim, n_particles = siminfo[1], gps_error = siminfo[2], system_noise = siminfo[3],
+            timestamp = as.POSIXct(timestamp, origin = "1970-01-01"))
+}))
+
+ggplot(times, aes(x = timestamp)) +
+    geom_point(aes(y = wall/1000, colour = as.factor(n_particles))) +
+    facet_grid(what~., scales = "free_y")
+    # geom_point(aes(y = cpu), colour = "orangered") +
+
+## between 13:45 and 14:00
+date <- format(times$timestamp[1], "%Y-%m-%d")
+trange <- as.POSIXct(paste(date, c("13:30", "14:00")))
+
+ttot <- times %>% group_by(sim, iteration) %>%
+    summarize(cpu = sum(cpu), wall = sum(wall), timestamp = first(timestamp), nvehicles = first(nvehicles),
+        n_particles = first(n_particles), gps_error = first(gps_error), system_noise = first(system_noise)) %>%
+    mutate(what = "all")
+tsmry <- times %>% bind_rows(ttot) %>%
+    filter(timestamp >= trange[1] & timestamp <= trange[2]) %>%
+    group_by(sim, what) %>%
+    summarize(cpu = mean(cpu), wall = mean(wall),
+        n_particles = first(n_particles), gps_error = first(gps_error), system_noise = first(system_noise)) %>%
+    ungroup() %>%
+    filter(what %in% c("all", "predicting ETAs", "updating vehicle states", "writing ETAs to protobuf feed"))
+
+ggplot(tsmry) +
+    geom_point(aes(n_particles, cpu, color = what)) +
+    ylab("Wall time (seconds)")
+
[…]

### 2018-10-23 anzjs2018/figures/04_model_results.R
dist fig

@@ -11,10 +11,10 @@ times <- do.call(bind_rows, lapply(list.files("../../transitr/simulations", patt
             timestamp = as.POSIXct(timestamp, origin = "1970-01-01"))
 }))
 
-ggplot(times, aes(x = timestamp)) +
-    geom_point(aes(y = wall/1000, colour = as.factor(n_particles))) +
-    facet_grid(what~., scales = "free_y")
-    # geom_point(aes(y = cpu), colour = "orangered") +
+# ggplot(times, aes(x = timestamp)) +
+#     geom_point(aes(y = wall/1000, colour = as.factor(n_particles))) +
+#     facet_grid(what~., scales = "free_y")
+#     # geom_point(aes(y = cpu), colour = "orangered") +
 
 ## between 13:45 and 14:00
 date <- format(times$timestamp[1], "%Y-%m-%d")
@@ -32,9 +32,9 @@ tsmry <- times %>% bind_rows(ttot) %>%
     ungroup() %>%
     filter(what %in% c("all", "predicting ETAs", "updating vehicle states", "writing ETAs to protobuf feed"))
 
-ggplot(tsmry) +
-    geom_point(aes(n_particles, cpu, color = what)) +
-    ylab("Wall time (seconds)")
+# ggplot(tsmry) +
+#     geom_point(aes(n_particles, cpu, color = what)) +
+#     ylab("Wall time (seconds)")
 
 lvls <- c("predicting ETAs", "updating vehicle states", "writing ETAs to protobuf feed", "all")
 tdat <- bind_rows(times, ttot) %>% filter(what %in% lvls) %>%
@@ -53,12 +53,43 @@ ggplot(tdat %>% summarize(cpu.hat = mean(cpu), cpu.var = sd(cpu))) +
     labs(shape = "")
 dev.off()
 
-## CPU timings
-ggplot(tsmry %>% group_by(what, n_particles) %>% 
-        summarize(wall.hat = mean(wall/1000), wall.var = sd(wall/1000))) +
-    geom_errorbar(aes(n_particles, 
-        ymin = wall.hat - wall.var, ymax = wall.hat + wall.var, 
-        color = what, shape = what), 
-        width = 100) +
-    xlab("Number of particles") +
-    ylab("CPU Time (seconds)")
+# ## CPU timings
+# ggplot(tsmry %>% group_by(what, n_particles) %>% 
+#         summarize(wall.hat = mean(wall/1000), wall.var = sd(wall/1000))) +
+#     geom_errorbar(aes(n_particles, 
[…]

### 2018-10-23 anzjs2018/sections/04_implementation.tex
some words

@@
 This allows us to run the model with different parameters over the same data.
-The parameters varied were the number of particles,
-GPS error $\epsilon^2$, and system noise $\sigma^2$.
+The parameters varied were the number of particles and system noise $\sigma^2$.
 The primary goal is to reduce the rate of degeneration and ``losing'' the bus 
@@
 and reducing the variance of parameter estimates; here we use vehicle speed as the example.
+
+The one parameter we can determine from the data is GPS error, $\epsilon^2$,
+by looking at the distance between observations and the shape.
+Figure~\ref{fig:gps_dist} shows the distribution of distance between the vehicle
+observation and the \emph{nearest point on the path}.
+Due to the way the observations are reported, 
+many observations will be ``exactly'' on the route at a bus stop,
+and we need to ignore these.
+
+\begin{figure}[tb]
+    \centering
+    % \includegraphics[width=0.7\textwidth]{figures/04_model_results_gps.pdf}
+    \caption{The distance between the observation and the nearest point on the shape.}
+    \label{fig:gps_dist}
+\end{figure}
 
@@
         The trend is approximately linear with the number of particles.}
-    \label{fig:figure1}
+    \label{fig:timings}
 \end{figure}
@@
 (i.e., the particle sample loses the actual bus and needs to be reinitialized)
-and the mean distance between the state estimate and the observation.
+and the mean distance between the state estimate and the observation,
 [We also explore the uncertainty of travel time estimates,

### 2018-10-24 anzjs2018/sections/04_implementation.tex
more words

@@
 Figure~\ref{fig:timings} shows the timing results for varying number of particles.
+Nothing too exciting going on here, 
+except that writing ETAs takes up the largest chunk of time,
+as this involves calculating quantiles which require the estimated ETAs to be sorted,
+which is computationally demanding. 
+The vehicle update takes the second longest time,
+and involves copying particles during the update resampling step.
+Finally, predicting the ETAs is pretty quick.
+The remaining parts of the application take negligible time,
+and all run on a single core to avoid mutex locks.
 
@@
 (i.e., the particle sample loses the actual bus and needs to be reinitialized)
-and the mean distance between the state estimate and the observation,
+and the mean distance between the state estimate and the observation; 
+we will use the distance between the observation and the route (cross track distance) 
+to determine how well the posterior distribution fits the data---%
+if the ratio between the CTD and RMSE is large,
+then the fit is ``bad''.
 [We also explore the uncertainty of travel time estimates,
 as well as the variability between consecutive buses]\footnote{assuming we can get the results in time}.
+
+From the Figure~X, 
+we see a trade-off between some of the parameters.
+Larger values of GPS error decrease the degeneration rate
+as well as the rate of resampling (which increases speed);
+however, the cost of this is that the mean distance between the posterior distribution
+and the observed location increases, giving a less precise estimate of the vehicle's state.
+Conversely, the system noise does not have the same effect,
+and only slightly effects the values.
+
+So the main way to determine how well everything is working is to look 
+at the RMSE of ETAs over time.
 

### 2018-10-25 anzjs2018/figures/04_model_results.R
pictures

@@ -66,8 +66,10 @@ dev.off()
 
 
 
-
 get_sim_files <- function(sim) {
+    if (file.exists(file.path("../../transitr/simulations", sim, "modeleval.rds"))) {
+        return(readRDS(file.path("../../transitr/simulations", sim, "modeleval.rds")))
+    }
     x <- try({
         siminfo <- strsplit(sim, "_")[[1]][-1]
         if (grepl("e", siminfo[3])) siminfo[3] <- format(as.numeric(siminfo[3]), scientific = FALSE)
@@ -76,20 +78,35 @@ get_sim_files <- function(sim) {
             lapply(list.files(file.path("../../transitr/simulations", sim, "modeleval"), pattern="vehicle_.*\\.csv", full.names = TRUE), 
                 function(x) 
                     read_csv(x, 
-                        col_names = c("vehicle_id", "trip_id", "ts", "prior_mse", "posterior_mse", 
-                                      "prior_speed_var", "posterior_speed_var", "dist_to_path", "Neff", "resample", "n_resample"),
-                        col_types = "cciddddddi", progress = FALSE) %>%
+                        col_names = c("vehicle_id", "trip_id", "ts", "prior_mse", "posterior_mse", #"sumwt", "varwt",
+                                      "post_speed", "prior_speed_var", "posterior_speed_var", "dist_to_path", 
+                                      "Neff", "resample", "n_resample", "bad_sample"),
+                        col_types = "ccidddddddiii", progress = FALSE) %>%
                     mutate(ts = as.POSIXct(ts, origin = "1970-01-01"))
             )
         ) %>% mutate(sim = sim, n_particles = siminfo[1], gps_error = siminfo[2], system_noise = siminfo[3])
     })
     if (inherits(x, "try-error")) return(NULL)
+    saveRDS(x, file.path("../../transitr/simulations", sim, "modeleval.rds"))
     x
 }
 
-res <- get_sim_files(list.files("../../transitr/simulations", pattern = "sim_")[1])
+
+res <- do.call(bind_rows, lapply(list.files("../../transitr/simulations", pattern = "sim_"), get_sim_files))
 
 ggplot(res %>% filter(dist_to_path < 5), aes(dist_to_path)) + geom_density()
-ggplot(res %>% filter(dist_to_path < 50 & dist_to_path > 0), aes(dist_to_path)) + geom_density()
 
[…]

### 2018-10-28 anzjs2018/sections/01_introduction.tex
Edits

--- 2018-08-27 anzjs2018/sections/01_introduction.tex
@@
 
-The need for realtime information.
-Where it goes wrong: 
-- bad schedule calibration;
-- no accounting for realtime traffic 
-- mention how only the delays are used (in Akl), which rely on (often inaccurate) schedules
-Literature review of historical and present methods?
+In public transport, \rt information (RTI), for example estimated times of arrival (ETAs),
+keeps transport users informed about their journey,
+and in some places make decisions on the best route to take.
+Unfortunately, here in Auckland, RTI is highly unreliable,
+and succumbs to several problems that public transport users
+are all too familiar with.
+One of these is inaccurate ETAs,
+which tend to either increase or decrease over time as the bus gets progressively later or earlier
+due to traffic congestion, or  because the schedule is poorly calibrated.
+A second is that buses are shown as on time but in fact are not running,
+and therefore have not registered on the \rt system.
+In this case, the scheduled times are used for ETAs and, 
+when the bus doesn't show up, 
+the bus disappeared from the \rt board and we get what passengers have described as ``Ghost Buses''.
 
-% Bus arrival-time prediction is far from a novel concept: 
-% ever since the first vehicle tracking technology was implemented
-% in transit vehicles in the late 1980's, 
-% real-time models have been implemented to provide \gls{rti} for commuters.
-% The simplest form of this is, the \gls{eta}
-% (usually displayed as \emph{minutes until arrival}),
-% was implemented in {{region}} by \cite{Wall99analgorithm} 
-% using their proposed \gls{kf} algorithm.
-% In their work, they used historical data to obtain \glspl{eta}.
+Arrival time prediction is only as accurate as the underlying model,
+and while a lot of work has gone into developing public transport vehicle models
+\citep{Cathey_2003,Jeong_2005,Yu_2011,Hans_2015},
+in many public transport systems no formal model is used.
+Instead, ETAs are solely based on the scheduled arrival time
+adjusted by the vehicle's delay at the most recent stop, if available.
+This assumes that the schedule is valid,
+and that there is no unusual congestion along the route,
+neither of which are valid assumptions,
+particularly in our test area of Auckland, New Zealand,
+where infrastructure for buses (such as priority lanes) is limited.
+This behaviour is exaggerated when there is a large distance between stops,
+as the ETA can change drastically.
 
-% Over the years, advances in computing power have enabled new
-% modeling approaches,
-% such as support vector machines \citep{Yu_2006},
-% artificial neural networks \citep{Yu_2011},
-% and more recently particle filters \citep{Hans_2015}.
+To avoid these problems and improve the reliability of ETAs,
+we first needed to develop a model to infer position and speed
+of transit vehicles from \rt GPS data,
+isntead of relying on delays reported only at stops.
+Several modeling approaches were explored, 
+such as the Kalman filter (used by \cite{Dailey_2001} and \cite{Cathey_2003}, ...),
+machine learning models \citep{Yu_2006,Chang_2010},
+and a particle filter approach demonstrated
+by \cite{Hans_2015}.
+We chose to use the last of these, the particle filter,
+as they are very flexible and aren't limited by assumptions
+such as unimodality;
+instead, they are constrained by computational complexity.
 
-Particle filter has proven useful to this kind of dynamic problem,
-where the state distribution is non-gaussian, non-symmetric, 
-multimodal...
+The \emph{reliablility} of arrival time estimates
+alo needs to be considered.
+Ideally, congestion along intermediate roads should be incorporated into the predictive model,
+and one way of doing this is to use the travel time of preceeding buses along the same roads.
+Several such approaches have been proposed.
+\cite{Yu_2010} used the travel time of vehilces on the same \emph{routes},
+which they showed to improve ETAs.
+However, this only works on routes with a high frequency.
+\citep{Yu_2011} extended this by looking at roads where several major rouets converged,
+again showing improvements but limited in scope.
+In general, however, no comprehensive network modeling approach has been proposed that used
+solely GPS position data to estimate arrival times
+while accounting for congestion.
 
-Any other recent attempts at network-estimation?
-There was a KF using point-to-point speed estimations?
-
-Computational aspect - historically, models had to be simple 
-(e.g., KF) so they could run in realtime.
-These days, running a somewhat powerful VM on a remote server
-is almost standard practice, so we can take full advantage 
-and use a more computationally demanding (particle filter)
-but more accurate estimation technique - paired with a more complex model.
-
-We first give an overview of the transit network,
-and how we construct it from raw GTFS shape data.
-This paper assumes a very basic model,
-with a focus on the implementation in a realtime setting.
+In this paper, we describe a \rt vehicle and network state
+model which uses the vehicles traveling throughout the transit road network
+to update congestion information in \rt.
+The vehicle model is implemented using a particle filter,
+so it is also necessary to demonstrate the faesibility of this as a valid real-time option.
+From the particle filter estimates of vehicle trajectories,  
+we estimate the travel time along roads update the road network's state.
+Future work will involve using this travel time information 
+in combination with historical data to develop an arrival time prediction model.

### 2018-10-28 anzjs2018/sections/02_gtfs.tex
Edits

@@
-\section{Working with GTFS data}
+\section{Working with \rt transit data}
 \label{sec:gtfs}
@@
 is an API (application programming interface) specification for transit data
-detailing it should be organised,
+detailing how it should be organised,
 making access easier for application developers.
@@
 (source \url{http://transitfeeds.com}).
-An advantage of this for us is that,
-provided our application depends solely on GTFS data,
+An advantage of this standardised format is that,
+provided an application depends solely on GTFS data,
 after developing it locally in Auckland it can be deployed to any other GTFS-based
@@
 There are two components to GTFS.
-The first, \emph{GTFS static}, includes information about stops, routes, trips, and shapes:
-a \emph{stop} is a physical location where passengers can embark and disembark the vehicle;
-a \emph{route} is a sequence of two or more stops displayed as a single service;
-a \emph{trip} is an instance of a route occuring at a specific time of day; and
-a \emph{shape} is the sequence of points defining a vehicle's path along a route
+The first, \emph{GTFS static}, includes information about
+\begin{itemize}
+\item \emph{stops}, a physical location where passengers can embark and disembark the vehicle;
+\item \emph{routes}, a sequence of two or more stops displayed as a single service;
+\item \emph{trips}, an instance of a route occuring at a specific time of day;
+\item \emph{schedules}, specifying the arrival (and departure) times for each bus at each of its stops; 
+\item \emph{shapes}, the sequence of points defining a vehicle's path along a route
+\end{itemize}
 \citep{GoogleDevelopers_2006}.
@@
 which is only available in a subset of the providers due to the requirement of 
-onboard GPS and a central server.
+onboard GPS tracking devices and a central server.
 It provides a standardised format for sharing vehicle positions and trip delays,
-which are often accessed via an API which developers can use in \rt applications.
+and are typically accessed by developers via an API can be used in \rt applications.
 
-Before we can develop our application,
-we need to modify the raw GTFS information to make it possible to detect common roads.
+As mentioned in Section~\ref{sec:intro},
+there are some major issues with the current ETA generation method.
+These are almost directly attributed to GTFS-realtime trip updates,
+which are currently the sole source of data for ETAs (in Auckland).
+Trip updates are reported whenever a vehicle arrives or departs a stop,
+and includes most importantly the delay between the scheduled and actual arrival times.
+This delay is then propagated on to all future stops to adjust the ETA.
+This assumes that the schedule is well calibrated and the time between scheduled arrivals
+is representative of the real-world travel time between stops. 
+Even more problematic is that, in the absense of trip updates,
+the delay defaults to zero,
+and unless the trip is manually cancelled,
+appears on-time to passengers waiting at the stop.
+If the bus is late or cancelled,
+the trip disappears from the \rt board and passengers
+are unable to distingush the two scenarios.
 
@@
 
-Arguably one of the most important predictors of arrival time is
-the travel time along intermediate roads,
-however in most applications this vital information is unavailable,
-at least directly. 
-Several predictive approaches using \emph{headway},
-the time between consecutive trips along the same route,
-have been proposed \citep{Hans_2015};
-however, while this is reasonable for high frequency routes,
-low frequncy routes will be unable to react quickly to changes in congestion,
-and it is for these routes which reliable ETAs are arguably more important,
-since the cost of missing a bus is greater.
+Arguably the most important predictor of arrival time is
+the cumulative travel time along intermediate roads.
+In most applications, however, this vital information is unavailable,
+at least directly,
+so here we attempt to construct a transit road network of intersections
+and the roads between them, 
+and map each route to a sequence of road segments.
 
-One solution would be to use information obtained from
-vehicles servicing other routes but traveling along the same roads
-to estimate arrival times.
-The simplest method of identifying overlapping routes is to compare
-the stop sequence: 
-routes with a common sub-sequence of stops must be traveling the same route between them.
-While there are several exceptions, for example express routes 
+The simplest way to construct such a network is to use the stop sequence:
+routes with a common sub-sequence of stops must be traveling the same route between them,
+so by defining a road as the path between two consecutive stops 
+we can obtain travel time estimates along that road segment.
+There are several places where this doesn't quite work, for example express routes 
 and multi-stop locations where there are high numbers of routes,
+resulting in some overlap between road segments,
 it is a viable enough simplication for the purposes of this paper.
@@
 along a road segment.
+Future work will look at using physical intersections as nodes in the network to improve it.
 
@@
 or intersection between them.
+It is also possible for a bus to remain stationary,
+so the possible vehicle positions rapidly increases with the time between observations.
 
@@
 bus stops and some major intersections.
-Often these positions appear to be preemptive 
+Often these updates are preemptive 
 (i.e., the bus is almost there, but not quite),
@@
 of the vehicle by finding the nearest point on the path to the observation;
-if this has decreased, the previous observation is rejected and the vehicle reverted
+if this has decreased, the current state is rejected and the vehicle reverted
 to its previous state.

### 2018-10-28 anzjs2018/sections/03_models.tex
Edits

@@
 real-time tracking applications.
-We are using two distinct models for the vehicle and network states,
+We are using a separate model for each of the vehicle and network states,
 and in each assume an underlying Markov process with state $\mathcal{X}_k$
@@
 with error $\nu_k\sim\mathrm{N}(0,\mathcal{R})$,
-giving the following general model below.
+giving the following general model:
 \begin{equation}
@@
 \end{equation}
-The transition function $f$ determines the relationship between consecutive states,
+The transition function $f$ describes the relationship between consecutive states
+(i.e., what we want to model),
 while the measurement function $h$ is a deterministic function describing the
@@
 The following sections describe the two models used in this application.
-The first is implemented using a particle filter to estimate vehicle states,
-with the primary objective of infering vehicle travel times along roads
-from a sequence of GPS positions,
-while the second uses these travel times to update road states 
+The first, implemented using a particle filter, estimates vehicle states
+to infer travel times along roads from a sequence of GPS positions;
+the second uses these travel times to update road states 
 and is implemented using an \emph{information filter},
-which is a variant of the Kalman filter \citep{}.
+a variant of the Kalman filter \citep{}.
 
@@
 $\bz = (z_1,\ldots,z_L)^\top$, in seconds,
-are independent of time but included in the state so they can be estimated
+are included in the state and estimated
 sequentially as the vehicle traverses the route.
@@
 transit vehicle modeling applications \citep{Hans_2015}.
-The primary advante of a particle filter is its handling of multimodality,
-as in figure~\ref{fig:pf_state_predict},
+The primary advantage of a particle filter is its handling of multimodality,
+as demonstrated in figure~\ref{fig:pf_state_predict},
 which is a common feature of the proposal distribution, particularly around bus stops.
-A second advantage is the intuitive likelihood function (see section~\ref{sec:pf_update}).
+Another advantage is the intuitive likelihood function (see section~\ref{sec:pf_update}).
 Conversely, particle filter methods are computationally demanding,
 requiring an increasing number of particles as the model complexity and
-number of parameters in increase \citep{}.
+number of parameters increases \citep{}.
 Section~\ref{sec:rt} describes our implementation of the particle filter in real-time 
-and it's timings.
+and its timings.
 
@@
 
-In a particle filter, the posterior distribution of the state at time $t_{k-1}$,
-is represented by a set of discrete points, or particles, (fig~\ref{fig:pf_state_prev}),
+In a particle filter, the posterior distribution of the state at time $t_{k-1}$
+is represented by a set of discrete points, or particles, (Figure~\ref{fig:pf_state_prev}),
 each with an associated weight $W_{k-1}^{(i)}$
@@
 p(\bX_{k-1} | \bY_{k-1}) \approx \tilde\bX_{k-1|k-1} 
-:= \{\bX_{k-1}^{(i)}, W_{k-1}^{(i)}\}_{i=1}^N
+:= \left\{\bX_{k-1}^{(i)}, W_{k-1}^{(i)}\right\}_{i=1}^N
 \end{equation}
@@
 After mutation, 
-the particles are reweighted using the likelihood $p(\bY_k | \bX_k^{(i)})$ 
-(sec.~\ref{sec:pf_update}) and standardised,
+the state is updated by reweighting the particles using the likelihood $p(\bY_k | \bX_k^{(i)})$ 
+(sec.~\ref{sec:pf_update}) and standardising so $\sum_{i=1}^N W_k^{(i)} = 1$,
 \begin{equation*}
@@
 p(\bX_k | \bY_k) \approx \tilde\bX_{k|k} := 
-\{\bX_{k}^{(i)}, W_{k}^{(i)}\}_{i=1}^N.
+\left\{\bX_{k}^{(i)}, W_{k}^{(i)}\right\}_{i=1}^N.
 \end{equation}
 
-In order to prevvent degeneration of the particle sample,
+In order to prevent degeneration of the particle filter state,
 \emph{importance resampling} is employed whenever the 
@@
 Afterwards, the weights are reinitialized to $W_k^{(i)} = N^{-1}$.
+Note this is a computationally demanding step,
+so reducing the rate of resampling is favourable.
 
@@
 
-Using a particle filter allows us to develop a vehicle model in which
-the transition function $f$ implements several key bus behaviours:
+The particle filter lets us to develop a model in which
+the vehicle transition function $f$ implements several key bus behaviours:
 \begin{itemize}
@@
 \item optional stopping and waiting at bus stops while passengers board and disembark, and
-\item optional stopping and waiting at intersections/traffic lights (future work).
+\item optional stopping and waiting at intersections or traffic lights (future work).
 \end{itemize}
@@
 using system noise parameter $\sigma^2$ which describes 
-how the vehicle's acceleration changes as a random process.
+how the vehicle's acceleration changes as a Markov process.
 The vehicle does not travel constantly along the route,
-however, as it needs to service bus stops along the way.
-Therefore, the transition function includes 
+however, as it needs to service bus stops along the way,
+so the transition function includes 
 stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
@@
 and the minimum dwell time at stops, $\gamma$,
-for the doors to open and close.
+for the doors to open and close,
+and account for deceleration and acceleration.
 For these parameters, we used constant values for all stops,
-and based the values on those used by \cite{Hans_2015};
+based on the values used by \cite{Hans_2015};
 future work will investigate modeling these in real-time similarly to Section~\ref{sec:kf}.
-No intersection model is implemented here,
-as the segment road segments are between stops,
-and the travel times predicted by the network model at time $t_k$,
-$\btheta(t_k) = (\theta_1(t_k), \ldots, \theta_L(t_k))^\top$
-(see section~\ref{sec:kf}).
+% No intersection model is implemented here,
+% as the segment road segments are between stops,
+% and the travel times predicted by the network model at time $t_k$,
+% $\btheta(t_k) = (\theta_1(t_k), \ldots, \theta_L(t_k))^\top$
+% (see section~\ref{sec:kf}).
 The transition function algorithm is shown in Figure~\ref{fig:algorithm},
@@
 After mutating the particle set, the posterior distribution $\tilde\bX_{k|k}$ 
-is updated by reweighting each particle using the likelihood.
+is obtained by reweighting each particle using the likelihood.
 The measurement function $h$,
@@
 as shown in figure~\ref{fig:pf_lhood}.
-The model for the observation generation is,
+The model for the observations is,
 assuming GPS error $\epsilon^2$,
@@
 The likelihood of the data given a particle's state estimate 
-is therefore easy to calculate using (\ref{eq:obs_exp})
+is therefore calculated using (\ref{eq:obs_exp})
 \begin{equation}
@@
 the likelihood cannot be calculated using the distance as the estimate
-is a distribution, and so instead a reverse non-deterministic transformation,
+is a distribution, and instead a reverse non-deterministic transformation,
 the inverse measurement function,
@@
 
-The purpose of the road network model is to provide a way of using
-spatial information about vehicle speeds to inform arrival time predictions.
-The simplest notion would be to map each vehicle speed estimate to a physical road;
-however, this has two problems
-\begin{enumerate}
-\item roads usually have two directions, so vehicle directionality would need to be computed,
-\item road data is not provided in GTFS, and can be difficult to obtain cleanly without
-    manual intervention.
-\end{enumerate}
-The most desirable solution would be to use the GTFS shape data to construct a network
-by merging overlapping routes and using merge points as intersections
-to define road segments.
-Initial efforts at applying this had poor results due to overly long or short segments,
-so for this current work we are approximating the road network by using the stops 
-along a route as intersection, and the legs between them as road segments,
-as shown in Figure~\ref{fig:network_creation}.
-
-To obtain reliable ETAs,
-road travel times need to be estimated in \rt,
-as this allows the model to react to congestion changes as detected by preceeding buses.
-Future work will aim to construct a more complex model incorporating historical trends 
-and correlations between roads, 
-however for the current work we are demonstrating our approach
-using a simple model,
-in which each road segment is treated independently.
-
-The model assums a Gaussian state distribution for travel time,
-truncated necessarily at zero;
-this will not be an issue as the state prediction and data will always be non-negative.
-However, the model must be able to handle multiple observations in a single update,
-which often occurs along major bus routes when several buses are traveling together.
-Therefore we have chosen the information filter as our estimation method.
-
-The information filter is a transormation of the Kalman filter in which the
-\emph{information matrix} and \emph{information vector} are used in place of 
-the covariance matrix and state vector, respectively.
-The advantage is that multiple observations can be added together when updating the state.
-Otherwise the information filter follows the same procedure as other recursive Bayesian models.
+The primary objective of the network model is to estimate the \rt traffic conditions
+(travel time) along roads in the transit network, 
+as well as make short-term predictions for estimating arrival times.
+In this paper,  we apply a very simple model in which each road is treated independently,
+future work will investigate using correlations with adjacend roads and historical data
+to improve the model and subsequent predictions.
 
 The network state $\boldsymbol\theta_c = \{\theta_c^\ell\}_{\ell = 1}^L$ is the travel time 
-of transit vehicles along road segment $\ell$ at time $t_c$.
-Observations of the state, $Z_c^\ell$, are obtained from the particle filter 
-in Section~\ref{sec:pf}.
-Due to the simplicity of the model,
-no transition matrix or measurement matrix is required,
-and the model from (\ref{eq:rbe_model}) reduces simply to
+of transit vehicles along road segment $\ell$ at time $t_c$,
+and is assumed to be constant with system noise $v_c^\ell \sim \mathrm{N}(0, \nu^2)$.
+Observations of the state, $Z_c^\ell$, are obtained from the particle filter in Section~\ref{sec:pf},
+and have measurement error $e_c^\ell \sim \mathrm{N}(0, R_c^\ell)$ as estimated 
+using uncertainty of particle estimates.
+The model from (\ref{eq:rbe_model}) now reduces to
 \begin{equation}
@@
 \end{equation}
-where $\Delta_c = t_c - t_{c-1}$,
-$v_c^\ell \sim \mathrm{N}(0, \nu^2)$ is the system noise,
-and $e_c^\ell \sim \mathrm{N}(0, R_c^\ell)$ is the measurement error
-as estimated from the vehicle model using uncertainty of particle estimates.
+where $\Delta_c = t_c - t_{c-1}$.
+
+Travel times are assumed to be non-negative but otherwise normally distributed,
+and constant over time---we do not attempt to model peak congestion in this paper.
+Additionally the observations are non-constact and may be multiple as several buses travel
+a road simultaneously.
+The information filter is a transformation of the Kalman filter in which the
+\emph{information matrix} and \emph{information vector} are used in place of 
+the covariance matrix and state vector, respectively.
+This allows multiple observations to be added together to update the state.
 
@@
 For the update step, the parameters are transformed into an information
-space parameterised by the information matrix $U^\ell_c$
-and the information vector $u^\ell_c$,
-\begin{equation}
-\label{eq:information_transform}
-U^\ell_{c|c-1} = \frac{1}{P_{c|c-1}}
-\quad\text{and}\quad
-\hat u^\ell_{c|c-1} = \frac{\hat \theta^\ell_{c|c-1}}{P^\ell_{c|c-1}}
-\end{equation}
-The measurement data obtained from the vehicle model are
-the travel time of vehicle $m$ along segment $\ell$,
-$\bar z_c^{m\ell}$, and the uncertainty $s^{m\ell}_c$.
-These are transformed to a measurement information covariance matrix $B_c^{m\ell}$
-and measurement information vector $b_c^{m\ell}$,
-\begin{equation*}
-B^{m\ell}_c = \frac{1}{(s^{m\ell}_c)^{2}}\quad\text{and}\quad
-b^{m\ell}_c = \frac{\bar z^{m\ell}_c}{(s^{m\ell}_c)^2}
-\end{equation*}
+space parameterised by the information matrix $U^\ell_c = (P_{c|c-1})^{-1}$
+and the information vector $u^\ell_c~=~\hat \theta^\ell_{c|c-1} (P^\ell_{c|c-1})^{-1}$.
+The travel time estimates from the vehicle model for vehicle $m$ along segment $\ell$
+is $\bar z_c^{m\ell}$, with uncertainty $s^{m\ell}_c$,
+can be transformed to a measurement information covariance matrix 
+$B_c^{m\ell} = (s^{m\ell}_c)^{-2}$
+and measurement information vector $b_c^{m\ell}~=~\bar z^{m\ell}_c (s^{m\ell}_c)^{-2}$.
 The information update is now the sum of the information over all $M$ vehicles
-that traversed the segment since the last update, so
+that traversed segment $\ell$ since the last update, so
 \begin{align*}
@@
 \end{align*}
-which are back transformed
-using the inverse of (\ref{eq:information_transform})
-to obtain the desired parameter estimates
+which are back transformed to obtain the desired parameter estimates
 \begin{equation*}

### 2018-10-28 anzjs2018/sections/04_implementation.tex
pictures
Edits

@@
 
-There are two components to the application:
-the static GTFS~schedule and network data,
-and the real-time modeling and prediction.
-We chose \verb+Rcpp+ to develop our program,
-giving us the advantages of an R~package for data manipulation 
-(notably \verb+RSQLite+ and \verb+dplyr+)
-and distribution, 
+The application consists of a
+static GTFS~schedule and network data component,
+and a \rt modeling and prediction component.
+We chose \verb+Rcpp+ to develop the program,
+providing access to other R packages for data manipulation 
+(notably \verb+RSQLite+ and \verb+dplyr+),
 as well as the speed and memory management capabilities of \verb|C++|. 
-As a result, all of the following has been implemented in the R package
+The following has been implemented in the R package
 \verb+transitr+, available on Github (\url{https://github.com/tmelliott/transitr}).
-Here our focus is on the real-time aplication, so we are only discussing the real-time 
-features of model's implementation.
+In this section, we discuss the features of the \rt component.
 
-The general structure of the main function is:
+The general structure of the \rt component is:
 \begin{enumerate}
@@
     \item Collect travel time information for each vehicle for any completed roads
-    \item Update road network with any completed road segments
-    \item Generate ETAs for vehicles
+    \item Update road network for any completed road segments
+    \item Generate ETAs for vehicles uing combination of particle filter and network state
     \item Write ETAs to extended Google Protobuf binary file for distribution
@@
 
-The main concern is speed: 
+The primary concern is speed: 
 we desire ETAs to be available as soon as possible after obtaining the data.
@@
 
+\begin{figure}[tb]
+    \centering
+    \includegraphics[width=0.7\textwidth]{figures/04_model_results_dist.pdf}
+    \caption{The timings for various parts of the application, and overall. %
+        The trend is approximately linear with the number of particles.}
+    \label{fig:dist_to_route}
+\end{figure}
+
+\begin{figure}[tb]
+    \centering
+    \includegraphics[width=0.7\textwidth]{figures/04_model_results_neff.pdf}
+    \caption{The timings for various parts of the application, and overall. %
+        The trend is approximately linear with the number of particles.}
+    \label{fig:neff}
+\end{figure}
+
 From the Figure~X, 
@@
 and only slightly effects the values.
-
-So the main way to determine how well everything is working is to look 
-at the RMSE of ETAs over time.
 

### 2018-10-29 anzjs2018/sections/04_implementation.tex
cpu timings also

@@
 
-The primary concern is speed: 
-we desire ETAs to be available as soon as possible after obtaining the data.
-To do this, we use raw pointers to read-only GTFS objects,
-and particles are only copied when resampling is required.
-This is also why we chose to use protobuf as the output format,
-since this is fast to write and distribute.
-Steps 2b--e are performed in parallel, 
-which in our test environment of a virtual machine with 8~cores, 
-significantly speeds up computation.
+During the development of the application,
+we were primarly concerned with ensuring each component of the program
+is as efficient as possible,
+allowing ETAs to be generated and distributed fast enough---our initial goal was 30~seconds.
+Using \verb|C++| provides enough control over memory management
+to ensure the program runs fast enough in \rt 
+while still implementing the models.
+We could also use OpenMP for parallelisation,
+allowing us to process multiple vehicles simultaneously
+depending on the number of cores available.
 
-As far as results, we have two categories:
-implementation and faesibility,
-and model performance;
-we do not have any complex arrival time estimation method,
-so no results for that are shown here.
-
-The simulations we discuss here use a subset of real data from 8~October, 2018.
-This allows us to run the model with different parameters over the same data.
-The parameters varied were the number of particles and system noise $\sigma^2$.
-The primary goal is to reduce the rate of degeneration and ``losing'' the bus 
-(i.e., no particles are near the observed location),
-and reducing the variance of parameter estimates; here we use vehicle speed as the example.
-
-The one parameter we can determine from the data is GPS error, $\epsilon^2$,
-by looking at the distance between observations and the shape.
-Figure~\ref{fig:gps_dist} shows the distribution of distance between the vehicle
-observation and the \emph{nearest point on the path}.
-Due to the way the observations are reported, 
-many observations will be ``exactly'' on the route at a bus stop,
-and we need to ignore these.
+The particle filter requires thousands of particles per vehicle,
+so it is necessary to explore the performance of the application
+with varying number of particles.
+It is also necessary to assess the performance of the particle filter,
+and compare values for the fixed model parameters.
+To do so, we implemented a simulated \rt environment
+in which we could analyse the same subset of real vehicle data from 8~October, 2018
+using a range of settings.
+These simulations were carried out of a Virtual Machine 
+with 8~cores and 32~Gb of memory, 
+running Ubuntu~16.04 and using R~3.4.1.
 
@@
 Since the number of vehicles traveling at any given time changes throughout the day,
-we used the average timings from 13:45--14:00 to compare.
+we used the average timings over a 15~minute window to compare.
 
 Figure~\ref{fig:timings} shows the timing results for varying number of particles.
+As expected,
+
 Nothing too exciting going on here, 
@@
 \label{sec:model_perf}
+
+The parameters varied were the number of particles,
+GPS error $\epsilon$, and system noise $\sigma^2$.
+The primary goal is to reduce the rate of degeneration and ``losing'' the bus 
+(i.e., no particles are near the observed location),
+and reducing the variance of parameter estimates; here we use vehicle speed as the example.
+
+The one parameter we can determine from the data is GPS error, $\epsilon^2$,
+by looking at the distance between observations and the shape.
+Figure~\ref{fig:gps_dist} shows the distribution of distance between the vehicle
+observation and the \emph{nearest point on the path}.
+Due to the way the observations are reported, 
+many observations will be ``exactly'' on the route at a bus stop,
+and we need to ignore these.
 

### 2018-10-29 anzjs2018/figures/04_model_results.R

@@ -44,13 +44,22 @@ tdat <- bind_rows(times, ttot) %>% filter(what %in% lvls) %>%
 
 ## Wall timings
 pdf('figures/04_model_results_timing.pdf', width = 6, height = 3)
+gridExtra::grid.arrange(
 ggplot(tdat %>% summarize(cpu.hat = mean(cpu), cpu.var = sd(cpu))) +
     geom_pointrange(aes(n_particles, cpu.hat,
         ymin = cpu.hat - cpu.var, ymax = cpu.hat + cpu.var, 
         shape = what)) +
     xlab("Number of particles") +
     ylab("Wall Time (seconds)") +
-    labs(shape = "")
+    labs(shape = ""),
+ggplot(tdat %>% summarize(wall.hat = mean(wall), wall.var = sd(wall))) +
+    geom_pointrange(aes(n_particles, wall.hat,
+        ymin = wall.hat - cpu.var, ymax = wall.hat + wall.var, 
+        shape = what)) +
+    xlab("Number of particles") +
+    ylab("CPU Time (seconds)") +
+    labs(shape = ""),
+    nrow = 1, widths = c(3, 2))
 dev.off()
 
 # ## CPU timings

### 2018-10-29 anzjs2018/figures/04_model_results.R
cpu timings also

@@ -44,13 +44,22 @@ tdat <- bind_rows(times, ttot) %>% filter(what %in% lvls) %>%
 
 ## Wall timings
 pdf('figures/04_model_results_timing.pdf', width = 6, height = 3)
+gridExtra::grid.arrange(
 ggplot(tdat %>% summarize(cpu.hat = mean(cpu), cpu.var = sd(cpu))) +
     geom_pointrange(aes(n_particles, cpu.hat,
         ymin = cpu.hat - cpu.var, ymax = cpu.hat + cpu.var, 
         shape = what)) +
     xlab("Number of particles") +
     ylab("Wall Time (seconds)") +
-    labs(shape = "")
+    labs(shape = ""),
+ggplot(tdat %>% summarize(wall.hat = mean(wall), wall.var = sd(wall))) +
+    geom_pointrange(aes(n_particles, wall.hat,
+        ymin = wall.hat - wall.var, ymax = wall.hat + wall.var, 
+        shape = what)) +
+    xlab("Number of particles") +
+    ylab("CPU Time (seconds)") +
+    labs(shape = ""),
+    nrow = 1, widths = c(3, 2))
 dev.off()
 
 # ## CPU timings

### 2018-10-29 anzjs2018/figures/04_model_results.R
plot code

@@ -36,42 +36,42 @@ tsmry <- times %>% bind_rows(ttot) %>%
 #     geom_point(aes(n_particles, cpu, color = what)) +
 #     ylab("Wall time (seconds)")
 
-lvls <- c("predicting ETAs", "updating vehicle states", "writing ETAs to protobuf feed", "all")
+lvls <- c("updating vehicle states", "predicting ETAs", "writing ETAs to protobuf feed", "all")
 tdat <- bind_rows(times, ttot) %>% filter(what %in% lvls) %>%
     mutate(what = factor(what, levels = lvls)) %>%
     group_by(what, n_particles)
 
 
-## Wall timings
+# Wall timings
 pdf('figures/04_model_results_timing.pdf', width = 6, height = 3)
-gridExtra::grid.arrange(
-ggplot(tdat %>% summarize(cpu.hat = mean(cpu), cpu.var = sd(cpu))) +
-    geom_pointrange(aes(n_particles, cpu.hat,
-        ymin = cpu.hat - cpu.var, ymax = cpu.hat + cpu.var, 
-        shape = what)) +
-    xlab("Number of particles") +
-    ylab("Wall Time (seconds)") +
-    labs(shape = ""),
-ggplot(tdat %>% summarize(wall.hat = mean(wall), wall.var = sd(wall))) +
+# gridExtra::grid.arrange(
+# ggplot(tdat %>% summarize(cpu.hat = mean(cpu), cpu.var = sd(cpu))) +
+#     geom_pointrange(aes(n_particles, cpu.hat,
+#         ymin = cpu.hat - cpu.var, ymax = cpu.hat + cpu.var, 
+#         shape = what)) +
+#     theme(legend.position = "none") +
+#     xlab("Number of particles") +
+#     ylab("Wall Time (seconds)") +
+#     labs(shape = ""),
+# ggplot(tdat %>% summarize(wall.hat = mean(wall), wall.var = sd(wall))) +
+#     geom_pointrange(aes(n_particles, wall.hat,
+#         ymin = wall.hat - wall.var, ymax = wall.hat + wall.var, 
+#         shape = what)) +
+#     xlab("Number of particles") +
+#     ylab("CPU Time (seconds)") +
+#     labs(shape = ""),
+#     nrow = 1, widths = c(3, 4))
+# dev.off()
+
+ggplot(tsmry %>% group_by(what, n_particles) %>% 
+        summarize(wall.hat = mean(wall/1000), wall.var = sd(wall/1000))) +
[…]

### 2018-10-29 anzjs2018/sections/04_implementation.tex
words

@@
 Figure~\ref{fig:timings} shows the timing results for varying number of particles.
-As expected,
-
-Nothing too exciting going on here, 
-except that writing ETAs takes up the largest chunk of time,
-as this involves calculating quantiles which require the estimated ETAs to be sorted,
-which is computationally demanding. 
-The vehicle update takes the second longest time,
-and involves copying particles during the update resampling step.
-Finally, predicting the ETAs is pretty quick.
-The remaining parts of the application take negligible time,
-and all run on a single core to avoid mutex locks.
+We see that there is a very slight quadratic trend to run times,
+as several steps are not of complexity $O(N)$,
+the main one being the estimation of ETA quantiles.
+While not discussed in this paper,
+the application generates ETAs for each particle
+and calculates ETA quantiles for each vehicle to demonstrate the computational faesibility.
+This step requires sorting a vector, which is $O(N\log(N))$.
+Clearly, our initial target of 30~seconds is obtainable given the number of particles
+is not too large, and there are enough cores.
 
@@
     \caption{The timings for various parts of the application, and overall. %
-        The trend is approximately linear with the number of particles.}
+        The trend is slightly non-linear due to the complexity of the ETA generation step, %
+        which requires sorting of the particle ETAs to obtain quantile estimates.}
     \label{fig:timings}

### 2018-10-29 anzjs2018/sections/04_implementation.tex
more writing and pictures

@@
 
-The parameters varied were the number of particles,
+The model was evaluated using several statistics,
+each calculated using a range of parameter values.
+The first is effective sample size,
+defined in (\ref{eq:neff}),
+which is used to determined if resampling is required.
+Another important statistic is the \emph{degeneration rate},
+or the percent of iterations where the particle filter
+is unable to find any plausible states given the observation,
+and the vehicle is reinitialised.
+Note, however, that the lower limit of this value 
+is attributed to \emph{invalid trips},
+however there is no simple way of filtering these trips.
+The last statistic we compute is the variance of road segment travel times
+over time.
+For this example, we varied the number of particles $N$,
 GPS error $\epsilon$, and system noise $\sigma^2$.
-The primary goal is to reduce the rate of degeneration and ``losing'' the bus 
-(i.e., no particles are near the observed location),
-and reducing the variance of parameter estimates; here we use vehicle speed as the example.
 
@@
 by looking at the distance between observations and the shape.
-Figure~\ref{fig:gps_dist} shows the distribution of distance between the vehicle
+Figure~\ref{fig:dist_to_route} shows the distribution of distance between the vehicle
 observation and the \emph{nearest point on the path}.
@@
 many observations will be ``exactly'' on the route at a bus stop,
-and we need to ignore these.
-
-To check the model results, 
-we need to realise that we are not trying to accurately predict
-future locations of the vehicle,
-but instead are trying to use a sequence of GPS positions to infer travel
-times along intermediate roads.
-Therefore, several tools used to assess the validity of a particle filter
-are unnecessary.
-Instead, we will be comparing the rate of degeneration 
-(i.e., the particle sample loses the actual bus and needs to be reinitialized)
-and the mean distance between the state estimate and the observation; 
-we will use the distance between the observation and the route (cross track distance) 
-to determine how well the posterior distribution fits the data---%
-if the ratio between the CTD and RMSE is large,
-then the fit is ``bad''.
-[We also explore the uncertainty of travel time estimates,
-as well as the variability between consecutive buses]\footnote{assuming we can get the results in time}.
+which explains the spike at just above zero~meters.
 
@@
     \includegraphics[width=0.7\textwidth]{figures/04_model_results_dist.pdf}
-    \caption{The timings for various parts of the application, and overall. %
-        The trend is approximately linear with the number of particles.}
+    \caption{The distance between the observation and the nearest point on the shape.}
     \label{fig:dist_to_route}
@@
 
+For the effective sample size, 
+increasing GPS error and system error both tend to
+increase $N_\text{eff}$, as shown in Figure~\ref{fig:degen_rate}.
+Increasing the GPS error distributes the weights more evenly,
+while increasing system noise allows more particles to reach
+the necessary speed,
+so these are as expected.
+
 \begin{figure}[tb]
     \centering
-    \includegraphics[width=0.7\textwidth]{figures/04_model_results_neff.pdf}
-    \caption{The timings for various parts of the application, and overall. %
-        The trend is approximately linear with the number of particles.}
-    \label{fig:neff}
+    \includegraphics[width=0.9\textwidth]{figures/04_model_results_degen.pdf}
+    \caption{The effective sample size and degeneration rate for a range of parameter 
+        values. The columns are of increasing number of particles,
+        and the rows are for increasing (downwards) GPS error.}
+    \label{fig:degen_rate}
 \end{figure}
 
-From the Figure~X, 
-we see a trade-off between some of the parameters.
-Larger values of GPS error decrease the degeneration rate
-as well as the rate of resampling (which increases speed);
-however, the cost of this is that the mean distance between the posterior distribution
-and the observed location increases, giving a less precise estimate of the vehicle's state.
-Conversely, the system noise does not have the same effect,
-and only slightly effects the values.
+The degeneration rate, 
+shown on the $y$-axis of Figure~\ref{fig:degen_rate},
+decreases as GPS error increases,
+as anomolous observations where the bus is far from the route
+are more likely.
+It also decreases with the number of particles,
+while system noise has no apparent effect.
 
-% How we implement it, choice of software (Rcpp = R + C++).
-% R: dealing with data structures is easier, maintainability, interfacing
-% C++: speed
+Lastly is the variability of travel times and their uncertainties.
+These are shown in Figure~\ref{fig:travel_times}.
+System noise again shows no effect on the variability of travel time estimates,
+however increasing GPS error tends to increase travel time variability.
+This means that there is a trade-off between the models performance
+and how well it estimates the main parameter of interest:
+larger values of $\epsilon$ result in less degeneration and larger $N_\text{eff}$,
+at the cost of more variable travel time estimates.
 
-% Overall structure:
-% - load
-% - fetch positions
-% - initialize or mutate+update
-% - update network
-% - make ETA predictions (vehicle state + network state)
-
-% Some of the key things:
-% - minimise copy, parallelisation using OMP
-% - moving as much computation ``outside'' of the main loop as possible
-%   (e.g., ``pre''-predict vehicle/network states so only update required)
-%   so ETAs are generated ASAP after retrieving data
-% - keeping the GTFS database up-to-date by fetching new data each morning
-% - distribution - a cloud database vs a single protobuf file with everything 
-%   (maintenance/reliability/speed/size)
+\begin{figure}[tb]
+    \centering
+    \includegraphics[width=0.9\textwidth]{figures/04_model_results_times.pdf}
+    \caption{Travel times.}
+    \label{fig:travel_times}
+\end{figure}

### 2018-10-29 anzjs2018/sections/06_discuss.tex
more writing and pictures

@@
 
-It all seems to work (hopefully I don't need to change this D:)
+The main goal of this work was to explore the faesibility
+of a particle filter as a \rt estimation method for transit vehicle states,
+and the use of estimated states to determine traffic conditions
+throughout the network.
+This allows vehicles to inform others---irrespective of route---of 
+how long it will take to reach future stops.
 
-Next steps include
-- generating useful priors for ``fall back'' (i.e., no data, future) prediction
-- develop and implement a more complex network update model (correlations, etc)
-- prediction estimates (point vs interval)
+Not only is the particle filter a viable choice in terms of speed,
+it has shown it can handle most situations without losing the vehicle,
+with the exception of when the vehicle is not traveling along
+the specified route.
+Hopefully with future development,
+the stopping behaviour can be more accurately modeled,
+perhaps incorporating trip updates into the likelihood function.
+
+The network model is still in the early stages of development,
+but given that the implementation is running quickly---twice as fast as our target---%
+there is room for developing a more complex model.
+Future work will explore the use of historical data to develop informative priors
+for travel times,
+which will provide a useful fallback in the case of no data,
+or for making more accurate long-term predictions.
+Once such a model has been developed, we plan to explore prediction estimates,
+particuarly comparing point and intervals as a means of communicating 
+uncertainty to commuters.

### 2018-10-29 anzjs2018/figures/04_model_results.R

@@ -106,7 +106,9 @@ res <- do.call(bind_rows, lapply(list.files("../../transitr/simulations", patter
 ggplot(res %>% filter(dist_to_path < 5), aes(dist_to_path)) + geom_density()
 
 pdf('figures/04_model_results_dist.pdf', width = 6, height = 3)
-ggplot(res %>% filter(dist_to_path < 20 & dist_to_path > 0), aes(dist_to_path)) + geom_density()
+ggplot(res %>% filter(dist_to_path < 20 & dist_to_path > 0), aes(dist_to_path)) + 
+    geom_density() +
+    xlab("Distance to path (m)") + ylab("Density")
 dev.off()
 
 sims <- res %>% filter(dist_to_path < 20)
@@ -118,4 +120,70 @@ ggplot(sims %>% filter(Neff <= n_particles & Neff > 0 & bad_sample == 0) %>%
     geom_point(aes(x = factor(system_noise), Neff / n_particles * 100, shape = factor(n_particles), colour = factor(gps_error))) +
     labs(x = "System Noise", y = "Effective Sample Size (% of Number of Particles)",
          shape = "Number of Particles", colour = "GPS Error (m)")
+dev.off()
+
+
+
+
+pdf('figures/04_model_results_degen.pdf', width = 10, height = 5)
+ggplot(sims %>% group_by(n_particles, gps_error, system_noise) %>%
+        summarize(p_bad = mean(bad_sample), p_good = mean(Neff < 1000),
+            Neff = mean(Neff))) + 
+    geom_point(aes(100 * Neff / n_particles, 100*p_bad, shape = as.factor(system_noise))) +
+    facet_grid(gps_error ~ n_particles) +
+    xlab("Effective Sample Size (% of N)") + ylab("Degeneration Rate (% of Iterations)") +
+    labs(shape = "System noise")
+# ggplot(sims %>% group_by(n_particles, gps_error, system_noise) %>%
+#         summarize(p_bad = mean(bad_sample))) + 
+#     geom_col(aes(as.factor(n_particles), y = p_bad, fill = as.factor(system_noise)), position = "dodge") +
+#     facet_grid( ~ gps_error) +
+#     xlab("Numer of particles") + ylab("Degeneration Rate") +
+#     labs(fill = "System Noise")
+dev.off()
+
+
+
+
+get_nw_times <- function(sim) {
+    siminfo <- strsplit(sim, "_")[[1]][-1]
[…]

### 2018-10-30 anzjs2018/anzjs2018.tex
Write some words about moels
more writing
Writing model information
models
network model
Implementation info
intro very drafty
Double spacing
Figures for network and pf
edits

@@
 \usepackage[UKenglish]{isodate}
+\usepackage{float}
+\usepackage{subcaption}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+% \newfloat{lstfloat}{htbp}{lop}
+% \floatname{lstfloat}{Listing}
+% \def\lstfloatautorefname{Listing}
+
+% \doublespace
 
 \input{../glossary.tex}
+
+\newcommand{\bv}{\boldsymbol{v}}
+\newcommand{\bz}{\boldsymbol{z}}
+\newcommand{\bw}{\boldsymbol{w}}
+\newcommand{\bR}{\boldsymbol{R}}
+\newcommand{\bT}{\boldsymbol{T}}
+\newcommand{\br}{\boldsymbol{r}}
+\newcommand{\btheta}{\boldsymbol{\theta}}
+
+\newcommand{\rt}{real-time\ }
 
@@
 \cleanlookdateon
-\runningheads{Short title of paper}{TOM~ELLIOTT AND THOMAS~LUMLEY}
-\title{Long name of the paper}
+\runningheads{Modeling transit vehicles in real-time}{TOM~ELLIOTT AND THOMAS~LUMLEY}
+\title{A real-time particle filter for modeling travel times of transit vehicles through a GTFS-based road network}
 \author{Tom Elliott\corrauth~and~Thomas~Lumley}
@@
 \begin{abstract}
-Predicting transit vehicle arrival time is a somewhat complex procress.
-Road networks are dynamic, and can change from free-flowing to highly congested
-very quickly.
-Any reliable prediction framework needs to be able to respond to these changes,
-but also to future trends, such as typical changes to traffic flow
-at peak times, based on hisorical data.
-Of course, the major constraint on any prediction framework is going to be
-computational efficiency.
-Due to the realtime nature of the problem, predictions need to be 
-available as soon as possible after observing the vehicles' positions,
-which are updated approximately every 30~seconds.
-In this paper, we describe our C++ implementation of a particle filter to estimate
-vehicle state in combination with a network state model,
-allowing us to model transit vehicle flow through the network.
-While our application here is being testing in Auckland, New Zealand,
-we hope that our methods are general enough to be applied to any
-transit network that uses GTFS.
-Preliminary results show that, with suitable hardware,
-the framework is fast enough while retaining the complexity necessary
-to incorporate as much information as necessary to predict arrival times.
+Predicting the arrival time of a transit vehicle involves not only
+knowledge of its current position and schedule adherance,
+but also information about traffic conditions along the remainder of its route.
+Road networks are dynamic, 
+and can change from free-flowing to highly congested very quickly,
+which can effect the arrival time of transit vehicles, 
+buses in particular,
+% This in turn leads to commuter frustration when they have to wait longer than
+% initially anticipated,
+so any reliable prediction framework needs to account for \rt traffic conditions.
+In this paper,
+we propose a particle filter that can be used to estimate travel times
+along roads within a transit network based solely on GTFS data.
+Due to the \rt nature of the problem and computational demands of such an approach,
+we demonstrate the faesibility of our program, 
+written in the R programming language and making extensive use of C++.
+The application provides the basis from which future work will be based,
+most notably for combining the \rt travel times with historical data and a 
+formal road network model to make reliable arrival time predictions.
+% Of course, the major constraint on any prediction framework is going to be
+% computational efficiency.
+% Due to the realtime nature of the problem, predictions need to be 
+% available as soon as possible after observing the vehicles' positions,
+% which are updated approximately every 30~seconds.
+% In this paper, we describe our C++ implementation of a particle filter to estimate
+% vehicle state in combination with a network state model,
+% allowing us to model transit vehicle flow through the network.
+Development and testing has taken place here in Auckland, New Zealand,
+although we expect our methods are general enough to be applied to any
+transit network using GTFS.
 \end{abstract}
 
-\keywords{particle filter; Kalman filter; transit modeling;
-          transit networks; statistical computing; applications; gtfs}
+\keywords{particle filter; transit modeling;
+          transit networks; applications; gtfs}
 
@@
 \input{sections/03_models}
-\input{sections/04_implement}
-\input{sections/05_results}
+\input{sections/04_implementation}
+% \input{sections/05_results}
 \input{sections/06_discuss}

### 2018-10-30 anzjs2018/sections/01_introduction.tex
edits

@@
 due to traffic congestion, or  because the schedule is poorly calibrated.
-A second is that buses are shown as on time but in fact are not running,
-and therefore have not registered on the \rt system.
-In this case, the scheduled times are used for ETAs and, 
-when the bus doesn't show up, 
-the bus disappeared from the \rt board and we get what passengers have described as ``Ghost Buses''.
+Another is that buses are shown as \emph{on-time} if they are not present
+in the \rt system,
+no matter whether this is due to technical problems,
+or that the bus has failed to show up;
+in this case, schedule times are used to determine ETAs and, 
+if the bus doesn't show up on time,
+the service is removed from the \rt board 
+and we get what passengers have described as ``Ghost Buses''.
 
@@
 This behaviour is exaggerated when there is a large distance between stops,
-as the ETA can change drastically.
+as the ETA is only updated at stops so there can be a long time between updates.
 
 To avoid these problems and improve the reliability of ETAs,
-we first needed to develop a model to infer position and speed
-of transit vehicles from \rt GPS data,
-isntead of relying on delays reported only at stops.
-Several modeling approaches were explored, 
+we required two things:
+first, a robust vehicle model to estimate the position and speed
+of transit vehicles from \rt GPS data;
+and secondly, a way to incorporate road congestion information 
+into the arrival time predictions.
+Several vehicle modeling approaches were explored, 
 such as the Kalman filter (used by \cite{Dailey_2001} and \cite{Cathey_2003}, ...),
@@
 by \cite{Hans_2015}.
-We chose to use the last of these, the particle filter,
-as they are very flexible and aren't limited by assumptions
-such as unimodality;
-instead, they are constrained by computational complexity.
+The last of these, the particle filter, was chosen for several reasons,
+discussed in Section~\ref{sec:pf}.
 
-The \emph{reliablility} of arrival time estimates
-alo needs to be considered.
-Ideally, congestion along intermediate roads should be incorporated into the predictive model,
-and one way of doing this is to use the travel time of preceeding buses along the same roads.
-Several such approaches have been proposed.
-\cite{Yu_2010} used the travel time of vehilces on the same \emph{routes},
-which they showed to improve ETAs.
-However, this only works on routes with a high frequency.
-\citep{Yu_2011} extended this by looking at roads where several major rouets converged,
-again showing improvements but limited in scope.
-In general, however, no comprehensive network modeling approach has been proposed that used
+In order to improve the reliability of ETAs,
+congestion along intermediate roads should be incorporated into the predictive model.
+One solution is to use the travel time of preceeding buses along the same roads
+to predict travel times for the current bus,
+such as proposed by \cite{Yu_2010},
+who were able to improve prediction accuracy
+by using the travel time of vehicles on the same \emph{route}.
+However, this only works well for high-frequency routes, so
+\cite{Yu_2011} modeled travel times along roads where several major routes converged,
+which showed further improvement however is limited in scope.
+In general, however, no comprehensive network modeling approach has been proposed us
 solely GPS position data to estimate arrival times
@@
 to update congestion information in \rt.
-The vehicle model is implemented using a particle filter,
-so it is also necessary to demonstrate the faesibility of this as a valid real-time option.
+As the vehicle model is implemented using a particle filter,
+we demonstrate its faesibility as a valid real-time option.
 From the particle filter estimates of vehicle trajectories,  
-we estimate the travel time along roads update the road network's state.
-Future work will involve using this travel time information 
+The particle filter provides estimates of vehicle travel times along roads,
+which can be used to update the road network's state.
+Future work involves using this travel time information 
 in combination with historical data to develop an arrival time prediction model.

### 2018-10-30 anzjs2018/sections/02_gtfs.tex
edits

@@
 Developed and maintained by Google \citep{GoogleDevelopers_2006},
-who make use of it in their Google Maps Transit Directions,
+who use it in Google Maps Transit Directions,
 it is used by over 900~transit providers around the world,
@@
 As mentioned in Section~\ref{sec:intro},
-there are some major issues with the current ETA generation method.
+there are some major issues with the current prediction method in Auckland.
 These are almost directly attributed to GTFS-realtime trip updates,
-which are currently the sole source of data for ETAs (in Auckland).
+which are currently the sole source of data for ETAs.
 Trip updates are reported whenever a vehicle arrives or departs a stop,
-and includes most importantly the delay between the scheduled and actual arrival times.
-This delay is then propagated on to all future stops to adjust the ETA.
+and includes the delay between the scheduled and actual arrival times,
+and is propagated to all future stops to adjust their ETAs.
 This assumes that the schedule is well calibrated and the time between scheduled arrivals
 is representative of the real-world travel time between stops. 
-Even more problematic is that, in the absense of trip updates,
-the delay defaults to zero,
-and unless the trip is manually cancelled,
-appears on-time to passengers waiting at the stop.
-If the bus is late or cancelled,
-the trip disappears from the \rt board and passengers
-are unable to distingush the two scenarios.
+In many places in Auckland, 
+traffic congestion is highly dynamic and changes from day to day,
+so even a well-calibrated schedule is an unreliable baseline.
 
@@
 
-Arguably the most important predictor of arrival time is
-the cumulative travel time along intermediate roads.
+The primary predictor of arrival time is 
+the travel time along roads between where the bus is \emph{now},
+and the stop at which the passenger is waiting.
 In most applications, however, this vital information is unavailable,
 at least directly,
-so here we attempt to construct a transit road network of intersections
-and the roads between them, 
-and map each route to a sequence of road segments.
+so here we construct a transit road network consisting of intersections
+and the roads between them.
+By generalising to physical roads,
+we remove any route-dependency,
+which increases the quantity of data available for each road.
+Then, each route can instead be represented as a sequence of road segments,
+each with a travel time state.
 
-The simplest way to construct such a network is to use the stop sequence:
-routes with a common sub-sequence of stops must be traveling the same route between them,
-so by defining a road as the path between two consecutive stops 
-we can obtain travel time estimates along that road segment.
-There are several places where this doesn't quite work, for example express routes 
-and multi-stop locations where there are high numbers of routes,
-resulting in some overlap between road segments,
-it is a viable enough simplication for the purposes of this paper.
-Figure~\ref{fig:network_creation} demonstrates how several overlapping routes 
-are merged to form a road network.
+To construct the network,
+we use stops as nodes, so that
+routes with a common sub-sequence of stops will share the roads between them;
+roads are then defined as the links between stops,
+as demonstrated in Figure~\ref{fig:network_creation}.
 In section~\ref{sec:kf} we present a model for estimating vehicle travel time
 along a road segment.
-Future work will look at using physical intersections as nodes in the network to improve it.
+There are several places where this doesn't quite work, 
+for example express routes which do not service certain stops,
+and multi-stop locations,
+which results in overlapping road segments.
+Future work will look at using intersection points of the route shapes
+as nodes in the network to improve the coverage and reduce overlap.
 
@@
 
-GTFS \rt allows developers to query the current positions of vehicles
-in the transit network.
+GTFS-realtime provides the current positions of vehicles in a transit network.
 The data consists of the time $t_k$ that the observation was made,
 the GPS position of the vehicle, $\bY_k$, 
-and some other information about the trip being serviced.
-Vehicle positions are updated with a frequency of anywhere between 10~seconds and several minutes,
-so there is often a lot of uncertainty about the trajectory
-between two observations, particularly when there is one or more bus stop
-or intersection between them.
+and information about the trip being serviced.
+In Auckland, vehicle positions are updated with a frequency of anywhere between 10~seconds and several minutes,
+so there is often a lot of uncertainty about a vehicle's intermediate trajectory,
+particularly when there are one or more bus stops along the way.
 It is also possible for a bus to remain stationary,
-so the possible vehicle positions rapidly increases with the time between observations.
+for example due to heavy congestion,
+so the number of possible trajectories rapidly increases with 
+the time between observations.
 
-Another complication with the Auckland Transport realtime feed is that
-the buses are programmed to report their location when arriving at
-bus stops and some major intersections.
-Often these updates are preemptive 
+One important consideration regarding the Auckland Transport's \rt implementation is that
+buses are programmed to report their location when arriving at or departing from
+bus stops, as well as some major intersections.
+To further complicate this,
+these positions can be preemtive,
 (i.e., the bus is almost there, but not quite),
-and subsequent observations place the bus \emph{behind} the stop or intersection
-(e.g., in a queue of traffic at traffic lights).
+so consecutive observations can show a bus traveling backwards.
 To handle this, we compute the approximate distance traveled, $\tilde x_k$,
@@
 if this has decreased, the current state is rejected and the vehicle reverted
-to its previous state.
+to its previous state become continuing.
+This way, we avoid degeneration of the particle fitler (Section~\ref{sec:model_perf})
+at the cost of losing one observation that is most likely invalid anyway.

### 2018-10-30 anzjs2018/sections/03_models.tex
more writing and pictures
edits

@@
 
-Due to the real-time nature of this application, we are using
-a recursive Bayesian filtering approach as is often used in
-real-time tracking applications.
-We are using a separate model for each of the vehicle and network states,
-and in each assume an underlying Markov process with state $\mathcal{X}_k$
-with system noise $\omega_k\sim\mathrm{N}(0,\mathcal{Q})$ 
+Real-time tracking applications such as ours often use
+a Bayesian filtering approach,
+or recursive Bayesian estimation,
+in which the data is processed sequentially instead of all at once,
+reducing computational demands.
+These models assume an underlying Markov process with state $\mathcal{X}_k$
+and system noise $\omega_k\sim\mathrm{N}(0,\mathcal{Q})$ 
 of which observations $\mathcal{Y}_k$ are made
@@
 \end{equation}
-The transition function $f$ describes the relationship between consecutive states
-(i.e., what we want to model),
+The transition function $f$ describes the relationship between consecutive states,
 while the measurement function $h$ is a deterministic function describing the
@@
 
-The following sections describe the two models used in this application.
+The following sections describe the two models used in this application,
+both of which are examples of a Bayesian filter.
 The first, implemented using a particle filter, estimates vehicle states
@@
 and is implemented using an \emph{information filter},
-a variant of the Kalman filter \citep{}.
+a variant of the Kalman filter \citep{Anderson_2012}.
 
@@
 the vehicle's distance traveled $x$ in meters along the route and
-it's speed $\dot x$ in meters per second.
+its speed $\dot x$ in meters per second.
 Additionally, travel times along road segments along the route, 
@@
 requiring an increasing number of particles as the model complexity and
-number of parameters increases \citep{}.
+number of parameters increases \citep{Carpenter_1999}.
 Section~\ref{sec:rt} describes our implementation of the particle filter in real-time 
@@
         \caption{The transition function $f$ models the behaviour of each particle,
-            including changes in acceleration and stopping behaviour at bus stops.}
+            including changes in acceleration and stopping behaviour at bus stops,
+            which can result in multiple modes.}
         \label{fig:pf_state_predict}
@@
         \includegraphics[width=\textwidth]{figures/03_particle_filter_6.pdf}
-        \caption{Each particle's distance from the true observation is calculated using
+        \caption{Each particle's distance from the true observation (red cross) 
+            is calculated using
             the geographic distance between two coordinates, which can then be used
@@
     \caption{The vehicles state is estimated using a set of discrete points, or particles,
-        each of which is independently mutated to make predictions of the future state. The history of the state estimates can be used to estimate parameters of interest, such as
-        travel times along roads.}
+        each of which is independently mutated to make predictions of the future state.
+        Multimodality is easily handled, as demonstrated here,
+        without degeneration of the filter.}
     \label{fig:pf_state}
@@
 \emph{importance resampling} is employed whenever the 
-effective sample size $N_{\text{eff}} = 1/(\sum_i (W_k^{(i)})^2)$
+effective sample size 
+\begin{equation}
+\label{eq:neff}
+N_{\text{eff}} = \frac{1}{\sum_i (W_k^{(i)})^2}
+\end{equation}
 falls below a specified threshold $N_{\text{thres}}$.
 The resampling is performed with replacement using importance weights $W_k^{(i)}$,
-as demonstrated in Figure~\ref{fig:pf_state_update}.
+as demonstrated in Figure~\ref{fig:pf_state_predict2}.
 Afterwards, the weights are reinitialized to $W_k^{(i)} = N^{-1}$.
@@
 
-The particle filter lets us to develop a model in which
-the vehicle transition function $f$ implements several key bus behaviours:
+The particle filter allows us to develop a model of several important bus behaviours:
 \begin{itemize}
@@
 future work will investigate modeling these in real-time similarly to Section~\ref{sec:kf}.
-% No intersection model is implemented here,
-% as the segment road segments are between stops,
-% and the travel times predicted by the network model at time $t_k$,
-% $\btheta(t_k) = (\theta_1(t_k), \ldots, \theta_L(t_k))^\top$
-% (see section~\ref{sec:kf}).
-The transition function algorithm is shown in Figure~\ref{fig:algorithm},
+
+The stop positions are represented by their distance into trip,
+$\boldsymbol{S} = (S_1, \ldots, S_J)^\top$.
+The $stop\_index$ function returns the index of a stop,
+given a distance traveled and set of stop distances.
+The transition function algorithm is shown in Algorithm~\ref{fig:algorithm},
 which implements those features discussed above.
+Lines~2--9 are necessary to decrease the degeneration rate of the filter,
+which can occur when there is a long delay between observations
+simply because the bus has not moved due to traffic or other reasons. 
+We therefore place positive probability for some particles to remain stationary.
+
+Lines~11--12 of Algorithm~\ref{fig:algorithm} implement the system noise,
+which involves changing the vehicle's speed each second (i.e., acceleration).
+Lines~13--25 handle arrival behaviour at stops:
+if the particle passes the next stop,
+it has a chance of stopping (lines~18--19),
+in which case the dwell time is sampled and deducted from the 
+time remaining to travel (lines~20--21).
+Finally, to track travel time along segments,
+at each point where the vehicle travels (lines 9 and 26),
+the travel time for that segment, $z_j$, is incremented.
+The mutation ends once the travel time $\delta$ reaches zero,
+or when the vehicle reaches the end of the route (lines 10, 14).
+Once finished, the particle's new state is returned (line 29).
+
+\renewcommand{\algorithmicrequire}{\textbf{Start:}}
+\newcommand{\algorithmicbreak}{\textbf{break}}
+\newcommand{\Break}{\State \algorithmicbreak}
+\begin{algorithm}[tb]
+    \caption{Particle mutation function.}
+    \label{fig:algorithm}
+    \begin{algorithmic}[2]
+    \Require $x_{k-1}, \dot x_{k-1}, \boldsymbol z$\\
+    Initialization $\delta \gets t_k - t_{k-1}$, 
+        $j \gets stop\_index(x, \boldsymbol{S})$
+    \If {$u \sim \mathrm{U}(0,1) < 0.5$ and $x = S_j$}
+        \Comment{Remain longer at stop}
+        \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
+        \State $\delta \gets \delta - \gamma - d_j$
+    \ElsIf {$u \sim \mathrm{U}(0,1) < 0.1$}
+        \Comment{Occasionally bus stopped due to traffic}
+        \State $w \gets \min(\delta, \mathrm{Exp}(1/\delta))$
+        \State $\delta \gets \delta - w$
+        \State $z_j \gets z_j + w$
+        \Comment{Congestion counts as travel time}
+    \EndIf
+    \While {$\delta > 0$ and $x < S_J$} 
+        \State $\dot x \sim \mathrm{N}(\dot x, \sigma^2)$
+        \Comment{System noise, trunctated to be non-negative}
+        \State $x \leftarrow x + \dot x$
+        \If {$x \geq S_{j+1}$}
+            \If {$j+1 = J$}
+                \Comment{Break at end of route}
+                \State $x \gets S_J$
+                \Break
+            \EndIf
+            \State $u_j \sim \mathrm{U}(0,1)$
+            \Comment{Simulate vehicle stopping}
+            \If {$u_j < \pi$}
+                \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
+                \State $\delta \gets \delta - \gamma - d_j$
+                \State $x \gets S_j$
+            \EndIf
+            \State $j\gets j+1$
+        \EndIf
+        \State $\delta \gets \delta - 1$
+        \State $z_j \gets z_j + 1$
+        \Comment{Track travel time along road}
+    \EndWhile
+    \\
+    \Return $x, \dot x, \boldsymbol z$
+    \end{algorithmic}
+\end{algorithm}
 
@@
 and an additional function $g$ which transforms GPS coordinates onto a flat
-surface (the geographical equirectangular projection),
-allow the placement of a bivate normal likelihood on the data
-as shown in figure~\ref{fig:pf_lhood}.
+surface (the geographical equirectangular projection, \cite{Snyder_1998}),
+allow the placement of a bivate normal likelihood on the data.
 The model for the observations is,
@@
 The likelihood of the data given a particle's state estimate 
-is therefore calculated using (\ref{eq:obs_exp})
+cane be calculated using (\ref{eq:obs_exp})
 \begin{equation}
@@
 in other approaches, such as the Kalman filter,
-the likelihood cannot be calculated using the distance as the estimate
-is a distribution, and instead a reverse non-deterministic transformation,
+a reverse non-deterministic transformation,
 the inverse measurement function,
@@
 as well as make short-term predictions for estimating arrival times.
-In this paper,  we apply a very simple model in which each road is treated independently,
-future work will investigate using correlations with adjacend roads and historical data
+In this paper, we develop simple model in which each road is treated independently.
+Future work will investigate using correlations with adjacent roads and historical data
 to improve the model and subsequent predictions.
@@
 
-Travel times are assumed to be non-negative but otherwise normally distributed,
-and constant over time---we do not attempt to model peak congestion in this paper.
-Additionally the observations are non-constact and may be multiple as several buses travel
-a road simultaneously.
+Since multiple vehilces can travel along a road at the same time,
+we have used an information filter to implement the network model.
 The information filter is a transformation of the Kalman filter in which the
@@
 the covariance matrix and state vector, respectively.
-This allows multiple observations to be added together to update the state.
+This allows multiple observations to be added together to update the state
+in a single iteration.
 
@@
 and the information vector $u^\ell_c~=~\hat \theta^\ell_{c|c-1} (P^\ell_{c|c-1})^{-1}$.
-The travel time estimates from the vehicle model for vehicle $m$ along segment $\ell$
+The travel time estimate of vehicle $m$ along segment $\ell$
 is $\bar z_c^{m\ell}$, with uncertainty $s^{m\ell}_c$,
-can be transformed to a measurement information covariance matrix 
+and is transformed to a measurement information covariance matrix 
 $B_c^{m\ell} = (s^{m\ell}_c)^{-2}$
 and measurement information vector $b_c^{m\ell}~=~\bar z^{m\ell}_c (s^{m\ell}_c)^{-2}$.
-The information update is now the sum of the information over all $M$ vehicles
-that traversed segment $\ell$ since the last update, so
+The total information is the sum of the information over all $M$ vehicles
+that traversed segment $\ell$ since the last update,
+so the update equation is
 \begin{align*}
 U^\ell_{c|c} &= U^\ell_{c|c-1} + \sum_{m=1}^M B^{m\ell}_{c} \\
-\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^M b^{m\ell}_{c}
+\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^M b^{m\ell}_{c}.
 \end{align*}
-which are back transformed to obtain the desired parameter estimates
+The desired parameter estimates (tavel time) are obtained 
+by the inverse transformations
 \begin{equation*}
@@
 \quad\text{and}\quad
-P^\ell_{c|c} = \frac{1}{U^\ell_{c|c}}.
+P^\ell_{c|c} = \frac{1}{U^\ell_{c|c}}
 \end{equation*}
+which can now be used to predict the travel times of upcoming vehicles.

### 2018-10-30 anzjs2018/sections/04_implementation.tex
edits

@@
-\section{Real-time implementation and preliminary results}
+\section{Real-time implementation}
 \label{sec:rt}
@@
 
-The general structure of the \rt component is:
+The general structure of the \rt component is shown below,
+with the bold steps being those discussed in this paper.
 \begin{enumerate}
@@
     \item Update or create new vehicle objects from new data
-    \item Run particle filter on each vehicle to estimate update state
-    \item Collect travel time information for each vehicle for any completed roads
-    \item Update road network for any completed road segments
+    \item \textbf{Run particle filter on each vehicle to update state}
+    \item \textbf{Update state of any roads for which vehicles 
+        have completed travel}
     \item Generate ETAs for vehicles uing combination of particle filter and network state
@@
 is as efficient as possible,
-allowing ETAs to be generated and distributed fast enough---our initial goal was 30~seconds.
-Using \verb|C++| provides enough control over memory management
+allowing ETAs to be generated and distributed fast enough to be faesible in real-time,
+with a target of 30~seconds or faster at peak time.
+Using C++ provides enough control over memory management
 to ensure the program runs fast enough in \rt 
 while still implementing the models.
-We could also use OpenMP for parallelisation,
-allowing us to process multiple vehicles simultaneously
-depending on the number of cores available.
+We also use OpenMP for parallelisation,
+allowing us to process multiple vehicles simultaneously,
+which are each independent so threadsafety is a non-issue.
 
@@
 with varying number of particles.
-It is also necessary to assess the performance of the particle filter,
-and compare values for the fixed model parameters.
+We also assess the performance of the particle filter
+for a range of fixed model parameters values,
+such as system noise and GPS error.
 To do so, we implemented a simulated \rt environment
@@
 
-\begin{figure}[tb]
-    \centering
-    % \includegraphics[width=0.7\textwidth]{figures/04_model_results_gps.pdf}
-    \caption{The distance between the observation and the nearest point on the shape.}
-    \label{fig:gps_dist}
-\end{figure}
+% \begin{figure}[tb]
+%     \centering
+%     % \includegraphics[width=0.7\textwidth]{figures/04_model_results_gps.pdf}
+%     \caption{The distance between the observation and the nearest point on the shape.}
+%     \label{fig:gps_dist}
+% \end{figure}
 
@@
 During development of our application,
-our primary concert was the computational faesibility of a particle filter.
+our primary concern was the computational faesibility of a particle filter.
 For each iteration, 
-the timings of the individual components in 2a--f above are recorded.
+the timings of the individual components are recorded.
 Since the number of vehicles traveling at any given time changes throughout the day,
@@
 and the vehicle is reinitialised.
-Note, however, that the lower limit of this value 
-is attributed to \emph{invalid trips},
-however there is no simple way of filtering these trips.
+Note, however, that some observations are \emph{invalid trips},
+in which the vehicle is logged in to the wrong trip,
+or ``off duty'' while still transmitting for a trip.
+Some of these can be filtered for being excessively far from the route,
+but others---for example traveling the wrong direction---cannot be.
 The last statistic we compute is the variance of road segment travel times
@@
 
-The one parameter we can determine from the data is GPS error, $\epsilon^2$,
+The one parameter we can investigate from the data is GPS error, $\epsilon^2$,
 by looking at the distance between observations and the shape.
@@
 observation and the \emph{nearest point on the path}.
-Due to the way the observations are reported, 
-many observations will be ``exactly'' on the route at a bus stop,
-which explains the spike at just above zero~meters.
+We see two peaks,
+one at about 1m, and another at 2.5m.
+The first of these is possibly an artefact of the data,
+for example automated waypoints at intersections.
+We have explored a range of GPS error values from 1 to 5~m.
 
@@
 
-Lastly is the variability of travel times and their uncertainties.
-These are shown in Figure~\ref{fig:travel_times}.
+Lastly is the uncertainty of travel time estimates,
+as shown in Figure~\ref{fig:travel_times}.
 System noise again shows no effect on the variability of travel time estimates,
@@
     \includegraphics[width=0.9\textwidth]{figures/04_model_results_times.pdf}
-    \caption{Travel times.}
+    \caption{The within-segment posterior travel time uncertainty (standard error), subset by number of particles.}
     \label{fig:travel_times}

### 2018-10-30 anzjs2018/sections/06_discuss.tex
edits

@@
 This allows vehicles to inform others---irrespective of route---of 
-how long it will take to reach future stops.
+current \rt traffic conditions,
+and provide a means of obtaining more reliable ETAs.
 
-Not only is the particle filter a viable choice in terms of speed,
-it has shown it can handle most situations without losing the vehicle,
-with the exception of when the vehicle is not traveling along
-the specified route.
-Hopefully with future development,
-the stopping behaviour can be more accurately modeled,
-perhaps incorporating trip updates into the likelihood function.
+We have shown that the particle filter a viable \rt option in terms of speed,
+and while it occasionally loses the bus,
+a lot of this is attributed to invalid data (incorrect route, direction, etc).
+Future development will incorporate more accurate stopping behaviour,
+perhaps using GTFS arrival and departure times in the likelihood function.
 
 The network model is still in the early stages of development,
-but given that the implementation is running quickly---twice as fast as our target---%
+but given that the implementation is running well within our 30~second target,
 there is room for developing a more complex model.

### 2018-10-30 anzjs2018/figures/03_particle_filter.R

@@ -69,7 +69,7 @@ grid.text("h(x)", 0.5, 0.5, gp = gpar(fontface = 3, fontfamily = 'HersheySerif')
 dev.off()
 
 
-x2 <- x + runif(N, 10, 20) * 120
+x2 <- x + runif(N, c(10, 20), c(20, 25)) * 120
 
 pdf('figures/03_particle_filter_2.pdf', width = WIDTH, height = HEIGHT)
 gridExtra::grid.arrange(
@@ -103,12 +103,12 @@ x3 <- sample(x2, replace = TRUE, prob = wt)
 pdf('figures/03_particle_filter_4.pdf', width = WIDTH, height = HEIGHT)
 gridExtra::grid.arrange(
     p0 + 
-        geom_point(aes(lon, lat), data = h(x), colour = 'gray') +
+        # geom_point(aes(lon, lat), data = h(x), colour = 'gray') +
         geom_point(aes(lon, lat), data = h(x2), colour = 'gray') +
         geom_point(aes(lon, lat), data = h(x3)) +
         geom_point(aes(lon, lat), data = y, col = "orangered", pch = 4, size = 2, stroke = 1.3),
     d0 + 
-        geom_dotplot(aes(x), data = tibble(x = x), binwidth = 50, color = 'gray', fill = 'gray') +
+        # geom_dotplot(aes(x), data = tibble(x = x), binwidth = 50, color = 'gray', fill = 'gray') +
         geom_dotplot(aes(x), data = tibble(x = x2), binwidth = 50, color = 'gray', fill = 'gray') +
         geom_dotplot(aes(x), data = tibble(x = x3), binwidth = 50),
     ncol = 1)
@@ -139,7 +139,8 @@ ggplot() +
     theme(panel.grid = element_blank(), axis.text = element_blank()) +
     scale_x_continuous(position = 'top') +
     ylab("Latitude") + xlab("Longitude") +
-    geom_point(aes(lon, lat), data = h(x2), col = ifelse(1:N == 1, 'black', 'gray')) +
+    geom_point(aes(lon, lat), data = h(x2[x2 < 11500]), col = 'black') +
+    # geom_point(aes(lon, lat), data = h(x2[x2 < 11500][1]), col = 'black') +
     geom_point(aes(lon, lat), data = y, col = "orangered", pch = 4, size = 2, stroke = 1.3) +
     geom_segment(aes(lon, lat, xend = y$lon, yend = y$lat), data = h(x2)[1,], lty = 2)
 dev.off()

### 2018-10-30 anzjs2018/figures/04_model_results.R

@@ -168,22 +168,52 @@ if (file.exists("nwtimes.rda")) {
     save(nwtimes, file = "nwtimes.rda")
 }
 
+segids <- table(nwtimes$segment_id) %>% sort %>% tail(100) %>% names
 
-nwtimes.smry1 <- nwtimes %>% group_by(n_particles, gps_error, system_noise, segment_id) %>%
-    summarize(sd.time = sd(var, na.rm = TRUE), mean.var = mean(var, na.rm = TRUE)) %>% ungroup
+nwtimes.smry1 <- nwtimes %>% filter(segment_id %in% segids) %>% 
+    group_by(n_particles, gps_error, system_noise, segment_id) %>%
+    summarize(mean.tt = mean(mean, na.rm = TRUE), var.tt = sd(mean, na.rm = TRUE),
+        sd.time = sd(var, na.rm = TRUE), mean.var = mean(var, na.rm = TRUE)) %>% ungroup
 
 nwtimes.smry <- nwtimes.smry1 %>% group_by(n_particles, gps_error, system_noise) %>%
-    summarize(mean = mean(sd.time, na.rm = TRUE), sd = mean(mean.var, na.rm = TRUE)) %>% ungroup()
+    summarize(mean = mean(sd.time, na.rm = TRUE), sd = mean(mean.var, na.rm = TRUE/mean.tt)) %>% ungroup()
 
-# ggplot(nwtimes.smry) +
-#     geom_point(aes(sqrt(mean.var/n_particles), var.var, color = as.factor(system_noise))) +
-#     facet_grid(~n_particles)
+# smry2 <- 
+#     nwtimes %>% filter(segment_id %in% segids) %>%
+#     group_by(n_particles, gps_error, system_noise, segment_id) %>%
+#     summarize(sd.time = sd(var, na.rm = TRUE), mean.var = mean(var, na.rm = TRUE)) %>%
+#     group_by(n_particles, gps_error, system_noise) %>%
+#     summarize(mean = mean(sd.time, na.rm = TRUE), sd = mean(mean.var, na.rm = TRUE)) %>% 
+#     ungroup()
 
-pdf('figures/04_model_results_times.pdf', width = 10, height = 5)
-ggplot(nwtimes.smry) +
-    geom_point(aes(mean / sqrt(n_particles), sd / sqrt(n_particles), 
+# ggplot(nwtimes %>% filter(segment_id %in% segids) %>% 
+#     group_by(n_particles, gps_error, system_noise, segment_id) %>%
[…]

### 2018-10-30 anzjs2018/sections/06_discuss.tex
some words + graphs

--- 2018-08-27 anzjs2018/sections/06_discuss.tex
@@
 
-It all seems to work (hopefully I don't need to change this D:)
+In this paper we have described an approach to transit vehicle modeling
+that enables the \rt estimation of road state from vehicle position data,
+which will in future be used to make what will hopefully be more reliable 
+arrival time predictions.
+The vehicle model, implemented using a particle filter,
+allows complex bus behaviours to be modeled,
+with future work planned to explore the other parameters---dwell time and stopping probabilities---%
+as well as to extend the model to also account for intersections.
+The travel times estimated from the particle filter then allow us to model \rt road state.
 
-Next steps include
-- generating useful priors for ``fall back'' (i.e., no data, future) prediction
-- develop and implement a more complex network update model (correlations, etc)
-- prediction estimates (point vs interval)
+The network model presented here is of the simplest nature,
+but provides the foundation for future work which will involve
+deveoping a formal model that combines \rt information with
+historical data, allowing predictions of future road state to be more accurate,
+for example incorporating peak time congestion forecasts.
+This in turn provides the basis for an arrival time prediction model,
+which will be a combination of vehicle state and road state forecasts.
+
+The primary purpose of this paper was to present our approach and 
+explore and demonstrate tis faesibility as a \rt option for RTI in transit systems.
+We have been able to construct an application that runs in \rt and capable
+of producing estimates within 30~seconds---although this can be reduced 
+by using fewer particles or, if necessary, increasing the number of cores.
+
+% The main goal of this work was to explore the faesibility
+% of a particle filter as a \rt estimation method for transit vehicle states,
+% and the use of estimated states to determine traffic conditions
+% throughout the network.
+% This allows vehicles to inform others---irrespective of route---of 
+% current \rt traffic conditions,
+% and provide a means of obtaining more reliable ETAs.
+
+% We have shown that the particle filter a viable \rt option in terms of speed,
+% and while it occasionally loses the bus,
+% a lot of this is attributed to invalid data (incorrect route, direction, etc).
+% Future development will incorporate more accurate stopping behaviour,
+% perhaps using GTFS arrival and departure times in the likelihood function.
+
+% The network model is still in the early stages of development,
+% but given that the implementation is running well within our 30~second target,
+% there is room for developing a more complex model.
+% Future work will explore the use of historical data to develop informative priors
+% for travel times,
+% which will provide a useful fallback in the case of no data,
+% or for making more accurate long-term predictions.
+% Once such a model has been developed, we plan to explore prediction estimates,
+% particuarly comparing point and intervals as a means of communicating 
+% uncertainty to commuters.

### 2018-10-30 anzjs2018/figures/04_model_results.R

@@ -168,7 +168,56 @@ if (file.exists("nwtimes.rda")) {
     save(nwtimes, file = "nwtimes.rda")
 }
 
-segids <- table(nwtimes$segment_id) %>% sort %>% tail(100) %>% names
+segids <- table(nwtimes$segment_id) %>% sort %>% tail(50) %>% names
+
+## plot just one segment
+# sdata <- nwtimes %>% filter(segment_id %in% segids) %>%
+#     group_by(n_particles, gps_error, system_noise) %>%
+#     do((.) %>% filter(timestamp > min((.)$timestamp))) %>%
+#     ungroup %>% group_by(segment_id, timestamp) %>% 
+#     do((.) %>% mutate(overall.mean = mean(mean))) %>% 
+#     ungroup %>% group_by(n_particles, gps_error, system_noise, segment_id) %>%
+#     summarize(meansqr = mean(((mean - overall.mean) / var)^2)) %>%
+#     ungroup %>% group_by(n_particles, gps_error, system_noise) %>%
+#     summarize(med.sqr = median(meansqr, na.rm = TRUE),
+#         low.sqr = quantile(meansqr, 0.05, na.rm = TRUE),
+#         high.sqr = quantile(meansqr, 0.75, na.rm = TRUE))
+
+# ggplot(sdata) +
+#     geom_pointrange(aes(as.factor(gps_error), med.sqr, ymin = low.sqr, ymax = high.sqr)) +
+#     facet_grid(n_particles~system_noise, scales = "free_y")
+
+# ## old
+# ggplot(sdata) + 
+#     geom_path(aes(timestamp, mean, colour = as.factor(n_particles))) +
+#     geom_linerange(aes(timestamp, ymin = mean - var, ymax = mean + var, colour = as.factor(n_particles))) +
+#     facet_grid(gps_error ~ system_noise)
+
+# ggplot(sdata) + 
+#     geom_path(aes(timestamp, mean, colour = as.factor(gps_error))) +
+#     geom_linerange(aes(timestamp, ymin = mean - 2*var, ymax = mean + 2*var, colour = as.factor(gps_error))) +
+#     facet_grid(n_particles ~ system_noise)
+
+# ggplot(sdata) + 
+#     geom_path(aes(timestamp, mean, colour = as.factor(system_noise))) +
+#     geom_linerange(aes(timestamp, ymin = mean - 2*var, ymax = mean + 2*var, colour = as.factor(system_noise))) +
[…]

### 2018-10-31 anzjs2018/figures/04_model_results.R
Update figure for ttimes

@@ -168,24 +168,30 @@ if (file.exists("nwtimes.rda")) {
     save(nwtimes, file = "nwtimes.rda")
 }
 
-segids <- table(nwtimes$segment_id) %>% sort %>% tail(50) %>% names
+segids <- table(nwtimes$segment_id) %>% sort %>% tail(10) %>% names
 
 ## plot just one segment
-# sdata <- nwtimes %>% filter(segment_id %in% segids) %>%
-#     group_by(n_particles, gps_error, system_noise) %>%
-#     do((.) %>% filter(timestamp > min((.)$timestamp))) %>%
-#     ungroup %>% group_by(segment_id, timestamp) %>% 
-#     do((.) %>% mutate(overall.mean = mean(mean))) %>% 
-#     ungroup %>% group_by(n_particles, gps_error, system_noise, segment_id) %>%
-#     summarize(meansqr = mean(((mean - overall.mean) / var)^2)) %>%
-#     ungroup %>% group_by(n_particles, gps_error, system_noise) %>%
-#     summarize(med.sqr = median(meansqr, na.rm = TRUE),
-#         low.sqr = quantile(meansqr, 0.05, na.rm = TRUE),
-#         high.sqr = quantile(meansqr, 0.75, na.rm = TRUE))
+sdata <- nwtimes %>% #filter(segment_id %in% segids) %>%
+    group_by(n_particles, gps_error, system_noise) %>%
+    do((.) %>% filter(timestamp > min((.)$timestamp))) %>%
+    ungroup %>% group_by(segment_id) %>% 
+    do((.) %>% mutate(tt.sd = sd(mean, na.rm = TRUE))) %>% 
+    ungroup %>% group_by(n_particles, gps_error, system_noise, segment_id) %>%
+    summarize(varp = sd(mean, na.rm = TRUE) / first(tt.sd)) %>%
+    ungroup %>% group_by(n_particles, gps_error, system_noise) %>%
+    summarize(varp.mean = mean(varp, na.rm = TRUE), varp.sd = sd(varp, na.rm = TRUE) / sqrt(n()))
 
-# ggplot(sdata) +
-#     geom_pointrange(aes(as.factor(gps_error), med.sqr, ymin = low.sqr, ymax = high.sqr)) +
-#     facet_grid(n_particles~system_noise, scales = "free_y")
+pdf('figures/04_model_results_times.pdf', width = 10, height = 2)
+ggplot(sdata) +
[…]

### 2018-10-31 anzjs2018/anzjs2018.tex
some words + graphs
More edits

@@
 
-% \doublespace
+\doublespace
 
@@
 }
+\ack{
+    We would like to thank Auckland Transport for providing free,
+    open access to the \rt data; and the Center for e-Research at the
+    University of Auckland for providing and maintaining the virtual
+    machine used in our work.
+}
 
@@
 knowledge of its current position and schedule adherance,
-but also information about traffic conditions along the remainder of its route.
+but also of traffic conditions along the remainder of the route.
 Road networks are dynamic, 
 and can change from free-flowing to highly congested very quickly,
-which can effect the arrival time of transit vehicles, 
-buses in particular,
-% This in turn leads to commuter frustration when they have to wait longer than
-% initially anticipated,
-so any reliable prediction framework needs to account for \rt traffic conditions.
-In this paper,
-we propose a particle filter that can be used to estimate travel times
-along roads within a transit network based solely on GTFS data.
-Due to the \rt nature of the problem and computational demands of such an approach,
-we demonstrate the faesibility of our program, 
-written in the R programming language and making extensive use of C++.
-The application provides the basis from which future work will be based,
-most notably for combining the \rt travel times with historical data and a 
-formal road network model to make reliable arrival time predictions.
-% Of course, the major constraint on any prediction framework is going to be
-% computational efficiency.
-% Due to the realtime nature of the problem, predictions need to be 
-% available as soon as possible after observing the vehicles' positions,
-% which are updated approximately every 30~seconds.
-% In this paper, we describe our C++ implementation of a particle filter to estimate
-% vehicle state in combination with a network state model,
-% allowing us to model transit vehicle flow through the network.
-Development and testing has taken place here in Auckland, New Zealand,
-although we expect our methods are general enough to be applied to any
-transit network using GTFS.
+which impacts the arrival time of transit vehicles,
+particularly buses which often share the road with other vehicles,
+so reliable predictions need to account for \rt and future traffic conditions.
+The first step in this process is to construct a framework with which road state
+(traffic conditions) can be estimated using \rt transit vehicle position data.
+We propose a framework which implements a vehicle model using a particle filter
+to estimate travel times along roads,
+used by a second model to estimate \rt traffic conditions which can
+later be used for arrival time prediction.
+We demonstrate the \rt faesibility of such an approach in \rt,
+and detail our implementation in R and C++.
+Development and testing took place in Auckland, New Zealand,
+with future work planned to incorporate historical data into the network model
+to improve the reliability of ETAs.
 \end{abstract}

### 2018-10-31 anzjs2018/sections/01_introduction.tex
More edits

@@
 
-In public transport, \rt information (RTI), for example estimated times of arrival (ETAs),
+In public transport, \rt information (RTI), such as estimated times of arrival (ETAs),
 keeps transport users informed about their journey,
-and in some places make decisions on the best route to take.
+and potentially make decisions on the best route to take.
 Unfortunately, here in Auckland, RTI is highly unreliable,
@@
 in the \rt system,
-no matter whether this is due to technical problems,
-or that the bus has failed to show up;
-in this case, schedule times are used to determine ETAs and, 
+regardless of whether this is due to technical problems,
+or the bus failing to show up;
+in either case, schedule times are used to determine ETAs and, 
 if the bus doesn't show up on time,
@@
 
-Arrival time prediction is only as accurate as the underlying model,
+Arrival time prediction is only as reliable as the underlying model,
 and while a lot of work has gone into developing public transport vehicle models
@@
 
-To avoid these problems and improve the reliability of ETAs,
-we required two things:
-first, a robust vehicle model to estimate the position and speed
+To avoid the problems associated with the arrival time predictions
+based solely on reported delays,
+a more robus modeling and prediction framework is required.
+This should consist of a robust vehicle model to estimate the position and speed
 of transit vehicles from \rt GPS data;
-and secondly, a way to incorporate road congestion information 
+and a way to incorporate road congestion information 
 into the arrival time predictions.
 Several vehicle modeling approaches were explored, 
-such as the Kalman filter (used by \cite{Dailey_2001} and \cite{Cathey_2003}, ...),
+including the Kalman filter \citep{Dailey_2001,Cathey_2003},
 machine learning models \citep{Yu_2006,Chang_2010},
-and a particle filter approach demonstrated
-by \cite{Hans_2015}.
-The last of these, the particle filter, was chosen for several reasons,
-discussed in Section~\ref{sec:pf}.
+and a particle filter \citep{Hans_2015}.
+The last of these was chosen for several reasons
+which are discussed in Section~\ref{sec:pf}.
 
-In order to improve the reliability of ETAs,
-congestion along intermediate roads should be incorporated into the predictive model.
-One solution is to use the travel time of preceeding buses along the same roads
-to predict travel times for the current bus,
-such as proposed by \cite{Yu_2010},
-who were able to improve prediction accuracy
-by using the travel time of vehicles on the same \emph{route}.
-However, this only works well for high-frequency routes, so
-\cite{Yu_2011} modeled travel times along roads where several major routes converged,
-which showed further improvement however is limited in scope.
-In general, however, no comprehensive network modeling approach has been proposed us
-solely GPS position data to estimate arrival times
-while accounting for congestion.
+The next step is to construct a method of estimating traffic conditions.
+This can be done by using the travel time of preceeding buses along the road
+to predict travel times for the current bus.
+\cite{Yu_2010} obtained improved prediction accuracy by using travel times
+from preceeding buses servicing the same route.
+\cite{Hans_2015} used a similar approach,
+instead referring to headway, the time between consecutive vehicles
+at a point on the route.
+However, this approach only works well for high frequency routes.
+\cite{Yu_2011} showed further improvements by combining travel times 
+from several routes which converged,
+but is limited to predefined, converging routes.
+In general, however, no comprehensive network modeling approach has been proposed using
+solely GPS position data to accounting for congestion when estimating arrival times.
+We describe a simple method of constructing a road network
+from GTFS data (section~\ref{sec:gtfs}),
+which allows all transit vehicles traveling along a road to contribute to tis state,
+regardless of which route they are serving.
 
@@
 model which uses the vehicles traveling throughout the transit road network
-to update congestion information in \rt.
+to update congestion information in real-time.
 As the vehicle model is implemented using a particle filter,
+a computationally intensive method,
 we demonstrate its faesibility as a valid real-time option.
-From the particle filter estimates of vehicle trajectories,  
-The particle filter provides estimates of vehicle travel times along roads,
+The vehicle model provides estimates of travel times along roads,
 which can be used to update the road network's state.
-Future work involves using this travel time information 
+Future work is planned to use this travel time information 
 in combination with historical data to develop an arrival time prediction model.

### 2018-10-31 anzjs2018/sections/02_gtfs.tex
More edits

@@
 is representative of the real-world travel time between stops. 
-In many places in Auckland, 
-traffic congestion is highly dynamic and changes from day to day,
-so even a well-calibrated schedule is an unreliable baseline.
+Therefore, an alternative estimation approach using \rt travel times 
+between stops is more adequate.
 
@@
 which increases the quantity of data available for each road.
-Then, each route can instead be represented as a sequence of road segments,
-each with a travel time state.
+Each route is then represeted as a sequence of road segments,
+each with an associated travel time.
 
-To construct the network,
-we use stops as nodes, so that
-routes with a common sub-sequence of stops will share the roads between them;
-roads are then defined as the links between stops,
-as demonstrated in Figure~\ref{fig:network_creation}.
-In section~\ref{sec:kf} we present a model for estimating vehicle travel time
-along a road segment.
-There are several places where this doesn't quite work, 
-for example express routes which do not service certain stops,
-and multi-stop locations,
-which results in overlapping road segments.
-Future work will look at using intersection points of the route shapes
-as nodes in the network to improve the coverage and reduce overlap.
+The transit network is constructed by splitting each route
+into spatially identifiable segments,
+each representative of a physical road.
+The simplest way to do this is to use bus stops as nodes in the network
+(Figure~\ref{fig:network_creation_1}),
+and connecting roads as edges.
+In this way, routes that serice the same subsequence of stops
+will all contribute to congestion information for the connecting roads,
+as shown in Figure~\ref{fig:network_creation_2}.
+There are several drawbacks to this method,
+for example routes merge at physical intersections, not bus stops,
+so some overlap will occur as shown.
+However, this is minimal, and our simplification is sufficent for the current work.
 
@@
 
-One important consideration regarding the Auckland Transport's \rt implementation is that
+One important consideration regarding Auckland Transport's \rt implementation is that
 buses are programmed to report their location when arriving at or departing from
@@
 To further complicate this,
-these positions can be preemtive,
+these positions can be pre-emtive,
 (i.e., the bus is almost there, but not quite),
@@
 if this has decreased, the current state is rejected and the vehicle reverted
-to its previous state become continuing.
+to its previous state before continuing.
 This way, we avoid degeneration of the particle fitler (Section~\ref{sec:model_perf})

### 2018-10-31 anzjs2018/sections/03_models.tex
More edits

@@
 
-Real-time tracking applications such as ours often use
+Real-time tracking applications often use
 a Bayesian filtering approach,
@@
 in which the data is processed sequentially instead of all at once,
-reducing computational demands.
+reducing computational demands
+\citep{Anderson_2012}.
 These models assume an underlying Markov process with state $\mathcal{X}_k$
@@
 
-We chose to employ a particle filter to estimate $\bX_k$,
+We use a particle filter to estimate $\bX_k$,
 due to do its high flexibility and use in recent 
@@
 which is a common feature of the proposal distribution, particularly around bus stops.
-Another advantage is the intuitive likelihood function (see section~\ref{sec:pf_update}).
+Another advantage is the intuitive likelihood function, 
+discussed in section~\ref{sec:pf_update}.
 Conversely, particle filter methods are computationally demanding,
@@
 \end{equation}
-falls below a specified threshold $N_{\text{thres}}$.
+falls below a specified threshold $N_{\text{thres}}$
+\citep{}.
 The resampling is performed with replacement using importance weights $W_k^{(i)}$,
@@
 
-The particle filter allows us to develop a model of several important bus behaviours:
+The particle filter allows us to model several important bus behaviours:
 \begin{itemize}
@@
 using system noise parameter $\sigma^2$ which describes 
-how the vehicle's acceleration changes as a Markov process.
+how the vehicle's speed changes as a Markov process.
 The vehicle does not travel constantly along the route,
@@
 
+The transition function algorithm is shown in Algorithm~\ref{fig:algorithm},
+which implements those features discussed above.
 The stop positions are represented by their distance into trip,
@@
 given a distance traveled and set of stop distances.
-The transition function algorithm is shown in Algorithm~\ref{fig:algorithm},
-which implements those features discussed above.
 Lines~2--9 are necessary to decrease the degeneration rate of the filter,
@@
 it has a chance of stopping (lines~18--19),
-in which case the dwell time is sampled and deducted from the 
+in which case the dwell time is sampled and subtracted from the 
 time remaining to travel (lines~20--21).
@@
 at each point where the vehicle travels (lines 9 and 26),
-the travel time for that segment, $z_j$, is incremented.
+the travel time for the current segment, $z_j$, is incremented.
 The mutation ends once the travel time $\delta$ reaches zero,
 or when the vehicle reaches the end of the route (lines 10, 14).
-Once finished, the particle's new state is returned (line 29).
+Once complete, the particle's new state is returned (line 29).
 
@@
 \end{equation}
-Now, the geographical distance between the observation and the true vehicle can be expressed
+The geographical distance between the observation and the true vehicle can be expressed
 in terms of two independent random variables $z_1, z_2 \sim \mathrm{N}(0,1)$
@@
 The likelihood of the data given a particle's state estimate 
-cane be calculated using (\ref{eq:obs_exp})
+is calculated using (\ref{eq:obs_exp})
 \begin{equation}
@@
 as well as make short-term predictions for estimating arrival times.
-In this paper, we develop simple model in which each road is treated independently.
-Future work will investigate using correlations with adjacent roads and historical data
+In this paper, we develop a simple model in which each road is treated independently.
+Future work will explore correlations between adjacent roads and using historical data
 to improve the model and subsequent predictions.
@@
 of transit vehicles along road segment $\ell$ at time $t_c$,
-and is assumed to be constant with system noise $v_c^\ell \sim \mathrm{N}(0, \nu^2)$.
+and is assumed to be constant over time 
+with system noise $v_c^\ell \sim \mathrm{N}(0, \nu^2)$.
 Observations of the state, $Z_c^\ell$, are obtained from the particle filter in Section~\ref{sec:pf},
 and have measurement error $e_c^\ell \sim \mathrm{N}(0, R_c^\ell)$ as estimated 
-using uncertainty of particle estimates.
+using the uncertainty of particle estimates.
 The model from (\ref{eq:rbe_model}) now reduces to
@@
 
-Since multiple vehilces can travel along a road at the same time,
+Since multiple vehicles can travel along a road simultaneously,
 we have used an information filter to implement the network model.
@@
 and uncertainty $P_c^\ell = \mathrm{var}(\theta_c^\ell)$,
-which are predicted from the previous state estimate using the prediction
-equations
+which are predicted from the previous state estimate using the predictive model
 \begin{align*}
@@
 and is transformed to a measurement information covariance matrix 
-$B_c^{m\ell} = (s^{m\ell}_c)^{-2}$
+$B_c^{m\ell}~=~(s^{m\ell}_c)^{-2}$
 and measurement information vector $b_c^{m\ell}~=~\bar z^{m\ell}_c (s^{m\ell}_c)^{-2}$.
-The total information is the sum of the information over all $M$ vehicles
+The total information is the sum of the information over all $M_c^\ell$ vehicles
 that traversed segment $\ell$ since the last update,
@@
 \begin{align*}
-U^\ell_{c|c} &= U^\ell_{c|c-1} + \sum_{m=1}^M B^{m\ell}_{c} \\
-\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^M b^{m\ell}_{c}.
+U^\ell_{c|c} &= U^\ell_{c|c-1} + \sum_{m=1}^{M_c^\ell} B^{m\ell}_{c} \\
+\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^{M_c^\ell} b^{m\ell}_{c}.
 \end{align*}
-The desired parameter estimates (tavel time) are obtained 
+The desired parameter estimates are obtained 
 by the inverse transformations

### 2018-10-31 anzjs2018/sections/04_implementation.tex
More edits

@@
 
-The application consists of a
-static GTFS~schedule and network data component,
-and a \rt modeling and prediction component.
+The application consists of two components:
+the first handles importing static GTFS data and constructing the network,
+while the second implements the \rt models and prediction.
 We chose \verb+Rcpp+ to develop the program,
-providing access to other R packages for data manipulation 
+which provides access to other R packages for data manipulation 
 (notably \verb+RSQLite+ and \verb+dplyr+),
 as well as the speed and memory management capabilities of \verb|C++|. 
-The following has been implemented in the R package
+The program is implemented in the R package
 \verb+transitr+, available on Github (\url{https://github.com/tmelliott/transitr}).
-In this section, we discuss the features of the \rt component.
+In this section, we discuss the features of the \rt component
+and assess its performance in \rt,
+both with respect to timing and parameter estimation.
 
@@
 \item Load GTFS data from database
-\item Indefinitely repeat when new data recieved \ldots
+\item Each time new data are recieved \ldots
 \begin{enumerate}
-    \item Update or create new vehicle objects from new data
-    \item \textbf{Run particle filter on each vehicle to update state}
+    \item Update or create new vehicle objects from the new data
+    \item \textbf{Run particle filter on each vehicle to update or initialise state}
     \item \textbf{Update state of any roads for which vehicles 
@@
 with a target of 30~seconds or faster at peak time.
-Using C++ provides enough control over memory management
-to ensure the program runs fast enough in \rt 
-while still implementing the models.
-We also use OpenMP for parallelisation,
-allowing us to process multiple vehicles simultaneously,
-which are each independent so threadsafety is a non-issue.
+Using C++ provides the memory management control necessary to make the application
+viable in real-time.
+We also make use of OpenMP for parallisation,
+since each vehicle is modelled independently so it is trivial to scale up without
+worrying about threadsafety.
 
-The particle filter requires thousands of particles per vehicle,
+The number of particles needed per vehicle 
+depends on many factors,
 so it is necessary to explore the performance of the application
 with varying number of particles.
-We also assess the performance of the particle filter
-for a range of fixed model parameters values,
+Application performance is also assessed
+for a range of fixed model parameter values,
 such as system noise and GPS error.
-To do so, we implemented a simulated \rt environment
-in which we could analyse the same subset of real vehicle data from 8~October, 2018
-using a range of settings.
+To enable comparisons, we implemented a simulated \rt environment
+in which the same subset of real vehicle data from 8~October, 2018
+could be analysed using a range of settings.
 These simulations were carried out of a Virtual Machine 
@@
 
-% \begin{figure}[tb]
-%     \centering
-%     % \includegraphics[width=0.7\textwidth]{figures/04_model_results_gps.pdf}
-%     \caption{The distance between the observation and the nearest point on the shape.}
-%     \label{fig:gps_dist}
-% \end{figure}
-
 \subsection{Program Timings}
@@
 
-During development of our application,
-our primary concern was the computational faesibility of a particle filter.
-For each iteration, 
-the timings of the individual components are recorded.
+In each iteration, 
+the timing of the various program components is recorded.
 Since the number of vehicles traveling at any given time changes throughout the day,
-we used the average timings over a 15~minute window to compare.
+we used the average timings over an off-peak 15~minute window.
+Figure~\ref{fig:timings} shows the average timings for 
+varying numbers of particles, $N$.
 
-Figure~\ref{fig:timings} shows the timing results for varying number of particles.
-We see that there is a very slight quadratic trend to run times,
-as several steps are not of complexity $O(N)$,
-the main one being the estimation of ETA quantiles.
-While not discussed in this paper,
-the application generates ETAs for each particle
-and calculates ETA quantiles for each vehicle to demonstrate the computational faesibility.
-This step requires sorting a vector, which is $O(N\log(N))$.
-Clearly, our initial target of 30~seconds is obtainable given the number of particles
-is not too large, and there are enough cores.
+The most time consuming component is ETA writing,
+which involves summarising the individual ETAs estimated for each particle.
+While not discussed here this involves quantiles which use a sorting algorithm,
+which has complexity $O(N \log N)$ 
+which explains the slightly quadratic trend in the timings.
+
+The next most intensive step is vehicle updating,
+and takes about 5~seconds for 8000 particles.
+The ETA prediction step only takes a few seconds,
+although this may increase once a more comprehensive model is developed.
+On our 8-core virtual machine the entire process is completed within 20~seconds
+when 8000~particles are used.
+In the next section we explore the effect of changing $N$
+to allow finding the optimal number.
 
@@
 
-The model was evaluated using several statistics,
-each calculated using a range of parameter values.
-The first is effective sample size,
-defined in (\ref{eq:neff}),
-which is used to determined if resampling is required.
-Another important statistic is the \emph{degeneration rate},
-or the percent of iterations where the particle filter
-is unable to find any plausible states given the observation,
-and the vehicle is reinitialised.
-Note, however, that some observations are \emph{invalid trips},
-in which the vehicle is logged in to the wrong trip,
-or ``off duty'' while still transmitting for a trip.
-Some of these can be filtered for being excessively far from the route,
-but others---for example traveling the wrong direction---cannot be.
-The last statistic we compute is the variance of road segment travel times
-over time.
-For this example, we varied the number of particles $N$,
-GPS error $\epsilon$, and system noise $\sigma^2$.
+For each simulation, 
+the model was evaluated using several statistics.
+These were effective sample size, $N_\text{eff}$ calculated using (\ref{eq:neff});
+degeneration rate, the percentage of iterations in which the vehicle was lost
+(no plausible particles);
+and the variance of segment travel time estimates.
 
-The one parameter we can investigate from the data is GPS error, $\epsilon^2$,
-by looking at the distance between observations and the shape.
-Figure~\ref{fig:dist_to_route} shows the distribution of distance between the vehicle
-observation and the \emph{nearest point on the path}.
-We see two peaks,
-one at about 1m, and another at 2.5m.
-The first of these is possibly an artefact of the data,
-for example automated waypoints at intersections.
-We have explored a range of GPS error values from 1 to 5~m.
+The GPS error parameter values were chosen by examining the distribution
+of the distance between observations and the nearest point on the route.
+This is shown in Figure~\ref{fig:dist_to_route},
+in which we see a peak at about 1m and another at 2.5m,
+so we have chosen GPS error values of $\epsilon \in \{1,2,3,5\}$.
+System error is the variation of speed per second,
+or acceleration rate,
+so our chosen values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$ 
+correspond to a variance of 0.09, 0.9, 9, and 45 meters per second over 
+30~seconds.
 
@@
 
-For the effective sample size, 
-increasing GPS error and system error both tend to
-increase $N_\text{eff}$, as shown in Figure~\ref{fig:degen_rate}.
-Increasing the GPS error distributes the weights more evenly,
-while increasing system noise allows more particles to reach
-the necessary speed,
-so these are as expected.
+From Figure~\ref{fig:degen_rate}, we see that $N_\text{eff}$ increases
+with GPS error and decreases with system noise.
+The degeneration rate decreases with GPS error and increases with $N$,
+but is relatively unaffected by system noise.
+Large GPS error affects the likelihood,
+and gives more weight to particles farther from the observation
+than does a smaller GPS error.
+Conversesly, increasing system noise spreads out the particle cloud,
+so fewer particles will be near the vehicle.
+Thus, these results are not surprising, 
+but show that the model is working as expected.
 
@@
 
-The degeneration rate, 
-shown on the $y$-axis of Figure~\ref{fig:degen_rate},
-decreases as GPS error increases,
-as anomolous observations where the bus is far from the route
-are more likely.
-It also decreases with the number of particles,
-while system noise has no apparent effect.
-
-Lastly is the uncertainty of travel time estimates,
-as shown in Figure~\ref{fig:travel_times}.
-System noise again shows no effect on the variability of travel time estimates,
-however increasing GPS error tends to increase travel time variability.
-This means that there is a trade-off between the models performance
-and how well it estimates the main parameter of interest:
-larger values of $\epsilon$ result in less degeneration and larger $N_\text{eff}$,
-at the cost of more variable travel time estimates.
+The last measure of performance is the variance of segment travel times.
+This is calculated by \_\_\_.
+Figure~\ref{fig:travel_times} shows that variance increases with GPS error,
+and decreases with $N$,
+with little effect made by system noise.
+There is therefore a tradeoff between particle filter performance
+($N_\text{eff}$ and degeneration rate) and parameter estimation,
+although the true test will be calculating the prediction error
+of ETAs once the prediction model has been implemented.
 

### 2018-10-31 anzjs2018/sections/06_discuss.tex
More edits

@@
 with future work planned to explore the other parameters---dwell time and stopping probabilities---%
-as well as to extend the model to also account for intersections.
-The travel times estimated from the particle filter then allow us to model \rt road state.
+as well as to extend the model to also account for physical intersections.
+The travel times estimated from the particle filter then allow modeling of
+\rt road state.
 
-The network model presented here is of the simplest nature,
+The network model presented here is simple,
 but provides the foundation for future work which will involve
-deveoping a formal model that combines \rt information with
-historical data, allowing predictions of future road state to be more accurate,
+developing a formal model that combines \rt information with
+historical data, allowing for more accurate predictions of future road state,
 for example incorporating peak time congestion forecasts.
-This in turn provides the basis for an arrival time prediction model,
-which will be a combination of vehicle state and road state forecasts.
+This in turn will provide the basis for an arrival time prediction model,
+a combination of vehicle state and road state forecasts.
 

### 2018-11-01 anzjs2018/sections/01_introduction.tex
edits

@@
 and potentially make decisions on the best route to take.
+It has also been shown that perceived waiting time is less
+when arrival time information is available \citep{TCRP_2003b}.
 Unfortunately, here in Auckland, RTI is highly unreliable,
@@
 
-To avoid the problems associated with the arrival time predictions
-based solely on reported delays,
-a more robus modeling and prediction framework is required.
+To avoid the problems discussed above,
+a more robust modeling and prediction framework is required.
 This should consist of a robust vehicle model to estimate the position and speed

### 2018-11-01 anzjs2018/sections/02_gtfs.tex
edits

@@
     \end{subfigure}
-    \caption{Construction of a route network involves combining routes at nodes %
-        (here we are using stops) which are connected by edges (roads). %
-        In (a), the three unique routes, represented by different line types, clearly %
-        overlap in several places. In (b) these have been merged, and the width of each line %
-        represents how many routes use that link.}
+    \caption{
+        Constructing a route network from the stops (dots) of three separate
+        routes (lines), as shown in (a).
+        Routes using the same subsequence of stops are assigned the same
+        road segments, so the road network shown 
+        consists of nodes (stops) and edges (roads),
+        as shown in (b), where the thickness of each lines
+        corresponds to how many routes share that road.
+    }
+    % Construction of a route network involves combining routes at nodes %
+    %     (here we are using stops) which are connected by edges (roads). %
+    %     In (a), the three unique routes, represented by different line types, clearly %
+    %     overlap in several places. In (b) these have been merged, and the width of each line %
+    %     represents how many routes use that link.}
     \label{fig:network_creation}

### 2018-11-01 anzjs2018/sections/03_models.tex
edits

@@
 
-\begin{figure}[tb]
+\begin{figure}[p]
     \centering
-    \begin{subfigure}[t]{0.48\textwidth}
+    \begin{subfigure}[t]{0.9\textwidth}
         \centering
-        \includegraphics[width=\textwidth]{figures/03_particle_filter_1.pdf}
-        \caption{The vehicle state represented by a sample of discrete points is related to the observable state (GPS position) through the transition function $h$.}
+        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_1.pdf}
+        \caption{
+            Vehicle state is mapped to observation space by the
+            measurement function $h$.   
+        }
+        % The vehicle state represented by a sample of discrete points is related to the observable state (GPS position) through the transition function $h$.}
         \label{fig:pf_state_prev}
-    \end{subfigure}\;\;
-    \begin{subfigure}[t]{0.48\textwidth}
+    \end{subfigure}\\
+    \begin{subfigure}[t]{0.9\textwidth}
         \centering
-        \includegraphics[width=\textwidth]{figures/03_particle_filter_2.pdf}
-        \caption{The transition function $f$ models the behaviour of each particle,
-            including changes in acceleration and stopping behaviour at bus stops,
-            which can result in multiple modes.}
+        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_2.pdf}
+        \caption{
+            The transition function $f$ predicts the future state 
+            of each particle.
+        }
+            % The transition function $f$ models the behaviour of each particle,
+            % including changes in acceleration and stopping behaviour at bus stops,
+            % which can result in multiple modes.}
         \label{fig:pf_state_predict}
     \end{subfigure}\\
-    \begin{subfigure}[t]{0.48\textwidth}
+    % \begin{subfigure}[t]{0.48\textwidth}
+    %     \centering
+    %     \includegraphics[width=\textwidth]{figures/03_particle_filter_6.pdf}
+    %     \caption{Each particle's distance from the true observation (red cross) 
+    %         is calculated using
+    %         the geographic distance between two coordinates, which can then be used
+    %         to calculate it's weight.}
+    %     \label{fig:pf_state_update}
+    % \end{subfigure}\;\;
+    \begin{subfigure}[t]{0.9\textwidth}
         \centering
-        \includegraphics[width=\textwidth]{figures/03_particle_filter_6.pdf}
-        \caption{Each particle's distance from the true observation (red cross) 
-            is calculated using
-            the geographic distance between two coordinates, which can then be used
-            to calculate it's weight.}
-        \label{fig:pf_state_update}
-    \end{subfigure}\;\;
-    \begin{subfigure}[t]{0.48\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figures/03_particle_filter_4.pdf}
-        \caption{Resampling occurs using likelihood weights that use the distance
-            between the particle's location and the vehicles reported location (red cross),
-            resulting in the updated state.}
+        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_4.pdf}
+        \caption{
+            The state is updated by weighted resampling based on distance
+            of particle to observation.
+        }
+        % Resampling occurs using likelihood weights that use the distance
+        %     between the particle's location and the vehicles reported location (red cross),
+        %     resulting in the updated state.}
         \label{fig:pf_state_predict2}
     \end{subfigure}
-    \caption{The vehicles state is estimated using a set of discrete points, or particles,
-        each of which is independently mutated to make predictions of the future state.
-        Multimodality is easily handled, as demonstrated here,
-        without degeneration of the filter.}
+    \caption{
+        The particle filter approximates vehicle state using a set of 
+        discrete points (a), which are each mutated indepently (b)
+        to predict the next state.
+        The measurement function $h$ allows state particle state
+        to be mapped to a GPS coordinate (a),
+        which can be compared to the observed location (c)
+        to calculate each particles likelihood and perform resampling.
+    }
+        % The vehicles state is estimated using a set of discrete points, or particles,
+        % each of which is independently mutated to make predictions of the future state.
+        % Multimodality is easily handled, as demonstrated here,
+        % without degeneration of the filter.}
     \label{fig:pf_state}
 \end{figure}
+
+\afterpage{\clearpage}
 

### 2018-11-01 anzjs2018/sections/04_implementation.tex
edits

@@
     \includegraphics[width=0.7\textwidth]{figures/04_model_results_timing.pdf}
-    \caption{The timings for various parts of the application, and overall. %
-        The trend is slightly non-linear due to the complexity of the ETA generation step, %
-        which requires sorting of the particle ETAs to obtain quantile estimates.}
+    \caption{
+        Application timings were recorded for each component of each
+        iteration, and averages over a 15~minute window are shown here.
+        The timings shown are \emph{wall time}, the actual time passed,
+        not CPU time which would be approximately 8~times larger 
+        (for using 8~cores).
+        Steps that only involve particle calculations are approximately
+        of complexity $O(N)$,
+        however some steps are a little more complex, 
+        so there is a slightly exponential trend.
+    }
+        % The timings for various parts of the application, and overall. %
+        % The trend is slightly non-linear due to the complexity of the ETA generation step, %
+        % which requires sorting of the particle ETAs to obtain quantile estimates.}
     \label{fig:timings}
@@
 
-For each simulation, 
-the model was evaluated using several statistics.
+To evaluate the performance of the model,
+the simulation was run using a range of values of GPS error, $\epsilon$,
+and system noise, $\sigma^2$.
+Figure \ref{fig:dist_to_route} shows the distribution of the distance
+between each observation and the route using a nearest point algorithm,
+so we chose to use GPS error values $\epsilon \in \{1,2,3,5\}$. 
+For the system noise we use values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$,
+which correspond to the average vehicle speed having a variance between 0.1 and 45~meters per second
+over a period of 30~seconds.
+For each of the combination of these parameter values, along with varying $N$,
+several values were computed.
 These were effective sample size, $N_\text{eff}$ calculated using (\ref{eq:neff});
@@
 
-The GPS error parameter values were chosen by examining the distribution
-of the distance between observations and the nearest point on the route.
-This is shown in Figure~\ref{fig:dist_to_route},
-in which we see a peak at about 1m and another at 2.5m,
-so we have chosen GPS error values of $\epsilon \in \{1,2,3,5\}$.
-System error is the variation of speed per second,
-or acceleration rate,
-so our chosen values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$ 
-correspond to a variance of 0.09, 0.9, 9, and 45 meters per second over 
-30~seconds.
-
 \begin{figure}[tb]
@@
     \includegraphics[width=0.7\textwidth]{figures/04_model_results_dist.pdf}
-    \caption{The distance between the observation and the nearest point on the shape.}
+    \caption{
+        The distance between the observation and the nearest point on the shape
+        is calculated by finding the shortest distance between point and path.
+        This gives an approximate idea of GPS error, however it should be noted
+        that many observations are made automatically at waypoints,
+        which accounts for the peak at 1m.
+        Note also that these observations have been truncated at 20m.
+    }
     \label{fig:dist_to_route}
@@
     \centering
+    \includegraphics[width=0.9\textwidth]{figures/04_model_results_neff.pdf}
+    \caption{
+        The effective sample size is an approximate measure of the variance
+        of the particle weights.
+        When all particles have equal weights, $N_\text{eff} = N$,
+        while if all the weight is on one particle, $N_\text{eff} = 1$. 
+        The figure is subset by system noise, which shows how $N_\text{eff}$ 
+        decreases as system noise increases,
+        while increasing GPS error (on the $x$-axis) increases $N_\text{eff}$.
+        Increasing $N$ has a negative effect on $N_\text{eff}$ but this is
+        only evident when $N_\text{eff}$ is larger.
+        The error bars show $\pm 2$SE.
+    }
+    \label{fig:neff}
+\end{figure}
+
+\begin{figure}[tb]
+    \centering
     \includegraphics[width=0.9\textwidth]{figures/04_model_results_degen.pdf}
-    \caption{The effective sample size and degeneration rate for a range of parameter 
-        values. The columns are of increasing number of particles,
-        and the rows are for increasing (downwards) GPS error.}
+    \caption{
+        The degeneration rate is the percentage of iterations where there were no
+        plausible particles, so the vehicle was ``lost'' by the model.
+        This decreases as GPS error and number of particles increase,
+        but there is no visible effect of system noise.
+        The error bars show $\pm 2$SE.
+    }
     \label{fig:degen_rate}
@@
 
-The last measure of performance is the variance of segment travel times.
-This is calculated by \_\_\_.
-Figure~\ref{fig:travel_times} shows that variance increases with GPS error,
-and decreases with $N$,
-with little effect made by system noise.
-There is therefore a tradeoff between particle filter performance
-($N_\text{eff}$ and degeneration rate) and parameter estimation,
-although the true test will be calculating the prediction error
-of ETAs once the prediction model has been implemented.
+The last measure of performance is the relative variance of segment travel times.
+This is calculated by computing the average variance of travel times over all simulations,
+and then for each simulation, 
+taking the average (over road segments) of the ratio between variance for that simulations
+and variance for all simualtions.
+Figure~\ref{fig:travel_times} shows that simulations with larger GPS error
+result in a higher relative variance,
+while simulations with more particles have lower relative variance.
+There is no visible effect of system noise on travel time variation.
+
+The results of the simulations
+demonstrate a tradeoff between particle filter performance
+($N_\text{eff}$ and degeneration rate) and parameter estimation.
+However, until the arrival time prediction model has been completed,
+it is difficult to make decisions about the optimal values to use:
+we are unable to say whether a higher variation of travel time estimates
+is going to have a significant effect on the final ETAs,
+particularly when compared to the uncertainty of forecasting road state.
 
@@
     \centering
-    \includegraphics[width=0.9\textwidth]{figures/04_model_results_times.pdf}
-    \caption{The within-segment posterior travel time uncertainty (standard error), subset by number of particles.}
+    \includegraphics[width=\textwidth]{figures/04_model_results_times.pdf}
+    \caption{
+        The variability of travel time within a segment increases with GPS error,
+        and decreases slightly with increasing number of particles.
+        There is no visible effect of system noise on travel time variability.
+        The error bars show $\pm 2$SE.
+    }
     \label{fig:travel_times}

### 2018-11-01 anzjs2018/sections/06_discuss.tex
edits

@@
 that enables the \rt estimation of road state from vehicle position data,
-which will in future be used to make what will hopefully be more reliable 
-arrival time predictions.
+which can later be used to make arrival time predictions
+that account for \rt traffic conditions.
 The vehicle model, implemented using a particle filter,
 allows complex bus behaviours to be modeled,
-with future work planned to explore the other parameters---dwell time and stopping probabilities---%
-as well as to extend the model to also account for physical intersections.
-The travel times estimated from the particle filter then allow modeling of
-\rt road state.
+with future work planned to extend the model to handle physical intersections,
+such as traffic lights.
+From the particle filter,
+vehicle travel times along roads are estimated which are in turn used 
+in the modeling of the road network state.
 
 The network model presented here is simple,
-but provides the foundation for future work which will involve
-developing a formal model that combines \rt information with
-historical data, allowing for more accurate predictions of future road state,
-for example incorporating peak time congestion forecasts.
-This in turn will provide the basis for an arrival time prediction model,
-a combination of vehicle state and road state forecasts.
-
-The primary purpose of this paper was to present our approach and 
-explore and demonstrate tis faesibility as a \rt option for RTI in transit systems.
-We have been able to construct an application that runs in \rt and capable
-of producing estimates within 30~seconds---although this can be reduced 
-by using fewer particles or, if necessary, increasing the number of cores.
+but provides the foundation for future work to
+devel a formal model combining \rt travel time information with historical data, 
+allowing short- and medium-term forecasts of road state,
+for example incorporating peak time congestion.
+This in turn will provide the basis for an arrival time prediction model
+that combines \rt vehicle state with travel time forecasts
+and provide commuters with reliable ETAs.
+These will likely consist of a point estimate along with a prediction interval,
+as it is impossible given the available data to accurately predict arrival time
+due to traffic lights, unknown dwell times at intermediate bus stops,
+and many other factors.
 

### 2018-11-01 anzjs2018/figures/04_model_results.R

@@ -20,6 +20,7 @@ times <- do.call(bind_rows, lapply(list.files("../../transitr/simulations", patt
 date <- format(times$timestamp[1], "%Y-%m-%d")
 trange <- as.POSIXct(paste(date, c("13:30", "14:00")))
 
+olvls <- c("all", "predicting ETAs", "updating vehicle states", "writing ETAs to protobuf feed")
 ttot <- times %>% group_by(sim, iteration) %>%
     summarize(cpu = sum(cpu), wall = sum(wall), timestamp = first(timestamp), nvehicles = first(nvehicles),
         n_particles = first(n_particles), gps_error = first(gps_error), system_noise = first(system_noise)) %>%
@@ -30,15 +31,16 @@ tsmry <- times %>% bind_rows(ttot) %>%
     summarize(cpu = mean(cpu), wall = mean(wall),
         n_particles = first(n_particles), gps_error = first(gps_error), system_noise = first(system_noise)) %>%
     ungroup() %>%
-    filter(what %in% c("all", "predicting ETAs", "updating vehicle states", "writing ETAs to protobuf feed"))
+    filter(what %in% olvls)
 
 # ggplot(tsmry) +
 #     geom_point(aes(n_particles, cpu, color = what)) +
 #     ylab("Wall time (seconds)")
 
-lvls <- c("updating vehicle states", "predicting ETAs", "writing ETAs to protobuf feed", "all")
-tdat <- bind_rows(times, ttot) %>% filter(what %in% lvls) %>%
-    mutate(what = factor(what, levels = lvls)) %>%
+ordlvls <- olvls[c(3,2,4,1)]
+lvls <- c("Updating Vehicle States", "Predict ETAs", "Write ETAs to file", "Total")
+tdat <- bind_rows(times, ttot) %>% filter(what %in% olvls) %>%
+    mutate(what = factor(what, levels = ordlvls, labels = lvls)) %>%
     group_by(what, n_particles)
 
 
@@ -64,13 +66,15 @@ pdf('figures/04_model_results_timing.pdf', width = 6, height = 3)
 # dev.off()
 
 ggplot(tsmry %>% group_by(what, n_particles) %>% 
-        summarize(wall.hat = mean(wall/1000), wall.var = sd(wall/1000))) +
+        summarize(wall.hat = mean(wall/1000), wall.var = sd(wall/1000)) %>%
+        ungroup() %>%
[…]

### 2018-11-01 anzjs2018/anzjs2018.tex
edits
run spellcheck

@@
 
-\doublespace
+% \doublespace
+\usepackage{afterpage}
 
@@
 Predicting the arrival time of a transit vehicle involves not only
-knowledge of its current position and schedule adherance,
+knowledge of its current position and schedule adherence,
 but also of traffic conditions along the remainder of the route.
@@
 later be used for arrival time prediction.
-We demonstrate the \rt faesibility of such an approach in \rt,
+We demonstrate the \rt feasibility of such an approach in \rt,
 and detail our implementation in R and C++.

### 2018-11-01 anzjs2018/sections/01_introduction.tex
run spellcheck

@@
 To avoid the problems discussed above,
-a more robust modeling and prediction framework is required.
+a more robust modelling and prediction framework is required.
 This should consist of a robust vehicle model to estimate the position and speed
@@
 into the arrival time predictions.
-Several vehicle modeling approaches were explored, 
+Several vehicle modelling approaches were explored, 
 including the Kalman filter \citep{Dailey_2001,Cathey_2003},
@@
 The next step is to construct a method of estimating traffic conditions.
-This can be done by using the travel time of preceeding buses along the road
+This can be done by using the travel time of preceding buses along the road
 to predict travel times for the current bus.
 \cite{Yu_2010} obtained improved prediction accuracy by using travel times
-from preceeding buses servicing the same route.
+from preceding buses servicing the same route.
 \cite{Hans_2015} used a similar approach,
@@
 but is limited to predefined, converging routes.
-In general, however, no comprehensive network modeling approach has been proposed using
+In general, however, no comprehensive network modelling approach has been proposed using
 solely GPS position data to accounting for congestion when estimating arrival times.
@@
 from GTFS data (section~\ref{sec:gtfs}),
-which allows all transit vehicles traveling along a road to contribute to tis state,
+which allows all transit vehicles travelling along a road to contribute to its state,
 regardless of which route they are serving.
@@
 In this paper, we describe a \rt vehicle and network state
-model which uses the vehicles traveling throughout the transit road network
+model which uses the vehicles travelling throughout the transit road network
 to update congestion information in real-time.
@@
 a computationally intensive method,
-we demonstrate its faesibility as a valid real-time option.
+we demonstrate its feasibility as a valid real-time option.
 The vehicle model provides estimates of travel times along roads,

### 2018-11-01 anzjs2018/sections/02_gtfs.tex
run spellcheck

@@
 \item \emph{routes}, a sequence of two or more stops displayed as a single service;
-\item \emph{trips}, an instance of a route occuring at a specific time of day;
+\item \emph{trips}, an instance of a route occurring at a specific time of day;
 \item \emph{schedules}, specifying the arrival (and departure) times for each bus at each of its stops; 
@@
 which is only available in a subset of the providers due to the requirement of 
-onboard GPS tracking devices and a central server.
+on board GPS tracking devices and a central server.
 It provides a standardised format for sharing vehicle positions and trip delays,
@@
 which increases the quantity of data available for each road.
-Each route is then represeted as a sequence of road segments,
+Each route is then represented as a sequence of road segments,
 each with an associated travel time.
@@
 and connecting roads as edges.
-In this way, routes that serice the same subsequence of stops
+In this way, routes that service the same subsequence of stops
 will all contribute to congestion information for the connecting roads,
@@
 so some overlap will occur as shown.
-However, this is minimal, and our simplification is sufficent for the current work.
+However, this is minimal, and our simplification is sufficient for the current work.
 
@@
 
-\subsection{Realtime vehicle locations}
+\subsection{Real-time vehicle locations}
 \label{sec:realtime_data}
@@
 To further complicate this,
-these positions can be pre-emtive,
+these positions can be preemptive,
 (i.e., the bus is almost there, but not quite),
-so consecutive observations can show a bus traveling backwards.
-To handle this, we compute the approximate distance traveled, $\tilde x_k$,
+so consecutive observations can show a bus travelling backwards.
+To handle this, we compute the approximate distance travelled, $\tilde x_k$,
 of the vehicle by finding the nearest point on the path to the observation;
@@
 to its previous state before continuing.
-This way, we avoid degeneration of the particle fitler (Section~\ref{sec:model_perf})
+This way, we avoid degeneration of the particle filter (Section~\ref{sec:model_perf})
 at the cost of losing one observation that is most likely invalid anyway.

### 2018-11-01 anzjs2018/sections/03_models.tex
run spellcheck

@@
 The underlying vehicle state at time $t_k$ consists of
-the vehicle's distance traveled $x$ in meters along the route and
+the vehicle's distance travelled $x$ in meters along the route and
 its speed $\dot x$ in meters per second.
@@
 due to do its high flexibility and use in recent 
-transit vehicle modeling applications \citep{Hans_2015}.
+transit vehicle modelling applications \citep{Hans_2015}.
 The primary advantage of a particle filter is its handling of multimodality,
@@
         The particle filter approximates vehicle state using a set of 
-        discrete points (a), which are each mutated indepently (b)
+        discrete points (a), which are each mutated independently (b)
         to predict the next state.
@@
 the state is updated by reweighting the particles using the likelihood $p(\bY_k | \bX_k^{(i)})$ 
-(sec.~\ref{sec:pf_update}) and standardising so $\sum_{i=1}^N W_k^{(i)} = 1$,
+(Section~\ref{sec:pf_update}) and standardising so $\sum_{i=1}^N W_k^{(i)} = 1$,
 \begin{equation*}
@@
 as demonstrated in Figure~\ref{fig:pf_state_predict2}.
-Afterwards, the weights are reinitialized to $W_k^{(i)} = N^{-1}$.
+Afterwards, the weights are reinitialised to $W_k^{(i)} = N^{-1}$.
 Note this is a computationally demanding step,
@@
 based on the values used by \cite{Hans_2015};
-future work will investigate modeling these in real-time similarly to Section~\ref{sec:kf}.
+future work will investigate modelling these in real-time similarly to Section~\ref{sec:kf}.
 
@@
 The $stop\_index$ function returns the index of a stop,
-given a distance traveled and set of stop distances.
+given a distance travelled and set of stop distances.
 Lines~2--9 are necessary to decrease the degeneration rate of the filter,
@@
     \Require $x_{k-1}, \dot x_{k-1}, \boldsymbol z$\\
-    Initialization $\delta \gets t_k - t_{k-1}$, 
+    Initialisation $\delta \gets t_k - t_{k-1}$, 
         $j \gets stop\_index(x, \boldsymbol{S})$
@@
         \State $\dot x \sim \mathrm{N}(\dot x, \sigma^2)$
-        \Comment{System noise, trunctated to be non-negative}
+        \Comment{System noise, truncated to be non-negative}
         \State $x \leftarrow x + \dot x$
@@
 surface (the geographical equirectangular projection, \cite{Snyder_1998}),
-allow the placement of a bivate normal likelihood on the data.
+allow the placement of a bivariate normal likelihood on the data.
 The model for the observations is,
@@
 It is worth noting that this representation of the likelihood is only
-possible due to the discretization of the state provided by the particle filter;
+possible due to the discrete nature of the particle filter state estimate;
 in other approaches, such as the Kalman filter,
@@
 the inverse measurement function,
-is required to estimate the ``observed'' distance traveled $\hat Z_k~=~h^{-1}(\bY_k)$, 
+is required to estimate the ``observed'' distance travelled $\hat Z_k~=~h^{-1}(\bY_k)$, 
 which introduces additional error and uncertainty into the model.

### 2018-11-01 anzjs2018/sections/04_implementation.tex
run spellcheck

@@
 \item Load GTFS data from database
-\item Each time new data are recieved \ldots
+\item Each time new data are received \ldots
 \begin{enumerate}
@@
         have completed travel}
-    \item Generate ETAs for vehicles uing combination of particle filter and network state
+    \item Generate ETAs for vehicles using combination of particle filter and network state
     \item Write ETAs to extended Google Protobuf binary file for distribution
@@
 During the development of the application,
-we were primarly concerned with ensuring each component of the program
+we were primarily concerned with ensuring each component of the program
 is as efficient as possible,
-allowing ETAs to be generated and distributed fast enough to be faesible in real-time,
+allowing ETAs to be generated and distributed fast enough to be feasible in real-time,
 with a target of 30~seconds or faster at peak time.
@@
 viable in real-time.
-We also make use of OpenMP for parallisation,
+We also make use of OpenMP for parallelisation,
 since each vehicle is modelled independently so it is trivial to scale up without
-worrying about threadsafety.
+worrying about thread safety.
 
@@
 the timing of the various program components is recorded.
-Since the number of vehicles traveling at any given time changes throughout the day,
+Since the number of vehicles travelling at any given time changes throughout the day,
 we used the average timings over an off-peak 15~minute window.
@@
         This gives an approximate idea of GPS error, however it should be noted
-        that many observations are made automatically at waypoints,
+        that many observations are made automatically at way points,
         which accounts for the peak at 1m.
@@
 than does a smaller GPS error.
-Conversesly, increasing system noise spreads out the particle cloud,
+Conversely, increasing system noise spreads out the particle cloud,
 so fewer particles will be near the vehicle.
@@
 taking the average (over road segments) of the ratio between variance for that simulations
-and variance for all simualtions.
+and variance for all simulations.
 Figure~\ref{fig:travel_times} shows that simulations with larger GPS error

### 2018-11-01 anzjs2018/sections/06_discuss.tex
run spellcheck

@@
 
-In this paper we have described an approach to transit vehicle modeling
+In this paper we have described an approach to transit vehicle modelling
 that enables the \rt estimation of road state from vehicle position data,
@@
 The vehicle model, implemented using a particle filter,
-allows complex bus behaviours to be modeled,
+allows complex bus behaviours to be modelled,
 with future work planned to extend the model to handle physical intersections,
@@
 vehicle travel times along roads are estimated which are in turn used 
-in the modeling of the road network state.
+in the modelling of the road network state.
 
@@
 but provides the foundation for future work to
-devel a formal model combining \rt travel time information with historical data, 
+develop a formal model combining \rt travel time information with historical data, 
 allowing short- and medium-term forecasts of road state,
@@
 
-% The main goal of this work was to explore the faesibility
+% The main goal of this work was to explore the feasibility
 % of a particle filter as a \rt estimation method for transit vehicle states,
@@
 % Once such a model has been developed, we plan to explore prediction estimates,
-% particuarly comparing point and intervals as a means of communicating 
+% particularly comparing point and intervals as a means of communicating 
 % uncertainty to commuters.

### 2018-11-01 anzjs2018/anzjs2018.tex
Clean up dir

@@
 \cleanlookdateon
-\runningheads{Modeling transit vehicles in real-time}{TOM~ELLIOTT AND THOMAS~LUMLEY}
-\title{A real-time particle filter for modeling travel times of transit vehicles through a GTFS-based road network}
+\runningheads{Modelling the travel time of transit vehicles}{TOM~ELLIOTT AND THOMAS~LUMLEY}
+\title{
+    Modelling the travel time of transit vehicles in \rt through a GTFS-based road network
+    using GPS vehicle locations
+}
+
+% Modelling transit vehicles to estimate the \rt traffic conditions in a public transport network}
 \author{Tom Elliott\corrauth~and~Thomas~Lumley}
@@
 (traffic conditions) can be estimated using \rt transit vehicle position data.
-We propose a framework which implements a vehicle model using a particle filter
-to estimate travel times along roads,
-used by a second model to estimate \rt traffic conditions which can
-later be used for arrival time prediction.
-We demonstrate the \rt feasibility of such an approach in \rt,
-and detail our implementation in R and C++.
+Our proposed framework implements a vehicle model using a particle filter
+to estimate road travel times,
+which are used in a second model to estimate \rt traffic conditions.
 Development and testing took place in Auckland, New Zealand,
-with future work planned to incorporate historical data into the network model
-to improve the reliability of ETAs.
+although attempts were made at generalising each component so 
+it will be compatible with other public transport systems around the world.
+We demonstrate the \rt feasibility and performance of our approach in real-time,
+where a combination of R and C++ was used to obtain the necessary performance results.
+Future work will use these estimated traffic conditions 
+in combination with historical data to obtain reliable
+arrival time predictions of transit vehicles.
 \end{abstract}

### 2018-11-01 anzjs2018/sections/01_introduction.tex
Clean up dir

@@
 keeps transport users informed about their journey,
-and potentially make decisions on the best route to take.
+and enabled them to make decisions about the best route.
 It has also been shown that perceived waiting time is less
@@
 are all too familiar with.
-One of these is inaccurate ETAs,
-which tend to either increase or decrease over time as the bus gets progressively later or earlier
-due to traffic congestion, or  because the schedule is poorly calibrated.
-Another is that buses are shown as \emph{on-time} if they are not present
-in the \rt system,
-regardless of whether this is due to technical problems,
-or the bus failing to show up;
-in either case, schedule times are used to determine ETAs and, 
-if the bus doesn't show up on time,
-the service is removed from the \rt board 
-and we get what passengers have described as ``Ghost Buses''.
+Inaccurate ETAs are a major source of frustration,
+as they tend to either increase or decrease over time 
+as the bus gets progressively later or earlier due to traffic congestion, 
+or simply due to poor schedule calibration.
+Furthermore, buses are shown as \emph{on-time} 
+if they are not present in the \rt system and have not been cancelled,
+so ETAs are based solely on scheduled arrival times;
+if the bus does not show up on time
+the service is removed from the \rt board,
+leaving passengers unsure as to when---if at all---their bus will show up,
+a phenomenon termed by bus users as ``ghost buses''.
 
@@
 where infrastructure for buses (such as priority lanes) is limited.
-This behaviour is exaggerated when there is a large distance between stops,
-as the ETA is only updated at stops so there can be a long time between updates.
 
-To avoid the problems discussed above,
+To avoid the problems mentioned above,
 a more robust modelling and prediction framework is required.
 This should consist of a robust vehicle model to estimate the position and speed
-of transit vehicles from \rt GPS data;
+of transit vehicles from \rt GPS data,
 and a way to incorporate road congestion information 
@@
 and a particle filter \citep{Hans_2015}.
-The last of these was chosen for several reasons
+We chose to use a particle filter for several reasons,
 which are discussed in Section~\ref{sec:pf}.
 
-The next step is to construct a method of estimating traffic conditions.
-This can be done by using the travel time of preceding buses along the road
-to predict travel times for the current bus.
-\cite{Yu_2010} obtained improved prediction accuracy by using travel times
-from preceding buses servicing the same route.
-\cite{Hans_2015} used a similar approach,
-instead referring to headway, the time between consecutive vehicles
-at a point on the route.
-However, this approach only works well for high frequency routes.
+The next step involves estimating traffic conditions throughtout the transport network
+using the information obtained from the GPS data. 
+\cite{Yu_2010} improved prediction accuracy by using the travel times
+of buses travelling along the same route.
+A similar method presented by \cite{Hans_2015}
+used headway, the time between consecutive vehicles at a point on the route,
+as a predictor of travel time.
+Since these approaches only work well on high frequency routes,
 \cite{Yu_2011} showed further improvements by combining travel times 
-from several routes which converged,
-but is limited to predefined, converging routes.
+from several routes;
+however, this was limited to predefined converging routes.
 In general, however, no comprehensive network modelling approach has been proposed using
-solely GPS position data to accounting for congestion when estimating arrival times.
-We describe a simple method of constructing a road network
-from GTFS data (section~\ref{sec:gtfs}),
-which allows all transit vehicles travelling along a road to contribute to its state,
-regardless of which route they are serving.
+solely GPS position data to model and account for congestion when estimating arrival times.
 
-In this paper, we describe a \rt vehicle and network state
-model which uses the vehicles travelling throughout the transit road network
-to update congestion information in real-time.
-As the vehicle model is implemented using a particle filter,
-a computationally intensive method,
-we demonstrate its feasibility as a valid real-time option.
-The vehicle model provides estimates of travel times along roads,
-which can be used to update the road network's state.
-Future work is planned to use this travel time information 
-in combination with historical data to develop an arrival time prediction model.
+In Section~\ref{sec:models} of this paper, 
+we describe the \rt vehicle and network state models
+used to estimate traffic congestion using the transit vehicles
+travelling through the network in real-time,
+and in Section~\ref{sec:rt} we demonstrate the feasibility of using a particle filter
+in \rt to model transit vehicles,
+and evaluate its performance.
+First, however, we descibe how we construct a
+road network from GTFS data in Section~\ref{sec:gtfs}.

### 2018-11-01 anzjs2018/sections/02_gtfs.tex
Clean up dir

@@
 \citep{GoogleDevelopers_2006}.
-The second component is \emph{GTFS realtime},
+The second component is \emph{GTFS-realtime},
 which is only available in a subset of the providers due to the requirement of 
@@
 there are some major issues with the current prediction method in Auckland.
-These are almost directly attributed to GTFS-realtime trip updates,
+These are almost directly attributed to \emph{GTFS-realtime} trip updates,
 which are currently the sole source of data for ETAs.
@@
 and includes the delay between the scheduled and actual arrival times,
-and is propagated to all future stops to adjust their ETAs.
-This assumes that the schedule is well calibrated and the time between scheduled arrivals
-is representative of the real-world travel time between stops. 
-Therefore, an alternative estimation approach using \rt travel times 
-between stops is more adequate.
+which is propagated to all future stops to update their ETAs.
+This assumes that the schedule is well calibrated 
+and the scheduled travel time between stops
+is representative of the real-world travel time between them. 
+Therefore, an alternative estimation approach that uses \rt travel times 
+between stops would seem more adequate.
 
@@
 at least directly,
-so here we construct a transit road network consisting of intersections
+so here we describe how we constructed a transit road network consisting of intersections
 and the roads between them.
-By generalising to physical roads,
-we remove any route-dependency,
-which increases the quantity of data available for each road.
-Each route is then represented as a sequence of road segments,
-each with an associated travel time.
+Generalising to physical roads removes route dependencies,
+which increases the quantity of data available for each road
+and allows each route to be represented as a sequence of road segments,
+which have individual travel time distributions (Section~\ref{sec:kf}).
 
-The transit network is constructed by splitting each route
+The transit network was constructed by splitting each route
 into spatially identifiable segments,
-each representative of a physical road.
-The simplest way to do this is to use bus stops as nodes in the network
-(Figure~\ref{fig:network_creation_1}),
-and connecting roads as edges.
+each representative of a physical road,
+by using bus stops as nodes in the network
+and the connecting roads as edges
+(Figure~\ref{fig:network_creation}).
 In this way, routes that service the same subsequence of stops
-will all contribute to congestion information for the connecting roads,
-as shown in Figure~\ref{fig:network_creation_2}.
-There are several drawbacks to this method,
-for example routes merge at physical intersections, not bus stops,
-so some overlap will occur as shown.
-However, this is minimal, and our simplification is sufficient for the current work.
+all contribute to congestion information for the connecting roads.
+Although there are several drawbacks to this method,
+such as road segment overlaps where routes merge between stops,
+this is minimal and our approach is sufficient for the current work.
+Future work will investigate methods of detecting where routes merge
+(physical intersections) and using these as nodes in the network.
 
@@
     }
-    % Construction of a route network involves combining routes at nodes %
-    %     (here we are using stops) which are connected by edges (roads). %
-    %     In (a), the three unique routes, represented by different line types, clearly %
-    %     overlap in several places. In (b) these have been merged, and the width of each line %
-    %     represents how many routes use that link.}
     \label{fig:network_creation}
@@
 
-GTFS-realtime provides the current positions of vehicles in a transit network.
-The data consists of the time $t_k$ that the observation was made,
+\emph{GTFS-realtime} provides the positions of vehicles in a transit network.
+The data consists of the time $t_k$ that the position was last updated,
 the GPS position of the vehicle, $\bY_k$, 
@@
 these positions can be preemptive,
-(i.e., the bus is almost there, but not quite),
+for example when approaching a queue of traffic at an intersection
+and the way-point is triggered before the bus physically gets there,
 so consecutive observations can show a bus travelling backwards.
@@
 to its previous state before continuing.
-This way, we avoid degeneration of the particle filter (Section~\ref{sec:model_perf})
+We are available to avoid degeneration of the particle filter (Section~\ref{sec:model_perf})
 at the cost of losing one observation that is most likely invalid anyway.

### 2018-11-01 anzjs2018/sections/03_models.tex
Clean up dir

@@
 sequentially as the vehicle traverses the route.
-These are combined into the state vector,
-\begin{equation}
-\label{eq:vehicle_state}
-\bX_k = 
-\begin{bmatrix}
-    x_k \\ \dot x_k \\ \bz
-\end{bmatrix}.
-\end{equation}
+These are combined into the state vector
+$\bX_k = \left[x_k\ \dot x_k\ \bz\right]^\top$.
+% \begin{equation*}
+% \label{eq:vehicle_state}
+% \bX_k = 
+% \begin{bmatrix}
+%     x_k \\ \dot x_k \\ \bz
+% \end{bmatrix}.
+% \end{equation*}
 Observations $\bY_k$ of the vehicle are made using a GPS at time $t_k$
 with GPS error $\epsilon^2$,
-giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle,
-\begin{equation}
-\label{eq:vehicle_obs}
-\bY_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}.
-\end{equation}
+giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle
+as $\bY_k = \left[\lambda_k\ \phi_k \right]^\top$
+% \begin{equation}
+% \label{eq:vehicle_obs}
+% \bY_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}.
+% \end{equation}
 
 We use a particle filter to estimate $\bX_k$,
-due to do its high flexibility and use in recent 
+due to its high flexibility and use in recent 
 transit vehicle modelling applications \citep{Hans_2015}.
@@
         to predict the next state.
-        The measurement function $h$ allows state particle state
+        The measurement function $h$ allows each particle state
         to be mapped to a GPS coordinate (a),
         which can be compared to the observed location (c)
-        to calculate each particles likelihood and perform resampling.
+        to calculate each particle's likelihood to compute resampling weights.
     }
@@
 each with an associated weight $W_{k-1}^{(i)}$
-\begin{equation}
+\begin{equation*}
 p(\bX_{k-1} | \bY_{k-1}) \approx \tilde\bX_{k-1|k-1} 
 := \left\{\bX_{k-1}^{(i)}, W_{k-1}^{(i)}\right\}_{i=1}^N
-\end{equation}
+\end{equation*}
 each of which is independently updated or \emph{mutated} using the transition function $f$ (fig~\ref{fig:pf_state_predict})
-\begin{equation}
+\begin{equation*}
 p(\bX_k | \bX_{k-1}) \approx \tilde\bX_{k|k-1} := 
 \left\{f(\bX_{k-1}^{(i)}, \psi), W_{k-1}^{(i)}\right\}_{i=1}^N
-\end{equation}
+\end{equation*}
 using parameter vector $\psi$ containing all of the necessary parameters
@@
 generating a posterior estimate of the state
-\begin{equation}
+\begin{equation*}
 p(\bX_k | \bY_k) \approx \tilde\bX_{k|k} := 
 \left\{\bX_{k}^{(i)}, W_{k}^{(i)}\right\}_{i=1}^N.
-\end{equation}
+\end{equation*}
 
@@
 \emph{importance resampling} is employed whenever the 
-effective sample size 
-\begin{equation}
-\label{eq:neff}
-N_{\text{eff}} = \frac{1}{\sum_i (W_k^{(i)})^2}
-\end{equation}
-falls below a specified threshold $N_{\text{thres}}$
-\citep{}.
+effective sample size $N_{\text{eff}} = 1 / \sum_i (W_k^{(i)})^2$
+falls below a specified threshold $N_{\text{thres}}$.
 The resampling is performed with replacement using importance weights $W_k^{(i)}$,
@@
 For these parameters, we used constant values for all stops,
-based on the values used by \cite{Hans_2015};
+based on those used by \cite{Hans_2015};
 future work will investigate modelling these in real-time similarly to Section~\ref{sec:kf}.
@@
 The stop positions are represented by their distance into trip,
-$\boldsymbol{S} = (S_1, \ldots, S_J)^\top$.
-The $stop\_index$ function returns the index of a stop,
-given a distance travelled and set of stop distances.
+$\boldsymbol{S} = (S_1, \ldots, S_J)^\top$, and
+the $stop\_index$ function returns the index of a stop,
+given a distance travelled and set of stop distances $\boldsymbol{S}$.
 Lines~2--9 are necessary to decrease the degeneration rate of the filter,
@@
 Finally, to track travel time along segments,
-at each point where the vehicle travels (lines 9 and 26),
+at each point where the vehicle travels (lines 7 and 26),
 the travel time for the current segment, $z_j$, is incremented.
@@
 \newcommand{\Break}{\State \algorithmicbreak}
-\begin{algorithm}[tb]
+\begin{algorithm}[t]
     \caption{Particle mutation function.}
@@
 
+\afterpage{\clearpage}
+
 \subsubsection{Updating state using the observation likelihood}
@@
 assuming GPS error $\epsilon^2$,
-\begin{equation}
+\begin{equation*}
 \label{eq:pf_obs_model}
@@
 \quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
-\end{equation}
+\end{equation*}
 The geographical distance between the observation and the true vehicle can be expressed
 in terms of two independent random variables $z_1, z_2 \sim \mathrm{N}(0,1)$
-\begin{equation}
+\begin{equation*}
 \label{eq:obs_dist}
@@
     = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}.
-\end{equation}
+\end{equation*}
 The sum of two independent, standard normal random variables 
@@
 which is itself exponential,
-so rearranging (\ref{eq:obs_dist}) yields
-\begin{equation}
+so rearranging gives
+\begin{equation*}
 \label{eq:obs_exp}
 \left(\frac{dist(\bY_k, h(\bX_k))}{\epsilon}\right)^2 =
-z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
-\end{equation}
+z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right).
+\end{equation*}
 
 The likelihood of the data given a particle's state estimate 
-is calculated using (\ref{eq:obs_exp})
-\begin{equation}
+can now be calculated as
+\begin{equation*}
 p(\bY_k | \bX_k^{(i)}, \epsilon) =
@@
 \right\}
-\end{equation}
+\end{equation*}
 It is worth noting that this representation of the likelihood is only
@@
 using the uncertainty of particle estimates.
-The model from (\ref{eq:rbe_model}) now reduces to
-\begin{equation}
+The model from (\ref{eq:rbe_model}) reduces to
+\begin{equation*}
 \begin{split}
@@
 \end{split}
-\end{equation}
+\end{equation*}
 where $\Delta_c = t_c - t_{c-1}$.
@@
 Since multiple vehicles can travel along a road simultaneously,
-we have used an information filter to implement the network model.
+we used an information filter to implement the network model.
 The information filter is a transformation of the Kalman filter in which the

### 2018-11-01 anzjs2018/sections/04_implementation.tex
Clean up dir

@@
 while the second implements the \rt models and prediction.
-We chose \verb+Rcpp+ to develop the program,
+We used Rcpp \citep{Rcpp}
+to develop the program,
 which provides access to other R packages for data manipulation 
-(notably \verb+RSQLite+ and \verb+dplyr+),
-as well as the speed and memory management capabilities of \verb|C++|. 
+as well as the speed and memory management capabilities of C++.
 The program is implemented in the R package
@@
 In this section, we discuss the features of the \rt component
-and assess its performance in \rt,
+and assess its performance 
 both with respect to timing and parameter estimation.
@@
     \item Update or create new vehicle objects from the new data
-    \item \textbf{Run particle filter on each vehicle to update or initialise state}
+    \item \textbf{Run particle filter on each vehicle to update or initialise its state}
     \item \textbf{Update state of any roads for which vehicles 
         have completed travel}
-    \item Generate ETAs for vehicles using combination of particle filter and network state
-    \item Write ETAs to extended Google Protobuf binary file for distribution
+    \item Generate ETAs for vehicles using combination of vehicle and network state
+    \item Write ETAs to a file for distribution
 \end{enumerate}
@@
 we were primarily concerned with ensuring each component of the program
-is as efficient as possible,
+was as efficient as possible,
 allowing ETAs to be generated and distributed fast enough to be feasible in real-time,
 with a target of 30~seconds or faster at peak time.
-Using C++ provides the memory management control necessary to make the application
-viable in real-time.
-We also make use of OpenMP for parallelisation,
+Using C++ provided the memory management control necessary to make this possible.
+We also made use of OpenMP for parallelisation,
 since each vehicle is modelled independently so it is trivial to scale up without
-worrying about thread safety.
+issues with thread safety.
 
@@
 depends on many factors,
-so it is necessary to explore the performance of the application
+so it was necessary to explore the performance of the application
 with varying number of particles.
@@
 In each iteration, 
-the timing of the various program components is recorded.
+the timing of the various program components was recorded.
 Since the number of vehicles travelling at any given time changes throughout the day,
@@
 
-The most time consuming component is ETA writing,
-which involves summarising the individual ETAs estimated for each particle.
-While not discussed here this involves quantiles which use a sorting algorithm,
-which has complexity $O(N \log N)$ 
+The most time consuming component was ETA writing,
+which involved summarising the individual ETAs estimated for each particle.
+While not discussed here this involved computing quantiles which uses a sorting algorithm
+with complexity $O(N \log N)$,
 which explains the slightly quadratic trend in the timings.
 
-The next most intensive step is vehicle updating,
-and takes about 5~seconds for 8000 particles.
-The ETA prediction step only takes a few seconds,
+The next most intensive step was vehicle updating,
+which took about 5~seconds for 8000 particles.
+The ETA prediction step only took a few seconds,
 although this may increase once a more comprehensive model is developed.
@@
 when 8000~particles are used.
-In the next section we explore the effect of changing $N$
-to allow finding the optimal number.
 
@@
         The timings shown are \emph{wall time}, the actual time passed,
-        not CPU time which would be approximately 8~times larger 
-        (for using 8~cores).
+        not CPU time.
         Steps that only involve particle calculations are approximately
         of complexity $O(N)$,
-        however some steps are a little more complex, 
+        however some steps are a little more complex (writing ETAs involves computing quantiles), 
         so there is a slightly exponential trend.
@@
 between each observation and the route using a nearest point algorithm,
-so we chose to use GPS error values $\epsilon \in \{1,2,3,5\}$. 
-For the system noise we use values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$,
-which correspond to the average vehicle speed having a variance between 0.1 and 45~meters per second
+which led us to use the values $\epsilon \in \{1,2,3,5\}$ for GPS error. 
+For the system noise we used values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$,
+which correspond to an average vehicle speed variance of between 0.1 and 45~meters per second
 over a period of 30~seconds.
-For each of the combination of these parameter values, along with varying $N$,
-several values were computed.
-These were effective sample size, $N_\text{eff}$ calculated using (\ref{eq:neff});
+For each of combination of these parameter values, including varying $N$,
+the following values were calculated:
+effective sample size, $N_\text{eff}$;
 degeneration rate, the percentage of iterations in which the vehicle was lost
 (no plausible particles);
-and the variance of segment travel time estimates.
+and the (relative) variance of segment travel time estimates.
 
@@
 
-From Figure~\ref{fig:degen_rate}, we see that $N_\text{eff}$ increases
-with GPS error and decreases with system noise.
-The degeneration rate decreases with GPS error and increases with $N$,
-but is relatively unaffected by system noise.
+Figure~\ref{fig:neff} shows that $N_\text{eff}$ increased
+with GPS error and decreased with system noise,
+while remaining unaffected by changing $N$.
+The degeneration rate decreased with GPS error and increased with $N$,
+and was unaffected by system noise,
+as seen in Figure~\ref{fig:degen_rate}.
 Large GPS error affects the likelihood,
-and gives more weight to particles farther from the observation
+giving more weight to particles farther from the observation
 than does a smaller GPS error.
@@
         but there is no visible effect of system noise.
+        Note, however, that some of these can be attributed to invalid data,
+        for example if the driver has entered the wrong trip direction.
         The error bars show $\pm 2$SE.
@@
 The last measure of performance is the relative variance of segment travel times.
-This is calculated by computing the average variance of travel times over all simulations,
-and then for each simulation, 
-taking the average (over road segments) of the ratio between variance for that simulations
-and variance for all simulations.
+We used relative variance as each segment has a very different travel time distribution,
+so to compare the variability between simulations we needed to 
+compute the average variance of travel times over all simulations,
+taking the average (over road segments) of the ratio between variance for
+each individual simulation and variance over all simulations.
 Figure~\ref{fig:travel_times} shows that simulations with larger GPS error
-result in a higher relative variance,
-while simulations with more particles have lower relative variance.
-There is no visible effect of system noise on travel time variation.
+resulted in a higher relative variance,
+while simulations with more particles had lower relative variance.
+There was no visible effect of system noise on the variability of travel times.
 
@@
 is going to have a significant effect on the final ETAs,
-particularly when compared to the uncertainty of forecasting road state.
+particularly when compared to the uncertainty of forecasting road state,
+or whether reducing the degeneration rate, 
+or increasing $N_\text{eff}$, would be more advantageous.
 

### 2018-11-01 anzjs2018/sections/06_discuss.tex
Clean up dir

@@
 
-In this paper we have described an approach to transit vehicle modelling
+In this paper we described an approach to transit vehicle modelling
 that enables the \rt estimation of road state from vehicle position data,
-which can later be used to make arrival time predictions
+which in the future will be used to make arrival time predictions
 that account for \rt traffic conditions.
@@
 that combines \rt vehicle state with travel time forecasts
-and provide commuters with reliable ETAs.
+and provide commuters with what will hopefully be reliable ETAs.
 These will likely consist of a point estimate along with a prediction interval,

### 2019-07-31 anzjs2018/sections/02_gtfs.tex
make a bunch of changes

@@
 who use it in Google Maps Transit Directions,
-it is used by over 900~transit providers around the world,
-including here in Auckland, New Zealand
-(source \url{http://transitfeeds.com}).
+it is used by over 1100~transit providers around the world\footnote{%
+source: \url{http://transitfeeds.com} as of August 2019},
+including here in Auckland, New Zealand.
 An advantage of this standardised format is that,
@@
 \item \emph{schedules}, specifying the arrival (and departure) times for each bus at each of its stops; 
-\item \emph{shapes}, the sequence of points defining a vehicle's path along a route
+\item and \emph{shapes}, the sequence of points defining a vehicle's path along a route.
 \end{itemize}
-\citep{GoogleDevelopers_2006}.
 The second component is \emph{GTFS-realtime},
@@
 and the stop at which the passenger is waiting.
-In most applications, however, this vital information is unavailable,
-at least directly,
-so here we describe how we constructed a transit road network consisting of intersections
-and the roads between them.
-Generalising to physical roads removes route dependencies,
-which increases the quantity of data available for each road
-and allows each route to be represented as a sequence of road segments,
-which have individual travel time distributions (Section~\ref{sec:kf}).
+In most applications, however, this vital information is unavailable, at least directly.
+In order to compute travel times efficiently
+and make use of all available data pooled across multiple routes,
+we constructed a \emph{road network},
+allowing us to model each physical road as a unique entity.
+In this way, every bus travelling along a road contributes to its state,
+and can be used to predice travel times for all upcoming buses 
+travelling along the road,
+irrespective of the route they are servicing.
 
@@
 this is minimal and our approach is sufficient for the current work.
-Future work will investigate methods of detecting where routes merge
-(physical intersections) and using these as nodes in the network.
 

### 2019-07-31 anzjs2018/sections/03_models.tex
make a bunch of changes

@@
 \citep{Anderson_2012}.
-These models assume an underlying Markov process with state $\mathcal{X}_k$
-and system noise $\omega_k\sim\mathrm{N}(0,\mathcal{Q})$ 
-of which observations $\mathcal{Y}_k$ are made
-with error $\nu_k\sim\mathrm{N}(0,\mathcal{R})$,
+These models assume an underlying Markov process with state $\boldsymbol{\mathcal{X}}_k$,
+an $n\times1$ column vector,
+and system noise $\boldsymbol{\omega}_k\sim\mathrm{N}(\boldsymbol{0},\mathcal{Q})$ 
+of which observations $\boldsymbol{\mathcal{Y}}_k$,
+an $m\times1$ column vector, are made
+with error $\boldsymbol{\nu}_k\sim\mathrm{N}(\boldsymbol{0},\mathcal{R})$,
 giving the following general model:
@@
 \begin{split}
-\mathcal{X}_k &= f(\mathcal{X}_{k-1}, \omega_k) \\
-\mathcal{Y}_k &= h(\mathcal{X}_k, \nu_k)
+\boldsymbol{\mathcal{X}}_k &= f(\boldsymbol{\mathcal{X}}_{k-1}, \boldsymbol{\omega}_k) \\
+\boldsymbol{\mathcal{Y}}_k &= h(\boldsymbol{\mathcal{X}}_k) + \boldsymbol{\nu}_k
 \end{split}
 \end{equation}
-The transition function $f$ describes the relationship between consecutive states,
-while the measurement function $h$ is a deterministic function describing the
-relationship between the underlying and observed states.
+The transition function $f:\mathbb{R}^n\mapsto\mathbb{R}^n$ 
+describes the relationship between consecutive states,
+while the measurement function $h:\mathbb{R}^n\mapsto\mathbb{R}^m$ is a deterministic function
+mapping the underlying state to the observable state.
 
@@
 its speed $\dot x$ in meters per second.
-Additionally, travel times along road segments along the route, 
-$\bz = (z_1,\ldots,z_L)^\top$, in seconds,
-are included in the state and estimated
-sequentially as the vehicle traverses the route.
+% The travel times along road segments, $\bz = (z_1,\ldots,z_L)^\top$, in seconds,
+% are estimated sequentially as the vehicle traverses the route,
+% and are not included explicitely in the state.
 These are combined into the state vector
-$\bX_k = \left[x_k\ \dot x_k\ \bz\right]^\top$.
-% \begin{equation*}
-% \label{eq:vehicle_state}
-% \bX_k = 
-% \begin{bmatrix}
-%     x_k \\ \dot x_k \\ \bz
-% \end{bmatrix}.
-% \end{equation*}
-Observations $\bY_k$ of the vehicle are made using a GPS at time $t_k$
+$\bX_k = \left[x_k\ \dot x_k\right]^\top$,
+of which observations $\bY_k$ are made using a GPS at time $t_k$
 with GPS error $\epsilon^2$,
 giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle
-as $\bY_k = \left[\lambda_k\ \phi_k \right]^\top$
-% \begin{equation}
-% \label{eq:vehicle_obs}
-% \bY_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}.
-% \end{equation}
+as $\bY_k = \left[\lambda_k\ \phi_k \right]^\top$.
 
@@
 due to its high flexibility and use in recent 
-transit vehicle modelling applications \citep{Hans_2015}.
+transit vehicle modelling applications \citep{Hans_2015},
+as well as providing an easy way to estimate travel times
+$\bz = (z_1,\ldots,z_L)^\top$, in seconds, along road segments
+as the vehicle traverses the route.
 The primary advantage of a particle filter is its handling of multimodality,
-as demonstrated in figure~\ref{fig:pf_state_predict},
+as demonstrated in Figure~\ref{fig:pf_state_predict},
 which is a common feature of the proposal distribution, particularly around bus stops.
-Another advantage is the intuitive likelihood function, 
-discussed in section~\ref{sec:pf_update}.
+Another advantage is that the likelihood function is intuitively based
+on the physical distance between the vehicle observation and the state's
+estimate of the true location, $h(\bX_k)$ (Section~\ref{sec:pf_update}),
+rather than the reverse of first estimating the ``observed'' distance travelled.
 Conversely, particle filter methods are computationally demanding,
@@
 is represented by a set of discrete points, or particles, (Figure~\ref{fig:pf_state_prev}),
-each with an associated weight $W_{k-1}^{(i)}$
+each with an associated weight $W_{k-1}^{(i)}$.
+The state can then be expressed using the Dirac delta function $\dot\delta_X(x)$ \citep{cn},
+so that
 \begin{equation*}
-p(\bX_{k-1} | \bY_{k-1}) \approx \tilde\bX_{k-1|k-1} 
-:= \left\{\bX_{k-1}^{(i)}, W_{k-1}^{(i)}\right\}_{i=1}^N
+p(\bX_{k-1} | \bY_{1:k-1}) \approx 
+    \sum_{i=1}^N W_{k-1} \dot\delta_{\bX_{k-1}}(\bX_{k-1}^{(i)})
 \end{equation*}
-each of which is independently updated or \emph{mutated} using the transition function $f$ (fig~\ref{fig:pf_state_predict})
+each of which is independently updated or \emph{mutated} using the transition function $f$ (Figure~\ref{fig:pf_state_predict})
 \begin{equation*}
-p(\bX_k | \bX_{k-1}) \approx \tilde\bX_{k|k-1} := 
-\left\{f(\bX_{k-1}^{(i)}, \psi), W_{k-1}^{(i)}\right\}_{i=1}^N
+p(\bX_k | \bX_{k-1}) \approx 
+    \sum_{i=1}^N W_{k-1} \dot\delta_{\bX_{k}}(f(\bX_{k-1}^{(i)}, \psi))
 \end{equation*}
@@
 \begin{equation*}
-p(\bX_k | \bY_k) \approx \tilde\bX_{k|k} := 
-\left\{\bX_{k}^{(i)}, W_{k}^{(i)}\right\}_{i=1}^N.
+p(\bX_k | \bY_k) \approx  
+    \sum_{i=1}^N W_{k} \dot\delta_{\bX_{k}}(\bX_{k}^{(i)})
 \end{equation*}
 
-In order to prevent degeneration of the particle filter state,
-\emph{importance resampling} is employed whenever the 
+One problem with particle filters is degeneration,
+which occurs when the particle sample no longer approximates the target distribution
+due to the particles becoming too diverse
+and all of the weight falling onto only a small number of them.
+This occurs naturally over time, but the rate at which it occurs depends 
+on a relationship between how well the transition function predicts future states,
+and the variance of those predictions (system noise).
+To avoid degeneration,
+particle filters use \emph{importance resampling}
+using importance weights $\{W_k^{(1)}, \ldots, W_k^{(N)}\}$,
+which removes improbable particles and replaces them with more probable ones,
+as demonstrated in Figure~\ref{fig:pf_state_predict2}.
+However, resampling requires sorting $N$ particles,
+which is of $\mathcal{O}(N\log N)$ complexity,
+so we can reduce computational costs by only resampling when required.
+To prevent degeneration of the particle filter state,
+the sample is resampled whenever the
 effective sample size $N_{\text{eff}} = 1 / \sum_i (W_k^{(i)})^2$
 falls below a specified threshold $N_{\text{thres}}$.
-The resampling is performed with replacement using importance weights $W_k^{(i)}$,
-as demonstrated in Figure~\ref{fig:pf_state_predict2}.
-Afterwards, the weights are reinitialised to $W_k^{(i)} = N^{-1}$.
-Note this is a computationally demanding step,
-so reducing the rate of resampling is favourable.
+We used a fixed threshold value of $N_{\text{thres}} = N/10$
+for the current work.
+\textbf{Check this}.
 

### 2019-07-31 anzjs2018/figures/03_particle_filter.R

@@ -30,7 +30,7 @@ d0 <- ggplot() +
     geom_path(aes(x = xx, y = yy), data = tibble(xx = c(8000, 15000), yy = c(0, 0))) +
     theme_minimal() + 
     theme(panel.grid = element_blank(), axis.text = element_blank()) +
-    xlab("Distance Traveled") + ylab("") +
+    xlab("Distance Travelled") + ylab("") +
     xlim(8000, 15000)
 
 h <- function(x, shape = shapef) {

### 2019-07-31 anzjs2018/figures/03_particle_filter.R
Update figures

@@ -9,12 +9,15 @@ route <- gtfs %>% tbl("routes") %>%
     filter(route_short_name == "110" && route_long_name %like% '%To City%') %>%
     arrange(desc(version)) %>% head(1) %>% collect()
 trip <- gtfs %>% tbl("trips") %>% 
-    filter(route_id == route$route_id) %>% head(1) %>% collect()
+    filter(route_id == !!route$route_id) %>% head(1) %>% collect()
+distgeo <- function(x) {
+    c(0, geosphere::distGeo(x))[1:nrow(x)]
+}
 shape <- gtfs %>% tbl("shapes") %>%
-    filter(shape_id == trip$shape_id) %>% arrange(shape_pt_sequence) %>% collect() %>%
+    filter(shape_id == !!trip$shape_id) %>% arrange(shape_pt_sequence) %>% collect() %>%
     mutate(shape_dist_traveled = (.) %>% 
         select(shape_pt_lon, shape_pt_lat) %>%
-        as.matrix() %>% geosphere::distGeo() %>% cumsum %>% c(0, .))
+        as.matrix() %>% distgeo() %>% cumsum)
 dbDisconnect(gtfs)
 
 shapef <- shape %>% filter(shape_dist_traveled > 8000 & shape_dist_traveled < 15000)
@@ -47,12 +50,12 @@ h <- function(x, shape = shapef) {
         geosphere::destPoint(p1, b, d) %>% as.numeric
     }))
     colnames(x) <- c('lon', 'lat')
-    as.tibble(x)
+    as_tibble(x)
 }
 
 WIDTH = 5
 HEIGHT = 2.5
-set.seed(873624)
+set.seed(8237315)
 N <- 20
 x <- sample(rnorm(N, 9000, 50), replace = TRUE)
 
@@ -69,7 +72,8 @@ grid.text("h(x)", 0.5, 0.5, gp = gpar(fontface = 3, fontfamily = 'HersheySerif')
 dev.off()
 
 
-x2 <- x + runif(N, c(10, 20), c(20, 25)) * 120
+# x2 <- x + runif(N, c(10, 23), c(20, 27)) * 120 + rnorm(N, 0, 10)
+x2 <- x + rnorm(N, c(14, 25), c(3, 1)) * 120
 
 pdf('figures/03_particle_filter_2.pdf', width = WIDTH, height = HEIGHT)
 gridExtra::grid.arrange(

### 2019-08-01 anzjs2018/sections/03_models.tex
edits to sec 3

@@
 
-The particle filter allows us to model several important bus behaviours:
+The particle filter allows us to flexibly model bus behaviour,
+which in our current model includes
 \begin{itemize}
-\item non-constant speed along roads,
-\item optional stopping and waiting at bus stops while passengers board and disembark, and
-\item optional stopping and waiting at intersections or traffic lights (future work).
+\item non-constant speed along roads (between observations), and
+\item bus stop behaviour, which involves optional stopping and waiting times
+    while passengers board and disembark.
 \end{itemize}
+To handle the above, the transition function consists of two models,
+one for each of the above behaviours.
+The first models the speed and distance travelled of the vehicle 
+by using Newton's Laws of Motion,
+which, letting $\delta = t_k - t_{k-1}$ and $\epsilon_k^{(i)}\sim\mathrm{N}(0, \sigma^2)$, is
+\begin{equation}
+\label{eq:trans_newton}
+\bX_k^{(i)} = f_1(\bX_{k-1}^{(i)}, \delta_k, \epsilon_k^{(i)}) = 
+    \begin{bmatrix}
+        x_{k-1}^{(i)} + \delta(\dot x_{k-1}^{(i)} + \epsilon_k^{(i)}) \\
+        \dot x_{k-1}^{(i)} + \epsilon_k^{(i)}
+    \end{bmatrix}.
+\end{equation}
 
-For each particle, the transition function generates a plausible trajectory,
-using system noise parameter $\sigma^2$ which describes 
-how the vehicle's speed changes as a Markov process.
-The vehicle does not travel constantly along the route,
-however, as it needs to service bus stops along the way,
-so the transition function includes 
-stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
-dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
-board and disembark (conditional on the bus stopping),
-and the minimum dwell time at stops, $\gamma$,
-for the doors to open and close,
-and account for deceleration and acceleration.
-For these parameters, we used constant values for all stops,
-based on those used by \cite{Hans_2015};
-future work will investigate modelling these in real-time similarly to Section~\ref{sec:kf}.
+The second vehicle behaviour requires a model of bus stop wait times.
+We assume the vehicle is approaching stops $j$,
+and will stop with some probability $\pi_j$.
+If it does not stop, the remaining travel time $\delta_k$ is unaltered;
+otherwise, the bus waits for a minimum of $\gamma$ seconds---this 
+accounts for deceleration, opening and closing of doors, and acceleration
+\citep{Hands_2015}---and then waits while passengers board and disembark,
+which is assumed to be exponentially distributed with mean $\tau_j$.
+This leads to the following model,
+\begin{equation}
+\label{eq:trans_stop}
+\begin{split}
+p_j^{(i)} &\sim \mathrm{Bernoulli}(\pi_j) \\
+\tilde d_j^{(i)} &\sim \mathrm{Exponential}(\tau_j) \\
+d_j^{(i)} &= p_j^{(i)}(\gamma + \tilde d_j^{(i)}).
+\end{split}
+\end{equation}
 
-The transition function algorithm is shown in Algorithm~\ref{fig:algorithm},
-which implements those features discussed above.
-The stop positions are represented by their distance into trip,
-$\boldsymbol{S} = (S_1, \ldots, S_J)^\top$, and
-the $stop\_index$ function returns the index of a stop,
-given a distance travelled and set of stop distances $\boldsymbol{S}$.
-Lines~2--9 are necessary to decrease the degeneration rate of the filter,
-which can occur when there is a long delay between observations
-simply because the bus has not moved due to traffic or other reasons. 
-We therefore place positive probability for some particles to remain stationary.
+The combination of (\ref{eq:trans_newton}) and (\ref{eq:trans_stop})
+is handled by the transition function algorithm,
+displayed in Algorithm~\ref{fig:algorithm}.
+While (\ref{eq:trans_stop}) is used directly in the algorithm,
+(\ref{eq:trans_newton}) is modified to allow the bus to approach a stop
+and conditionally wait before continuing,
+until $\delta_k$ reaches zero.
+The algorithm also keeps track of each particle's travel time $z_\ell^{(i)}$
+along each segment $\ell$ of the route,
+storing the times so that the posterior distribution of the travel time
+can be computed once all particles have completed segment $\ell$:
+\begin{equation}
+\label{eq:post_tt}
+p(z_\ell | \bY_{1:k}) \approx
+    \sum_{i=1}^N W_k^{(i)} \dot\delta_{z_\ell}(z_\ell^{(i)}).
+\end{equation}
 
-Lines~11--12 of Algorithm~\ref{fig:algorithm} implement the system noise,
-which involves changing the vehicle's speed each second (i.e., acceleration).
-Lines~13--25 handle arrival behaviour at stops:
-if the particle passes the next stop,
-it has a chance of stopping (lines~18--19),
-in which case the dwell time is sampled and subtracted from the 
-time remaining to travel (lines~20--21).
-Finally, to track travel time along segments,
-at each point where the vehicle travels (lines 7 and 26),
-the travel time for the current segment, $z_j$, is incremented.
-The mutation ends once the travel time $\delta$ reaches zero,
-or when the vehicle reaches the end of the route (lines 10, 14).
-Once complete, the particle's new state is returned (line 29).
+% For each particle, the transition function generates a plausible trajectory,
+% using system noise parameter $\sigma^2$ which describes 
+% how the vehicle's speed changes as a Markov process.
+% The vehicle does not travel constantly along the route,
+% however, as it needs to service bus stops along the way,
+% so the transition function includes 
+% stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
+% dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
+% board and disembark (conditional on the bus stopping),
+% and the minimum dwell time at stops, $\gamma$,
+% for the doors to open and close,
+% and account for deceleration and acceleration.
+% For these parameters, we used constant values for all stops,
+% based on those used by \cite{Hans_2015};
+% future work will investigate modelling these in real-time similarly to Section~\ref{sec:kf}.
+
+% The transition function algorithm is shown in Algorithm~\ref{fig:algorithm},
+% which implements those features discussed above.
+% The stop positions are represented by their distance into trip,
+% $\boldsymbol{S} = (S_1, \ldots, S_J)^\top$, and
+% the $stop\_index$ function returns the index of a stop,
+% given a distance travelled and set of stop distances $\boldsymbol{S}$.
+% Lines~2--9 are necessary to decrease the degeneration rate of the filter,
+% which can occur when there is a long delay between observations
+% simply because the bus has not moved due to traffic or other reasons. 
+% We therefore place positive probability for some particles to remain stationary.
+
+% Lines~11--12 of Algorithm~\ref{fig:algorithm} implement the system noise,
+% which involves changing the vehicle's speed each second (i.e., acceleration).
+% Lines~13--25 handle arrival behaviour at stops:
+% if the particle passes the next stop,
+% it has a chance of stopping (lines~18--19),
+% in which case the dwell time is sampled and subtracted from the 
+% time remaining to travel (lines~20--21).
+% Finally, to track travel time along segments,
+% at each point where the vehicle travels (lines 7 and 26),
+% the travel time for the current segment, $z_j$, is incremented.
+% The mutation ends once the travel time $\delta$ reaches zero,
+% or when the vehicle reaches the end of the route (lines 10, 14).
+% Once complete, the particle's new state is returned (line 29).
 
@@
 
-After mutating the particle set, the posterior distribution $\tilde\bX_{k|k}$ 
-is obtained by reweighting each particle using the likelihood.
-The measurement function $h$,
-and an additional function $g$ which transforms GPS coordinates onto a flat
-surface (the geographical equirectangular projection, \cite{Snyder_1998}),
-allow the placement of a bivariate normal likelihood on the data.
-The model for the observations is,
-assuming GPS error $\epsilon^2$,
-\begin{equation*}
-\label{eq:pf_obs_model}
-g(\bY_k) = g(h(\bX_k)) + \br_k,
-\quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
-\end{equation*}
-The geographical distance between the observation and the true vehicle can be expressed
-in terms of two independent random variables $z_1, z_2 \sim \mathrm{N}(0,1)$
-\begin{equation*}
+To compute the likelihood of the observation given each particle state,
+$p(\bY_k|\bX_k^{(i)})$,
+the measurement function $h$ is needed along with an
+\emph{geographical projection} $g$ to allow comparison of $\bY_k$,
+a GPS observation, with $\bX_k$, a distance travelled along the route in meters.
+
+The measurement function $h$ computes the GPS coordinates by using the 
+shape information provided by GTFS and travelling $x_k$ meters along 
+this two-dimensional line.
+Meanwhile, we use the \emph{Equirectangular projection} \citep{Snyder_1998},
+$\br = g(\bY_1 | \bY_0)$,
+so that the magnitude of $\boldsymbol{r}$ is equal to the physical distance
+between $\bY_0$ and $\bY_1$, with both dimensions in meters.
+Let $\bY_i = [\lambda_i, \psi_i]^\top$,
+with $\lambda_i$ and $\psi_i$ measured in radians,  
+and $R$ is the Earth's radius, in meters, then
+\begin{equation}
+\br = 
+g(\bY_1 | \bY_0) = 
+    \begin{bmatrix}
+        r_1 \\ r_2
+    \end{bmatrix} =
+    R \begin{bmatrix}
+        (\lambda_1 - \lambda_0)\cos\phi_0 \\
+        \phi_1 - \phi_0
+    \end{bmatrix}
+\end{equation}
+and, more importantly, the geographical distance $\mathcal{D}$ between the two 
+GPS coordinates is given by
+\begin{equation}
+\label{eq:dist}
+\mathcal{D}(\bY_0, \bY_1) = ||g(\bY_0 | \bY_1)|| = \sqrt{r_1^2 + r_2^2}
+\end{equation}
+
+Now we assume GPS observations have a known error of $\omega^2$ meters,
+and that \mbox{$\br \sim \mathrm{N}(\boldsymbol{0}, \omega^2\mathbf{I})$},
+which allows us to express $\br$ in terms of two independent
+standard normal random variables $z_1, z_2 \sim \mathrm{N}(0,1)$,
+which lets us express the distance between a vehicle observation $\bY_k$
+and a state $\bX_k$ using (\ref{eq:dist}) and the measurement function $h$,
+\begin{equation}
 \label{eq:obs_dist}
-dist(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
-    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}.
-\end{equation*}
-The sum of two independent, standard normal random variables 
-is $\chi^2$ distributed with 2~degrees of freedom,
-which is itself exponential,
-so rearranging gives
-\begin{equation*}
+\mathcal{D}(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
+    = \sqrt{(\omega z_1)^2 + (\omega z_2)^2}
+    = \sqrt{\omega^2 (z_1^2 + z_2^2)}
+\end{equation}
+
+Finally, it is known that the distribution of the sum of two indendent 
+standard normal variables is exponentially distributed with rate $\frac{1}{2}$,
+and that if $Z \sim \mathrm{Exponential}(\theta)$ then
+$cZ \sim \mathrm{Exponential}(\frac{\theta}{c})$,
+so the \emph{squared} distance from (\ref{eq:obs_dist}) has exponential distribution
+\begin{equation}
 \label{eq:obs_exp}
-\left(\frac{dist(\bY_k, h(\bX_k))}{\epsilon}\right)^2 =
-z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right).
-\end{equation*}
+\mathcal{D}(\bY_k, h(\bX_k))^2 =
+\omega^2(z_1^2 + z_2^2) \sim \mathrm{Exponential}\left(\frac{1}{2\omega^2}\right).
+\end{equation}
 
-The likelihood of the data given a particle's state estimate 
-can now be calculated as
-\begin{equation*}
+The likelihood of the observation $\bY_k$ given a state estimate $\bX_k^{(i)}$
+can now be expressed using (\ref{eq:obs_exp}) and the PDF of the exponential distribution,
+\begin{equation}
+\label{eq:lhood}
 p(\bY_k | \bX_k^{(i)}, \epsilon) =
-\frac{1}{2}\exp\left\{
--\frac{1}{2} \left(\frac{dist(\bY_k, h(\bX_k^{(i)}))}{\epsilon}\right)^2
-\right\}
-\end{equation*}
+\frac{1}{2\omega^2}\exp\left\{
+    -\frac{\mathcal{D}(\bY_k, h(\bX_k^{(i)}))^2}{2\omega^2}
+\right\},
+\end{equation}
+where the distance between two GPS coordinates is easily computed,
+making (\ref{eq:lhood}) easy and fast to evaluate within the particle filter algorithm.
+
 It is worth noting that this representation of the likelihood is only
 possible due to the discrete nature of the particle filter state estimate;
-in other approaches, such as the Kalman filter,
-a reverse non-deterministic transformation,
-the inverse measurement function,
-is required to estimate the ``observed'' distance travelled $\hat Z_k~=~h^{-1}(\bY_k)$, 
-which introduces additional error and uncertainty into the model.
+in the Kalman filter, which has often been used in vehicle tracking,
+the measurement \emph{matrix} is used, and required that there is a linear
+transformation between the state and its observations.
+To enable this, applications first estimate the \emph{observed distance travelled}
+by snapping the GPS observations to the route,
+which introduces unnecessary error and uncertainty into the model.
+Our approach avoids this, which makes it more stable in locations where two 
+parts of the route are close to each other,
+such as at loops, where a single GPS observation might have two likely ``snapping'' points.
 

### 2019-08-01 anzjs2018/sections/03_models.tex
more changes

@@
 
-% For each particle, the transition function generates a plausible trajectory,
-% using system noise parameter $\sigma^2$ which describes 
-% how the vehicle's speed changes as a Markov process.
-% The vehicle does not travel constantly along the route,
-% however, as it needs to service bus stops along the way,
-% so the transition function includes 
-% stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
-% dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
-% board and disembark (conditional on the bus stopping),
-% and the minimum dwell time at stops, $\gamma$,
-% for the doors to open and close,
-% and account for deceleration and acceleration.
-% For these parameters, we used constant values for all stops,
-% based on those used by \cite{Hans_2015};
-% future work will investigate modelling these in real-time similarly to Section~\ref{sec:kf}.
-
-% The transition function algorithm is shown in Algorithm~\ref{fig:algorithm},
-% which implements those features discussed above.
-% The stop positions are represented by their distance into trip,
-% $\boldsymbol{S} = (S_1, \ldots, S_J)^\top$, and
-% the $stop\_index$ function returns the index of a stop,
-% given a distance travelled and set of stop distances $\boldsymbol{S}$.
-% Lines~2--9 are necessary to decrease the degeneration rate of the filter,
-% which can occur when there is a long delay between observations
-% simply because the bus has not moved due to traffic or other reasons. 
-% We therefore place positive probability for some particles to remain stationary.
-
-% Lines~11--12 of Algorithm~\ref{fig:algorithm} implement the system noise,
-% which involves changing the vehicle's speed each second (i.e., acceleration).
-% Lines~13--25 handle arrival behaviour at stops:
-% if the particle passes the next stop,
-% it has a chance of stopping (lines~18--19),
-% in which case the dwell time is sampled and subtracted from the 
-% time remaining to travel (lines~20--21).
-% Finally, to track travel time along segments,
-% at each point where the vehicle travels (lines 7 and 26),
-% the travel time for the current segment, $z_j$, is incremented.
-% The mutation ends once the travel time $\delta$ reaches zero,
-% or when the vehicle reaches the end of the route (lines 10, 14).
-% Once complete, the particle's new state is returned (line 29).
-
 \renewcommand{\algorithmicrequire}{\textbf{Start:}}
@@
 as well as make short-term predictions for estimating arrival times.
-In this paper, we develop a simple model in which each road is treated independently.
-Future work will explore correlations between adjacent roads and using historical data
-to improve the model and subsequent predictions.
+In this paper, we model each road segment independently,
+as including correlations simplifies the model to not only a one-dimensional Kalman filter,
+but it also means computations can be run in parallel,
+greatly increasing the real-time performance of the model.
 
@@
 of transit vehicles along road segment $\ell$ at time $t_c$,
-and is assumed to be constant over time 
-with system noise $v_c^\ell \sim \mathrm{N}(0, \nu^2)$.
-Observations of the state, $Z_c^\ell$, are obtained from the particle filter in Section~\ref{sec:pf},
-and have measurement error $e_c^\ell \sim \mathrm{N}(0, R_c^\ell)$ as estimated 
-using the uncertainty of particle estimates.
-The model from (\ref{eq:rbe_model}) reduces to
+of which observations $Z_c^{\ell m}$
+are obtained from the particle filter for vehicle $m$ as defined in
+(\ref{eq:post_tt}), such that 
+(after adding the $m$ superscript to identify unique vehicles), 
+\begin{equation*}
+Z_c^{\ell m} = \mathrm{E}(p(z_c^{\ell m} | \bY^m_{1:k})).
+\end{equation*}
+The measurement error is also estimated from the particle filter,
+\begin{equation*}
+R_c^{\ell m} = \mathrm{Var}(p(z_c^{\ell m} | \bY^m_{1:k})).
+\end{equation*}
+
+The model of travel times becomes a simple reduction of (\ref{eq:rbe_model}) 
+to the one dimensional case, 
+with $f$ and $h$ both unity,
+system noise $v_c^\ell \sim \mathrm{N}(0, \nu^2)$
+with $\Delta_c = t_c - t_{c-1}$
+and $\nu_\ell^2$ is the variance of travel times per second,
+and the observation error $e_c^{\ell m} \sim \mathrm{N}(0, R_c^{\ell m})$:
 \begin{equation*}
@@
 \theta_c^\ell &= \theta_{c-1}^\ell + \Delta_c^\ell v_c^\ell \\
-Z_c^\ell &= \theta_c^\ell + e_c^\ell
+Z_c^{\ell m} &= \theta_c^\ell + e_c^{\ell m}
 \end{split}
 \end{equation*}
-where $\Delta_c = t_c - t_{c-1}$.
 
@@
 This allows multiple observations to be added together to update the state
-in a single iteration.
+in a single iteration,
+which is the case when multiple vehicles travel along a road at the same time.
 
-The state is estimated by its mean $\hat \theta_c^\ell = \mathrm{E}(\theta_c^\ell)$
-and uncertainty $P_c^\ell = \mathrm{var}(\theta_c^\ell)$,
+The state at time $t_{c-1}$ is parameterised by its mean and variance,
+conditional on all of the data from all vehicles up until time $t_{c-1}$,
+$\hat \theta_{c-1|c-1}^\ell = \mathrm{E}(\theta_c^\ell | Z_{1:c-1}^{\ell\boldsymbol{\cdot}})$
+and \mbox{$P_{c-1|c-1}^\ell = \mathrm{var}(\theta_{c-1}^\ell | Z_{1:c-1}^{\ell\boldsymbol{\cdot}})$}, respectively,
 which are predicted from the previous state estimate using the predictive model
@@
 \label{eq:kf_transition}
-\hat \theta^\ell_{c|c-1} &= \hat \theta^\ell_{c|c-1} \\
-P^\ell_{c|c-1} &= P^\ell_{c-1|c-1} + \Delta_c^\ell \nu_c^\ell
+\hat \theta^\ell_{c|c-1} &= \hat \theta^\ell_{c-1|c-1} \\
+P^\ell_{c|c-1} &= P^\ell_{c-1|c-1} + (\Delta_c^\ell \nu_c^\ell)^2
 \end{align*}
@@
 and the information vector $u^\ell_c~=~\hat \theta^\ell_{c|c-1} (P^\ell_{c|c-1})^{-1}$.
-The travel time estimate of vehicle $m$ along segment $\ell$
-is $\bar z_c^{m\ell}$, with uncertainty $s^{m\ell}_c$,
-and is transformed to a measurement information covariance matrix 
-$B_c^{m\ell}~=~(s^{m\ell}_c)^{-2}$
-and measurement information vector $b_c^{m\ell}~=~\bar z^{m\ell}_c (s^{m\ell}_c)^{-2}$.
+The travel time estimate of vehicle $m$ along segment $\ell$,
+$Z_c^{\ell m}$, along with its uncertainty $R^{\ell m}_c$,
+are transformed to a measurement information covariance matrix 
+$B_c^{\ell m}~=~(R^{\ell m}_c)^{-2}$
+and measurement information vector $b_c^{\ell m}~=~Z^{\ell m}_c (R^{\ell m}_c)^{-2}$.
 The total information is the sum of the information over all $M_c^\ell$ vehicles
@@
 \end{equation*}
-which can now be used to predict the travel times of upcoming vehicles.
+which can now be used in the prediction of travel times
+for upcoming buses.

### 2019-08-01 anzjs2018/sections/04_implementation.tex
more changes

@@
 between each observation and the route using a nearest point algorithm,
-which led us to use the values $\epsilon \in \{1,2,3,5\}$ for GPS error. 
+which led us to use the values $\epsilon \in \{1,2,3,5\}$ for GPS error in meters.
 For the system noise we used values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$,

### 2019-08-03 anzjs2018/sections/04_implementation.tex
altering graphs

@@
 to develop the program,
-which provides access to other R packages for data manipulation 
+which provides access to R \citep{rcore},
+which contains many useful tools for data manipulation and package development,
 as well as the speed and memory management capabilities of C++.

### 2019-08-03 anzjs2018/figures/04_model_results.R

@@ -79,16 +79,89 @@ dev.off()
 
 
 
+## vehicle history files
+res <- do.call(bind_rows, 
+    lapply(
+        list.files("../../transitr/simulations", pattern = "sim_"), 
+        function(sim) {
+            siminfo <- as.numeric(strsplit(strsplit(sim, "_")[[1]][3], "-")[[1]])
+            do.call(
+                bind_rows,
+                lapply(
+                    list.files(
+                        file.path("../../transitr/simulations", sim, "history"), 
+                        pattern = "vehicle_.*\\.csv", full.names = TRUE
+                    ),
+                    function(x) {
+                        read_csv(x,
+                            col_types = "cccicddddiiidddddddc"
+                        )
+                    }
+                )
+            ) %>% mutate(
+                sim = sim, 
+                n_particles = siminfo[1], 
+                gps_error = siminfo[3], 
+                system_noise = siminfo[2]
+            )
+        }
+    )
+)
+
+pdf('figures/04_model_results_dist.pdf', width = 6, height = 3)
+ggplot(
+    res %>% filter(event_dist_to_route < 20 & event_dist_to_route > 0), 
+    aes(event_dist_to_route)
+) + 
+    geom_density() +
+    xlab("Distance to path (m)") + ylab("Density")
+dev.off()
+
+
+sims <- res %>% filter(event_dist_to_route < 20)
+
+pdf('figures/04_model_results_neff.pdf', width = 14, height = 3)
+ggplot(sims %>% group_by(n_particles, gps_error, system_noise) %>%
+        summarize(Neff.hat = mean(Neff, na.rm = TRUE), 
+                  Neff.sd = sd(Neff, na.rm = TRUE) / sqrt(n()))) +
+    geom_pointrange(aes(as.integer(as.factor(gps_error)), #+ as.numeric(as.factor(n_particles)) / 10 - 0.25, 
+        100*Neff.hat / n_particles, 
+        ymin = 100*(Neff.hat - 2*Neff.sd) / n_particles, 
+        ymax = 100*(Neff.hat + 2*Neff.sd) / n_particles,
+        shape = as.factor(n_particles))) +
+    facet_grid(~system_noise) +
+    scale_x_continuous(
+        breaks = seq_along(unique(sims$gps_error)), 
+        labels = levels(as.factor(sims$gps_error))
+    ) +
[…]

### 2019-08-03 anzjs2018/sections/04_implementation.tex
Fix graphs

@@
     \centering
-    \includegraphics[width=0.7\textwidth]{figures/04_model_results_timing.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/04_model_results_timing.pdf}
     \caption{
@@
     \centering
-    \includegraphics[width=0.7\textwidth]{figures/04_model_results_dist.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/04_model_results_dist.pdf}
     \caption{
@@
 
-Figure~\ref{fig:neff} shows that $N_\text{eff}$ increased
+Figure~\ref{fig:perf_stats} shows that $N_\text{eff}$ increased
 with GPS error and decreased with system noise,
@@
 The degeneration rate decreased with GPS error and increased with $N$,
-and was unaffected by system noise,
-as seen in Figure~\ref{fig:degen_rate}.
+and was unaffected by system noise.
 Large GPS error affects the likelihood,
@@
     \centering
-    \includegraphics[width=0.9\textwidth]{figures/04_model_results_neff.pdf}
+    \includegraphics[width=\textwidth]{figures/04_model_results_neff.pdf}
     \caption{
@@
     }
-    \label{fig:neff}
+    \label{fig:perf_stats}
 \end{figure}
 
-\begin{figure}[tb]
-    \centering
-    \includegraphics[width=0.9\textwidth]{figures/04_model_results_degen.pdf}
-    \caption{
-        The degeneration rate is the percentage of iterations where there were no
-        plausible particles, so the vehicle was ``lost'' by the model.
-        This decreases as GPS error and number of particles increase,
-        but there is no visible effect of system noise.
-        Note, however, that some of these can be attributed to invalid data,
-        for example if the driver has entered the wrong trip direction.
-        The error bars show $\pm 2$SE.
-    }
-    \label{fig:degen_rate}
-\end{figure}
+% \begin{figure}[tb]
+%     \centering
+%     \includegraphics[width=0.9\textwidth]{figures/04_model_results_degen.pdf}
+%     \caption{
+%         The degeneration rate is the percentage of iterations where there were no
+%         plausible particles, so the vehicle was ``lost'' by the model.
+%         This decreases as GPS error and number of particles increase,
+%         but there is no visible effect of system noise.
+%         Note, however, that some of these can be attributed to invalid data,
+%         for example if the driver has entered the wrong trip direction.
+%         The error bars show $\pm 2$SE.
+%     }
+%     \label{fig:degen_rate}
+% \end{figure}
 
@@
 each individual simulation and variance over all simulations.
+If $\tilde z_\ell^{\sigma,\epsilon}$ is a vector of all travel times
+along segment $\ell$ during the simulation with GPS error $\sigma$ and system noise $\epsilon$,
+then the relative variance $\bar v^{\sigma,\epsilon}$ is calculated by
+\begin{equation*}
+\begin{split}
+v_\ell^{\sigma,\epsilon} &=
+\frac{\mathrm{Var}(\tilde z_\ell^{\sigma,\epsilon})}{\mathrm{Var}(\tilde z_\ell)} \\
+\bar v^{\sigma,\epsilon} &=
+    \frac{1}{L}\sum_{\ell=1}^L v_\ell^{\sigma,\epsilon}
+\end{split}
+\end{equation*}
 Figure~\ref{fig:travel_times} shows that simulations with larger GPS error

### 2019-08-03 anzjs2018/figures/04_model_results.R

@@ -20,32 +20,52 @@ times <- do.call(bind_rows, lapply(list.files("../../transitr/simulations", patt
 date <- format(times$timestamp[1], "%Y-%m-%d")
 trange <- as.POSIXct(paste(date, c("13:30", "14:00")))
 
-olvls <- c("all", "predicting ETAs", "updating vehicle states", "writing ETAs to protobuf feed")
 ttot <- times %>% group_by(sim, iteration) %>%
-    summarize(cpu = sum(cpu), wall = sum(wall), timestamp = first(timestamp), nvehicles = first(nvehicles),
-        n_particles = first(n_particles), gps_error = first(gps_error), system_noise = first(system_noise)) %>%
+    summarize(
+        cpu = sum(cpu),
+        wall = sum(wall), 
+        timestamp = first(timestamp), 
+        nvehicles = first(nvehicles),
+        n_particles = first(n_particles), 
+        gps_error = first(gps_error), 
+        system_noise = first(system_noise)
+    ) %>%
     mutate(what = "all")
 tsmry <- times %>% bind_rows(ttot) %>%
     filter(timestamp >= trange[1] & timestamp <= trange[2]) %>%
     group_by(sim, what) %>%
-    summarize(cpu = mean(cpu), wall = mean(wall),
-        n_particles = first(n_particles), gps_error = first(gps_error), system_noise = first(system_noise)) %>%
-    ungroup() %>%
-    filter(what %in% olvls)
+    summarize(
+        cpu = mean(cpu), 
+        wall = mean(wall),
+        n_particles = first(n_particles), 
+        gps_error = first(gps_error), 
+        system_noise = first(system_noise)
+    ) %>%
+    ungroup() #%>%
+    #filter(what %in% olvls)
+olvls <- sort(unique(tsmry$what)) #c("all", "predicting ETAs", "updating vehicle states", "writing ETAs to protobuf feed")
 
 # ggplot(tsmry) +
 #     geom_point(aes(n_particles, cpu, color = what)) +
 #     ylab("Wall time (seconds)")
 
-ordlvls <- olvls[c(3,2,4,1)]
-lvls <- c("Updating Vehicle States", "Predict ETAs", "Write ETAs to file", "Total")
+ordlvls <- olvls[c(2, 5, 6, 4, 3, 7, 1)]
+lvls <- c(
+    "Load Vehicle Data",
+    "Update Vehicle Information",
+    "Update Vehicle States", 
+    "Update Network States",
[…]

### 2019-08-04 anzjs2018/sections/02_gtfs.tex
Fix figure captions + labels

@@
     \caption{
-        Constructing a route network from the stops (dots) of three separate
-        routes (lines), as shown in (a).
-        Routes using the same subsequence of stops are assigned the same
-        road segments, so the road network shown 
-        consists of nodes (stops) and edges (roads),
-        as shown in (b), where the thickness of each lines
-        corresponds to how many routes share that road.
+        Bus stops (shown as dots above) can be serviced by more than one route, 
+        as shown in (a), with different line types representing three separate routes.
+        The constructed road network, using stops as nodes, is shown in (b),
+        in which the connecting lines represent physical road segments,
+        with line width proportional to the number of individual routes 
+        travelling along that road.
     }

### 2019-08-04 anzjs2018/sections/04_implementation.tex
Fix figure captions + labels

@@
     \caption{
-        Application timings were recorded for each component of each
-        iteration, and averages over a 15~minute window are shown here.
-        The timings shown are \emph{wall time}, the actual time passed,
-        not CPU time.
-        Steps that only involve particle calculations are approximately
-        of complexity $O(N)$,
-        however some steps are a little more complex (writing ETAs involves computing quantiles), 
-        so there is a slightly exponential trend.
+        The real-time running time of each step was recorded within each iteration,
+        and the average times over a 15~minute off-peak window are shown above.
+        Total time increases as the number of particles $N$ increases.
+        The reason for the large contribution of writing ETAs is that
+        this step involves estimation of quantiles,
+        which requires sorting the particles, 
+        an operation of $\mathcal{O}(N\log N)$ complexity.
+        The vehicle update step also involves sorting whenever the sample
+        degenerations and resampling is required.
+        Total time also includes fetching the data from the API,
+        which is constant for any $N$, which explains the increase
+        in standardised total time for $N=500$.
     }
-        % The timings for various parts of the application, and overall. %
-        % The trend is slightly non-linear due to the complexity of the ETA generation step, %
-        % which requires sorting of the particle ETAs to obtain quantile estimates.}
     \label{fig:timings}

### 2019-08-04 anzjs2018/figures/03_particle_filter.R

@@ -33,7 +33,7 @@ d0 <- ggplot() +
     geom_path(aes(x = xx, y = yy), data = tibble(xx = c(8000, 15000), yy = c(0, 0))) +
     theme_minimal() + 
     theme(panel.grid = element_blank(), axis.text = element_blank()) +
-    xlab("Distance Travelled") + ylab("") +
+    xlab("Distance travelled (meters)") + ylab("") +
     xlim(8000, 15000)
 
 h <- function(x, shape = shapef) {

### 2019-08-05 anzjs2018/sections/04_implementation.tex
Update captions

--- 2019-08-03 anzjs2018/sections/04_implementation.tex
@@
     \caption{
-        Application timings were recorded for each component of each
-        iteration, and averages over a 15~minute window are shown here.
-        The timings shown are \emph{wall time}, the actual time passed,
-        not CPU time.
-        Steps that only involve particle calculations are approximately
-        of complexity $O(N)$,
-        however some steps are a little more complex (writing ETAs involves computing quantiles), 
-        so there is a slightly exponential trend.
+        The average real-time running time of each step within each iteration
+        over a 15~minute off-peak window.
+        The vehicle update and writing ETA steps involve sorting the particles,
+        which is an operation of $\mathcal{O}(N\log N)$ complexity.
+        Total time is for the entire iteration, which includes fetching the data from the API.
+        % Total time increases as the number of particles $N$ increases.
+        % The reason for the large contribution of writing ETAs is that
+        % this step involves estimation of quantiles,
+        % which requires sorting the particles, 
+        % The vehicle update step also involves sorting whenever the sample
+        % degenerations and resampling is required.
+        % Total time also includes fetching the data from the API,
+        % which is constant for any $N$, which explains the increase
+        % in standardised total time for $N=500$.
     }
-        % The timings for various parts of the application, and overall. %
-        % The trend is slightly non-linear due to the complexity of the ETA generation step, %
-        % which requires sorting of the particle ETAs to obtain quantile estimates.}
     \label{fig:timings}
@@
     \caption{
-        The distance between the observation and the nearest point on the shape
-        is calculated by finding the shortest distance between point and path.
-        This gives an approximate idea of GPS error, however it should be noted
-        that many observations are made automatically at way points,
-        which accounts for the peak at 1m.
-        Note also that these observations have been truncated at 20m.
+        The distribution of minimum distance between observed vehicle location
+        and the route path, truncated to 20~meters.
     }
@@
     \caption{
-        The effective sample size is an approximate measure of the variance
-        of the particle weights.
-        When all particles have equal weights, $N_\text{eff} = N$,
-        while if all the weight is on one particle, $N_\text{eff} = 1$. 
-        The figure is subset by system noise, which shows how $N_\text{eff}$ 
-        decreases as system noise increases,
-        while increasing GPS error (on the $x$-axis) increases $N_\text{eff}$.
-        Increasing $N$ has a negative effect on $N_\text{eff}$ but this is
-        only evident when $N_\text{eff}$ is larger.
-        The error bars show $\pm 2$SE.
+        The effective sample size, $N_\text{eff}$,
+        and degeneration rate for varying values of GPS error,
+        system noise, and number of particles $N$.
+        % increases with GPS error and decreases with system noise,
+        % however both effects are reduced for larger $N$.
+        % Conversely, degeneration rate decreases with GPS error and
+        % system noise, although the effect of the latter is very small,
+        % and again the relationship is less pronounced for larger $N$.
     }
@@
     \caption{
-        The variability of travel time within a segment increases with GPS error,
-        and decreases slightly with increasing number of particles.
-        There is no visible effect of system noise on travel time variability.
-        The error bars show $\pm 2$SE.
+        The relative variability of travel time estimates for varying
+        values of GPS error, system noise, and number of particles, $N$,
+        with error bars $\pm 2$ standard errors.
+        % as GPS error increases,
+        % with no clear relationship with system noise.
+        % Increasing $N$ decreases variability.
     }

### 2019-08-07 anzjs2018/sections/01_introduction.tex
Fix typo
Update end edit words

@@
 
-In public transport, \rt information (RTI), such as estimated times of arrival (ETAs),
-keeps transport users informed about their journey,
-and enabled them to make decisions about the best route.
-It has also been shown that perceived waiting time is less
-when arrival time information is available \citep{TCRP_2003b}.
-Unfortunately, here in Auckland, RTI is highly unreliable,
-and succumbs to several problems that public transport users
-are all too familiar with.
-Inaccurate ETAs are a major source of frustration,
-as they tend to either increase or decrease over time 
-as the bus gets progressively later or earlier due to traffic congestion, 
-or simply due to poor schedule calibration.
+In public transport, \rt information (RTI)---%
+most notably estimated times of arrival (ETAs)---%
+keeps transport users informed of changes to their commute,
+and plan their journey accordingly.
+Previous research has shown that perceived waiting time is less
+when arrival time information is available \citep{TCRP_2003b},
+but here in Auckland, RTI is highly unreliable,
+resulting in frustration and deterence of public transport users.
+Inaccurate ETAs are a major source of this frustration,
+as they fluctuate---sometimes dramatically---over time.
+The cause of this can be complex, such as unexpected traffic congestion,
+or simply because of poor schedule calibration.
 Furthermore, buses are shown as \emph{on-time} 
-if they are not present in the \rt system and have not been cancelled,
-so ETAs are based solely on scheduled arrival times;
-if the bus does not show up on time
+if they are not present in the \rt system and have not been manually cancelled,
+in which case ETAs are based solely on scheduled arrival times.
+Once the scheduled arrival time has passed,
 the service is removed from the \rt board,
 leaving passengers unsure as to when---if at all---their bus will show up,
-a phenomenon termed by bus users as ``ghost buses''.
+a phenomenon referred to as ``ghost buses'' by transport bloggers.
 
@@
 \citep{Cathey_2003,Jeong_2005,Yu_2011,Hans_2015},
-in many public transport systems no formal model is used.
-Instead, ETAs are solely based on the scheduled arrival time
-adjusted by the vehicle's delay at the most recent stop, if available.
-This assumes that the schedule is valid,
-and that there is no unusual congestion along the route,
-neither of which are valid assumptions,
+many public transport providers---notably Auckland Transport---useno formal model is used.
+Instead, ETAs are based solely on the scheduled arrival time
+adjusted by the vehicle's arrival or departure delay at the most recent stop, 
+\emph{if available},
+the primary assumption being that the schedule is valid,
+and secondly that there is no unusual congestion along the route;
+neither of these are valid,
 particularly in our test area of Auckland, New Zealand,
-where infrastructure for buses (such as priority lanes) is limited.
+where infrastructure for buses (such as priority lanes) is limited,
+and bus drivers, in general, 
+do not actively adhere to the schedule.
 
-To avoid the problems mentioned above,
-a more robust modelling and prediction framework is required.
-This should consist of a robust vehicle model to estimate the position and speed
+A more robust modelling and prediction framework 
+based on \rt congestion information would avoid making these assumptions.
+Such a framework should consist of a robust vehicle model to estimate the position and speed
 of transit vehicles from \rt GPS data,
-and a way to incorporate road congestion information 
-into the arrival time predictions.
+and secondly a means of combining speed information from vehicles
+to model traffic flows,
+which can then be used to improve arrival time predictions.
 Several vehicle modelling approaches were explored, 
@@
 machine learning models \citep{Yu_2006,Chang_2010},
-and a particle filter \citep{Hans_2015}.
-We chose to use a particle filter for several reasons,
-which are discussed in Section~\ref{sec:pf}.
+and the particle filter,
+which has successfully been used in vehicle tracking applications
+\citep{cn},
+as well as in transport modelling \citep{Hans_2015}.
+For several reasons, discussed further in Section~\ref{sec:pf},
+the particle filter was used in the current application.s
 
-The next step involves estimating traffic conditions throughtout the transport network
-using the information obtained from the GPS data. 
+The second component involves estimating traffic conditions throughout the transport network
+using the information obtained from the vehicle model.
 \cite{Yu_2010} improved prediction accuracy by using the travel times
@@
 First, however, we descibe how we construct a
-road network from GTFS data in Section~\ref{sec:gtfs}.
+transit road network from GTFS data in Section~\ref{sec:gtfs}.

### 2019-08-07 anzjs2018/sections/02_gtfs.tex
Update end edit words

@@
 It provides a standardised format for sharing vehicle positions and trip delays,
-and are typically accessed by developers via an API can be used in \rt applications.
+and are typically accessed by developers via an API that can be used in \rt applications.
 
 As mentioned in Section~\ref{sec:intro},
-there are some major issues with the current prediction method in Auckland.
-These are almost directly attributed to \emph{GTFS-realtime} trip updates,
-which are currently the sole source of data for ETAs.
-Trip updates are reported whenever a vehicle arrives or departs a stop,
+there are some major issues with the arrival time prediction method currently
+deployed in Auckland,
+which is based entirely on \emph{GTFS-realtime} trip updates---%
+vehicle positions are used only to display \emph{where} the bus is.
+Trip updates are reported whenever a vehicle arrives at or departs from a stop,
 and includes the delay between the scheduled and actual arrival times,
 which is propagated to all future stops to update their ETAs.
-This assumes that the schedule is well calibrated 
-and the scheduled travel time between stops
+As already mentioned, this assumes that the schedule is well calibrated,
+and that the time between stops
 is representative of the real-world travel time between them. 
-Therefore, an alternative estimation approach that uses \rt travel times 
+Figure~\ref{fig:gtfs-delays} demonstrates the fallibility of the current approach,
+where we see the progressive lateness causing the passengers' ETA to 
+decrease much slower than expected.
+An estimation approach using \rt travel times 
 between stops would seem more adequate.
@@
 the travel time along roads between where the bus is \emph{now},
-and the stop at which the passenger is waiting.
+and the stop at which a passenger is waiting.
 In most applications, however, this vital information is unavailable, at least directly.
@@
 we constructed a \emph{road network},
-allowing us to model each physical road as a unique entity.
-In this way, every bus travelling along a road contributes to its state,
-and can be used to predice travel times for all upcoming buses 
+allowing us to model roads directly,
+such that every bus travelling along a road contributes to its state,
+which can in turn be used to predice travel times for all upcoming buses 
 travelling along the road,
@@
 
-The transit network was constructed by splitting each route
-into spatially identifiable segments,
+To construct the transit network,
+routes were split into spatially identifiable segments,
 each representative of a physical road,
-by using bus stops as nodes in the network
+using bus stops as nodes in the network
 and the connecting roads as edges
@@
 In this way, routes that service the same subsequence of stops
-all contribute to congestion information for the connecting roads.
+all contribute to traffic flow information for the connecting roads.
 Although there are several drawbacks to this method,
 such as road segment overlaps where routes merge between stops,
-this is minimal and our approach is sufficient for the current work.
+our approach is sufficient for the purposes of assessing the \rt 
+faesibilty of our proposed prediction framework.
 
@@
 these positions can be preemptive,
-for example when approaching a queue of traffic at an intersection
-and the way-point is triggered before the bus physically gets there,
-so consecutive observations can show a bus travelling backwards.
+for example when approaching a queue of traffic at an intersection,
+the way-point may be triggered before the bus physically gets there,
+so consecutive observations may show what appears to be a bus travelling backwards.
 To handle this, we compute the approximate distance travelled, $\tilde x_k$,
 of the vehicle by finding the nearest point on the path to the observation;
-if this has decreased, the current state is rejected and the vehicle reverted
+if this has decreased, the current state (based on the preemtive observation)
+is rejected and the vehicle reverted
 to its previous state before continuing.
-We are available to avoid degeneration of the particle filter (Section~\ref{sec:model_perf})
-at the cost of losing one observation that is most likely invalid anyway.
+So at the cost of losing one (likely invalid) observation,
+we are able to improve particle filter performance
+when this situation is encountered.

### 2019-08-07 anzjs2018/sections/03_models.tex
Update end edit words

@@
 an $n\times1$ column vector,
-and system noise $\boldsymbol{\omega}_k\sim\mathrm{N}(\boldsymbol{0},\mathcal{Q})$ 
+with system noise $\boldsymbol{\omega}_k\sim\mathrm{N}(\boldsymbol{0},\mathcal{Q})$ 
 of which observations $\boldsymbol{\mathcal{Y}}_k$,
@@
 while the measurement function $h:\mathbb{R}^n\mapsto\mathbb{R}^m$ is a deterministic function
-mapping the underlying state to the observable state.
+mapping the underlying state to the observable state,
+of which noisy measurements are made.
 
@@
 The first, implemented using a particle filter, estimates vehicle states
-to infer travel times along roads from a sequence of GPS positions;
+with the primary goal of estimating road travel times from a sequence of GPS positions;
 the second uses these travel times to update road states 
@@
 its speed $\dot x$ in meters per second.
-% The travel times along road segments, $\bz = (z_1,\ldots,z_L)^\top$, in seconds,
-% are estimated sequentially as the vehicle traverses the route,
-% and are not included explicitely in the state.
 These are combined into the state vector
 $\bX_k = \left[x_k\ \dot x_k\right]^\top$,
-of which observations $\bY_k$ are made using a GPS at time $t_k$
-with GPS error $\epsilon^2$,
+of which observations $\bY_k$ are made using a GPS
+with GPS error $\omega^2$,
 giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle
@@
 transit vehicle modelling applications \citep{Hans_2015},
-as well as providing an easy way to estimate travel times
+as well as allowing estimation of travel times
 $\bz = (z_1,\ldots,z_L)^\top$, in seconds, along road segments
 as the vehicle traverses the route.
-The primary advantage of a particle filter is its handling of multimodality,
+The primary advantage of the particle filter is its handling of multimodality,
 as demonstrated in Figure~\ref{fig:pf_state_predict},
@@
 on the physical distance between the vehicle observation and the state's
-estimate of the true location, $h(\bX_k)$ (Section~\ref{sec:pf_update}),
-rather than the reverse of first estimating the ``observed'' distance travelled.
+estimate of the true location, $h(\bX_k)$ (Section~\ref{sec:pf_update}).
 Conversely, particle filter methods are computationally demanding,
-requiring an increasing number of particles as the model complexity and
+requiring an increasing number of particles as model complexity and
 number of parameters increases \citep{Carpenter_1999}.
-Section~\ref{sec:rt} describes our implementation of the particle filter in real-time 
-and its timings.
+Section~\ref{sec:rt} describes our implementation of the particle filter
+and its \rt performance.
 
@@
 each with an associated weight $W_{k-1}^{(i)}$.
-The state can then be expressed using the Dirac delta function $\dot\delta_X(x)$ \citep{cn},
+The state can then be expressed using the Dirac delta function $\dot\delta(\cdot)$,
 so that
@@
 p(\bX_{k-1} | \bY_{1:k-1}) \approx 
-    \sum_{i=1}^N W_{k-1} \dot\delta_{\bX_{k-1}}(\bX_{k-1}^{(i)})
+    \sum_{i=1}^N W_{k-1} \dot\delta(\bX_{k-1} - \bX_{k-1}^{(i)})
 \end{equation*}
@@
 p(\bX_k | \bX_{k-1}) \approx 
-    \sum_{i=1}^N W_{k-1} \dot\delta_{\bX_{k}}(f(\bX_{k-1}^{(i)}, \psi))
+    \sum_{i=1}^N W_{k-1} \dot\delta(\bX_{k} - f(\bX_{k-1}^{(i)}, \boldsymbol{\psi}))
 \end{equation*}
-using parameter vector $\psi$ containing all of the necessary parameters
+using parameter vector $\boldsymbol{\psi}$ containing all of the necessary parameters
 for the model (including system noise).
@@
 p(\bX_k | \bY_k) \approx  
-    \sum_{i=1}^N W_{k} \dot\delta_{\bX_{k}}(\bX_{k}^{(i)})
+    \sum_{i=1}^N W_{k} \dot\delta(\bX_{k} - \bX_{k}^{(i)})
 \end{equation*}
@@
 due to the particles becoming too diverse
-and all of the weight falling onto only a small number of them.
+and all of the weight falls onto only a small number of them.
 This occurs naturally over time, but the rate at which it occurs depends 
@@
 which is of $\mathcal{O}(N\log N)$ complexity,
-so we can reduce computational costs by only resampling when required.
-To prevent degeneration of the particle filter state,
-the sample is resampled whenever the
-effective sample size $N_{\text{eff}} = 1 / \sum_i (W_k^{(i)})^2$
+so to reduce computational costs, resampling is performed only when
+the effective sample size $N_{\text{eff}} = 1 / \sum_i (W_k^{(i)})^2$
 falls below a specified threshold $N_{\text{thres}}$.
@@
 for the current work.
-\textbf{Check this}.
 
@@
 The particle filter allows us to flexibly model bus behaviour,
-which in our current model includes
-\begin{itemize}
+which in the current model includes
+\begin{enumerate}
 \item non-constant speed along roads (between observations), and
@@
     while passengers board and disembark.
-\end{itemize}
-To handle the above, the transition function consists of two models,
+\end{enumerate}
+The transition function consists of two models,
 one for each of the above behaviours.
@@
 by using Newton's Laws of Motion,
-which, letting $\delta = t_k - t_{k-1}$ and $\epsilon_k^{(i)}\sim\mathrm{N}(0, \sigma^2)$, is
+which, letting $\delta_k = t_k - t_{k-1}$ and $\epsilon_k^{(i)}\sim\mathrm{N}(0, \sigma^2)$, 
+truncated to ensure $\dot x_{k-1}^{(i)} + \epsilon_k^{(i)} > 0$,
+is
 \begin{equation}
@@
     \begin{bmatrix}
-        x_{k-1}^{(i)} + \delta(\dot x_{k-1}^{(i)} + \epsilon_k^{(i)}) \\
+        x_{k-1}^{(i)} + \delta_k(\dot x_{k-1}^{(i)} + \epsilon_k^{(i)}) \\
         \dot x_{k-1}^{(i)} + \epsilon_k^{(i)}
@@
 The second vehicle behaviour requires a model of bus stop wait times.
-We assume the vehicle is approaching stops $j$,
-and will stop with some probability $\pi_j$.
+When the vehicle is approaching stop $j$,
+it stops with some probability $\pi_j$.
 If it does not stop, the remaining travel time $\delta_k$ is unaltered;
@@
 accounts for deceleration, opening and closing of doors, and acceleration
-\citep{Hands_2015}---and then waits while passengers board and disembark,
+\citep{Hans_2015}---and then waits while passengers board and disembark,
 which is assumed to be exponentially distributed with mean $\tau_j$.
@@
 While (\ref{eq:trans_stop}) is used directly in the algorithm,
-(\ref{eq:trans_newton}) is modified to allow the bus to approach a stop
+(\ref{eq:trans_newton}) is evaluated iteratively,
+allowing the bus to approach a stop
 and conditionally wait before continuing,
@@
 $p(\bY_k|\bX_k^{(i)})$,
-the measurement function $h$ is needed along with an
-\emph{geographical projection} $g$ to allow comparison of $\bY_k$,
+the measurement function $h$ is needed along with 
+\emph{geographical projection} function $g$ to allow comparison of $\bY_k$,
 a GPS observation, with $\bX_k$, a distance travelled along the route in meters.
 
-The measurement function $h$ computes the GPS coordinates by using the 
+The measurement function $h$ computes GPS coordinates using the 
 shape information provided by GTFS and travelling $x_k$ meters along 
@@
 which allows us to express $\br$ in terms of two independent
-standard normal random variables $z_1, z_2 \sim \mathrm{N}(0,1)$,
-which lets us express the distance between a vehicle observation $\bY_k$
-and a state $\bX_k$ using (\ref{eq:dist}) and the measurement function $h$,
+standard normal random variables $z_1, z_2 \sim \mathrm{N}(0,1)$.
+Now the distance between a vehicle observation $\bY_k$
+and a state $\bX_k$ using (\ref{eq:dist}) and the measurement function $h$
+is expressed as
 \begin{equation}
@@
 The likelihood of the observation $\bY_k$ given a state estimate $\bX_k^{(i)}$
-can now be expressed using (\ref{eq:obs_exp}) and the PDF of the exponential distribution,
+can now be expressed using (\ref{eq:obs_exp}) 
+and the probability density function of the exponential distribution,
 \begin{equation}
@@
 where the distance between two GPS coordinates is easily computed,
-making (\ref{eq:lhood}) easy and fast to evaluate within the particle filter algorithm.
+allowing for fast evaluation of (\ref{eq:lhood}) within the particle filter algorithm.
 
@@
 possible due to the discrete nature of the particle filter state estimate;
-in the Kalman filter, which has often been used in vehicle tracking,
-the measurement \emph{matrix} is used, and required that there is a linear
+in the Kalman filter, which has often been used in transit vehicle tracking,
+the measurement \emph{matrix} is used, and requires a linear
 transformation between the state and its observations.
@@
 (travel time) along roads in the transit network, 
-as well as make short-term predictions for estimating arrival times.
+and make short-term forecasts for estimating arrival times.
 In this paper, we model each road segment independently,
-as including correlations simplifies the model to not only a one-dimensional Kalman filter,
+as excluding correlations simplifies the model to not only a one-dimensional Kalman filter,
 but it also means computations can be run in parallel,
@@
 with $f$ and $h$ both unity,
-system noise $v_c^\ell \sim \mathrm{N}(0, \nu^2)$
+system noise $v_c^\ell \sim \mathrm{N}(0, \nu_\ell^2)$,
 with $\Delta_c = t_c - t_{c-1}$
-and $\nu_\ell^2$ is the variance of travel times per second,
+where $\nu_\ell^2$ is the average change in travel time per second
+along road segment $\ell$,
 and the observation error $e_c^{\ell m} \sim \mathrm{N}(0, R_c^{\ell m})$:
@@
 \emph{information matrix} and \emph{information vector} are used in place of 
-the covariance matrix and state vector, respectively.
-This allows multiple observations to be added together to update the state
-in a single iteration,
-which is the case when multiple vehicles travel along a road at the same time.
+the covariance matrix and state vector, respectively,
+allowing multiple observations to be added together to update the state
+in a single iteration.
 
@@
 for upcoming buses. 
+It should be noted that, if segment correlations were to be included,
+the full state $\boldsymbol{\theta}_c$ would need updating in a single step,
+which would require inverting a large $L\times L$ matrix,
+significantly increasing the time of updating the state.

### 2019-08-07 anzjs2018/sections/04_implementation.tex
Update end edit words

@@
 and assess its performance 
-both with respect to timing and parameter estimation.
+both with respect to timing and travel time estimation.
 
@@
 since each vehicle is modelled independently so it is trivial to scale up without
-issues with thread safety.
+thread safety concerns.
 
@@
 In each iteration, 
-the timing of the various program components was recorded.
+the timings of the various program components were recorded.
 Since the number of vehicles travelling at any given time changes throughout the day,
@@
 with complexity $O(N \log N)$,
-which explains the slightly quadratic trend in the timings.
+which explains non-linear relationship with $N$.
 
@@
 although this may increase once a more comprehensive model is developed.
-On our 8-core virtual machine the entire process is completed within 20~seconds
+On our 8-core virtual machine,
+the average time to process one iteration during off-peak is under 20~seconds
 when 8000~particles are used.
+Figure~\ref{fig:day-timings} shows iteration timings over the course of a day.
 
@@
 To evaluate the performance of the model,
-the simulation was run using a range of values of GPS error, $\epsilon$,
+the simulation was run using a range of values of GPS error, $\omega^2$,
 and system noise, $\sigma^2$.
@@
 between each observation and the route using a nearest point algorithm,
-which led us to use the values $\epsilon \in \{1,2,3,5\}$ for GPS error in meters.
+which suggested using values $\omega \in \{1,2,3,5\}$ for GPS error in meters.
 For the system noise we used values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$,
-which correspond to an average vehicle speed variance of between 0.1 and 45~meters per second
+which correspond to an average vehicle speed variation of between 0.1 and 45~meters per second
 over a period of 30~seconds.
-For each of combination of these parameter values, including varying $N$,
+For each combination of these parameter values, including varying $N$,
 the following values were calculated:
@@
 
-Figure~\ref{fig:perf_stats} shows that $N_\text{eff}$ increased
-with GPS error and decreased with system noise,
+Figure~\ref{fig:perf_stats} shows that $N_\text{eff}$ increases
+with GPS error and decreases with system noise,
 while remaining unaffected by changing $N$.
-The degeneration rate decreased with GPS error and increased with $N$,
-and was unaffected by system noise.
+The degeneration rate decreases with GPS error and increases with $N$,
+but is mostly unaffected by system noise.
 Large GPS error affects the likelihood,
@@
 Conversely, increasing system noise spreads out the particle cloud,
-so fewer particles will be near the vehicle.
+so fewer particles will be near the vehicle,
+decreasing $N_\text{eff}$,
+but more plausible trajectories are sampled,
+so we see a slight decrease in degeneration rate.
 Thus, these results are not surprising, 
@@
 so to compare the variability between simulations we needed to 
-compute the average variance of travel times over all simulations,
+compute the average variance of estimated travel times over all simulations,
 taking the average (over road segments) of the ratio between variance for
 each individual simulation and variance over all simulations.
-If $\tilde z_\ell^{\sigma,\epsilon}$ is a vector of all travel times
-along segment $\ell$ during the simulation with GPS error $\sigma$ and system noise $\epsilon$,
-then the relative variance $\bar v^{\sigma,\epsilon}$ is calculated by
+If $\tilde z_\ell^{\omega,\epsilon}$ is a vector of all travel times
+along segment $\ell$ during the simulation with GPS error $\omega$ and system noise $\epsilon$,
+then the relative variance $\bar v^{\omega,\epsilon}$ is calculated by
 \begin{equation*}
 \begin{split}
-v_\ell^{\sigma,\epsilon} &=
-\frac{\mathrm{Var}(\tilde z_\ell^{\sigma,\epsilon})}{\mathrm{Var}(\tilde z_\ell)} \\
-\bar v^{\sigma,\epsilon} &=
-    \frac{1}{L}\sum_{\ell=1}^L v_\ell^{\sigma,\epsilon}
+v_\ell^{\omega,\epsilon} &=
+\frac{\mathrm{Var}(\tilde z_\ell^{\omega,\epsilon})}{\mathrm{Var}(\tilde z_\ell)} \\
+\bar v^{\omega,\epsilon} &=
+    \frac{1}{L}\sum_{\ell=1}^L v_\ell^{\omega,\epsilon}
 \end{split}

### 2019-08-08 anzjs2018/sections/01_introduction.tex
more edits

@@
 machine learning models \citep{Yu_2006,Chang_2010},
-and the particle filter,
-which has successfully been used in vehicle tracking applications
-\citep{cn},
-as well as in transport modelling \citep{Hans_2015}.
-For several reasons, discussed further in Section~\ref{sec:pf},
-the particle filter was used in the current application.s
+and the particle filter \citep{Hans_2015}.
+
+The particle filter has proven itself as a robust option for
+\rt vehicle tracking applications
+\citep{Gustafsson_2002,Davidson_2011},
+and as computational power has increased over recent years,
+a strong competitor for \rt applications.
+\cite{Ulmke_2006} showed it to handle situations where traditional
+(i.e., Kalman filter) methods break down:
+the particle filter retained the accurate uncertainty about the location
+where there was a lack of information,
+an important property for our model in which we expect
+many situations where there are sparse observations
+and a strongly multimodal distribution.
+By using a particle filter,
+we were able to develop a simple, flexible framework
+which is robust to the type of data we expect to observe.
 

### 2019-08-08 anzjs2018/sections/02_gtfs.tex
more edits

@@
 Figure~\ref{fig:gtfs-delays} demonstrates the fallibility of the current approach,
-where we see the progressive lateness causing the passengers' ETA to 
-decrease much slower than expected.
+where we see the progressive lateness causing the passenger's ETA to 
+jump each time the bus arrives at a stop.
 An estimation approach using \rt travel times 
 between stops would seem more adequate.
+
+\begin{figure}[tb]
+    \centering
+    \includegraphics[width=0.9\textwidth]{figures/02_gtfs_delays.pdf}
+    \caption{
+        On the left, scheduled arrival times are shown as solid points,
+        for several stops along a route. The actual arrival times
+        are shown with crosses,
+        and the GTFS-based ETAs for stop 5, $\hat A_1$, $\hat A_2$, and $\hat A_3$, are also shown.
+        On the right,
+        for a passenger arriving at stop 5 at 12:04,
+        the ETA over time is shown, demonstrating the fluctuation in
+        arrival time as the bus arrives at each stop.
+    }
+    \label{fig:gtfs-delays}
+\end{figure}
 

### 2019-08-08 anzjs2018/sections/03_models.tex
more edits

@@
 as $\bY_k = \left[\lambda_k\ \phi_k \right]^\top$.
-
 We use a particle filter to estimate $\bX_k$,
@@
 transit vehicle modelling applications \citep{Hans_2015},
-as well as allowing estimation of travel times
+as well as many other more general vehicle tracking 
+applications \citep{}.
+In addition, it makes it trivial to estimate travel times
 $\bz = (z_1,\ldots,z_L)^\top$, in seconds, along road segments
 as the vehicle traverses the route.
+
 The primary advantage of the particle filter is its handling of multimodality,
@@
 
-The combination of (\ref{eq:trans_newton}) and (\ref{eq:trans_stop})
-is handled by the transition function algorithm,
-displayed in Algorithm~\ref{fig:algorithm}.
-While (\ref{eq:trans_stop}) is used directly in the algorithm,
-(\ref{eq:trans_newton}) is evaluated iteratively,
-allowing the bus to approach a stop
-and conditionally wait before continuing,
-until $\delta_k$ reaches zero.
+The model components are implemented as an algorithm,
+in which (\ref{eq:trans_newton}) is evaluated iteratively
+one second at a time, decreasing $\delta_k$ by one each time,
+with $\epsilon_k^{(i)}$ only sampled only at the beginning
+of each time update.
+This allows the particle to approach the next stop,
+at which point (\ref{eq:trans_stop}) is evaluated for the particle
+and the dwell time subtracted from $\delta_k$,
+and continuing until $\delta_k$ reaches zero.
+Iterating this way ensures that 
+the algorithm works regardless of how many
+stops are encountered during the time step.
 The algorithm also keeps track of each particle's travel time $z_\ell^{(i)}$
@@
 p(z_\ell | \bY_{1:k}) \approx
-    \sum_{i=1}^N W_k^{(i)} \dot\delta_{z_\ell}(z_\ell^{(i)}).
+    \sum_{i=1}^N W_k^{(i)} \dot\delta(z_\ell - z_\ell^{(i)}).
 \end{equation}
 
-\renewcommand{\algorithmicrequire}{\textbf{Start:}}
-\newcommand{\algorithmicbreak}{\textbf{break}}
-\newcommand{\Break}{\State \algorithmicbreak}
-\begin{algorithm}[t]
-    \caption{Particle mutation function.}
-    \label{fig:algorithm}
-    \begin{algorithmic}[2]
-    \Require $x_{k-1}, \dot x_{k-1}, \boldsymbol z$\\
-    Initialisation $\delta \gets t_k - t_{k-1}$, 
-        $j \gets stop\_index(x, \boldsymbol{S})$
-    \If {$u \sim \mathrm{U}(0,1) < 0.5$ and $x = S_j$}
-        \Comment{Remain longer at stop}
-        \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
-        \State $\delta \gets \delta - \gamma - d_j$
-    \ElsIf {$u \sim \mathrm{U}(0,1) < 0.1$}
-        \Comment{Occasionally bus stopped due to traffic}
-        \State $w \gets \min(\delta, \mathrm{Exp}(1/\delta))$
-        \State $\delta \gets \delta - w$
-        \State $z_j \gets z_j + w$
-        \Comment{Congestion counts as travel time}
-    \EndIf
-    \While {$\delta > 0$ and $x < S_J$} 
-        \State $\dot x \sim \mathrm{N}(\dot x, \sigma^2)$
-        \Comment{System noise, truncated to be non-negative}
-        \State $x \leftarrow x + \dot x$
-        \If {$x \geq S_{j+1}$}
-            \If {$j+1 = J$}
-                \Comment{Break at end of route}
-                \State $x \gets S_J$
-                \Break
-            \EndIf
-            \State $u_j \sim \mathrm{U}(0,1)$
-            \Comment{Simulate vehicle stopping}
-            \If {$u_j < \pi$}
-                \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
-                \State $\delta \gets \delta - \gamma - d_j$
-                \State $x \gets S_j$
-            \EndIf
-            \State $j\gets j+1$
-        \EndIf
-        \State $\delta \gets \delta - 1$
-        \State $z_j \gets z_j + 1$
-        \Comment{Track travel time along road}
-    \EndWhile
-    \\
-    \Return $x, \dot x, \boldsymbol z$
-    \end{algorithmic}
-\end{algorithm}
+% \renewcommand{\algorithmicrequire}{\textbf{Start:}}
+% \newcommand{\algorithmicbreak}{\textbf{break}}
+% \newcommand{\Break}{\State \algorithmicbreak}
+% \begin{algorithm}[t]
+%     \caption{Particle mutation function.}
+%     \label{fig:algorithm}
+%     \begin{algorithmic}[2]
+%     \Require $x_{k-1}, \dot x_{k-1}, \boldsymbol z$\\
+%     Initialisation $\delta \gets t_k - t_{k-1}$, 
+%         $j \gets stop\_index(x, \boldsymbol{S})$
+%     \If {$u \sim \mathrm{U}(0,1) < 0.5$ and $x = S_j$}
+%         \Comment{Remain longer at stop}
+%         \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
+%         \State $\delta \gets \delta - \gamma - d_j$
+%     \ElsIf {$u \sim \mathrm{U}(0,1) < 0.1$}
+%         \Comment{Occasionally bus stopped due to traffic}
+%         \State $w \gets \min(\delta, \mathrm{Exp}(1/\delta))$
+%         \State $\delta \gets \delta - w$
+%         \State $z_j \gets z_j + w$
+%         \Comment{Congestion counts as travel time}
+%     \EndIf
+%     \While {$\delta > 0$ and $x < S_J$} 
+%         \State $\dot x \sim \mathrm{N}(\dot x, \sigma^2)$
+%         \Comment{System noise, truncated to be non-negative}
+%         \State $x \leftarrow x + \dot x$
+%         \If {$x \geq S_{j+1}$}
+%             \If {$j+1 = J$}
+%                 \Comment{Break at end of route}
+%                 \State $x \gets S_J$
+%                 \Break
+%             \EndIf
+%             \State $u_j \sim \mathrm{U}(0,1)$
+%             \Comment{Simulate vehicle stopping}
+%             \If {$u_j < \pi$}
+%                 \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
+%                 \State $\delta \gets \delta - \gamma - d_j$
+%                 \State $x \gets S_j$
+%             \EndIf
+%             \State $j\gets j+1$
+%         \EndIf
+%         \State $\delta \gets \delta - 1$
+%         \State $z_j \gets z_j + 1$
+%         \Comment{Track travel time along road}
+%     \EndWhile
+%     \\
+%     \Return $x, \dot x, \boldsymbol z$
+%     \end{algorithmic}
+% \end{algorithm}
 
-\afterpage{\clearpage}
+% \afterpage{\clearpage}
 

### 2019-08-08 anzjs2018/sections/04_implementation.tex
more edits

@@
 On our 8-core virtual machine,
-the average time to process one iteration during off-peak is under 20~seconds
+the average time to process one iteration during off-peak is about 16~seconds
 when 8000~particles are used.
-Figure~\ref{fig:day-timings} shows iteration timings over the course of a day.
+% Figure~\ref{fig:day-timings} shows iteration timings over the course of a day.
 

### 2019-08-08 anzjs2018/sections/05_discuss.tex
more edits

\section{Discussion and Future Work}
\label{sec:discussion}

In this paper we described an approach to transit vehicle modelling
that enables the \rt estimation of road state from vehicle position data,
which will later be used to make arrival time predictions
that account for \rt traffic conditions.
The vehicle model, implemented using a particle filter,
allows complex bus behaviours to be modelled,
from which 
vehicle travel times along roads are estimated which are in turn used 
in the modelling of the road network state.

We have focussed on making a general framework
in which the existing models used for specific components
can easily be changed---%
for example, using a more complex model of stop dwell times
based on historical data in place of the exponential---%
or new ones added, such as implementing intersection-specific behaviour
if intersecion locations are available.
Future work is planned to detect route meeting points
and use these as intersection nodes,
with the goal of further reducing road segment overlap.

The network model presented here is simple, 
but provides the foundation for future work to
develop a formal model combining \rt travel time information with historical data, 
allowing short- and medium-term forecasts of road state,
for example incorporating peak time congestion.
% This in turn will provide the basis for an arrival time prediction model
% that combines \rt vehicle state with travel time forecasts
% in order to provide commuters with more reliable ETAs.
As for the arrival times, 
we plan to estimate these from a combination of historical 
and \rt data.
The model presented in this paper provides the starting point
for a transit modelling and predictive framework
that can be adapted to specific transit providers,
combined with historical data,
and run in \rt to provide commuters with
improved ETAs with which they can more reliably
plan their commutes.

The results of our simulations show that the framework is a 
faesible \rt alternative.
While the timings presented were for off-peak times,
we hope futher investigation of choosing $N$ and $N_\text{eff}$,
as well as work removing some redudancy from the transition algorithm,
will mean times during peak periods, when there are 2-3~times
as many vehicles, will come in under our 30~second target.

### 2019-08-08 anzjs2018/figures/04_model_results.R

@@ -111,7 +111,8 @@ egg::ggarrange(
         xlab("Number of particles") +
         ylab("Average iteration time (s)") + 
         coord_flip() +
-        theme_minimal(),
+        theme_classic() +
+        theme(strip.background = element_blank()),
     ggplot(figdat, aes((as.factor(n_particles)))) +
         geom_point(aes(y = 1000 * wall.hat / n_particles)) +
         geom_linerange(aes(ymin = 0, ymax = 1000 * wall.hat / n_particles)) +
@@ -119,10 +120,15 @@ egg::ggarrange(
         xlab("Number of particles") +
         ylab("Standardised time (ms / particle)") + 
         coord_flip() +
-        theme_minimal(),
+        theme_classic() +
+        theme(strip.background = element_blank()),
     nrow = 2
 )
 dev.off()
+
+# ## total times by time of day
+# ggplot(ttot, aes(timestamp, wall / 1000)) +
+#     geom_point()
 
 
 
@@ -158,9 +164,10 @@ res <- do.call(bind_rows, lapply(list.files("../../transitr/simulations", patter
 
 pdf('figures/04_model_results_dist.pdf', width = 6, height = 3)
 ggplot(res %>% filter(dist_to_path < 20 & dist_to_path > 0), aes(dist_to_path)) + 
-    geom_density(fill = "gray") +
+    geom_density() +
     xlab("Distance to path (m)") + ylab("Density") +
-    theme_minimal()
+    theme_classic() +
+    theme(strip.background = element_blank())
 dev.off()
 
 sims <- res %>% filter(dist_to_path < 20) %>% 
@@ -181,7 +188,8 @@ egg::ggarrange(
         facet_grid(~n_particles, 
             labeller = as_labeller(function(x) paste("N =", x))) +
         labs(lty = "System noise") +
-        theme_minimal() +
+        theme_classic() +
+        theme(strip.background = element_blank()) +
         xlab("GPS error (m)") +
         ylab(expression(bar(N)[eff]/N)),
     ggplot(sims, aes(gps_error, p_bad, 
@@ -193,7 +201,8 @@ egg::ggarrange(
             labeller = as_labeller(function(x) paste("N =", x))) +
         labs(lty = "System noise") +
         theme(legend.position = "none") +
-        theme_minimal() +
+        theme_classic() +
+        theme(strip.background = element_blank()) +
         xlab("GPS error (m)") +
[…]

### 2019-08-08 anzjs2018/sections/01_introduction.tex
changes ... oops

--- 2019-08-07 anzjs2018/sections/01_introduction.tex
@@
 keeps transport users informed of changes to their commute,
-and plan their journey accordingly.
+allowing them to plan their journey accordingly.
 Previous research has shown that perceived waiting time is less
@@
 but here in Auckland, RTI is highly unreliable,
-resulting in frustration and deterence of public transport users.
+leading to frustration and ultimately detering public transport users.
 Inaccurate ETAs are a major source of this frustration,
 as they fluctuate---sometimes dramatically---over time.
-The cause of this can be complex, such as unexpected traffic congestion,
-or simply because of poor schedule calibration.
+This can be caused by complex changes to traffic conditions,
+but equally can be due simply to poor schedule calibration.
 Furthermore, buses are shown as \emph{on-time} 

### 2019-08-09 anzjs2018/sections/01_introduction.tex
More edits

--- 2019-08-08 anzjs2018/sections/01_introduction.tex
@@
 as they fluctuate---sometimes dramatically---over time.
-This can be caused by complex changes to traffic conditions,
-but equally can be due simply to poor schedule calibration.
+While this may be caused by unpredictable changes to traffic conditions,
+it can equally simply be a result of poor schedule calibration.
 Furthermore, buses are shown as \emph{on-time} 
@@
 \citep{Cathey_2003,Jeong_2005,Yu_2011,Hans_2015},
-many public transport providers---notably Auckland Transport---useno formal model is used.
+many public transport providers---notably Auckland Transport---no formal model is used.
 Instead, ETAs are based solely on the scheduled arrival time
 adjusted by the vehicle's arrival or departure delay at the most recent stop, 
-\emph{if available},
-the primary assumption being that the schedule is valid,
+\emph{if available}.
+The primary assumption here is that the schedule is valid,
 and secondly that there is no unusual congestion along the route;
@@
 do not actively adhere to the schedule.
-
 A more robust modelling and prediction framework 
@@
 which can then be used to improve arrival time predictions.
+
 Several vehicle modelling approaches were explored, 
@@
 machine learning models \citep{Yu_2006,Chang_2010},
-and the particle filter,
-which has successfully been used in vehicle tracking applications
-\citep{cn},
-as well as in transport modelling \citep{Hans_2015}.
-For several reasons, discussed further in Section~\ref{sec:pf},
-the particle filter was used in the current application.s
+and the particle filter \citep{Hans_2015},
+which has proven itself as a robust option for
+\rt vehicle tracking applications
+\citep{Gustafsson_2002,Davidson_2011}.
+\cite{Ulmke_2006} compared used both a Gaussian sum approximation and a particle filter
+to track a vehicle in real-time,
+and although not their primary aim,
+showed how the particle filter behaved better in some difficult situations.
+In particular, the particle filter accurately estimated the \emph{uncertainty}
+of the vehicle's location when the observations were unavailable for a time,
+which is a situation that often occurs with transit data---%
+the bus may be stopped a a bus stop, or it could be continuing along the route but,
+for some other reason, failing to report its position.
+Choosing to implement a particle filter has allowed us
+to develop a simple, flexible framework
+which is robust to the type of data we expect to observe.
 
-The second component involves estimating traffic conditions throughout the transport network
+The second component of our framework 
+involves estimating traffic conditions throughout the transport network
 using the information obtained from the vehicle model.
 \cite{Yu_2010} improved prediction accuracy by using the travel times
-of buses travelling along the same route.
+of previous buses travelling along the same route.
 A similar method presented by \cite{Hans_2015}
@@
 
-In Section~\ref{sec:models} of this paper, 
-we describe the \rt vehicle and network state models
-used to estimate traffic congestion using the transit vehicles
-travelling through the network in real-time,
-and in Section~\ref{sec:rt} we demonstrate the feasibility of using a particle filter
-in \rt to model transit vehicles,
-and evaluate its performance.
-First, however, we descibe how we construct a
-transit road network from GTFS data in Section~\ref{sec:gtfs}.
+In this paper, we propose a simple framework for modelling the predicting
+transit vehicle arrival in real-time.
+Section~\ref{sec:gtfs} describes how we construct a transit network,
+which allows us to model travel times along physical roads,
+while Section~\ref{sec:models} presents two \rt models---%
+one for each of the vehicle and network states---%
+with the primary goal of estimating \rt traffic states throughout the network.
+Finally, in Section~\ref{sec:rt}, we discuss the \rt faesibility and performance
+of using a particle filter to model transit vehicles in real-time.

### 2019-08-09 anzjs2018/sections/02_gtfs.tex
More edits

--- 2019-08-07 anzjs2018/sections/02_gtfs.tex
@@
 and includes the delay between the scheduled and actual arrival times,
-which is propagated to all future stops to update their ETAs.
+which is propagated to all future stops to update their ETAs,
+as demonstrated in Figure~\ref{fig:gtfs-delays}.
 As already mentioned, this assumes that the schedule is well calibrated,
@@
 is representative of the real-world travel time between them. 
-Figure~\ref{fig:gtfs-delays} demonstrates the fallibility of the current approach,
-where we see the progressive lateness causing the passengers' ETA to 
-decrease much slower than expected.
-An estimation approach using \rt travel times 
-between stops would seem more adequate.
+Figure~\ref{fig:gtfs-delays} demonstrates the fallibility of this prediction method,
+where we see the progressive lateness causing the passenger's ETA to 
+jump each time the bus arrives at a stop.
+By using \rt travel time information,
+we can better predict the travel time between stops,
+which we hope will result in more accurate and reliable arrival time estimates.
+
+\begin{figure}[tb]
+    \centering
+    \includegraphics[width=0.8\textwidth]{figures/02_gtfs_delays.pdf}
+    \caption{
+        On the left, scheduled arrival times are shown as solid points,
+        for several stops along a route. The actual arrival times
+        are shown with crosses,
+        and the GTFS-based ETAs for stop 5, $\{\hat A_1, \hat A_2, \hat A_3\}$, are shown.
+        On the right,
+        for a passenger arriving at stop 5 at 12:04,
+        ETA predictions
+        $\mathrm{ETA}_j = \mathrm{Time} - \hat A_j$ are shown
+        to demonstrate the fluctuation in
+        arrival time as the bus arrives at each stop.
+    }
+    \label{fig:gtfs-delays}
+\end{figure}
 
@@
 such that every bus travelling along a road contributes to its state,
-which can in turn be used to predice travel times for all upcoming buses 
+which can in turn be used to predict travel times for all upcoming buses 
 travelling along the road,
@@
         \centering
-        \includegraphics[width=0.95\textwidth]{figures/02_network_segments_1.pdf}
+        \includegraphics[width=0.9\textwidth]{figures/02_network_segments_1.pdf}
         \caption{Raw GTFS route shapes}
@@
         \centering
-        \includegraphics[width=0.95\textwidth]{figures/02_network_segments_2.pdf}
+        \includegraphics[width=0.9\textwidth]{figures/02_network_segments_2.pdf}
         \caption{GTFS-based road network}
@@
 of the vehicle by finding the nearest point on the path to the observation;
-if this has decreased, the current state (based on the preemtive observation)
+if this has decreased, the current state (based on the preemptive observation)
 is rejected and the vehicle reverted

### 2019-08-09 anzjs2018/sections/03_models.tex
More edits

@@
 of which observations $\bY_k$ are made using a GPS
-with GPS error $\omega^2$,
+with error $\omega^2$,
 giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle
@@
 We use a particle filter to estimate $\bX_k$,
-due to its high flexibility and use in recent 
-transit vehicle modelling applications \citep{Hans_2015},
-as well as many other more general vehicle tracking 
-applications \citep{}.
+due to its flexibility and proven robustness
+in recent vehicle modelling applications \citep{Ulmke_2006,Hans_2015}.
 In addition, it makes it trivial to estimate travel times
@@
     \centering
-    \begin{subfigure}[t]{0.9\textwidth}
+    \begin{subfigure}[t]{1\textwidth}
         \centering
@@
         }
-        % The vehicle state represented by a sample of discrete points is related to the observable state (GPS position) through the transition function $h$.}
         \label{fig:pf_state_prev}
@@
         }
-            % The transition function $f$ models the behaviour of each particle,
-            % including changes in acceleration and stopping behaviour at bus stops,
-            % which can result in multiple modes.}
         \label{fig:pf_state_predict}
     \end{subfigure}\\
-    % \begin{subfigure}[t]{0.48\textwidth}
-    %     \centering
-    %     \includegraphics[width=\textwidth]{figures/03_particle_filter_6.pdf}
-    %     \caption{Each particle's distance from the true observation (red cross) 
-    %         is calculated using
-    %         the geographic distance between two coordinates, which can then be used
-    %         to calculate it's weight.}
-    %     \label{fig:pf_state_update}
-    % \end{subfigure}\;\;
     \begin{subfigure}[t]{0.9\textwidth}
@@
         }
-        % Resampling occurs using likelihood weights that use the distance
-        %     between the particle's location and the vehicles reported location (red cross),
-        %     resulting in the updated state.}
         \label{fig:pf_state_predict2}
@@
     }
-        % The vehicles state is estimated using a set of discrete points, or particles,
-        % each of which is independently mutated to make predictions of the future state.
-        % Multimodality is easily handled, as demonstrated here,
-        % without degeneration of the filter.}
     \label{fig:pf_state}
@@
 each with an associated weight $W_{k-1}^{(i)}$.
-The state can then be expressed using the Dirac delta function $\dot\delta(\cdot)$,
+The state is expressed using the Dirac delta function $\dot\delta(\cdot)$,
 so that
@@
 \end{equation*}
-each of which is independently updated or \emph{mutated} using the transition function $f$ (Figure~\ref{fig:pf_state_predict})
+each of which is independently updated or \emph{mutated} using the transition function $f$ (Figure~\ref{fig:pf_state_predict}),
 \begin{equation*}
 p(\bX_k | \bX_{k-1}) \approx 
-    \sum_{i=1}^N W_{k-1} \dot\delta(\bX_{k} - f(\bX_{k-1}^{(i)}, \boldsymbol{\psi}))
+    \sum_{i=1}^N W_{k-1} \dot\delta(\bX_{k} - f(\bX_{k-1}^{(i)}, \boldsymbol{\psi})),
 \end{equation*}
-using parameter vector $\boldsymbol{\psi}$ containing all of the necessary parameters
-for the model (including system noise).
+where $\boldsymbol{\psi}$ is a vector containing all of the necessary model parameters
+(Section~\ref{sec:pf_prediction}).
 After mutation, 
@@
 \end{equation*}
-generating a posterior estimate of the state
+which gives us the final posterior state estimate
 \begin{equation*}
 p(\bX_k | \bY_k) \approx  
-    \sum_{i=1}^N W_{k} \dot\delta(\bX_{k} - \bX_{k}^{(i)})
+    \sum_{i=1}^N W_{k} \dot\delta(\bX_{k} - \bX_{k}^{(i)}).
 \end{equation*}
@@
 using importance weights $\{W_k^{(1)}, \ldots, W_k^{(N)}\}$,
-which removes improbable particles and replaces them with more probable ones,
+thereby removing improbable particles and replacing them with more probable ones,
 as demonstrated in Figure~\ref{fig:pf_state_predict2}.
@@
 The particle filter allows us to flexibly model bus behaviour,
-which in the current model includes
+which currently includes
 \begin{enumerate}
@@
 by using Newton's Laws of Motion,
-which, letting $\delta_k = t_k - t_{k-1}$ and $\epsilon_k^{(i)}\sim\mathrm{N}(0, \sigma^2)$, 
-truncated to ensure $\dot x_{k-1}^{(i)} + \epsilon_k^{(i)} > 0$,
-is
+which, letting $\delta_k = t_k - t_{k-1}$, is
 \begin{equation}
 \label{eq:trans_newton}
-\bX_k^{(i)} = f_1(\bX_{k-1}^{(i)}, \delta_k, \epsilon_k^{(i)}) = 
+\bX_k^{(i)} = f(\bX_{k-1}^{(i)}, \delta_k, \epsilon_k^{(i)}) = 
     \begin{bmatrix}
@@
         \dot x_{k-1}^{(i)} + \epsilon_k^{(i)}
-    \end{bmatrix}.
+    \end{bmatrix},
 \end{equation}
+where $\epsilon_k^{(i)}\sim\mathrm{N}_T(0, \sigma^2)$, truncated to the interval
+$(-\dot x_{k-1}^{(i)}, 30 - \dot x_{k-1}^{(i)})$,
+which ensures vehicle speed is positive and less than 30~m/s
+(approximately 110~km/h).
 
@@
 The model components are implemented as an algorithm,
-in which (\ref{eq:trans_newton}) is evaluated iteratively
-one second at a time, decreasing $\delta_k$ by one each time,
-with $\epsilon_k^{(i)}$ only sampled only at the beginning
-of each time update.
-This allows the particle to approach the next stop,
+in which (\ref{eq:trans_newton}) is evaluated iteratively one second at a time,
+decreasing $\delta_k^{(i)}$ by one each time,
+after initializing $\delta_k^{(i)} = \delta_k$.
+In order to obtain the desired non-constant speed between observations,
+$\epsilon_k^{(i)}$ is re-sampled each second,
+allong the particle's speed to vary.
+This iterative approach allows the particle to approach the next stop,
 at which point (\ref{eq:trans_stop}) is evaluated for the particle
-and the dwell time subtracted from $\delta_k$,
-and continuing until $\delta_k$ reaches zero.
-Iterating this way ensures that 
-the algorithm works regardless of how many
-stops are encountered during the time step.
+and the dwell time subtracted from $\delta_k^{(i)}$.
+This process continues until $\delta_k^{(i)}$ reaches zero.
 The algorithm also keeps track of each particle's travel time $z_\ell^{(i)}$
@@
 
-% \renewcommand{\algorithmicrequire}{\textbf{Start:}}
-% \newcommand{\algorithmicbreak}{\textbf{break}}
-% \newcommand{\Break}{\State \algorithmicbreak}
-% \begin{algorithm}[t]
-%     \caption{Particle mutation function.}
-%     \label{fig:algorithm}
-%     \begin{algorithmic}[2]
-%     \Require $x_{k-1}, \dot x_{k-1}, \boldsymbol z$\\
-%     Initialisation $\delta \gets t_k - t_{k-1}$, 
-%         $j \gets stop\_index(x, \boldsymbol{S})$
-%     \If {$u \sim \mathrm{U}(0,1) < 0.5$ and $x = S_j$}
-%         \Comment{Remain longer at stop}
-%         \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
-%         \State $\delta \gets \delta - \gamma - d_j$
-%     \ElsIf {$u \sim \mathrm{U}(0,1) < 0.1$}
-%         \Comment{Occasionally bus stopped due to traffic}
-%         \State $w \gets \min(\delta, \mathrm{Exp}(1/\delta))$
-%         \State $\delta \gets \delta - w$
-%         \State $z_j \gets z_j + w$
-%         \Comment{Congestion counts as travel time}
-%     \EndIf
-%     \While {$\delta > 0$ and $x < S_J$} 
-%         \State $\dot x \sim \mathrm{N}(\dot x, \sigma^2)$
-%         \Comment{System noise, truncated to be non-negative}
-%         \State $x \leftarrow x + \dot x$
-%         \If {$x \geq S_{j+1}$}
-%             \If {$j+1 = J$}
-%                 \Comment{Break at end of route}
-%                 \State $x \gets S_J$
-%                 \Break
-%             \EndIf
-%             \State $u_j \sim \mathrm{U}(0,1)$
-%             \Comment{Simulate vehicle stopping}
-%             \If {$u_j < \pi$}
-%                 \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
-%                 \State $\delta \gets \delta - \gamma - d_j$
-%                 \State $x \gets S_j$
-%             \EndIf
-%             \State $j\gets j+1$
-%         \EndIf
-%         \State $\delta \gets \delta - 1$
-%         \State $z_j \gets z_j + 1$
-%         \Comment{Track travel time along road}
-%     \EndWhile
-%     \\
-%     \Return $x, \dot x, \boldsymbol z$
-%     \end{algorithmic}
-% \end{algorithm}
-
-% \afterpage{\clearpage}
-
 \subsubsection{Updating state using the observation likelihood}
@@
 the measurement function $h$ is needed along with 
-\emph{geographical projection} function $g$ to allow comparison of $\bY_k$,
+a \emph{geographical projection} function $g$ to allow comparison of $\bY_k$,
 a GPS observation, with $\bX_k$, a distance travelled along the route in meters.
 
-The measurement function $h$ computes GPS coordinates using the 
+The measurement function $h$ computes GPS coordinates by using the 
 shape information provided by GTFS and travelling $x_k$ meters along 
@@
 $\br = g(\bY_1 | \bY_0)$,
-so that the magnitude of $\boldsymbol{r}$ is equal to the physical distance
-between $\bY_0$ and $\bY_1$, with both dimensions in meters.
+such that the magnitude of $\boldsymbol{r}$ is equal to the physical distance
+between $\bY_0$ and $\bY_1$ and both dimensions are in meters.
 Let $\bY_i = [\lambda_i, \psi_i]^\top$,
@@
 \label{eq:dist}
-\mathcal{D}(\bY_0, \bY_1) = ||g(\bY_0 | \bY_1)|| = \sqrt{r_1^2 + r_2^2}
+\mathcal{D}(\bY_0, \bY_1) = ||g(\bY_0 | \bY_1)|| = \sqrt{r_1^2 + r_2^2}.
 \end{equation}
@@
 Now the distance between a vehicle observation $\bY_k$
-and a state $\bX_k$ using (\ref{eq:dist}) and the measurement function $h$
+and a state $\bX_k$, using (\ref{eq:dist}) and the measurement function $h$,
 is expressed as
@@
     = \sqrt{(\omega z_1)^2 + (\omega z_2)^2}
-    = \sqrt{\omega^2 (z_1^2 + z_2^2)}
+    = \sqrt{\omega^2 (z_1^2 + z_2^2)}.
 \end{equation}
@@
 $cZ \sim \mathrm{Exponential}(\frac{\theta}{c})$,
-so the \emph{squared} distance from (\ref{eq:obs_dist}) has exponential distribution
+so the \emph{squared} distance from (\ref{eq:obs_dist}) is distributed as
 \begin{equation}
@@
 \label{eq:lhood}
-p(\bY_k | \bX_k^{(i)}, \epsilon) =
+p(\bY_k | \bX_k^{(i)}, \omega) =
 \frac{1}{2\omega^2}\exp\left\{
@@
 of transit vehicles along road segment $\ell$ at time $t_c$,
-of which observations $Z_c^{\ell m}$
-are obtained from the particle filter for vehicle $m$ as defined in
-(\ref{eq:post_tt}), such that 
+of which observations $Z_{c\ell}^{m}$
+are obtained from the particle filter for vehicle $m$ at time $t_c$,
+as defined in (\ref{eq:post_tt}), such that 
 (after adding the $m$ superscript to identify unique vehicles), 
-\begin{equation*}
-Z_c^{\ell m} = \mathrm{E}(p(z_c^{\ell m} | \bY^m_{1:k})).
-\end{equation*}
+\begin{equation}
+\label{eq:tt_obs_mean}
+Z_{c\ell}^{m} = \mathrm{E}(z_{\ell}^{m} | \bY^m_{1:c}) = 
+\sum_{i=1}^N (W_c^{(i)})^m (z_\ell^{(i)})^{m}.
+\end{equation}
 The measurement error is also estimated from the particle filter,
-\begin{equation*}
-R_c^{\ell m} = \mathrm{Var}(p(z_c^{\ell m} | \bY^m_{1:k})).
-\end{equation*}
+\begin{equation}
+\label{eq:tt_obs_err}
+R_{c\ell}^m = \mathrm{Var}(z_{\ell}^{m} | \bY^m_{1:c}) = 
+\sum_{i=1}^N (W_c^{(i)})^m \left((z_\ell^{(i)})^{m} - Z_{c\ell}^{m}\right)^2.
+\end{equation}
 
@@
 with $f$ and $h$ both unity,
-system noise $v_c^\ell \sim \mathrm{N}(0, \nu_\ell^2)$,
-with $\Delta_c = t_c - t_{c-1}$
+system noise $v_{c\ell} \sim \mathrm{N}(0, \nu_\ell^2)$,
 where $\nu_\ell^2$ is the average change in travel time per second
 along road segment $\ell$,
-and the observation error $e_c^{\ell m} \sim \mathrm{N}(0, R_c^{\ell m})$:
+the observation error $e_{c\ell}^{m} \sim \mathrm{N}(0, R_{c\ell}^{m})$,
+and $\Delta_c = t_c - t_{c-1}$
 \begin{equation*}
 \begin{split}
-\theta_c^\ell &= \theta_{c-1}^\ell + \Delta_c^\ell v_c^\ell \\
-Z_c^{\ell m} &= \theta_c^\ell + e_c^{\ell m}
+\theta_c^\ell &= \theta_{c-1}^\ell + \Delta_{c\ell} v_{c\ell} \\
+Z_{c\ell}^{m} &= \theta_c^\ell + e_{c\ell}^{m}
 \end{split}
@@
 The state at time $t_{c-1}$ is parameterised by its mean and variance,
-conditional on all of the data from all vehicles up until time $t_{c-1}$,
-$\hat \theta_{c-1|c-1}^\ell = \mathrm{E}(\theta_c^\ell | Z_{1:c-1}^{\ell\boldsymbol{\cdot}})$
-and \mbox{$P_{c-1|c-1}^\ell = \mathrm{var}(\theta_{c-1}^\ell | Z_{1:c-1}^{\ell\boldsymbol{\cdot}})$}, respectively,
+conditional on $\boldsymbol{Z}_{1:c-1,\ell}$, 
+the data from all vehicles up until time $t_{c-1}$,
+\begin{equation*}
+\hat \theta_{c-1|c-1}^\ell = 
+\mathrm{E}(\theta_{c-1}^\ell | \boldsymbol{Z}_{1:c-1,\ell})
+\quad\text{and}\quad
+P_{c-1|c-1}^\ell = 
+\mathrm{Var}(\theta_{c-1}^\ell | \boldsymbol{Z}_{1:c-1,\ell})
+\end{equation*}
+respectively,
 which are predicted from the previous state estimate using the predictive model
@@
 \hat \theta^\ell_{c|c-1} &= \hat \theta^\ell_{c-1|c-1} \\
-P^\ell_{c|c-1} &= P^\ell_{c-1|c-1} + (\Delta_c^\ell \nu_c^\ell)^2
+P^\ell_{c|c-1} &= P^\ell_{c-1|c-1} + (\Delta_{c\ell} \nu_{c\ell})^2
 \end{align*}
@@
 For the update step, the parameters are transformed into an information
-space parameterised by the information matrix $U^\ell_c = (P_{c|c-1})^{-1}$
+space parameterised by the information matrix $U^\ell_c = (P_{c|c-1}^\ell)^{-1}$
 and the information vector $u^\ell_c~=~\hat \theta^\ell_{c|c-1} (P^\ell_{c|c-1})^{-1}$.
 The travel time estimate of vehicle $m$ along segment $\ell$,
-$Z_c^{\ell m}$, along with its uncertainty $R^{\ell m}_c$,
+along with its uncertainty, 
+as defined in (\ref{eq:tt_obs_mean}) and (\ref{eq:tt_obs_err}), respectively,
 are transformed to a measurement information covariance matrix 
-$B_c^{\ell m}~=~(R^{\ell m}_c)^{-2}$
-and measurement information vector $b_c^{\ell m}~=~Z^{\ell m}_c (R^{\ell m}_c)^{-2}$.
-The total information is the sum of the information over all $M_c^\ell$ vehicles
-that traversed segment $\ell$ since the last update,
+$B_{c\ell}^{m}~=~(R_{c\ell}^m)^{-2}$
+and measurement information vector $b_{c\ell}^{m}~=~Z_{c\ell}^{m} (R_{c\ell}^{m})^{-2}$.
+The total information is the sum of the information over all $M_{c\ell}$ vehicles
+that traversed segment $\ell$ in the time interval $(t_{c-1}, t_c]$,
 so the update equation is
 \begin{align*}
-U^\ell_{c|c} &= U^\ell_{c|c-1} + \sum_{m=1}^{M_c^\ell} B^{m\ell}_{c} \\
-\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^{M_c^\ell} b^{m\ell}_{c}.
+U^\ell_{c|c} &= U^\ell_{c|c-1} + \sum_{m=1}^{M_{c\ell}} B_{c\ell}^{m} \\
+\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^{M_{c\ell}} b_{c\ell}^{m}.
 \end{align*}

### 2019-08-09 anzjs2018/sections/04_implementation.tex
More edits

@@
 with complexity $O(N \log N)$,
-which explains non-linear relationship with $N$.
+which explains the non-linear relationship with $N$.
 
@@
         Total time is for the entire iteration, which includes fetching the data from the API.
-        % Total time increases as the number of particles $N$ increases.
-        % The reason for the large contribution of writing ETAs is that
-        % this step involves estimation of quantiles,
-        % which requires sorting the particles, 
-        % The vehicle update step also involves sorting whenever the sample
-        % degenerations and resampling is required.
-        % Total time also includes fetching the data from the API,
-        % which is constant for any $N$, which explains the increase
-        % in standardised total time for $N=500$.
     }
@@
 degeneration rate, the percentage of iterations in which the vehicle was lost
-(no plausible particles);
+(no plausible particles remained);
 and the (relative) variance of segment travel time estimates.
@@
     \centering
-    \includegraphics[width=0.8\textwidth]{figures/04_model_results_dist.pdf}
+    \includegraphics[width=0.7\textwidth]{figures/04_model_results_dist.pdf}
     \caption{
@@
     \label{fig:dist_to_route}
+\end{figure}
+
+\begin{figure}[tb]
+    \centering
+    \includegraphics[width=\textwidth]{figures/04_model_results_neff.pdf}
+    \caption{
+        The effective sample size, $N_\text{eff}$,
+        and degeneration rate for varying values of GPS error,
+        system noise, and number of particles $N$.
+    }
+    \label{fig:perf_stats}
 \end{figure}
@@
 
-\begin{figure}[tb]
-    \centering
-    \includegraphics[width=\textwidth]{figures/04_model_results_neff.pdf}
-    \caption{
-        The effective sample size, $N_\text{eff}$,
-        and degeneration rate for varying values of GPS error,
-        system noise, and number of particles $N$.
-        % increases with GPS error and decreases with system noise,
-        % however both effects are reduced for larger $N$.
-        % Conversely, degeneration rate decreases with GPS error and
-        % system noise, although the effect of the latter is very small,
-        % and again the relationship is less pronounced for larger $N$.
-    }
-    \label{fig:perf_stats}
-\end{figure}
-
-% \begin{figure}[tb]
-%     \centering
-%     \includegraphics[width=0.9\textwidth]{figures/04_model_results_degen.pdf}
-%     \caption{
-%         The degeneration rate is the percentage of iterations where there were no
-%         plausible particles, so the vehicle was ``lost'' by the model.
-%         This decreases as GPS error and number of particles increase,
-%         but there is no visible effect of system noise.
-%         Note, however, that some of these can be attributed to invalid data,
-%         for example if the driver has entered the wrong trip direction.
-%         The error bars show $\pm 2$SE.
-%     }
-%     \label{fig:degen_rate}
-% \end{figure}
-
 The last measure of performance is the relative variance of segment travel times.
@@
 each individual simulation and variance over all simulations.
-If $\tilde z_\ell^{\omega,\epsilon}$ is a vector of all travel times
-along segment $\ell$ during the simulation with GPS error $\omega$ and system noise $\epsilon$,
-then the relative variance $\bar v^{\omega,\epsilon}$ is calculated by
+If $\tilde \bz_\ell^{\omega,\sigma}$ is a vector of all travel times
+along segment $\ell$ during the simulation with GPS error $\omega$ and system noise $\sigma$,
+then the relative variance $\bar v^{\omega,\sigma}$ is calculated by
 \begin{equation*}
 \begin{split}
-v_\ell^{\omega,\epsilon} &=
-\frac{\mathrm{Var}(\tilde z_\ell^{\omega,\epsilon})}{\mathrm{Var}(\tilde z_\ell)} \\
-\bar v^{\omega,\epsilon} &=
-    \frac{1}{L}\sum_{\ell=1}^L v_\ell^{\omega,\epsilon}
+v_\ell^{\omega,\sigma} &=
+\frac{\mathrm{Var}(\tilde \bz_\ell^{\omega,\sigma})}{\mathrm{Var}(\tilde \bz_\ell)} \\
+\bar v^{\omega,\sigma} &=
+    \frac{1}{L}\sum_{\ell=1}^L v_\ell^{\omega,\sigma}
 \end{split}
@@
 while simulations with more particles had lower relative variance.
-There was no visible effect of system noise on the variability of travel times.
+There was no clear pattern to the effect of system noise.
 
@@
 is going to have a significant effect on the final ETAs,
-particularly when compared to the uncertainty of forecasting road state,
-or whether reducing the degeneration rate, 
-or increasing $N_\text{eff}$, would be more advantageous.
+particularly when compared to the uncertainty of forecasting road state.
 

### 2019-08-09 anzjs2018/sections/05_discuss.tex
More edits

@@
 for example incorporating peak time congestion.
-% This in turn will provide the basis for an arrival time prediction model
-% that combines \rt vehicle state with travel time forecasts
-% in order to provide commuters with more reliable ETAs.
 As for the arrival times, 
@@
 While the timings presented were for off-peak times,
-we hope futher investigation of choosing $N$ and $N_\text{eff}$,
+futher investigation into the choice of $N$ and $N_\text{eff}$,
 as well as work removing some redudancy from the transition algorithm,
-will mean times during peak periods, when there are 2-3~times
-as many vehicles, will come in under our 30~second target.
+should allow us to model all vehicles during peak periods---%
+when there are 2-3~times as many vehicles---%
+within our 30~second timeframe target.

### 2019-08-09 anzjs2018/figures/02_gtfs_delays.R

@@ -6,7 +6,7 @@ s <- paste0("2019-01-01 12:",
         ":00") %>%
     as.POSIXct
 
-delay <- c(30, 60, 90)
+delay <- c(30, 60, 120)
 a <- s[1:3] + delay
 
 pdf('figures/02_gtfs_delays.pdf', width = 7, height = 2.5, onefile = FALSE)
@@ -30,26 +30,29 @@ egg::ggarrange(
         geom_point(aes(s[4] + delay, d[4]),
             shape = 21, fill = "white"),
     ggplot() +
+        geom_path(
+            aes(
+                c(s[2], a[2], a[2], a[3], a[3], a[3] + 60),
+                c(4.5, 3.5, 4, 1, 2, 1)
+            ),
+            lty = 2
+        ) +
         geom_point(
             aes(
                 c(s[2], a[-1]),
                 c(4.5, 4, 2)
-            )
-        ) +
-        geom_path(
-            aes(
-                c(s[2], a[2], a[2], a[3], a[3]),
-                c(4.5, 3.5, 4, 1.5, 2)
-                # as.integer(c(
-                #     s[4] - a[1], 
-                #     s[4] - a[1] - (a[2] - s[1]),
-                #     s[4] - a[2],
-                #     s[4] - a[2] - (a[3] - s[2]),
-                #     s[4] - a[3]
-                # ) / 60)
             ),
-            lty = 2
+            shape = 21, fill = "white"
         ) +
+        geom_text(
+            aes(
+                c(s[2], a[-1]), 
+                c(4.5, 4, 2),
+                label = paste0("ETA[", 1:3, "]")
+            ),
+            parse = TRUE,
+            nudge_y = 0.3, nudge_x = 10) +
+        ylim(0, 5) +
         xlab("Time") +
         ylab("ETA (minutes)") +
         theme_classic(),

### 2019-08-09 anzjs2018/sections/01_introduction.tex
more edits

@@
 \citep{Cathey_2003,Jeong_2005,Yu_2011,Hans_2015},
-many public transport providers---notably Auckland Transport---no formal model is used.
+many public transport providers---notably Auckland Transport---use no formal model.
 Instead, ETAs are based solely on the scheduled arrival time
@@
 \citep{Gustafsson_2002,Davidson_2011}.
-\cite{Ulmke_2006} compared used both a Gaussian sum approximation and a particle filter
+\cite{Ulmke_2006} compared both a Gaussian sum approximation and a particle filter
 to track a vehicle in real-time,
@@
 which is a situation that often occurs with transit data---%
-the bus may be stopped a a bus stop, or it could be continuing along the route but,
+the bus may be stopped at a bus stop, or it could be continuing along the route but,
 for some other reason, failing to report its position.
@@
 
-In this paper, we propose a simple framework for modelling the predicting
-transit vehicle arrival in real-time.
+In this paper, we propose a simple framework for modelling transit vehicles and predicting
+their arrival times in real-time.
 Section~\ref{sec:gtfs} describes how we construct a transit network,
@@
 with the primary goal of estimating \rt traffic states throughout the network.
-Finally, in Section~\ref{sec:rt}, we discuss the \rt faesibility and performance
-of using a particle filter to model transit vehicles in real-time.
+Finally, we discuss the \rt faesibility and performance
+of the particle filter in Section~\ref{sec:rt}.

### 2019-08-09 anzjs2018/sections/03_models.tex
more edits

@@
 p(\bX_{k-1} | \bY_{1:k-1}) \approx 
-    \sum_{i=1}^N W_{k-1} \dot\delta(\bX_{k-1} - \bX_{k-1}^{(i)})
+    \sum_{i=1}^N W_{k-1}^{(i)} \dot\delta(\bX_{k-1} - \bX_{k-1}^{(i)})
 \end{equation*}
@@
 p(\bX_k | \bX_{k-1}) \approx 
-    \sum_{i=1}^N W_{k-1} \dot\delta(\bX_{k} - f(\bX_{k-1}^{(i)}, \boldsymbol{\psi})),
+    \sum_{i=1}^N W_{k-1}^{(i)} \dot\delta(\bX_{k} - f(\bX_{k-1}^{(i)}, \boldsymbol{\psi})),
 \end{equation*}
@@
 p(\bX_k | \bY_k) \approx  
-    \sum_{i=1}^N W_{k} \dot\delta(\bX_{k} - \bX_{k}^{(i)}).
+    \sum_{i=1}^N W_{k}^{(i)} \dot\delta(\bX_{k} - \bX_{k}^{(i)}).
 \end{equation*}
@@
 the effective sample size $N_{\text{eff}} = 1 / \sum_i (W_k^{(i)})^2$
-falls below a specified threshold $N_{\text{thres}}$.
-We used a fixed threshold value of $N_{\text{thres}} = N/10$
+falls below a specified threshold $N_{\text{thres}}$
+\citep{Gustafsson_2002}.
+We used a fixed threshold value of $N_{\text{thres}} = N/4$
 for the current work.
@@
 one for each of the above behaviours.
-The first models the speed and distance travelled of the vehicle 
+The first models the speed and distance travelled by the vehicle 
 by using Newton's Laws of Motion,
@@
 $(-\dot x_{k-1}^{(i)}, 30 - \dot x_{k-1}^{(i)})$,
-which ensures vehicle speed is positive and less than 30~m/s
+which ensures vehicle speed is always positive and less than 30~m/s
 (approximately 110~km/h).
@@
 $\epsilon_k^{(i)}$ is re-sampled each second,
-allong the particle's speed to vary.
+allowing the particle's speed to vary.
 This iterative approach allows the particle to approach the next stop,

### 2019-08-09 anzjs2018/sections/05_discuss.tex
more edits

@@
 or new ones added, such as implementing intersection-specific behaviour
-if intersecion locations are available.
+if intersection locations are available.
 Future work is planned to detect route meeting points

### 2019-08-10 anzjs2018/anzjs.tex
remove date from filename
Grammar and sentence/paragraph restructuring

--- 2018-11-01 anzjs2018/anzjs2018.tex
@@
 
-\input{../glossary.tex}
-
 \newcommand{\bv}{\boldsymbol{v}}
@@
 \def\volumeyear{2018}
-\linenumbers
+% \linenumbers
 
@@
 knowledge of its current position and schedule adherence,
-but also of traffic conditions along the remainder of the route.
-Road networks are dynamic, 
-and can change from free-flowing to highly congested very quickly,
+but also traffic conditions along the remainder of the route.
+Road networks are dynamic 
+and can quickly change from free-flowing to highly congested,
 which impacts the arrival time of transit vehicles,
@@
 which are used in a second model to estimate \rt traffic conditions.
-Development and testing took place in Auckland, New Zealand,
-although attempts were made at generalising each component so 
-it will be compatible with other public transport systems around the world.
+Although development and testing took place in Auckland, New Zealand,
+we generalised each component to make the framework compatible 
+with other public transport systems around the world.
 We demonstrate the \rt feasibility and performance of our approach in real-time,
@@
 \input{sections/04_implementation}
-% \input{sections/05_results}
-\input{sections/06_discuss}
+\input{sections/05_discuss}
 \bibliographystyle{anzsj}
-\bibliography{../reflist.bib}
+\bibliography{reflist.bib}
 

### 2019-08-10 anzjs2018/sections/01_introduction.tex
Grammar and sentence/paragraph restructuring

@@
 when arrival time information is available \citep{TCRP_2003b},
-but here in Auckland, RTI is highly unreliable,
-leading to frustration and ultimately detering public transport users.
-Inaccurate ETAs are a major source of this frustration,
+but in Auckland, RTI is highly unreliable,
+leading to frustration and ultimately deterring public transport users.
+Inaccurate ETAs are the primary source of this frustration,
 as they fluctuate---sometimes dramatically---over time.
-While this may be caused by unpredictable changes to traffic conditions,
-it can equally simply be a result of poor schedule calibration.
+While unpredictable changes to traffic conditions may cause this,
+it can equally be due to poor schedule calibration.
 Furthermore, buses are shown as \emph{on-time} 
-if they are not present in the \rt system and have not been manually cancelled,
+if they are not present in the \rt system and 
+the service provider has not manually cancelled them,
 in which case ETAs are based solely on scheduled arrival times.
@@
 Arrival time prediction is only as reliable as the underlying model,
-and while a lot of work has gone into developing public transport vehicle models
+and while considerable work has gone into developing public transport vehicle models
 \citep{Cathey_2003,Jeong_2005,Yu_2011,Hans_2015},
@@
 \emph{if available}.
-The primary assumption here is that the schedule is valid,
-and secondly that there is no unusual congestion along the route;
-neither of these are valid,
+This method assumes that the schedule is accurate
+and there is no unusual congestion along the route.
+Neither of these is a valid assumption,
 particularly in our test area of Auckland, New Zealand,
 where infrastructure for buses (such as priority lanes) is limited,
-and bus drivers, in general, 
-do not actively adhere to the schedule.
+and bus drivers do not actively adhere to the schedule.
 A more robust modelling and prediction framework 
@@
 of transit vehicles from \rt GPS data,
-and secondly a means of combining speed information from vehicles
-to model traffic flows,
-which can then be used to improve arrival time predictions.
+and a means of combining speed information from vehicles
+to model traffic flow.
+This information can then be used to improve arrival time predictions.
 
@@
 In particular, the particle filter accurately estimated the \emph{uncertainty}
-of the vehicle's location when the observations were unavailable for a time,
-which is a situation that often occurs with transit data---%
+of the vehicle's location when the observations were unavailable for a time.
+This situation often occurs with transit data---%
 the bus may be stopped at a bus stop, or it could be continuing along the route but,
@@
 as a predictor of travel time.
-Since these approaches only work well on high frequency routes,
+Since these approaches only work well on high-frequency routes,
 \cite{Yu_2011} showed further improvements by combining travel times 
 from several routes;
-however, this was limited to predefined converging routes.
+however, their method was limited to predefined converging routes.
 In general, however, no comprehensive network modelling approach has been proposed using
-solely GPS position data to model and account for congestion when estimating arrival times.
+only GPS position data to model and account for congestion when estimating arrival times.
 
@@
 with the primary goal of estimating \rt traffic states throughout the network.
-Finally, we discuss the \rt faesibility and performance
+Finally, we discuss the \rt feasibility and performance
 of the particle filter in Section~\ref{sec:rt}.

### 2019-08-10 anzjs2018/sections/02_gtfs.tex
Grammar and sentence/paragraph restructuring

@@
 making access easier for application developers.
-Developed and maintained by Google \citep{GoogleDevelopers_2006},
+GTFS is developed and maintained by Google \citep{GoogleDevelopers_2006},
 who use it in Google Maps Transit Directions,
-it is used by over 1100~transit providers around the world\footnote{%
+and over 1100~transit providers use it around the world\footnote{%
 source: \url{http://transitfeeds.com} as of August 2019},
@@
 provided an application depends solely on GTFS data,
-after developing it locally in Auckland it can be deployed to any other GTFS-based
+after developing it locally in Auckland it will be deployable to any other GTFS-based
 public transport system with minimal modification.
@@
 \end{itemize}
-The second component is \emph{GTFS-realtime},
-which is only available in a subset of the providers due to the requirement of 
-on board GPS tracking devices and a central server.
+The second component, \emph{GTFS-realtime},
+is only available in a subset of providers due to the requirement of 
+onboard GPS tracking devices and a central server.
 It provides a standardised format for sharing vehicle positions and trip delays,
-and are typically accessed by developers via an API that can be used in \rt applications.
+and are typically accessed by developers via an API that \rt applications can use.
 
 As mentioned in Section~\ref{sec:intro},
-there are some major issues with the arrival time prediction method currently
+there are some fundamental issues with the arrival time prediction method currently
 deployed in Auckland,
@@
 vehicle positions are used only to display \emph{where} the bus is.
-Trip updates are reported whenever a vehicle arrives at or departs from a stop,
-and includes the delay between the scheduled and actual arrival times,
-which is propagated to all future stops to update their ETAs,
+Vehicles report trip updates when they arrive at or depart from a stop,
+including the delay between the scheduled and actual arrival times.
+The delay is then propagated to all future stops to update their ETAs,
 as demonstrated in Figure~\ref{fig:gtfs-delays}.
-As already mentioned, this assumes that the schedule is well calibrated,
+As already mentioned, this assumes that the schedule is well-calibrated
 and that the time between stops
@@
 Figure~\ref{fig:gtfs-delays} demonstrates the fallibility of this prediction method,
-where we see the progressive lateness causing the passenger's ETA to 
+where we see the incremental lateness causing the passenger's ETA to 
 jump each time the bus arrives at a stop.
@@
 the travel time along roads between where the bus is \emph{now},
-and the stop at which a passenger is waiting.
+and the stop where a passenger waits.
 In most applications, however, this vital information is unavailable, at least directly.
@@
 such that every bus travelling along a road contributes to its state,
-which can in turn be used to predict travel times for all upcoming buses 
+which can, in turn, be used to predict travel times for all upcoming buses 
 travelling along the road,
@@
 
-To construct the transit network,
-routes were split into spatially identifiable segments,
+Constructing the transit network involved splitting routes
+into spatially identifiable segments,
 each representative of a physical road,
@@
 such as road segment overlaps where routes merge between stops,
-our approach is sufficient for the purposes of assessing the \rt 
-faesibilty of our proposed prediction framework.
+our approach is sufficient for assessing the \rt 
+feasibility of our proposed prediction framework.
 
@@
 the GPS position of the vehicle, $\bY_k$, 
-and information about the trip being serviced.
-In Auckland, vehicle positions are updated with a frequency of anywhere between 10~seconds and several minutes,
-so there is often a lot of uncertainty about a vehicle's intermediate trajectory,
-particularly when there are one or more bus stops along the way.
+and information about the serviced trip.
+In Auckland, vehicle positions are updated with a frequency 
+of anywhere between 10~seconds and several minutes,
+so there is often considerable uncertainty about a vehicle's intermediate trajectory,
+in particular when there are one or more bus stops along the way.
 It is also possible for a bus to remain stationary,
-for example due to heavy congestion,
+such as when there is heavy congestion,
 so the number of possible trajectories rapidly increases with 
@@
 
-One important consideration regarding Auckland Transport's \rt implementation is that
+An important consideration regarding Auckland Transport's \rt implementation is that
 buses are programmed to report their location when arriving at or departing from
@@
 these positions can be preemptive,
-for example when approaching a queue of traffic at an intersection,
-the way-point may be triggered before the bus physically gets there,
+for example when approaching a queue of traffic at an intersection;
+the way-point may trigger before the bus physically gets to it,
 so consecutive observations may show what appears to be a bus travelling backwards.
@@
 of the vehicle by finding the nearest point on the path to the observation;
-if this has decreased, the current state (based on the preemptive observation)
-is rejected and the vehicle reverted
+if this has decreased, we reject the current state (based on the preemptive observation)
+and reset the vehicle 
 to its previous state before continuing.
 So at the cost of losing one (likely invalid) observation,
-we are able to improve particle filter performance
-when this situation is encountered.
+we can improve predictive performance
+when this event occurs.

### 2019-08-10 anzjs2018/sections/03_models.tex
Grammar and sentence/paragraph restructuring

@@
 
-The following sections describe the two models used in this application,
-both of which are examples of a Bayesian filter.
+The following sections describe two Bayesian filters used in this application.
 The first, implemented using a particle filter, estimates vehicle states
-with the primary goal of estimating road travel times from a sequence of GPS positions;
-the second uses these travel times to update road states 
+with the primary goal of estimating road travel times from a sequence of GPS positions.
+The second uses these travel times to update road states 
 and is implemented using an \emph{information filter},
@@
 as $\bY_k = \left[\lambda_k\ \phi_k \right]^\top$.
-We use a particle filter to estimate $\bX_k$,
+We use a particle filter to estimate $\bX_k$
 due to its flexibility and proven robustness
 in recent vehicle modelling applications \citep{Ulmke_2006,Hans_2015}.
-In addition, it makes it trivial to estimate travel times
+It also makes it trivial to estimate travel times
 $\bz = (z_1,\ldots,z_L)^\top$, in seconds, along road segments
@@
 which is a common feature of the proposal distribution, particularly around bus stops.
-Another advantage is that the likelihood function is intuitively based
+Another advantage is that the likelihood function is intuitive, based
 on the physical distance between the vehicle observation and the state's
-estimate of the true location, $h(\bX_k)$ (Section~\ref{sec:pf_update}).
+estimate of the actual location, $h(\bX_k)$ (Section~\ref{sec:pf_update}).
 Conversely, particle filter methods are computationally demanding,
 requiring an increasing number of particles as model complexity and
-number of parameters increases \citep{Carpenter_1999}.
+the number of parameters increases \citep{Carpenter_1999}.
 Section~\ref{sec:rt} describes our implementation of the particle filter
@@
         \caption{
-            The state is updated by weighted resampling based on distance
-            of particle to observation.
+            The state is updated by resampling with likelihood weights based on the distance
+            between the particle and the observation, shown as a (red) cross.
         }
@@
         to predict the next state.
-        The measurement function $h$ allows each particle state
-        to be mapped to a GPS coordinate (a),
+        The measurement function $h$ maps each particle state
+        to a GPS coordinate (a),
         which can be compared to the observed location (c)
@@
 
-One problem with particle filters is degeneration,
-which occurs when the particle sample no longer approximates the target distribution
-due to the particles becoming too diverse
-and all of the weight falls onto only a small number of them.
-This occurs naturally over time, but the rate at which it occurs depends 
-on a relationship between how well the transition function predicts future states,
+One problem with particle filters is degeneration:
+when the particles become too spread out with respect to the predictive distribution,
+the weight will be concentrated on to just a few of them.
+This occurs naturally over time, but the rate at which it does so depends 
+on a relationship between how well the transition function predicts future states
 and the variance of those predictions (system noise).
@@
 one for each of the above behaviours.
-The first models the speed and distance travelled by the vehicle 
-by using Newton's Laws of Motion,
-which, letting $\delta_k = t_k - t_{k-1}$, is
+The first models the speed and distance travelled by the vehicle,
 \begin{equation}
@@
 \end{equation}
-where $\epsilon_k^{(i)}\sim\mathrm{N}_T(0, \sigma^2)$, truncated to the interval
-$(-\dot x_{k-1}^{(i)}, 30 - \dot x_{k-1}^{(i)})$,
-which ensures vehicle speed is always positive and less than 30~m/s
+where $\delta_k = t_k - t_{k-1}$
+and $\epsilon_k^{(i)}\sim\mathrm{N}_T(0, \sigma^2)$, truncated to the interval
+$(-\dot x_{k-1}^{(i)}, 30 - \dot x_{k-1}^{(i)})$
+to ensure vehicle speed is always positive and less than 30~m/s
 (approximately 110~km/h).
@@
 which is assumed to be exponentially distributed with mean $\tau_j$.
-This leads to the following model,
+This behaviour is expressed in the following model:
 \begin{equation}
@@
 
-To compute the likelihood of the observation given each particle state,
+Computing the likelihood of the observation given each particle state,
 $p(\bY_k|\bX_k^{(i)})$,
-the measurement function $h$ is needed along with 
+uses the measurement function $h$ and 
 a \emph{geographical projection} function $g$ to allow comparison of $\bY_k$,
@@
 
-Finally, it is known that the distribution of the sum of two indendent 
-standard normal variables is exponentially distributed with rate $\frac{1}{2}$,
-and that if $Z \sim \mathrm{Exponential}(\theta)$ then
-$cZ \sim \mathrm{Exponential}(\frac{\theta}{c})$,
-so the \emph{squared} distance from (\ref{eq:obs_dist}) is distributed as
+Finally, the sum of two independent 
+standard normal random variables has a Chi-square distribution with two degrees of freedom,
+which is also an exponential distribution with rate~$\frac{1}{2}$,
+and if $Z \sim \mathrm{Exponential}(\theta)$ then
+$cZ \sim \mathrm{Exponential}(\frac{\theta}{c})$;
+therefore, the \emph{squared} distance from (\ref{eq:obs_dist}) is distributed as
 \begin{equation}
@@
 possible due to the discrete nature of the particle filter state estimate;
-in the Kalman filter, which has often been used in transit vehicle tracking,
-the measurement \emph{matrix} is used, and requires a linear
+the Kalman filter---which has often been used for transit vehicle tracking---%
+uses the measurement \emph{matrix}, requiring a linear
 transformation between the state and its observations.
-To enable this, applications first estimate the \emph{observed distance travelled}
+To do so, applications first estimate the \emph{observed distance travelled}
 by snapping the GPS observations to the route,
@@
 The primary objective of the network model is to estimate the \rt traffic conditions
-(travel time) along roads in the transit network, 
+(travel time) along roads in the transit network
 and make short-term forecasts for estimating arrival times.
 In this paper, we model each road segment independently,
-as excluding correlations simplifies the model to not only a one-dimensional Kalman filter,
+as excluding correlations not only simplifies the model to a one-dimensional Kalman filter,
 but it also means computations can be run in parallel,
-greatly increasing the real-time performance of the model.
+further improving the real-time performance of the model.
 
@@
 The model of travel times becomes a simple reduction of (\ref{eq:rbe_model}) 
-to the one dimensional case, 
+to the one-dimensional case, 
 with $f$ and $h$ both unity,
@@
 for upcoming buses. 
-It should be noted that, if segment correlations were to be included,
+Note that, if we included segment correlations,
 the full state $\boldsymbol{\theta}_c$ would need updating in a single step,

### 2019-08-10 anzjs2018/sections/04_implementation.tex
more edits
Grammar and sentence/paragraph restructuring

@@
 and assess its performance 
-both with respect to timing and travel time estimation.
+with respect to timing and travel time estimation.
 
-The general structure of the \rt component is shown below,
-with the bold steps being those discussed in this paper.
+Below is the general structure of the \rt component, 
+where the bold steps are those discussed in this paper.
 \begin{enumerate}
@@
 During the development of the application,
-we were primarily concerned with ensuring each component of the program
+our primary concern was to ensure each component of the program
 was as efficient as possible,
-allowing ETAs to be generated and distributed fast enough to be feasible in real-time,
+so ETA generation and distribution is fast enough to be feasible in real-time,
 with a target of 30~seconds or faster at peak time.
 Using C++ provided the memory management control necessary to make this possible.
-We also made use of OpenMP for parallelisation,
-since each vehicle is modelled independently so it is trivial to scale up without
-thread safety concerns.
+Since we modelled each vehicle independently, 
+it was straightforward to spread the vehicle updates 
+over multiple cores using OpenMP without thread-safety concerns.
 
@@
 so it was necessary to explore the performance of the application
-with varying number of particles.
-Application performance is also assessed
-for a range of fixed model parameter values,
-such as system noise and GPS error.
+with a varying number of particles.
+We assessed the application's performance
+using a range of values for
+system noise and GPS error.
 To enable comparisons, we implemented a simulated \rt environment
-in which the same subset of real vehicle data from 8~October, 2018
+in which the same subset of real vehicle data from 8~October 2018
 could be analysed using a range of settings.
-These simulations were carried out of a Virtual Machine 
-with 8~cores and 32~Gb of memory, 
+These simulations were carried out on a virtual machine 
+with 8~Intel Xeon 3.00GHz CPU cores and 32~GB of memory, 
 running Ubuntu~16.04 and using R~3.4.1.
@@
 
-In each iteration, 
-the timings of the various program components were recorded.
+Within each iteration, the program recorded timings of its various components.
 Since the number of vehicles travelling at any given time changes throughout the day,
-we used the average timings over an off-peak 15~minute window.
+we used the average timings over an off-peak 15-minute window.
 Figure~\ref{fig:timings} shows the average timings for 
@@
 
-The most time consuming component was ETA writing,
+The most time-consuming component was ETA writing,
 which involved summarising the individual ETAs estimated for each particle.
@@
 The ETA prediction step only took a few seconds,
-although this may increase once a more comprehensive model is developed.
+although this may increase once a more comprehensive model has been developed.
 On our 8-core virtual machine, 
-the average time to process one iteration during off-peak is about 16~seconds
-when 8000~particles are used.
-% Figure~\ref{fig:day-timings} shows iteration timings over the course of a day.
+the average time to process one iteration during off-peak 
+was about 16 seconds when using 8000 particles.
 
@@
         The average real-time running time of each step within each iteration
-        over a 15~minute off-peak window.
+        over a 15-minute off-peak window.
         The vehicle update and writing ETA steps involve sorting the particles,
@@
 
-To evaluate the performance of the model,
-the simulation was run using a range of values of GPS error, $\omega^2$,
+Evaluating the performance of the model involved
+running the simulation using a range of values of GPS error, $\omega^2$,
 and system noise, $\sigma^2$.
@@
 which suggested using values $\omega \in \{1,2,3,5\}$ for GPS error in meters.
-For the system noise we used values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$,
+For the system noise, we used values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$,
 which correspond to an average vehicle speed variation of between 0.1 and 45~meters per second
-over a period of 30~seconds.
+over 30~seconds.
 For each combination of these parameter values, including varying $N$,
@@
     \caption{
-        The distribution of minimum distance between observed vehicle location
+        The distribution of the minimum distance between observed vehicle location
         and the route path, truncated to 20~meters.
@@
         and degeneration rate for varying values of GPS error,
-        system noise, and number of particles $N$.
+        system noise, and the number of particles $N$.
     }
@@
 Figure~\ref{fig:perf_stats} shows that $N_\text{eff}$ increases
-with GPS error and decreases with system noise,
+with GPS error and decreases with system noise
 while remaining unaffected by changing $N$.
@@
 so we see a slight decrease in degeneration rate.
-Thus, these results are not surprising, 
-but show that the model is working as expected.
+Thus, these results are not surprising
+and show that the model is working as expected.
 
@@
 taking the average (over road segments) of the ratio between variance for
-each individual simulation and variance over all simulations.
+each simulation and variance over all simulations.
 If $\tilde \bz_\ell^{\omega,\sigma}$ is a vector of all travel times
@@
     \frac{1}{L}\sum_{\ell=1}^L v_\ell^{\omega,\sigma}
-\end{split}
+\end{split}.
 \end{equation*}
@@
 resulted in a higher relative variance,
-while simulations with more particles had lower relative variance.
+while simulations with more particles had a smaller relative variance.
 There was no clear pattern to the effect of system noise.
@@
         The relative variability of travel time estimates for varying
-        values of GPS error, system noise, and number of particles, $N$,
+        values of GPS error, system noise, and the number of particles, $N$,
         with error bars $\pm 2$ standard errors.

### 2019-08-10 anzjs2018/sections/05_discuss.tex
Grammar and sentence/paragraph restructuring

@@
 
-In this paper we described an approach to transit vehicle modelling
+In this paper, we described an approach to transit vehicle modelling
 that enables the \rt estimation of road state from vehicle position data,
@@
 that account for \rt traffic conditions.
-The vehicle model, implemented using a particle filter,
-allows complex bus behaviours to be modelled,
-from which 
-vehicle travel times along roads are estimated which are in turn used 
-in the modelling of the road network state.
+The particle filter allows the modelling of complex bus behaviours,
+providing estimates of vehicle travel times along roads
+which are in turn used to estimate the road network state.
 
@@
 Future work is planned to detect route meeting points
-and use these as intersection nodes,
-with the goal of further reducing road segment overlap.
+and use these as intersection nodes
+ to further reduce road segment overlap.
 
@@
 The results of our simulations show that the framework is a 
-faesible \rt alternative.
+feasible \rt alternative.
 While the timings presented were for off-peak times,
-futher investigation into the choice of $N$ and $N_\text{eff}$,
-as well as work removing some redudancy from the transition algorithm,
+further investigation into the choice of $N$ and $N_\text{eff}$,
+as well as work removing redundancy from the transition algorithm,
 should allow us to model all vehicles during peak periods---%
 when there are 2-3~times as many vehicles---%
-within our 30~second timeframe target.
+within our 30-second timeframe target.

### 2019-08-10 anzjs2018/sections/02_gtfs.tex
More spelling mistakes

@@
 is only available in a subset of providers due to the requirement of 
-onboard GPS tracking devices and a central server.
+on-board GPS tracking devices and a central server.
 It provides a standardised format for sharing vehicle positions and trip delays,
@@
 To further complicate this,
-these positions can be preemptive,
+these positions can be pre-emptive,
 for example when approaching a queue of traffic at an intersection;
@@
 of the vehicle by finding the nearest point on the path to the observation;
-if this has decreased, we reject the current state (based on the preemptive observation)
+if this has decreased, we reject the current state (based on the pre-emptive observation)
 and reset the vehicle 

### 2019-08-10 anzjs2018/sections/04_implementation.tex
More spelling mistakes

@@
 The results of the simulations
-demonstrate a tradeoff between particle filter performance
+demonstrate a trade-off between particle filter performance
 ($N_\text{eff}$ and degeneration rate) and parameter estimation.

### 2019-08-10 anzjs2018/sections/05_discuss.tex
More spelling mistakes

@@
 when there are 2-3~times as many vehicles---%
-within our 30-second timeframe target.
+within our 30-second time frame target.

### 2019-08-10 anzjs2018/anzjs.tex
more ...

@@
     We would like to thank Auckland Transport for providing free,
-    open access to the \rt data; and the Center for e-Research at the
+    open access to the \rt data; and the Centre for e-Research at the
     University of Auckland for providing and maintaining the virtual
@@
 
-\keywords{particle filter; transit modeling;
+\keywords{particle filter; transit modelling;
           transit networks; applications; gtfs}

### 2019-08-10 anzjs2018/sections/01_introduction.tex
more grammar

@@
 Previous research has shown that perceived waiting time is less
-when arrival time information is available \citep{TCRP_2003b},
-but in Auckland, RTI is highly unreliable,
+when arrival time information is available \citep{TCRP_2003b};
+however, in Auckland, RTI is highly unreliable,
 leading to frustration and ultimately deterring public transport users.

### 2019-08-10 anzjs2018/sections/03_models.tex
Use british speling

@@
 decreasing $\delta_k^{(i)}$ by one each time,
-after initializing $\delta_k^{(i)} = \delta_k$.
+after initialising $\delta_k^{(i)} = \delta_k$.
 In order to obtain the desired non-constant speed between observations,

### 2019-08-11 anzjs2018/sections/01_introduction.tex
Final (?) edits

@@
 their arrival times in real-time.
-Section~\ref{sec:gtfs} describes how we construct a transit network,
-which allows us to model travel times along physical roads,
+Section~\ref{sec:gtfs} describes how we constructed a transit network,
+allowing us to model travel times along physical roads,
 while Section~\ref{sec:models} presents two \rt models---%

### 2019-08-11 anzjs2018/sections/02_gtfs.tex
A makefile
Final (?) edits

@@
 making access easier for application developers.
-GTFS is developed and maintained by Google \citep{GoogleDevelopers_2006},
-who use it in Google Maps Transit Directions,
-and over 1100~transit providers use it around the world\footnote{%
+GTFS, developed and maintained by Google \citep{GoogleDevelopers_2006},
+% who use it in Google Maps Transit Directions,
+is used by over 1100~transit providers around the world\footnote{%
 source: \url{http://transitfeeds.com} as of August 2019},
@@
 It provides a standardised format for sharing vehicle positions and trip delays,
-and are typically accessed by developers via an API that \rt applications can use.
+which developers access via an API for use in \rt applications.
 
@@
 vehicle positions are used only to display \emph{where} the bus is.
-Vehicles report trip updates when they arrive at or depart from a stop,
-including the delay between the scheduled and actual arrival times.
+Vehicles report a trip update,
+which includes the delay between scheduled and actual arrival time,
+when they arrive at or depart from a stop.
 The delay is then propagated to all future stops to update their ETAs,
@@
 we can better predict the travel time between stops,
-which we hope will result in more accurate and reliable arrival time estimates.
+from which more accurate and, ideally, more reliable arrival esimates
+can be made.
 
@@
 which can, in turn, be used to predict travel times for all upcoming buses 
-travelling along the road,
+travelling along the same road,
 irrespective of the route they are servicing.
@@
 these positions can be pre-emptive,
-for example when approaching a queue of traffic at an intersection;
+for example when approaching a queue of traffic at an intersection:
 the way-point may trigger before the bus physically gets to it,

### 2019-08-11 anzjs2018/sections/03_models.tex
Final (?) edits

@@
 The underlying vehicle state at time $t_k$ consists of
-the vehicle's distance travelled $x$ in meters along the route and
-its speed $\dot x$ in meters per second.
+the vehicle's distance travelled $x_k$ in meters along the route and
+its speed $\dot x_k$ in meters per second.
 These are combined into the state vector
@@
 in recent vehicle modelling applications \citep{Ulmke_2006,Hans_2015}.
-It also makes it trivial to estimate travel times
+It also allows us to estimate travel times
 $\bz = (z_1,\ldots,z_L)^\top$, in seconds, along road segments
@@
 The state is expressed using the Dirac delta function $\dot\delta(\cdot)$,
-so that
 \begin{equation*}
@@
 \end{equation*}
-each of which is independently updated or \emph{mutated} using the transition function $f$ (Figure~\ref{fig:pf_state_predict}),
+which is updated of \emph{mutated} by applying the transition function $f$
+to each particle independently (Figure~\ref{fig:pf_state_predict}),
+giving us the proposal distribution
 \begin{equation*}
@@
 \end{equation*}
-which gives us the final posterior state estimate
+which leads to the final posterior state estimate
 \begin{equation*}
@@
 This occurs naturally over time, but the rate at which it does so depends 
-on a relationship between how well the transition function predicts future states
+on the relationship between how well the transition function predicts future states
 and the variance of those predictions (system noise).
-To avoid degeneration,
+If left unchecked, the predictive distribution reaches a point
+where all particles have weights which are effectively zero,
+at which point we say the particle filter has degenerated.
+To avoid this situation,
 particle filters use \emph{importance resampling}
@@
 
-The particle filter allows us to flexibly model bus behaviour,
-which currently includes
+The particle filter provides us with a flexible method
+with which to model bus behaviour.
+Currently, this model includes
 \begin{enumerate}
@@
 When the vehicle is approaching stop $j$,
-it stops with some probability $\pi_j$.
+it stops with probability $\pi_j$.
 If it does not stop, the remaining travel time $\delta_k$ is unaltered;
@@
 \citep{Hans_2015}---and then waits while passengers board and disembark,
-which is assumed to be exponentially distributed with mean $\tau_j$.
-This behaviour is expressed in the following model:
+which we assume to follow an exponential distribution with mean $\tau_j$.
+Stopping behaviour is expressed in the following model:
 \begin{equation}
@@
 in which (\ref{eq:trans_newton}) is evaluated iteratively one second at a time,
-decreasing $\delta_k^{(i)}$ by one each time,
+each step decreasing $\delta_k^{(i)}$ by one second,
 after initialising $\delta_k^{(i)} = \delta_k$.
@@
 along each segment $\ell$ of the route,
-storing the times so that the posterior distribution of the travel time
+storing the times so that the posterior distribution of travel time
 can be computed once all particles have completed segment $\ell$:
@@
 
-Computing the likelihood of the observation given each particle state,
+Computing the likelihood of the observation given a particle state,
 $p(\bY_k|\bX_k^{(i)})$,
@@
 a \emph{geographical projection} function $g$ to allow comparison of $\bY_k$,
-a GPS observation, with $\bX_k$, a distance travelled along the route in meters.
+a GPS observation, with $\bX_k$, 
+which includes the distance, in meters, travelled along the route.
 
@@
 and $R$ is the Earth's radius, in meters, then
-\begin{equation}
+\begin{equation*}
 \br = 
@@
     \end{bmatrix}
-\end{equation}
+\end{equation*}
 and, more importantly, the geographical distance $\mathcal{D}$ between the two 
@@
 and that \mbox{$\br \sim \mathrm{N}(\boldsymbol{0}, \omega^2\mathbf{I})$},
-which allows us to express $\br$ in terms of two independent
+allowing us to express $\br$ in terms of two independent
 standard normal random variables $z_1, z_2 \sim \mathrm{N}(0,1)$.
@@
 the observation error $e_{c\ell}^{m} \sim \mathrm{N}(0, R_{c\ell}^{m})$,
-and $\Delta_c = t_c - t_{c-1}$
+and $\Delta_c = t_c - t_{c-1}$:
 \begin{equation*}
@@
 \theta_c^\ell &= \theta_{c-1}^\ell + \Delta_{c\ell} v_{c\ell} \\
-Z_{c\ell}^{m} &= \theta_c^\ell + e_{c\ell}^{m}
+Z_{c\ell}^{m} &= \theta_c^\ell + e_{c\ell}^{m}.
 \end{split}
@@
 \hat \theta^\ell_{c|c-1} &= \hat \theta^\ell_{c-1|c-1} \\
-P^\ell_{c|c-1} &= P^\ell_{c-1|c-1} + (\Delta_{c\ell} \nu_{c\ell})^2
+P^\ell_{c|c-1} &= P^\ell_{c-1|c-1} + (\Delta_{c\ell} \nu_{c\ell})^2.
 \end{align*}
@@
 that traversed segment $\ell$ in the time interval $(t_{c-1}, t_c]$,
-so the update equation is
+so the update equations are simply
 \begin{align*}
@@
 \end{equation*}
-which can now be used in the prediction of travel times
+which can be used in the prediction of travel times
 for upcoming buses. 

### 2019-08-12 anzjs2018/response.txt
minor edits

Thank you for all of the very constructive feedback, I hope I've covered them adequately. Below I've replied to each point following the order/layout of the original comments.

Referee 1:
The issue of constant stopping probabilities and dwell times is something we're planning to explore with historical data, which I've now mentioned in the discussion (lines 331-333). After the model specification, I have now addressed the likely consequences of making the assumption (lines 187-191).

- I've clarified dimensions of the state vector (lines 131-133).
- The measurement function was a function of X and a sampled value of noise for generality, but I've changed this to h(x) + e in (1) to avoid confusion.
- I've re-expressed the likelihood equations to more clearly describe the relationship between X and Y, through the functions g and h (section 3.1.2)
- Degeneration has been described in more detail (para starting line 158), and PF described through model formulae instead of as an algorithm (section 3.1.1); included specification of N_eff value we used (line 170).
- Continuity improved: "intuitive likelihood" is followed with reasoning (line 153)

- spelling fixed, now using Figure/Section etc consistently, and using all UK English.
- description of network construction (removing route dependencies) has been clarified (section 2.1)
- figures improved to be consistent and clearer
- added citations for R, dplyr, and ggplot2 (all used to summarise and present simulation results)

Referee 2:
1. I've added a reference to the Ulmke 2006 paper (para starting line 35), which demonstrates the advantage of the PF when there's sparse sampling.

2. The main advantage of our approach is that we don't need to snap points to the route. I've included an example where "snapping" can fall over in previous (i.e., KF) approaches do highlight the advantage of avoiding it (bottom of page 5).

3. I've removed z from the vehicle state X and explained how we take advantage of the PF to estimate these values in real-time using eq (4).

4. Added state equations to show that Z is simply a measurement of theta, which is obtained from the PF (top of page 12)

5. Added missing units to parameters.

6. We're in the process of developing a simulation to test travel time estimation as part of the arrival time prediction component of the project, so these aren't available for the current paper.

7. Clarified that no correlation structure is used (line 230).

8. All figures improved/restructured to make them much easier to read and remove any ambiguity.

AE:
I've respecified the PF state equations using Dirac, which I agree is much better, thank you.

Instead of writing out the model algorithm, I've described the model using the suggested notation and briefly mentioned how this is implemented using an iterative algorithm. I don't think there was any advantage to including the full algorithm description, so I decided to remove it.

I've re-written the likelihood section to clarify g, h, and the relationship between X and Y, as well as fixed the definition to keep epsilon on the RHS.

Fixed typos in equation subscript (page 12, eq 3), and the noise term (page 12, eq 1).

And moved future work comments to the end.

### 2019-08-12 anzjs2018/sections/03_models.tex
minor edits

@@
 \item non-constant speed along roads (between observations), and
-\item bus stop behaviour, which involves optional stopping and waiting times
-    while passengers board and disembark.
+\item bus stop behaviour, (optional stopping and wait times while passengers board and disembark).
 \end{enumerate}
@@
 The second vehicle behaviour requires a model of bus stop wait times.
-When the vehicle is approaching stop $j$,
-it stops with probability $\pi_j$.
+When approaching stop $j$,
+the vehicle stops with probability $\pi_j$.
 If it does not stop, the remaining travel time $\delta_k$ is unaltered;
-otherwise, the bus waits for a minimum of $\gamma$ seconds---this 
-accounts for deceleration, opening and closing of doors, and acceleration
+otherwise, the bus waits for a minimum of $\gamma$ seconds---% 
+accounting for deceleration, opening and closing of doors, and acceleration
 \citep{Hans_2015}---and then waits while passengers board and disembark,
@@
 \tilde d_j^{(i)} &\sim \mathrm{Exponential}(\tau_j) \\
-d_j^{(i)} &= p_j^{(i)}(\gamma + \tilde d_j^{(i)}).
+d_j^{(i)} &= p_j^{(i)}(\gamma + \tilde d_j^{(i)}),
 \end{split}
 \end{equation}
+using constant values of $\pi_j = 0.5$ and $\tau_j = 10$~seconds for all stops;
+if available, historical dwell time information can easily be used
+to specify each $\tau_j$,
+or even replace the exponential prior altogether.
+As it stands, we found these values to be vague enough that
+the actual vehicle trajectory was sampled by enough of the particles
+to prevent degeneration of the filter;
+the consequence being increased variance of travel time estimates.
 
@@
 after initialising $\delta_k^{(i)} = \delta_k$.
-In order to obtain the desired non-constant speed between observations,
-$\epsilon_k^{(i)}$ is re-sampled each second,
-allowing the particle's speed to vary.
+% To allow non-constant speed between observations,
+We also re-sampled
+$\epsilon_k^{(i)}$ each second
+to allow the particle's speed to vary between observations.
 This iterative approach allows the particle to approach the next stop,

### 2019-08-12 anzjs2018/sections/05_discuss.tex
minor edits

@@
 We have focussed on making a general framework
-in which the existing models used for specific components
-can easily be changed---%
-for example, using a more complex model of stop dwell times
-based on historical data in place of the exponential---%
-or new ones added, such as implementing intersection-specific behaviour
-if intersection locations are available.
-Future work is planned to detect route meeting points
-and use these as intersection nodes
- to further reduce road segment overlap.
+into which new models can be incorporated,
+such as implementing intersection-specific behaviour
+once intersection locations are available,
+or exiting models changed.
+For example, we plan to model stop dwell time using historical data,
+which will allow us to replace the exponential distribution
+with something more appropriate.
 

### 2020-03-31 anzjs2018/anzjs.tex
Add line numbers
Editor changes

@@
 \def\volumeyear{2018}
-% \linenumbers
+\linenumbers
 
@@
 We demonstrate the \rt feasibility and performance of our approach in real-time,
-where a combination of R and C++ was used to obtain the necessary performance results.
+where a combination of \textsf{R} and \textsf{C++} was used to obtain the necessary performance results.
 Future work will use these estimated traffic conditions

### 2020-03-31 anzjs2018/sections/01_introduction.tex
Editor changes

@@
 leaving passengers unsure as to when---if at all---their bus will show up,
-a phenomenon referred to as ``ghost buses'' by transport bloggers.
+a phenomenon referred to as `ghost buses' by transport bloggers.
 

### 2020-03-31 anzjs2018/sections/02_gtfs.tex
Editor changes

@@
 % who use it in Google Maps Transit Directions,
-is used by over 1100~transit providers around the world\footnote{%
-source: \url{http://transitfeeds.com} as of August 2019},
+is used by over 1100~transit providers around the world
+(sourced from \url{http://transitfeeds.com} on August 2019),
 including here in Auckland, New Zealand.
@@
 There are two components to GTFS.
-The first, \emph{GTFS static}, includes information about
+The first, \emph{GTFS static}, includes information about:
 \begin{itemize}

### 2020-03-31 anzjs2018/sections/04_implementation.tex
Add line numbers
minor edits
Editor changes

@@
 while the second implements the \rt models and prediction.
-We used Rcpp \citep{Rcpp}
+We used \textsf{Rcpp} \citep{Rcpp}
 to develop the program,
-which provides access to R \citep{rcore},
+which provides access to \textsf{R} \citep{rcore},
 which contains many useful tools for data manipulation and package development,
-as well as the speed and memory management capabilities of C++.
-The program is implemented in the R package
+as well as the speed and memory management capabilities of \textsf{C++}.
+The program is implemented in the \textsf{R} package
 \verb+transitr+, available on Github (\url{https://github.com/tmelliott/transitr}).
@@
 Below is the general structure of the \rt component,
-where the bold steps are those discussed in this paper.
+where the bold steps are those discussed in this paper:
 \begin{enumerate}
 \item Load GTFS data from database
-\item Each time new data are received \ldots
+\item Each time new data are received, do:
 \begin{enumerate}
@@
 with a target of 30~seconds or faster at peak time.
-Using C++ provided the memory management control necessary to make this possible.
+Using \textsf{C++} provided the memory management control necessary to make this possible.
 Since we modelled each vehicle independently,
 it was straightforward to spread the vehicle updates
-over multiple cores using OpenMP without thread-safety concerns.
+over multiple cores using \textsf{OpenMP} without thread-safety concerns.
 
@@
 with 8~Intel Xeon 3.00GHz CPU cores and 32~GB of memory,
-running Ubuntu~16.04 and using R~3.4.1.
+running \textsf{Ubuntu}~16.04 and using \textsf{R}~3.4.1.
+The simulations were processed locally using \textsf{R}~3.6.0,
+and made use of the \textsf{R} packages \verb+dplyr+ \citep{dplyr}
+for manipulating and summarising the results,
+and \verb+ggplot2+ \citep{ggplot2} to graph them.
 
@@
 While not discussed here this involved computing quantiles which uses a sorting algorithm
-with complexity $O(N \log N)$,
+with complexity $\mathcal{O}(N \log N)$,
 which explains the non-linear relationship with $N$.
@@
 v_\ell^{\omega,\sigma} &=
-\frac{\mathrm{Var}(\tilde \bz_\ell^{\omega,\sigma})}{\mathrm{Var}(\tilde \bz_\ell)} \\
+\frac{\mathrm{Var}(\tilde \bz_\ell^{\omega,\sigma})}{\mathrm{Var}(\tilde \bz_\ell)}, \\
 \bar v^{\omega,\sigma} &=
-    \frac{1}{L}\sum_{\ell=1}^L v_\ell^{\omega,\sigma}
-\end{split}.
+    \frac{1}{L}\sum_{\ell=1}^L v_\ell^{\omega,\sigma}.
+\end{split}
 \end{equation*}

### 2020-03-31 anzjs2018/sections/05_discuss.tex
Editor changes

@@
-\section{Discussion and Future Work}
+\section{Discussion and future work}
 \label{sec:discussion}
@@
 should allow us to model all vehicles during peak periods---%
-when there are 2-3~times as many vehicles---%
+when there are 2--3~times as many vehicles---%
 within our 30-second time frame target.

### 2020-04-13 anzjs2018/anzjs.tex
Prep submission

@@
 
-\def\volumeyear{2018}
+\def\volumeyear{2020}
 \linenumbers
@@
 \maketitle
-\input{sections/01_introduction}
-\input{sections/02_gtfs}
-\input{sections/03_models}
-\input{sections/04_implementation}
-\input{sections/05_discuss}
+
+\section{Introduction}
+\label{sec:intro}
+
+In public transport, \rt information (RTI)---%
+most notably estimated times of arrival (ETAs)---%
+keeps transport users informed of changes to their commute,
+allowing them to plan their journey accordingly.
+Previous research has shown that perceived waiting time is less
+when arrival time information is available \citep{TCRP_2003b};
+however, in Auckland, RTI is highly unreliable,
+leading to frustration and ultimately deterring public transport users.
+Inaccurate ETAs are the primary source of this frustration,
+as they fluctuate---sometimes dramatically---over time.
+While unpredictable changes to traffic conditions may cause this,
+it can equally be due to poor schedule calibration.
+Furthermore, buses are shown as \emph{on-time}
+if they are not present in the \rt system and
+the service provider has not manually cancelled them,
+in which case ETAs are based solely on scheduled arrival times.
+Once the scheduled arrival time has passed,
+the service is removed from the \rt board,
+leaving passengers unsure as to when---if at all---their bus will show up,
+a phenomenon referred to as `ghost buses' by transport bloggers.
+
+Arrival time prediction is only as reliable as the underlying model,
+and while considerable work has gone into developing public transport vehicle models
+\citep{Cathey_2003,Jeong_2005,Yu_2011,Hans_2015},
+many public transport providers---notably Auckland Transport---use no formal model.
+Instead, ETAs are based solely on the scheduled arrival time
+adjusted by the vehicle's arrival or departure delay at the most recent stop,
+\emph{if available}.
+This method assumes that the schedule is accurate
+and there is no unusual congestion along the route.
+Neither of these is a valid assumption,
+particularly in our test area of Auckland, New Zealand,
+where infrastructure for buses (such as priority lanes) is limited,
+and bus drivers do not actively adhere to the schedule.
+A more robust modelling and prediction framework
+based on \rt congestion information would avoid making these assumptions.
+Such a framework should consist of a robust vehicle model to estimate the position and speed
+of transit vehicles from \rt GPS data,
+and a means of combining speed information from vehicles
+to model traffic flow.
+This information can then be used to improve arrival time predictions.
+
+Several vehicle modelling approaches were explored,
+including the Kalman filter \citep{Dailey_2001,Cathey_2003},
+machine learning models \citep{Yu_2006,Chang_2010},
+and the particle filter \citep{Hans_2015},
+which has proven itself as a robust option for
+\rt vehicle tracking applications
+\citep{Gustafsson_2002,Davidson_2011}.
+\cite{Ulmke_2006} compared both a Gaussian sum approximation and a particle filter
+to track a vehicle in real-time,
+and although not their primary aim,
+showed how the particle filter behaved better in some difficult situations.
+In particular, the particle filter accurately estimated the \emph{uncertainty}
+of the vehicle's location when the observations were unavailable for a time.
+This situation often occurs with transit data---%
+the bus may be stopped at a bus stop, or it could be continuing along the route but,
+for some other reason, failing to report its position.
+Choosing to implement a particle filter has allowed us
+to develop a simple, flexible framework
+which is robust to the type of data we expect to observe.
+
+The second component of our framework
+involves estimating traffic conditions throughout the transport network
+using the information obtained from the vehicle model.
+\cite{Yu_2010} improved prediction accuracy by using the travel times
+of previous buses travelling along the same route.
+A similar method presented by \cite{Hans_2015}
+used headway, the time between consecutive vehicles at a point on the route,
+as a predictor of travel time.
+Since these approaches only work well on high-frequency routes,
+\cite{Yu_2011} showed further improvements by combining travel times
+from several routes;
+however, their method was limited to predefined converging routes.
+In general, however, no comprehensive network modelling approach has been proposed using
+only GPS position data to model and account for congestion when estimating arrival times.
+
+In this paper, we propose a simple framework for modelling transit vehicles and predicting
+their arrival times in real-time.
+Section~\ref{sec:gtfs} describes how we constructed a transit network,
+allowing us to model travel times along physical roads,
+while Section~\ref{sec:models} presents two \rt models---%
+one for each of the vehicle and network states---%
+with the primary goal of estimating \rt traffic states throughout the network.
+Finally, we discuss the \rt feasibility and performance
+of the particle filter in Section~\ref{sec:rt}.
+
+\section{Working with \rt transit data}
+\label{sec:gtfs}
+
+GTFS (general transit feed specification)
+is an API (application programming interface) specification for transit data
+detailing how it should be organised,
+making access easier for application developers.
+GTFS, developed and maintained by Google \citep{GoogleDevelopers_2006},
+% who use it in Google Maps Transit Directions,
+is used by over 1100~transit providers around the world
+(sourced from \url{http://transitfeeds.com} on August 2019),
+including here in Auckland, New Zealand.
+An advantage of this standardised format is that,
+provided an application depends solely on GTFS data,
+after developing it locally in Auckland it will be deployable to any other GTFS-based
+public transport system with minimal modification.
+
+There are two components to GTFS.
+The first, \emph{GTFS static}, includes information about:
+\begin{itemize}
+\item \emph{stops}, a physical location where passengers can embark and disembark the vehicle;
+\item \emph{routes}, a sequence of two or more stops displayed as a single service;
+\item \emph{trips}, an instance of a route occurring at a specific time of day;
+\item \emph{schedules}, specifying the arrival (and departure) times for each bus at each of its stops;
+\item and \emph{shapes}, the sequence of points defining a vehicle's path along a route.
+\end{itemize}
+The second component, \emph{GTFS-realtime},
+is only available in a subset of providers due to the requirement of
+on-board GPS tracking devices and a central server.
+It provides a standardised format for sharing vehicle positions and trip delays,
+which developers access via an API for use in \rt applications.
+
+As mentioned in Section~\ref{sec:intro},
+there are some fundamental issues with the arrival time prediction method currently
+deployed in Auckland,
+which is based entirely on \emph{GTFS-realtime} trip updates---%
+vehicle positions are used only to display \emph{where} the bus is.
+Vehicles report a trip update,
+which includes the delay between scheduled and actual arrival time,
+when they arrive at or depart from a stop.
+The delay is then propagated to all future stops to update their ETAs,
+as demonstrated in Figure~\ref{fig:gtfs-delays}.
+As already mentioned, this assumes that the schedule is well-calibrated
+and that the time between stops
+is representative of the real-world travel time between them.
+Figure~\ref{fig:gtfs-delays} demonstrates the fallibility of this prediction method,
+where we see the incremental lateness causing the passenger's ETA to
+jump each time the bus arrives at a stop.
+By using \rt travel time information,
+we can better predict the travel time between stops,
+from which more accurate and, ideally, more reliable arrival esimates
+can be made.
+
+\begin{figure}[tb]
+    \centering
+    \includegraphics[width=0.8\textwidth]{figures/02_gtfs_delays.pdf}
+    \caption{
+        On the left, scheduled arrival times are shown as solid points,
+        for several stops along a route. The actual arrival times
+        are shown with crosses,
+        and the GTFS-based ETAs for stop 5, $\{\hat A_1, \hat A_2, \hat A_3\}$, are shown.
+        On the right,
+        for a passenger arriving at stop 5 at 12:04,
+        ETA predictions
+        $\mathrm{ETA}_j = \mathrm{Time} - \hat A_j$ are shown
+        to demonstrate the fluctuation in
+        arrival time as the bus arrives at each stop.
+    }
+    \label{fig:gtfs-delays}
+\end{figure}
+
+\subsection{Transit network construction}
+\label{sec:network_build}
+
+The primary predictor of arrival time is
+the travel time along roads between where the bus is \emph{now},
+and the stop where a passenger waits.
+In most applications, however, this vital information is unavailable, at least directly.
+In order to compute travel times efficiently
+and make use of all available data pooled across multiple routes,
+we constructed a \emph{road network},
+allowing us to model roads directly,
+such that every bus travelling along a road contributes to its state,
+which can, in turn, be used to predict travel times for all upcoming buses
+travelling along the same road,
+irrespective of the route they are servicing.
+
+Constructing the transit network involved splitting routes
+into spatially identifiable segments,
+each representative of a physical road,
+using bus stops as nodes in the network
+and the connecting roads as edges
+(Figure~\ref{fig:network_creation}).
+In this way, routes that service the same subsequence of stops
+all contribute to traffic flow information for the connecting roads.
+Although there are several drawbacks to this method,
+such as road segment overlaps where routes merge between stops,
+our approach is sufficient for assessing the \rt
+feasibility of our proposed prediction framework.
+
+\begin{figure}[tb]
+    \centering
+    \begin{subfigure}{0.7\textwidth}
+        \centering
+        \includegraphics[width=0.9\textwidth]{figures/02_network_segments_1.pdf}
+        \caption{Raw GTFS route shapes}
+        \label{fig:network_creation_1}
+    \end{subfigure} \\
+    \begin{subfigure}{0.7\textwidth}
+        \centering
+        \includegraphics[width=0.9\textwidth]{figures/02_network_segments_2.pdf}
+        \caption{GTFS-based road network}
+        \label{fig:network_creation_2}
+    \end{subfigure}
+    \caption{
+        Bus stops (shown as dots above) can be serviced by more than one route,
+        as shown in (a), with different line types representing three separate routes.
+        The constructed road network, using stops as nodes, is shown in (b),
+        in which the connecting lines represent physical road segments,
+        with line width proportional to the number of individual routes
+        travelling along that road.
+    }
+    \label{fig:network_creation}
+\end{figure}
+
+\subsection{Real-time vehicle locations}
+\label{sec:realtime_data}
+
+\emph{GTFS-realtime} provides the positions of vehicles in a transit network.
+The data consists of the time $t_k$ that the position was last updated,
+the GPS position of the vehicle, $\bY_k$,
+and information about the serviced trip.
+In Auckland, vehicle positions are updated with a frequency
+of anywhere between 10~seconds and several minutes,
+so there is often considerable uncertainty about a vehicle's intermediate trajectory,
+in particular when there are one or more bus stops along the way.
+It is also possible for a bus to remain stationary,
+such as when there is heavy congestion,
+so the number of possible trajectories rapidly increases with
+the time between observations.
+
+An important consideration regarding Auckland Transport's \rt implementation is that
+buses are programmed to report their location when arriving at or departing from
+bus stops, as well as some major intersections.
+To further complicate this,
+these positions can be pre-emptive,
+for example when approaching a queue of traffic at an intersection:
+the way-point may trigger before the bus physically gets to it,
+so consecutive observations may show what appears to be a bus travelling backwards.
+To handle this, we compute the approximate distance travelled, $\tilde x_k$,
+of the vehicle by finding the nearest point on the path to the observation;
+if this has decreased, we reject the current state (based on the pre-emptive observation)
+and reset the vehicle
+to its previous state before continuing.
+So at the cost of losing one (likely invalid) observation,
+we can improve predictive performance
+when this event occurs.
+
+\section{The models}
+\label{sec:models}
+
+Real-time tracking applications often use
+a Bayesian filtering approach,
+or recursive Bayesian estimation,
+in which the data is processed sequentially instead of all at once,
+reducing computational demands
+\citep{Anderson_2012}.
+These models assume an underlying Markov process with state $\boldsymbol{\mathcal{X}}_k$,
+an $n\times1$ column vector,
+with system noise $\boldsymbol{\omega}_k\sim\mathrm{N}(\boldsymbol{0},\mathcal{Q})$
+of which observations $\boldsymbol{\mathcal{Y}}_k$,
+an $m\times1$ column vector, are made
+with error $\boldsymbol{\nu}_k\sim\mathrm{N}(\boldsymbol{0},\mathcal{R})$,
+giving the following general model:
+\begin{equation}
+\label{eq:rbe_model}
+\begin{split}
+\boldsymbol{\mathcal{X}}_k &= f(\boldsymbol{\mathcal{X}}_{k-1}, \boldsymbol{\omega}_k), \\
+\boldsymbol{\mathcal{Y}}_k &= h(\boldsymbol{\mathcal{X}}_k) + \boldsymbol{\nu}_k.
+\end{split}
+\end{equation}
+The transition function $f:\mathbb{R}^n\mapsto\mathbb{R}^n$
+describes the relationship between consecutive states,
+while the measurement function $h:\mathbb{R}^n\mapsto\mathbb{R}^m$ is a deterministic function
+mapping the underlying state to the observable state,
+of which noisy measurements are made.
+
+The following sections describe two Bayesian filters used in this application.
+The first, implemented using a particle filter, estimates vehicle states
+with the primary goal of estimating road travel times from a sequence of GPS positions.
+The second uses these travel times to update road states
+and is implemented using an \emph{information filter},
+a variant of the Kalman filter \citep{Anderson_2012}.
+
+\subsection{Real-time vehicle model}
+\label{sec:pf}
+
+The underlying vehicle state at time $t_k$ consists of
+the vehicle's distance travelled $x_k$ in meters along the route and
+its speed $\dot x_k$ in meters per second.
+These are combined into the state vector
+$\bX_k = \left[x_k\ \dot x_k\right]^\top$,
+of which observations $\bY_k$ are made using a GPS
+with error $\omega^2$,
+giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle
+as $\bY_k = \left[\lambda_k\ \phi_k \right]^\top$.
+We use a particle filter to estimate $\bX_k$
+due to its flexibility and proven robustness
+in recent vehicle modelling applications \citep{Ulmke_2006,Hans_2015}.
+It also allows us to estimate travel times
+$\bz = (z_1,\ldots,z_L)^\top$, in seconds, along road segments
+as the vehicle traverses the route.
+
+The primary advantage of the particle filter is its handling of multimodality,
+as demonstrated in Figure~\ref{fig:pf_state_predict},
+which is a common feature of the proposal distribution, particularly around bus stops.
+Another advantage is that the likelihood function is intuitive, based
+on the physical distance between the vehicle observation and the state's
+estimate of the actual location, $h(\bX_k)$ (Section~\ref{sec:pf_update}).
+Conversely, particle filter methods are computationally demanding,
+requiring an increasing number of particles as model complexity and
+the number of parameters increases \citep{Carpenter_1999}.
+Section~\ref{sec:rt} describes our implementation of the particle filter
+and its \rt performance.
+
+\begin{figure}[p]
+    \centering
+    \begin{subfigure}[t]{1\textwidth}
+        \centering
+        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_1.pdf}
+        \caption{
+            Vehicle state is mapped to observation space by the
+            measurement function $h$.
+        }
+        \label{fig:pf_state_prev}
+    \end{subfigure}\\
+    \begin{subfigure}[t]{0.9\textwidth}
+        \centering
+        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_2.pdf}
+        \caption{
+            The transition function $f$ predicts the future state
+            of each particle.
+        }
+        \label{fig:pf_state_predict}
+    \end{subfigure}\\
+    \begin{subfigure}[t]{0.9\textwidth}
+        \centering
+        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_4.pdf}
+        \caption{
+            The state is updated by resampling with likelihood weights based on the distance
+            between the particle and the observation, shown as a (red) cross.
+        }
+        \label{fig:pf_state_predict2}
+    \end{subfigure}
+    \caption{
+        The particle filter approximates vehicle state using a set of
+        discrete points (a), which are each mutated independently (b)
+        to predict the next state.
+        The measurement function $h$ maps each particle state
+        to a GPS coordinate (a),
+        which can be compared to the observed location (c)
+        to calculate each particle's likelihood to compute resampling weights.
+    }
+    \label{fig:pf_state}
+\end{figure}
+
+\afterpage{\clearpage}
+
+In a particle filter, the posterior distribution of the state at time $t_{k-1}$
+is represented by a set of discrete points, or particles, (Figure~\ref{fig:pf_state_prev}),
+each with an associated weight $W_{k-1}^{(i)}$.
+The state is expressed using the Dirac delta function $\dot\delta(\cdot)$,
+\begin{equation*}
+p(\bX_{k-1} | \bY_{1:k-1}) \approx
+    \sum_{i=1}^N W_{k-1}^{(i)} \dot\delta(\bX_{k-1} - \bX_{k-1}^{(i)}),
+\end{equation*}
+which is updated of \emph{mutated} by applying the transition function $f$
+to each particle independently (Figure~\ref{fig:pf_state_predict}),
+giving us the proposal distribution
+\begin{equation*}
+p(\bX_k | \bX_{k-1}) \approx
+    \sum_{i=1}^N W_{k-1}^{(i)} \dot\delta(\bX_{k} - f(\bX_{k-1}^{(i)}, \boldsymbol{\psi})),
+\end{equation*}
+where $\boldsymbol{\psi}$ is a vector containing all of the necessary model parameters
+(Section~\ref{sec:pf_prediction}).
+After mutation,
+the state is updated by reweighting the particles using the likelihood $p(\bY_k | \bX_k^{(i)})$
+(Section~\ref{sec:pf_update}) and standardising so $\sum_{i=1}^N W_k^{(i)} = 1$,
+\begin{equation*}
+W_k^{(i)} = \frac{W_{k-1}^{(i)} p(\bY_k | \bX_k^{(i)})}{
+    \sum_{j=1}^N W_{k-1}^{(j)} p(\bY_k | \bX_k^{(j)})
+},
+\end{equation*}
+which leads to the final posterior state estimate
+\begin{equation*}
+p(\bX_k | \bY_k) \approx
+    \sum_{i=1}^N W_{k}^{(i)} \dot\delta(\bX_{k} - \bX_{k}^{(i)}).
+\end{equation*}
+
+One problem with particle filters is degeneration:
+when the particles become too spread out with respect to the predictive distribution,
+the weight will be concentrated on to just a few of them.
+This occurs naturally over time, but the rate at which it does so depends
+on the relationship between how well the transition function predicts future states
+and the variance of those predictions (system noise).
+If left unchecked, the predictive distribution reaches a point
+where all particles have weights which are effectively zero,
+at which point we say the particle filter has degenerated.
+To avoid this situation,
+particle filters use \emph{importance resampling}
+using importance weights $\{W_k^{(1)}, \ldots, W_k^{(N)}\}$,
+thereby removing improbable particles and replacing them with more probable ones,
+as demonstrated in Figure~\ref{fig:pf_state_predict2}.
+However, resampling requires sorting $N$ particles,
+which is of $\mathcal{O}(N\log N)$ complexity,
+so to reduce computational costs, resampling is performed only when
+the effective sample size $N_{\text{eff}} = 1 / \sum_i (W_k^{(i)})^2$
+falls below a specified threshold $N_{\text{thres}}$
+\citep{Gustafsson_2002}.
+We used a fixed threshold value of $N_{\text{thres}} = N/4$
+for the current work.
+
+\subsubsection{Vehicle transition function}
+\label{sec:pf_prediction}
+
+The particle filter provides us with a flexible method
+with which to model bus behaviour.
+Currently, this model includes:
+\begin{enumerate}
+\item non-constant speed along roads (between observations), and
+\item bus stop behaviour, (optional stopping and wait times while passengers board and disembark).
+\end{enumerate}
+The transition function consists of two models,
+one for each of the above behaviours.
+The first models the speed and distance travelled by the vehicle,
+\begin{equation}
+\label{eq:trans_newton}
+\bX_k^{(i)} = f(\bX_{k-1}^{(i)}, \delta_k, \epsilon_k^{(i)}) =
+    \begin{bmatrix}
+        x_{k-1}^{(i)} + \delta_k(\dot x_{k-1}^{(i)} + \epsilon_k^{(i)}) \\
+        \dot x_{k-1}^{(i)} + \epsilon_k^{(i)}
+    \end{bmatrix},
+\end{equation}
+where $\delta_k = t_k - t_{k-1}$
+and $\epsilon_k^{(i)}\sim\mathrm{N}_T(0, \sigma^2)$, truncated to the interval
+$(-\dot x_{k-1}^{(i)}, 30 - \dot x_{k-1}^{(i)})$
+to ensure vehicle speed is always positive and less than 30~m/s
+(approximately 110~km/h).
+
+The second vehicle behaviour requires a model of bus stop wait times.
+When approaching stop $j$,
+the vehicle stops with probability $\pi_j$.
+If it does not stop, the remaining travel time $\delta_k$ is unaltered;
+otherwise, the bus waits for a minimum of $\gamma$ seconds---%
+accounting for deceleration, opening and closing of doors, and acceleration
+\citep{Hans_2015}---and then waits while passengers board and disembark,
+which we assume to follow an exponential distribution with mean $\tau_j$.
+Stopping behaviour is expressed in the following model:
+\begin{equation}
+\label{eq:trans_stop}
+\begin{split}
+p_j^{(i)} &\sim \mathrm{Bernoulli}(\pi_j), \\
+\tilde d_j^{(i)} &\sim \mathrm{Exponential}(\tau_j), \\
+d_j^{(i)} &= p_j^{(i)}(\gamma + \tilde d_j^{(i)}),
+\end{split}
+\end{equation}
+using constant values of $\pi_j = 0.5$ and $\tau_j = 10$~seconds for all stops;
+if available, historical dwell time information can easily be used
+to specify each $\tau_j$,
+or even replace the exponential prior altogether.
+As it stands, we found these values to be vague enough that
+the actual vehicle trajectory was sampled by enough of the particles
+to prevent degeneration of the filter;
+the consequence being increased variance of travel time estimates.
+
+The model components are implemented as an algorithm,
+in which (\ref{eq:trans_newton}) is evaluated iteratively one second at a time,
+each step decreasing $\delta_k^{(i)}$ by one second,
+after initialising $\delta_k^{(i)} = \delta_k$.
+% To allow non-constant speed between observations,
+We also re-sampled
+$\epsilon_k^{(i)}$ each second
+to allow the particle's speed to vary between observations.
+This iterative approach allows the particle to approach the next stop,
+at which point (\ref{eq:trans_stop}) is evaluated for the particle
+and the dwell time subtracted from $\delta_k^{(i)}$.
+This process continues until $\delta_k^{(i)}$ reaches zero.
+The algorithm also keeps track of each particle's travel time $z_\ell^{(i)}$
+along each segment $\ell$ of the route,
+storing the times so that the posterior distribution of travel time
+can be computed once all particles have completed segment $\ell$:
+\begin{equation}
+\label{eq:post_tt}
+p(z_\ell | \bY_{1:k}) \approx
+    \sum_{i=1}^N W_k^{(i)} \dot\delta(z_\ell - z_\ell^{(i)}).
+\end{equation}
+
+\subsubsection{Updating state using the observation likelihood}
+\label{sec:pf_update}
+
+Computing the likelihood of the observation given a particle state,
+$p(\bY_k|\bX_k^{(i)})$,
+uses the measurement function $h$ and
+a \emph{geographical projection} function $g$ to allow comparison of $\bY_k$,
+a GPS observation, with $\bX_k$,
+which includes the distance, in meters, travelled along the route.
+
+The measurement function $h$ computes GPS coordinates by using the
+shape information provided by GTFS and travelling $x_k$ meters along
+this two-dimensional line.
+Meanwhile, we use the \emph{Equirectangular projection} \citep{Snyder_1998},
+$\br = g(\bY_1 | \bY_0)$,
+such that the magnitude of $\boldsymbol{r}$ is equal to the physical distance
+between $\bY_0$ and $\bY_1$ and both dimensions are in meters.
+Let $\bY_i = [\lambda_i, \psi_i]^\top$,
+with $\lambda_i$ and $\psi_i$ measured in radians,
+and $R$ is the Earth's radius, in meters, then
+\begin{equation*}
+\br =
+g(\bY_1 | \bY_0) =
+    \begin{bmatrix}
+        r_1 \\ r_2
+    \end{bmatrix} =
+    R \begin{bmatrix}
+        (\lambda_1 - \lambda_0)\cos\phi_0 \\
+        \phi_1 - \phi_0
+    \end{bmatrix},
+\end{equation*}
+and, more importantly, the geographical distance $\mathcal{D}$ between the two
+GPS coordinates is given by
+\begin{equation}
+\label{eq:dist}
+\mathcal{D}(\bY_0, \bY_1) = ||g(\bY_0 | \bY_1)|| = \sqrt{r_1^2 + r_2^2}.
+\end{equation}
+
+Now we assume GPS observations have a known error of $\omega^2$ meters,
+and that \mbox{$\br \sim \mathrm{N}(\boldsymbol{0}, \omega^2\mathbf{I})$},
+allowing us to express $\br$ in terms of two independent
+standard normal random variables $z_1, z_2 \sim \mathrm{N}(0,1)$.
+Now the distance between a vehicle observation $\bY_k$
+and a state $\bX_k$, using (\ref{eq:dist}) and the measurement function $h$,
+is expressed as
+\begin{equation}
+\label{eq:obs_dist}
+\mathcal{D}(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2}
+    = \sqrt{(\omega z_1)^2 + (\omega z_2)^2}
+    = \sqrt{\omega^2 (z_1^2 + z_2^2)}.
+\end{equation}
+
+Finally, the sum of two independent
+standard normal random variables has a Chi-square distribution with two degrees of freedom,
+which is also an exponential distribution with rate~$\frac{1}{2}$,
+and if $Z \sim \mathrm{Exponential}(\theta)$ then
+$cZ \sim \mathrm{Exponential}(\frac{\theta}{c})$;
+therefore, the \emph{squared} distance from (\ref{eq:obs_dist}) is distributed as
+\begin{equation}
+\label{eq:obs_exp}
+\mathcal{D}(\bY_k, h(\bX_k))^2 =
+\omega^2(z_1^2 + z_2^2) \sim \mathrm{Exponential}\left(\frac{1}{2\omega^2}\right).
+\end{equation}
+
+The likelihood of the observation $\bY_k$ given a state estimate $\bX_k^{(i)}$
+can now be expressed using (\ref{eq:obs_exp})
+and the probability density function of the exponential distribution,
+\begin{equation}
+\label{eq:lhood}
+p(\bY_k | \bX_k^{(i)}, \omega) =
+\frac{1}{2\omega^2}\exp\left\{
+    -\frac{\mathcal{D}(\bY_k, h(\bX_k^{(i)}))^2}{2\omega^2}
+\right\},
+\end{equation}
+where the distance between two GPS coordinates is easily computed,
+allowing for fast evaluation of (\ref{eq:lhood}) within the particle filter algorithm.
+
+It is worth noting that this representation of the likelihood is only
+possible due to the discrete nature of the particle filter state estimate;
+the Kalman filter---which has often been used for transit vehicle tracking---%
+uses the measurement \emph{matrix}, requiring a linear
+transformation between the state and its observations.
+To do so, applications first estimate the \emph{observed distance travelled}
+by snapping the GPS observations to the route,
+which introduces unnecessary error and uncertainty into the model.
+Our approach avoids this, which makes it more stable in locations where two
+parts of the route are close to each other,
+such as at loops, where a single GPS observation might have two likely ``snapping'' points.
+
+\subsection{Network model}
+\label{sec:kf}
+
+The primary objective of the network model is to estimate the \rt traffic conditions
+(travel time) along roads in the transit network
+and make short-term forecasts for estimating arrival times.
+In this paper, we model each road segment independently,
+as excluding correlations not only simplifies the model to a one-dimensional Kalman filter,
+but it also means computations can be run in parallel,
+further improving the real-time performance of the model.
+
+The network state $\boldsymbol\theta_c = \{\theta_c^\ell\}_{\ell = 1}^L$ is the travel time
+of transit vehicles along road segment $\ell$ at time $t_c$,
+of which observations $Z_{c\ell}^{m}$
+are obtained from the particle filter for vehicle $m$ at time $t_c$,
+as defined in (\ref{eq:post_tt}), such that
+(after adding the $m$ superscript to identify unique vehicles),
+\begin{equation}
+\label{eq:tt_obs_mean}
+Z_{c\ell}^{m} = \mathrm{E}(z_{\ell}^{m} | \bY^m_{1:c}) =
+\sum_{i=1}^N (W_c^{(i)})^m (z_\ell^{(i)})^{m}.
+\end{equation}
+The measurement error is also estimated from the particle filter,
+\begin{equation}
+\label{eq:tt_obs_err}
+R_{c\ell}^m = \mathrm{var}(z_{\ell}^{m} | \bY^m_{1:c}) =
+\sum_{i=1}^N (W_c^{(i)})^m \left((z_\ell^{(i)})^{m} - Z_{c\ell}^{m}\right)^2.
+\end{equation}
+
+The model of travel times becomes a simple reduction of (\ref{eq:rbe_model})
+to the one-dimensional case,
+with $f$ and $h$ both unity,
+system noise $v_{c\ell} \sim \mathrm{N}(0, \nu_\ell^2)$,
+where $\nu_\ell^2$ is the average change in travel time per second
+along road segment $\ell$,
+the observation error $e_{c\ell}^{m} \sim \mathrm{N}(0, R_{c\ell}^{m})$,
+and $\Delta_c = t_c - t_{c-1}$:
+\begin{equation*}
+\begin{split}
+\theta_c^\ell &= \theta_{c-1}^\ell + \Delta_{c\ell} v_{c\ell} \\
+Z_{c\ell}^{m} &= \theta_c^\ell + e_{c\ell}^{m}.
+\end{split}
+\end{equation*}
+
+Since multiple vehicles can travel along a road simultaneously,
+we used an information filter to implement the network model.
+The information filter is a transformation of the Kalman filter in which the
+\emph{information matrix} and \emph{information vector} are used in place of
+the covariance matrix and state vector, respectively,
+allowing multiple observations to be added together to update the state
+in a single iteration.
+
+The state at time $t_{c-1}$ is parameterised by its mean and variance,
+conditional on $\boldsymbol{Z}_{1:c-1,\ell}$,
+the data from all vehicles up until time $t_{c-1}$,
+\begin{equation*}
+\hat \theta_{c-1|c-1}^\ell =
+\mathrm{E}(\theta_{c-1}^\ell | \boldsymbol{Z}_{1:c-1,\ell})
+\quad\text{and}\quad
+P_{c-1|c-1}^\ell =
+\mathrm{var}(\theta_{c-1}^\ell | \boldsymbol{Z}_{1:c-1,\ell}),
+\end{equation*}
+respectively,
+which are predicted from the previous state estimate using the predictive model
+\begin{align*}
+\label{eq:kf_transition}
+\hat \theta^\ell_{c|c-1} &= \hat \theta^\ell_{c-1|c-1}, \\
+P^\ell_{c|c-1} &= P^\ell_{c-1|c-1} + (\Delta_{c\ell} \nu_{c\ell})^2.
+\end{align*}
+
+For the update step, the parameters are transformed into an information
+space parameterised by the information matrix $U^\ell_c = (P_{c|c-1}^\ell)^{-1}$
+and the information vector $u^\ell_c~=~\hat \theta^\ell_{c|c-1} (P^\ell_{c|c-1})^{-1}$.
+The travel time estimate of vehicle $m$ along segment $\ell$,
+along with its uncertainty,
+as defined in (\ref{eq:tt_obs_mean}) and (\ref{eq:tt_obs_err}), respectively,
+are transformed to a measurement information covariance matrix
+$B_{c\ell}^{m}~=~(R_{c\ell}^m)^{-2}$
+and measurement information vector $b_{c\ell}^{m}~=~Z_{c\ell}^{m} (R_{c\ell}^{m})^{-2}$.
+The total information is the sum of the information over all $M_{c\ell}$ vehicles
+that traversed segment $\ell$ in the time interval $(t_{c-1}, t_c]$,
+so the update equations are simply
+\begin{align*}
+U^\ell_{c|c} &= U^\ell_{c|c-1} + \sum_{m=1}^{M_{c\ell}} B_{c\ell}^{m}, \\
+\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^{M_{c\ell}} b_{c\ell}^{m}.
+\end{align*}
+The desired parameter estimates are obtained
+by the inverse transformations
+\begin{equation*}
+\hat \theta^\ell_{c|c} = \frac{\hat u^\ell_{c|c}}{U^\ell_{c|c}}
+\quad\text{and}\quad
+P^\ell_{c|c} = \frac{1}{U^\ell_{c|c}},
+\end{equation*}
+which can be used in the prediction of travel times
+for upcoming buses.
+Note that, if we included segment correlations,
+the full state $\boldsymbol{\theta}_c$ would need updating in a single step,
+which would require inverting a large $L\times L$ matrix,
+significantly increasing the time of updating the state.
+
+\section{Real-time implementation}
+\label{sec:rt}
+
+The application consists of two components:
+the first handles importing static GTFS data and constructing the network,
+while the second implements the \rt models and prediction.
+We used \textsf{Rcpp} \citep{Rcpp}
+to develop the program,
+which provides access to \textsf{R} \citep{rcore},
+which contains many useful tools for data manipulation and package development,
+as well as the speed and memory management capabilities of \textsf{C++}.
+The program is implemented in the \textsf{R} package
+\verb+transitr+, available on Github (\url{https://github.com/tmelliott/transitr}).
+In this section, we discuss the features of the \rt component
+and assess its performance
+with respect to timing and travel time estimation.
+
+Below is the general structure of the \rt component,
+where the bold steps are those discussed in this paper:
+\begin{enumerate}
+\item Load GTFS data from database
+\item Each time new data are received, do:
+\begin{enumerate}
+    \item Update or create new vehicle objects from the new data
+    \item \textbf{Run particle filter on each vehicle to update or initialise its state}
+    \item \textbf{Update state of any roads for which vehicles
+        have completed travel}
+    \item Generate ETAs for vehicles using combination of vehicle and network state
+    \item Write ETAs to a file for distribution
+\end{enumerate}
+\end{enumerate}
+
+During the development of the application,
+our primary concern was to ensure each component of the program
+was as efficient as possible,
+so ETA generation and distribution is fast enough to be feasible in real-time,
+with a target of 30~seconds or faster at peak time.
+Using \textsf{C++} provided the memory management control necessary to make this possible.
+Since we modelled each vehicle independently,
+it was straightforward to spread the vehicle updates
+over multiple cores using \textsf{OpenMP} without thread-safety concerns.
+
+The number of particles needed per vehicle
+depends on many factors,
+so it was necessary to explore the performance of the application
+with a varying number of particles.
+We assessed the application's performance
+using a range of values for
+system noise and GPS error.
+To enable comparisons, we implemented a simulated \rt environment
+in which the same subset of real vehicle data from 8~October 2018
+could be analysed using a range of settings.
+These simulations were carried out on a virtual machine
+with 8~Intel Xeon 3.00GHz CPU cores and 32~GB of memory,
+running \textsf{Ubuntu}~16.04 and using \textsf{R}~3.4.1.
+The simulations were processed locally using \textsf{R}~3.6.0,
+and made use of the \textsf{R} packages \verb+dplyr+ \citep{dplyr}
+for manipulating and summarising the results,
+and \verb+ggplot2+ \citep{ggplot2} to graph them.
+
+\subsection{Program Timings}
+\label{sec:timings}
+
+Within each iteration, the program recorded timings of its various components.
+Since the number of vehicles travelling at any given time changes throughout the day,
+we used the average timings over an off-peak 15-minute window.
+Figure~\ref{fig:timings} shows the average timings for
+varying numbers of particles, $N$.
+
+The most time-consuming component was ETA writing,
+which involved summarising the individual ETAs estimated for each particle.
+While not discussed here this involved computing quantiles which uses a sorting algorithm
+with complexity $\mathcal{O}(N \log N)$,
+which explains the non-linear relationship with $N$.
+
+The next most intensive step was vehicle updating,
+which took about 5~seconds for 8000 particles.
+The ETA prediction step only took a few seconds,
+although this may increase once a more comprehensive model has been developed.
+On our 8-core virtual machine,
+the average time to process one iteration during off-peak
+was about 16 seconds when using 8000 particles.
+
+\begin{figure}[tb]
+    \centering
+    \includegraphics[width=0.8\textwidth]{figures/04_model_results_timing.pdf}
+    \caption{
+        The average real-time running time of each step within each iteration
+        over a 15-minute off-peak window.
+        The vehicle update and writing ETA steps involve sorting the particles,
+        which is an operation of $\mathcal{O}(N\log N)$ complexity.
+        Total time is for the entire iteration, which includes fetching the data from the API.
+    }
+    \label{fig:timings}
+\end{figure}
+
+\subsection{Model performance}
+\label{sec:model_perf}
+
+Evaluating the performance of the model involved
+running the simulation using a range of values of GPS error, $\omega^2$,
+and system noise, $\sigma^2$.
+Figure \ref{fig:dist_to_route} shows the distribution of the distance
+between each observation and the route using a nearest point algorithm,
+which suggested using values $\omega \in \{1,2,3,5\}$ for GPS error in meters.
+For the system noise, we used values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$,
+which correspond to an average vehicle speed variation of between 0.1 and 45~meters per second
+over 30~seconds.
+For each combination of these parameter values, including varying $N$,
+the following values were calculated:
+effective sample size, $N_\text{eff}$;
+degeneration rate, the percentage of iterations in which the vehicle was lost
+(no plausible particles remained);
+and the (relative) variance of segment travel time estimates.
+
+\begin{figure}[tb]
+    \centering
+    \includegraphics[width=0.7\textwidth]{figures/04_model_results_dist.pdf}
+    \caption{
+        The distribution of the minimum distance between observed vehicle location
+        and the route path, truncated to 20~meters.
+    }
+    \label{fig:dist_to_route}
+\end{figure}
+
+\begin{figure}[tb]
+    \centering
+    \includegraphics[width=\textwidth]{figures/04_model_results_neff.pdf}
+    \caption{
+        The effective sample size, $N_\text{eff}$,
+        and degeneration rate for varying values of GPS error,
+        system noise, and the number of particles $N$.
+    }
+    \label{fig:perf_stats}
+\end{figure}
+
+Figure~\ref{fig:perf_stats} shows that $N_\text{eff}$ increases
+with GPS error and decreases with system noise
+while remaining unaffected by changing $N$.
+The degeneration rate decreases with GPS error and increases with $N$,
+but is mostly unaffected by system noise.
+Large GPS error affects the likelihood,
+giving more weight to particles farther from the observation
+than does a smaller GPS error.
+Conversely, increasing system noise spreads out the particle cloud,
+so fewer particles will be near the vehicle,
+decreasing $N_\text{eff}$,
+but more plausible trajectories are sampled,
+so we see a slight decrease in degeneration rate.
+Thus, these results are not surprising
+and show that the model is working as expected.
+
+The last measure of performance is the relative variance of segment travel times.
+We used relative variance as each segment has a very different travel time distribution,
+so to compare the variability between simulations we needed to
+compute the average variance of estimated travel times over all simulations,
+taking the average (over road segments) of the ratio between variance for
+each simulation and variance over all simulations.
+If $\tilde \bz_\ell^{\omega,\sigma}$ is a vector of all travel times
+along segment $\ell$ during the simulation with GPS error $\omega$ and system noise $\sigma$,
+then the relative variance $\bar v^{\omega,\sigma}$ is calculated by
+\begin{equation*}
+\begin{split}
+v_\ell^{\omega,\sigma} &=
+\frac{\mathrm{Var}(\tilde \bz_\ell^{\omega,\sigma})}{\mathrm{Var}(\tilde \bz_\ell)}, \\
+\bar v^{\omega,\sigma} &=
+    \frac{1}{L}\sum_{\ell=1}^L v_\ell^{\omega,\sigma}.
+\end{split}
+\end{equation*}
+Figure~\ref{fig:travel_times} shows that simulations with larger GPS error
+resulted in a higher relative variance,
+while simulations with more particles had a smaller relative variance.
+There was no clear pattern to the effect of system noise.
+
+The results of the simulations
+demonstrate a trade-off between particle filter performance
+($N_\text{eff}$ and degeneration rate) and parameter estimation.
+However, until the arrival time prediction model has been completed,
+it is difficult to make decisions about the optimal values to use:
+we are unable to say whether a higher variation of travel time estimates
+is going to have a significant effect on the final ETAs,
+particularly when compared to the uncertainty of forecasting road state.
+
+\begin{figure}[tb]
+    \centering
+    \includegraphics[width=\textwidth]{figures/04_model_results_times.pdf}
+    \caption{
+        The relative variability of travel time estimates for varying
+        values of GPS error, system noise, and the number of particles, $N$,
+        with error bars $\pm 2$ standard errors.
+        % as GPS error increases,
+        % with no clear relationship with system noise.
+        % Increasing $N$ decreases variability.
+    }
+    \label{fig:travel_times}
+\end{figure}
+
+\section{Discussion and future work}
+\label{sec:discussion}
+
+In this paper, we described an approach to transit vehicle modelling
+that enables the \rt estimation of road state from vehicle position data,
+which will later be used to make arrival time predictions
+that account for \rt traffic conditions.
+The particle filter allows the modelling of complex bus behaviours,
+providing estimates of vehicle travel times along roads
+which are in turn used to estimate the road network state.
+
+We have focussed on making a general framework
+into which new models can be incorporated,
+such as implementing intersection-specific behaviour
+once intersection locations are available,
+or exiting models changed.
+For example, we plan to model stop dwell time using historical data,
+which will allow us to replace the exponential distribution
+with something more appropriate.
+
+The network model presented here is simple,
+but provides the foundation for future work to
+develop a formal model combining \rt travel time information with historical data,
+allowing short- and medium-term forecasts of road state,
+for example incorporating peak time congestion.
+As for the arrival times,
+we plan to estimate these from a combination of historical
+and \rt data.
+The model presented in this paper provides the starting point
+for a transit modelling and predictive framework
+that can be adapted to specific transit providers,
+combined with historical data,
+and run in \rt to provide commuters with
+improved ETAs with which they can more reliably
+plan their commutes.
+
+The results of our simulations show that the framework is a
+feasible \rt alternative.
+While the timings presented were for off-peak times,
+further investigation into the choice of $N$ and $N_\text{eff}$,
+as well as work removing redundancy from the transition algorithm,
+should allow us to model all vehicles during peak periods---%
+when there are 2--3~times as many vehicles---%
+within our 30-second time frame target.
+
 \bibliographystyle{anzsj}

### 2020-07-27 new2020/main.Rnw
Framework for building paper (with rmarkdown)

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}

\documentclass{article}

\title{Doing things with particle filters}
\author{Tom Elliott and Thomas Lumley}

\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Section One},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\usepackage{amsmath}
\usepackage{cleveref}

\begin{document}

\maketitle

%% Content goes here

\input{sections/sec01.tex}

\bibliographystyle{acm}
\bibliography{allrefs}

\end{document}

### 2020-07-27 new2020/sections/sec01.Rmd
Framework for building paper (with rmarkdown)

---
title: Section One
header-includes:
   - \usepackage{cleveref}
---

# Introduction

```{=latex}
\begin{equation}
\label{eq:no}
y = 2x
\end{equation}
```

Predicting bus arrival is ... quite something \( \cite{Elliott_2020} \)

I can just type things here, like ... yeah!
```{=latex}
\begin{equation}
\label{eq:hello}
x = \int 2x\; \mathrm{d}x
\end{equation}
```

We see in `\cref{eq:hello}`{=latex} that ... blah.

### 2020-07-27 new2020/clean_tex.R

@@ -1,0 +1,9 @@
+ca <- commandArgs(trailingOnly = TRUE)
+
+f <- ca[1]
+
+txt <- readLines(f)
+from <- grep("\\maketitle", txt, fixed = TRUE) + 1L
+to <- grep("\\end{document}", txt, fixed = TRUE) - 1L
+txt <- txt[from:to]
+writeLines(txt, f)

### 2020-07-27 new2020/sections/sec02.Rmd
Get subfiles working

---
title: Section Two
header-includes:
   - \usepackage{cleveref}
---

# Background

## A transit road network

* information about GTFS
* the concept of converting it into a network (as per `\cite{Elliott_2020}`{=latex})
* end product is real-time estimates of traffic state

## Estimating network and vehicle state using particle filtering

* particle filter on vehicle state
* used to estimate vehicle speed, position
* obtain distribution of travel times for each vehicle along each road

### 2020-07-27 new2020/sections/sec03.Rmd
Get subfiles working

---
title: Section Three
header-includes:
   - \usepackage{cleveref}
---

# Predicting arrival time

## Particle filter etas

* use particle filter and network state to obtain ETA estiamtes for each particle/stop
* include dwell time uncertainties, etc
* result is a distribution estimated by $N$ particles

## Simplified ETA CDF

* round to minutes
* compute the CDF by definition ``number of particles arriving within $x$ minutes''

### 2020-07-27 new2020/sections/sec04.Rmd
Get subfiles working

---
title: Section Four
header-includes:
   - \usepackage{cleveref}
---

# Journey planning applications

* CDF makes it possible to answer many (often complex) jouney planning questions
* P(catch)
* P(arrive on time)
* P(transfer)
* this is a simple computation - can be done client side (i.e., on a user's phone) by passing CDF (small size, as e.g., JSON)

### 2020-07-28 new2020/sections/sec01.Rmd
Better build system

@@
 
-```{=latex}
-\begin{equation}
-\label{eq:no}
-y = 2x
-\end{equation}
-```
-
-Predicting bus arrival is ... quite something \( \cite{Elliott_2020} \)
-
-I can just type things here, like ... yeah!
-```{=latex}
-\begin{equation}
-\label{eq:hello}
-x = \int 2x\; \mathrm{d}x
-\end{equation}
-```
-
-We see in `\cref{eq:hello}`{=latex} that ... blah.
+* the problem of predicting bus arrival - it needs to use real-time traffic information @cite:cn;
+* however, many deployed methods are either specific to a provider/city, or don't make use of real-time data (only vehicle position and/or arrival delays, e.g., in Auckland)
+* since the only logical source of ``traffic data'' in this setting is the transit vehicles themselves, makes sense to develop framework that uses them to estimate real-time network state @cite:Elliott_2020;
+* in this example, particle filter is used to obtain a sample of points from the arrival time distribution - many many points, which cannot possibly be distributed or stored efficiently
+* we propose a method of reducing this to a simple discrete CDF of arrival time (in minutes)

### 2020-07-28 new2020/clean_tex.R

@@ -6,4 +6,20 @@ txt <- readLines(f)
 from <- grep("\\maketitle", txt, fixed = TRUE) + 1L
 to <- grep("\\end{document}", txt, fixed = TRUE) - 1L
 txt <- txt[from:to]
+
+# convert @cite:Paper_2000,Paper2_2001; to
+# `\cite{Paper_2000, Paper2_2001}`{=latex}
+
+regx <- "@([a-z]+):([a-zA-Z0-9,_\\\\]+);"
+m <- do.call(rbind, stringr::str_match_all(txt, regx))
+if (nrow(m)) {
+    for (i in seq_len(nrow(m))) {
+        rep <- paste0("\\", m[i, 2], "{", m[i, 3], "}")
+        rep <- gsub("[\\]+_", "_", rep)
+        cat("\n ", m[i, 2], "->", rep)
+        txt <- stringr::str_replace_all(txt, stringr::fixed(m[i, 1]), rep)
+    }
+    cat("")
+}
+
 writeLines(txt, f)

### 2020-09-28 new2020/main.Rnw
Get subfiles working
Use statsoc theme style

@@
 % Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
+% \PassOptionsToPackage{unicode}{hyperref}
+% \PassOptionsToPackage{hyphens}{url}
 
-\documentclass{article}
+\documentclass{statsoc}
 
-\title{Doing things with particle filters}
-\author{Tom Elliott and Thomas Lumley}
+\title[Particle buses]{Doing things with particle filters}
+
+\author{Tom Elliott}
+\address{University of Auckland,
+         Auckland,
+         New Zealand}
+\email{tom.elliott@auckland.ac.nz}
+
+\author[Elliott, T \& Lumley, T]{Thomas Lumley}
+\address{University of Auckland,
+         Auckland,
+         New Zealand}
+\email{t.lumley@auckland.ac.nz}
 
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+% \setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 
@@
 
-%% Content goes here
+\begin{abstract}
+The prediction of transit vehicle arrival time is a complex problem with the added difficulty of its real-time nature.
+Reliable predictions require the combination of both vehicle and network states, implying that real-time traffic congestion information is incorporated into the predictions.
+Inevitably, however, there is a lot of uncertainty around arrival times: road speeds fluctuate, particularly around peak time, and multimodality is added with every stop passed (which may or may not cause the bus to stop depending on passenger demand).
+Therefore, not only do we need to predict arrival time, but also assess and report the uncertainties involved.
+We present a method of using a particle filter to combine the states of both vehicle and network to obtain arrival time distributions.
+These are then discretised in such a way that allows real-time calculation of event probabilities for use in journey planning applications.
+\end{abstract}
+\keywords{particle filter, transport, real-time, GTFS}
 
-\input{sections/sec01.tex}
+Hello \dots
 
-\bibliographystyle{acm}
+<<include-sections,echo=FALSE,results="asis">>=
+for (file in list.files('sections', pattern = "*.tex", full.names = TRUE)) {
+  cat(sprintf("\\input{%s}", file), "\n")
+}
+@
+
+\bibliographystyle{chicago}
 \bibliography{allrefs}
@@
 \end{document}
+
+% royal statistical society - journal a
+% computational statistics & data analysis
+% transportation research part c
+
+%% inzight paper:
+% what for, and then how
+
+% american statistician (education)
+% JSS (design/modular etc)
+
+% NZSA conference

### 2020-09-29 new2020/main.Rnw
Use rss.bst
Basic outline thing
begin adding structure

@@
 
-\title[Particle buses]{Doing things with particle filters}
+\title[A real-time particle filter for bus arrival prediction]{A real-time particle filter for bus arrival prediction with discrete CDFs for journey planning}
 
@@
 
-\usepackage{lmodern}
+% \usepackage{lmodern}
 \usepackage{amssymb,amsmath}
@@
 
+\input{symbols.tex}
+
 \begin{document}
@@
 \end{abstract}
-\keywords{particle filter, transport, real-time, GTFS}
-
-Hello \dots
+\keywords{particle filter, transport, real-time, GTFS, transit networks, journey planning}
 
@@
 
-\bibliographystyle{chicago}
+\bibliographystyle{rss}
 \bibliography{allrefs}

### 2020-09-29 new2020/sections/sec01.Rmd
Use statsoc theme style
begin adding structure

--- 2020-07-27 new2020/sections/sec01.Rmd
@@
 
-```{=latex}
-\begin{equation}
-\label{eq:no}
-y = 2x
-\end{equation}
-```
+* the problem of predicting bus arrival - it needs to use real-time traffic information @cite:cn;
+* however, many deployed methods are either specific to a provider/city, or don't make use of real-time data (only vehicle position and/or arrival delays, e.g., in Auckland)
+* since the only logical source of "traffic data" in this setting is the transit vehicles themselves, makes sense to develop framework that uses them to estimate real-time network state @cite:Elliott_2020;
+* in this example, particle filter is used to obtain a sample of points from the arrival time distribution - many many points, which cannot possibly be distributed or stored efficiently
+* we propose a method of reducing this to a simple discrete CDF of arrival time (in minutes)
+* we show that these results can be used to answer some common journey planning questions, in real-time, which outperform current "method"
 
-Predicting bus arrival is ... quite something \( \cite{Elliott_2020} \)
+Some of the references include:
 
-I can just type things here, like ... yeah!
-```{=latex}
-\begin{equation}
-\label{eq:hello}
-x = \int 2x\; \mathrm{d}x
-\end{equation}
-```
-
-We see in `\cref{eq:hello}`{=latex} that ... blah.
+* @cite:Cathey_2003;
+* @cite:Yu_2006,Yu_2010,Yu_2011;
+* @cite:Hans_2015;

### 2020-09-29 new2020/sections/sec02.Rmd
Better build system
begin adding structure

@@
 
+Before describing the process of obtaining arrival time distributions, we must first define the framework with which we obtain vehicle and network state estimates.
+@cite:Elliott_2020; present a process for constructing a transit road network from raw GTFS data.
+
 ## A transit road network
@@
 * information about GTFS
-* the concept of converting it into a network (as per `\cite{Elliott_2020}`{=latex})
+* the concept of converting it into a network (as per @cite:Elliott_2020;)
 * end product is real-time estimates of traffic state

### 2020-09-29 new2020/sections/sec03.Rmd
begin adding structure

@@
    - \usepackage{cleveref}
+   - \input{../symbols.tex}
 ---
@@
 
-## Particle filter etas
+Accurate, reliable prediction of bus arrival time requires knowledge of both vehicle and network states, represented by $\Vstate_{k}$ and $\NWstate_{c}$, respectively.
+Since there is a high degree in the uncertainty of these, particularly when forecasting future network states around peak times, it is crucial to incorporate this uncertainty into the arrival time distribution and, ultimately, decision making processes.
 
-* use particle filter and network state to obtain ETA estiamtes for each particle/stop
-* include dwell time uncertainties, etc
-* result is a distribution estimated by $N$ particles
+The particle filter presents a robust method of sampling all of the possible trajectories the vehicle might take @citep:Hans_2015;.
 
-## Simplified ETA CDF
+* initial state incoporates uncertainty (shape, multimodality) of vehicle state
+* trajectory of each particle = one possible path the bus might take
+* accounts for uncertainty in (forecasted) road speed, correlations, etc
+* result makes no assumptions about shape of distribution
 
-* round to minutes
-* compute the CDF by definition ``number of particles arriving within $x$ minutes''
+The main downside of the particle filter is the computational demand of it, often requiring 5000-10000 particles per bus.
+
+* @cite:Elliott_2020; showed the pf is feasible in real-time for modelling
+* need to also show predicions can be done quickly and usefully
+* reduce number of particles where possible
+
+Our implementation is described in @cref:pf_etas;.
+To examine the effectiveness of our approach, we run our method on a full day and compared the predictions with the actual arrivals, as well as the currently used "GTFS" predictions.
+These are discussed in @cref:pf_results;.
+
+## Particle filter ETAs
+@label:pf_etas;
+
+In our application, the vehicle state is already represented by a set of $\Np$\ particles,
+\[
+p(\Vstate_{k|k}\cond{}\Vobs_k) \approx \sum_{i=1}^\Np \Pwt_k \DiracMeasure{\Vstate_k\vi}{\Vstate_k}
+\]
+which we use directly.
+However, were vehicle states available in some other form, one would simply take a sample from the posterior state estimate $p(\Vstate_{k|k}\cond{}\Vobs_k)$.
+This gives us a sample of plausible bus states for which we can predict individual arrival times at upcoming stops.
+
+As with the vehicle model described by @citet:Elliott_2020;, we can iteratively forecast each particle's arrival at all upcoming stops.
+This involves incorporating network state (vehicle speeds along roads) as well as bus stopping behaviour.
+These two aspects are described individually below, and each particle simply iterates between them until it reaches the end of the route.
+
+[introduce trip state]
+
+### Road segment travel times
+
+[mostly copy 5-step process from thesis section 5.2.1]
+
+Each particle begins having travelled $\Vdist_k\vi$\ meters along the route, with a current speed of $\Vspeed_k\vi$.
+The first travel time required is, therefore, the time to reach the end of the current road segment, $\TripSeg\vi$, assuming the bus maintains its current speed.
+From the transit network construction, we know that segment $\TripSeg\vi$ starts $\Tsegd_{\TripSeg\vi}$\ meters along the route and has a length of $\Tseglen_{\TripSeg\vi}$\ meters, so the distance remaining is
+\[
+   \bar z\vi = \Tsegd_{\TripSeg\vi} + \Tseglen_{\TripSeg\vi} - \Vdist_k\vi
+\]
+which will take the particle
+\[
+   z\vi = \frac{\bar z\vi}{\Vspeed_k\vi}\ \text{seconds}
+\]
+to complete.
+
+[ once trip state introduced, the above becomes:]
+
+...
+particle has completed 100$p_\ell\vi$% of the segment.
+If the vehicle is nearing the end of the segment (less than 200\ meters remaining), we keep the particle's initial speed $\Vspeed_k\vi$.
+Otherwise, we simulate a speed from the network state with mean $\NWstate_\ell$, uncertainty $\NWstatevar_\ell$, and between-vehicle variability $\NWvar_\ell$, as follows:
+\begin{equation*}
+   \Vspeed_\ell{\vi}^\star =
+   \begin{cases}
+      \Vspeed_k\vi & (1-\SegProg_k)\Tseglen_\ell < 200, \\
+      v_\ell{\vi} \sim \TNormal{\NWstate_\ell}{\NWstatevar_\ell^2 + \NWvar_\ell^2}{0}{\MaxSpeed_\ell}, & \text{otherwise}.
+   \end{cases}
+\end{equation*}
+
+Now the travel time to the end of the current segment can simply be obtained as
+\[
+   \tilde z\vi = \frac{(1-\SegProg_k\vi)\Tseglen_\ell}{\Vspeed_\ell{\vi}^\star}.
+\]
+We now store an iterative variable *travel-time-so-far* with the time taken to reach the end of the current segment, $\ttsofar\vi = \tilde z\vi$.
+
+### Bus stops
+
+### Arrival time distribution
+
+## Results
+@label:pf_results;

### 2020-09-29 new2020/clean_tex.R

@@ -16,10 +16,10 @@ if (nrow(m)) {
     for (i in seq_len(nrow(m))) {
         rep <- paste0("\\", m[i, 2], "{", m[i, 3], "}")
         rep <- gsub("[\\]+_", "_", rep)
-        cat("\n ", m[i, 2], "->", rep)
+        # cat("\n ", m[i, 1], "->", rep)
         txt <- stringr::str_replace_all(txt, stringr::fixed(m[i, 1]), rep)
     }
-    cat("")
+    # cat("")
 }
 
 writeLines(txt, f)

### 2020-09-30 new2020/sections/sec02.Rmd
Putting down words / thoughts ... most likely unnecessary

@@
 
+**most of this section will be removed/cut-down to just the basic concepts**
+
+## GTFS static and real-time
+@label:gtfs;
+
+The structure of public transport systems is very similar across the globe, so much so that Google developed GTFS @citep:GoogleDevelopers_2006; to standardise the way transport agencies organise and distribute transit data.
+GTFS consists of two core components: *static*, which contains pre-defined information, such as routes and stops, and *real-time*, which contains vehicle location and other data collected in real-time.
+
+Static GTFS contains *structural* data, descibing information one might observe on a printed timetable.
+This includes route information about trips (instances of a route at a specific time), the geographical path the vehicle takes, and the stops it services.
+This provides the basis of transit data, and plays a key role in developing the aforementioned *transit network*.
+
+Real-time GTFS descibes how live data, such as the location of a bus, or its time of arrival at or departure from a bus stop, is formatted.
+Vehicle locations allow users to track the progress of the vehicle along its route, and are often updated every 10--30\ seconds.
+In some implementations, such as in Auckland, they are also triggered by events such as arrival at a bus stop or (major) intersection.
+Arrival and departure events, on the other hand, are triggered whenever the bus arrives at or departs from a bus stop (or at least thinks it does), and include the time of the event and, if scheduled arrival times are available, the *delay* between scheduled and actual arrival (or departure).
+
+A huge benefit of GTFS being adopted widely around the world is that developers can easily access transit feeds for different providers in different cities and countries using the same structure.
+Any application developed using (only) GTFS information can, technically, be redeployed to any other system also using GTFS.
+For this reason, we opted to develop our application framework based solely on GTFS data, as described next.
+
 ## A transit road network
+@label:network;
 
-* information about GTFS
-* the concept of converting it into a network (as per @cite:Elliott_2020;)
-* end product is real-time estimates of traffic state
+One important predictor of bus arrival is travel time @citep:Shalaby_2004;.
+However, previous research has focused on situations where either only a single route is considered, or else different routes and combined by manually locating common road segments.
+This is, of course, not easy or even possible to automate.
+@cite:Elliott_2020; proposed a method of developing a transit road network from only GTFS data by identifying *nodes* (bus stops and intersections) and the *edges* (roads) between them.
 
-## Estimating network and vehicle state using particle filtering
+Here, we use the simplest version of this method, which places nodes at bus stops.
+Thus, any two routes that share the same sub-sequence of bus stops are assumed to follow the same path between them (this is almost always the case).
+By expressing the route as a series of nodes and road segments, we can begin to share information across routes.
 
-* particle filter on vehicle state
-* used to estimate vehicle speed, position
-* obtain distribution of travel times for each vehicle along each road
+The goal is to model vehicles as they travel along their respective routes, estimating their average speed along each road segment.
+Having traversed a road segment, the data is passed to the network to update the real-time traffic state.
+This can in turn be used, optionally with a forecasting function, to predict a future travel time of a bus along the road, which is used in arrival time prediction.
+
+## Particle filters estimating vehicle state
+@label:pf;
+
+Estimation of vehicle state in real-time commonly uses *recursive Bayesian estimation*, in which the posterior becomes the prior for the next iteration.
+Particle filters are a powerful method of estimating successive states, as they make few assumptions and, most importantly, cover a large range of possible trajectories, even when there is a high degree of multimodality as is the case with transit data (e.g., around bus stops).
+Here, I give a brief overview of the particle filter which was used by @citet:Elliott_2020; to estimate vehicle state.
+
+In a recurisive Bayesian model, the underlying state we are interested in is denoted $\vec{x} = \vec x_{0:k} = \left[\vec x_0, \vec x_1, \ldots, \vec x_k\right]$.
+It can rarely be observed directly, and instead we obtain measurements $\vec y = \vec y_{1:k} = \left[\vec y_1, \ldots, \vec y_k\right]$.
+In a particle filter @citep:Gordon_1993;, this state is represented by a weighted sample of *points*,
+and makes use of the delta measure $\delta$,
+\[
+   p(\vec x_{k-1} | \vec y_{0:k-1}) \approx \sum_{i=1}^N w_{k-1}^{(i)} \DiracMeasure{\vec x_{k-1}^{(i)}}{\vec x_{k-1}}.
+\]
+@citet:Elliott_2020; describe the particle filter in detail, with discussion of its use for modelling transit vehicles in real-time.
+However, here we only discuss the merits of the particle filter, and need no more details than given already.
+
+The main advantage of the particle filter is, as mentioned, its ability to cope with a wide range of plausible, discontinuous trajectories.
+This often happens in transit data as the bus passes bus stops at which it may or may not stop.
+It's main downside is the computational demand, since we need 1000s of particles in memory for each vehicle, and perform transitions and updates to each of them.
+However, @citet:Elliott_2020; have shown that, so long as it is programmed efficiently, the particle filter can run fast enough to be feasible in a real-time setting, at least for updating vehicle states.
+
+In the next section, we make use of the flexible nature of the particle filter in that each particle is transitioned independently, making it possible to account for a wide variety of complex behaviours.
+These most notably include bus stops, which introduce a lot of uncertainty, but also layovers, which can result in a wide temporal gap between possible arrival times.
+
+The particle filter allows us to estimate the state of a vehicle and, most importantly, it's average speed along road segments:
+\[
+   p(b_\ell | \vec y_{1:k}) \approx
+   \sum_{i=1}^N w_{k}^{(i)} \DiracMeasure{b_{\ell}^{(i)}}{b_{\ell}}.
+\]
+
+## Estimating network state
+@label:state_estimates;
+
+Having obtained a estimate of the speed of each vehicle along a road, we can proceed to updating the state of the road network itself.
+This uses a Kalman filter to update the state, $\vec \beta_c$, using the estimated vehicle speeds.
+
+This all provides a flexible method for modelling real-time traffic conditions in any network using only GTFS, and not rely on other data sources (for example taxis or video cameras).
+
+We use a hierarchical model on vehicle travel times, as individual vehicles have different average speeds, even if they travel at the same time, due to different driver behavior and other factors.
+The between-vehicle noise parameter $\NWvar_\ell$ describes this, and is estimated from historical data.
+
+The observations are also uncertain, as they are estimated from the particle filter.
+We can assume between-vehicle and measurement uncertainties are independent, so the total variance of observations is
+\[ \Var{b_\ell | \vec y_{1:k}} = \NWerr_\ell^2 + \NWvar_\ell^2. \]
+The system noise $\NWnoise$ is also estimated from historical data.
+Together we can quickly update the network state using the information variance of the Kalman filter (since this allows data from multiple sources simultaneously, which can happen when two or more buses are travelling one behind the other).
+
+The result is an estimate of the state of the network at time $t_c$,
+\[
+   p(\NWstate_c | \cup_m \cup_\ell\ \NWobs_{\ell,c,m}) \sim
+   \Normal{\vec\NWstate_\ell}{\NWstatevar^2}.
+\]
+This can be used to predict the travel time of buses.
+
+If a forecast function is available, we could also forecast future travel time based on current state and historical trends, and this would be incorporated.

### 2020-09-30 new2020/clean_tex.R

@@ -10,7 +10,7 @@ txt <- txt[from:to]
 # convert @cite:Paper_2000,Paper2_2001; to
 # `\cite{Paper_2000, Paper2_2001}`{=latex}
 
-regx <- "@([a-z]+):([a-zA-Z0-9,_\\\\]+);"
+regx <- "@([a-zA-Z]+):([a-zA-Z0-9,_\\\\]+);"
 m <- do.call(rbind, stringr::str_match_all(txt, regx))
 if (nrow(m)) {
     for (i in seq_len(nrow(m))) {

### 2020-10-01 new2020/sections/sec03.Rmd
more (mostly silly) words, but moving towards the right direction

@@
 
-[introduce trip state]
+To begin, we define the state of an active trip at time $\Tript_k$ in which the vehicle is at or last visited stop $\TripStop_k$, together with an indicator $\TripDep_k$ of whether or not the bus has departed the stop, the index of the current road segment $\TripSeg_k$, and the progress along that segment $\SegProg_k$:
+\[
+   \Tripr_k = \tvec{\Tript_k, \TripStop_k, \TripDep_k, \TripSeg_k, \SegProg_k}.
+\]
 
-### Road segment travel times
+Next, we approximate the distribution of trip state by generating a sample of particles from $p(\Tripr_k | \Vobs_{1:k})$.
+In our case, we already have a weighted sample of particles, but the sample could be taken from any posterior distribution, for example if vehicle state is estimated using a Kalman filter the particles can be sampled from the Gaussian state estimate.
 
-[mostly copy 5-step process from thesis section 5.2.1]
+We wish to take a sample of size $\tilde N \leq N$, which can depend on the number of remaining stops and the level of accuracy desired.
+However, in the next section we approximate the predictive distributions using a discrete CDF, so accuracy only needs to be at about the 30\ second level.
+Thus, we pull sample with replacement from $p(\Vstate_k|\Vobs_{1:k})$, yielding the trip state estimate
+\[
+   p(\Tripr_k | \Vobs_{1:k}) \approx
+   \frac{1}{\tilde N} \sum_{i=1}^{\tilde N} \DiracMeasure{\Tripr_k\vi}{\Tripr_k}
+\]
+where the components of $\Tripr_k\vi$ are computed from the vehicle state particles. This doesn't make sense. Gosh darnit.
 
-Each particle begins having travelled $\Vdist_k\vi$\ meters along the route, with a current speed of $\Vspeed_k\vi$.
-The first travel time required is, therefore, the time to reach the end of the current road segment, $\TripSeg\vi$, assuming the bus maintains its current speed.
-From the transit network construction, we know that segment $\TripSeg\vi$ starts $\Tsegd_{\TripSeg\vi}$\ meters along the route and has a length of $\Tseglen_{\TripSeg\vi}$\ meters, so the distance remaining is
-\[
-   \bar z\vi = \Tsegd_{\TripSeg\vi} + \Tseglen_{\TripSeg\vi} - \Vdist_k\vi
-\]
-which will take the particle
-\[
-   z\vi = \frac{\bar z\vi}{\Vspeed_k\vi}\ \text{seconds}
-\]
-to complete.
+Assuming I can rewrite the above to make sense ...
 
-[ once trip state introduced, the above becomes:]
+Now we iterate each trip particle to the end of the route, storing the arrival times at all stops along the way. This is a five-step process.
 
-...
-particle has completed 100$p_\ell\vi$% of the segment.
+Note: need to add current speed to trip state ???
+
+**Step 1: complete the current segment**
+
+Each particle has completed 100$p_\ell\vi$% of the segment.
 If the vehicle is nearing the end of the segment (less than 200\ meters remaining), we keep the particle's initial speed $\Vspeed_k\vi$.
@@
 \begin{equation*}
-   \Vspeed_\ell{\vi}^\star =
+   v_\ell{\vi}^\star =
    \begin{cases}
       \Vspeed_k\vi & (1-\SegProg_k)\Tseglen_\ell < 200, \\
-      v_\ell{\vi} \sim \TNormal{\NWstate_\ell}{\NWstatevar_\ell^2 + \NWvar_\ell^2}{0}{\MaxSpeed_\ell}, & \text{otherwise}.
+      v_\ell{\vi} \sim \TNormal{\NWstate_\ell}{\RouteNWstatevarseg_\ell^2 + \NWvar_\ell^2}{0}{\MaxSpeed_\ell}, & \text{otherwise}.
    \end{cases}
@@
 
-### Bus stops
+**Step 2: compute arrival time at next stop**
 
-### Arrival time distribution
+If there are any remaining segments between the current segment $\TripSeg_k\vi$ and the next stop $j_k\vi + 1$, their travel times are sampled from the network state with variance increased by network system noise to account for forecasting changes.
+\[
+   v_j\vi \sim
+   \TNormal{\hat \NWstate_j}{\left[(\RouteNWstatevarseg_j + \ttsofar\vi\NWnoise)^2 + \NWvar_j^2\right]\wedge P_{\text{max}}}{0}{\MaxSpeed_j},
+\]
+allowing uncertainty to increase up to a maximum $P_\text{max}$ which is the overall variance of all speeds along all segments estimated from historical data.
+The travel time along each segment is added to the travel-time-so-far,
+\[
+   \ttsofar\vi = \ttsofar\vi + \frac{\Tseglen_j}{v_j\vi}.
+\]
+
+Now, arrival time at stop $j$ is the sum of travel time along all segments up to the next stop. In the iterative process, this is simply $\Tarr_j\vi = t_k + \ttsofar\vi$.
+
+**Step 3: compute stop dwell time**
+
+Once at a stop, a bus either stops and waits while passengers board and disembark, or travels past without stopping.
+This *dwell time* needs to be modelled @citep:cn;, but adds multimodality to the arrival time distribution.
+
+The dwell time model is ... [from chapter 3], as used by @citet:Hans_2015,Elliott_2020;. This now gets added to the travel-time-so-far,
+\[
+   \ttsofar\vi = \ttsofar\vi + \pdwell_j\vi
+\]
+unless the stop is a layover. In that case, we allow the particle---with some probability---to remain at the stop until the end of the dwell time, or the scheduled departure time, whichever is later.
+
+**Step 4: repeat steps 2--3**
+
+We now repeat each of the steps, sampling travel times for intermediary road segments as the particle arrives, and storing stop arrival times, until the particle reaches the end of the route.
+
+**Step 5: obtain arrival time distributions**
+
+Since each particle stores its arrival at all stops, we can very easily obtain the predictive distribution of arrival time for the bus using the delta measure,
+\[
+   p(\Tarr_j | \Vstate_k, \vec\NWstate_k) \approx
+   \frac{1}{\tilde N}\sum_{i=1}^{\tilde N}
+      \DiracMeasure{\Tarr_j\vi}{\Tarr_j}.
+\]
+We may obtain any summary statistics from this distribution, including the mean and prediction intervals.
 
@@
 @label:pf_results;
+
+* we ran model to full day of data
+* predicted arrival times
+* at end of day, compare to actual arrival times
+* also computed GTFS prediction (schedule + delay) as comparison
+
+* the results show our method works better even without forecasting overall
+* this picture shows during off-peak, we outperform and capture the uncertainty inherent in arrival time prediction

### 2020-10-01 new2020/sections/sec03.Rmd
Results words

@@
 
-* we ran model to full day of data
-* predicted arrival times
-* at end of day, compare to actual arrival times
-* also computed GTFS prediction (schedule + delay) as comparison
+To assess the feasibility and reliability of this approach, we ran the full model on one day of data from DATE.
+This used the particle filter and network model described by @cite:Elliott_2020;, and the arrival time predictions described above.
+At the end of the day, we could compare the predicted with the actual arrival times.
+In addition, we compare to the GTFS-based predictions which are simply the scheduled arrival time plus the delay at the most recent stop.
 
-* the results show our method works better even without forecasting overall
-* this picture shows during off-peak, we outperform and capture the uncertainty inherent in arrival time prediction
+To assess, we use RMSE, MAE, MAPE, and PICP.
+RMSE and MAE provide an assessment of the absolute prediction reliability, while MAPE scales predictions by the time-until-arrival; that is, it is more affected by short-term predictions when the bus is near, giving us a comparative measure of short-term reliability (which is useful when deciding between walking, jogging, or sprinting to the bus stop).
+PICP (which is only available for the particle filter estimates) assess how well the uncertainty is captured (we would expect a 95% prediction interval to capture the true value in 95% of cases).
+
+The overall values are shown in @cref:tab:results1;.
+The particle filter predictions are slightly more accurate than GTFS on average, and have about 20% error versus GTFS's 27%.
+However, only 77% of prediction intervals contain the actual arrival time, indicating that uncertainties are being under-predicted (the nominal coverage is 85%).
+
+Since traffic and bus numbers change over the course of a day, we also computed the assessment statistics over 15\ minute intervals.
+These results are shown in @cref:fig:results2;, and immediately we can see the sizeable affect of peak time of predictive accuracy.
+This suggests that, during off-peak time when traffic conditions are stable, the real-time predictions are a vast improvment on the status quo.
+Since the underlying network model is not yet enabled to handle forecasts, its not surprising that accuracy and uncertainty coverage is lacking.

### 2020-10-01 new2020/clean_tex.R

@@ -10,7 +10,7 @@ txt <- txt[from:to]
 # convert @cite:Paper_2000,Paper2_2001; to
 # `\cite{Paper_2000, Paper2_2001}`{=latex}
 
-regx <- "@([a-zA-Z]+):([a-zA-Z0-9,_\\\\]+);"
+regx <- "@([a-zA-Z]+):([a-zA-Z0-9,_\\\\:]+);"
 m <- do.call(rbind, stringr::str_match_all(txt, regx))
 if (nrow(m)) {
     for (i in seq_len(nrow(m))) {

### 2020-10-02 new2020/sections/sec03.Rmd
Add main pf results

@@
 ---
+
+```{r,echo=FALSE}
+knitr::opts_chunk$set(fig.path="figures/")
+```
 
@@
 Since the underlying network model is not yet enabled to handle forecasts, its not surprising that accuracy and uncertainty coverage is lacking.
+
+```{r day_results,echo=FALSE,fig.width=8,fig.height=4,message=FALSE,warning=FALSE,cache=TRUE,fig.cap="Prediction accuracy results for the particle filter and GTFS methods."}
+# library(RSQLite)
+library(tidyverse)
+# library(dbplyr)
+library(patchwork)
+
+load("../data/etadata.rda")
+res_timeofday <- eta_talk %>%
+    mutate(
+        hour = timestamp %>% format("%H") %>% as.integer,
+        minute = timestamp %>% format("%M") %>% as.integer,
+        hour = hour + (floor(minute / 15) * 15 / 60)
+    ) %>%
+    filter(time_until_arrival > 0) %>%
+    group_by(hour) %>%
+    summarize(
+        pf_rmse = sqrt(mean(pf_error^2)),
+        pf_mae = mean(abs(pf_error)),
+        pf_mape = mean(abs(pf_error / time_until_arrival)),
+        pf_ci_cov = mean(pf_ci),
+        pf_ci_width = mean(pf_upper - pf_lower),
+        pf_ci_lower = quantile(pf_upper - pf_lower, 0.025),
+        pf_ci_upper = quantile(pf_upper - pf_lower, 0.975),
+        gtfs_rmse = sqrt(mean(gtfs_error^2)),
+        gtfs_mae = mean(abs(gtfs_error)),
+        gtfs_mape = mean(abs(gtfs_error / time_until_arrival))
+    ) %>% arrange(hour)
+
+pal <- viridis::magma(3)[2:1] #RColorBrewer::brewer.pal(2, "Set1")
+names(pal) <- c("Particle filter", "GTFS")
+
+plot_base <- ggplot(res_timeofday, aes(hour)) +
+    theme_classic() +
+    theme(legend.position = "none") +
+    scale_colour_manual(values = pal, name = "") +
+    xlab("Time of day (hour)")
+
+plot_rmse <- plot_base +
+    geom_path(aes(y = pf_rmse, colour = "Particle filter")) +
+    geom_path(aes(y = gtfs_rmse, colour = "GTFS"), alpha = 0.5) +
+    scale_y_continuous(
+        "RMSE (s)"
+        # trans = "log10"
+    ) +
+    theme(legend.position = "bottom")
+
+plot_mae <- plot_base +
+    geom_path(aes(y = pf_mae, colour = "Particle filter")) +
+    geom_path(aes(y = gtfs_mae, colour = "GTFS"), alpha = 0.5) +
+    scale_y_continuous(
+        "MAE (s)"
+        # trans = "log10"
+    )
+
+plot_mape <- plot_base +
+    geom_path(aes(y = 100 * pf_mape, colour = "Particle filter")) +
+    geom_path(aes(y = 100 * gtfs_mape, colour = "GTFS"), alpha = 0.5) +
+    scale_y_continuous(
+        "MAPE (%)"
+        # trans = "log10"
+    )
+
+plot_ci <- plot_base +
+    geom_hline(aes(yintercept = 85), lty = 2) +
+    geom_path(aes(y = 100 * pf_ci_cov, colour = "Particle filter")) +
+    scale_y_continuous(
+        name = "PICP (%)"
+    )
+
+plot_rmse + plot_mae +
+    plot_mape + plot_ci +
+    guide_area() +
+    plot_layout(
+        guides = "collect",
+        design = "
+            13
+            24
+            55
+        ",
+        heights = c(1, 1, 0.1)
+    )
+```

### 2020-10-02 new2020/sections/sec03.Rmd
Add table

@@
    - \usepackage{cleveref}
+   - \usepackage{booktabs}
    - \input{../symbols.tex}
@@
 
+```{r load_results,echo=FALSE,cache=TRUE}
+library(tidyverse)
+library(tidyselect)
+load("../data/etadata.rda")
+```
+
 To assess the feasibility and reliability of this approach, we ran the full model on one day of data from DATE.
@@
 
-The overall values are shown in @cref:tab:results1;.
+The overall values are shown in @cref:tab:overall_results;.
 The particle filter predictions are slightly more accurate than GTFS on average, and have about 20% error versus GTFS's 27%.
@@
 
+```{r overall_results,echo=FALSE,cache=TRUE,results="asis",dependson="load_results"}
+
+res_overall <- eta_talk %>%
+   filter(time_until_arrival > 0) %>%
+   summarize(
+      pf_rmse = sqrt(mean(pf_error^2)),
+      pf_mae = mean(abs(pf_error)),
+      pf_mape = mean(abs(pf_error / time_until_arrival)) * 100,
+      pf_cicov = mean(pf_ci) * 100,
+   #   pf_ci_width = mean(pf_upper - pf_lower),
+   #   pf_ci_lower = quantile(pf_upper - pf_lower, 0.025),
+   #   pf_ci_upper = quantile(pf_upper - pf_lower, 0.975),
+      gtfs_rmse = sqrt(mean(gtfs_error^2)),
+      gtfs_mae = mean(abs(gtfs_error)),
+      gtfs_mape = mean(abs(gtfs_error / time_until_arrival)) * 100
+   ) %>%
+   pivot_longer(everything(),
+      names_to = c("model", ".value"),
+      names_pattern = "([a-z]*)_(.*)"
+   ) %>%
+   mutate(
+      model = c("Particle filter", "Schedule-delay")
+   ) %>%
+   rename(
+      " " = "model",
+      "RMSE (s)" = "rmse",
+      "MAE (s)" = "mae",
+      "MAPE (%)" = "mape",
+      "PICP (%)" = "cicov"
+   )
+
+opts <- options(knitr.kable.NA = "")
+knitr::kable(res_overall, #"latex",
+   booktabs = TRUE,
+   # table.envir = "table",
+   digits = 0,
+   caption = "Results for particle filter and other."
+)
+options(opts)
+```
+
 Since traffic and bus numbers change over the course of a day, we also computed the assessment statistics over 15\ minute intervals.
-These results are shown in @cref:fig:results2;, and immediately we can see the sizeable affect of peak time of predictive accuracy.
-This suggests that, during off-peak time when traffic conditions are stable, the real-time predictions are a vast improvment on the status quo.
+These results are shown in @cref:fig:day_results;, and immediately we can see the sizeable affect of peak time of predictive accuracy.
+This suggests that, during off-peak time when traffic conditions are stable, the real-time predictions are a vast improvement on the status quo.
 Since the underlying network model is not yet enabled to handle forecasts, its not surprising that accuracy and uncertainty coverage is lacking.
 
-```{r day_results,echo=FALSE,fig.width=8,fig.height=4,message=FALSE,warning=FALSE,cache=TRUE,fig.cap="Prediction accuracy results for the particle filter and GTFS methods."}
-# library(RSQLite)
-library(tidyverse)
-# library(dbplyr)
+```{r day_results,echo=FALSE,fig.width=8,fig.height=4,message=FALSE,warning=FALSE,cache=TRUE,fig.cap="\\label{fig:day_results}Prediction accuracy results for the particle filter and GTFS methods.",dependson="load_results"}
 library(patchwork)
 
-load("../data/etadata.rda")
 res_timeofday <- eta_talk %>%

### 2020-10-04 new2020/sections/sec03.Rmd
texalpha in subfile

@@
    - \usepackage{booktabs}
+   - \usepackage{texalpha}
    - \input{../symbols.tex}
@@
 ```{r overall_results,echo=FALSE,cache=TRUE,results="asis",dependson="load_results"}
-
 res_overall <- eta_talk %>%

### 2020-10-04 new2020/sections/sec03.Rmd
texTalpha pkg
Try find the bug

--- 2020-10-02 new2020/sections/sec03.Rmd
@@
 title: Section Three
+
 header-includes:
    - \usepackage{cleveref}
-   - \input{../symbols.tex}
+   - \usepackage{booktabs}
 ---
@@
 
+```{r overall_results,echo=FALSE,cache=TRUE,results="asis",dependson="load_results"}
+# res_overall <- eta_talk %>%
+#    filter(time_until_arrival > 0) %>%
+#    summarize(
+#       pf_rmse = sqrt(mean(pf_error^2)),
+#       pf_mae = mean(abs(pf_error)),
+#       pf_mape = mean(abs(pf_error / time_until_arrival)) * 100,
+#       pf_cicov = mean(pf_ci) * 100,
+#    #   pf_ci_width = mean(pf_upper - pf_lower),
+#    #   pf_ci_lower = quantile(pf_upper - pf_lower, 0.025),
+#    #   pf_ci_upper = quantile(pf_upper - pf_lower, 0.975),
+#       gtfs_rmse = sqrt(mean(gtfs_error^2)),
+#       gtfs_mae = mean(abs(gtfs_error)),
+#       gtfs_mape = mean(abs(gtfs_error / time_until_arrival)) * 100
+#    ) %>%
+#    pivot_longer(everything(),
+#       names_to = c("model", ".value"),
+#       names_pattern = "([a-z]*)_(.*)"
+#    ) %>%
+#    mutate(
+#       model = c("Particle filter", "Schedule-delay")
+#    ) %>%
+#    rename(
+#       " " = "model",
+#       "RMSE (s)" = "rmse",
+#       "MAE (s)" = "mae",
+#       "MAPE (%)" = "mape",
+#       "PICP (%)" = "cicov"
+#    )
+
+# opts <- options(knitr.kable.NA = "")
+# knitr::kable(res_overall, #"latex",
+#    booktabs = TRUE,
+#    # table.envir = "table",
+#    digits = 0,
+#    caption = "Results for particle filter and other."
+# )
+# options(opts)
+```
+
 Since traffic and bus numbers change over the course of a day, we also computed the assessment statistics over 15\ minute intervals.
-These results are shown in @cref:fig:results2;, and immediately we can see the sizeable affect of peak time of predictive accuracy.
-This suggests that, during off-peak time when traffic conditions are stable, the real-time predictions are a vast improvment on the status quo.
+These results are shown in @cref:fig:day_results;, and immediately we can see the sizeable affect of peak time of predictive accuracy.
+This suggests that, during off-peak time when traffic conditions are stable, the real-time predictions are a vast improvement on the status quo.
 Since the underlying network model is not yet enabled to handle forecasts, its not surprising that accuracy and uncertainty coverage is lacking.
 
-```{r day_results,echo=FALSE,fig.width=8,fig.height=4,message=FALSE,warning=FALSE,cache=TRUE,fig.cap="Prediction accuracy results for the particle filter and GTFS methods."}
-# library(RSQLite)
-library(tidyverse)
-# library(dbplyr)
-library(patchwork)
+```{r day_results,echo=FALSE,fig.width=8,fig.height=4,message=FALSE,warning=FALSE,cache=TRUE,fig.cap="\\label{fig:day_results}Prediction accuracy results for the particle filter and GTFS methods.",dependson="load_results"}
+# library(patchwork)
 
-load("../data/etadata.rda")
-res_timeofday <- eta_talk %>%
-    mutate(
-        hour = timestamp %>% format("%H") %>% as.integer,
-        minute = timestamp %>% format("%M") %>% as.integer,
-        hour = hour + (floor(minute / 15) * 15 / 60)
-    ) %>%
-    filter(time_until_arrival > 0) %>%
-    group_by(hour) %>%
-    summarize(
-        pf_rmse = sqrt(mean(pf_error^2)),
-        pf_mae = mean(abs(pf_error)),
-        pf_mape = mean(abs(pf_error / time_until_arrival)),
-        pf_ci_cov = mean(pf_ci),
-        pf_ci_width = mean(pf_upper - pf_lower),
-        pf_ci_lower = quantile(pf_upper - pf_lower, 0.025),
-        pf_ci_upper = quantile(pf_upper - pf_lower, 0.975),
-        gtfs_rmse = sqrt(mean(gtfs_error^2)),
-        gtfs_mae = mean(abs(gtfs_error)),
-        gtfs_mape = mean(abs(gtfs_error / time_until_arrival))
-    ) %>% arrange(hour)
+# res_timeofday <- eta_talk %>%
+#     mutate(
+#         hour = timestamp %>% format("%H") %>% as.integer,
+#         minute = timestamp %>% format("%M") %>% as.integer,
+#         hour = hour + (floor(minute / 15) * 15 / 60)
+#     ) %>%
+#     filter(time_until_arrival > 0) %>%
+#     group_by(hour) %>%
+#     summarize(
+#         pf_rmse = sqrt(mean(pf_error^2)),
+#         pf_mae = mean(abs(pf_error)),
+#         pf_mape = mean(abs(pf_error / time_until_arrival)),
+#         pf_ci_cov = mean(pf_ci),
+#         pf_ci_width = mean(pf_upper - pf_lower),
+#         pf_ci_lower = quantile(pf_upper - pf_lower, 0.025),
+#         pf_ci_upper = quantile(pf_upper - pf_lower, 0.975),
+#         gtfs_rmse = sqrt(mean(gtfs_error^2)),
+#         gtfs_mae = mean(abs(gtfs_error)),
+#         gtfs_mape = mean(abs(gtfs_error / time_until_arrival))
+#     ) %>% arrange(hour)
 
-pal <- viridis::magma(3)[2:1] #RColorBrewer::brewer.pal(2, "Set1")
-names(pal) <- c("Particle filter", "GTFS")
+# pal <- viridis::magma(3)[2:1] #RColorBrewer::brewer.pal(2, "Set1")
+# names(pal) <- c("Particle filter", "GTFS")
 
-plot_base <- ggplot(res_timeofday, aes(hour)) +
-    theme_classic() +
-    theme(legend.position = "none") +
-    scale_colour_manual(values = pal, name = "") +
-    xlab("Time of day (hour)")
+# plot_base <- ggplot(res_timeofday, aes(hour)) +
+#     theme_classic() +
+#     theme(legend.position = "none") +
+#     scale_colour_manual(values = pal, name = "") +
+#     xlab("Time of day (hour)")
 
-plot_rmse <- plot_base +
-    geom_path(aes(y = pf_rmse, colour = "Particle filter")) +
-    geom_path(aes(y = gtfs_rmse, colour = "GTFS"), alpha = 0.5) +
-    scale_y_continuous(
-        "RMSE (s)"
-        # trans = "log10"
-    ) +
-    theme(legend.position = "bottom")
+# plot_rmse <- plot_base +
+#     geom_path(aes(y = pf_rmse, colour = "Particle filter")) +
+#     geom_path(aes(y = gtfs_rmse, colour = "GTFS"), alpha = 0.5) +
+#     scale_y_continuous(
+#         "RMSE (s)"
+#         # trans = "log10"
+#     ) +
+#     theme(legend.position = "bottom")
 
-plot_mae <- plot_base +
-    geom_path(aes(y = pf_mae, colour = "Particle filter")) +
-    geom_path(aes(y = gtfs_mae, colour = "GTFS"), alpha = 0.5) +
-    scale_y_continuous(
-        "MAE (s)"
-        # trans = "log10"
-    )
+# plot_mae <- plot_base +
+#     geom_path(aes(y = pf_mae, colour = "Particle filter")) +
+#     geom_path(aes(y = gtfs_mae, colour = "GTFS"), alpha = 0.5) +
+#     scale_y_continuous(
+#         "MAE (s)"
+#         # trans = "log10"
+#     )
 
-plot_mape <- plot_base +
-    geom_path(aes(y = 100 * pf_mape, colour = "Particle filter")) +
-    geom_path(aes(y = 100 * gtfs_mape, colour = "GTFS"), alpha = 0.5) +
-    scale_y_continuous(
-        "MAPE (%)"
-        # trans = "log10"
-    )
+# plot_mape <- plot_base +
+#     geom_path(aes(y = 100 * pf_mape, colour = "Particle filter")) +
+#     geom_path(aes(y = 100 * gtfs_mape, colour = "GTFS"), alpha = 0.5) +
+#     scale_y_continuous(
+#         "MAPE (%)"
+#         # trans = "log10"
+#     )
 
-plot_ci <- plot_base +
-    geom_hline(aes(yintercept = 85), lty = 2) +
-    geom_path(aes(y = 100 * pf_ci_cov, colour = "Particle filter")) +
-    scale_y_continuous(
-        name = "PICP (%)"
-    )
+# plot_ci <- plot_base +
+#     geom_hline(aes(yintercept = 85), lty = 2) +
+#     geom_path(aes(y = 100 * pf_ci_cov, colour = "Particle filter")) +
+#     scale_y_continuous(
+#         name = "PICP (%)"
+#     )
 
-plot_rmse + plot_mae +
-    plot_mape + plot_ci +
-    guide_area() +
-    plot_layout(
-        guides = "collect",
-        design = "
-            13
-            24
-            55
-        ",
-        heights = c(1, 1, 0.1)
-    )
+# plot_rmse + plot_mae +
+#     plot_mape + plot_ci +
+#     guide_area() +
+#     plot_layout(
+#         guides = "collect",
+#         design = "
+#             13
+#             24
+#             55
+#         ",
+#         heights = c(1, 1, 0.1)
+#     )
 ```

### 2020-10-04 new2020/sections/sec03.Rmd
show some results

@@
    - \usepackage{booktabs}
+   - \input{../symbols.tex}
 ---
@@
 ```{r overall_results,echo=FALSE,cache=TRUE,results="asis",dependson="load_results"}
-# res_overall <- eta_talk %>%
-#    filter(time_until_arrival > 0) %>%
-#    summarize(
-#       pf_rmse = sqrt(mean(pf_error^2)),
-#       pf_mae = mean(abs(pf_error)),
-#       pf_mape = mean(abs(pf_error / time_until_arrival)) * 100,
-#       pf_cicov = mean(pf_ci) * 100,
-#    #   pf_ci_width = mean(pf_upper - pf_lower),
-#    #   pf_ci_lower = quantile(pf_upper - pf_lower, 0.025),
-#    #   pf_ci_upper = quantile(pf_upper - pf_lower, 0.975),
-#       gtfs_rmse = sqrt(mean(gtfs_error^2)),
-#       gtfs_mae = mean(abs(gtfs_error)),
-#       gtfs_mape = mean(abs(gtfs_error / time_until_arrival)) * 100
-#    ) %>%
-#    pivot_longer(everything(),
-#       names_to = c("model", ".value"),
-#       names_pattern = "([a-z]*)_(.*)"
-#    ) %>%
-#    mutate(
-#       model = c("Particle filter", "Schedule-delay")
-#    ) %>%
-#    rename(
-#       " " = "model",
-#       "RMSE (s)" = "rmse",
-#       "MAE (s)" = "mae",
-#       "MAPE (%)" = "mape",
-#       "PICP (%)" = "cicov"
-#    )
+res_overall <- eta_talk %>%
+   filter(time_until_arrival > 0) %>%
+   summarize(
+      pf_rmse = sqrt(mean(pf_error^2)),
+      pf_mae = mean(abs(pf_error)),
+      pf_mape = mean(abs(pf_error / time_until_arrival)) * 100,
+      pf_cicov = mean(pf_ci) * 100,
+   #   pf_ci_width = mean(pf_upper - pf_lower),
+   #   pf_ci_lower = quantile(pf_upper - pf_lower, 0.025),
+   #   pf_ci_upper = quantile(pf_upper - pf_lower, 0.975),
+      gtfs_rmse = sqrt(mean(gtfs_error^2)),
+      gtfs_mae = mean(abs(gtfs_error)),
+      gtfs_mape = mean(abs(gtfs_error / time_until_arrival)) * 100
+   ) %>%
+   pivot_longer(everything(),
+      names_to = c("model", ".value"),
+      names_pattern = "([a-z]*)_(.*)"
+   ) %>%
+   mutate(
+      model = c("Particle filter", "Schedule-delay")
+   ) %>%
+   rename(
+      " " = "model",
+      "RMSE (s)" = "rmse",
+      "MAE (s)" = "mae",
+      "MAPE (%)" = "mape",
+      "PICP (%)" = "cicov"
+   )
 

### 2020-10-04 new2020/sections/sec03.Rmd
comment more; add caching
hide output

--- 2020-10-04 new2020/sections/sec03.Rmd
@@
 
-```{r load_results,echo=FALSE,cache=TRUE}
+```{r load_results,echo=FALSE,cache=TRUE,message=FALSE}
 library(tidyverse)

### 2020-10-04 new2020/sections/sec03.Rmd
Uncomment-  bug was in message output

--- 2020-10-04 new2020/sections/sec03.Rmd
@@
 
-# opts <- options(knitr.kable.NA = "")
-# knitr::kable(res_overall, #"latex",
-#    booktabs = TRUE,
-#    # table.envir = "table",
-#    digits = 0,
-#    caption = "Results for particle filter and other."
-# )
-# options(opts)
+opts <- options(knitr.kable.NA = "")
+knitr::kable(res_overall, #"latex",
+   booktabs = TRUE,
+   # table.envir = "table",
+   digits = 0,
+   caption = "Results for particle filter and other."
+)
+options(opts)
 ```
@@
 ```{r day_results,echo=FALSE,fig.width=8,fig.height=4,message=FALSE,warning=FALSE,cache=TRUE,fig.cap="\\label{fig:day_results}Prediction accuracy results for the particle filter and GTFS methods.",dependson="load_results"}
-# library(patchwork)
+library(patchwork)
 
-# res_timeofday <- eta_talk %>%
-#     mutate(
-#         hour = timestamp %>% format("%H") %>% as.integer,
-#         minute = timestamp %>% format("%M") %>% as.integer,
-#         hour = hour + (floor(minute / 15) * 15 / 60)
-#     ) %>%
-#     filter(time_until_arrival > 0) %>%
-#     group_by(hour) %>%
-#     summarize(
-#         pf_rmse = sqrt(mean(pf_error^2)),
-#         pf_mae = mean(abs(pf_error)),
-#         pf_mape = mean(abs(pf_error / time_until_arrival)),
-#         pf_ci_cov = mean(pf_ci),
-#         pf_ci_width = mean(pf_upper - pf_lower),
-#         pf_ci_lower = quantile(pf_upper - pf_lower, 0.025),
-#         pf_ci_upper = quantile(pf_upper - pf_lower, 0.975),
-#         gtfs_rmse = sqrt(mean(gtfs_error^2)),
-#         gtfs_mae = mean(abs(gtfs_error)),
-#         gtfs_mape = mean(abs(gtfs_error / time_until_arrival))
-#     ) %>% arrange(hour)
+res_timeofday <- eta_talk %>%
+    mutate(
+        hour = timestamp %>% format("%H") %>% as.integer,
+        minute = timestamp %>% format("%M") %>% as.integer,
+        hour = hour + (floor(minute / 15) * 15 / 60)
+    ) %>%
+    filter(time_until_arrival > 0) %>%
+    group_by(hour) %>%
+    summarize(
+        pf_rmse = sqrt(mean(pf_error^2)),
+        pf_mae = mean(abs(pf_error)),
+        pf_mape = mean(abs(pf_error / time_until_arrival)),
+        pf_ci_cov = mean(pf_ci),
+        pf_ci_width = mean(pf_upper - pf_lower),
+        pf_ci_lower = quantile(pf_upper - pf_lower, 0.025),
+        pf_ci_upper = quantile(pf_upper - pf_lower, 0.975),
+        gtfs_rmse = sqrt(mean(gtfs_error^2)),
+        gtfs_mae = mean(abs(gtfs_error)),
+        gtfs_mape = mean(abs(gtfs_error / time_until_arrival))
+    ) %>% arrange(hour)
 
-# pal <- viridis::magma(3)[2:1] #RColorBrewer::brewer.pal(2, "Set1")
-# names(pal) <- c("Particle filter", "GTFS")
+pal <- viridis::magma(3)[2:1] #RColorBrewer::brewer.pal(2, "Set1")
+names(pal) <- c("Particle filter", "GTFS")
 
-# plot_base <- ggplot(res_timeofday, aes(hour)) +
-#     theme_classic() +
-#     theme(legend.position = "none") +
-#     scale_colour_manual(values = pal, name = "") +
-#     xlab("Time of day (hour)")
+plot_base <- ggplot(res_timeofday, aes(hour)) +
+    theme_classic() +
+    theme(legend.position = "none") +
+    scale_colour_manual(values = pal, name = "") +
+    xlab("Time of day (hour)")
 
-# plot_rmse <- plot_base +
-#     geom_path(aes(y = pf_rmse, colour = "Particle filter")) +
-#     geom_path(aes(y = gtfs_rmse, colour = "GTFS"), alpha = 0.5) +
-#     scale_y_continuous(
-#         "RMSE (s)"
-#         # trans = "log10"
-#     ) +
-#     theme(legend.position = "bottom")
+plot_rmse <- plot_base +
+    geom_path(aes(y = pf_rmse, colour = "Particle filter")) +
+    geom_path(aes(y = gtfs_rmse, colour = "GTFS"), alpha = 0.5) +
+    scale_y_continuous(
+        "RMSE (s)"
+        # trans = "log10"
+    ) +
+    theme(legend.position = "bottom")
 
-# plot_mae <- plot_base +
-#     geom_path(aes(y = pf_mae, colour = "Particle filter")) +
-#     geom_path(aes(y = gtfs_mae, colour = "GTFS"), alpha = 0.5) +
-#     scale_y_continuous(
-#         "MAE (s)"
-#         # trans = "log10"
-#     )
+plot_mae <- plot_base +
+    geom_path(aes(y = pf_mae, colour = "Particle filter")) +
+    geom_path(aes(y = gtfs_mae, colour = "GTFS"), alpha = 0.5) +
+    scale_y_continuous(
+        "MAE (s)"
+        # trans = "log10"
+    )
 
-# plot_mape <- plot_base +
-#     geom_path(aes(y = 100 * pf_mape, colour = "Particle filter")) +
-#     geom_path(aes(y = 100 * gtfs_mape, colour = "GTFS"), alpha = 0.5) +
-#     scale_y_continuous(
-#         "MAPE (%)"
-#         # trans = "log10"
-#     )
+plot_mape <- plot_base +
+    geom_path(aes(y = 100 * pf_mape, colour = "Particle filter")) +
+    geom_path(aes(y = 100 * gtfs_mape, colour = "GTFS"), alpha = 0.5) +
+    scale_y_continuous(
+        "MAPE (%)"
+        # trans = "log10"
+    )
 
-# plot_ci <- plot_base +
-#     geom_hline(aes(yintercept = 85), lty = 2) +
-#     geom_path(aes(y = 100 * pf_ci_cov, colour = "Particle filter")) +
-#     scale_y_continuous(
-#         name = "PICP (%)"
-#     )
+plot_ci <- plot_base +
+    geom_hline(aes(yintercept = 85), lty = 2) +
+    geom_path(aes(y = 100 * pf_ci_cov, colour = "Particle filter")) +
+    scale_y_continuous(
+        name = "PICP (%)"
+    )
 
-# plot_rmse + plot_mae +
-#     plot_mape + plot_ci +
-#     guide_area() +
-#     plot_layout(
-#         guides = "collect",
-#         design = "
-#             13
-#             24
-#             55
-#         ",
-#         heights = c(1, 1, 0.1)
-#     )
+plot_rmse + plot_mae +
+    plot_mape + plot_ci +
+    guide_area() +
+    plot_layout(
+        guides = "collect",
+        design = "
+            13
+            24
+            55
+        ",
+        heights = c(1, 1, 0.1)
+    )
 ```

### 2020-10-05 new2020/sections/sec04.Rmd
begin adding structure
Some words for Cdf stuff - and ready for more data/graphs/tables

@@
    - \usepackage{cleveref}
+   - \input{../symbols.tex}
 ---
 
-# Journey planning applications
+# A discretized CDF approximation for journey planning
+
+One of the main problems with particle filter results is that we have a large sample of points, rather than the description of a posterior distribution (as would be the case with a Kalman filter, for example).
+Trying to do any real-time processing, for example computing event probabilities, would be very computationally intensive, and probably not feasible on a user's phone, which would require the server to not only generate the arrival time distributions, but to process individual journey planning queries too.
+
+To consider a simpler approximation, we start by noting that arrival times are generally displayed to the user in one-minute accuracy.
+Therefore, we can consider the concept of discretizing arrival times at the one-minute level and obtaining a CDF which for the probability that the bus arrives within $a$\ minutes.
+
+Computing a CDF approximation using particles is in fact a trivial exercise, and can be done in real-time without any negative affect on performance.
+This is because the method described below *does not* require sorting of the particles, which we would need to do to compute, for example, quantiles (median, or a prediction interval).
+This essentially consists of counting the number of particles in each one-minute interval and expressing it as a probability.
+Here, the particle arrival time $\alpha\vi$ is in seconds, is transformed to minutes and rounded down using the floor operator, while the discrete CDF is reported in $a$\ minutes.
+\begin{equation}
+   \label{eq:eta_count}
+   \Pr(A \in [a, a+1]) = 
+   \frac{1}{\tilde N} \sum_{i=1}^{\tilde N} I\left\{ \lfloor\frac{\alpha\vi}{60}\rfloor = a \right\},
+\end{equation}
+where the indicator $I\{a=b\}$ is 1 if $a$ equals $b$, and zero otherwise.
+
+Using @cref:eq:eta_count; the CDF for bus arrival at a given stop is
+\[
+   P(A < a) = \sum_{x=0}^{x=a-1} \Pr{A \in [x, x+1]},
+\]
+and is demonsrated graphically in @cref:fig:pf_cdf;.
+
+```{r pf_cdf,echo=FALSE,fig.width=6,fig.height=4,message=FALSE,warning=FALSE,cache=TRUE,fig.cap="\\label{fig:pf_cdf}Particle filter estimates of arrival time and the process to obtain a discretized CDF approximation."}
+library(tidyverse)
+set.seed(346)
+N <- 2e3
+x <- rlnorm(N, log(14), 0.14)
+d <- data.frame(x = x)
+
+cdf <- table(floor(x))
+cdf <- c(structure(0, .Names = min(floor(x)) - 1), cdf)
+d2 <- data.frame(a = as.integer(names(cdf)), n = as.integer(cdf)) %>%
+   mutate(
+      n2 = cumsum(n),
+      p = n2 / N
+   )
+
+p1 <- ggplot(d, aes(x)) +
+   geom_dotplot(dotsize = 1, binwidth = 0.09, fill = "#333333", colour = "white") +
+   geom_histogram(
+      fill = "transparent", 
+      colour = "black",
+      binwidth = 1,
+      boundary = 10
+   ) +
+   theme_classic() +
+   xlab("") +
+   ylab("Number of particles") +
+   xlim(min(d2$a), max(d2$a))
+
+p2 <- ggplot(d2, aes(a, p)) +
+   geom_step() +
+   theme_classic() +
+   xlab("Arrival time, a (minutes") +
+   ylab("Pr(A < a)")
+
+library(patchwork)
+p1 / p2
+```
+
+The CDF can easily be distributed to users' phones, since the data structure is simple and consists of only a few values.
+Now we assess the validity of the simplification by predicting event probabilities and comparing those to the truth, as well as the GTFS binary estimate of the event outcomes.
+
+Of course, we can obtain simple summary statistics for the distribution, such as the median or other quantiles. For a quantile $q\in(0,1)$, we calculate
+\[
+   \hat A_q = \max\{a\in\{0,1,\ldots\} : \Pr(A < a) \leq q\}
+\]
+which provides the smallest arrival time, in minutes, with a probability of no more than $q$ of the bus arriving before that time.
+
+Another question, however, is to compute probabilities of events.
+The simplest is, of course, the probability that the bus arrives before or after a specific time, $\Pr{A < a}$ or $\Pr{A \geq a}$.
+This is useful for answer questions such as
+
+* what is the probability of catching the bus if I arrive by 9:30am?
+* what is the probability the bus will arrive before 1pm for my meeting?
+
+Calculating these events is straighforward given the CDF, and can be performed on a user's phone or quickly on the server.
+We performed a journey planning exercise to assess the reliability of these results by computing the probablity of a bus arriving before a specific time and assessing the outcome.
+Due to restrains on the current implementation, we only have arrival time distributions for buses which have begun the route.
+The journey planning problem is this: *given two alternative bus stop options requiring walking in opposite directions, which is the best to take?* We account for the walking time to both stops A and B, and compute the probabilities of catching the next few buses.
 
@@
 * this is a simple computation - can be done client side (i.e., on a user's phone) by passing CDF (small size, as e.g., JSON)
+
+## Simplified ETA CDF
+@label:eta_cdf;
+
+* round to minutes
+* compute the CDF by definition ``number of particles arriving within $x$ minutes''

### 2020-10-06 new2020/sections/sec03.Rmd
Some words for Cdf stuff - and ready for more data/graphs/tables
More results + finally figure out there's a bug in the statsoc class re tables...

@@
 ## Particle filter ETAs
-@label:pf_etas;
 
@@
 ## Results
-@label:pf_results;
 
@@
 ```{r overall_results,echo=FALSE,cache=TRUE,results="asis",dependson="load_results"}
+
 res_overall <- eta_talk %>%
@@
    booktabs = TRUE,
-   # table.envir = "table",
+   format = "latex",
+   # label = "overall_results",
    digits = 0,
-   caption = "Results for particle filter and other."
+   caption = "Results for particle filter and other.",
+   caption.short = "Short cap"
 )
@@
 ```
-
-Since traffic and bus numbers change over the course of a day, we also computed the assessment statistics over 15\ minute intervals.
-These results are shown in @cref:fig:day_results;, and immediately we can see the sizeable affect of peak time of predictive accuracy.
-This suggests that, during off-peak time when traffic conditions are stable, the real-time predictions are a vast improvement on the status quo.
-Since the underlying network model is not yet enabled to handle forecasts, its not surprising that accuracy and uncertainty coverage is lacking.
-
-```{r day_results,echo=FALSE,fig.width=8,fig.height=4,message=FALSE,warning=FALSE,cache=TRUE,fig.cap="\\label{fig:day_results}Prediction accuracy results for the particle filter and GTFS methods.",dependson="load_results"}
-library(patchwork)
-
-res_timeofday <- eta_talk %>%
-    mutate(
-        hour = timestamp %>% format("%H") %>% as.integer,
-        minute = timestamp %>% format("%M") %>% as.integer,
-        hour = hour + (floor(minute / 15) * 15 / 60)
-    ) %>%
-    filter(time_until_arrival > 0) %>%
-    group_by(hour) %>%
-    summarize(
-        pf_rmse = sqrt(mean(pf_error^2)),
-        pf_mae = mean(abs(pf_error)),
-        pf_mape = mean(abs(pf_error / time_until_arrival)),
-        pf_ci_cov = mean(pf_ci),
-        pf_ci_width = mean(pf_upper - pf_lower),
-        pf_ci_lower = quantile(pf_upper - pf_lower, 0.025),
-        pf_ci_upper = quantile(pf_upper - pf_lower, 0.975),
-        gtfs_rmse = sqrt(mean(gtfs_error^2)),
-        gtfs_mae = mean(abs(gtfs_error)),
-        gtfs_mape = mean(abs(gtfs_error / time_until_arrival))
-    ) %>% arrange(hour)
-
-pal <- viridis::magma(3)[2:1] #RColorBrewer::brewer.pal(2, "Set1")
-names(pal) <- c("Particle filter", "GTFS")
-
-plot_base <- ggplot(res_timeofday, aes(hour)) +
-    theme_classic() +
-    theme(legend.position = "none") +
-    scale_colour_manual(values = pal, name = "") +
-    xlab("Time of day (hour)")
-
-plot_rmse <- plot_base +
-    geom_path(aes(y = pf_rmse, colour = "Particle filter")) +
-    geom_path(aes(y = gtfs_rmse, colour = "GTFS"), alpha = 0.5) +
-    scale_y_continuous(
-        "RMSE (s)"
-        # trans = "log10"
-    ) +
-    theme(legend.position = "bottom")
-
-plot_mae <- plot_base +
-    geom_path(aes(y = pf_mae, colour = "Particle filter")) +
-    geom_path(aes(y = gtfs_mae, colour = "GTFS"), alpha = 0.5) +
-    scale_y_continuous(
-        "MAE (s)"
-        # trans = "log10"
-    )
-
-plot_mape <- plot_base +
-    geom_path(aes(y = 100 * pf_mape, colour = "Particle filter")) +
-    geom_path(aes(y = 100 * gtfs_mape, colour = "GTFS"), alpha = 0.5) +
-    scale_y_continuous(
-        "MAPE (%)"
-        # trans = "log10"
-    )
-
-plot_ci <- plot_base +
-    geom_hline(aes(yintercept = 85), lty = 2) +
-    geom_path(aes(y = 100 * pf_ci_cov, colour = "Particle filter")) +
-    scale_y_continuous(
-        name = "PICP (%)"
-    )
-
-plot_rmse + plot_mae +
-    plot_mape + plot_ci +
-    guide_area() +
-    plot_layout(
-        guides = "collect",
-        design = "
-            13
-            24
-            55
-        ",
-        heights = c(1, 1, 0.1)
-    )
-```

### 2020-10-06 new2020/sections/sec04.Rmd
More results + finally figure out there's a bug in the statsoc class re tables...

@@
    - \usepackage{cleveref}
+   - \usepackage{booktabs}
    - \input{../symbols.tex}
 ---
+
+```{r,echo=FALSE}
+knitr::opts_chunk$set(fig.path="figures/")
+```
 
@@
    \label{eq:eta_count}
-   \Pr(A \in [a, a+1]) = 
+   \Pr{A \in [a, a+1]} =
    \frac{1}{\tilde N} \sum_{i=1}^{\tilde N} I\left\{ \lfloor\frac{\alpha\vi}{60}\rfloor = a \right\},
@@
 
-```{r pf_cdf,echo=FALSE,fig.width=6,fig.height=4,message=FALSE,warning=FALSE,cache=TRUE,fig.cap="\\label{fig:pf_cdf}Particle filter estimates of arrival time and the process to obtain a discretized CDF approximation."}
+```{r pf_cdf,echo=FALSE,fig.width=6,fig.height=4,out.width="\\textwidth",message=FALSE,warning=FALSE,cache=TRUE,fig.cap="Particle filter estimates of arrival time and the process to obtain a discretized CDF approximation."}
 library(tidyverse)
@@
 \[
-   \hat A_q = \max\{a\in\{0,1,\ldots\} : \Pr(A < a) \leq q\}
+   \hat A_q = \max\{a\in\{0,1,\ldots\} : \Pr{A < a} \leq q\}
 \]
@@
 Due to restrains on the current implementation, we only have arrival time distributions for buses which have begun the route.
-The journey planning problem is this: *given two alternative bus stop options requiring walking in opposite directions, which is the best to take?* We account for the walking time to both stops A and B, and compute the probabilities of catching the next few buses.
+The journey planning problem is this: *given two alternative bus stop options requiring walking in opposite directions, which is the best to take?* We account for the walking time $W_1$ and $W_2$ to both stops $R_1$ and $R_2$, respectively, and compute the probabilities of catching the next few buses.
+The probability of catching $R_j$, $j=1,2$, if leaving now at time $T$, is given by
+\[
+   \Pr{\text{catch} | \text{choose } R_j} =
+   \Pr{A_j \geq T + W_j} =
+   1 - \Pr{A_j < T + W_j}
+\]
+which is easily obtained from the CDF.
+To simulate, we chose a sequence of times from 9\ am til 3\ pm in 15\ minute intervals, and for each calculated the probability of catching upcoming buses at the two alternative stops.
+This was compared to the outcome, and the GTFS binary prediction.
+The results are shown in @cref:fig:jp_catch;.
+
+```{r jp_prep,echo=FALSE,message=FALSE,cache=TRUE}
+library(tidyverse)
+library(patchwork)
+load("../data/all_results.rda")
+```
+
+```{r jp_catch,cache=TRUE,dependson="jp_prep",echo=FALSE,fig.width=8,fig.height=2.5,out.width="\\textwidth",message=FALSE,warning=FALSE,fig.cap="Predicted outcomes of journey planning process."}
+## Now plot p-catch versus did_catch
+mdata <- map(all_results, "summary") %>%
+  bind_rows %>%
+  mutate(
+    did_catch = as.integer(did_catch),
+    did_arrive = as.integer(did_arrive),
+    gtfs_catch = ifelse(gtfs_catch == 1, "Yes", "No"),
+    gtfs_arrive = ifelse(gtfs_arrive == 1, "Yes", "No")
+  )
+
+mdata %>%
+   mutate(did_catch = ifelse(did_catch, "Bus was caught", "Bus was missed")) %>%
+   ggplot(aes(p_catch + rnorm(length(p_catch), 0, 0.01), gtfs_catch)) +
+      ggbeeswarm::geom_beeswarm(
+         groupOnX = FALSE,
+         priority = "random"
+      ) +
+      theme_classic() +
+      scale_x_continuous("Predicted probabilty of catching the bus") +
+      scale_y_discrete("GTFS prediction") +
+      facet_grid(~did_catch) +
+      theme(
+         strip.background = element_blank(),
+         strip.text = element_text(hjust = 0)
+      )
+
+gtfs_catch_tab <- table(
+  gtfs = mdata$gtfs_catch,
+  outcome = ifelse(mdata$did_catch, "Yes", "No")
+)
+gtfs_catch_pc <-
+  round(sweep(gtfs_catch_tab, 1, rowSums(gtfs_catch_tab), "/"), 2)
+
+pf_catch_tab50 <- table(
+   pf = ifelse(mdata$p_catch > 0.5, "Yes", "No"),
+   outcome = ifelse(mdata$did_catch, "Yes", "No")
+)
+pf_catch_pc50 <-
+   round(sweep(pf_catch_tab50, 1, rowSums(pf_catch_tab50), "/"), 2)
+
+pf_catch_tab80 <- table(
+   pf = ifelse(mdata$p_catch > 0.8, "Yes", "No"),
+   outcome = ifelse(mdata$did_catch, "Yes", "No")
+)
+pf_catch_pc80 <-
+   round(sweep(pf_catch_tab80, 1, rowSums(pf_catch_tab80), "/"), 2)
+
+tab <- cbind(
+  "Predicted outcome" = c("No", "Yes"),
+  format(gtfs_catch_pc, nsmall = 2L, digits = 2L),
+  NA,
+  format(pf_catch_pc50, nsmall = 2L, digits = 2L),
+  NA,
+  format(pf_catch_pc80, nsmall = 2L, digits = 2L)
+)
+
+# roc curve?
+pc <- seq(0, 1, by = 0.05)
+roc <- lapply(pc,
+   function(p) {
+      tab <- table(
+         pf = factor(ifelse(mdata$p_catch > p, "Yes", "No"), levels = c("No", "Yes")),
+         outcome = factor(ifelse(mdata$did_catch, "Yes", "No"), levels = c("No", "Yes"))
+      )
+      P <- sum(tab[,2])
+      N <- sum(tab[,1])
+      TP <- tab[2,2]
+      TN <- tab[1,1]
+      FP <- tab[2,1]
+      FN <- tab[1,2]
+      c(
+         p = p,
+         tpr = TP / P,
+         tnr = TN / N,
+         ppv = TP / (TP + FP),
+         fnr = FN / P,
+         fpr = FP / N
+      )
+   }
+) %>% bind_rows
+
+gtfs_roc <- data.frame(
+   fpr = gtfs_catch_tab[2,1] / sum(gtfs_catch_tab[,1]),
+   tpr = gtfs_catch_tab[2,2] / sum(gtfs_catch_tab[,2])
+)
+```
+
+```{r jp_catch_dec,cache=TRUE,dependson="jp_catch",echo=FALSE,fig.width=5,fig.height=2.5,out.width="0.5\\textwidth",fig.align="center",message=FALSE,warning=FALSE,fig.cap="False positive rate for varying decision level.
+Increasing the threshold for predicting a positive outcome reduces the *miss rate*, at the cost of not trying to catch potential buses."}
+gp <- 0.175
+ggplot(roc, aes(p, fpr)) +
+   geom_path(lwd = 1.2) +
+   coord_cartesian(expand = FALSE) +
+   geom_segment(aes(x = gp, xend = 0, yend = fpr),
+      data = gtfs_roc, colour = "red", lty = 2
+   ) +
+   geom_segment(aes(x = gp, xend = gp, yend = 0),
+      data = gtfs_roc, colour = "red", lty = 2) +
+   xlab("Decision threshold") +
+   ylab("False positive / miss rate") +
+   theme_classic()
+```
+
+```{r jp_catch_tab,cache=TRUE,dependson="jp_catch",echo=FALSE,results='asis'}
+options(knitr.kable.NA = "")
+knitr::kable(
+  tab,
+  row.names = FALSE,
+  format = "latex",
+  booktabs = TRUE,
+#   label = "pf_catch_tab",
+  caption = "Prediction results for all journeys, with the displayed values representing the proportion of outcomes in each cell (rows are conditioned by the prediction).",
+  caption.short = "Prediction results for all journeys",
+  escape = FALSE,
+  linesep = ""
+) %>%
+  kableExtra::add_header_above(
+    c(" ", "Schedule-delay" = 2, " ", "P > 0.5" = 2, " ", "P > 0.8" = 2)
+  ) %>%
+  kableExtra::add_header_above(
+    c("", "Observed outcome" = 8)
+  ) %>%
+  kableExtra::kable_styling(font_size = 8)
+```
+
+Discussion of results!!
+
+* we can adjust decision to fit the situation - coffee with friend vs job interview
+* doesn't account for *wait time* (i.e., if you miss the bus, how long till the next one)
+
+The second journey planning scenario is *arriving before a specified time*, which requires forecasting further into the future.
+Such a decision might be important for someone, allowing them to either inform their friend they'll be late for that coffee, or catching a taxi to get to the interview.
+As before, we selected a stop and this time asked if the bus would arrive on-time (where on time was 15--45\ minutes in the future).
+The results, as before, are displayed in, where the successfull event is now that the bus arrived before the specified time $A$.
+
+```{r jp_arrive,cache=TRUE,dependson="jp_catch",echo=FALSE,fig.width=8,fig.height=2.5,out.width="\\textwidth",message=FALSE,warning=FALSE,fig.cap="Predicted outcomes of journey planning process - arrive on time."}
+## Now plot p-catch versus did_catch
+
+mdata %>%
+   mutate(did_arrive = ifelse(did_arrive, "Arrived on time", "Arrived late")) %>%
+   ggplot(aes(p_arrive + rnorm(length(p_arrive), 0, 0.01), gtfs_arrive)) +
+      ggbeeswarm::geom_beeswarm(
+         groupOnX = FALSE,
+         priority = "random"
+      ) +
+      theme_classic() +
+      scale_x_continuous("Predicted probabilty of bus arriving on time") +
+      scale_y_discrete("GTFS prediction") +
+      facet_grid(~did_arrive) +
+      theme(
+         strip.background = element_blank(),
+         strip.text = element_text(hjust = 0)
+      )
+
+gtfs_arrive_tab <- table(
+  gtfs = mdata$gtfs_arrive,
+  outcome = ifelse(mdata$did_arrive, "Yes", "No")
+)
+gtfs_arrive_pc <-
+  round(sweep(gtfs_arrive_tab, 1, rowSums(gtfs_arrive_tab), "/"), 2)
+
+pf_arrive_tab50 <- table(
+   pf = ifelse(mdata$p_arrive > 0.5, "Yes", "No"),
+   outcome = ifelse(mdata$did_arrive, "Yes", "No")
+)
+pf_arrive_pc50 <-
+   round(sweep(pf_arrive_tab50, 1, rowSums(pf_arrive_tab50), "/"), 2)
+
+pf_arrive_tab80 <- table(
+   pf = ifelse(mdata$p_arrive > 0.8, "Yes", "No"),
+   outcome = ifelse(mdata$did_arrive, "Yes", "No")
+)
+pf_arrive_pc80 <-
+   round(sweep(pf_arrive_tab80, 1, rowSums(pf_arrive_tab80), "/"), 2)
+
+tab <- cbind(
+  "Predicted outcome" = c("No", "Yes"),
+  format(gtfs_arrive_pc, nsmall = 2L, digits = 2L),
+  NA,
+  format(pf_arrive_pc50, nsmall = 2L, digits = 2L),
+  NA,
+  format(pf_arrive_pc80, nsmall = 2L, digits = 2L)
+)
+
+# roc curve?
+pc <- seq(0, 1, by = 0.05)
+roc <- lapply(pc,
+   function(p) {
+      tab <- table(
+         pf = factor(ifelse(mdata$p_arrive > p, "Yes", "No"), levels = c("No", "Yes")),
+         outcome = factor(ifelse(mdata$did_arrive, "Yes", "No"), levels = c("No", "Yes"))
+      )
+      P <- sum(tab[,2])
+      N <- sum(tab[,1])
+      TP <- tab[2,2]
+      TN <- tab[1,1]
+      FP <- tab[2,1]
+      FN <- tab[1,2]
+      c(
+         p = p,
+         tpr = TP / P,
+         tnr = TN / N,
+         ppv = TP / (TP + FP),
+         fnr = FN / P,
+         fpr = FP / N
+      )
+   }
+) %>% bind_rows
+
+gtfs_roc <- data.frame(
+   fpr = gtfs_arrive_tab[2,1] / sum(gtfs_arrive_tab[,1]),
+   tpr = gtfs_arrive_tab[2,2] / sum(gtfs_arrive_tab[,2])
+)
+```
+
+```{r jp_arrive_dec,cache=TRUE,dependson="jp_arrive",echo=FALSE,fig.width=5,fig.height=2.5,out.width="0.5\\textwidth",fig.align="center",message=FALSE,warning=FALSE,fig.cap="False positive rate for varying decision level.
+Increasing the threshold for predicting a positive outcome reduces the arrival late rate, at the cost of not trying to catch potential buses."}
+gp <- 0.175
+ggplot(roc, aes(p, fpr)) +
+   geom_path(lwd = 1.2) +
+   coord_cartesian(expand = FALSE) +
+   # geom_segment(aes(x = gp, xend = 0, yend = fpr),
+   #    data = gtfs_roc, colour = "red", lty = 2
+   # ) +
+   # geom_segment(aes(x = gp, xend = gp, yend = 0),
+   #    data = gtfs_roc, colour = "red", lty = 2) +
+   xlab("Decision threshold") +
+   ylab("False positive") +
+   theme_classic()
+```
+
+```{r jp_arrive_tab,cache=TRUE,dependson="jp_arrive",echo=FALSE,results='asis'}
+options(knitr.kable.NA = "")
+knitr::kable(
+  tab,
+  row.names = FALSE,
+  format = "latex",
+  booktabs = TRUE,
+#   label = "pf_catch_tab",
+  caption = "Prediction results for all journeys 2, with the displayed values representing the proportion of outcomes in each cell (rows are conditioned by the prediction).",
+  caption.short = "Prediction results for all journeys 2",
+  escape = FALSE,
+  linesep = ""
+) %>%
+  kableExtra::add_header_above(
+    c(" ", "Schedule-delay" = 2, " ", "P > 0.5" = 2, " ", "P > 0.8" = 2)
+  ) %>%
+  kableExtra::add_header_above(
+    c("", "Observed outcome" = 8)
+  ) %>%
+  kableExtra::kable_styling(font_size = 8)
+```
+
+The final scenario is for a journey which requires transfering between two services, in which case we want to predict the probability that a transfer between two particular trips at the transfer stop will be successful (that is, the first bus arrives before the second).
 
@@
 * this is a simple computation - can be done client side (i.e., on a user's phone) by passing CDF (small size, as e.g., JSON)
-
-## Simplified ETA CDF
-@label:eta_cdf;
-
-* round to minutes
-* compute the CDF by definition ``number of particles arriving within $x$ minutes''

### 2020-10-06 new2020/get_deps.R
Automatically (?) keep track of and install dependencies

@@ -1,0 +1,17 @@
+ca <- commandArgs(TRUE)
+files <- list.files("sections", pattern = ".Rmd", full.names = TRUE)
+text <- suppressWarnings(do.call(c, sapply(files, readLines)))
+
+if (!requireNamespace("stringr", quietly = TRUE))
+    install.packages("stringr")
+
+pkgs <- c(
+    stringr::str_match(text, "library\\((.+)\\)")[,2],
+    stringr::str_match(text, "([a-zA-Z0-9]+)::")[,2]
+)
+pkgs <- unique(pkgs[!is.na(pkgs)])
+if (length(ca)) saveRDS(pkgs, file = ca[1], version = 2)
+pkgs <- pkgs[!sapply(pkgs, requireNamespace, quietly = TRUE)]
+
+if (length(pkgs))
+    install.packages(pkgs)

### 2020-10-06 new2020/get_deps.R
Install if no dep file specified

@@ -10,8 +10,10 @@ pkgs <- c(
     stringr::str_match(text, "([a-zA-Z0-9]+)::")[,2]
 )
 pkgs <- unique(pkgs[!is.na(pkgs)])
-if (length(ca)) saveRDS(pkgs, file = ca[1], version = 2)
-pkgs <- pkgs[!sapply(pkgs, requireNamespace, quietly = TRUE)]
-
-if (length(pkgs))
-    install.packages(pkgs)
+if (length(ca)) {
+    saveRDS(pkgs, file = ca[1], version = 2)
+} else {
+    pkgs <- pkgs[!sapply(pkgs, requireNamespace, quietly = TRUE)]
+    if (length(pkgs))
+        install.packages(pkgs)
+}

### 2020-10-11 new2020/sections/sec03.Rmd
Last(?) of the main results

@@
       pf_cicov = mean(pf_ci) * 100,
-   #   pf_ci_width = mean(pf_upper - pf_lower),
-   #   pf_ci_lower = quantile(pf_upper - pf_lower, 0.025),
-   #   pf_ci_upper = quantile(pf_upper - pf_lower, 0.975),
       gtfs_rmse = sqrt(mean(gtfs_error^2)),
@@
 ```
+
+```{r time_results,echo=FALSE,message=FALSE,warning=FALSE,cache=TRUE,dependson="overall_results",fig.width=8,fig.height=5,out.width="\\textwidth",fig.align="center",fig.cap="Results by time of day."}
+res_timeofday <- eta_talk %>%
+    mutate(
+        hour = timestamp %>% format("%H") %>% as.integer,
+        minute = timestamp %>% format("%M") %>% as.integer,
+        hour = hour + (floor(minute / 15) * 15 / 60)
+    ) %>%
+    filter(time_until_arrival > 0) %>%
+    group_by(hour) %>%
+    summarize(
+        pf_rmse = sqrt(mean(pf_error^2)),
+        pf_mae = mean(abs(pf_error)),
+        pf_mape = mean(abs(pf_error / time_until_arrival)),
+        pf_ci_cov = mean(pf_ci),
+        gtfs_rmse = sqrt(mean(gtfs_error^2)),
+        gtfs_mae = mean(abs(gtfs_error)),
+        gtfs_mape = mean(abs(gtfs_error / time_until_arrival))
+    ) %>% arrange(hour)
+
+pal <- viridis::magma(3)[2:1]
+names(pal) <- c("Particle filter", "GTFS")
+
+plot_base <- ggplot(res_timeofday, aes(hour)) +
+    theme_classic() +
+    theme(legend.position = "none") +
+    scale_colour_manual(values = pal, name = "") +
+    xlab("Time of day (hour)")
+
+plot_rmse <- plot_base +
+    geom_path(aes(y = pf_rmse, colour = "Particle filter")) +
+    geom_path(aes(y = gtfs_rmse, colour = "GTFS"), alpha = 0.5) +
+    scale_y_continuous(
+        "RMSE (s)"
+    ) +
+    theme(legend.position = "bottom")
+
+plot_mae <- plot_base +
+    geom_path(aes(y = pf_mae, colour = "Particle filter")) +
+    geom_path(aes(y = gtfs_mae, colour = "GTFS"), alpha = 0.5) +
+    scale_y_continuous(
+        "MAE (s)"
+    )
+
+plot_mape <- plot_base +
+    geom_path(aes(y = 100 * pf_mape, colour = "Particle filter")) +
+    geom_path(aes(y = 100 * gtfs_mape, colour = "GTFS"), alpha = 0.5) +
+    scale_y_continuous(
+        "MAPE (%)"
+    )
+
+plot_ci <- plot_base +
+    geom_hline(aes(yintercept = 85), lty = 2) +
+    geom_path(aes(y = 100 * pf_ci_cov, colour = "Particle filter")) +
+    scale_y_continuous(
+        name = "PICP (%)"
+    )
+
+plot_rmse + plot_mae +
+    plot_mape + plot_ci +
+    guide_area() +
+    plot_layout(
+        guides = "collect",
+        design = "
+            13
+            24
+            55
+        ",
+        heights = c(1, 1, 0.1)
+    )
+
+```

### 2020-10-11 new2020/sections/sec04.Rmd
Grab transfer results data
Last(?) of the main results

@@
 The final scenario is for a journey which requires transfering between two services, in which case we want to predict the probability that a transfer between two particular trips at the transfer stop will be successful (that is, the first bus arrives before the second).
+Given two CDFs for the arrival time of the first and second trips, $\Pr{L_1 < x}$ and $\Pr{L2 < y}$, respectively, the probability of a successfull transfer between them is
+\[
+\begin{split}
+   \mathbb{P}_{\text{transfer}} =
+   \Pr{L_1 < L_2} &=
+   \sum_{x=1}^\infty \Pr{L_1 < L_2 | L_2 = x} \Pr{L_2 = x}\\
+   &= \sum_{x=1}^\infty \Pr{L_1 < x} \Pr{L_2 = x}\\
+   &= \sum_{x=1}^\infty \Pr{L_1 < x}
+      \left[ \Pr{L_2 < x+1} - \Pr{L_2 < x} \right].
+\end{split}
+\]
+This is easily obtained from the CDFs and is a simple enough computation which could be done on a user's device.
 
-* CDF makes it possible to answer many (often complex) jouney planning questions
-* P(catch)
-* P(arrive on time)
-* P(transfer)
-* this is a simple computation - can be done client side (i.e., on a user's phone) by passing CDF (small size, as e.g., JSON)
+```{r jp_transfer_prep,echo=FALSE,message=FALSE,cache=TRUE}
+library(tidyverse)
+library(patchwork)
+load("../data/all_results_transfer.rda")
+```
+
+```{r jp_transfer,cache=TRUE,dependson="jp_transfer_prep",echo=FALSE,fig.width=8,fig.height=2.5,out.width="\\textwidth",message=FALSE,warning=FALSE,fig.cap="Predicted outcomes of journey planning process for a transfer between services."}
+
+mdata <- map(all_results, "tab") %>% bind_rows
+mdata %>%
+   mutate(
+      did_catch = ifelse(did_catch == "Y", "Transfer made", "Transfer missed"),
+      gtfs_catch = ifelse(gtfs_eta == "Y", "Yes", "No")
+   ) %>%
+   ggplot(aes(p_catch + rnorm(length(p_catch), 0, 0.01), gtfs_catch)) +
+      ggbeeswarm::geom_beeswarm(
+         groupOnX = FALSE,
+         priority = "random"
+      ) +
+      theme_classic() +
+      scale_x_continuous("Predictive probabilty of making transfer") +
+      scale_y_discrete("GTFS prediction") +
+      facet_grid(~did_catch) +
+      theme(
+         strip.background = element_blank(),
+         strip.text = element_text(hjust = 0)
+      )
+
+gtfs_catch_tab <- table(
+  gtfs = ifelse(mdata$gtfs_eta == "Y", "Yes", "No"),
+  outcome = ifelse(mdata$did_catch == "Y", "Yes", "No")
+)
+gtfs_catch_pc <-
+  round(sweep(gtfs_catch_tab, 1, rowSums(gtfs_catch_tab), "/"), 2)
+
+pf_catch_tab50 <- table(
+   pf = ifelse(mdata$p_catch > 0.5, "Yes", "No"),
+   outcome = ifelse(mdata$did_catch == "Y", "Yes", "No")
+)
+pf_catch_pc50 <-
+   round(sweep(pf_catch_tab50, 1, rowSums(pf_catch_tab50), "/"), 2)
+
+pf_catch_tab80 <- table(
+   pf = ifelse(mdata$p_catch > 0.8, "Yes", "No"),
+   outcome = ifelse(mdata$did_catch == "Y", "Yes", "No")
+)
+pf_catch_pc80 <-
+   round(sweep(pf_catch_tab80, 1, rowSums(pf_catch_tab80), "/"), 2)
+
+tab <- cbind(
+  "Predicted outcome" = c("No", "Yes"),
+  format(gtfs_catch_pc, nsmall = 2L, digits = 2L),
+  NA,
+  format(pf_catch_pc50, nsmall = 2L, digits = 2L),
+  NA,
+  format(pf_catch_pc80, nsmall = 2L, digits = 2L)
+)
+
+# roc curve?
+pc <- seq(0, 1, by = 0.05)
+roc <- lapply(pc,
+   function(p) {
+      tab <- table(
+         pf = factor(
+            ifelse(mdata$p_catch > p, "Yes", "No"),
+            levels = c("No", "Yes")),
+         outcome = factor(
+            ifelse(mdata$did_catch == "Y", "Yes", "No"),
+            levels = c("No", "Yes")
+         )
+      )
+      P <- sum(tab[,2])
+      N <- sum(tab[,1])
+      TP <- tab[2,2]
+      TN <- tab[1,1]
+      FP <- tab[2,1]
+      FN <- tab[1,2]
+      c(
+         p = p,
+         tpr = TP / P,
+         tnr = TN / N,
+         ppv = TP / (TP + FP),
+         fnr = FN / P,
+         fpr = FP / N
+      )
+   }
+) %>% bind_rows
+
+gtfs_roc <- data.frame(
+   fpr = gtfs_catch_tab[2,1] / sum(gtfs_catch_tab[,1]),
+   tpr = gtfs_catch_tab[2,2] / sum(gtfs_catch_tab[,2])
+)
+
+```
+
+```{r jp_transfer_dec,cache=TRUE,dependson="jp_transfer",echo=FALSE,fig.width=5,fig.height=2.5,out.width="0.5\\textwidth",fig.align="center",message=FALSE,warning=FALSE,fig.cap="False positive rate for varying decision level of transfer.
+Increasing the threshold for ...
+.
+The GTFS rate is fixed at about 10\\%, which corresponds to using a 37.5\\% decision threshold."}
+gp <- 0.375
+ggplot(roc, aes(p, fpr)) +
+   geom_path(lwd = 1.2) +
+   coord_cartesian(expand = FALSE) +
+   geom_segment(aes(x = gp, xend = 0, yend = fpr),
+      data = gtfs_roc, colour = "red", lty = 2
+   ) +
+   geom_segment(aes(x = gp, xend = gp, yend = 0),
+      data = gtfs_roc, colour = "red", lty = 2) +
+   xlab("Decision threshold") +
+   ylab("False positive / miss rate") +
+   theme_classic()
+```
+
+```{r jp_transfer_tab,cache=TRUE,dependson="jp_transfer",echo=FALSE,results='asis'}
+options(knitr.kable.NA = "")
+knitr::kable(
+  tab,
+  row.names = FALSE,
+  format = "latex",
+  booktabs = TRUE,
+  caption = "Prediction results for all transfer journeys, with the displayed values representing the proportion of outcomes in each cell (rows are conditioned by the prediction).",
+  caption.short = "Prediction results for all journeys",
+  escape = FALSE,
+  linesep = ""
+) %>%
+  kableExtra::add_header_above(
+    c(" ", "Schedule-delay" = 2, " ", "P > 0.5" = 2, " ", "P > 0.8" = 2)
+  ) %>%
+  kableExtra::add_header_above(
+    c("", "Observed outcome" = 8)
+  ) %>%
+  kableExtra::kable_styling(font_size = 8)
+```
+
+* PF methods allows decision making to include risk
+* GTFS is a fixed rate; cannot adapt to specifications of the journey planning application (i.e., when it's pouring with rain, want high chance of successful transfer)
+* Of course, the wait time is also important (i.e., how long waiting for the transfer?) which can be calculated as well... ?
+
+\[
+\begin{split}
+   \mathbb{E}_{\text{wait}} &= \E{L_2 - L_1 \cond L_1 < L_2} \\
+   &= \sum_{y=1}^\infty \sum_{x=0}^{y-1} \left(y - x\right)
+      \Pr{L_1 = x \cap L_2 = y} \\
+   &= \sum_{y=1}^\infty \sum_{x=0}^{y-1} \left(y - x\right)
+      \Pr{L_1 = x} \Pr{L_2 = y}
+\end{split}
+\]
+where $\Pr{A = a} = \Pr{A < a + 1} - \Pr{A < a}$.
+This lets us calculate the expected waiting time given the transfer is successful, and accounts for uncertainties in both vehicle's arrival times.
+A similar formula could be used to calcualte, for example, the variance of waiting time---again, all on a user's phone, which removes compuational demand from the server.
+
+```{r, echo=FALSE, eval = FALSE}
+N <- 1000L
+x <- pmax(1, rnorm(N, 15, 2))
+y <- pmax(1, rnorm(N, 25, 2))
+
+xr <- floor(x)
+yr <- floor(y)
+xcdf <- cumsum(table(xr)) / N
+ycdf <- cumsum(table(yr)) / N
+
+Fx <- numeric(max(xr))
+Fx[as.integer(names(xcdf))] <- xcdf
+Fy <- numeric(max(yr))
+Fy[as.integer(names(ycdf))] <- ycdf
+
+plot(Fx, type = "S", xlim = range(c(xr, yr)))
+lines(Fy, type = "S", col = "red")
+
+Px <- 0
+for (j in 2:(max(yr)-1))
+   for (i in 1:min(j-1, max(xr)-1))
+      Px <- Px + (j - i) * (Fx[i+1] - Fx[i]) * (Fy[j+1] - Fy[j])
+Px
+
+mean(ifelse(y >= x, y - x, NA), na.rm = TRUE)
+```

### 2020-10-11 new2020/sections/sec03.Rmd
load patchwork

@@
 ```{r time_results,echo=FALSE,message=FALSE,warning=FALSE,cache=TRUE,dependson="overall_results",fig.width=8,fig.height=5,out.width="\\textwidth",fig.align="center",fig.cap="Results by time of day."}
+library(patchwork)
+
 res_timeofday <- eta_talk %>%

### 2020-10-11 new2020/main.Rnw
Add main pf results
Load texalpha package for characters in bib
texTalpha pkg
Try find the bug
Remove tables fix (rss)

@@
 \email{t.lumley@auckland.ac.nz}
+
+%%%% FIX FOR TABLES
+% \usepackage{etoolbox}
+
+% \makeatletter
+% \patchcmd{\@makecaption}
+%   {\parbox}
+%   {\advance\@tempdima-\fontdimen2\font\parbox} % decrease the width!
+%   {}{}
+% \makeatother
+%%%% END FIX
 
@@
 \usepackage{cleveref}
+\usepackage{longtable}
+\usepackage{booktabs}
 
 \input{symbols.tex}
+
+\graphicspath{{sections/}}
 
@@
 <<include-sections,echo=FALSE,results="asis">>=
-for (file in list.files('sections', pattern = "*.tex", full.names = TRUE)) {
+for (file in list.files('sections', pattern = "*.tex$", full.names = TRUE)) {
   cat(sprintf("\\input{%s}", file), "\n")

### 2020-10-13 new2020/sections/sec03.Rmd
Writing up section 3

@@
 
+## Trip state
+
+The state of a trip at time $t_k$, $\Tripr_k$, includes the necessary information to input into the arrival time prediction algorithm. The state includes:
+
+* $t_k$, the time of the most recent observation associated with the trip; 
+* $\Vspeed_k$, the speed of the vehicle servicing the trip;
+* $\TripStop_k$, the index of the current or most recently stop visited;
+* $\TripDep_k$, an indicator taking the value 1 if the vehicle has departed stop $\TripStop_k$, or zero otherwise;
+* $\TripSeg_k$, the index of the current road segment; and
+* $\SegProg_k$, the proportion of the current segment travelled so far.
+
+Together these are expressed in the vector
+\[
+   \Tripr_k = \tvec{\Tript_k, \TripStop_k, \TripDep_k, \TripSeg_k, \SegProg_k}.
+\]
+
+To predict bus arrival using a particle filter, the trip state needs expressing via a sample of individual particles,
+\begin{equation}\label{eq:pf_trip_state}
+   p(\Tripr_k | \Vobs_{1:k}) \approx
+   \frac{1}{\tilde N} \sum_{i=1}^{\tilde N} \DiracMeasure{\Tripr_k\vi}{\Tripr_k}.
+\end{equation}
+In this work, we are using a particle filter to estimate vehicle state @citep:Elliott_2020;.
+However, if vehicle state was available in some other way, the first step would be to sample $N$\ particles from the posterior.
+We assume vehicle state includes at least *distance travelled*, $x_k$, and *speed*, $\dot x_k$.
+Each particle's state is easily converted into a trip state: the speed $\dot x_k\vi$ is used directly; $\TripStop_k$, $\TripSeg_k$, and $\SegProg_k$ can be calculated based on the particle's distance and static route information (node distances $\Tsegd$ and lengths $\Tseglen$); and $\TripDep_k$ is unity if the most recent observation was a departure, and zero otherwise.
+
 ## Particle filter ETAs
 
+Given a sample of $N$ particles represent a trip's state, as shown in @cref:eq:pf_trip_state;, road network state $\NWstate_c$, and optionally a travel time forecast function $\mathcal{F}(b;t,\Theta)$ [[section 2]], we can begin to iteratively predict a vehicle's arrival time at upcoming stops.
+To simplify notation, we only consider a single trip, and assume that the network state $\NWstate_c$ is a subset of the full network state containing only the segments used by the trip.
+The following iterative steps are applied independently to each particle, each of which is initialized with a *travel-time-so-far* parameter $\ttsofar\vi=0$\ seconds.
+
+#### **Step one: complete the current segment**
+
+Perhaps the most complicated step, it involves calculating the time taken to reach the next node, and depends on how far along the segment the particle is.
+If the particle is at a node ($\SegProg_k\vi = 0$), then we must allow it to remain there for a short time before continuing (using model definitions from [[section 3]]):
+
+* if the particle is at a stop node (stop index $j$),
+    * if $\TripDep_k = 1$, do nothing,
+    * else wait time is simulated from a modified dwell time distribution, \[\ttsofar\vi = \ttsofar\vi + \Istop_j\vi \left( \frac{\gamma}{2} + \pdwell_j\vi \right);\]
+* if the particle is at an intersection node (index $m$), 
+    * wait time is simulated from the intersection model, \[\ttsofar\vi = \ttsofar\vi + \Iint\vi_m\pwait\vi_m.\]
+
+However, if the particle is not at a node, then we calculate the remaining travel time along the segment.
+This depends on the speed of the particle, $\Vspeed_k\vi$, the remaining distance along the segment, and the road state.
+If the particle is nearing the end of a road segment of length $\Tseglen_m$ (less than 200\ meters remaining) we assume the particle will retain approximately the same speed until it reaches the next node.
+Conversely, if the particle is not near the end, we expect the particle will adhere more to the average speed along the road segment with average speed $\NWstate_m$, uncertainty $\NWstatevar_{mm}$, and between-vehicle variance $\NWvar_m$.
+Thus, the average speed of the particle for the remainder of segment $m = \TripSeg_k\vi$ with maximum speed $\MaxSpeed_m$ is
+\begin{equation*}
+v\vi \sim
+\begin{cases}
+   \TNormal{\Vspeed_k\vi}{\Vnoise}{0}{\MaxSpeed_m},
+      & (1-\SegProg_k)\Tseglen_m < 200, \\
+   \TNormal{\NWstate_m}{\NWstatevar_{mm}^2 + \NWvar_m^2}{0}{\MaxSpeed_m}, 
+      & \text{otherwise}.
+\end{cases}
+\end{equation*}
+
+The travel time to the end of the current segment is now added to travel-time-so-far,
+\[
+\ttsofar\vi = \ttsofar\vi + 
+   \frac{(1-\SegProg_k\vi)\Tseglen_m}{v\vi}.
+\]
+In [[figure:X]], this is the time until arrival at the node marked I.
+
+<!-- 
 In our application, the vehicle state is already represented by a set of $\Np$\ particles,
@@
 
-Note: need to add current speed to trip state ???
+Note: need to add current speed to trip state ??? -->
 
-**Step 1: complete the current segment**
+<!-- **Step 1: complete the current segment**
 
@@
 \]
-We now store an iterative variable *travel-time-so-far* with the time taken to reach the end of the current segment, $\ttsofar\vi = \tilde z\vi$.
+We now store an iterative variable *travel-time-so-far* with the time taken to reach the end of the current segment, $\ttsofar\vi = \tilde z\vi$. -->
 

### 2020-10-13 new2020/sections/sec03.Rmd
pf prediction steps

@@
 
-**Step 2: compute arrival time at next stop**
+#### **Step 2: compute arrival time at next stop**
 
-If there are any remaining segments between the current segment $\TripSeg_k\vi$ and the next stop $j_k\vi + 1$, their travel times are sampled from the network state with variance increased by network system noise to account for forecasting changes.
+If the next node is an intersection, as is the case in [[figure X]], we iteratively calculate travel time along each segment until the particle arrives at a stop.
+As before we simulate vehicle speed from the network state, but now allowing for increased variability over time.
+Here, a forecast function could be used if available, otherwise a simpler approach can be used, as we have here.
+We use a similar distribution as in step\ 1, but allow speed to increase over time with a rate of $q$\ m/s$^2$, up to a maximum of $P_\text{max}$, which is the overall variance of vehicle speeds based on historical data.
+The predicted particle speed is thus
 \[
-   v_j\vi \sim
-   \TNormal{\hat \NWstate_j}{\left[(\RouteNWstatevarseg_j + \ttsofar\vi\NWnoise)^2 + \NWvar_j^2\right]\wedge P_{\text{max}}}{0}{\MaxSpeed_j},
+   v_m\vi \sim
+   \TNormal{\hat \NWstate_m}{\left[(\NWstatevar_{mm} + \ttsofar\vi\NWnoise)^2 + \NWvar_m^2\right]\wedge P_{\text{max}}}{0}{\MaxSpeed_m}.
 \]
-allowing uncertainty to increase up to a maximum $P_\text{max}$ which is the overall variance of all speeds along all segments estimated from historical data.
-The travel time along each segment is added to the travel-time-so-far,
+Travel-time-so-far is increased by the particle's travel time,
 \[
-   \ttsofar\vi = \ttsofar\vi + \frac{\Tseglen_j}{v_j\vi}.
+   \ttsofar\vi = \ttsofar\vi + \frac{\Tseglen_m}{v_m\vi}.
 \]
 
-Now, arrival time at stop $j$ is the sum of travel time along all segments up to the next stop. In the iterative process, this is simply $\Tarr_j\vi = t_k + \ttsofar\vi$.
+This process is repeated until the particle arrives at a stop, as is indicated by III in [[figureX]]. The arrival time at the stop is then stored as
+\[
+\Tarr_m = t_k + \ttsofar\vi.
+\]
 
-**Step 3: compute stop dwell time**
+#### **Step 3: stop dwell time**
 
-Once at a stop, a bus either stops and waits while passengers board and disembark, or travels past without stopping.
+When at a stop, the particle first determines its dwell time.
+Using the model parameters described in \cref{particle-filters-estimating-vehicle-state}, the dwell time at stop $j$ is added to the travel-time-so-far,
+\begin{equation*}
+\begin{split}
+   p_j\vi &\sim \Bern{\pi_j},\\
+   d_j\vi &\sim \TNormal{\dwell_j}{\dwellvar_j}{0}{\MaxSpeed_j},\\
+   \ttsofar\vi &= \ttsofar\vi + p_j\vi(\gamma + d_j\vi).
+\end{split}
+\end{equation*}
+
+However, we also need to consider *layovers*, which are stops at which the bus is expected to wait if it is running early.
+Adherence is shown in [[figureY]], which shows about 60% driver compliance.
+So, with probability 0.6, the particle will wait until at least the scheduled departure time at stop $j$, $D_j$:
+\[
+   \ttsofar\vi =
+   \begin{cases}
+      \max\{\ttsofar\vi, D_j - t_k\} & \text{w.p. } 0.6,\\
+      \ttsofar\vi & \text{otherwise}.
+   \end{cases}
+\]
+
+<!-- Once at a stop, a bus either stops and waits while passengers board and disembark, or travels past without stopping.
 This *dwell time* needs to be modelled @citep:cn;, but adds multimodality to the arrival time distribution.
@@
 \]
-unless the stop is a layover. In that case, we allow the particle---with some probability---to remain at the stop until the end of the dwell time, or the scheduled departure time, whichever is later.
+unless the stop is a layover.
+In that case, we allow the particle---with some probability---to remain at the stop until the end of the dwell time, or the scheduled departure time, whichever is later.
+-->
 
-**Step 4: repeat steps 2--3**
+#### **Step 4: repeat steps 2--3**
 
-We now repeat each of the steps, sampling travel times for intermediary road segments as the particle arrives, and storing stop arrival times, until the particle reaches the end of the route.
+Next, the particle continues along the route, sampling travel times (step 2) for each road segment, and dwell times (step 3) for each stop, while recording its arrival time at each stop along the way until it reaches the end of the route.
 
-**Step 5: obtain arrival time distributions**
+#### **Step 5: obtain arrival time distributions**
 
-Since each particle stores its arrival at all stops, we can very easily obtain the predictive distribution of arrival time for the bus using the delta measure,
+The predictive distribution of arrival time at stop $j$, given trip and network states at time $t_k$, is easily obtained using the delta measure (section XX),
 \[
-   p(\Tarr_j | \Vstate_k, \vec\NWstate_k) \approx
+   p(\Tarr_j | \Tripr_k, \vec\NWstate_k) \approx
    \frac{1}{\tilde N}\sum_{i=1}^{\tilde N}
@@
 \]
-We may obtain any summary statistics from this distribution, including the mean and prediction intervals.
+Means, quantiles, and other summary statistics can easily be obtained from the sample; however, some, such as quantiles, require sorting all $\tilde N$ particles.
+In \cref{chapter4} I present an approximation which avoids sorting the particles which is a computationally intensive procedure.
 

### 2020-10-19 new2020/sections/sec03.Rmd
some typing

@@
 
-To assess the feasibility and reliability of this approach, we ran the full model on one day of data from DATE.
+The particle filter vehicle model and Kalman filter network model described by @citet:Elliott_2020; were combined with the particle filter prediction model described in \cref{particle-filter-etas} and applied to a full day of data from [[date]].
+For each prediction of arrival time at each stop along all routes, we stored the mean along with the 5 and 90% quantiles, which could be compared to the bus's actual arrival time at the end of the day.
+We also computed the schedule-delay prediction using the real-time trip updates to compare our results to the status quo.
+
+To assess and compare predictive performance, we use RMSE, MAE, MAPE, and PICP. For random variables $\{X_n: n=1,\ldots,M\}$ with estimates $\{\hat X_n: n=1,\ldots,M\}$,
+\begin{equation*}
+\begin{split}
+\text{RMSE} &= \sqrt{\frac{1}{M} \sum_{n=1}^M \left(\hat X_n - X_n\right)^2},\\
+\text{MAE} &= \frac{1}{M}\sum_{n=1}^M \left|\hat X_n - X_n\right|,\\
+\text{MAPE} &= \frac{1}{M}\sum_{n=1}^M \frac{\left|\hat X_n - X_n\right|}{X_n}.
+\end{split}
+\end{equation*}
+For prediction intervals $\{Q_n = (X_{na},X_{nb}):n=1,\ldots,M\}$, the indicator variable $I_n$ is 1 if $X\in Q_n$, and zero otherwise. Thus,
+\[
+\text{PICP} = \frac{1}{M} \sum_{i=1}^M I_n.
+\]
+[[ copy this definition from thesis ...]].
+
+RMSE and MAE each evaluate the accuracy of the predictor, with MAE using absolute error is less influenced by extreme prediction errors.
+MAPE evaluates the proportional error of the predictor, which is appropriate for arrival time prediction as we expect errors to increase for stops farther away.
+Finally, PICP examines how well the model incorporates the underlying arrival time uncertainty of the model: we expect the PICP for an 85% prediction interval to be approximately 0.85.
+This value is not available for the schedule-delay predictor which only provides point estimates of arrival time.
+
+<!-- To assess the feasibility and reliability of this approach, we ran the full model on one day of data from DATE.
 This used the particle filter and network model described by @cite:Elliott_2020;, and the arrival time predictions described above.
@@
 However, only 77% of prediction intervals contain the actual arrival time, indicating that uncertainties are being under-predicted (the nominal coverage is 85%).
+-->
 

### 2020-10-20 new2020/sections/sec02.Rmd
More results + finally figure out there's a bug in the statsoc class re tables...
ch2 words

@@
 
-**most of this section will be removed/cut-down to just the basic concepts**
+## Transit data, GTFS, and a road network
 
-## GTFS static and real-time
-@label:gtfs;
+The general structure of transit data has remained fairly standard, tracing back to an initial prescription by @citet:Cathey_2003;.
+This work introduced terms such as *trip* and *route*, which are continually used today.
+However, the stucture of the data was massively formalised in 2006 when Google released their first version of *General Transit Feed Specification (GTFS)* @citep:GoogleDevelopers_2006;.
 
+Transit data requires information about *routes*, journeys from an origin to a destination servicing a sequence of *stops* where passengers may board or disembark, and *trips* which are time-specific instances of a route.
+The path or *shape* travelled by the vehicle is described by a sequence of GPS coordinates.
+This information is contained within *GTFS static* along with other information one might find on a printed timetable, such as scheduled arrival and departures.
+
+Real-time transit data, on the other hand, includes information about the position of a vehicle.
+This might be in the form of a *distance travelled* along the route @citep:Wall_1999,Dailey_2001; or, more commonly in modern systems, as GPS coordinates (latitude and longitude) @citep:Cathey_2003,Hans_2015;.
+The structure and format of this information is described by *GTFS-realtime*.
+
+<!--
 The structure of public transport systems is very similar across the globe, so much so that Google developed GTFS @citep:GoogleDevelopers_2006; to standardise the way transport agencies organise and distribute transit data.
@@
 For this reason, we opted to develop our application framework based solely on GTFS data, as described next.
+-->
 
 ## A transit road network
-@label:network;
 
@@
 ## Particle filters estimating vehicle state
-@label:pf;
 
@@
 ## Estimating network state
-@label:state_estimates;
 

### 2020-10-21 new2020/sections/sec02.Rmd
Some sec2 words

@@
 
+An important aspect missing from the setup, however, is information about *physical interactions* between separate routes (which may share common roads) which is critical to accurate prediction @citep:Jeong_2005,;.
+@citet:Elliott_2020; describe a simple process for combining route information from static GTFS data to obtain a *transit road network*, expressing each route as a sequence of physical roads.
+The benefit of this is that vehicles servicing different routes but using the same road(s) can share information about their speed.
+This greatly enhances the amount of data available for forecasting arrival times.
+
+```{r road_nw,echo=FALSE}
+# figure for the network
+```
+
+In the approach, stops are considered *nodes* in the network, while the roads between them are *edges*.
+Any buses travelling between the same two (or more) stops share the road between them, thus providing information about the speeds.
+This increases the rate of update and is particularly beneficial for predicting arrival times for low-frequency routes (see \cref{predicting-arrival-time}).
+
 <!--
@@
 
-## A transit road network
+<!-- ## A transit road network
 
@@
 This can in turn be used, optionally with a forecasting function, to predict a future travel time of a bus along the road, which is used in arrival time prediction.
+-->
 
-## Particle filters estimating vehicle state
+## Recursive Bayes and particle filtering
+
+In a real-time application, performing a full update of the posterior distribution of some parameters becomes hugely inefficient.
+It is possible to recursively update the posterior, however, by way of recursive Bayesian filtering (alternatively called sequential Monte Carlo).
+Given a random variable $\vec x = \vec x_{0:k} = \tvec{\vec x_0,\vec x_1,\ldots\,\vec x_k}$ at time $t_k$ following a Markov process with associated observation $\vec y = \vec y_{1:k} = \tvec{\vec y_1,\ldots,\vec y_k}$, the posterior distribution at time $t_k$ is expressed via the recursive formula
+\[
+p(\vec x_{0:k}\cond{} \vec y_{1:k}) \propto
+   p(\vec x_{0:k-1}\cond{} \vec y_{1:k-1})
+   p(\vec x_k\cond{}\vec x_{k-1})
+   p(\vec y_k\cond{}\vec x_k)
+\]
+which involves the previous state, next state prediction, and the likelihood, respectively.
+We therefore need to compute $p(\vec x_k\cond{}\vec x_{k-1})$, which involves the *transition function* between consecutive states, and the likelihood $p(\vec y_k\cond{}\vec x_k)$.
+The specific details of each of these depends on the model, which is discussed in \cref{estimating-road-state-from-vehicle-observations}.
+
+Estimation of recursive Bayesian models takes many forms, most notably the Kalman filter @citep:; which has been widely used in transit modelling @citep:Wall_1999,Dailey_2001,Cathey_2003,Jeong_2005;.
+However, as discussed by @citet:Hans_2015,Elliott_2020;, it makes several assumptions which do not work so well with transit vehicles, namely that the errors are normal with a single mode.
+The particle filter @citep:Gordon_1993; provides a numerical approach to solving recursive Bayesian models, much the same as Markov chain Monte Carlo provides estimates of standard posterior distributions.
+In a particle filter, the state $\vec x_k$ is represented by a sample of $N$\ particles $\{\vec x_k\vi\}_{i=1}^N$ with associated weights $\{w_k\vi\}_{i=1}^N$.
+We use the delta measure $\delta$ to express the approximated state
+\[
+   p(\vec x_{k-1} | \vec y_{0:k-1}) \approx \sum_{i=1}^N w_{k-1}^{(i)} \DiracMeasure{\vec x_{k-1}^{(i)}}{\vec x_{k-1}}.
+\]
+
+Particle filter estimates are updated in two steps.
+First the state prediction of $p(\vec x_k\cond{}\vec x_{k-1})$ is made using the transition function $f$, which implements the vehicle model (\cref{vehicle-model}):
+\[
+   p(\vec x_{k} | \vec x_{k-1}) \approx
+      \sum_{i=1}^N w_{k-1}^{(i)}
+         \DiracMeasure{\vec x_{k}^{(i)}}{\vec x_{k}},\quad
+   \vec x_k\vi = f(\vec x_{k-1}\vi, \Delta_k, \sigma).
+\]
+
+Second is the likelihood update, which quantifies the probability of the observed vehicle position given each of the particle predictions.
+The particle weights are updated according to their likelihood,
+\[
+w_k\vi =
+   \frac{w_{k-1}\vi p(\vec y_k\cond{}\vec x_k\vi)}{
+      \sum_{j=1}^N w_{k-1}\vi[j] p(\vec y_k\cond{}\vec x_k\vi[j])
+   }.
+\]
+For the likelihood function itself, we use an exponential distribution on the geographic distance between the particle and the observed location, as proposed by @citet:Elliott_2020;.
+
+To ensure the *effective sample size* remains adequately large (greater than $\frac{1}{4}N)$, importance resampling of the particles, with replacemnet, is performed whenever the effective sample size $N_{\text{eff}} = \frac{1}{\sum_{i=1}^N (w_k\vi)^2}$ drops below the threshold.
+The particle filter was for this reason originally called the *bootstrap filter* @citep:Gordon_1993;.
+
+## Estimating road state from vehicle observations
+
+In a transit setup, we observe vehicle locations $\vec y_k$ consisting of the vehicle's latitude and longitude at time $t_k$.
+The goal is to estimate the vehicle's distance travelled $x$ and speed $\dot x$, from which we can calculate its average speed along individual road segments.
+The underlying, unobservable state of the vehicle is $\vec x_k = \tvec{x_k,\dots x_k}$.
+
+Estimation of network state involves three steps: vehicle state estimation, estimation of average speed along roads, and updating the network state model.
+
+### Vehicle model
+
+Estimating vehicle state requires a vehicle model consisting of several key behaviours.
+@cite:Elliott_2020; give details, however of particular importance to arrival time prediction is the *dwell time* model for stop behaviour.
+When the bus arrives at a stop, it can either drive past, or stop and allow passengers to board and disembark.
+
+<!--
 
@@
    \sum_{i=1}^N w_{k}^{(i)} \DiracMeasure{b_{\ell}^{(i)}}{b_{\ell}}.
-\]
+\] -->
 
+<!--
 ## Estimating network state
@@
 
-If a forecast function is available, we could also forecast future travel time based on current state and historical trends, and this would be incorporated.
+If a forecast function is available, we could also forecast future travel time based on current state and historical trends, and this would be incorporated. -->

### 2020-10-22 new2020/sections/sec02.Rmd
Vehicle model bits

@@
 This increases the rate of update and is particularly beneficial for predicting arrival times for low-frequency routes (see \cref{predicting-arrival-time}).
-
-<!--
-The structure of public transport systems is very similar across the globe, so much so that Google developed GTFS @citep:GoogleDevelopers_2006; to standardise the way transport agencies organise and distribute transit data.
-GTFS consists of two core components: *static*, which contains pre-defined information, such as routes and stops, and *real-time*, which contains vehicle location and other data collected in real-time.
-
-Static GTFS contains *structural* data, descibing information one might observe on a printed timetable.
-This includes route information about trips (instances of a route at a specific time), the geographical path the vehicle takes, and the stops it services.
-This provides the basis of transit data, and plays a key role in developing the aforementioned *transit network*.
-
-Real-time GTFS descibes how live data, such as the location of a bus, or its time of arrival at or departure from a bus stop, is formatted.
-Vehicle locations allow users to track the progress of the vehicle along its route, and are often updated every 10--30\ seconds.
-In some implementations, such as in Auckland, they are also triggered by events such as arrival at a bus stop or (major) intersection.
-Arrival and departure events, on the other hand, are triggered whenever the bus arrives at or departs from a bus stop (or at least thinks it does), and include the time of the event and, if scheduled arrival times are available, the *delay* between scheduled and actual arrival (or departure).
-
-A huge benefit of GTFS being adopted widely around the world is that developers can easily access transit feeds for different providers in different cities and countries using the same structure.
-Any application developed using (only) GTFS information can, technically, be redeployed to any other system also using GTFS.
-For this reason, we opted to develop our application framework based solely on GTFS data, as described next.
--->
-
-<!-- ## A transit road network
-
-One important predictor of bus arrival is travel time @citep:Shalaby_2004;.
-However, previous research has focused on situations where either only a single route is considered, or else different routes and combined by manually locating common road segments.
-This is, of course, not easy or even possible to automate.
-@cite:Elliott_2020; proposed a method of developing a transit road network from only GTFS data by identifying *nodes* (bus stops and intersections) and the *edges* (roads) between them.
-
-Here, we use the simplest version of this method, which places nodes at bus stops.
-Thus, any two routes that share the same sub-sequence of bus stops are assumed to follow the same path between them (this is almost always the case).
-By expressing the route as a series of nodes and road segments, we can begin to share information across routes.
-
-The goal is to model vehicles as they travel along their respective routes, estimating their average speed along each road segment.
-Having traversed a road segment, the data is passed to the network to update the real-time traffic state.
-This can in turn be used, optionally with a forecasting function, to predict a future travel time of a bus along the road, which is used in arrival time prediction.
--->
 
@@
 When the bus arrives at a stop, it can either drive past, or stop and allow passengers to board and disembark.
+The model presented here is based directly off that described by @cite:Elliott_2020;, which is itself based of the model described by @cite:Hans_2015;.
+
+When a bus approaches a bus stop, it stops with probability $\pi$ to pick up and drop off passengers.
+If there are no passengers, the vehicle does not stop and loses no time.
+However, if it does stop, the time lost doing so is referred to as *dwell time*.
+Dwell time is composed of two parts:
+
+1. deceleration, doors open and close, acceleration; and
+2. servicing passengers
+
+The first is assumed constant for all stops, and is parameterised as $\gamma$, the *minimum dwell time*.
+The stop service time is given a distribution for each stop.
+The mean and variance of dwell time at stop $j$, estimated from historical data, are $\tau_j$ and $\omega_j$, respectively.
+The distribution is assumed to follow a (truncated) normal distribution,
+\[
+d_j \sim \TNormal{\tau_j}{\omega_j^2}{0}{\infty}.
+\]
+The total dwell time for a particle is
+\[
+z_j\vi = p\vi\left(\gamma + d_j\vi \right),\quad
+p\vi \sim \Bern{\pi}.
+\]
+The dwell time model will be used in the arrival time prediction framework described in [[section 3]].
 
 <!--
-
-Estimation of vehicle state in real-time commonly uses *recursive Bayesian estimation*, in which the posterior becomes the prior for the next iteration.
-Particle filters are a powerful method of estimating successive states, as they make few assumptions and, most importantly, cover a large range of possible trajectories, even when there is a high degree of multimodality as is the case with transit data (e.g., around bus stops).
-Here, I give a brief overview of the particle filter which was used by @citet:Elliott_2020; to estimate vehicle state.
-
-In a recurisive Bayesian model, the underlying state we are interested in is denoted $\vec{x} = \vec x_{0:k} = \left[\vec x_0, \vec x_1, \ldots, \vec x_k\right]$.
-It can rarely be observed directly, and instead we obtain measurements $\vec y = \vec y_{1:k} = \left[\vec y_1, \ldots, \vec y_k\right]$.
-In a particle filter @citep:Gordon_1993;, this state is represented by a weighted sample of *points*,
-and makes use of the delta measure $\delta$,
-\[
-   p(\vec x_{k-1} | \vec y_{0:k-1}) \approx \sum_{i=1}^N w_{k-1}^{(i)} \DiracMeasure{\vec x_{k-1}^{(i)}}{\vec x_{k-1}}.
-\]
-@citet:Elliott_2020; describe the particle filter in detail, with discussion of its use for modelling transit vehicles in real-time.
-However, here we only discuss the merits of the particle filter, and need no more details than given already.
-
-The main advantage of the particle filter is, as mentioned, its ability to cope with a wide range of plausible, discontinuous trajectories.
-This often happens in transit data as the bus passes bus stops at which it may or may not stop.
-It's main downside is the computational demand, since we need 1000s of particles in memory for each vehicle, and perform transitions and updates to each of them.
-However, @citet:Elliott_2020; have shown that, so long as it is programmed efficiently, the particle filter can run fast enough to be feasible in a real-time setting, at least for updating vehicle states.
-
-In the next section, we make use of the flexible nature of the particle filter in that each particle is transitioned independently, making it possible to account for a wide variety of complex behaviours.
-These most notably include bus stops, which introduce a lot of uncertainty, but also layovers, which can result in a wide temporal gap between possible arrival times.
-
-The particle filter allows us to estimate the state of a vehicle and, most importantly, it's average speed along road segments:
-\[
-   p(b_\ell | \vec y_{1:k}) \approx
-   \sum_{i=1}^N w_{k}^{(i)} \DiracMeasure{b_{\ell}^{(i)}}{b_{\ell}}.
-\] -->
 

### 2020-10-26 new2020/sections/sec02.Rmd
Chapter 2 wordss

@@
 
+### Travel time estimation
+
+The goal of the vehicle model is to estimate not only their real-time state (including speed), but also to estimate their *average speed along individual road segments*.
+How this is done will depend upon the estimation technique being used, and in some cases will not be possible.
+The particle filter, however, enables easy estimation of many additional parameters, including average speed along roads @citep:Elliott_2020;.
+
+For each particle, the average speed (m/s) is calculated by first computing the *travel time* (seconds) along each road segment.
+To do so, the application stores each particle's start and end times, $T_{\ell}$ and $T'_{\ell}$, respectively, along segment $\ell$ of the route.
+This is easily accomplished due to the particle filter's iterative nature.
+The average speed of a particle along a road segment of length $\Tseglen_\ell$ is simply
+\[
+\Vttobs_\ell\vi =
+   \frac{\Tseglen_\ell}{{T'}_\ell\vi - T_\ell\vi}.
+\]
+
+As with the vehicle state, the vehicle's posterior distribution of average speed is estimated from the particles using the delta measure,
+\[
+p(\Vttobs_\ell\cond{}\Vstate_{0:k},\Vobs_{1:k}) \approx
+   \sum_{i=1}^N w_k\vi \DiracMeasure{\Vttobs_\ell}{\Vttobs_\ell\vi},
+\]
+from which the mean $\bar\NWobs_\ell$ and variance $\NWerr_\ell^2$ may be computed:
+\begin{equation*}
+\begin{split}
+\bar\NWobs_\ell &=
+   \E{\Vttobs_\ell\cond{}\Vstate_{0:k},\Vobs_{1:k}} =
+   \sum_{i=1}^N w_k\vi \Vttobs_\ell\vi,\\
+\NWerr_\ell^2 &=
+   \Var{\Vttobs_\ell\cond{}\Vstate_{0:k},\Vobs_{1:k}} =
+   \sum_{i=1}^N w_k\vi \left(\Vttobs_\ell - \bar\NWobs_\ell\vi\right)^2.
+\end{split}
+\end{equation*}
+These can be used in the next step to update the network state.
+
+### Network state update
+
+The network state model described by @citet:Elliott_2020; expresses the vehicle speed along roads using a hierarchical structure consisting of: the underlying, average speed $\RouteNWstateseg_{\ellc}$ of vehicles travelling along a road $\ell$ at time $c$ with uncertainty $\NWstatevar_{\ellc}$; the variance $\NWvar_\ell$ between individual vehicles $m\in V_m$ with speed $\Vtt_{\ellc}^m$; and the measurement error $\NWerr_\ell^m$ of each observation $\bar\NWobs_\ell^m$.
+Thus, for a road segment with maximum speed $\MaxSpeed_\ell$,
+\begin{equation*}
+\begin{split}
+\Vtt_{\ellc}^m &\sim
+   \TNormal{\NWstate_{\ellc}}{\NWstatevar_{\ellc} + \NWvar_\ell^2}{0}{\MaxSpeed_\ell}, \\
+\Vttobs_{\ellc}^m &\sim
+   \Normal{\Vtt_{\ellc}^m}{\NWerr_\ell^m}.
+\end{split}
+\end{equation*}
+
+The network state update need therefore account for multiple observations during any single update, due to the possiblility of multiple buses completing travel along a road segment at approximately the same time.
+@citet:Elliott_2020; used the information filter variant of the Kalman filter to update the network state, making the inclusion of multiple sources of information a simple addition of information to the current state.
+The end result is an estimate of real-time traffic conditions along every road within the network, updated whenever a vehicle servicing *any* route travels along it.
+
+The important component here is the multiple sources of error: namely the uncertainty of the state at time $t_c$, $\NWstatevar$, which increases over time with a rate of $\NWnoise$\ m/s/s; and the between-vehicle uncertainty $\NWvar_\ell$.
+The additional noise parameters $\NWnoise$ and $\NWvar_\ell$ are calculated from historical data using JAGS @citep:jags;.
+
+We now have everything we need to predict arrival times based on real-time vehicle and network states, accounting for uncertainties from traffic and bus stop dwell times.
+
 <!--

### 2020-11-08 new2020/sections/sec02.Rmd
remove old

@@
 We now have everything we need to predict arrival times based on real-time vehicle and network states, accounting for uncertainties from traffic and bus stop dwell times.
-
-<!--
-
-<!--
-## Estimating network state
-
-Having obtained a estimate of the speed of each vehicle along a road, we can proceed to updating the state of the road network itself.
-This uses a Kalman filter to update the state, $\vec \beta_c$, using the estimated vehicle speeds.
-
-This all provides a flexible method for modelling real-time traffic conditions in any network using only GTFS, and not rely on other data sources (for example taxis or video cameras).
-
-We use a hierarchical model on vehicle travel times, as individual vehicles have different average speeds, even if they travel at the same time, due to different driver behavior and other factors.
-The between-vehicle noise parameter $\NWvar_\ell$ describes this, and is estimated from historical data.
-
-The observations are also uncertain, as they are estimated from the particle filter.
-We can assume between-vehicle and measurement uncertainties are independent, so the total variance of observations is
-\[ \Var{b_\ell | \vec y_{1:k}} = \NWerr_\ell^2 + \NWvar_\ell^2. \]
-The system noise $\NWnoise$ is also estimated from historical data.
-Together we can quickly update the network state using the information variance of the Kalman filter (since this allows data from multiple sources simultaneously, which can happen when two or more buses are travelling one behind the other).
-
-The result is an estimate of the state of the network at time $t_c$,
-\[
-   p(\NWstate_c | \cup_m \cup_\ell\ \NWobs_{\ell,c,m}) \sim
-   \Normal{\vec\NWstate_\ell}{\NWstatevar^2}.
-\]
-This can be used to predict the travel time of buses.
-
-If a forecast function is available, we could also forecast future travel time based on current state and historical trends, and this would be incorporated. -->

### 2020-12-01 new2020/sections/sec03.Rmd

@@
 
+The overall results are shown in @cref:tab:overall_results;.
+This shows that the particle filter estimates have smaller error on average, but only slightly; with the biggest difference being in MAPE, suggesting that the particle filter is more accurate in proportion to the actual time-until-arrival.
+That is, when the bus is 10\ minutes away, the error is about 2\ minutes versus the schedule-delay's 3\ minute error; when the bus is 30\ minutes away these errors are 6\ minutes and 9\ minutes, respectively.
+The particle filter does appear to be underestimating uncertainty, as the 85% prediction interval has a coverage of only 77%.
+
 ```{r time_results,echo=FALSE,message=FALSE,warning=FALSE,cache=TRUE,dependson="overall_results",fig.width=8,fig.height=5,out.width="\\textwidth",fig.align="center",fig.cap="Results by time of day."}
@@
 ```
+
+We now calculate the four summary statistics by time of day in 15\ minute intervals, shown in @cref:fig:time_results;, where we see a different pattern.
+During peak-time, the particle filter has a larger error than the schedule-delay predictions according to RMSE and MAE, but consistently performs better in relative terms (MAPE).
+This implies that, during peak times when traffic congestion changes quickly, the particle filter has trouble forecasting arrival time accurately (our model does not yet account for peak time effects).
+The schedule-delay errors are proportionally about the same throughout the day.
+During daytime off-peak, however, the particle filter is consistently better at prediction arrival times under all metrics.
+
+The most poignant feature is, however, PICP: during daytime off-peak, traffic conditions are stable and uncertainty is modelled well by the particle filter, with the 85% prediction interval having very close to 85% coverage.
+This suggests that the underlying model is accurately modelling traffic conditions, allowing it to provide a useful prediction interval for the bus's arrival time.
+Since we have not yet implemented a forecast function, it is not surprising that the model fails under peak time when traffic conditions generally get worse and better in a short time frame.

### 2020-12-01 new2020/sections/sec02.Rmd
fix missing cite

@@
 
-An important aspect missing from the setup, however, is information about *physical interactions* between separate routes (which may share common roads) which is critical to accurate prediction @citep:Jeong_2005,;.
+An important aspect missing from the setup, however, is information about *physical interactions* between separate routes (which may share common roads) which is critical to accurate prediction @citep:Jeong_2005;.
 @citet:Elliott_2020; describe a simple process for combining route information from static GTFS data to obtain a *transit road network*, expressing each route as a sequence of physical roads.
@@
 
-Estimation of recursive Bayesian models takes many forms, most notably the Kalman filter @citep:; which has been widely used in transit modelling @citep:Wall_1999,Dailey_2001,Cathey_2003,Jeong_2005;.
+Estimation of recursive Bayesian models takes many forms, most notably the Kalman filter @citep:kf; which has been widely used in transit modelling @citep:Wall_1999,Dailey_2001,Cathey_2003,Jeong_2005;.
 However, as discussed by @citet:Hans_2015,Elliott_2020;, it makes several assumptions which do not work so well with transit vehicles, namely that the errors are normal with a single mode.

### 2020-12-02 new2020/sections/sec01.Rmd
intro

--- 2020-07-27 new2020/sections/sec01.Rmd
@@
 
-```{=latex}
-\begin{equation}
-\label{eq:no}
-y = 2x
-\end{equation}
-```
+Public transport is an essential service for any large city, and key to making it successful is its reliability.
+While extreme and costly measures exist---such as building new infrastructure---a simpler, cost-effective approach is to simply provide reliable real-time information.
+This makes it easier for commuters to use and should hopefully improve ridership.
 
-Predicting bus arrival is ... quite something \( \cite{Elliott_2020} \)
+Predicting bus arrival times is nothing new---indeed, many unique methods have been presented and deployed around the world.
+The difficulty remains, however, in accounting for real-time traffic conditions in a reliable and generalised way.
 
-I can just type things here, like ... yeah!
-```{=latex}
-\begin{equation}
-\label{eq:hello}
-x = \int 2x\; \mathrm{d}x
-\end{equation}
-```
+@citet:Dailey_2001; used historical travel time data to predict a vehicle's arrival time based on its current location, while @citet:Jeong_2005; used a Kalman filter to model real-time travel times along links of a route.
+Other examples, such as @citet:Yu_2006,Yu_2010,Yu_2011; have used machine learning models and some *route pooling* to estimate traffic conditions along specific high-flow routes.
+Even more recently, taxi data was used by @citet:cn; to model real-time traffic, which was then used to predict arrival times.
+The concensus in the literature is that reliable predicts depend on real-time traffic estimates, however the methods used to do so have been varied and not generally applicable to services outside the target location.
 
-We see in `\cref{eq:hello}`{=latex} that ... blah.
+In many cities, including out test locatio of Auckland, New Zealand, only have access to real-time transit vehicle location data, but no means of observing traffic conditions in real-time.
+@citet:Elliott_2020; proposed a simple method of constructing a *transit network* from GTFS @citep:GoogleDevelopers_2006; data, and using this along with real-time data to estimate real-time traffic state.
+In their paper, @citeauthor:Elliott_2020; used a particle filter to model a transit vehicle, as this has been shown @citep:Hans_2015; to be an effective method of estimating vehicle states given a wide range of plausible trajectories.
+
+In this paper, we build on the work of @citet:Elliott_2020; and use the particle filter to estimate bus arrival times at stops using a real-time network state estimated from the buses themselves.
+Since the particle filter results in a large sample of estimates, we also propose a simplification of the CDF that can easily be distributed to users and used in journey planning applications.
+
+<!--
+* the problem of predicting bus arrival - it needs to use real-time traffic information @cite:cn;
+* however, many deployed methods are either specific to a provider/city, or don't make use of real-time data (only vehicle position and/or arrival delays, e.g., in Auckland)
+* since the only logical source of "traffic data" in this setting is the transit vehicles themselves, makes sense to develop framework that uses them to estimate real-time network state @cite:Elliott_2020;
+* in this example, particle filter is used to obtain a sample of points from the arrival time distribution - many many points, which cannot possibly be distributed or stored efficiently
+* we propose a method of reducing this to a simple discrete CDF of arrival time (in minutes)
+* we show that these results can be used to answer some common journey planning questions, in real-time, which outperform current "method"
+
+Some of the references include:
+
+* @cite:Cathey_2003;
+* @cite:Yu_2006,Yu_2010,Yu_2011;
+* @cite:Hans_2015;
+-->

### 2020-12-02 new2020/sections/sec02.Rmd
opening of ch2

@@
 
-Before describing the process of obtaining arrival time distributions, we must first define the framework with which we obtain vehicle and network state estimates.
-@cite:Elliott_2020; present a process for constructing a transit road network from raw GTFS data.
+Reliable prediction of bus arrival times requires up-to-date or *real-time* information about current traffic conditions @citep:Jeong_2005,cn;.
+This is, however, rarely available in the context of public transport data.
+@citet:Elliott_2020; present a general method of using static and real-time GTFS data to estimate vehicle speeds along roads as a proxy for estimating traffic conditions.
+Their work provides the basic for this current work, which uses the real-time network state to make predictions of bus arrival which account for traffic conditions.
 
-## Transit data, GTFS, and a road network
+In this section, we provide a review of the work by @cite:Elliott_2020;, and give some extra details on another major source of uncertainty: intermediary bus stops.
+Finally, we introduce the particle filter, describe its implementation, and discuss why it proves beneficial for predicting vehicle arrival times reliably.
+
+## Estimating real-time road state
+
+It is well documented in the transit research literature that one of the key pieces of information requied for accurate arrival time prediction is *real-time traffic conditions* @citep:cn;.
+Unfortunately, this information is largely difficult or impossible to obtain, making globally applicable prediction methods hard to come by.
+However, there does exist a standard for transit data, GTFS @citep:GoogleDevelopers_2006;, wihch is used globally---a prediction system derived from GTFS would thus be useful on a worldwide scale.
+
+The general process implemented by @citet:Elliott_2020; consists of three components.
+The first involves using the available GTFS information to construct a *transit network* consisting of *nodes* (stops or intersections) and *edges* (the roads connecting them).
+The key point is that sections of each route can be mapped to their physical road, allowing traffic information to be obtained *independent* of the route.
+Next vehicles are modelled as they travel along their respective routes using real-time GTFS data, from which we estimate their *average speed along each road*.
+Finally the individual,likely noisy, speed 'observations' are used to update the underlying *road state*, which is what we will be using in \cref{predicting-arrival-time} to predict arrival time.
+
+### GTFS network
+
+TO DO: trim this down
 
@@
 This increases the rate of update and is particularly beneficial for predicting arrival times for low-frequency routes (see \cref{predicting-arrival-time}).
+[mentioned in subsection intro]
 
-## Recursive Bayes and particle filtering
+### Particle filter estimates of road speeds
+
+* introduce the recursive estimation of states in real-time
+* describe the basic model: predict state, compare to observation, update state
+* why the pf is well suited to buses (multimodality, etc)
+* how the pf works (very, very simply) -> details in section 2.3
+* added bonus: easy likelihood (Elliott & Lumley)
+* another bonus: each particle records its travel time along road of known length -> average speed *distribution* (according to particles)
+* can't really do this with, say, a Kalman filter
+
+### Real-time network state model
+
+* another recursive state estimation problem
+* this time a Gaussian makes sense (single mode, usually)
+* KF: proven way of estimating network state (Jeong? and some others ...?)
+* much simpler estimation procedure: just some matrix calculations
+* input: noisy 'observations' of vehicle speeds
+* complication: vehicles travel with different speeds, between-vehicle variance, but want to estimate this *separately* from uncertainty of the mean
+* B ~ Normal(beta, (Delta*q)^2 + Psi) (assume independence)
+* these estimated from historical data (see section 3)
+* since we're allowing multiple vehicles to travel simultaneously, Elliott & Lumley used Information filter - becomes a simple addition problem (add new information to current)
+
+## Modelling uncertainty at bus stops
+
+## Prediction using a particle filter
+
+* @citet:Hans_2015; paper
+* light details of how it works: how 'state' is pepresented
+
+## OLD
+
+## Recursive Bayes and particle filtering - skip?
 
